---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis


# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
library(rstatix)

source("functions.R")

# Set path to save plots
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# ---------------------------------------------------------------

# Load data (Kfoury)

```{r}

ds_name <- "Kfoury_2021_34719426"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")

seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))

seurat@misc$label_col <- "Status"
seurat@misc$low_res_ct_col <- "cells"
seurat@misc$hi_res_ct_col <- "cells"


seurat <- seurat |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors() #|>
  #RunUMAP()

labels <- get_labels(seurat, seurat@misc$label_col)
```


# Clustering

```{r}
seurat <- FindClusters_multi(seurat)
```



# Plot bio separation

```{r}
res_broad <- c(0.1, 0.5, 1, 2, 5, 10, 20, 50, 100)
res_broad_2sub <- c() # c(0.1, 0.5)
res_sub <- c() # c(0.1)

# Fractions of highly variable cell types (hvc) to use. Number if int > 1, fraction if < 1
hvc_fracs <- c(5, 10, 20, 0.1, 0.2, 0.5, 0.8, 1)

res_list <- list()


pca_dims <- c(2, 3, 5, 10)

for (p in pca_dims) {
  res_list[[paste0("authors", "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = seurat@misc$hi_res_ct_col, title = "authors", pca_dims = p)
  
  for (rb in res_broad){
    res_list[[paste0(as.character(rb), "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = paste0("RNA_snn_res.", as.character(rb)), title = paste0(as.character(rb), "_PCAdims_", p), pca_dims = p)
    
    if (rb %in% res_broad_2sub){
      for (rs in res_sub) {
        subcluster_col_name <- paste0("subcluster_broad_res", rb, "_sub_res", rs)
        
        res_list[[paste0(subcluster_col_name, "_PCAdims_", p)]] <- process_coda_fig(
          seurat, labels = labels,
          ct_col = subcluster_col_name,
          title = paste0(subcluster_col_name, "_PCAdims_", p), pca_dims = p)
      }
    }
  }
}

p <- NULL
res_list[[paste0("authors", "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = seurat@misc$hi_res_ct_col, title = "authors", pca_dims = p)


for (rb in res_broad){
  res_list[[paste0(as.character(rb), "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = paste0("RNA_snn_res.", as.character(rb)), title = paste0(as.character(rb), "_PCAdims_", p), pca_dims = p)
  
  if (rb %in% res_broad_2sub){
    for (rs in res_sub) {
      subcluster_col_name <- paste0("subcluster_broad_res", rb, "_sub_res", rs)
      
      res_list[[paste0(subcluster_col_name, "_PCAdims_", p)]] <- process_coda_fig(
        seurat, labels = labels,
        ct_col = subcluster_col_name,
        title = paste0(subcluster_col_name, "_PCAdims_", p), pca_dims = p)
    }
  }
  
  df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = paste0("RNA_snn_res.", as.character(rb))) %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()
  df_var <- get_ct_var(df, show_plot = FALSE)
  df_var <- df_var[order(df_var$Variance, decreasing = TRUE), ]
  
  for (hvcf in hvc_fracs) {
    if (hvcf > 1) {
      nhvc <- hvcf
    } else if (hvcf <= 1) {
      nhvc <- round(nrow(df_var) * hvcf)
    }
    if (nhvc > ncol(df) | nhvc == 1) {next}
    
    feat_mat <- df[, df_var$celltype[1:nhvc]]
    
    res_list[[paste0(as.character(rb), "nhvc", nhvc, "_PCAdims_", p)]] <- process_coda_fig(feat_mat = feat_mat, labels = labels, title = paste0(as.character(rb), "nhvc", nhvc, "_PCAdims_", p), pca_dims = NULL)
  }
}

```

# Plot results

## Save and flatten the nested list into a DataFrame

```{r}
df_results <- imap_dfr(res_list, ~ {
  method_name <- .y
  method_content <- .x # This is the list for method1, method2, etc.

  # Extract n_features here, before the inner loop
  n_features <- NA_integer_ # Initialize with NA_integer_ (for numeric NA)
  if ("feat_mat" %in% names(method_content)) {
    n_features <- ncol(method_content$feat_mat)
  }
  
  if ("scores" %in% names(.x)) {
    imap_dfr(.x$scores, ~ {
      tibble(
        method = method_name,
        score = .y,
        value = .x,
        n_features = n_features
      )
    })
  } else {
    tibble() # Return an empty tibble if "scores" is not found
  }
})



```



## Plot results

```{r fig.height=6, fig.width=10}
library(scales)

df1 <- df_results %>%
  filter(score == "cluster_score",
         !grepl("nhvc", method)) %>%
  mutate(PCA = ifelse(endsWith(method, "_"), "raw", "PCA")) %>%
  separate(method, into = c("label_col", "dummy"), sep = "_", extra = "merge", remove = FALSE) %>%
  separate(method, into = c("method1", "method2"), sep = "_", extra = "merge") %>%
  unite(new_column, method1, "PCA", sep = "_")
df1$new_column[grepl("raw", df1$new_column)] <- "raw"
levs <- str_sort(as.character(unique(df1$new_column)), numeric = TRUE)
df1$new_column <- factor(df1$new_column, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = new_column, linetype = score), alpha = 1, linewidth = 1.5) +
    geom_text(
    data = . %>% filter(grepl("raw", new_column)), # Filter the data here
    aes(y = value, label = label_col, color = new_column), # Map label and color
    vjust = -0.8, # Adjust vertical position of labels
    hjust = 0.5,  # Adjust horizontal position
    size = 3.5,   # Label font size
    show.legend = FALSE # Don't add a redundant legend for the labels
  ) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Cluster score")



df1 <- df_results %>%
  filter(score == "mod_score",
         !grepl("nhvc", method)) %>%
  mutate(PCA = ifelse(endsWith(method, "_"), "raw", "PCA")) %>%
  separate(method, into = c("label_col", "dummy"), sep = "_", extra = "merge", remove = FALSE) %>%
  separate(method, into = c("method1", "method2"), sep = "_", extra = "merge") %>%
  unite(new_column, method1, "PCA", sep = "_")
df1$new_column[grepl("raw", df1$new_column)] <- "raw"
levs <- str_sort(as.character(unique(df1$new_column)), numeric = TRUE)
df1$new_column <- factor(df1$new_column, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = new_column, linetype = score), alpha = 1, linewidth = 1.5) +
    geom_text(
    data = . %>% filter(grepl("raw", new_column)), # Filter the data here
    aes(y = value, label = label_col, color = new_column), # Map label and color
    vjust = -0.8, # Adjust vertical position of labels
    hjust = 0.5,  # Adjust horizontal position
    size = 3.5,   # Label font size
    show.legend = FALSE # Don't add a redundant legend for the labels
  ) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Modularity score")



df1 <- df_results %>%
  filter(score == "sil_score",
         !grepl("nhvc", method)) %>%
  mutate(PCA = ifelse(endsWith(method, "_"), "raw", "PCA")) %>%
  separate(method, into = c("label_col", "dummy"), sep = "_", extra = "merge", remove = FALSE) %>%
  separate(method, into = c("method1", "method2"), sep = "_", extra = "merge") %>%
  unite(new_column, method1, "PCA", sep = "_")
df1$new_column[grepl("raw", df1$new_column)] <- "raw"
levs <- str_sort(as.character(unique(df1$new_column)), numeric = TRUE)
df1$new_column <- factor(df1$new_column, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = new_column, linetype = score), alpha = 1, linewidth = 1.5) +
    geom_text(
    data = . %>% filter(grepl("raw", new_column)), # Filter the data here
    aes(y = value, label = label_col, color = new_column), # Map label and color
    vjust = -0.8, # Adjust vertical position of labels
    hjust = 0.5,  # Adjust horizontal position
    size = 3.5,   # Label font size
    show.legend = FALSE # Don't add a redundant legend for the labels
  ) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Silhouette score")



df1 <- df_results %>%
  filter(score == "cluster_score",
         grepl("nhvc", method)) %>%
  mutate(PCA = !endsWith(method, "_")) %>%
  separate(method, into = c("method1", "method2"), sep = "nhvc", extra = "merge")
levs <- str_sort(as.character(unique(df1$method1)), numeric = TRUE)
df1$method1 <- factor(df1$method1, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = method1, linetype = score), alpha = 1, linewidth = 1.5) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Based on highly variable cell types")



df1 <- df_results %>%
  filter(score == "mod_score",
         grepl("nhvc", method)) %>%
  mutate(PCA = !endsWith(method, "_")) %>%
  separate(method, into = c("method1", "method2"), sep = "nhvc", extra = "merge")
levs <- str_sort(as.character(unique(df1$method1)), numeric = TRUE)
df1$method1 <- factor(df1$method1, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = method1, linetype = score), alpha = 1, linewidth = 1.5) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Based on highly variable cell types")
```



# ---------------------------------------------------------------

# Plot variance and correlations

```{r}
ct_col <- "RNA_snn_res.0.5"
# ct_col <- "cells"

df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = ct_col) %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()

df_var <- get_ct_var(df, plot_title = ct_col)
```



```{r}
cor_matrix <- cor(df)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix)
```



```{r fig.height=12, fig.width=18}
# 1. Create the confusion matrix
conf_matrix <- table(seurat$cells, seurat$subcluster_annotation)

# 2. Convert to a data frame for ggplot2
conf_df <- as.data.frame(conf_matrix)

# 3. Rename columns for clarity (optional but good practice)
colnames(conf_df) <- c("Author_Annotation", "Seurat_Cluster", "Count")

# 4. Create the heatmap plot using ggplot2
ggplot(conf_df, aes(x = Seurat_Cluster, y = Author_Annotation, fill = Count)) +
  geom_tile(color = "black") + # Adds borders to tiles
  geom_text(aes(label = Count), color = "white", size = 3) + # Add counts as text on tiles
  scale_fill_gradient(low = "white", high = "steelblue") + # Adjust color scale
  labs(title = "Confusion Matrix: Author Annotation vs. Seurat Clustering (res=0.5)",
       x = "Seurat Clusters (Resolution 0.5)",
       y = "Author's Annotation",
       fill = "Number of Cells") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels
        panel.grid.major = element_blank(), # Remove major grid lines
        panel.grid.minor = element_blank(), # Remove minor grid lines
        plot.title = element_text(hjust = 0.5)) # Center plot title
```



# ---------------------------------------------------------------

# Load data (Stephenson)

```{r}

ds_name <- "MitchelJ_2023_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")

seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))

seurat <- subset(seurat, subset = Site == "Ncl")
seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))
seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)

seurat$sample_id <- as.character(seurat$sample_id)
samples_exclude <- c("BGCV10_CV0198", "MH8919230")
samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% samples_exclude])
seurat <- subset(seurat, subset = sample_id %in% samples_subset)

seurat$Sample <- seurat$sample_id

seurat@misc$label_col <- "Status"
seurat@misc$low_res_ct_col <- "initial_clustering"
seurat@misc$hi_res_ct_col <- "full_clustering"

seurat <- seurat |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors() # |>
  # RunUMAP()

labels <- get_labels(seurat, seurat@misc$label_col)
```


# Clustering

```{r}
seurat <- FindClusters_multi(seurat)
```


# Plot bio separation

```{r}
res_broad <- c(0.1, 0.5, 1, 2, 5, 10, 20, 50, 100)
res_broad_2sub <- c() # c(0.1, 0.5)
res_sub <- c() # c(0.1)

# Fractions of highly variable cell types (hvc) to use. Number if int > 1, fraction if < 1
hvc_fracs <- c(5, 10, 20, 0.1, 0.2, 0.5, 0.8, 1)

res_list <- list()


pca_dims <- c(2, 3, 5, 10)

for (p in pca_dims) {
  res_list[[paste0("authors", "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = seurat@misc$hi_res_ct_col, title = "authors", pca_dims = p)
  
  for (rb in res_broad){
    res_list[[paste0(as.character(rb), "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = paste0("RNA_snn_res.", as.character(rb)), title = paste0(as.character(rb), "_PCAdims_", p), pca_dims = p)
    
    if (rb %in% res_broad_2sub){
      for (rs in res_sub) {
        subcluster_col_name <- paste0("subcluster_broad_res", rb, "_sub_res", rs)
        
        res_list[[paste0(subcluster_col_name, "_PCAdims_", p)]] <- process_coda_fig(
          seurat, labels = labels,
          ct_col = subcluster_col_name,
          title = paste0(subcluster_col_name, "_PCAdims_", p), pca_dims = p)
      }
    }
  }
}

p <- NULL
res_list[[paste0("authors", "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = seurat@misc$hi_res_ct_col, title = "authors", pca_dims = p)


for (rb in res_broad){
  res_list[[paste0(as.character(rb), "_PCAdims_", p)]] <- process_coda_fig(seurat, labels = labels, ct_col = paste0("RNA_snn_res.", as.character(rb)), title = paste0(as.character(rb), "_PCAdims_", p), pca_dims = p)
  
  if (rb %in% res_broad_2sub){
    for (rs in res_sub) {
      subcluster_col_name <- paste0("subcluster_broad_res", rb, "_sub_res", rs)
      
      res_list[[paste0(subcluster_col_name, "_PCAdims_", p)]] <- process_coda_fig(
        seurat, labels = labels,
        ct_col = subcluster_col_name,
        title = paste0(subcluster_col_name, "_PCAdims_", p), pca_dims = p)
    }
  }
  
  df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = paste0("RNA_snn_res.", as.character(rb))) %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()
  df_var <- get_ct_var(df, show_plot = FALSE)
  df_var <- df_var[order(df_var$Variance, decreasing = TRUE), ]
  
  for (hvcf in hvc_fracs) {
    if (hvcf > 1) {
      nhvc <- hvcf
    } else if (hvcf <= 1) {
      nhvc <- round(nrow(df_var) * hvcf)
    }
    if (nhvc > ncol(df) | nhvc == 1) {next}
    
    feat_mat <- df[, df_var$celltype[1:nhvc]]
    
    res_list[[paste0(as.character(rb), "nhvc", nhvc, "_PCAdims_", p)]] <- process_coda_fig(feat_mat = feat_mat, labels = labels, title = paste0(as.character(rb), "nhvc", nhvc, "_PCAdims_", p), pca_dims = NULL)
  }
}

```

```{r}
df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = "RNA_snn_res.100") %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()

process_coda_fig(feat_mat = df, labels = get_labels(seurat, "Status_on_day_collection_summary"), title = "", pca_dims = NULL)

plot_pca(df, get_labels(seurat, "Status_on_day_collection_summary"), plotly_3d = TRUE)
```



# Plot results

## Save and flatten the nested list into a DataFrame

```{r}
df_results <- imap_dfr(res_list, ~ {
  method_name <- .y
  method_content <- .x # This is the list for method1, method2, etc.

  # Extract n_features here, before the inner loop
  n_features <- NA_integer_ # Initialize with NA_integer_ (for numeric NA)
  if ("feat_mat" %in% names(method_content)) {
    n_features <- ncol(method_content$feat_mat)
  }
  
  if ("scores" %in% names(.x)) {
    imap_dfr(.x$scores, ~ {
      tibble(
        method = method_name,
        score = .y,
        value = .x,
        n_features = n_features
      )
    })
  } else {
    tibble() # Return an empty tibble if "scores" is not found
  }
})
```



## Plot results

```{r fig.height=6, fig.width=10}
library(scales)

df1 <- df_results %>%
  filter(score == "cluster_score",
         !grepl("nhvc", method)) %>%
  mutate(PCA = ifelse(endsWith(method, "_"), "raw", "PCA")) %>%
  separate(method, into = c("label_col", "dummy"), sep = "_", extra = "merge", remove = FALSE) %>%
  separate(method, into = c("method1", "method2"), sep = "_", extra = "merge") %>%
  unite(new_column, method1, "PCA", sep = "_")
df1$new_column[grepl("raw", df1$new_column)] <- "raw"
levs <- str_sort(as.character(unique(df1$new_column)), numeric = TRUE)
df1$new_column <- factor(df1$new_column, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = new_column, linetype = score), alpha = 1, linewidth = 1.5) +
    geom_text(
    data = . %>% filter(grepl("raw", new_column)), # Filter the data here
    aes(y = value, label = label_col, color = new_column), # Map label and color
    vjust = -0.8, # Adjust vertical position of labels
    hjust = 0.5,  # Adjust horizontal position
    size = 3.5,   # Label font size
    show.legend = FALSE # Don't add a redundant legend for the labels
  ) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Cluster score")



df1 <- df_results %>%
  filter(score == "mod_score",
         !grepl("nhvc", method)) %>%
  mutate(PCA = ifelse(endsWith(method, "_"), "raw", "PCA")) %>%
  separate(method, into = c("label_col", "dummy"), sep = "_", extra = "merge", remove = FALSE) %>%
  separate(method, into = c("method1", "method2"), sep = "_", extra = "merge") %>%
  unite(new_column, method1, "PCA", sep = "_")
df1$new_column[grepl("raw", df1$new_column)] <- "raw"
levs <- str_sort(as.character(unique(df1$new_column)), numeric = TRUE)
df1$new_column <- factor(df1$new_column, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = new_column, linetype = score), alpha = 1, linewidth = 1.5) +
    geom_text(
    data = . %>% filter(grepl("raw", new_column)), # Filter the data here
    aes(y = value, label = label_col, color = new_column), # Map label and color
    vjust = -0.8, # Adjust vertical position of labels
    hjust = 0.5,  # Adjust horizontal position
    size = 3.5,   # Label font size
    show.legend = FALSE # Don't add a redundant legend for the labels
  ) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Modularity score")



df1 <- df_results %>%
  filter(score == "sil_score",
         !grepl("nhvc", method)) %>%
  mutate(PCA = ifelse(endsWith(method, "_"), "raw", "PCA")) %>%
  separate(method, into = c("label_col", "dummy"), sep = "_", extra = "merge", remove = FALSE) %>%
  separate(method, into = c("method1", "method2"), sep = "_", extra = "merge") %>%
  unite(new_column, method1, "PCA", sep = "_")
df1$new_column[grepl("raw", df1$new_column)] <- "raw"
levs <- str_sort(as.character(unique(df1$new_column)), numeric = TRUE)
df1$new_column <- factor(df1$new_column, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = new_column, linetype = score), alpha = 1, linewidth = 1.5) +
    geom_text(
    data = . %>% filter(grepl("raw", new_column)), # Filter the data here
    aes(y = value, label = label_col, color = new_column), # Map label and color
    vjust = -0.8, # Adjust vertical position of labels
    hjust = 0.5,  # Adjust horizontal position
    size = 3.5,   # Label font size
    show.legend = FALSE # Don't add a redundant legend for the labels
  ) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Silhouette score")



df1 <- df_results %>%
  filter(score == "cluster_score",
         grepl("nhvc", method)) %>%
  mutate(PCA = !endsWith(method, "_")) %>%
  separate(method, into = c("method1", "method2"), sep = "nhvc", extra = "merge")
levs <- str_sort(as.character(unique(df1$method1)), numeric = TRUE)
df1$method1 <- factor(df1$method1, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = method1, linetype = score), alpha = 1, linewidth = 1.5) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Based on highly variable cell types")



df1 <- df_results %>%
  filter(score == "mod_score",
         grepl("nhvc", method)) %>%
  mutate(PCA = !endsWith(method, "_")) %>%
  separate(method, into = c("method1", "method2"), sep = "nhvc", extra = "merge")
levs <- str_sort(as.character(unique(df1$method1)), numeric = TRUE)
df1$method1 <- factor(df1$method1, levels = levs)

ggplot(df1, aes(x = n_features)) + # Remove group=score from global aes if you want lines by method1
  geom_line(aes(y = value, color = method1, linetype = score), alpha = 1, linewidth = 1.5) +
  labs(x = "Number of features (log2 scale)", y = "Separation score") + # Adjusted y-axis label to match primary
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y.left = element_text(color = "black"), # Default color for primary axis label
        axis.title.y.right = element_text(color = "darkgrey")) + # Match color for secondary axis label
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n=10),
                     labels = label_number()) + # Changed to 10^exponent format
  # Add a color scale if you mapped method1 to color
  scale_color_discrete(name = "Method 1 Type") +
  ggtitle("Based on highly variable cell types")
```

```{r}
rstudioapi::getActiveDocumentContext()$path
```

# ---------------------------------------------------------------

# Plot variance and correlations

```{r}
# tdf <- as.data.frame.matrix(table(seurat$Sample, seurat$RNA_snn_res.200))

ct_col <- "RNA_snn_res.20"
# ct_col <- "RNA_snn_res.100"

df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = ct_col) %>%
    impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
    calc_perc_df() %>%
    clr()

df_var <- get_ct_var(df, plot_title = ct_col)
plot(sort(df_var$Variance))


df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = ct_col) %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()

df_var <- get_ct_var(df, plot_title = ct_col)
plot(sort(df_var$Variance))


df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = ct_col) %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.01) %>%
    calc_perc_df() %>%
    clr()

df_var <- get_ct_var(df, plot_title = ct_col)
plot(sort(df_var$Variance))


df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = ct_col) %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.001) %>%
    calc_perc_df() %>%
    clr()

df_var <- get_ct_var(df, plot_title = ct_col)
plot(sort(df_var$Variance))
```

```{r}
data_points <- df1$Variance

hist(data_points, breaks = 30)

# Calculate 95% quantile
quantile_95 <- quantile(data_points, 0.95)

# Identify outliers (points larger than 95th percentile)
outliers <- data_points[data_points > quantile_95]

# Print outliers
cat("Outliers (larger than 95th percentile):\n")
print(outliers)


# Plot histogram with outliers
hist(data_points, breaks = 30, col = "lightblue", main = "Histogram with Outliers")
abline(v = quantile_95, col = "red", lwd = 2)  # Add line for 95th percentile
points(outliers, rep(0, length(outliers)), col = "red", pch = 19)  # Add points for outliers



# Create a histogram with density scale
h <- hist(data_points, breaks = 30, plot = FALSE) # First, compute histogram without plotting
hist(data_points,
     freq = FALSE, # Plot probability densities, not frequencies
     breaks = 30,
     main = "Distribution with Outliers",
     xlab = "Value",
     ylab = "Density",
     col = "lightblue",
     border = "white",
     ylim = c(0, max(h$density) * 1.1) # Adjust ylim to make space for points
     )

# Add rug plot (small lines at the bottom indicating individual data points)
# This helps visualize point density but doesn't show outliers distinctly
rug(data_points)

# Add a density curve (kernel density estimate)
lines(density(data_points), col = "blue", lwd = 2)

# Optional: Add a theoretical normal distribution curve for comparison
# assuming the mean and SD of your data
xfit <- seq(min(data_points), max(data_points), length=100)
yfit <- dnorm(xfit, mean = mean(data_points), sd = sd(data_points))
lines(xfit, yfit, col = "red", lty = 2, lwd = 2) # Dashed red line for normal curve

legend("topright",
       legend = c("Kernel Density", "Normal Distribution"),
       col = c("blue", "red"),
       lwd = 2,
       lty = c(1, 2))



data <- df1$Variance

# Calculate mean and standard deviation of the data
# For robust outlier detection, consider using robust measures like median and MAD (Median Absolute Deviation)
# if you suspect many outliers might distort mean/sd.
data_mean <- mean(data)
data_sd <- sd(data)

# Define significance level (e.g., alpha = 0.05 for 95% CI)
alpha <- 0.05
# Critical Z-score for two-tailed test
# For 95% CI (alpha=0.05), we want the Z-score for alpha/2 = 0.025 in each tail
z_critical_lower <- qnorm(alpha / 2, mean = data_mean, sd = data_sd)
z_critical_upper <- qnorm(1 - (alpha / 2), mean = data_mean, sd = data_sd)

# Identify outliers
outliers_values <- data[data < z_critical_lower | data > z_critical_upper]
outliers_indices <- which(data < z_critical_lower | data > z_critical_upper)

cat("Outliers based on", (1-alpha)*100, "% confidence interval (assuming Normal Distribution):\n")
print(outliers_values)
cat("Indices of outliers:", outliers_indices, "\n")

# --- You can also use formal outlier tests if a package is available ---
# For example, the 'outliers' package provides Grubbs' test or Dixon's test.
# install.packages("outliers")
# library(outliers)
# grubbs.test(data) # Tests for a single outlier

# Or use IQR method (often used with box plots)
Q1 <- quantile(data, 0.25)
Q3 <- quantile(data, 0.75)
IQR_val <- Q3 - Q1
iqr_outliers <- data[data < (Q1 - 1.5 * IQR_val) | data > (Q3 + 1.5 * IQR_val)]
cat("\nOutliers based on IQR method:\n")
print(iqr_outliers)



# Re-create the dataframe, adding an 'is_outlier' column
df <- data.frame(Value = data)
df$is_outlier <- ifelse(df$Value < z_critical_lower | df$Value > z_critical_upper,
                        "Outlier", "Normal")

# Plotting with outliers highlighted
ggplot(df, aes(x = Value)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 0.2,
                 fill = "lightblue",
                 color = "white") +
  geom_density(color = "blue", linewidth = 1) +
  geom_rug(aes(color = is_outlier), alpha = 0.7) + # Color rug points based on outlier status
  geom_point(data = filter(df, is_outlier == "Outlier"), # Plot outlier points distinctly
             aes(x = Value, y = -0.005), # Position them slightly below the axis
             color = "red",
             size = 3,
             shape = 8) + # Use a star shape for outliers
  labs(title = paste0("Distribution with Outliers (", (1-alpha)*100, "% CI)"),
       x = "Value",
       y = "Density") +
  scale_color_manual(values = c("Normal" = "darkgreen", "Outlier" = "red")) + # Define colors for rug
  theme_minimal() +
  theme(legend.position = "none") # Hide the rug legend if desired

# Optionally, add lines for the critical thresholds
ggplot(df, aes(x = Value)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 0.2,
                 fill = "lightblue",
                 color = "white") +
  geom_density(color = "blue", linewidth = 1) +
  geom_rug(alpha = 0.3) +
  geom_point(data = filter(df, is_outlier == "Outlier"),
             aes(x = Value, y = 0), # Place outliers on the x-axis for this plot
             color = "red",
             size = 3,
             shape = 8,
             position = position_jitter(height = 0.002)) + # Jitter to avoid overlap
  # Add vertical lines for the critical thresholds
  geom_vline(xintercept = z_critical_lower, linetype = "dashed", color = "purple", linewidth = 1) +
  geom_vline(xintercept = z_critical_upper, linetype = "dashed", color = "purple", linewidth = 1) +
  annotate("text", x = z_critical_lower, y = max(ggplot_build(ggplot(df, aes(x = Value)) + geom_histogram(aes(y = after_stat(density)), binwidth = 5))$data[[1]]$density) * 0.9,
           label = paste0("Lower CI (", round(z_critical_lower, 2), ")"), color = "purple", hjust = 1.1) +
  annotate("text", x = z_critical_upper, y = max(ggplot_build(ggplot(df, aes(x = Value)) + geom_histogram(aes(y = after_stat(density)), binwidth = 5))$data[[1]]$density) * 0.9,
           label = paste0("Upper CI (", round(z_critical_upper, 2), ")"), color = "purple", hjust = -0.1) +
  labs(title = paste0("Distribution with Outliers (", (1-alpha)*100, "% CI)"),
       x = "Value",
       y = "Density") +
  theme_minimal()
```



```{r}
ct_col <- "RNA_snn_res.20"
# ct_col <- "subcluster_broad_res0.1_sub_res0.1"

df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = ct_col) %>%
    impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
    calc_perc_df() %>%
    clr()

df_var <- get_ct_var(df, plot_title = ct_col)

df_var <- df_var[order(df_var$Variance, decreasing = TRUE), ]

feat_mat <- df[, df_var$celltype[1:100]]
# res.pca <- prcomp(feat_mat, center = TRUE, scale. = FALSE)
# feat_mat <- res.pca[["x"]][, 1:3]

plot_pca(feat_mat, labels = get_labels(seurat, "Status"), plotly_3d = FALSE, title = "Test")
plot_pca(feat_mat, labels = get_labels(seurat, "Status"), plotly_3d = FALSE, title = "Test", invisible = c("quali"))
```



```{r}
process_coda_fig(
        seurat, labels = get_labels(seurat, "Status_on_day_collection_summary"),
        ct_col = "RNA_snn_res.20",
        title = paste0("RNA_snn_res.20", "_PCAdims_", 2), pca_dims = 2)
```

```{r}
length(table(seurat$subcluster_broad_res0.1_sub_res0.1))

ct_col <- "subcluster_broad_res0.1_sub_res0.1"
# ct_col <- "RNA_snn_res.100"
# ct_col <- "RNA_snn_res.200"

df <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = ct_col) %>%
    impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
    calc_perc_df() %>%
    clr()

df_var <- get_ct_var(df, plot_title = ct_col)
plot(sort(df_var$Variance))

```



# ---------------------------------------------------------------

# PCA of Stephenson

```{r}
ds_name <- "MitchelJ_2023_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")

seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))

# seurat <- subset(seurat, subset = Site == "Ncl")
# seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))
seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
# seurat <- subset(seurat, subset = Days_from_onset < 10)
# 
# seurat$sample_id <- as.character(seurat$sample_id)
# samples_exclude <- c("BGCV10_CV0198", "MH8919230")
# samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% samples_exclude])
# seurat <- subset(seurat, subset = sample_id %in% samples_subset)

seurat$Sample <- seurat$sample_id

seurat@misc$label_col <- "Status"
seurat@misc$low_res_ct_col <- "initial_clustering"
seurat@misc$hi_res_ct_col <- "full_clustering"

dims <- 1:30

seurat <- seurat |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA()
hvg <- seurat@assays[["RNA"]]@var.features

labels <- get_labels(seurat, seurat@misc$label_col)

pb <- get_pb_deseq2(seurat, sample_col = "Sample", hvg = hvg)
```

```{r}
process_coda_fig(seurat, labels = labels, ct_col = seurat@misc$hi_res_ct_col, title = "Status", pca_dims = NULL)
process_coda_fig(seurat, labels = get_labels(seurat, "Site"), ct_col = seurat@misc$hi_res_ct_col, title = "Site", pca_dims = NULL)

p <- plot_pca(pb, labels, plotly_3d = TRUE)
p <- plot_pca(pb, get_labels(seurat, "Site"), plotly_3d = TRUE)
```
