---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
source("src/R/load_all_functions.R")
source("Processed_dataset_metadata.R")

# Set paths
path_nas <- file.path(
  "",
  "Volumes",
  "Shared",
  "DataCollections",
  "Standardized_SingleCell_Datasets"
)
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"

set.seed(123)

`%notin%` <- Negate(`%in%`)
```



# ---------------------------------------------------------------

# Batch effect

## Joanito dataset

## Load/preprocess

```{r}
seurat <- readRDS(file.path(
  path_data,
  "JoaI_2022_35773407_Nofilt_whole_ECODAprocessed.rds"
))

seurat@meta.data$cell.type_new <- seurat@meta.data$cell.type
iCMS <- !is.na(seurat@meta.data$iCMS)
seurat@meta.data$cell.type_new[iCMS] <- paste0(
  seurat@meta.data$cell.type[iCMS],
  "_",
  ifelse(seurat@meta.data$iCMS[iCMS] == "Normal", "Normal", "Cancer")
)

meta <- seurat@meta.data %>%
  group_by(sample.ID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample.ID
meta$sample.origin <- gsub("-2", "", meta$sample.origin)
meta$seqtec <- ifelse(
  meta$dataset %in% c("CRC-SG1", "KUL5"),
  "5' seq",
  "3' seq"
)

bio_col <- "sample.origin"
batch_col <- "seqtec"

plot_list <- list()
```

## Plot PCA

### ECODA

```{r}
dist_mat <- as.data.frame.matrix(table(
  seurat@meta.data$sample.ID,
  seurat@meta.data$cell.type_new
)) %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 0.5
  ) %>%
  clr() %>%
  dist()

plot_list[["p_ecoda_bio"]] <- plot_mds(
  dist_mat,
  labels = meta[[bio_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T,
  title = "ECODA - Biological groups"
)
print(plot_list[["p_ecoda_bio"]])


plot_list[["p_ecoda_batch"]] <- plot_mds(
  dist_mat,
  labels = meta[[batch_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T,
  title = "ECODA - Batch groups"
)
print(plot_list[["p_ecoda_batch"]])
```

### PB

#### Without blacklist

```{r}
pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "none")

plot_list[["p_pb_bio"]] <- plot_mds(
  dist(pb_norm),
  labels = meta[[bio_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T,
  title = "Pseudobulk - Biological groups"
)
print(plot_list[["p_pb_bio"]])

plot_list[["p_pb_batch"]] <- plot_mds(
  dist(pb_norm),
  labels = meta[[batch_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T,
  title = "Pseudobulk - Batch groups"
)
print(plot_list[["p_pb_batch"]])
```

# Figure 5: Combine all plots

```{r fig.height=6.5, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = 'A') #&
# theme(
#   # legend.position='bottom',
#   plot.tag = element_text(size = 16, face = "bold")
# )
ggsave(
  file.path(path_plots, "Figure_5_batcheffect_Joanito_PCA.pdf"),
  width = 10,
  height = 6.5
)
```

## Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$sample.ID))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste(
  "Number of cells per sample:",
  round(mean(table((seurat@meta.data$sample.ID)))),
  "+-",
  round(sd(table((seurat@meta.data$sample.ID))))
))
print(paste(
  "Number of cell types:",
  length(unique(seurat@meta.data$cell.type))
))
```

# ---------------------------------------------------------------

# Stephenson dataset

## Load/preprocess

```{r}
p <- file.path(
  path_nas,
  "StephensonE_2021_39313646",
  "output",
  "StephensonE_2021_33879890_preprocessed.rds"
)

seurat <- readRDS(p)

seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))

seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset[seurat$Days_from_onset == "nan"] <- 999
seurat$Days_from_onset[seurat$Days_from_onset == "Not_known"] <- 999
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)


meta <- seurat@meta.data %>%
  group_by(Sample) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$Sample

pb <- get_pb(seurat, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

bio_col <- "Status"
batch_col <- "Site"


ct_remove <- c(
  "Platelets",
  "RBC",
  "C1_CD16_mono",
  "CD14_mono",
  "CD16_mono"
)
seurat_sub <- subset(seurat, subset = full_clustering %notin% ct_remove)
pb <- get_pb(seurat_sub, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm_sub <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

plot_list <- list()
```

## Plot PCA

### ECODA

```{r}
dist_mat <- as.data.frame.matrix(table(
  seurat@meta.data$Sample,
  seurat@meta.data$full_clustering
)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 0.5
  ) %>%
  clr() %>%
  dist()

plot_mds(
  dist_mat,
  labels = meta[[bio_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)


plot_mds(
  dist_mat,
  labels = meta[[batch_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
```

#### Subset to remove batch effect cell types

```{r}
dist_mat <- as.data.frame.matrix(table(
  seurat_sub@meta.data$Sample,
  seurat_sub@meta.data$full_clustering
)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 0.5
  ) %>%
  clr() %>%
  dist()

plot_list[["p_ecoda_bio_sub"]] <- plot_mds(
  dist_mat,
  labels = meta[[bio_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_ecoda_bio_sub"]])
ggsave(
  file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_bio.pdf"),
  width = 4,
  height = 3
)


plot_list[["p_ecoda_batch_sub"]] <- plot_mds(
  dist_mat,
  labels = meta[[batch_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_ecoda_batch_sub"]])
ggsave(
  file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_batch.pdf"),
  width = 4,
  height = 3
)
```

### PB

```{r}
plot_mds(
  dist(pb_norm),
  labels = meta[[bio_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)

plot_mds(
  dist(pb_norm),
  labels = meta[[batch_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
```

#### Subset to remove batch effect cell types

```{r}
plot_list[["p_pb_bio_sub"]] <- plot_mds(
  dist(pb_norm_sub),
  labels = meta[[bio_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_pb_bio_sub"]])
ggsave(
  file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_bio.pdf"),
  width = 4,
  height = 3
)

plot_list[["p_pb_batch_sub"]] <- plot_mds(
  dist(pb_norm_sub),
  labels = meta[[batch_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_pb_batch_sub"]])
ggsave(
  file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_batch.pdf"),
  width = 4,
  height = 3
)
```

#### With blacklist

```{r}
pb_norm_sub <- get_pb_deseq2(
  seurat_sub,
  sample_col = "Sample",
  black_list = "default"
)

plot_mds(
  dist(pb_norm_sub),
  labels = meta[[bio_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
#        width = 4,
#        height = 3)

plot_mds(
  dist(pb_norm_sub),
  labels = meta[[batch_col]],
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
#        width = 4,
#        height = 3)
```

# Supp fig 22: Combine all plots

```{r fig.height=15, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(
  file.path(path_plots, "Supp_fig_22_batcheffect_Stephenson_PCA_combined.pdf"),
  width = 10,
  height = 8
)
```


# ---------------------------------------------------------------

# Combined PBMC (Stephenson, GongSharma, Zhu)

## Aggregate pseudobulk

### Stephenson

```{r}
ds <- "Stephenson"
seurat <- readRDS(file.path(
  path_data,
  paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)
))

pb_stephenson <- get_pb(seurat, sample_col = "Sample")
metadata_stephenson <- get_metadata(seurat, sample_col = "Sample")

feat_mat_stephenson <- as.data.frame.matrix(table(
  seurat@meta.data$Sample,
  seurat@meta.data$layer2
)) %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 0.5
  ) %>%
  clr()
```

#### Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$Sample))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste(
  "Number of cells per sample:",
  round(mean(table((seurat@meta.data$Sample)))),
  "+-",
  round(sd(table((seurat@meta.data$Sample))))
))
print(paste("Number of cell types:", length(unique(seurat@meta.data$layer2))))
```


### GongSharma

```{r}
ds <- "Gongsharma_cmv_young_males"
seurat <- readRDS(file.path(
  path_data,
  paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)
))
set.seed(123)
subsample <- sample(
  unique(seurat$Sample),
  size = 15,
  replace = FALSE,
  prob = NULL
)
seurat <- subset(seurat, subset = Sample %in% subsample)

pb_gs <- get_pb(seurat, sample_col = "Sample")
metadata_gs <- get_metadata(seurat, sample_col = "Sample")

feat_mat_gs <- as.data.frame.matrix(table(
  seurat@meta.data$Sample,
  seurat@meta.data$layer2
)) %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 0.5
  ) %>%
  clr()
```

#### Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$Sample))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste(
  "Number of cells per sample:",
  round(mean(table((seurat@meta.data$Sample)))),
  "+-",
  round(sd(table((seurat@meta.data$Sample))))
))
print(paste("Number of cell types:", length(unique(seurat@meta.data$layer2))))
```


### Zhu

```{r}
seurat <- readRDS(
  "/Users/christianhalter/Desktop/ECODA_paper/data/ZhuH_2023_37379396whole_ECODAprocessed.rds"
)

sample_col <- "Sample"

feat_mat_zhu <- get_ct_comp_df_seurat(
  seurat,
  sample_col = sample_col,
  ct_col = "layer2"
) %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 0.5
  ) %>%
  clr()

pb_zhu <- get_pb(seurat, sample_col = sample_col, hvg = NULL)
metadata_zhu <- get_metadata(seurat, sample_col = sample_col)
```

#### Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$Sample))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste(
  "Number of cells per sample:",
  round(mean(table((seurat@meta.data$Sample)))),
  "+-",
  round(sd(table((seurat@meta.data$Sample))))
))
print(paste("Number of cell types:", length(unique(seurat@meta.data$layer2))))
```


## Combine pb matrices

```{r}
# 1. Put your three matrices into a list
matrix_list <- list(pb_stephenson, pb_gs, pb_zhu)

# 2. Find the intersection of all row names
common_genes <- Reduce(intersect, lapply(matrix_list, rownames))

# 3. Subset and combine in one step
# This ensures rows are in the exact same order before binding
combined_pseudobulk <- do.call(
  cbind,
  lapply(matrix_list, function(x) {
    x[common_genes, , drop = FALSE]
  })
)
print(dim(combined_pseudobulk))

metadata_stephenson$ds <- "Stephenson"
metadata_stephenson$cond <- metadata_stephenson$Status
metadata_gs$ds <- "GongSharma"
metadata_gs$cond <- "Healthy"
metadata_zhu$ds <- "Zhu"
metadata_zhu$cond <- "Healthy"

metadata_list <- list(
  metadata_stephenson,
  metadata_gs,
  metadata_zhu
)
combined_metadata <- do.call(
  rbind,
  lapply(metadata_list, function(x) {
    x
  })
)

pb_norm <- t(DESeq2.normalize(
  combined_pseudobulk,
  metadata = combined_metadata,
  n_hvg = 2000
))

feat_mat_list <- list(
  feat_mat_stephenson,
  feat_mat_gs,
  feat_mat_zhu
)
# 1. Find columns common to all dataframes
common_cols <- Reduce(intersect, lapply(feat_mat_list, colnames))

# 2. Subset each and rbind
combined_feat_mat <- do.call(
  rbind,
  lapply(feat_mat_list, function(x) {
    x[, common_cols, drop = FALSE]
  })
)
```


# Number of cell types (combined)

```{r}
print(paste("Number of cell types:", ncol(combined_feat_mat)))
```


# ECODA

```{r}
plot_list <- list()

plot_list[["p_ecoda_bio"]] <- plot_mds(
  dist(combined_feat_mat),
  labels = combined_metadata$cond,
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_ecoda_bio"]])
ggsave(
  file.path(path_plots, "batcheffect_combined_PCA_ECODA_bio.pdf"),
  width = 4,
  height = 4
)


plot_list[["p_ecoda_batch"]] <- plot_mds(
  dist(combined_feat_mat),
  labels = combined_metadata$ds,
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_ecoda_batch"]])
ggsave(
  file.path(path_plots, "batcheffect_combined_PCA_ECODA_batch.pdf"),
  width = 4,
  height = 4
)
```

# PB

```{r}
plot_list[["p_pb_bio"]] <- plot_mds(
  dist(pb_norm),
  labels = combined_metadata$cond,
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_pb_bio"]])
ggsave(
  file.path(path_plots, "batcheffect_combined_PCA_PB_bio.pdf"),
  width = 4,
  height = 3
)

plot_list[["p_pb_batch"]] <- plot_mds(
  dist(pb_norm),
  labels = combined_metadata$ds,
  cluster_score = F,
  mod_score = F,
  anosim_score = T
)
print(plot_list[["p_pb_batch"]])
ggsave(
  file.path(path_plots, "batcheffect_combined_PCA_PB_batch.pdf"),
  width = 4,
  height = 3
)
```



# Supp fig 22: Combine all plots

```{r fig.height=6.5, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(
  file.path(path_plots, "Supp_fig_22_batcheffect_multistudy.pdf"),
  width = 10,
  height = 6.5
)
```

