---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
source("functions.R")
source("Processed_dataset_metadata.R")

library(forcats)
library(patchwork)

# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
py_install("mofapy2", envname = "my_env", method="auto")
import("mofapy2")
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
# install.packages("GloScope")
library(GloScope)

# Set paths
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"

set.seed(123)

`%notin%` <- Negate(`%in%`)
```



# ---------------------------------------------------------------

# Combined PBMC (Stephenson, GongSharma, Zhu)

## Aggregate pseudobulk

### Stephenson

```{r}
ds <- "Stephenson"
seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))

pb_stephenson <- get_pb(seurat, sample_col = "Sample")
metadata_stephenson <- get_metadata(seurat, sample_col = "Sample")

feat_mat_stephenson <- as.data.frame.matrix(table(seurat@meta.data$Sample, seurat@meta.data$layer2)) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 0.5) %>%
  clr()
```

#### Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$Sample))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste("Number of cells per sample:", round(mean(table((seurat@meta.data$Sample)))), "+-", round(sd(table((seurat@meta.data$Sample))))))
print(paste("Number of cell types:", length(unique(seurat@meta.data$layer2))))
```


### GongSharma

```{r}
ds <- "Gongsharma_cmv_young_males"
seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
set.seed(123)
subsample <- sample(unique(seurat$Sample), size = 15, replace = FALSE, prob = NULL)
seurat <- subset(seurat, subset = Sample %in% subsample)

pb_gs <- get_pb(seurat, sample_col = "Sample")
metadata_gs <- get_metadata(seurat, sample_col = "Sample")

feat_mat_gs <- as.data.frame.matrix(table(seurat@meta.data$Sample, seurat@meta.data$layer2)) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 0.5) %>%
  clr()
```

#### Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$Sample))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste("Number of cells per sample:", round(mean(table((seurat@meta.data$Sample)))), "+-", round(sd(table((seurat@meta.data$Sample))))))
print(paste("Number of cell types:", length(unique(seurat@meta.data$layer2))))
```


### Zhu

```{r}
seurat <- readRDS("/Users/christianhalter/Desktop/ECODA_paper/data/ZhuH_2023_37379396whole_ECODAprocessed.rds")

sample_col <- "Sample"

feat_mat_zhu <- get_ct_comp_df_seurat(seurat, sample_col = sample_col, ct_col = "layer2") %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 0.5) %>%
  clr()

pb_zhu <- get_pb(seurat, sample_col = sample_col, hvg = NULL)
metadata_zhu <- get_metadata(seurat, sample_col = sample_col)
```

#### Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$Sample))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste("Number of cells per sample:", round(mean(table((seurat@meta.data$Sample)))), "+-", round(sd(table((seurat@meta.data$Sample))))))
print(paste("Number of cell types:", length(unique(seurat@meta.data$layer2))))
```


## Combine pb matrices

```{r}
# 1. Put your three matrices into a list
matrix_list <- list(pb_stephenson, pb_gs, pb_zhu)

# 2. Find the intersection of all row names
common_genes <- Reduce(intersect, lapply(matrix_list, rownames))

# 3. Subset and combine in one step
# This ensures rows are in the exact same order before binding
combined_pseudobulk <- do.call(cbind, lapply(matrix_list, function(x) {
  x[common_genes, , drop = FALSE]
}))
print(dim(combined_pseudobulk))

metadata_stephenson$ds <- "Stephenson"
metadata_stephenson$cond <- metadata_stephenson$Status
metadata_gs$ds <- "GongSharma"
metadata_gs$cond <- "Healthy"
metadata_zhu$ds <- "Zhu"
metadata_zhu$cond <- "Healthy"

metadata_list <- list(
  metadata_stephenson,
  metadata_gs,
  metadata_zhu
)
combined_metadata <- do.call(rbind, lapply(metadata_list, function(x) {
  x
}))

pb_norm <- t(DESeq2.normalize(combined_pseudobulk, metadata = combined_metadata, n_hvg = 2000))

feat_mat_list <- list(
  feat_mat_stephenson,
  feat_mat_gs,
  feat_mat_zhu
)
# 1. Find columns common to all dataframes
common_cols <- Reduce(intersect, lapply(feat_mat_list, colnames))

# 2. Subset each and rbind
combined_feat_mat <- do.call(rbind, lapply(feat_mat_list, function(x) {
  x[, common_cols, drop = FALSE]
}))

```


# Number of cell types (combined)

```{r}
print(paste("Number of cell types:", ncol(combined_feat_mat)))
```


# ECODA

```{r}
plot_list <- list()

plot_list[["p_ecoda_bio"]] <- plot_mds(dist(combined_feat_mat), labels = combined_metadata$cond,
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_ecoda_bio"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_ECODA_bio.pdf"),
       width = 4,
       height = 4)


plot_list[["p_ecoda_batch"]] <- plot_mds(dist(combined_feat_mat), labels = combined_metadata$ds,
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_ecoda_batch"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_ECODA_batch.pdf"),
       width = 4,
       height = 4)
```

# PB

```{r}
plot_list[["p_pb_bio"]] <- plot_mds(dist(pb_norm), labels = combined_metadata$cond,
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_pb_bio"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_PB_bio.pdf"),
       width = 4,
       height = 3)

plot_list[["p_pb_batch"]] <- plot_mds(dist(pb_norm), labels = combined_metadata$ds,
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_pb_batch"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_PB_batch.pdf"),
       width = 4,
       height = 3)
```



# Supp fig 22: Combine all plots

```{r fig.height=6.5, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(file.path(path_plots, "Supp_fig_22_batcheffect_multistudy.pdf"),
       width = 10,
       height = 6.5)
```

