---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
source("functions.R")
source("Processed_dataset_metadata.R")

library(forcats)

# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
py_install("mofapy2", envname = "my_env", method="auto")
import("mofapy2")
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
# install.packages("GloScope")
library(GloScope)

# Set paths
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"

set.seed(123)

`%notin%` <- Negate(`%in%`)
```



# Batch effect in gene expression and cell type composition

# ---------------------------------------------------------------

# Sikkema dataset

File was downloaded and pre-processed with script "Preprocess_sikkema.ipynb"


## Load/preprocess

```{r}
sikkema_processed_file <- file.path(path_data, "Sikkema.h5ad")
# 
# if (!file.exists())
# sikkema_ds = list(
#   file_path = file.path(path_data, "Sikkema.h5ad"),
#   tissue = "Airway_Lung",
#   normal_tissue = TRUE, # Whether the dataset contains cancer cells or only normal cells
#   sample_col = "Sample",
#   meta_cols_keep = c(
#     "Sample", # Sample identifier
#     "tissue", # Main biological grouping
#     "suspension_type", "BMI", "age_or_mean_of_age_range", "dataset", "fresh_or_frozen", "lung_condition", # Additional biological groupings
#     "study", "tissue_dissociation_protocol", "tissue_sampling_method", "assay", "tissue_level_2", "sex", # Additional biological groupings
#     "ann_finest_level", "ann_level_1", "ann_level_2", "ann_level_3", "ann_level_4", "ann_level_5", "original_ann_nonharmonized" # Author annotations
#   )
# )
seurat <- load_h5ad_to_seurat(file_name = sikkema_processed_file)
```


```{r}
tissues <- as.data.frame.matrix(table(seurat@meta.data$study, seurat@meta.data$tissue_level_2))
tissues

diss_prot <- as.data.frame.matrix(table(seurat@meta.data$study, seurat@meta.data$tissue_dissociation_protocol))
diss_prot

# Possible combinations:
# Barbry_Leroy_2020
# Jain_Misharin_2021
# 
# Meyer_2019
# Misharin_Budinger_2018
# Teichmann_Meyer_2019
```


## Plot PCA

### ECODA

```{r}
metadata <- get_metadata(seurat)

df_counts <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = "ann_finest_level")

df_imp <- df_counts %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 1
  )

feat_mat <- clr(df_imp)
```

```{r}
label_cols <- c(
  "tissue",
  # "suspension_type",
  "dataset",
  "BMI",
  "age_or_mean_of_age_range",
  "fresh_or_frozen",
  "lung_condition",
  "study",
  "tissue_dissociation_protocol",
  "tissue_sampling_method",
  "assay",
  "tissue_level_2",
  "sex"
)

for (label_col in label_cols) {
  labels <- get_labels(seurat, label_col = label_col)
  labels <- as.character(labels)
  labels[is.na(labels)] <- "NA"
  p <- plot_pca(feat_mat, labels, title = label_col)
  print(p)
}
```


```{r}
label_cols <- c(
  "tissue",
  "suspension_type",
  "dataset",
  # "BMI",
  "age_or_mean_of_age_range",
  "fresh_or_frozen",
  "lung_condition",
  "study",
  "tissue_dissociation_protocol",
  "tissue_sampling_method",
  "assay",
  "tissue_level_2",
  "sex"
)

for (label_col in label_cols) {
  labels <- meta[[label_col]]
  if (length(unique(labels)) == 1) {
    next
  }
  labels <- as.character(labels)
  labels[is.na(labels)] <- "NA"
  p <- plot_pca(fmat, labels, title = label_col)
  print(p)
}
```


```{r}
samples <- metadata$donor_id[metadata$study %in% c("Barbry_Leroy_2020", "Jain_Misharin_2021")]

meta <- metadata[metadata$donor_id %in% samples, ]
fmat <- feat_mat[rownames(feat_mat) %in% samples, ]

label_cols <- c(
  "tissue",
  "suspension_type",
  "dataset",
  # "BMI",
  "age_or_mean_of_age_range",
  "fresh_or_frozen",
  "lung_condition",
  "study",
  "tissue_dissociation_protocol",
  "tissue_sampling_method",
  "assay",
  "tissue_level_2",
  "sex"
)

for (label_col in label_cols) {
  labels <- meta[[label_col]]
  if (length(unique(labels)) == 1) {
    next
  }
  labels <- as.character(labels)
  labels[is.na(labels)] <- "NA"
  p <- plot_pca(fmat, labels, title = label_col)
  print(p)
}
```


```{r}
samples <- metadata$donor_id[metadata$study %in% c("Meyer_2019", "Misharin_Budinger_2018", "Teichmann_Meyer_2019")]

meta <- metadata[metadata$donor_id %in% samples, ]
fmat <- feat_mat[rownames(feat_mat) %in% samples, ]

label_cols <- c(
  "tissue",
  "dataset",
  # "BMI",
  "age_or_mean_of_age_range",
  "fresh_or_frozen",
  "lung_condition",
  "study",
  "tissue_dissociation_protocol",
  "tissue_sampling_method",
  "assay",
  "tissue_level_2",
  "sex"
)

for (label_col in label_cols) {
  labels <- meta[[label_col]]
  if (length(unique(labels)) == 1) {
    next
  }
  labels <- as.character(labels)
  labels[is.na(labels)] <- "NA"
  p <- plot_pca(fmat, labels, title = label_col)
  print(p)
}
```


### PB

```{r}
pb_norm <- get_pb_deseq2(seurat, sample_col = "Sample", hvg = NULL, n_hvg = 2000)
```

```{r}
label_cols <- c(
  "tissue",
  # "suspension_type",
  "dataset",
  "BMI",
  "age_or_mean_of_age_range",
  "fresh_or_frozen",
  "lung_condition",
  "study",
  "tissue_dissociation_protocol",
  "tissue_sampling_method",
  "assay",
  "tissue_level_2",
  "sex"
)

for (label_col in label_cols) {
  labels <- get_labels(seurat, label_col = label_col)
  labels <- as.character(labels)
  labels[is.na(labels)] <- "NA"
  p <- plot_pca(pb_norm, labels, title = label_col)
  print(p)
}
```


# ---------------------------------------------------------------


# Joanito dataset

## Load/preprocess

```{r}
seurat <- readRDS(file.path(path_data, "JoaI_2022_35773407_Nofilt_whole_ECODAprocessed.rds"))


seurat@meta.data$cell.type_new <- seurat@meta.data$cell.type
iCMS <- !is.na(seurat@meta.data$iCMS)
seurat@meta.data$cell.type_new[iCMS] <- paste0(seurat@meta.data$cell.type[iCMS], "_", ifelse(seurat@meta.data$iCMS[iCMS] == "Normal", "Normal", "Cancer"))

meta <- seurat@meta.data %>%
  group_by(sample.ID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample.ID
meta$sample.origin <- gsub("-2", "", meta$sample.origin)
meta$seqtec <- ifelse(meta$dataset %in% c("CRC-SG1", "KUL5"), "5' seq", "3' seq")

pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "default")

bio_col <- "sample.origin"
batch_col <- "seqtec"

plot_list <- list()
```


## Plot PCA

### ECODA

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[bio_col]])
plot_list[["p_ecoda_bio"]] <- plot_pca(feat_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_bio"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_ECODA_bio.pdf"),
       width = 4,
       height = 4)


avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[batch_col]])
plot_list[["p_ecoda_batch"]] <- plot_pca(feat_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_batch"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_ECODA_batch.pdf"),
       width = 4,
       height = 4)



plot_pca(feat_mat, labels = meta$dataset, title = "Authors annotation")


feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$layer2)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

plot_pca(feat_mat, labels = meta[[bio_col]], title = "HiTME annotation")
plot_pca(feat_mat, labels = meta$dataset, title = "HiTME annotation")

feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$layer_6)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

plot_pca(feat_mat, labels = meta[[bio_col]], title = "scATOMIC annotation")
plot_pca(feat_mat, labels = meta$dataset, title = "scATOMIC annotation")
```


### PB


#### Without blacklist

```{r}
pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "none")

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[bio_col]])
plot_list[["p_pb_bio"]] <- plot_pca(pb_norm, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_bio"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
       width = 4,
       height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[batch_col]])
plot_list[["p_pb_batch"]] <- plot_pca(pb_norm, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_batch"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
       width = 4,
       height = 3)


plot_pca(pb_norm, labels = meta$dataset)
```


#### With blacklist

```{r}
pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "default")

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[bio_col]])
plot_pca(pb_norm, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
#        width = 4,
#        height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[batch_col]])
plot_pca(pb_norm, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
#        width = 4,
#        height = 3)


plot_pca(pb_norm, labels = meta$dataset)
```


### Split by seq technology

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

prime3 <- meta$dataset %in% c("SMC", "KUL3", "CRC-SG2")
prime5 <- meta$dataset %in% c("CRC-SG1", "KUL5")
plot_pca(pb_norm[prime3, ], labels = meta$sample.origin[prime3])
plot_pca(pb_norm[prime5, ], labels = meta$sample.origin[prime5])

plot_pca(feat_mat[prime3, ], labels = meta$sample.origin[prime3])
plot_pca(feat_mat[prime5, ], labels = meta$sample.origin[prime5])
```

### Scores bio vs batch

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

labels <- meta$sample.origin
scores_ecoda_bio <- calc_sep_score(feat_mat, labels)
scores_pb_bio <- calc_sep_score(pb_norm, labels)

labels <- meta$dataset
scores_ecoda_batch <- calc_sep_score(feat_mat, labels)
scores_pb_batch <- calc_sep_score(pb_norm, labels)

print(paste("ECODA: Bio ANOSIM", round(scores_ecoda_bio$anosim_score, 3),
            "/ Batch ANOSIM", round(scores_ecoda_batch$anosim_score, 3),
            "=", round(scores_ecoda_bio$anosim_score / scores_ecoda_batch$anosim_score, 3)))
print(paste("PB: Bio ANOSIM", round(scores_pb_bio$anosim_score, 3),
            "/ Batch ANOSIM", round(scores_pb_batch$anosim_score, 3),
            "=", round(scores_pb_bio$anosim_score / scores_pb_batch$anosim_score, 3)))
```


# Combine all plots

```{r fig.height=6.5, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2)+
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_combined.pdf"),
       width = 10,
       height = 6.5)
```


# ---------------------------------------------------------------



# Stephenson dataset

## Load/preprocess

```{r}
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
p <- file.path(path_nas, "StephensonE_2021_39313646", "output", "StephensonE_2021_33879890_preprocessed.rds")

seurat <- readRDS(p)

seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))

seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset[seurat$Days_from_onset == "nan"] <- 999
seurat$Days_from_onset[seurat$Days_from_onset == "Not_known"] <- 999
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)


meta <- seurat@meta.data %>%
  group_by(Sample) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$Sample

pb <- get_pb(seurat, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

bio_col <- "Status"
batch_col <- "Site"


ct_remove <- c(
  "Platelets",
  "RBC",
  "C1_CD16_mono",
  "CD14_mono",
  "CD16_mono"
)
seurat_sub <- subset(seurat, subset = full_clustering %notin% ct_remove)
pb <- get_pb(seurat_sub, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm_sub <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

plot_list <- list()
```


## Plot PCA

### ECODA

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$Sample, seurat@meta.data$full_clustering)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[bio_col]])
plot_list[["p_ecoda_bio"]] <- plot_pca(feat_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_bio"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_bio.pdf"),
       width = 4,
       height = 4)


avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[batch_col]])
plot_list[["p_ecoda_batch"]] <- plot_pca(feat_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_batch"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_batch.pdf"),
       width = 4,
       height = 4)
```


#### Subset to remove batch effect cell types

```{r}
feat_mat <- as.data.frame.matrix(table(seurat_sub@meta.data$Sample, seurat_sub@meta.data$full_clustering)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[bio_col]])
plot_list[["p_ecoda_bio_sub"]] <- plot_pca(feat_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_bio_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_bio.pdf"),
       width = 4,
       height = 3)


avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[batch_col]])
plot_list[["p_ecoda_batch_sub"]] <- plot_pca(feat_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_batch_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_batch.pdf"),
       width = 4,
       height = 3)
```




### PB

```{r}
avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[bio_col]])
plot_list[["p_pb_bio"]] <- plot_pca(pb_norm, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_bio"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_bio.pdf"),
       width = 4,
       height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[batch_col]])
plot_list[["p_pb_batch"]] <- plot_pca(pb_norm, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_batch"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_batch.pdf"),
       width = 4,
       height = 3)
```

#### Subset to remove batch effect cell types

```{r}
avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[bio_col]])
plot_list[["p_pb_bio_sub"]] <- plot_pca(pb_norm_sub, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_bio_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_bio.pdf"),
       width = 4,
       height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[batch_col]])
plot_list[["p_pb_batch_sub"]] <- plot_pca(pb_norm_sub, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_batch_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_batch.pdf"),
       width = 4,
       height = 3)
```


#### With blacklist

```{r}
pb_norm_sub <- get_pb_deseq2(seurat_sub, sample_col = "Sample", black_list = "default")

avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[bio_col]])
plot_pca(pb_norm_sub, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
#        width = 4,
#        height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[batch_col]])
plot_pca(pb_norm_sub, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
#        width = 4,
#        height = 3)


plot_pca(pb_norm, labels = meta$dataset)
```

# Combine all plots

```{r fig.height=15, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 4)+
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_combined.pdf"),
       width = 10,
       height = 13)
```


# ---------------------------------------------------------------

# Combined PBMC (Stephenson, GongSharma, Zhu)

## Load result_list from MAIN_Analysis.Rmd

```{r}
result_file <- "result_list.rds"
result_list <- readRDS(result_file)
```

```{r}
seurat <- readRDS("/Users/christianhalter/Desktop/ECODA_paper/data/ZhuH_2023_37379396whole_ECODAprocessed.rds")

sample_col <- "Sample"

fmat_zhu <- get_ct_comp_df_seurat(seurat, sample_col = sample_col, ct_col = "layer2") %>%
  impute_zeros(
    clr_zero_impute_method = "counts_all",
    clr_zero_impute_num = 1
  ) %>%
  clr()

pb_zhu <- get_pb(seurat, sample_col = sample_col, hvg = NULL)
metadata <- get_metadata(seurat, sample_col = sample_col)
metadata[sample_col] <- gsub("-", "_", metadata[sample_col])
```


```{r}
for (ds in names(datasets)) {
  seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
  
  seurat <- replace_HiTMElayer3_annot(seurat, thresh = 0.1)
  
  seurat@misc$label_col <- datasets[[ds]][["label_col"]]
  seurat@misc$low_res_ct_col <- datasets[[ds]][["low_res_ct_col"]]
  seurat@misc$hi_res_ct_col <- datasets[[ds]][["hi_res_ct_col"]]
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots)
}
```



```{r}
labels_ds <- c()
labels_covid <- c()
temp_list <- list()

for (ds in c("GongSharma_all_age", "Stephenson")) {
  temp_list[[ds]] <- as.data.frame(result_list$bmark[[ds]]$ECODA_HiTME_HR_layer2$feat_mat)
  # temp_list[[ds]] <- as.data.frame(result_list$bmark[[ds]]$Pseudobulk_hvg2000$feat_mat)
  labels_ds <- c(labels_ds, rep(ds, nrow(temp_list[[ds]])))
  
  if (ds == "Stephenson") {
    labels_covid <- c(labels_covid, as.character(result_list$bmark[[ds]]$ECODA_HiTME_HR_layer2$labels))
  } else {
    labels_covid <- c(labels_covid, rep("Healthy", nrow(temp_list[[ds]])))
  }
}

ds <- "Zhu"
temp_list[[ds]] <- as.data.frame(fmat_zhu)
labels_ds <- c(labels_ds, rep(ds, nrow(temp_list[[ds]])))
labels_covid <- c(labels_covid, rep("Healthy", nrow(temp_list[[ds]])))

feat_mat <- plyr::rbind.fill(temp_list)
feat_mat <- feat_mat[ , colSums(is.na(feat_mat)) == 0]

plot_pca(feat_mat = feat_mat, labels = labels_ds, n_ct_show = 5, invisible = "quali")
ggsave(file.path(path_plots, "batcheffect_CombinedPBMC_PCA_ECODA_batch.pdf"),
       width = 4,
       height = 4)
plot_pca(feat_mat = feat_mat, labels = labels_covid, n_ct_show = 5, invisible = "quali")
ggsave(file.path(path_plots, "batcheffect_CombinedPBMC_PCA_ECODA_bio.pdf"),
       width = 4,
       height = 4)
plot_pca(feat_mat = feat_mat, labels = labels_ds, plotly_3d = T)
plot_pca(feat_mat = feat_mat, labels = labels_covid, plotly_3d = T)
```

