---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
source("functions.R")
source("Processed_dataset_metadata.R")

library(forcats)
library(patchwork)

# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
py_install("mofapy2", envname = "my_env", method="auto")
import("mofapy2")
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
# install.packages("GloScope")
library(GloScope)

# Set paths
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"

set.seed(123)

`%notin%` <- Negate(`%in%`)
```



# ---------------------------------------------------------------


# Joanito dataset

## Load/preprocess

```{r}
seurat <- readRDS(file.path(path_data, "JoaI_2022_35773407_Nofilt_whole_ECODAprocessed.rds"))


seurat@meta.data$cell.type_new <- seurat@meta.data$cell.type
iCMS <- !is.na(seurat@meta.data$iCMS)
seurat@meta.data$cell.type_new[iCMS] <- paste0(seurat@meta.data$cell.type[iCMS], "_", ifelse(seurat@meta.data$iCMS[iCMS] == "Normal", "Normal", "Cancer"))

meta <- seurat@meta.data %>%
  group_by(sample.ID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample.ID
meta$sample.origin <- gsub("-2", "", meta$sample.origin)
meta$seqtec <- ifelse(meta$dataset %in% c("CRC-SG1", "KUL5"), "5' seq", "3' seq")

pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "default")

bio_col <- "sample.origin"
batch_col <- "seqtec"

plot_list <- list()
```


## Plot PCA

### ECODA

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[bio_col]])
plot_list[["p_ecoda_bio"]] <- plot_pca(feat_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_bio"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_ECODA_bio.pdf"),
       width = 4,
       height = 4)


avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[batch_col]])
plot_list[["p_ecoda_batch"]] <- plot_pca(feat_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_batch"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_ECODA_batch.pdf"),
       width = 4,
       height = 4)



plot_pca(feat_mat, labels = meta$dataset, title = "Authors annotation")


feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$layer2)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

plot_pca(feat_mat, labels = meta[[bio_col]], title = "HiTME annotation")
plot_pca(feat_mat, labels = meta$dataset, title = "HiTME annotation")

feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$layer_6)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

plot_pca(feat_mat, labels = meta[[bio_col]], title = "scATOMIC annotation")
plot_pca(feat_mat, labels = meta$dataset, title = "scATOMIC annotation")
```


### PB


#### Without blacklist

```{r}
pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "none")

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[bio_col]])
plot_list[["p_pb_bio"]] <- plot_pca(pb_norm, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_bio"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
       width = 4,
       height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[batch_col]])
plot_list[["p_pb_batch"]] <- plot_pca(pb_norm, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_batch"]])
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
       width = 4,
       height = 3)


plot_pca(pb_norm, labels = meta$dataset)
```


#### With blacklist

```{r}
pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "default")

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[bio_col]])
plot_pca(pb_norm, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
#        width = 4,
#        height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[batch_col]])
plot_pca(pb_norm, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
#        width = 4,
#        height = 3)


plot_pca(pb_norm, labels = meta$dataset)
```


### Split by seq technology

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

prime3 <- meta$dataset %in% c("SMC", "KUL3", "CRC-SG2")
prime5 <- meta$dataset %in% c("CRC-SG1", "KUL5")
plot_pca(pb_norm[prime3, ], labels = meta$sample.origin[prime3])
plot_pca(pb_norm[prime5, ], labels = meta$sample.origin[prime5])

plot_pca(feat_mat[prime3, ], labels = meta$sample.origin[prime3])
plot_pca(feat_mat[prime5, ], labels = meta$sample.origin[prime5])
```

### Scores bio vs batch

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

labels <- meta$sample.origin
scores_ecoda_bio <- calc_sep_score(feat_mat, labels)
scores_pb_bio <- calc_sep_score(pb_norm, labels)

labels <- meta$dataset
scores_ecoda_batch <- calc_sep_score(feat_mat, labels)
scores_pb_batch <- calc_sep_score(pb_norm, labels)

print(paste("ECODA: Bio ANOSIM", round(scores_ecoda_bio$anosim_score, 3),
            "/ Batch ANOSIM", round(scores_ecoda_batch$anosim_score, 3),
            "=", round(scores_ecoda_bio$anosim_score / scores_ecoda_batch$anosim_score, 3)))
print(paste("PB: Bio ANOSIM", round(scores_pb_bio$anosim_score, 3),
            "/ Batch ANOSIM", round(scores_pb_batch$anosim_score, 3),
            "=", round(scores_pb_bio$anosim_score / scores_pb_batch$anosim_score, 3)))
```


# Combine all plots

```{r fig.height=6.5, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2)+
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_combined.pdf"),
       width = 10,
       height = 6.5)
```


# ---------------------------------------------------------------



# Stephenson dataset

## Load/preprocess

```{r}
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
p <- file.path(path_nas, "StephensonE_2021_39313646", "output", "StephensonE_2021_33879890_preprocessed.rds")

seurat <- readRDS(p)

seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))

seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset[seurat$Days_from_onset == "nan"] <- 999
seurat$Days_from_onset[seurat$Days_from_onset == "Not_known"] <- 999
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)


meta <- seurat@meta.data %>%
  group_by(Sample) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$Sample

pb <- get_pb(seurat, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

bio_col <- "Status"
batch_col <- "Site"


ct_remove <- c(
  "Platelets",
  "RBC",
  "C1_CD16_mono",
  "CD14_mono",
  "CD16_mono"
)
seurat_sub <- subset(seurat, subset = full_clustering %notin% ct_remove)
pb <- get_pb(seurat_sub, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm_sub <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

plot_list <- list()
```


## Plot PCA

### ECODA

```{r}
feat_mat <- as.data.frame.matrix(table(seurat@meta.data$Sample, seurat@meta.data$full_clustering)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[bio_col]])
plot_list[["p_ecoda_bio"]] <- plot_pca(feat_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_bio"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_bio.pdf"),
       width = 4,
       height = 4)


avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[batch_col]])
plot_list[["p_ecoda_batch"]] <- plot_pca(feat_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_batch"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_batch.pdf"),
       width = 4,
       height = 4)
```


#### Subset to remove batch effect cell types

```{r}
feat_mat <- as.data.frame.matrix(table(seurat_sub@meta.data$Sample, seurat_sub@meta.data$full_clustering)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[bio_col]])
plot_list[["p_ecoda_bio_sub"]] <- plot_pca(feat_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_bio_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_bio.pdf"),
       width = 4,
       height = 3)


avg_sep_score <- calc_avg_sep_score(feat_mat, labels = meta[[batch_col]])
plot_list[["p_ecoda_batch_sub"]] <- plot_pca(feat_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_batch_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_batch.pdf"),
       width = 4,
       height = 3)
```




### PB

```{r}
avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[bio_col]])
plot_list[["p_pb_bio"]] <- plot_pca(pb_norm, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_bio"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_bio.pdf"),
       width = 4,
       height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = meta[[batch_col]])
plot_list[["p_pb_batch"]] <- plot_pca(pb_norm, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_batch"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_batch.pdf"),
       width = 4,
       height = 3)
```

#### Subset to remove batch effect cell types

```{r}
avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[bio_col]])
plot_list[["p_pb_bio_sub"]] <- plot_pca(pb_norm_sub, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_bio_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_bio.pdf"),
       width = 4,
       height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[batch_col]])
plot_list[["p_pb_batch_sub"]] <- plot_pca(pb_norm_sub, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_batch_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_batch.pdf"),
       width = 4,
       height = 3)
```


#### With blacklist

```{r}
pb_norm_sub <- get_pb_deseq2(seurat_sub, sample_col = "Sample", black_list = "default")

avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[bio_col]])
plot_pca(pb_norm_sub, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
#        width = 4,
#        height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm_sub, labels = meta[[batch_col]])
plot_pca(pb_norm_sub, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
#        width = 4,
#        height = 3)

```

# Combine all plots

```{r fig.height=15, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 4)+
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_combined.pdf"),
       width = 10,
       height = 13)
```


# ---------------------------------------------------------------

# Combined PBMC (Stephenson, GongSharma, Zhu)

## Aggregate pseudobulk

### Stephenson

```{r}
ds <- "Stephenson"
seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))

pb_stephenson <- get_pb(seurat, sample_col = "Sample")
metadata_stephenson <- get_metadata(seurat, sample_col = "Sample")

feat_mat_stephenson <- as.data.frame.matrix(table(seurat@meta.data$Sample, seurat@meta.data$layer2)) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()
```

### GongSharma

```{r}
ds <- "GongSharma_all_age"
seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))

pb_gs <- get_pb(seurat, sample_col = "Sample")
metadata_gs <- get_metadata(seurat, sample_col = "Sample")

feat_mat_gs <- as.data.frame.matrix(table(seurat@meta.data$Sample, seurat@meta.data$layer2)) %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()
```

### Zhu

```{r}
seurat <- readRDS("/Users/christianhalter/Desktop/ECODA_paper/data/ZhuH_2023_37379396whole_ECODAprocessed.rds")

sample_col <- "Sample"

feat_mat_zhu <- get_ct_comp_df_seurat(seurat, sample_col = sample_col, ct_col = "layer2") %>%
  impute_zeros(clr_zero_impute_method = "counts_zeros", clr_zero_impute_num = 2/3) %>%
  clr()

pb_zhu <- get_pb(seurat, sample_col = sample_col, hvg = NULL)
metadata_zhu <- get_metadata(seurat, sample_col = sample_col)
```

## Combine pb matrices

```{r}
# 1. Put your three matrices into a list
matrix_list <- list(pb_stephenson, pb_gs, pb_zhu)

# 2. Find the intersection of all row names
common_genes <- Reduce(intersect, lapply(matrix_list, rownames))

# 3. Subset and combine in one step
# This ensures rows are in the exact same order before binding
combined_pseudobulk <- do.call(cbind, lapply(matrix_list, function(x) {
  x[common_genes, , drop = FALSE]
}))
print(dim(combined_pseudobulk))

metadata_stephenson$ds <- "Stephenson"
metadata_stephenson$cond <- metadata_stephenson$Status
metadata_gs$ds <- "GongSharma"
metadata_gs$cond <- "Healthy"
metadata_zhu$ds <- "Zhu"
metadata_zhu$cond <- "Healthy"

metadata_list <- list(
  metadata_stephenson,
  metadata_gs,
  metadata_zhu
)
combined_metadata <- do.call(rbind, lapply(metadata_list, function(x) {
  x
}))

pb_norm <- t(DESeq2.normalize(combined_pseudobulk, metadata = combined_metadata, n_hvg = 2000))

feat_mat_list <- list(
  feat_mat_stephenson,
  feat_mat_gs,
  feat_mat_zhu
)
# 1. Find columns common to all dataframes
common_cols <- Reduce(intersect, lapply(feat_mat_list, colnames))

# 2. Subset each and rbind
combined_feat_mat <- do.call(rbind, lapply(feat_mat_list, function(x) {
  x[, common_cols, drop = FALSE]
}))

```


# ECODA

```{r}
plot_list <- list()

avg_sep_score <- calc_avg_sep_score(combined_feat_mat, labels = combined_metadata$cond)
plot_list[["p_ecoda_bio"]] <- plot_pca(combined_feat_mat, labels = combined_metadata$cond,
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_bio"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_ECODA_bio.pdf"),
       width = 4,
       height = 4)


avg_sep_score <- calc_avg_sep_score(combined_feat_mat, labels = combined_metadata$ds)
plot_list[["p_ecoda_batch"]] <- plot_pca(combined_feat_mat, labels = combined_metadata$ds,
         cluster_score = F, mod_score = F, anosim_score = F, invisible = "quali", n_ct_show = 5, repel = TRUE,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_ecoda_batch"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_ECODA_batch.pdf"),
       width = 4,
       height = 4)
```

# PB

```{r}
avg_sep_score <- calc_avg_sep_score(pb_norm, labels = combined_metadata$cond)
plot_list[["p_pb_bio"]] <- plot_pca(pb_norm, labels = combined_metadata$cond,
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_bio"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_PB_bio.pdf"),
       width = 4,
       height = 3)

avg_sep_score <- calc_avg_sep_score(pb_norm, labels = combined_metadata$ds)
plot_list[["p_pb_batch"]] <- plot_pca(pb_norm, labels = combined_metadata$ds,
         cluster_score = F, mod_score = F, anosim_score = F,
         title = paste("Separation score:", avg_sep_score)
)
print(plot_list[["p_pb_batch"]])
ggsave(file.path(path_plots, "batcheffect_combined_PCA_PB_batch.pdf"),
       width = 4,
       height = 3)
```



# Combine all plots

```{r fig.height=6.5, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(file.path(path_plots, "batcheffect_combined_PCA_combined.pdf"),
       width = 10,
       height = 6.5)
```

