---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script compares different methods for exploratory data analysis of scRNA-seq data.


# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(factoextra)
library(plotly)
library(Hotelling)
library(compositions)
source("functions.R")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)

# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
library(ncdf4)
library(arrow)

# remotes::install_github("carmonalab/scooter", ref="f31eab3")
library(scooter)

my_seed <- 123
set.seed(my_seed)


# Set path to save plots
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# ---------------------------------------------------------------


```{r}
result_list <- list()
```


# Zhang et al. dataset

## Set paths

```{r}
ds <- "Zhang"

ds_name <- "ZhangY_2022_34653365"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
path_output_samples <- file.path(path_output, "samples")
```


### Load seurat object

```{r}
obj.list <- read_objs(path_output_samples)

# Exclude samples that consist mainly of one cell type as outliers
samples_exclude <- c("Pre_P012_b", "Pre_P013_t")
obj.list <- obj.list[!names(obj.list) %in% samples_exclude]

samples_clearcut <- c()

for (i in names(obj.list)) {
  if (obj.list[[i]]@meta.data$Treatment[1] %in% c("Pre") &
      !(obj.list[[i]]@meta.data$Clinical.efficacy.[1] %in% c("-"))) {
    samples_clearcut <- c(samples_clearcut, i)
  }
}

metacols_to_keep <- c("Sample", "Tissue", "layer1", "layer2")

for (i in samples_clearcut) {
  print(i)
  
  obj <- obj.list[[i]]
  Idents(obj) <- "Sample"
  obj@meta.data <- obj@meta.data[colnames(obj@assays$RNA$counts), metacols_to_keep]

  if (i == samples_clearcut[1]) {
    seurat <- obj
  } else {
    seurat <- merge(seurat, obj)
  }
}

rm(obj.list); gc()


seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features

label_col <- "Tissue"
low_res_ct_col <- "layer1"
hi_res_ct_col <- "layer2"

# Convert Seurat object to anndata for interoperability with python
# SaveH5Seurat(seurat, filename = "zhang_clearcut.h5Seurat")
# Convert("zhang_clearcut.h5Seurat", dest = "h5ad")
```


### Create plots

```{r}
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
result_list[[ds]][["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg = pseudobulk_hvg, labels = labels, hvg = hvg)

# Deconvolute using EPIC
result_list[[ds]][["deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["coda_lowres"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_collow_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
result_list[[ds]][["coda_hires"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_colhi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
result_list[[ds]][["mofa"]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, save_file_path = file.path("data/gongsharma_all_mofa_object.rds"))

# scITD
result_list[[ds]][["scitd"]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg)

# MrVI
result_list[[ds]][["mrvi"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/zhang_dists.nc", ct_col = low_res_ct_col, label_col = label_col)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_zhang.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Stephenson et al. dataset

## Set paths

```{r}
ds <- "Stephenson"

ds_name <- "MitchelJ_2023_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


### Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))

seurat <- subset(seurat, subset = Site == "Ncl")
seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))
seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)

seurat$sample_id <- as.character(seurat$sample_id)
samples_exclude <- c("BGCV10_CV0198", "MH8919230")
samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% samples_exclude])
seurat <- subset(seurat, subset = sample_id %in% samples_subset)

seurat$Sample <- seurat$sample_id
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features

label_col <- "Status"
low_res_ct_col <- "initial_clustering"
hi_res_ct_col <- "full_clustering"

# Convert Seurat object to anndata for interoperability with python
# SaveH5Seurat(seurat, filename = "mitchel_clearcut.h5Seurat")
# Convert("mitchel_clearcut.h5Seurat", dest = "h5ad")
```

### Create plots

```{r}
result_list[[ds]] <- list()
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
result_list[[ds]][["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg = pseudobulk_hvg, labels = labels, hvg = hvg)


# Deconvolute using EPIC
result_list[[ds]][["deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["coda_lowres"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_collow_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
result_list[[ds]][["coda_hires"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_colhi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
result_list[[ds]][["mofa"]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, save_file_path = file.path("data/mitchel_mofa_object.rds"))

# scITD
result_list[[ds]][["scitd"]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg)

# MrVI
result_list[[ds]][["mrvi"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/mitchel_dists.nc", "layer1", label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_mitchel.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Gong and Sharma et al. dataset

## Set paths

```{r}
ds <- "GongSharma_all"

ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## All CMV

### Load seurat object

```{r}
seurat <- load_seurat_object("gongsharma_all.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```

### Create plots

```{r}
label_col <- "subject.cmv"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

result_list[[ds]] <- list()
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
result_list[[ds]][["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg = pseudobulk_hvg, labels = labels, hvg = hvg)


# Deconvolute using EPIC
result_list[[ds]][["deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["coda_lowres"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_collow_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
result_list[[ds]][["coda_hires"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_colhi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
result_list[[ds]][["mofa"]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, save_file_path = file.path("data/gongsharma_all_mofa_object.rds"))

# scITD
result_list[[ds]][["scitd"]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg)

# MrVI
result_list[[ds]][["mrvi"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_all_dists.nc", ct_col = low_res_ct_col, label_col = label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_all_cmv.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## All gender

### Create plots

```{r}
ds <- "GongSharma_all_gender"

label_col <- "subject.biologicalSex"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

result_list[[ds]] <- list()
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
result_list[[ds]][["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg = pseudobulk_hvg, labels = labels, hvg = hvg)


# Deconvolute using EPIC
result_list[[ds]][["deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["coda_lowres"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_collow_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
result_list[[ds]][["coda_hires"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_colhi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
result_list[[ds]][["mofa"]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, save_file_path = file.path("data/gongsharma_all_mofa_object.rds"))

# scITD
result_list[[ds]][["scitd"]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg)

# MrVI
result_list[[ds]][["mrvi"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_all_dists.nc", ct_col = low_res_ct_col, label_col = label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_all_gender.pdf"),
       plot = p,
       width = 15,
       height = 10)
```




## Gender (CMV negative)

### Load seurat objects

```{r}
ds <- "GongSharma_all_gender"

seurat <- load_seurat_object("gongsharma_gender_cmvneg.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```


### Create plots

```{r}
label_col <- "subject.biologicalSex"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

result_list[[ds]] <- list()
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
result_list[[ds]][["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg = pseudobulk_hvg, labels = labels, hvg = hvg)


# Deconvolute using EPIC
result_list[[ds]][["deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["coda_lowres"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_collow_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
result_list[[ds]][["coda_hires"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_colhi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
result_list[[ds]][["mofa"]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, save_file_path = file.path("data/gongsharma_gender_cmvneg_mofa_object.rds"))

# scITD
result_list[[ds]][["scitd"]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg)

# MrVI
result_list[[ds]][["mrvi"]] <- process_mrvi_fig(seurat = seurat, dist_file = dist_file = "data/gongsharma_gender_cmvneg_dists.nc", ct_col = low_res_ct_col, label_col = label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_gender_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## CMV (females)

### Load seurat objects

```{r}
ds <- "GongSharma_females_cmv"

seurat <- load_seurat_object("gongsharma_cmv_females.h5ad")

seurat$Sample <- seurat$specimen.specimenGuid
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```



### Create plots

```{r}
label_col <- "subject.cmv"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

result_list[[ds]] <- list()
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
result_list[[ds]][["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg = pseudobulk_hvg, labels = labels, hvg = hvg)

# Deconvolute using EPIC
result_list[[ds]][["deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["coda_lowres"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_collow_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
result_list[[ds]][["coda_hires"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_colhi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
seurat <- UpdateSeuratObject(seurat)
result_list[[ds]][["mofa"]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, save_file_path = file.path("data/gongsharma_cmv_females_mofa_object.rds"))

# scITD
result_list[[ds]][["scitd"]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg)

# MrVI
result_list[[ds]][["mrvi"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_cmv_females_dists.nc", ct_col = low_res_ct_col, label_col = label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_cmv_females.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## Age (females cmvneg)

### Load seurat objects

```{r}
ds <- "GongSharma_females_cmvneg_age"

seurat <- load_seurat_object("gongsharma_age_females_cmvneg.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```



### Create plots

```{r}
label_col <- "age"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

result_list[[ds]] <- list()
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
result_list[[ds]][["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg = pseudobulk_hvg, labels = labels, hvg = hvg)


# Deconvolute using EPIC
result_list[[ds]][["deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["coda_lowres"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_collow_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
result_list[[ds]][["coda_hires"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_colhi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
seurat <- UpdateSeuratObject(seurat)
result_list[[ds]][["mofa"]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, save_file_path = file.path("data/gongsharma_age_females_cmvneg_mofa_object.rds"))

# scITD
result_list[[ds]][["scitd"]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg)

# MrVI
result_list[[ds]][["mrvi"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_age_females_cmvneg_dists.nc", ct_col = low_res_ct_col, label_col = label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_age_females_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 10)
```
