---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script compares different methods for exploratory data analysis of scRNA-seq data.


# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(factoextra)
library(plotly)
library(Hotelling)
library(compositions)
library(rstatix)
source("functions.R")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)

# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
py_install("mofapy2", envname = "my_env", method="auto")
import("mofapy2")
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
library(ncdf4)
library(arrow)

# remotes::install_github("carmonalab/scooter", ref="f31eab3")
library(scooter)

my_seed <- 123
set.seed(my_seed)


# Set path to save plots
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# ---------------------------------------------------------------


```{r}
result_list <- list()

factors_test <- c(2, 5, 10)
```


# Zhang et al. dataset

## Set paths

```{r}
ds <- "Zhang"

ds_name <- "ZhangY_2022_34653365"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
path_output_samples <- file.path(path_output, "samples")
```


### Load seurat object

```{r}
obj.list <- read_objs(path_output_samples)

# Exclude samples that consist mainly of one cell type as outliers
samples_exclude <- c("Pre_P012_b", "Pre_P013_t")
obj.list <- obj.list[!names(obj.list) %in% samples_exclude]

samples_clearcut <- c()

for (i in names(obj.list)) {
  if (obj.list[[i]]@meta.data$Treatment[1] %in% c("Pre") &
      !(obj.list[[i]]@meta.data$Clinical.efficacy.[1] %in% c("-"))) {
    samples_clearcut <- c(samples_clearcut, i)
  }
}

metacols_to_keep <- c("Sample", "Tissue", "layer1", "layer2")

for (i in samples_clearcut) {
  print(i)
  
  obj <- obj.list[[i]]
  Idents(obj) <- "Sample"
  obj@meta.data <- obj@meta.data[colnames(obj@assays$RNA$counts), metacols_to_keep]

  if (i == samples_clearcut[1]) {
    seurat <- obj
  } else {
    seurat <- merge(seurat, obj)
  }
}

rm(obj.list); gc()


seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features

label_col <- "Tissue"
low_res_ct_col <- "layer1"
hi_res_ct_col <- "layer2"

# Convert Seurat object to anndata for interoperability with python
# SaveH5Seurat(seurat, filename = "zhang_clearcut.h5Seurat")
# Convert("zhang_clearcut.h5Seurat", dest = "h5ad")
```


### Create plots

```{r}
result_list[[ds]] <- list()
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
colnames(pseudobulk) <- str_replace_all(colnames(pseudobulk), "-", "_")
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))
result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)

# Deconvolute using EPIC
result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition\nlow res.")

## layer2: high res. cell types
result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition\nhigh res.")


for (i in factors_test) {
  # Pseudobulk with PCA
  result_list[[ds]][[paste0("Pseudobulk_", i, "_PCA_dims")]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels, pca_dims = i,, title = paste0("Pseudobulk gene expression\n+ PCA (", i, " dims)"))
  
  # Hires CODA with PCA
  result_list[[ds]][[paste0("ECODA_high_res_", i, "_PCA_dims")]] <- process_coda_fig(seurat = seurat, labels = labels, pca_dims = i, ct_col = hi_res_ct_col, title = paste0("Cell type composition\nhigh res. + PCA (", i, " dims)"))
  
  # MOFA
  result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_bulk_fig(pb_norm = pb_norm, metadata = metadata, labels = labels, num_factors = i)
  
  # result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/zhang_mofa_object_", i, "factors.rds")))
  
  # result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/zhang_mofa_object_", i, "factors.rds")))
  
  # scITD
  result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
}

# MrVI
result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/zhang_dists.nc", ct_col = low_res_ct_col, label_col = label_col)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_zhang.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Stephenson et al. dataset

## Set paths

```{r}
ds <- "Stephenson"

ds_name <- "MitchelJ_2023_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


### Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))

seurat <- subset(seurat, subset = Site == "Ncl")
seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))
seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)

seurat$sample_id <- as.character(seurat$sample_id)
samples_exclude <- c("BGCV10_CV0198", "MH8919230")
samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% samples_exclude])
seurat <- subset(seurat, subset = sample_id %in% samples_subset)

seurat$Sample <- seurat$sample_id
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features

label_col <- "Status"
low_res_ct_col <- "initial_clustering"
hi_res_ct_col <- "full_clustering"

# Convert Seurat object to anndata for interoperability with python
# SaveH5Seurat(seurat, filename = "mitchel_clearcut.h5Seurat")
# Convert("mitchel_clearcut.h5Seurat", dest = "h5ad")
```

### Create plots

```{r}
result_list[[ds]] <- list()
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))
result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)

# Deconvolute using EPIC
result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition\nlow res.")

## layer2: high res. cell types
result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition\nhigh res.")


for (i in factors_test) {
  # Pseudobulk with PCA
  result_list[[ds]][[paste0("Pseudobulk_", i, "_PCA_dims")]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels, pca_dims = i,, title = paste0("Pseudobulk gene expression\n+ PCA (", i, " dims)"))
  
  # Hires CODA with PCA
  result_list[[ds]][[paste0("ECODA_high_res_", i, "_PCA_dims")]] <- process_coda_fig(seurat = seurat, labels = labels, pca_dims = i, ct_col = hi_res_ct_col, title = paste0("Cell type composition\nhigh res. + PCA (", i, " dims)"))
  
  # MOFA
  result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_bulk_fig(pb_norm = pb_norm, metadata = metadata, labels = labels, num_factors = i)
  
  # result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/mitchel_mofa_object_", i, "factors.rds")))
  
  # result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/mitchel_mofa_object_", i, "factors.rds")))
  
  # scITD
  result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
}

# MrVI
result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/mitchel_dists.nc", "layer1", label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_mitchel.pdf"),
       plot = p,
       width = 15,
       height = 10)
```



# ---------------------------------------------------------------

# Lee et al. dataset

## Set paths

```{r}
ds <- "Lee"

ds_name <- "LeeA_2021_34836966"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


### Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%sp_tumor_seurat.rds", ds_name)))

seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
varf <- seurat@assays[["RNA"]]@meta.data[["var.features"]]
hvg <- varf[!is.na(varf)]

label_col <- "condition"
low_res_ct_col <- "scGate_multi"
hi_res_ct_col <- "functional.cluster"

# # Convert Seurat object to anndata for interoperability with python
# object <- seurat
# object[["RNA3"]] <- as(object = object[["RNA"]], Class = "Assay")
# DefaultAssay(object) <- "RNA3"
# object[["RNA"]] <- NULL
# object <- RenameAssays(object = object, RNA3 = 'RNA')
# SaveH5Seurat(object, filename = "lee.h5Seurat")
# Convert("lee.h5Seurat", dest = "h5ad")
# rm(object); gc()
```

### Create plots

```{r}
result_list[[ds]] <- list()
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))
result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)

# Deconvolute using EPIC
result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition\nlow res.")

## layer2: high res. cell types
result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition\nhigh res.")


for (i in factors_test) {
  # Pseudobulk with PCA
  result_list[[ds]][[paste0("Pseudobulk_", i, "_PCA_dims")]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels, pca_dims = i,, title = paste0("Pseudobulk gene expression\n+ PCA (", i, " dims)"))
  
  # Hires CODA with PCA
  result_list[[ds]][[paste0("ECODA_high_res_", i, "_PCA_dims")]] <- process_coda_fig(seurat = seurat, labels = labels, pca_dims = i, ct_col = hi_res_ct_col, title = paste0("Cell type composition\nhigh res. + PCA (", i, " dims)"))
  
  # MOFA
  result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_bulk_fig(pb_norm = pb_norm, metadata = metadata, labels = labels, num_factors = i)
  
  # result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/lee_mofa_object_", i, "factors.rds")))
  
  # result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/lee_mofa_object_", i, "factors.rds")))
  
  # scITD
  result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
}

# MrVI
result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/lee_dists.nc", "scGate_multi", label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_lee.pdf"),
       plot = p,
       width = 15,
       height = 10)
```



# ---------------------------------------------------------------

# Gong and Sharma et al. dataset

## Set paths

```{r}
ds <- "GongSharma_all_cmv"

ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


# ----> removed

## CMV (all)

### Load seurat object

```{r}
# seurat <- load_seurat_object("data/gongsharma_all.h5ad")
# seurat$Sample <- seurat$specimen.specimenGuid
# seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
# hvg <- seurat@assays[["RNA"]]@var.features
# 
# seurat@version <- obj@version
```

### Create plots

```{r}
# label_col <- "subject.cmv"
# low_res_ct_col <- "AIFI_L1"
# hi_res_ct_col <- "AIFI_L3"
# 
# result_list[[ds]] <- list()
# labels <- get_labels(seurat, label_col)
# 
# # Pseudobulk
# pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
# pseudobulk_hvg <- pseudobulk[hvg,]
# result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)
# 
# 
# # Deconvolute using EPIC
# result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)
# 
# # CoDA
# ## layer1: low res. cell types
# result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition - low res. cell types")
# 
# ## layer2: high res. cell types
# result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition - high res. cell types")
# 
# for (i in factors_test) {
#   # MOFA
#   result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/gongsharma_all_cmv_mofa_object_", i, "factors.rds")))
#   
#   result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/gongsharma_all_cmv_mofa_object_", i, "factors.rds")))
#   
#   # scITD
#   result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
# }
# 
# # MrVI
# result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_all_dists.nc", ct_col = low_res_ct_col, label_col = label_col)
# 
# 
# plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
# p <- cowplot::plot_grid(plotlist = plot_list)
# p
# ggsave(file.path(path_plots, "benchmark_pca_gongsharma_all_cmv.pdf"),
#        plot = p,
#        width = 15,
#        height = 10)
```


## All gender

### Create plots

```{r}
# ds <- "GongSharma_all_gender"
# 
# label_col <- "subject.biologicalSex"
# low_res_ct_col <- "AIFI_L1"
# hi_res_ct_col <- "AIFI_L3"
# 
# result_list[[ds]] <- list()
# labels <- get_labels(seurat, label_col)
# 
# # Pseudobulk
# pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
# pseudobulk_hvg <- pseudobulk[hvg,]
# result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)
# 
# 
# # Deconvolute using EPIC
# result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)
# 
# # CoDA
# ## layer1: low res. cell types
# result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition - low res. cell types")
# 
# ## layer2: high res. cell types
# result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition - high res. cell types")
# 
# for (i in factors_test) {
#   # MOFA
#   result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/gongsharma_all_gender_mofa_object_", i, "factors.rds")))
#   
#   result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/gongsharma_all_gender_mofa_object_", i, "factors.rds")))
#   
#   # scITD
#   result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
# }
# 
# # MrVI
# result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_all_dists.nc", ct_col = low_res_ct_col, label_col = label_col)
# 
# 
# plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
# p <- cowplot::plot_grid(plotlist = plot_list)
# p
# ggsave(file.path(path_plots, "benchmark_pca_gongsharma_all_gender.pdf"),
#        plot = p,
#        width = 15,
#        height = 10)
```




## Gender (CMV negative)

### Load seurat objects

```{r}
# ds <- "GongSharma_cmvneg_gender"
# 
# seurat <- load_seurat_object("data/gongsharma_gender_cmvneg.h5ad")
# seurat$Sample <- seurat$specimen.specimenGuid
# seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
# hvg <- seurat@assays[["RNA"]]@var.features
# 
# seurat@version <- obj@version
```


### Create plots

```{r}
# label_col <- "subject.biologicalSex"
# low_res_ct_col <- "AIFI_L1"
# hi_res_ct_col <- "AIFI_L3"
# 
# result_list[[ds]] <- list()
# labels <- get_labels(seurat, label_col)
# 
# # Pseudobulk
# pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
# pseudobulk_hvg <- pseudobulk[hvg,]
# result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)
# 
# 
# # Deconvolute using EPIC
# result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)
# 
# # CoDA
# ## layer1: low res. cell types
# result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition - low res. cell types")
# 
# ## layer2: high res. cell types
# result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition - high res. cell types")
# 
# for (i in factors_test) {
#   # MOFA
#   result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/gongsharma_cmvneg_gender_mofa_object_", i, "factors.rds")))
#   
#   result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/gongsharma_cmvneg_gender_mofa_object_", i, "factors.rds")))
#   
#   # scITD
#   result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
# }
# 
# # MrVI
# result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_gender_cmvneg_dists.nc", ct_col = low_res_ct_col, label_col = label_col)
# 
# 
# plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
# p <- cowplot::plot_grid(plotlist = plot_list)
# p
# ggsave(file.path(path_plots, "benchmark_pca_gongsharma_cmvneg_gender.pdf"),
#        plot = p,
#        width = 15,
#        height = 10)
```

# <---- removed

## CMV (females)

### Load seurat objects

```{r}
ds <- "GongSharma_females_cmv"

seurat <- load_seurat_object("data/gongsharma_cmv_females.h5ad")

seurat$Sample <- seurat$specimen.specimenGuid
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features

seurat@version <- obj@version
```



### Create plots

```{r}
label_col <- "subject.cmv"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

result_list[[ds]] <- list()
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))
result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)

# Deconvolute using EPIC
result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition\nlow res.")

## layer2: high res. cell types
result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition\nhigh res.")


for (i in factors_test) {
  # Pseudobulk with PCA
  result_list[[ds]][[paste0("Pseudobulk_", i, "_PCA_dims")]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels, pca_dims = i,, title = paste0("Pseudobulk gene expression\n+ PCA (", i, " dims)"))
  
  # Hires CODA with PCA
  result_list[[ds]][[paste0("ECODA_high_res_", i, "_PCA_dims")]] <- process_coda_fig(seurat = seurat, labels = labels, pca_dims = i, ct_col = hi_res_ct_col, title = paste0("Cell type composition\nhigh res. + PCA (", i, " dims)"))
  
  # MOFA
  result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_bulk_fig(pb_norm = pb_norm, metadata = metadata, labels = labels, num_factors = i)
  
  # result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/gongsharma_females_cmv_mofa_object_", i, "factors.rds")))
  
  # result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/gongsharma_females_cmv_mofa_object_", i, "factors.rds")))
  
  # scITD
  result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
}

# MrVI
result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_cmv_females_dists.nc", ct_col = low_res_ct_col, label_col = label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_females_cmv.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## Age (females cmvneg)

### Load seurat objects

```{r}
ds <- "GongSharma_females_cmvneg_age"

seurat <- load_seurat_object("data/gongsharma_age_females_cmvneg.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features

seurat@version <- obj@version
```



### Create plots

```{r}
label_col <- "age"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

result_list[[ds]] <- list()
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))
result_list[[ds]][["Pseudobulk"]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels)

# Deconvolute using EPIC
result_list[[ds]][["ECODA_deconv"]] <- process_deconv_fig(pseudobulk = pseudobulk, labels = labels)

# CoDA
## layer1: low res. cell types
result_list[[ds]][["ECODA_low_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = low_res_ct_col, title = "Cell type composition\nlow res.")

## layer2: high res. cell types
result_list[[ds]][["ECODA_high_res"]] <- process_coda_fig(seurat = seurat, labels = labels, ct_col = hi_res_ct_col, title = "Cell type composition\nhigh res.")


for (i in factors_test) {
  # Pseudobulk with PCA
  result_list[[ds]][[paste0("Pseudobulk_", i, "_PCA_dims")]] <- process_pseudobulk_fig(pb_norm = pb_norm, labels = labels, pca_dims = i,, title = paste0("Pseudobulk gene expression\n+ PCA (", i, " dims)"))
  
  # Hires CODA with PCA
  result_list[[ds]][[paste0("ECODA_high_res_", i, "_PCA_dims")]] <- process_coda_fig(seurat = seurat, labels = labels, pca_dims = i, ct_col = hi_res_ct_col, title = paste0("Cell type composition\nhigh res. + PCA (", i, " dims)"))
  
  # MOFA
  result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_bulk_fig(pb_norm = pb_norm, metadata = metadata, labels = labels, num_factors = i)
  
  # result_list[[ds]][[paste0("MOFA_", i, "_factors")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, save_file_path = file.path(paste0("data/gongsharma_females_cmvneg_age_mofa_object_", i, "factors.rds")))
  
  # result_list[[ds]][[paste0("mofa_", i, "_factors_log_abs")]] <- process_mofa_fig(seurat = seurat, labels = labels, hvg = hvg, pseudobulk_hvg = pseudobulk_hvg, num_factors = i, log_abs_transform = TRUE, save_file_path = file.path(paste0("data/gongsharma_females_cmvneg_age_mofa_object_", i, "factors.rds")))
  
  # scITD
  result_list[[ds]][[paste0("scITD_", i, "_factors")]] <- process_scitd_fig(seurat = seurat, ct_col = low_res_ct_col, label_col = label_col, hvg = hvg, num_factors = i)
}

# MrVI
result_list[[ds]][["MrVI"]] <- process_mrvi_fig(seurat = seurat, dist_file = "data/gongsharma_age_females_cmvneg_dists.nc", ct_col = low_res_ct_col, label_col = label_col)


plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_females_cmvneg_age.pdf"),
       plot = p,
       width = 15,
       height = 10)
```

# ---------------------------------------------------------------


# Save and flatten the nested list into a DataFrame

```{r}
saveRDS(result_list, file = "data/benchmark_result_list.rds")

df_results <- imap_dfr(result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    if ("scores" %in% names(.x)) {
      imap_dfr(.x$scores, ~ {
        tibble(
          dataset = dataset_name,
          method = method_name,
          score = .y,
          value = .x
        )
      })
    } else {
      tibble() # Return an empty tibble if "scores" is not found
    }
  })
})
```


# Plot results

```{r fig.height=10, fig.width=14}
levs <- sort(as.character(unique(df_results$method)))
levs <- c(
  "ECODA_high_res",
  "ECODA_high_res_2_PCA_dims",
  "ECODA_high_res_5_PCA_dims",
  "ECODA_high_res_10_PCA_dims",
  "Pseudobulk",
  "Pseudobulk_2_PCA_dims",
  "Pseudobulk_5_PCA_dims",
  "Pseudobulk_10_PCA_dims",
  "MrVI",
  "MOFA_2_factors",
  "MOFA_5_factors",
  "MOFA_10_factors",
  "scITD_2_factors",
  "scITD_5_factors",
  "scITD_10_factors",
  "ECODA_low_res",
  "ECODA_deconv"
)

df_results$method <- factor(df_results$method, levels = levs)

df_results$dataset[df_results$dataset == "GongSharma_females_cmv"] <- "GongSharma_cmv"
df_results$dataset[df_results$dataset == "GongSharma_females_cmvneg_age"] <- "GongSharma_age"

df_results_clustering <- df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

# df_results_clustering <- df_results_clustering %>%
#   arrange(desc(mean))
df_results_clustering$method <- factor(df_results_clustering$method, levels = df_results_clustering$method)

df_results$method <- factor(df_results$method, levels = df_results_clustering$method)


df_results_modularity <- df_results %>%
  filter(score == "mod_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_modularity <- df_results_modularity %>%
  arrange(desc(mean))
df_results_modularity$method <- factor(df_results_modularity$method, levels = df_results_modularity$method)



# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "Silhouette score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- c(
  "Gong & Sharma\nSubset: females\nSeparation by: CMV infection",
  "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age",
  "Lee",
  "Stephenson",
  "Zhang")
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            labeller = labeller(score = score.labs, dataset = datas.labs),
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "benchmark_detailed_barplot.pdf"),
       width = 10,
       height = 6)


df_results %>%
  filter(score == "mod_score" | score == "cluster_score") %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_mod_clust.pdf"),
       width = 2.5,
       height = 4)



df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_bars.pdf"),
       width = 5,
       height = 7)

ggplot(df_results, aes(x = method, y = value)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores.pdf"),
       width = 7,
       height = 7)



df_results_clustering %>%
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_clustering)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)


ggplot(df_results_clustering, aes(x = method, y = value)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Adjusted Rand Index") +
  stat_compare_means(method = "anova", label.x.npc = 0.1, label.y.npc = 0.9) +
  # stat_compare_means(
  #   comparisons = my_comparisons,
  #   method = "t.test", 
  #   paired = F,
  #   p.adjust.method = "BH" #,
  #   # symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), symbols = c("****", "***", "**", "*", "ns"))
  # ) +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #   y.position = c(max(df_results_clustering$values)+0.1, max(df_results_clustering$values)+0.2), hide.ns = TRUE) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_clustering.pdf"),
       width = 6,
       height = 4.5)

df_results_modularity %>%
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Modularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity.pdf"),
       width = 2.5,
       height = 2.5)

```

# Show example PCAs for publication

```{r fig.height=6, fig.width=12}

levs <- c(
  "ECODA_high_res",
  "ECODA_low_res",
  "ECODA_deconv",
  "Pseudobulk",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI"
)


plot_list <- lapply(result_list[["GongSharma_females_cmv"]], function(method) method[["plot"]])
scores <- lapply(result_list[["GongSharma_females_cmv"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_females_cmv.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["GongSharma_females_cmvneg_age"]], function(method) method[["plot"]])
scores <- lapply(result_list[["GongSharma_females_cmvneg_age"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_females_cmvneg_age.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["Lee"]], function(method) method[["plot"]])
scores <- lapply(result_list[["Lee"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Lee.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["Zhang"]], function(method) method[["plot"]])
scores <- lapply(result_list[["Zhang"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Zhang.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["Stephenson"]], function(method) method[["plot"]])
scores <- lapply(result_list[["Stephenson"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Stephenson.pdf"),
       plot = p,
       width = 15,
       height = 7)

```
