---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script compares different methods for exploratory data analysis of scRNA-seq data.


# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

library(rstatix)
source("functions.R")

my_seed <- 123
set.seed(my_seed)


# Set path to save plots
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# ---------------------------------------------------------------


```{r}
result_list <- list()

factors_test <- c(2, 5, 10)
```


# Zhang et al. dataset

## Set paths

```{r}
ds <- "Zhang"

ds_name <- "ZhangY_2022_34653365"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
path_output_samples <- file.path(path_output, "samples")
```


## Load seurat object

```{r}
obj.list <- read_objs(path_output_samples)

# Exclude samples that consist mainly of one cell type as outliers
samples_exclude <- c("Pre_P012_b", "Pre_P013_t")
obj.list <- obj.list[!names(obj.list) %in% samples_exclude]

samples_clearcut <- c()

for (i in names(obj.list)) {
  if (obj.list[[i]]@meta.data$Treatment[1] %in% c("Pre") &
      !(obj.list[[i]]@meta.data$Clinical.efficacy.[1] %in% c("-"))) {
    samples_clearcut <- c(samples_clearcut, i)
  }
}

metacols_to_keep <- c("Sample", "Tissue", "layer1", "layer2")

for (i in samples_clearcut) {
  print(i)
  
  obj <- obj.list[[i]]
  Idents(obj) <- "Sample"
  obj@meta.data <- obj@meta.data[colnames(obj@assays$RNA$counts), metacols_to_keep]
  
  if (i == samples_clearcut[1]) {
    seurat <- obj
  } else {
    seurat <- merge(seurat, obj)
  }
}

rm(obj.list); gc()

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "Tissue"
low_res_ct_col <- "layer1"
hi_res_ct_col <- "layer2"
```


## Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")
result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)
plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Stephenson et al. dataset

## Set paths

```{r}
ds <- "Stephenson"

ds_name <- "MitchelJ_2023_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))

seurat <- subset(seurat, subset = Site == "Ncl")
seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))
seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)

seurat$sample_id <- as.character(seurat$sample_id)
samples_exclude <- c("BGCV10_CV0198", "MH8919230")
samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% samples_exclude])
seurat <- subset(seurat, subset = sample_id %in% samples_subset)

seurat$Sample <- seurat$sample_id

# Convert Seurat object to anndata for interoperability with python
# var_genes <- VariableFeatures(seurat)
# seurat_df <- GetAssayData(seurat)[var_genes,]
object <- CreateSeuratObject(counts = seurat@assays$RNA$counts, meta.data = seurat@meta.data[, c("Sample", "initial_clustering")])
object[["RNA3"]] <- as(object = object[["RNA"]], Class = "Assay")
DefaultAssay(object) <- "RNA3"
object[["RNA"]] <- NULL
object <- RenameAssays(object = object, RNA3 = 'RNA')
convert_seurat_to_h5ad(object, ds)
rm(object); gc()


label_col <- "Status"
low_res_ct_col <- "initial_clustering"
hi_res_ct_col <- "full_clustering"
```

## Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Gong and Sharma et al. dataset

## Set paths

```{r}
ds <- "GongSharma_all_cmv"

ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```



## CMV (females)

### Load seurat objects

```{r}
ds <- "GongSharma_females_cmv"

seurat <- load_seurat_object("data/gongsharma_cmv_females.h5ad")

seurat$Sample <- seurat$specimen.specimenGuid

seurat@version <- obj@version

label_col <- "subject.cmv"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"
```



### Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_", ds, ".pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## Age (females cmvneg)

### Load seurat objects

```{r}
ds <- "GongSharma_females_cmvneg_age"

seurat <- load_seurat_object("data/gongsharma_age_females_cmvneg.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))

seurat@version <- obj@version

label_col <- "age"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"
```



### Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Lee et al. dataset

## Set paths

```{r}
ds <- "Lee"

ds_name <- "LeeA_2021_34836966"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%sp_tumor_seurat.rds", ds_name)))

label_col <- "condition"
low_res_ct_col <- "scGate_multi"
hi_res_ct_col <- "functional.cluster"

# # Convert Seurat object to anndata for interoperability with python
# object <- seurat
# object[["RNA3"]] <- as(object = object[["RNA"]], Class = "Assay")
# DefaultAssay(object) <- "RNA3"
# object[["RNA"]] <- NULL
# object <- RenameAssays(object = object, RNA3 = 'RNA')
# SaveH5Seurat(object, filename = "lee.h5Seurat")
# Convert("lee.h5Seurat", dest = "h5ad")
# rm(object); gc()
```

## Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```




# ---------------------------------------------------------------

# Pelka et al. dataset

## Set paths

```{r}
ds <- "Pelka"

ds_name <- "PelkaK_2022_34450029"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name)))

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "Tissue"
low_res_ct_col <- "OriginalAnnotationLevel1"
hi_res_ct_col <- "OriginalAnnotationLevel3"
```

## Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Bassez et al. dataset

## Set paths

```{r}
ds <- "Bassez"

ds_name <- "BassezA_2021_33958794"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Treatment cohort

### Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name)))

seurat <- subset(seurat, subset = timepoint == "Pre")

seurat$Sample <- seurat$sample
seurat$cellSubType[is.na(seurat$cellSubType)] <- seurat$cellType[is.na(seurat$cellSubType)]

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "cohort"
low_res_ct_col <- "cellType"
hi_res_ct_col <- "cellSubType"
```

### Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Wu et al. dataset

## Set paths

```{r}
ds <- "Wu"

ds_name <- "WuS_2021_34493872"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))

label_col <- "subtype"
low_res_ct_col <- "celltype_major"
hi_res_ct_col <- "celltype_subset"

# # Convert Seurat object to anndata for interoperability with python
# object <- seurat
# object[["RNA3"]] <- as(object = object[["RNA"]], Class = "Assay")
# DefaultAssay(object) <- "RNA3"
# object[["RNA"]] <- NULL
# object <- RenameAssays(object = object, RNA3 = 'RNA')
# convert_seurat_to_h5ad(seurat, ds)
# rm(object); gc()
```

## Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```



# ---------------------------------------------------------------

# Smillie et al. dataset

## Set paths

```{r}
ds <- "Smillie"

ds_name <- "SmillieC_2019_31348891"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))

stot <- seurat
# seurat <- subset(stot, subset = Tissue_compartment == "Epithelial")
# seurat <- subset(stot, subset = Tissue_compartment == "Lamina_propria")
seurat <- subset(stot, subset = stot$OriginalAnnotationLevel1 == "Immune")

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "Tissue_compartment"
low_res_ct_col <- "OriginalAnnotationLevel2"
hi_res_ct_col <- "OriginalAnnotationLevel3"
```

## Create plots

#### DELETE <XXXXXXXX

```{r}
label_col <- "Tissue_compartment"

metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))

run_benchmark_analysis_test(seurat,
                            labels,
                            low_res_ct_col,
                            hi_res_ct_col)
```


```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```



# ---------------------------------------------------------------

# Adams et al. dataset
# Has also other non-immune cells

## Set paths

```{r}
ds <- "Adams"

ds_name <- "AdamsT_2020_32832599"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "Condition"
low_res_ct_col <- "OriginalAnnotationLevel1"
hi_res_ct_col <- "OriginalAnnotationLevel2"
```

## Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file,
                                            ECODA_deconv = FALSE)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------

# Kfoury et al. dataset

## Set paths

```{r}
ds <- "Kfoury"

ds_name <- "Kfoury_2021_34719426"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "Status"
low_res_ct_col <- "cells"
hi_res_ct_col <- "cells"
```

## Create plots

```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```



# ---------------------------------------------------------------

# Nathan et al. dataset
# Main separation by CD4 toxicity but no metadata available that correlates
# -> study mentions correlation with European ethnicity
# -> Corresponding author contacted
# -> Waiting for response
# Weak correlation with age

<!-- ## Set paths -->

<!-- ```{r} -->
<!-- ds <- "Nathan" -->

<!-- ds_name <- "NathanA_2021_34031617" -->

<!-- path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name) -->
<!-- path_output <- file.path(path_root, "output") -->
<!-- ``` -->


<!-- ## Load seurat object -->

<!-- ```{r} -->
<!-- seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name))) -->

<!-- # stot <- seurat -->
<!-- # tcells <- unique(seurat$celltype_subset)[grepl(pattern = "T_cells", x = unique(seurat$celltype_subset))] -->
<!-- # seurat <- subset(stot, subset = celltype_subset %in% tcells) -->

<!-- # Convert Seurat object to anndata for interoperability with python -->
<!-- convert_seurat_to_h5ad(seurat, ds) -->

<!-- # label_col <- "TB_status" -->
<!-- label_col <- "sex" -->
<!-- low_res_ct_col <- "OriginalAnnotationLevel1" -->
<!-- hi_res_ct_col <- "OriginalAnnotationLevel2" -->
<!-- ``` -->

<!-- ## Create plots -->

<!-- #### DELETE <XXXXXXXX -->

<!-- ```{r} -->
<!-- seurat$age_binned <- ifelse(seurat$age <= 30, "young", ifelse(seurat$age >= 60, "old", "middle")) -->
<!-- # seurat$age_binned <- ifelse(seurat$age <= 50, "young", "old") -->
<!-- # seurat$age_binned <- ifelse(seurat$age <= 25, "very_young", -->
<!-- #                             ifelse(seurat$age <= 50, "young", -->
<!-- #                                    ifelse(seurat$age <= 75, "middle", "old"))) -->
<!-- #  -->
<!-- # seurat$age_binned <- NULL -->
<!-- #  -->
<!-- # meta <- seurat@meta.data -->
<!-- #  -->
<!-- # # 1. Get the unique ages per sample -->
<!-- # unique_sample_ages <- meta %>% -->
<!-- #   distinct(Sample, age) %>% -->
<!-- #   pull(age) -->
<!-- #  -->
<!-- # # 2. Determine the breaks for two roughly equal sized bins based on the number of unique samples -->
<!-- # number_of_bins <- 2 -->
<!-- # age_breaks <- unique(quantile(unique_sample_ages, probs = seq.int(0, 1, by = 1 / number_of_bins))) -->
<!-- #  -->
<!-- # # 3. Assign the age bin to each unique sample -->
<!-- # sample_age_bins <- meta %>% -->
<!-- #   distinct(Sample, age) %>% -->
<!-- #   mutate(age_binned_sample = cut(age, breaks = age_breaks, include.lowest = TRUE)) %>% -->
<!-- #   dplyr::select(Sample, age_binned_sample) -->
<!-- #  -->
<!-- # # 4. Merge the sample-level age bins back into the Seurat object's metadata -->
<!-- # meta <- meta %>% -->
<!-- #   left_join(sample_age_bins, by = "Sample") %>% -->
<!-- #   # Optional: Remove the temporary 'age' column if you only want 'age_binned_sample' -->
<!-- #   # select(-age) -->
<!-- #   rename(age_binned = age_binned_sample) # Rename for consistency -->
<!-- #  -->
<!-- # # 5. Verify the number of samples in each bin -->
<!-- # meta %>% -->
<!-- #   distinct(Sample, age_binned) %>% -->
<!-- #   group_by(age_binned) %>% -->
<!-- #   count() -->
<!-- #  -->
<!-- # row.names(meta) <- row.names(seurat@meta.data) -->
<!-- # seurat@meta.data$age_binned <- meta$age_binned -->

<!-- label_col <- "age_binned" -->

<!-- metadata <- get_metadata(seurat) -->
<!-- labels <- get_labels(seurat, label_col) -->

<!-- pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]]) -->
<!-- pseudobulk_hvg <- pseudobulk[hvg,] -->
<!-- pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata)) -->

<!-- run_benchmark_analysis_test(seurat, -->
<!--                       labels, -->
<!--                       low_res_ct_col, -->
<!--                       hi_res_ct_col) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- mrvi_dist_file <- paste0("data/", ds, "_dists.nc") -->

<!-- result_list[[ds]] <- run_benchmark_analysis(seurat, -->
<!--                                       label_col, -->
<!--                                       low_res_ct_col, -->
<!--                                       hi_res_ct_col, -->
<!--                                       factors_test, -->
<!--                                       mrvi_dist_file) -->

<!-- plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]]) -->
<!-- p <- cowplot::plot_grid(plotlist = plot_list) -->
<!-- p -->
<!-- ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")), -->
<!--        plot = p, -->
<!--        width = 15, -->
<!--        height = 10) -->
<!-- ``` -->




# ---------------------------------------------------------------

# TEMP et al. dataset

## Set paths

```{r}
counts <- data.table::fread("data/raw.tsv")
counts <- as.data.frame(counts)
row.names(counts) <- counts$V1
row.names(counts) <- gsub("-", "_", row.names(counts))
counts$V1 <- NULL
counts <- Matrix(as.matrix(counts), sparse = TRUE)

meta <- data.table::fread("data/meta.tsv")
meta <- as.data.frame(meta)
meta <- meta[-1,]
row.names(meta) <- meta$NAME
row.names(meta) <- gsub("-", "_", row.names(meta))
meta$NAME <- NULL

seurat <- CreateSeuratObject(counts = t(counts), meta.data = meta)

seurat$Sample <- seurat$biosample_id

saveRDS(seurat, "data/Yasumizu.rds")
```



```{r}
ds <- "Chu"

ds_name <- "ChuY_2023_37248301"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))

# stot <- seurat
# tcells <- unique(seurat$celltype_subset)[grepl(pattern = "T_cells", x = unique(seurat$celltype_subset))]
# seurat <- subset(stot, subset = celltype_subset %in% tcells)

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "disease__ontology_label"
low_res_ct_col <- "cell_type__ontology_custom"
hi_res_ct_col <- "cell_type__ontology_custom"
```

## Create plots

#### DELETE <XXXXXXXX

```{r}
label_col <- "disease__ontology_label"

metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))

run_benchmark_analysis_test(seurat,
                            labels,
                            low_res_ct_col,
                            hi_res_ct_col)
```


```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```



# ---------------------------------------------------------------

# TEMP et al. dataset

## Set paths

```{r}
counts <- data.table::fread("data/raw.tsv")
counts <- as.data.frame(counts)
row.names(counts) <- counts$V1
row.names(counts) <- gsub("-", "_", row.names(counts))
counts$V1 <- NULL
counts <- Matrix(as.matrix(counts), sparse = TRUE)

meta <- data.table::fread("data/meta.tsv")
meta <- as.data.frame(meta)
meta <- meta[-1,]
row.names(meta) <- meta$NAME
row.names(meta) <- gsub("-", "_", row.names(meta))
meta$NAME <- NULL

seurat <- CreateSeuratObject(counts = t(counts), meta.data = meta)

seurat$Sample <- seurat$biosample_id

saveRDS(seurat, "data/Yasumizu.rds")
```



```{r}
ds <- "Chu"

ds_name <- "ChuY_2023_37248301"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))

# stot <- seurat
# tcells <- unique(seurat$celltype_subset)[grepl(pattern = "T_cells", x = unique(seurat$celltype_subset))]
# seurat <- subset(stot, subset = celltype_subset %in% tcells)

# Convert Seurat object to anndata for interoperability with python
convert_seurat_to_h5ad(seurat, ds)

label_col <- "disease__ontology_label"
low_res_ct_col <- "cell_type__ontology_custom"
hi_res_ct_col <- "cell_type__ontology_custom"
```

## Create plots

#### DELETE <XXXXXXXX

```{r}
label_col <- "disease__ontology_label"

metadata <- get_metadata(seurat)
labels <- get_labels(seurat, label_col)

pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
pb_norm <- t(DESeq2.normalize(pseudobulk_hvg, metadata))

run_benchmark_analysis_test(seurat,
                            labels,
                            low_res_ct_col,
                            hi_res_ct_col)
```


```{r}
mrvi_dist_file <- paste0("data/", ds, "_dists.nc")

result_list[[ds]] <- run_benchmark_analysis(seurat,
                                            label_col,
                                            low_res_ct_col,
                                            hi_res_ct_col,
                                            factors_test,
                                            mrvi_dist_file)

plot_list <- lapply(result_list[[ds]], function(method) method[["plot"]])
p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
       plot = p,
       width = 15,
       height = 10)
```



# ---------------------------------------------------------------
# ---------------------------------------------------------------


# Save and flatten the nested list into a DataFrame

```{r}
saveRDS(result_list, file = "data/benchmark_result_list.rds")

df_results <- imap_dfr(result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    if ("scores" %in% names(.x)) {
      imap_dfr(.x$scores, ~ {
        tibble(
          dataset = dataset_name,
          method = method_name,
          score = .y,
          value = .x
        )
      })
    } else {
      tibble() # Return an empty tibble if "scores" is not found
    }
  })
})
```


# Plot results

```{r fig.height=10, fig.width=14}
levs <- sort(as.character(unique(df_results$method)))
levs <- c(
  "ECODA_high_res",
  "ECODA_high_res_2_PCA_dims",
  "ECODA_high_res_5_PCA_dims",
  "ECODA_high_res_10_PCA_dims",
  "Pseudobulk",
  "Pseudobulk_2_PCA_dims",
  "Pseudobulk_5_PCA_dims",
  "Pseudobulk_10_PCA_dims",
  "MrVI",
  "MOFA_2_factors",
  "MOFA_5_factors",
  "MOFA_10_factors",
  "scITD_2_factors",
  "scITD_5_factors",
  "scITD_10_factors",
  "ECODA_low_res",
  "ECODA_deconv"
)

df_results$method <- factor(df_results$method, levels = levs)

df_results$dataset[df_results$dataset == "GongSharma_females_cmv"] <- "GongSharma_cmv"
df_results$dataset[df_results$dataset == "GongSharma_females_cmvneg_age"] <- "GongSharma_age"

df_results_clustering <- df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

# df_results_clustering <- df_results_clustering %>%
#   arrange(desc(mean))
df_results_clustering$method <- factor(df_results_clustering$method, levels = df_results_clustering$method)

df_results$method <- factor(df_results$method, levels = df_results_clustering$method)


df_results_modularity <- df_results %>%
  filter(score == "mod_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_modularity <- df_results_modularity %>%
  arrange(desc(mean))
df_results_modularity$method <- factor(df_results_modularity$method, levels = df_results_modularity$method)



# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "Silhouette score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- c(
  "Gong & Sharma\nSubset: females\nSeparation by: CMV infection",
  "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age",
  "Lee",
  "Stephenson",
  "Zhang")
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
# facet_grid(scales = "free",
#            labeller = labeller(score = score.labs, dataset = datas.labs),
#            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "benchmark_detailed_barplot.pdf"),
       width = 10,
       height = 6)


df_results %>%
  filter(score == "mod_score" | score == "cluster_score") %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_mod_clust.pdf"),
       width = 2.5,
       height = 4)



df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_bars.pdf"),
       width = 5,
       height = 7)

ggplot(df_results, aes(x = method, y = value)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores.pdf"),
       width = 7,
       height = 7)



df_results_clustering %>%
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_clustering)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)


ggplot(df_results_clustering, aes(x = method, y = value)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Adjusted Rand Index") +
  stat_compare_means(method = "anova", label.x.npc = 0.1, label.y.npc = 0.9) +
  # stat_compare_means(
  #   comparisons = my_comparisons,
  #   method = "t.test", 
  #   paired = F,
  #   p.adjust.method = "BH" #,
  #   # symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), symbols = c("****", "***", "**", "*", "ns"))
  # ) +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #   y.position = c(max(df_results_clustering$values)+0.1, max(df_results_clustering$values)+0.2), hide.ns = TRUE) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_clustering.pdf"),
       width = 6,
       height = 4.5)

df_results_modularity %>%
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Modularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity.pdf"),
       width = 2.5,
       height = 2.5)

```

# Show example PCAs for publication

```{r fig.height=6, fig.width=12}

levs <- c(
  "ECODA_high_res",
  "ECODA_low_res",
  "ECODA_deconv",
  "Pseudobulk",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI"
)


plot_list <- lapply(result_list[["GongSharma_females_cmv"]], function(method) method[["plot"]])
scores <- lapply(result_list[["GongSharma_females_cmv"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_females_cmv.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["GongSharma_females_cmvneg_age"]], function(method) method[["plot"]])
scores <- lapply(result_list[["GongSharma_females_cmvneg_age"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_females_cmvneg_age.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["Lee"]], function(method) method[["plot"]])
scores <- lapply(result_list[["Lee"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Lee.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["Zhang"]], function(method) method[["plot"]])
scores <- lapply(result_list[["Zhang"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Zhang.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(result_list[["Stephenson"]], function(method) method[["plot"]])
scores <- lapply(result_list[["Stephenson"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Stephenson.pdf"),
       plot = p,
       width = 15,
       height = 7)

```
