---
title: "ZhangY_2022_34653365 dataset - broad cell type clustering with scGate"
author: "Massimo Andreatta, Santiago Carmona"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script compares different methods for exploratory data analysis of scRNA-seq data.


# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(factoextra)
library(plotly)
library(Hotelling)
library(compositions)
source("functions.R")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)

# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
library(ncdf4)
library(arrow)

my_seed <- 123
set.seed(my_seed)
```



# Set path to save plots

```{r}
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Zhang et al. dataset

## Set paths


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Stephenson et al. dataset

## Set paths


# Set paths

```{r}
ds_name <- "MitchelJ_2023_PrePrintTBD"
ID <- "MitJ"
species <- "human"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_data <- file.path(path_root, "data")
path_output <- file.path(path_root, "output")
```


### Load seurat object

```{r}
seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))

seurat <- subset(seurat, subset = Site == "Ncl")
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)

seurat$sample_id <- as.character(seurat$sample_id)
samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% c("BGCV10_CV0198", "MH8919230")])
seurat <- subset(seurat, subset = sample_id %in% samples_subset)

seurat$Sample <- seurat$sample_id
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features

label_condition <- "Status"
low_res_ct_col <- "initial_clustering"
hi_res_ct_col <- "full_clustering"

# Convert Seurat object to anndata for interoperability with python
SaveH5Seurat(seurat, filename = "mitchel_clearcut.h5Seurat")
Convert("mitchel_clearcut.h5Seurat", dest = "h5ad")
```

### Create plots

```{r}
plot_list <- list()
labels <- get_labels(seurat, label_condition)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat_object, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
plot_list[["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg, labels, hvg)


# Deconvolute using EPIC
plot_list[["deconv"]] <- process_deconv_fig(pseudobulk, labels)

# CoDA
## layer1: low res. cell types
plot_list[["coda_lowres"]] <- process_coda_fig(seurat, labels, low_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
plot_list[["coda_hires"]] <- process_coda_fig(seurat, labels, hi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
plot_list[["mofa"]] <- process_mofa_fig(seurat, labels, hvg, pseudobulk_hvg, file.path("mitchel_mofa_object.rds"))

# scITD
plot_list[["scitd"]] <- process_scitd_fig(seurat, low_res_ct_col, label_condition, hvg)

# MrVI
plot_list[["mrvi"]] <- process_mrvi_fig(seurat, "./mitchel_dists.nc", low_res_ct_col, label_condition)


p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_mitchel.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Gong and Sharma et al. dataset

## Set paths

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"
ID <- "GoSH"
species <- "human"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
dir.create(path_root, showWarnings = F)
path_data <- file.path(path_root, "data")
dir.create(path_data, showWarnings = F)
path_plots <- file.path(path_root, "plots")
dir.create(path_plots, showWarnings = F)
path_output <- file.path(path_root, "output")
dir.create(path_output, showWarnings = F)
```


## All CMV

### Load seurat object

```{r}
seurat <- load_seurat_object("gongsharma_all.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```

### Create plots

```{r}
label_condition <- "subject.cmv"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

plot_list <- list()
labels <- get_labels(seurat, label_condition)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat_object, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
plot_list[["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg, labels, hvg)


# Deconvolute using EPIC
plot_list[["deconv"]] <- process_deconv_fig(pseudobulk, labels)

# CoDA
## layer1: low res. cell types
plot_list[["coda_lowres"]] <- process_coda_fig(seurat, labels, low_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
plot_list[["coda_hires"]] <- process_coda_fig(seurat, labels, hi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
plot_list[["mofa"]] <- process_mofa_fig(seurat, labels, hvg, pseudobulk_hvg, file.path("gongsharma_all_mofa_object.rds"))

# scITD
plot_list[["scitd"]] <- process_scitd_fig(seurat, low_res_ct_col, label_condition, hvg)

# MrVI
plot_list[["mrvi"]] <- process_mrvi_fig(seurat, "./gongsharma_all_dists.nc", low_res_ct_col, label_condition)


p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_all_cmv.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## All gender

### Create plots

```{r}
label_condition <- "subject.biologicalSex"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

plot_list <- list()
labels <- get_labels(seurat, label_condition)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
plot_list[["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg, labels, hvg)


# Deconvolute using EPIC
plot_list[["deconv"]] <- process_deconv_fig(pseudobulk, labels)

# CoDA
## layer1: low res. cell types
plot_list[["coda_lowres"]] <- process_coda_fig(seurat, labels, low_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
plot_list[["coda_hires"]] <- process_coda_fig(seurat, labels, hi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
plot_list[["mofa"]] <- process_mofa_fig(seurat, labels, hvg, pseudobulk_hvg, file.path("gongsharma_all_mofa_object.rds"))

# scITD
plot_list[["scitd"]] <- process_scitd_fig(seurat, low_res_ct_col, label_condition, hvg)

# MrVI
plot_list[["mrvi"]] <- process_mrvi_fig(seurat, "./gongsharma_all_dists.nc", low_res_ct_col, label_condition)


p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_all_gender.pdf"),
       plot = p,
       width = 15,
       height = 10)
```




## Gender (CMV negative)

### Load seurat objects

```{r}
seurat <- load_seurat_object("gongsharma_gender_cmvneg.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```


### Create plots

```{r}
label_condition <- "subject.biologicalSex"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

plot_list <- list()
labels <- get_labels(seurat, label_condition)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
plot_list[["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg, labels, hvg)


# Deconvolute using EPIC
plot_list[["deconv"]] <- process_deconv_fig(pseudobulk, labels)

# CoDA
## layer1: low res. cell types
plot_list[["coda_lowres"]] <- process_coda_fig(seurat, labels, low_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
plot_list[["coda_hires"]] <- process_coda_fig(seurat, labels, hi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
plot_list[["mofa"]] <- process_mofa_fig(seurat, labels, hvg, pseudobulk_hvg, file.path("gongsharma_gender_cmvneg_mofa_object.rds"))

# scITD
plot_list[["scitd"]] <- process_scitd_fig(seurat, low_res_ct_col, label_condition, hvg)

# MrVI
plot_list[["mrvi"]] <- process_mrvi_fig(seurat, "./gongsharma_gender_cmvneg_dists.nc", low_res_ct_col, label_condition)


p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_gender_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## CMV (females)

### Load seurat objects

```{r}
seurat <- load_seurat_object("gongsharma_cmv_females.h5ad")

seurat$Sample <- seurat$specimen.specimenGuid
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```



### Create plots

```{r}
label_condition <- "subject.cmv"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

plot_list <- list()
labels <- get_labels(seurat, label_condition)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
plot_list[["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg, labels, hvg)

# Deconvolute using EPIC
plot_list[["deconv"]] <- process_deconv_fig(pseudobulk, labels)

# CoDA
## layer1: low res. cell types
plot_list[["coda_lowres"]] <- process_coda_fig(seurat, labels, low_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
plot_list[["coda_hires"]] <- process_coda_fig(seurat, labels, hi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
seurat <- UpdateSeuratObject(seurat)
plot_list[["mofa"]] <- process_mofa_fig(seurat, labels, hvg, pseudobulk_hvg, file.path("gongsharma_cmv_females_mofa_object.rds"))

# scITD
plot_list[["scitd"]] <- process_scitd_fig(seurat, low_res_ct_col, label_condition, hvg)

# MrVI
plot_list[["mrvi"]] <- process_mrvi_fig(seurat, "./gongsharma_cmv_females_dists.nc", low_res_ct_col, label_condition)


p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_cmv_females.pdf"),
       plot = p,
       width = 15,
       height = 10)
```


## Age (females cmvneg)

### Load seurat objects

```{r}
seurat <- load_seurat_object("gongsharma_age_females_cmvneg.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))
seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000)
hvg <- seurat@assays[["RNA"]]@var.features
```



### Create plots

```{r}
label_condition <- "age"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

plot_list <- list()
labels <- get_labels(seurat, label_condition)

# Pseudobulk
pseudobulk <- as.matrix(AggregateExpression(seurat, group.by = "Sample", assays = "RNA")[["RNA"]])
pseudobulk_hvg <- pseudobulk[hvg,]
plot_list[["pseudobulk"]] <- process_pseudobulk_fig(pseudobulk_hvg, labels, hvg)


# Deconvolute using EPIC
plot_list[["deconv"]] <- process_deconv_fig(pseudobulk, labels)

# CoDA
## layer1: low res. cell types
plot_list[["coda_lowres"]] <- process_coda_fig(seurat, labels, low_res_ct_col, title = "Cell type composition - low res. cell types")

## layer2: high res. cell types
plot_list[["coda_hires"]] <- process_coda_fig(seurat, labels, hi_res_ct_col, title = "Cell type composition - high res. cell types")

# MOFA
seurat <- UpdateSeuratObject(seurat)
plot_list[["mofa"]] <- process_mofa_fig(seurat, labels, hvg, pseudobulk_hvg, file.path("gongsharma_age_females_cmvneg_mofa_object.rds"))

# scITD
plot_list[["scitd"]] <- process_scitd_fig(seurat, low_res_ct_col, label_condition, hvg)

# MrVI
plot_list[["mrvi"]] <- process_mrvi_fig(seurat, "./gongsharma_age_females_cmvneg_dists.nc", low_res_ct_col, label_condition)


p <- cowplot::plot_grid(plotlist = plot_list)
p
ggsave(file.path(path_plots, "benchmark_pca_gongsharma_age_females_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 10)
```
