---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
source("functions.R")
source("Processed_dataset_metadata.R")

library(forcats)

# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
py_install("mofapy2", envname = "my_env", method="auto")
import("mofapy2")
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
# install.packages("GloScope")
library(GloScope)

# Set paths
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"

set.seed(123)
```



# ---------------------------------------------------------------

# Colorectal cancer from Joanito et al.

Single-cell and bulk transcriptome sequencing identifies two epithelial tumor cell states and refines the consensus molecular classification of colorectal cancer

Data from:
https://www.nature.com/articles/s41588-022-01100-4


## Load data

```{r fig.height=8, fig.width=8}
seurat <- readRDS("/Volumes/Shared/DataCollections/Standardized_SingleCell_Datasets/JoaI_2022_35773407/output/JoaI_2022_35773407_Nofilt_whole.rds")

seurat@meta.data$cell.type_new <- seurat@meta.data$cell.type
iCMS <- !is.na(seurat@meta.data$iCMS)
seurat@meta.data$cell.type_new[iCMS] <- paste0(seurat@meta.data$cell.type[iCMS], "_", ifelse(seurat@meta.data$iCMS[iCMS] == "Normal", "Normal", "Cancer"))

meta <- seurat@meta.data %>%
  group_by(sample.ID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample.ID
meta$sample.origin <- gsub("-2", "", meta$sample.origin)


pb <- get_pb(seurat, sample_col = "sample.ID")
metadata <- get_metadata(seurat, sample_col = "sample.ID")
pb_norm <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))


feat_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  clr()

plot_pca(feat_mat, labels = meta$sample.origin)
plot_pca(feat_mat, labels = meta$dataset)


plot_pca(pb_norm, labels = meta$sample.origin)
plot_pca(pb_norm, labels = meta$dataset)
# plot_pca(pb_norm, labels = meta$Gender)

prime3 <- meta$dataset %in% c("SMC", "KUL3", "CRC-SG2")
prime5 <- meta$dataset %in% c("CRC-SG1", "KUL5")
plot_pca(pb_norm[prime3, ], labels = meta$sample.origin[prime3])
plot_pca(pb_norm[prime5, ], labels = meta$sample.origin[prime5])

plot_pca(feat_mat[prime3, ], labels = meta$sample.origin[prime3])
plot_pca(feat_mat[prime5, ], labels = meta$sample.origin[prime5])
```




# ---------------------------------------------------------------

# Brain from Macnair et al.

snRNA-seq stratifies multiple sclerosis patients into distinct white matter glial responses

Data from:
https://www.sciencedirect.com/science/article/pii/S0896627324008730#sec8



## Load data

```{r}
library(readxl)

df_mac_main <- data.frame(fread(file.path(path_data, "other", "MacnairW_2025_Neuron", "ms_lesions_snRNAseq_col_data_2023-09-12.txt.gz")))
df_mac_val <- data.frame(fread(file.path(path_data, "other", "MacnairW_2025_Neuron", "ms_lesions_snRNAseq_validation_col_data_2023-09-12.txt.gz")))

feat_mat <- as.data.frame.matrix(table(df_mac_main$sample_id_anon, df_mac_main$type_fine)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_mac_main %>%
    group_by(sample_id_anon) %>%
    slice(1)
row.names(meta) <- meta$sample_id_anon

p <- plot_pca(feat_mat, labels = meta$matter)


rsub <- meta$matter == "GM"
feat_mat_gm <- feat_mat[rsub, ]
meta_gm <- meta[rsub, ]

p <- plot_pca(feat_mat_gm, labels = meta_gm$lesion_type)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$seq_pool)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$sex)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$sample_source)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$diagnosis)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$age_cat)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$pmi_cat)


rsub <- meta$matter == "WM"
feat_mat_wm <- feat_mat[rsub, ]
meta_wm <- meta[rsub, ]

p <- plot_pca(feat_mat_wm, labels = meta_wm$lesion_type)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$seq_pool)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$sex)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$sample_source)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$diagnosis)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$age_cat)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$pmi_cat)
```


# ---------------------------------------------------------------

# Tumor immune cell atlas from Nieto et al.

A single-cell tumor immune atlas for precision oncology

Data from:
https://genome.cshlp.org/content/31/10/1913.long#sec-13


## Load data

```{r fig.height=12, fig.width=12}
df_ni <- data.frame(fread(file.path(path_data, "other", "TICAtlas_metadata.csv")))


feat_mat <- as.data.frame.matrix(table(df_ni$patient, df_ni$lv1_annot)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_ni %>%
    group_by(patient) %>%
    slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$patient

meta$kmeans_cluster[is.na(meta$kmeans_cluster)] <- 7

p <- plot_pca(feat_mat, labels = meta$source)
# p <- plot_pca(feat_mat, labels = meta$subtype, plotly_3d = TRUE)


p <- plot_pca(feat_mat, labels = meta$subtype, invisible = "none", labelsize = 4)


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(meta$subtype),
  dataset = as.factor(meta$source)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)






feat_mat <- as.data.frame.matrix(table(df_ni$patient, df_ni$lv2_annot)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_ni %>%
    group_by(patient) %>%
    slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$patient

meta$kmeans_cluster[is.na(meta$kmeans_cluster)] <- 7

p <- plot_pca(feat_mat, labels = meta$source)
# p <- plot_pca(feat_mat, labels = meta$subtype, plotly_3d = TRUE)


p <- plot_pca(feat_mat, labels = meta$subtype, invisible = "none", labelsize = 4)


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(meta$subtype),
  dataset = as.factor(meta$source)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```




# ---------------------------------------------------------------

# NSCLC response atlas from Liu et al.

A single-cell atlas reveals immune heterogeneity in anti-PD-1-treated non-small cell lung cancer

Data from:
https://www.cell.com/cell/fulltext/S0092-8674(25)00291-0#sec-4


## Load data

```{r fig.height=12, fig.width=12}
df <- data.frame(fread(file.path(path_data, "other", "GSE243013_NSCLC_immune_scRNA_metadata.csv.gz")))

# Keep only samples with a reasonable amount of cells
samples_keep <- sort(table(df$sampleID)) > 1000
samples_keep_names <- names(samples_keep[samples_keep])
df <- df[df$sampleID %in% samples_keep_names, ]

feat_mat <- as.data.frame.matrix(table(df$sampleID, df$sub_cell_type)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  # impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  calc_perc_df() %>%
  clr()

meta <- df %>%
  group_by(sampleID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sampleID

p <- plot_pca(feat_mat, labels = meta$gender, plotly_3d = TRUE)

p <- plot_pca(feat_mat, labels = meta$pathological_response)
p <- plot_pca(feat_mat, labels = meta$pathological_response, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$pathological_response, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$pathological_response, invisible = "none", labelsize = 4, axes = c(1,3))

p <- plot_pca(feat_mat, labels = meta$radiological_response)
p <- plot_pca(feat_mat, labels = meta$radiological_response, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4, axes = c(1,3))


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  response = as.factor(meta$pathological_response),
  cancer_type = as.factor(meta$cancer_type),
  gender = as.factor(meta$gender),
  stage = as.factor(meta$pre_treatment_staging),
  smoking = as.factor(meta$smoking_history)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, grepl("CD4|CD8", colnames(feat_mat))]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "B")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "M")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```


```{r fig.width=18}
df_plot <- feat_mat
df_plot$sampleID <- meta$sampleID
df_plot$pathological_response <- meta$pathological_response
df_plot <- df_plot %>%
  pivot_longer(
    cols = colnames(feat_mat),
    names_to = "cts",
    values_to = "values"
  )
ggplot(df_plot, aes(x = cts, y = values, fill = pathological_response)) +
  geom_boxplot(outlier.shape = NA) +
  # stat_compare_means(
  #   # comparisons = my_comparisons,
  #   # ref.group = "Pseudobulk",
  #   method = "t.test",
  #   # paired = TRUE,
  #   angle = 15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```




# ---------------------------------------------------------------

# BCC and SCC response from Yost et al.

Clonal replacement of tumor-specific T cells following PD-1 blockade

Data from:
https://www.nature.com/articles/s41591-019-0522-3#data-availability


## Load data

```{r fig.height=12, fig.width=12}
df1 <- data.frame(fread(file.path(path_data, "other", "GSE123813_bcc_all_metadata.txt.gz")))
df1$sort <- NULL
df1$type <- "bcc"
df1$sample <- paste0(df1$patient, "_", df1$treatment)
df2 <- data.frame(fread(file.path(path_data, "other", "GSE123813_scc_metadata.txt.gz")))
df2$type <- "scc"
df2$sample <- paste0(df2$patient, "_", df2$treatment)

df <- rbind(df1, df2)

sort(table(df$sample))

table(df$sample, df$cluster)

feat_mat <- as.data.frame.matrix(table(df$sample, df$cluster)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  # impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  calc_perc_df() %>%
  clr()

meta <- df %>%
  group_by(sample) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample


p <- plot_pca(feat_mat, labels = meta$type)
p <- plot_pca(feat_mat, labels = meta$treatment)

p <- plot_pca(feat_mat, labels = meta$type, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$type, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$type, invisible = "none", labelsize = 4, axes = c(1,3))

# p <- plot_pca(feat_mat, labels = meta$radiological_response)
# p <- plot_pca(feat_mat, labels = meta$radiological_response, plotly_3d = TRUE)
# p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4)
# p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4, axes = c(1,3))


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  response = as.factor(meta$pathological_response),
  cancer_type = as.factor(meta$cancer_type),
  gender = as.factor(meta$gender),
  stage = as.factor(meta$pre_treatment_staging),
  smoking = as.factor(meta$smoking_history)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, grepl("CD4|CD8", colnames(feat_mat))]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "B")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "M")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```



# ---------------------------------------------------------------

# CART from Haradhvala et al.

Clonal replacement of tumor-specific T cells following PD-1 blockade

Data from:
https://www.nature.com/articles/s41591-019-0522-3#data-availability


## Load data

```{r}
seurat <- readRDS("/Users/christianhalter/Library/CloudStorage/Dropbox/CSI/Standardized_SingleCell_Datasets/HarN_2022_36097221/output/HarN_2022_36097221whole.rds")

library(HiTME)

# Fetch reference maps
ref.maps <- get.reference.maps(collection = "human")

# Cell type classification on a Seurat object
query <- Run.HiTME(query,
                   ref.maps = ref.maps[["human"]],
                   species = "human" # for mouse data set to "mouse"
                   )

```


# ---------------------------------------------------------------

# Explanation of negative ARI and ANOSIM scores

```{r}
# --- Load necessary packages ---
library(vegan)
library(mclust)
library(ggplot2)


# --- Main function to analyze clusters and plot results ---
analyze_clusters <- function(n_samples, cluster_specs) {
  # Generate the data for all clusters
  data_list <- list()
  true_labels <- c()
  group_labels <- c()
  for (i in 1:length(cluster_specs)) {
    spec <- cluster_specs[[i]]
    group_name <- LETTERS[i] # Assign simple group names A, B, C, ...
    
    # Generate the descriptive label for the legend
    mean_str <- paste0("Mean: (", paste(spec$mean, collapse = ", "), ")")
    sd_str <- paste0("SD: (", paste(spec$sd, collapse = ", "), ")")
    descriptive_label <- paste(group_name, mean_str, sd_str, sep = " | ")

    # Generate the data
    if (length(spec$sd) == 1) {
      cluster_data <- data.frame(
        x = rnorm(n_samples, mean = spec$mean[1], sd = spec$sd),
        y = rnorm(n_samples, mean = spec$mean[2], sd = spec$sd)
      )
    } else {
      cluster_data <- data.frame(
        x = rnorm(n_samples, mean = spec$mean[1], sd = spec$sd[1]),
        y = rnorm(n_samples, mean = spec$mean[2], sd = spec$sd[2])
      )
    }

    cluster_data$group <- group_name
    data_list[[i]] <- cluster_data
    true_labels <- c(true_labels, rep(group_name, n_samples))
    group_labels <- c(group_labels, rep(descriptive_label, n_samples))
  }
  
  # Combine data into a single data frame
  data_df <- do.call(rbind, data_list)

  # Add the descriptive labels to the dataframe
  data_df$group_label <- factor(group_labels, levels = unique(group_labels))

  # Generate shuffled labels for both statistical test and plot
  shuffled_labels <- sample(true_labels)
  shuffled_group_labels <- sample(group_labels) # This was the flawed part. Corrected below.
  
  # The fix: create a shuffled version of the descriptive labels that correctly corresponds to the shuffled simple labels
  shuffled_labels_map <- setNames(group_labels, true_labels)
  shuffled_group_labels <- as.character(shuffled_labels_map[shuffled_labels])

  # Calculate distance matrix
  dist_matrix <- dist(data_df[, c("x", "y")], method = "euclidean")
  
  # Calculate ANOSIM and ARI for correct labels
  anosim_correct <- anosim(dist_matrix, true_labels)
  ari_correct <- adjustedRandIndex(true_labels, true_labels)

  # Calculate ANOSIM and ARI for shuffled labels
  anosim_shuffled <- anosim(dist_matrix, shuffled_labels)
  ari_shuffled <- adjustedRandIndex(true_labels, shuffled_labels)

  # Create plot titles with rounded metrics
  title_correct <- paste0(
    "Correct Labels\n",
    "ANOSIM R: ", round(anosim_correct$statistic, 4),
    " | ARI: ", round(ari_correct, 4)
  )
  
  title_shuffled <- paste0(
    "Shuffled Labels\n",
    "ANOSIM R: ", round(anosim_shuffled$statistic, 4),
    " | ARI: ", round(ari_shuffled, 4)
  )
  
  # Create plots
  plot_correct <- ggplot(data_df, aes(x = x, y = y, color = group_label)) +
    geom_point(size = 2) +
    labs(title = title_correct, x = "Feature 1", y = "Feature 2", color = "Group Specs") +
    theme_minimal()
  print(plot_correct)
  
  # The fix: create the shuffled dataframe by re-assigning the labels to the original data points
  data_df_shuffled_plot <- data_df
  data_df_shuffled_plot$group_label <- factor(shuffled_group_labels, levels = unique(group_labels))
  
  plot_shuffled <- ggplot(data_df_shuffled_plot, aes(x = x, y = y, color = group_label)) +
    geom_point(size = 2) +
    labs(title = title_shuffled, x = "Feature 1", y = "Feature 2", color = "Group Specs") +
    theme_minimal()
  print(plot_shuffled)
  
  # Create a data frame to store the results
  results_df <- data.frame(
    Scenario = c("Correct", "Shuffled"),
    Anosim_R = c(anosim_correct$statistic, anosim_shuffled$statistic),
    ARI = c(ari_correct, ari_shuffled)
  )
  
  return(results_df)
}


# --- Define Scenarios ---
set.seed(4)



# Scenario 1: Well-separated clusters
n_samples <- 50
cluster_specs <- list(
  list(mean = c(0, 0), sd = 1),
  list(mean = c(5, 5), sd = 1)
)

# --- Run and print results for each scenario ---
cat("--- Scenario 1: Well-Separated Clusters (50 samples) ---\n")
results1 <- analyze_clusters(n_samples, cluster_specs)
print(results1)


# Scenario 2: Well-separated clusters few samples
n_samples <- 5
cluster_specs <- list(
  list(mean = c(0, 0), sd = 1),
  list(mean = c(5, 5), sd = 1)
)

# --- Run and print results for each scenario ---
cat("--- Scenario 2: Well-separated clusters (10 samples)---\n")
results2 <- analyze_clusters(n_samples, cluster_specs)
print(results2)


# Scenario 3: Well-separated clusters few samples
n_samples <- 2
cluster_specs <- list(
  list(mean = c(0, 0), sd = 1),
  list(mean = c(5, 5), sd = 1)
)

# --- Run and print results for each scenario ---
cat("--- Scenario 3: Well-separated clusters (4 samples) ---\n")
results3 <- analyze_clusters(n_samples, cluster_specs)
print(results3)


```


```{r}
# --- Load necessary packages ---
library(vegan)
library(mclust)
library(ggplot2)

# --- Define scenario parameters ---
# set.seed(42)
n_samples <- 2 # Number of samples per cluster

# Create two clearly separated Gaussian clusters
cluster1 <- data.frame(
  x = rnorm(n_samples, mean = 0, sd = 1),
  y = rnorm(n_samples, mean = 0, sd = 1)
)
cluster2 <- data.frame(
  x = rnorm(n_samples, mean = 5, sd = 1),
  y = rnorm(n_samples, mean = 5, sd = 1)
)

# Combine the data
data_df <- rbind(cluster1, cluster2)

# Create the correct, true labels
true_labels <- factor(c(rep("Cluster A", n_samples), rep("Cluster B", n_samples)))

# Create the "worst-case" labels by assigning them in an alternating pattern
worst_labels <- factor(rep(c("Worst A", "Worst B"), length.out = nrow(data_df)))

# --- Plot the scenarios ---
# Plot for Correct Labels
plot_correct <- ggplot(data_df, aes(x = x, y = y, color = true_labels)) +
  geom_point(size = 2) +
  labs(title = "Data with Correct Cluster Labels", x = "Feature 1", y = "Feature 2") +
  theme_minimal()
print(plot_correct)

# Plot for Worst-Case Labels
plot_worst <- ggplot(data_df, aes(x = x, y = y, color = worst_labels)) +
  geom_point(size = 2) +
  labs(title = "Data with Worst-Case (Alternating) Labels", x = "Feature 1", y = "Feature 2") +
  theme_minimal()
print(plot_worst)

# --- Calculate ANOSIM and ARI ---
# Calculate Euclidean distance matrix
dist_matrix <- dist(data_df[, c("x", "y")], method = "euclidean")

# ANOSIM: Test for separation based on correct vs. worst-case groups
cat("--- ANOSIM Results ---\n")
anosim_correct <- anosim(dist_matrix, true_labels)
cat("ANOSIM R-statistic (Correct Labels):", round(anosim_correct$statistic, 4), "\n")
# A high positive value indicates strong separation.

anosim_worst <- anosim(dist_matrix, worst_labels)
cat("ANOSIM R-statistic (Worst-Case Labels):", round(anosim_worst$statistic, 4), "\n\n")
# This value should be highly negative, indicating that the within-group
# dissimilarity is much greater than the between-group dissimilarity.

# ARI: Compare the worst-case labels to the correct labels
cat("--- Adjusted Rand Index Results ---\n")
# A perfect clustering (correct vs correct) gives a score of 1.
ari_perfect <- adjustedRandIndex(true_labels, true_labels)
cat("ARI (Perfect Match):", round(ari_perfect, 4), "\n")

# Comparing the correct labels to the worst-case labels should
# yield a highly negative score, indicating the worst possible agreement.
ari_worst <- adjustedRandIndex(true_labels, worst_labels)
cat("ARI (Correct vs. Worst-Case):", round(ari_worst, 4), "\n")
```


# Export cell type compositions

```{r}
get_sample_metadata <- function(cell_data_df, sample_col) {
  # Step 1: Group the data by `sample_col` and check for uniqueness
  distinct_counts <- cell_data_df %>%
    dplyr::group_by(!!rlang::sym(sample_col)) %>%
    dplyr::summarise(across(everything(), ~ n_distinct(.x))) %>%
    dplyr::ungroup()

  # Step 2: Identify columns that are constant across all samples
  constant_cols <- distinct_counts %>%
    dplyr::select(where(~ max(.x) == 1)) %>%
    names()

  # Step 3: Select and return the constant columns
  metadata <- cell_data_df %>%
    dplyr::select(all_of(c(sample_col, constant_cols))) %>%
    dplyr::distinct() %>%
    as.data.frame() # Convert back to data.frame

  # Set row names to the sample column for easier access
  rownames(metadata) <- metadata[[sample_col]]
  metadata[[sample_col]] <- NULL

  return(metadata)
}


ds_data_list <- list()

for (ds in names(datasets)) {
  print(ds)
  
  seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
  seurat@misc$label_col <- datasets[[ds]][["label_col"]]
  
  ds_data_list[[ds]] <- list(
    cell_counts_highresolution = get_ct_comp_df_seurat(
      seurat,
      sample_col = "Sample",
      ct_col = datasets[[ds]][["hi_res_ct_col"]]
    ),
    cell_counts_lowresolution = get_ct_comp_df_seurat(
      seurat,
      sample_col = "Sample",
      ct_col = datasets[[ds]][["low_res_ct_col"]]
    ),
    metadata = get_sample_metadata(seurat@meta.data, "Sample"),
    main_biologicalcondition_columnname = datasets[[ds]][["label_col"]]
  )
}
  
saveRDS(
  object = ds_data_list,
  file = file.path(path_data, "example_data.rds")
)
```
