---
title: "Compositional data analysis of the Gong & Sharma et al. dataset"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

Using exploratory compositional data analysis on large healthy data set

paper: <https://www.biorxiv.org/content/10.1101/2024.09.10.612119v1> accession: TBD

This script takes output of "<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>" see "Cell type frequency data"


# For additional sample metadata see: "<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>"

# General study description from authors:
```{r}
# Immune aging is a dynamic process driven by complex internal and external interactions. We examined the healthy human immune system across age using deep molecular profiling .

# To do so, we used our Human Immune Health Atlas to label peripheral blood mononuclear cells (PBMCs) from a prospective, longitudinally sampled cohort of healthy young adults (n=49, 25-35 years) and older adults (n=47, 55-65 years) followed over the course of 2-3 years. This cohort was observed under homeostatic conditions, as well as prior to, and at 7 and 90 days following, annual influenza vaccinations, resulting in a scRNA-seq dataset with over 13.5 million PBMCs.

# Our study highlights that gradual, age-related alterations in the homeostatic transcriptional networks of immune cells lead to shifts in immune responsiveness as we age.
```

# Cell type data description
```{r}
# Cell type frequency data
# After labeling cell types, we tabulated the cell count for each of the 868 samples utilized in our study at each level of resolution in our Immune Health Atlas. Below, we provide these cell counts, the fraction of counts for each sample, and the centered log ratio (CLR) transformation of those fractions that we utilized for our analyses.

# In addition, we utilized the absolute lymphocyte counts (ALC) provided in our clinical blood count lab results for each sample (available below) to compute estimated cell type abundance based on normalization to ALC. For more details, see our Data Analysis Methods.
```

# Metadata columns description
```{r}
# cohort.cohortGuid: Cohort ID (BR1 or BR2)
# subject.subjectGuid: Subject ID
# subject.biologicalSex: Subject Sex (Female or Male)
# subject.cmv: Subject CMV Status (Negative or Positive)
# subject.bmi: Subject BMI (integer)
# subject.race: Subject race
# subject.ethnicity: Subject ethnicity
# subject.birthYear: Subject Birth Year
# subject.ageAtFirstDraw: Subject Age at earliest blood draw in study
# sample.sampleKitGuid: Sample Kit ID
# sample.visitName: Sample Visit Name
# sample.drawDate: Sample Draw Date (Year-Month)
# sample.subjectAgeAtDraw: Subject age at time of draw, based on year of Draw Date and Birth Year
# specimen.specimenGuid: Specimen ID (pbmc_sample_id in .h5 files)

# Frequency-related columns
# (for AIFI_L1 as an example; AIFI_L1 is replaced with AIFI_L2 and AIFI_L3 for those levels)

# AIFI_L1: Cell Type assignment
# AIFI_L1_count: Count of cells within this sample with cell type assignment
# total_cells: Total cells within this sample
# scrna.lymphocyte_count: Sum of T, NK, and B cells
# bc.lymphocyte_count: Absolute Lymphocyte Count (ALC) from clinical Blood Counts (bc.)
# alc_ratio: ALC per scRNA Lymphocyte Count
# AIFI_L1_frac_total: Fraction of cells with cell type assignment divided by Total cells for this sample
# AIFI_L1_alc: ALC estimate for this cell type assignment
# AIFI_L1_clr: Centered Log Ratio computed using AIFI_L1_frac_total for all types within this sample
```



# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

# install.packages("/Users/christianhalter/Desktop/scooter_1.0.0.tar.gz", repos = NULL, type = "source")

library(scooter)
library(tidyverse)
library(plotly)
library(ggplot2)
library(ggpubr)
library(factoextra)
library(compositions)
library(pROC)
library(mclust)
library(Rfast)
library(zoo)
library(pheatmap)
source("functions.R")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)

# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
library(ncdf4)
library(arrow)

library(dplyr)

my_seed <- 123
set.seed(my_seed)
```

# Set paths

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets", ds_name)
path_data <- file.path(path_root, "data")
path_plots <- file.path("/Users/christianhalter/Library/CloudStorage/OneDrive-SharedLibraries-UniversitédeGenève/Santiago Carmona - CSI/LabPapers/ECODA/plots")
```


# Load dataset
Manually downloaded from:
"<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>"

```{r}
files <- list(L1 = file.path(path_data, "sound_life_AIFI_L1_frequencies.csv"),
              L2 = file.path(path_data, "sound_life_AIFI_L2_frequencies.csv"),
              L3 = file.path(path_data, "sound_life_AIFI_L3_frequencies.csv"))

data <- list()

for (f in names(files)) {
  data[[f]] <- read.csv(files[[f]])
}
```


# Cell types per layer

```{r}
t1 <- as.data.frame.matrix(t(table(data$L3$AIFI_L1, data$L3$AIFI_L2)))

for (i in ncol(t1):1) {
  t1 <- t1[order(t1[, i], decreasing = TRUE), ]
}

colnames(t1) <- str_replace_all(colnames(t1), " ", "_")
row.names(t1) <- str_replace_all(row.names(t1), " ", "_")

t2 <- as.data.frame.matrix(t(table(data$L3$AIFI_L2, data$L3$AIFI_L3)))

for (i in ncol(t2):1) {
  t2 <- t2[order(t2[, i], decreasing = TRUE), ]
}

colnames(t2) <- str_replace_all(colnames(t2), " ", "_")
row.names(t2) <- str_replace_all(row.names(t2), " ", "_")
```


# Plot number of cells per sample

```{r}
for (layer in names(data)) {
  df <- data[[layer]] %>%
    dplyr::select(paste0("AIFI_", layer),
           paste0("AIFI_", layer, "_frac_total"),
           paste0("total_cells"),
           "specimen.specimenGuid") %>%
    group_by(specimen.specimenGuid) %>%
    reframe(cells_sum = min(total_cells),
            annotated_cells_perc = sum(!!sym(paste0("AIFI_", layer, "_frac_total"))))
  
  p <- ggplot(data = df, aes(x = reorder(specimen.specimenGuid, cells_sum), y = cells_sum)) +
    geom_bar(stat="identity", fill="steelblue") +
    xlab("") +
    ylab("Cells") +
    theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
    ggtitle(paste("Cell counts per sample - Layer: ", layer))
  print(p)
  
  p <- ggplot(data = df, aes(x = reorder(specimen.specimenGuid, cells_sum), y = annotated_cells_perc)) +
    geom_bar(stat="identity", fill="steelblue") +
    xlab("") +
    ylab("Cells") +
    theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
    ggtitle(paste("Percent annotated cells - Layer: ", layer))
  print(p)
}
```

# Process data into wide dataframe

```{r}
dproc <- list()

for (layer in names(data)) {

  dproc$meta <- data[[layer]] %>%
    dplyr::select(c(1:14)) %>%
    unique() %>% 
    mutate_at(1:14, as.factor)
  
  ctable <- data[[layer]] %>%
    dplyr::select(c(14,
             !!sym(paste0("AIFI_", layer)),
             !!sym(paste0("AIFI_", layer, "_count")),
             !!sym(paste0("AIFI_", layer, "_frac_total")),
             !!sym(paste0("AIFI_", layer, "_clr")))) %>%
    pivot_wider(names_from = !!sym(paste0("AIFI_", layer)),
                values_from = c(!!sym(paste0("AIFI_", layer, "_count")),
                                !!sym(paste0("AIFI_", layer, "_frac_total")),
                                !!sym(paste0("AIFI_", layer, "_clr")))) %>% 
    column_to_rownames(var = "specimen.specimenGuid")
  
  i <- 1
  n_cts <- length(unique(data[[layer]][[paste0("AIFI_", layer)]]))
  
  dproc[[layer]]$cell_counts <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_count_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)

  i <- i + n_cts
  dproc[[layer]]$freq <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_frac_total_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)
  i <- i + n_cts
  dproc[[layer]]$clr_authors <- ctable[, i:(i + n_cts - 1)] %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_clr_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)
  
  df_counts <- dproc[[layer]]$cell_counts

  df_counts_1count_zero <- df_counts
  df_counts_1count_zero[df_counts_1count_zero == 0] <- 1
  dproc[[layer]]$clr_1count_zero <- df_counts_1count_zero %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_1count_all <- (df_counts + 1) %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_0.1perc <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()
  
  dproc[[layer]]$clr_0.01perc <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.01) %>%
    calc_perc_df() %>%
    clr()
}

saveRDS(dproc, file.path(path_data, "cell_type_compositions.rds"))
```

## Create scoot objects based on author annotation

```{r}
for (layer in names(data)) {
  df_counts <- dproc[[layer]]$cell_counts
  df_freq <- dproc[[layer]]$freq * 100
  df_clr <- dproc[[layer]]$clr_0.1perc
  
  cols <- colnames(df_counts)
  
  df_counts$scoot_sample <- rownames(df_counts)
  df_freq$scoot_sample <- rownames(df_freq)
  df_clr$scoot_sample <- rownames(df_clr)

  df_counts_long <- df_counts %>%
    pivot_longer(cols = cols,
                 names_to = "celltype",
                 values_to = "cell_counts")
  df_freq_long <- df_freq %>%
    pivot_longer(cols = cols,
                 names_to = "celltype",
                 values_to = "freq")
  df_clr_long <- df_clr %>%
    pivot_longer(cols = cols,
                 names_to = "celltype",
                 values_to = "clr")
  df_long <- dplyr::left_join(df_counts_long,
                              df_freq_long,
                              by = c("scoot_sample", "celltype"))
  df_long <- dplyr::left_join(df_long,
                              df_clr_long,
                              by = c("scoot_sample", "celltype"))
  
  ctable <- list(layer_1 = df_long)
  
  meta <- dproc$meta
  colnames(meta)[colnames(meta) == "specimen.specimenGuid"] <- "scoot_sample"
  
  assign(paste0("scoot_summary_", layer, "_authors"),
         methods::new("scoot",
                      data = list(composition = ctable),
                      metadata = meta,
                      version = packageVersion("scooter"))
  )
}
```



# ---------------------------------------------------------------

# Introduction schematic
GongSharma highest resolution -> best clustering example so far

```{r}
feat_mat <- dproc$L3$clr_0.1perc
res.pca <- prcomp(feat_mat)

eigenvalues <- res.pca$sdev^2
round(eigenvalues/sum(eigenvalues),3)
plot(cumsum(eigenvalues/sum(eigenvalues)))

# Create a scree plot
plot(eigenvalues, type = "b",
     xlab = "Principal Component",
     ylab = "Eigenvalue")
plot(eigenvalues/sum(eigenvalues)*100, type = "b",
     xlab = "Principal Component",
     ylab = "Percentage of Variance Explained")


feat_mat <- res.pca$x[, 1:3]

# dist_mat <- dist(feat_mat)
# # Perform hierarchical clustering
# hc <- stats::hclust(dist_mat, method = "ward.D2")
# labels <- stats::cutree(hc, k = 4)

# labels <- kmeans(feat_mat, 2)$cluster
labels <- cluster::pam(feat_mat, k = 4)$cluster

# adj_graph <- igraph::as.undirected(cccd::nng(feat_mat, k = 2))
# r <- quantile(igraph::strength(adj_graph))[2] / (igraph::gorder(adj_graph) - 1) / 4
# r <- 0.0000000001
# labels <- igraph::cluster_leiden(adj_graph, resolution = r)$membership
# 
# dbscan_clustres <- dbscan::hdbscan(feat_mat, minPts = 20)
# plot(dbscan_clustres, show_flat = TRUE)
# labels <- dbscan_clustres$cluster

factoextra::fviz_pca(res.pca,
                     habillage = labels,
                     # addEllipses = TRUE,
                     # label = "none",
                     pointsize = 1,
                     # invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Cluster"), color=guide_legend(title="Cluster")) +
  ggtitle("Unsupervised clustering")
ggsave(file.path(path_plots, "workflow_schema_clustering_unsupervised.pdf"),
       width = 12,
       height = 10)

factoextra::fviz_pca(res.pca,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     geom = "point") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Cluster"), color=guide_legend(title="Cluster")) +
  ggtitle("Unsupervised clustering")
ggsave(file.path(path_plots, "workflow_schema_clustering_unsupervised.pdf"),
       width = 4,
       height = 3)

factoextra::fviz_pca(res.pca,
                     habillage = dproc$meta$subject.cmv,
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     geom = "point") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="CMV"), color=guide_legend(title="CMV")) +
  ggtitle("Biological variable 1\nCMV infection status")
ggsave(file.path(path_plots, "workflow_schema_clustering_CMV.pdf"),
       width = 4,
       height = 3)

factoextra::fviz_pca(res.pca,
                     habillage = factor(ifelse(dproc$meta$cohort.cohortGuid == "BR1", "Young", "Old"), levels = c("Young", "Old")),
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     geom = "point") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Age"), color=guide_legend(title="Age")) +
  ggtitle("Biological variable 2\nAge")
ggsave(file.path(path_plots, "workflow_schema_clustering_age.pdf"),
       width = 4,
       height = 3)
```

```{r fig.height=6, fig.width=16}
composition_boxplot(scoot_object = scoot_summary_L3_authors)
composition_boxplot(scoot_object = scoot_summary_L3_authors,
                    group_by = "subject.cmv")
composition_boxplot(scoot_object = scoot_summary_L3_authors,
                    group_by = "cohort.cohortGuid")
composition_boxplot(scoot_object = scoot_summary_L3_authors,
                    group_by = "cohort.cohortGuid",
                    facet_by = "subject.cmv")
```



# Example plot of normalization for schema

```{r}

layer <- "L1"
hist(dproc[[layer]]$cell_counts %>% unlist(), breaks=50)
hist(dproc[[layer]]$freq %>% unlist() * 100, breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_1count_all %>% unlist(), breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_0.1perc %>% unlist(), breaks=50, main="", xlab="")

layer <- "L2"
hist(dproc[[layer]]$cell_counts %>% unlist(), breaks=50)
hist(dproc[[layer]]$freq %>% unlist() * 100, breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_1count_all %>% unlist(), breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_0.1perc %>% unlist(), breaks=50, main="", xlab="")



layer <- "L3"
hist(dproc[[layer]]$cell_counts %>% unlist(), breaks=50)
hist(dproc[[layer]]$freq %>% unlist() * 100, breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_1count_all %>% unlist(), breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_0.1perc %>% unlist(), breaks=50, main="", xlab="")

df <- dproc[[layer]]$cell_counts %>% unlist() %>% as.data.frame()
p1 <- ggplot(df, aes(.))+
  geom_histogram(color="darkblue", fill="lightblue", bins = 15) +
  xlab("Counts") + ylab("") + ggtitle("Cell counts") +
  theme_classic() +
  scale_x_continuous(trans='sqrt') +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())

df <- dproc[[layer]]$freq %>% unlist() %>% as.data.frame() %>% mutate(across(everything(), ~ . * 100))
p2 <- ggplot(df, aes(.))+
  geom_histogram(color="darkblue", fill="lightblue", bins = 15) +
  xlab("%") + ylab("") + ggtitle("Frequency") +
  theme_classic() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())

df <- dproc[[layer]]$clr_1count_all %>% unlist() %>% as.data.frame()
p3 <- ggplot(df, aes(.))+
  geom_histogram(color="darkblue", fill="lightblue", bins = 15) +
  xlab("CLR") + ylab("") + ggtitle("Centered\nlogratio") +
  theme_classic() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())

patchwork::wrap_plots(p1, p2, p3, nrow = 1)

ggsave(file.path(path_plots, "workflow_schema_normalization_example.pdf"),
       width = 4,
       height = 2.5)
```



# ---------------------------------------------------------------

# Plot number of samples and cell types lost per granularity

```{r}
# Apply function to each dataframe
df1 <- dproc[["L1"]][["cell_counts"]]
df2 <- dproc[["L2"]][["cell_counts"]]
df3 <- dproc[["L3"]][["cell_counts"]]

# Assume df1, df2, df3 are your three dataframes
dfs <- list(Low = df1, Medium = df2, High = df3)


# Function to calculate percentage of rows/columns with at least one zero
compute_zero_percentage <- function(df, granularity) {
  total_samples <- nrow(df)
  total_celltypes <- ncol(df)
  
  samples_with_zeros <- sum(rowSums(df == 0) > 0) / total_samples * 100
  celltypes_with_zeros <- sum(colSums(df == 0) > 0) / total_celltypes * 100
  total_zeros <- sum(df == 0) / (nrow(df) * ncol(df)) * 100
  
  granularity_labels <- paste0(granularity, "\n(", total_celltypes, " cell types)")
  
  data.frame(
    Granularity = factor(granularity_labels, levels = granularity_labels),
    Category = c("Samples", "Cell Types", "Total"),
    Percentage = c(samples_with_zeros, celltypes_with_zeros, total_zeros)
  )
}

# Compute percentages for all dataframes
zero_percentages <- bind_rows(
  lapply(names(dfs), function(n) compute_zero_percentage(dfs[[n]], n))
)

# Plot
ggplot(zero_percentages, aes(x = Granularity, y = Percentage, group = Category, color = Category, label = round(Percentage, 1))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(vjust=-1, show.legend  = F) +
  theme_minimal() +
  labs(title = "",
       x = "Granularity",
       y = "Percentage (%)",
       color = "Number of zeros") +
  theme(text = element_text(size = 14)) +
  ylim(0,100)
ggsave(file.path(path_plots, "coda_numberzeros.pdf"), width = 6, height = 4)
```




# ---------------------------------------------------------------

# Example PCAs

## Based on cell type composition

### Figure 2

```{r}
p_pca <- function(res.pca, labels, title = "") {
  p <- factoextra::fviz_pca(res.pca,
                            habillage = labels,
                            label = "var",
                            pointsize = 3,
                            invisible = c("var", "quali"),
                            geom = "point") +
    ggtitle(title) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
    coord_equal()
  print(p)
}

feat_mat <- dproc[["L3"]]$clr_0.1perc
meta <- dproc$meta
meta$age <- ifelse(as.numeric(as.character(meta$sample.subjectAgeAtDraw)) < 50, "young", "old")
res.pca <- prcomp(feat_mat, scale. = FALSE)

labels <- meta$subject.cmv
p_pca(res.pca, labels, title = "CMV")

labels <- meta$age
p_pca(res.pca, labels, title = "Age")

# Perform hierarchical clustering
dist_mat <- dist(feat_mat)
hc <- stats::hclust(dist_mat, method = "ward.D2")
clust_labels <- stats::cutree(hc, k = 4)
p_pca(res.pca, clust_labels, title = "hclust")

# Perform PAM clustering
clust_labels <- cluster::pam(feat_mat, k = 4)$cluster
p_pca(res.pca, clust_labels, title = "pamclust")

p_pca(res.pca, meta$subject.subjectGuid, title = "Individual")


feat_mat_females <- feat_mat[meta$subject.biologicalSex == "Female", ]
meta_females <- meta[meta$subject.biologicalSex == "Female", ]
res.pca <- prcomp(feat_mat_females, scale. = FALSE)

labels <- meta_females$subject.cmv
title <- "Biological variable 1"
mod_score <- round(calc_modularity(feat_mat_females, labels), 3)
title <- paste0(title, "\nModularity score: ", mod_score)
sil_score <- round(calc_sil(feat_mat_females, labels), 3)
title <- paste0(title, "\nSilhouette score: ", sil_score)
factoextra::fviz_pca(res.pca,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "none",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="CMV"), color=guide_legend(title="CMV")) +
  ggtitle("") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
ggsave(file.path(path_plots, "pca_ctcomp_females_cmv.pdf"),
       width = 4,
       height = 3)

labels <- ifelse(as.numeric(as.character(meta_females$sample.subjectAgeAtDraw)) < 50, "young", "old")
title <- "Biological variable 2"
mod_score <- round(calc_modularity(feat_mat_females, labels), 3)
title <- paste0(title, "\nModularity score: ", mod_score)
sil_score <- round(calc_sil(feat_mat_females, labels), 3)
title <- paste0(title, "\nSilhouette score: ", sil_score)
factoextra::fviz_pca(res.pca,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "none",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Age"), color=guide_legend(title="Age")) +
  ggtitle("") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
ggsave(file.path(path_plots, "pca_ctcomp_females_age.pdf"),
       width = 4,
       height = 3)

dist_mat <- dist(feat_mat_females)
# Perform hierarchical clustering
hc <- stats::hclust(dist_mat, method = "ward.D2")
clust_labels <- as.factor(stats::cutree(hc, k = 4))
p_pca(res.pca, clust_labels, title = "hclust")

# Perform PAM clustering
clust_labels <- cluster::pam(feat_mat_females, k = 4)$cluster
p_pca(res.pca, clust_labels, title = "pamclust")

clust_labels_reordered <- clust_labels
clust_labels_reordered[clust_labels == 3] <- 1
clust_labels_reordered[clust_labels == 1] <- 2
clust_labels_reordered[clust_labels == 2] <- 3

factoextra::fviz_pca(res.pca,
                     habillage = clust_labels_reordered,
                     # addEllipses = TRUE, , ellipse.level=0.80,
                     label = "none",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(clust_labels)))) +
  guides(shape=guide_legend(title="Immune type"), color=guide_legend(title="Immune type")) +
  ggtitle("") +
  theme_classic() +
  theme(
    # axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    # axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank()
  )
ggsave(file.path(path_plots, "pca_ctcomp_females_unsupervised_clustering.pdf"),
       width = 4,
       height = 3)


df_clust <- as.data.frame(cbind(as.character(meta_females$subject.cmv), meta_females$age, clust_labels))
colnames(df_clust) <- c("cmv", "age", "cluster")

cmv_counts <- df_clust %>%
  group_by(cluster, cmv) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

# Calculate counts for age (V2) and percentage within each cluster
age_counts <- df_clust %>%
  group_by(cluster, age) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

# Plot CMV percentages
cmv_plot <- ggplot(cmv_counts, aes(x = factor(cluster), y = percentage, fill = cmv)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Cluster", y = "Percentage", fill = "CMV Status", title = "CMV Status by Cluster (Percentage)") +
  scale_y_continuous(labels = scales::percent) + # Format y-axis as percentages
  theme_minimal()

# Plot age percentages
age_plot <- ggplot(age_counts, aes(x = factor(cluster), y = percentage, fill = age)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Cluster", y = "Percentage", fill = "Age", title = "Age by Cluster (Percentage)") +
  scale_y_continuous(labels = scales::percent) + # Format y-axis as percentages
  theme_minimal()

# Display the plots
print(cmv_plot)
ggsave(file.path(path_plots, "pca_ctcomp_females_unsupervised_clustering_barplot_cmv.pdf"),
       width = 4,
       height = 3)
print(age_plot)
ggsave(file.path(path_plots, "pca_ctcomp_females_unsupervised_clustering_barplot_age.pdf"),
       width = 4,
       height = 3)


df1 <- feat_mat
df1 <- df1[ , sqrt(colVars(feat_mat)) > 0.5]
df1 %>%
  prcomp() %>%
  p_pca(labels = meta$subject.cmv)
df1 %>%
  prcomp() %>%
  p_pca(labels = meta$age)
```


### Figure 3

```{r}
feat_mat <- dproc[["L3"]]$clr_0.1perc
res.pca <- prcomp(feat_mat, scale. = FALSE)


labels <- paste0("CMV ", meta$subject.cmv, " ", ifelse(as.numeric(as.character(meta$sample.subjectAgeAtDraw)) < 50, "young", "old"))
factoextra::fviz_pca(res.pca,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "all",
                     pointsize = 1,
                     labelsize = 2,
                     # invisible = c("var", "quali"),
                     repel = F,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Ground truth"), color=guide_legend(title="Ground truth")) +
  ggtitle("") +
  theme_classic() +
  theme(
    # axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    # axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank()
  )
ggsave(file.path(path_plots, "pca_ctcomp_all_ground_truth.pdf"),
       width = 6,
       height = 4)
```


### Heatmp (ct comp)

```{r fig.height=12, fig.width=12}

clust_labels <- cluster::pam(feat_mat_females, k = 4)$cluster

get_label <- function(string) {
  if (grepl("Immune Variation|Stand|Day 90", string)) {
    return("Baseline")
  } else {
    match <- "Vax"
    # match <- regmatches(string, regexpr("Day \\d+", string))
    return(match)
  }
}
meta$flu_vax_day <- as.character(meta$sample.visitName)
meta$flu_vax_day <- sapply(meta$flu_vax_day, get_label)

df_heatmap <- feat_mat_females %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  Immune_type = clust_labels,
  CMV = meta_females$subject.cmv,
  Age = meta_females$age,
  BMI = ifelse(na.aggregate(as.numeric(as.character(meta_females$subject.bmi)), FUN = mean) >30, ">30", "<30")
) %>% as.data.frame()

pheatmap(df_heatmap, annotation_col = my_sample_col, scale = "none", clustering_method = "ward.D2")


df1 <- feat_mat_females
df1 <- df1[ , sqrt(colVars(feat_mat_females)) > 0.5]
df1 %>%
  prcomp() %>%
  p_pca(labels = meta_females$subject.cmv)
df1 %>%
  prcomp() %>%
  p_pca(labels = meta_females$age)

df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  Immune_type = clust_labels,
  CMV = meta_females$subject.cmv,
  Age = meta_females$age
) %>% as.data.frame()

pheatmap(df_heatmap, annotation_col = my_sample_col, scale = "none", clustering_method = "ward.D2")


feat_mat <- dproc[["L3"]]$clr_0.1perc
plot(sqrt(colVars(feat_mat)))

df1 <- feat_mat

dist_mat <- dist(df1)
hc <- stats::hclust(dist_mat, method = "ward.D2")
clust_labels <- stats::cutree(hc, k = 4)
# clust_labels <- cluster::pam(df1, k = 4)$cluster

clust_labels_reordered <- clust_labels
clust_labels_reordered[clust_labels == 4] <- 1
clust_labels_reordered[clust_labels == 1] <- 2
clust_labels_reordered[clust_labels == 2] <- 4
clust_labels_reordered <- as.factor(clust_labels_reordered)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
gg_colors <- gg_color_hue(4)

ann_colors = list(
    # Vaccination_day = c(Vax = "red", Baseline = "black")
  # BMI = c(Adipose = "red", Lean = "black")
  Immune_type = c("1" = gg_colors[1], "2" = gg_colors[2], "3" = gg_colors[3], "4" = gg_colors[4])
)

df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  # Race = meta$subject.race,
  # BMI = ifelse(na.aggregate(as.numeric(as.character(meta$subject.bmi)), FUN = mean) >30, "Adipose", "Lean"),
  # Vaccination_day = meta$flu_vax_day,
  Age = meta$age,
  CMV = meta$subject.cmv,
  Immune_type = clust_labels_reordered
) %>% as.data.frame()

p <- pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  scale = "none",
  clustering_method = "ward.D2",
  treeheight_row = 0,
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4,
  annotation_colors = ann_colors
  )
pdf(file.path(path_plots, "heatmap_gongsharma_ctcomp_full.pdf"), width=10, height=10)
grid::grid.newpage()
grid::grid.draw(p$gtable)
dev.off()



df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  # Race = meta$subject.race,
  # BMI = ifelse(na.aggregate(as.numeric(as.character(meta$subject.bmi)), FUN = mean) >30, "Adipose", "Lean"),
  # Vaccination_day = meta$flu_vax_day,
  Age = meta$age,
  CMV = meta$subject.cmv,
  Immune_type = clust_labels_reordered
) %>% as.data.frame()

p <- pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  scale = "none",
  clustering_method = "ward.D2",
  # treeheight_row = 0,
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4 ,
  annotation_colors = ann_colors
  )
pdf(file.path(path_plots, "heatmap_gongsharma_ctcomp_full_withrowclustering.pdf"), width=10, height=10)
grid::grid.newpage()
grid::grid.draw(p$gtable)
dev.off()



row_order <- p[["tree_row"]][["order"]]
col_order <- p[["tree_col"]][["order"]]
cts <- p[["tree_row"]][["labels"]][row_order]


df1 <- feat_mat[col_order, cts[1:17]]
# df1 <- df1[ , sqrt(colVars(feat_mat)) > 0.5]


df1 %>%
  prcomp() %>%
  p_pca(labels = meta$subject.cmv[col_order])
df1 %>%
  prcomp() %>%
  p_pca(labels = meta$age[col_order])

df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  Immune_type = clust_labels,
  CMV = meta$subject.cmv,
  Age = meta$age #,
  # Sex = meta$subject.biologicalSex
) %>% as.data.frame()

p <- pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 3,
  cutree_cols = 4)
pdf(file.path(path_plots, "heatmap_gongsharma_ctcomp_subset_hvc.pdf"), width=10, height=5)
grid::grid.newpage()
grid::grid.draw(p$gtable)
dev.off()

```




### Calculate variance for compositions across bio conditions

```{r}
df <- dproc[["L3"]]$clr_0.01perc
meta <- mutate_all(dproc$meta, as.character)
meta$age <- ifelse(as.numeric(as.character(dproc$meta$sample.subjectAgeAtDraw)) < 50, "young", "old")
meta$all <- 1

list_vars <- list()

cond_cols <- c(
  "subject.subjectGuid",
  # "subject.cmv",
  # "age",
  "all"
)
for (cond in cond_cols) {
  col_vars <- c()
  for (i in unique(meta[[cond]])) {
    cond_rows <- which(meta[[cond]] == i)
    for (col in 1:ncol(df)) {
      if(length(cond_rows) >= 2) {
        # col_vars <- c(col_vars, cv(df[cond_rows, col]))
        col_vars <- c(col_vars, var(df[cond_rows, col]))
      }
    }
  }
  list_vars[[cond]] <- col_vars
}



# Find the maximum length of the vectors
max_length <- max(lengths(list_vars))

# Pad shorter vectors with NA to match the maximum length
padded_list <- lapply(list_vars, function(x) {
  length(x) <- max_length
  return(x)
})

# Convert the padded list to a data frame
df_vars <- as.data.frame(padded_list)

median(df_vars$subject.subjectGuid, na.rm = TRUE)
median(df_vars$all, na.rm = TRUE)
median(df_vars$subject.subjectGuid, na.rm = TRUE) / median(df_vars$all, na.rm = TRUE)
mean(df_vars$subject.subjectGuid, na.rm = TRUE)
mean(df_vars$all, na.rm = TRUE)

new_names <- c("Inter-sample", "Across all samples")
colnames(df_vars) <- new_names

df_vars %>%
  pivot_longer(cols = new_names) %>%
  
  ggplot(aes(x=name, y=value)) + 
  geom_violin(trim=FALSE, fill="gray")+
  labs(title="",x="", y = "Variance (in CLR)")+
  geom_boxplot(width = 0.1, outlier.shape=NA)+
  theme_classic() +
  stat_compare_means() +
  scale_fill_brewer(palette="Blues") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic()
ggsave(file.path(path_plots, "gongsharma_variance_violinplot.pdf"),
       width = 3,
       height = 3)
```

## Based on pseudobulk

```{r}
pseudobulk <- readRDS(file.path(path_data, "all_bulk_sample.rds"))
meta <- readRDS(file.path(path_data, "all_bulk_meta.rds"))
meta <- meta %>%
  arrange((specimen.specimenGuid))
meta$age <- ifelse(as.numeric(as.character(meta$sample.subjectAgeAtDraw)) < 50, "young", "old")

pb_normed <- DESeq2.normalize(t(as.data.frame(pseudobulk)[, 2:ncol(pseudobulk)]), meta)
pb_normed <- t(pb_normed)
row.names(pb_normed) <- pseudobulk$specimen.specimenGuid

res.pca <- prcomp(pb_normed, scale. = FALSE)
labels <- meta$subject.biologicalSex
p_pca(res.pca, labels, title = "Sex")
ggsave(file.path(path_plots, "pca_pseudobulk_by_sex.pdf"),
       width = 4,
       height = 3)

labels <- meta$subject.cmv
p_pca(res.pca, labels, title = "CMV")
ggsave(file.path(path_plots, "pca_pseudobulk_by_cmv.pdf"),
       width = 4,
       height = 3)

labels <- meta$age
p_pca(res.pca, labels, title = "Age")
ggsave(file.path(path_plots, "pca_pseudobulk_by_age.pdf"),
       width = 4,
       height = 3)


pb_normed_females <- pb_normed[meta$subject.biologicalSex == "Female", ]
meta_females <- meta[meta$subject.biologicalSex == "Female"]

res.pca_females <- prcomp(pb_normed_females, scale. = FALSE)

# Perform PAM clustering
clust_labels <- cluster::pam(pb_normed_females, k = 4)$cluster
p_pca(res.pca_females, clust_labels, title = "pamclust")
factoextra::fviz_pca(res.pca_females,
                     habillage = clust_labels,
                     # addEllipses = TRUE,
                     label = "none",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(clust_labels)))) +
  guides(shape=guide_legend(title="Pseudobulk\ncluster"), color=guide_legend(title="Pseudobulk\ncluster")) +
  ggtitle("") +
  theme_classic() +
  theme(
    # axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    # axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank()
  )
ggsave(file.path(path_plots, "pca_pseudobulk_females_unsupervised_clustering.pdf"),
       width = 4,
       height = 3)


factoextra::fviz_pca(res.pca_females,
                     # habillage = NA,
                     # addEllipses = TRUE,
                     label = "none",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(clust_labels)))) +
  guides(shape=guide_legend(title="Pseudobulk\ncluster"), color=guide_legend(title="Pseudobulk\ncluster")) +
  ggtitle("") +
  theme_classic() +
  theme(
    # axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    # axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank()
  )
ggsave(file.path(path_plots, "pca_pseudobulk_females_unsupervised_clustering_nolabels.pdf"),
       width = 3,
       height = 3)




labels <- meta_females$subject.cmv
title <- "Biological variable 1"
mod_score <- round(calc_modularity(pb_normed_females, labels), 3)
title <- paste0(title, "\nModularity score: ", mod_score)
sil_score <- round(calc_sil(pb_normed_females, labels), 3)
title <- paste0(title, "\nSilhouette score: ", sil_score)
factoextra::fviz_pca(res.pca_females,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "none",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="CMV"), color=guide_legend(title="CMV")) +
  ggtitle("") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
ggsave(file.path(path_plots, "pca_pseudobulk_females_by_cmv.pdf"),
       width = 4,
       height = 3)

labels <- meta_females$age
title <- "Biological variable 2"
mod_score <- round(calc_modularity(pb_normed_females, labels), 3)
title <- paste0(title, "\nModularity score: ", mod_score)
sil_score <- round(calc_sil(pb_normed_females, labels), 3)
title <- paste0(title, "\nSilhouette score: ", sil_score)
factoextra::fviz_pca(res.pca_females,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "none",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     # repel = TRUE,
                     geom = "point",
                     axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Age"), color=guide_legend(title="Age")) +
  ggtitle("") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
ggsave(file.path(path_plots, "pca_pseudobulk_females_by_age.pdf"),
       width = 4,
       height = 3)



df_clust <- as.data.frame(cbind(as.character(meta_females$subject.cmv), meta_females$age, clust_labels))
colnames(df_clust) <- c("cmv", "age", "cluster")

cmv_counts <- df_clust %>%
  group_by(cluster, cmv) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

# Calculate counts for age (V2) and percentage within each cluster
age_counts <- df_clust %>%
  group_by(cluster, age) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

# Plot CMV percentages
cmv_plot <- ggplot(cmv_counts, aes(x = factor(cluster), y = percentage, fill = cmv)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Cluster", y = "Percentage", fill = "CMV Status", title = "CMV Status by Cluster (Percentage)") +
  scale_y_continuous(labels = scales::percent) + # Format y-axis as percentages
  theme_minimal()

# Plot age percentages
age_plot <- ggplot(age_counts, aes(x = factor(cluster), y = percentage, fill = age)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Cluster", y = "Percentage", fill = "Age", title = "Age by Cluster (Percentage)") +
  scale_y_continuous(labels = scales::percent) + # Format y-axis as percentages
  theme_minimal()

# Display the plots
print(cmv_plot)
ggsave(file.path(path_plots, "pca_pb_females_unsupervised_clustering_barplot_cmv.pdf"),
       width = 4,
       height = 3)
print(age_plot)
ggsave(file.path(path_plots, "pca_pb_females_unsupervised_clustering_barplot_age.pdf"),
       width = 4,
       height = 3)
```


### Calculate variance for expression across bio conditions

```{r}
df <- pb_normed
meta <- mutate_all(dproc$meta, as.character)
meta$age <- ifelse(as.numeric(as.character(dproc$meta$sample.subjectAgeAtDraw)) < 50, "young", "old")
meta$all <- 1

list_vars <- list()

cond_cols <- c(
  "subject.subjectGuid",
  # "subject.cmv",
  # "age",
  "all"
)
for (cond in cond_cols) {
  col_vars <- c()
  for (i in unique(meta[[cond]])) {
    cond_rows <- which(meta[[cond]] == i)
    for (col in 1:ncol(df)) {
      if(length(cond_rows) >= 2) {
        # col_vars <- c(col_vars, cv(df[cond_rows, col]))
        col_vars <- c(col_vars, var(df[cond_rows, col]))
      }
    }
  }
  list_vars[[cond]] <- col_vars
}



# Find the maximum length of the vectors
max_length <- max(lengths(list_vars))

# Pad shorter vectors with NA to match the maximum length
padded_list <- lapply(list_vars, function(x) {
  length(x) <- max_length
  return(x)
})

# Convert the padded list to a data frame
df_vars <- as.data.frame(padded_list)

median(df_vars$subject.subjectGuid, na.rm = TRUE)
median(df_vars$all, na.rm = TRUE)
median(df_vars$subject.subjectGuid, na.rm = TRUE) / median(df_vars$all, na.rm = TRUE)
mean(df_vars$subject.subjectGuid, na.rm = TRUE)
mean(df_vars$all, na.rm = TRUE)

new_names <- c("Inter-sample", "Across all samples")
colnames(df_vars) <- new_names

df_vars %>%
  pivot_longer(cols = new_names) %>%
  
  ggplot(aes(x=name, y=value)) + 
  geom_violin(trim=FALSE, fill="gray")+
  labs(title="",x="", y = "Variance (in noramlized counts)")+
  geom_boxplot(width = 0.1, outlier.shape=NA)+
  theme_classic() +
  stat_compare_means() +
  scale_fill_brewer(palette="Blues") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic()
ggsave(file.path(path_plots, "gongsharma_variance_violinplot_expression.pdf"),
       width = 3,
       height = 3)
```

# ---------------------------------------------------------------

# Correlation plot (ct comp)

```{r}
cor_matrix <- cor(dproc[["L1"]][["clr_0.01perc"]])
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix)

cor_matrix <- cor(dproc[["L2"]][["clr_0.01perc"]])
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix)

variances <- apply(dproc[["L3"]][["clr_0.01perc"]], 2, var)
plot(variances)
abline(mean(variances))
hist(variances, breaks = 20)
```

```{r fig.height=14, fig.width=14}
cor_matrix <- cor(dproc[["L3"]][["clr_0.01perc"]])
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))
```



# ---------------------------------------------------------------


# Variance

```{r fig.height=8, fig.width=10}

ggplotRegression <- function (data, title, ylim = c(0.01, 5)) {
  
  require(ggplot2)
  require(ggrepel)
  
  lmod <- lm(Variance ~ Relative_abundance, data = data)
  
  ggplot(data, aes(x = Relative_abundance, y = Variance)) + 
    geom_point() +
    # geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
    labs(title = paste(
      title #,
      # "\nAdj R2 = ",signif(summary(lmod)$adj.r.squared, 2),
      # "Intercept =",signif(lmod$coef[[1]], 2),
      # " Slope =",signif(lmod$coef[[2]], 2),
      # " P =",signif(summary(lmod)$coef[2,4], 2)
    )) +
    theme_classic() +
    # geom_text_repel(
    #   aes(label = celltype),
    #   family = "Poppins",
    #   size = 3,
    #   min.segment.length = 0, 
    #   seed = 42, 
    #   box.padding = 0.5,
    #   max.overlaps = Inf,
    #   # arrow = arrow(length = unit(0.010, "npc")),
    #   # nudge_x = .15,
    #   # nudge_y = .5,
    #   color = "grey50"
    # ) +
    xlab("CLR-transformed relative abundance") + ylab("Variance")
}


plot_list <- list()


# 
# df <- dproc$L3$cell_counts %>%
#   log() %>%
#   pivot_longer(
#     cols = everything(),
#     names_to = "celltype",
#     values_to = "values"
#   )
# 
# df1 <- df %>%
#     group_by(celltype) %>%
#     dplyr::summarize(
#       Relative_abundance = mean(values, na.rm=TRUE),
#       Variance = var(values, na.rm=TRUE)
#     )
# 
# p <- ggplotRegression(df1, "High granularity annotation counts")
# print(p)
# plot_list <- append(plot_list, list(p))
# 
# 
# df <- dproc$L3$freq %>%
#   log() %>%
#   pivot_longer(
#     cols = everything(),
#     names_to = "celltype",
#     values_to = "values"
#   )
# 
# df1 <- df %>%
#     group_by(celltype) %>%
#     dplyr::summarize(
#       Relative_abundance = mean(values, na.rm=TRUE),
#       Variance = var(values, na.rm=TRUE)
#     )
# 
# p <- ggplotRegression(df1, "High granularity annotation freq")
# print(p)
# plot_list <- append(plot_list, list(p))



df <- dproc$L1$clr_0.01perc %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )

p <- ggplotRegression(df1, "Low granularity annotation\nZero replacement: + 0.01%")
print(p)
plot_list <- append(plot_list, list(p))

df <- dproc$L2$clr_0.01perc %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )

p <- ggplotRegression(df1, "Medium granularity annotation\nZero replacement: + 0.01%")
print(p)
plot_list <- append(plot_list, list(p))



df <- dproc$L3$clr_0.01perc %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )

ggplotRegression(df1, "High granularity annotation\nZero replacement: + 0.01%") +
  geom_text(aes(label=celltype), hjust=0, vjust=0)
p <- ggplotRegression(df1, "High granularity annotatio\nZero replacement: + 0.01%")
print(p)
plot_list <- append(plot_list, list(p))


df <- dproc$L3$clr_1count_zero %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )

p <- ggplotRegression(df1, "High granularity annotation\nZero replacement: +1 zeros")
print(p)
plot_list <- append(plot_list, list(p))


df <- dproc$L3$cell_counts %>%
  mutate_all(~ ./10) %>%
  round() %>%
  calc_perc_df() %>%
  impute_zeros("counts_zeros") %>%
  clr() %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )

p <- ggplotRegression(df1, "High granularity annotation\n(low cell counts)\nZero replacement: +1 zeros")
print(p)
plot_list <- append(plot_list, list(p))


df <- dproc$L3$clr_0.1perc %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )

ggplotRegression(df1, "High granularity annotation\nZero replacement: + 0.1%") +
  geom_text(aes(label=celltype), hjust=0, vjust=0)
p <- ggplotRegression(df1, "High granularity annotation\nZero replacement: + 0.1%")
print(p)
plot_list <- append(plot_list, list(p))

patchwork::wrap_plots(plot_list, ncol = 3)
ggsave(file.path(path_plots, "variance_gongsharma.pdf"),
       width = 12,
       height = 8)


```

```{r}
clearcut_samples <- dproc$meta$specimen.specimenGuid[dproc$meta$subject.biologicalSex == "Female" & dproc$meta$subject.cmv == "Negative" & as.numeric(dproc$meta$subject.ageAtFirstDraw) < 50]
df <- dproc$L3$clr_0.01perc[row.names(dproc$L3$clr_0.1perc) %in% clearcut_samples, ]
df1 <- df %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df1 %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )

ggplotRegression(df1, "CLR\nZero replacement: + 0.01%")



df <- dproc$L3$freq[row.names(dproc$L3$freq) %in% clearcut_samples, ] * 100
df1 <- df %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "values"
  )

df1 <- df1 %>%
    group_by(celltype) %>%
    dplyr::summarize(
      Relative_abundance = mean(values, na.rm=TRUE),
      Variance = var(values, na.rm=TRUE)
    )


# lmod <- lm(Variance ~ Relative_abundance, data = df1)

ggplot(df1, aes(x = Relative_abundance, y = Variance)) + 
  geom_point() +
  labs(title = "Frequency") +
  theme_classic() +
  xlab("Relative abundance (%)") + ylab("Variance")
```


# ---------------------------------------------------------------

# Blood from Terekhova et al.

Single-cell atlas of healthy human blood unveils age-related loss of NKG2C+GZMB−CD8+ memory T cells and accumulation of type 2 memory T cells
Marina Terekhova1,6 ∙ Amanda Swain1,6 ∙ Pavla Bohacova1,6 ∙ Ekaterina Aladyeva1 ∙ Laura Arthur1 ∙ Anwesha Laha1 ∙ Denis A. Mogilenko1,2 ∙ Samantha Burdess1 ∙ Vladimir Sukhov1,3 ∙ Denis Kleverov1,3 ∙ Barbora Echalar1 ∙ Petr Tsurinov1,4 ∙ Roman Chernyatchik1,5 ∙ Kamila Husarcikova1 ∙ Maxim N. Artyomov1,7 martyomov@wustl.edu Show less

Data from:
https://www.cell.com/immunity/fulltext/S1074-7613(23)00453-3?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1074761323004533%3Fshowall%3Dtrue



## Load data

```{r}
library(readxl)

df_ter <- data.frame(read_xlsx(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "TerekhovaM_2023_CellImmu", "mmc7.xlsx")))
meta_ter <- data.frame(read_xlsx(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "TerekhovaM_2023_CellImmu", "mmc2.xlsx")))

df_ter <- df_ter[order(df_ter$Donor_id), ]
meta_ter <- meta_ter[order(meta_ter$Donor_id), ]

identical(df_ter$Donor_id, meta_ter$Donor_id)

meta_ter <- meta_ter[, c("Donor_id", "Gender", "Ethnicity", "Age_group")]
meta_ter$Age <- df_ter$Age
meta_ter$BMI <- df_ter$BMI
meta_ter$Age_bin <- ifelse(meta_ter$Age > 50, ">50", "<=50")

df_ter_lowres <- df_ter[, 8:16]

df_ter[, 17:20] <- df_ter[, 17:20] * df_ter_lowres$Myeloid.cells / 100
df_ter[, 21:36] <- df_ter[, 21:36] * df_ter_lowres$CD4..T.cells / 100
df_ter[, 37:48] <- df_ter[, 37:48] * df_ter_lowres$TRAV1.2...CD8..T.cells / 100
df_ter[, 49:54] <- df_ter[, 49:54] * df_ter_lowres$NK.cells / 100
df_ter[, 55:59] <- df_ter[, 55:59] * df_ter_lowres$gd.T.cells / 100
df_ter[, 60:68] <- df_ter[, 60:68] * df_ter_lowres$B.cells / 100
df_ter_hires <- df_ter[, c(10, 12, 16, 17:(ncol(df_ter)-2))]


```


```{r fig.height=9, fig.width=9}
feat_mat <- df_ter_lowres %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  clr()
p <- plot_pca(feat_mat, labels = meta_ter$Age_group)

feat_mat <- df_ter_hires %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  clr()
p <- plot_pca(feat_mat, labels = meta_ter$Age_group)
p <- plot_pca(feat_mat, labels = meta_ter$Age_group, invisible = "var", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta_ter$Age_group, invisible = "none", labelsize = 4)

p <- plot_pca(feat_mat, labels = meta_ter$Age_bin, plotly_3d = TRUE)


p <- plot_pca(feat_mat[meta_ter$Age_group %in% c("A", "D", "E"), ], labels = meta_ter$Age_group[meta_ter$Age_group %in% c("A", "D", "E")])
p <- plot_pca(feat_mat[meta_ter$Age_group %in% c("A", "E"), ], labels = meta_ter$Age_group[meta_ter$Age_group %in% c("A", "E")], invisible = "none", labelsize = 4)

```


