---
title: "Compositional data analysis of the Gong & Sharma et al. dataset"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

Using exploratory compositional data analysis on large healthy data set

paper: <https://www.biorxiv.org/content/10.1101/2024.09.10.612119v1> accession: TBD

This script takes output of "<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>" see "Cell type frequency data"


# For additional sample metadata see: "<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>"

# General study description from authors:
```{r}
# Immune aging is a dynamic process driven by complex internal and external interactions. We examined the healthy human immune system across age using deep molecular profiling .

# To do so, we used our Human Immune Health Atlas to label peripheral blood mononuclear cells (PBMCs) from a prospective, longitudinally sampled cohort of healthy young adults (n=49, 25-35 years) and older adults (n=47, 55-65 years) followed over the course of 2-3 years. This cohort was observed under homeostatic conditions, as well as prior to, and at 7 and 90 days following, annual influenza vaccinations, resulting in a scRNA-seq dataset with over 13.5 million PBMCs.

# Our study highlights that gradual, age-related alterations in the homeostatic transcriptional networks of immune cells lead to shifts in immune responsiveness as we age.
```

# Cell type data description
```{r}
# Cell type frequency data
# After labeling cell types, we tabulated the cell count for each of the 868 samples utilized in our study at each level of resolution in our Immune Health Atlas. Below, we provide these cell counts, the fraction of counts for each sample, and the centered log ratio (CLR) transformation of those fractions that we utilized for our analyses.

# In addition, we utilized the absolute lymphocyte counts (ALC) provided in our clinical blood count lab results for each sample (available below) to compute estimated cell type abundance based on normalization to ALC. For more details, see our Data Analysis Methods.
```

# Metadata columns description
```{r}
# cohort.cohortGuid: Cohort ID (BR1 or BR2)
# subject.subjectGuid: Subject ID
# subject.biologicalSex: Subject Sex (Female or Male)
# subject.cmv: Subject CMV Status (Negative or Positive)
# subject.bmi: Subject BMI (integer)
# subject.race: Subject race
# subject.ethnicity: Subject ethnicity
# subject.birthYear: Subject Birth Year
# subject.ageAtFirstDraw: Subject Age at earliest blood draw in study
# sample.sampleKitGuid: Sample Kit ID
# sample.visitName: Sample Visit Name
# sample.drawDate: Sample Draw Date (Year-Month)
# sample.subjectAgeAtDraw: Subject age at time of draw, based on year of Draw Date and Birth Year
# specimen.specimenGuid: Specimen ID (pbmc_sample_id in .h5 files)

# Frequency-related columns
# (for AIFI_L1 as an example; AIFI_L1 is replaced with AIFI_L2 and AIFI_L3 for those levels)

# AIFI_L1: Cell Type assignment
# AIFI_L1_count: Count of cells within this sample with cell type assignment
# total_cells: Total cells within this sample
# scrna.lymphocyte_count: Sum of T, NK, and B cells
# bc.lymphocyte_count: Absolute Lymphocyte Count (ALC) from clinical Blood Counts (bc.)
# alc_ratio: ALC per scRNA Lymphocyte Count
# AIFI_L1_frac_total: Fraction of cells with cell type assignment divided by Total cells for this sample
# AIFI_L1_alc: ALC estimate for this cell type assignment
# AIFI_L1_clr: Centered Log Ratio computed using AIFI_L1_frac_total for all types within this sample
```



# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

library(tidyverse)
library(plotly)
library(ggplot2)
library(ggpubr)
library(forcats)
library(patchwork)
library(factoextra)
library(compositions)
library(pROC)
library(mclust)
library(Rfast)
library(zoo)
library(pheatmap)
source("functions.R")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)

# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
library(ncdf4)
library(arrow)

library(dplyr)

my_seed <- 123
set.seed(my_seed)
```

# Set paths

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets", ds_name)
path_data <- file.path(path_root, "data")
path_plots <- file.path(getwd(), "plots")
```


# Load dataset
Manually downloaded from:
"<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>"

```{r}
files <- list(L1 = file.path(path_data, "sound_life_AIFI_L1_frequencies.csv"),
              L2 = file.path(path_data, "sound_life_AIFI_L2_frequencies.csv"),
              L3 = file.path(path_data, "sound_life_AIFI_L3_frequencies.csv"))

data <- list()

for (f in names(files)) {
  data[[f]] <- read.csv(files[[f]])
}
```


# Cell types per layer

```{r}
t1 <- as.data.frame.matrix(t(table(data$L3$AIFI_L1, data$L3$AIFI_L2)))

for (i in ncol(t1):1) {
  t1 <- t1[order(t1[, i], decreasing = TRUE), ]
}

colnames(t1) <- str_replace_all(colnames(t1), " ", "_")
row.names(t1) <- str_replace_all(row.names(t1), " ", "_")

t2 <- as.data.frame.matrix(t(table(data$L3$AIFI_L2, data$L3$AIFI_L3)))

for (i in ncol(t2):1) {
  t2 <- t2[order(t2[, i], decreasing = TRUE), ]
}

colnames(t2) <- str_replace_all(colnames(t2), " ", "_")
row.names(t2) <- str_replace_all(row.names(t2), " ", "_")
```


# Plot number of cells per sample

```{r}
for (layer in names(data)) {
  df <- data[[layer]] %>%
    dplyr::select(paste0("AIFI_", layer),
           paste0("AIFI_", layer, "_frac_total"),
           paste0("total_cells"),
           "specimen.specimenGuid") %>%
    group_by(specimen.specimenGuid) %>%
    reframe(cells_sum = min(total_cells),
            annotated_cells_perc = sum(!!sym(paste0("AIFI_", layer, "_frac_total"))))
  
  p <- ggplot(data = df, aes(x = reorder(specimen.specimenGuid, cells_sum), y = cells_sum)) +
    geom_bar(stat="identity", fill="steelblue") +
    xlab("") +
    ylab("Cells") +
    theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
    ggtitle(paste("Cell counts per sample - Layer: ", layer))
  print(p)
  
  p <- ggplot(data = df, aes(x = reorder(specimen.specimenGuid, cells_sum), y = annotated_cells_perc)) +
    geom_bar(stat="identity", fill="steelblue") +
    xlab("") +
    ylab("Cells") +
    theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
    ggtitle(paste("Percent annotated cells - Layer: ", layer))
  print(p)
}
```

# Process data into wide dataframe

```{r}
dproc <- list()

for (layer in names(data)) {

  dproc$meta <- data[[layer]] %>%
    dplyr::select(c(1:14)) %>%
    unique() %>% 
    mutate_at(1:14, as.factor)
  
  ctable <- data[[layer]] %>%
    dplyr::select(c(14,
             !!sym(paste0("AIFI_", layer)),
             !!sym(paste0("AIFI_", layer, "_count")),
             !!sym(paste0("AIFI_", layer, "_frac_total")),
             !!sym(paste0("AIFI_", layer, "_clr")))) %>%
    pivot_wider(names_from = !!sym(paste0("AIFI_", layer)),
                values_from = c(!!sym(paste0("AIFI_", layer, "_count")),
                                !!sym(paste0("AIFI_", layer, "_frac_total")),
                                !!sym(paste0("AIFI_", layer, "_clr")))) %>% 
    column_to_rownames(var = "specimen.specimenGuid")
  
  i <- 1
  n_cts <- length(unique(data[[layer]][[paste0("AIFI_", layer)]]))
  
  dproc[[layer]]$cell_counts <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_count_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)

  i <- i + n_cts
  dproc[[layer]]$freq <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_frac_total_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)
  i <- i + n_cts
  dproc[[layer]]$clr_authors <- ctable[, i:(i + n_cts - 1)] %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_clr_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)
  
  df_counts <- dproc[[layer]]$cell_counts

  df_counts_1count_zero <- df_counts
  df_counts_1count_zero[df_counts_1count_zero == 0] <- 1
  dproc[[layer]]$clr_1count_zero <- df_counts_1count_zero %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_1count_all <- (df_counts + 1) %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_1count <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
    calc_perc_df() %>%
    clr()
  
  dproc[[layer]]$clr_0.1perc <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()
  
  dproc[[layer]]$clr_0.01perc <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.01) %>%
    calc_perc_df() %>%
    clr()
}

dproc$meta$age <- factor(ifelse(dproc$meta$cohort.cohortGuid == "BR1", "Young", "Old"), levels = c("Young", "Old"))
dproc$meta$combined_cmv_age <- factor(paste0("CMV", ifelse(as.character(dproc$meta$subject.cmv) == "Positive", "+", "-"), " ", as.character(dproc$meta$age)),
                                      levels = c("CMV- Old",
                                                 "CMV+ Old",
                                                 "CMV- Young",
                                                 "CMV+ Young")
)

saveRDS(dproc, file.path(path_data, "cell_type_compositions.rds"))
```


# ---------------------------------------------------------------

# Figure 1 A

## UMAP example

```{r}
source("Processed_dataset_metadata.R")

ds <- "Zhang"
seurat <- readRDS(file.path(file.path(getwd(), "data"), paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
seurat_sub <- subset(seurat, downsample = 500)

seurat_sub <- RunUMAP(seurat_sub, dims = 1:10)
Idents(seurat_sub) <- "RNA_snn_res.0.1"
DimPlot(seurat_sub, reduction = "umap") + theme_void() & NoLegend()
ggsave(file.path(path_plots, "Figure_1_A_umap.pdf"),
       width = 4,
       height = 4)

df <- table(seurat_sub$Sample, seurat_sub$RNA_snn_res.0.1) %>%
  as.data.frame.matrix() %>%
  calc_perc_df()

# 1. Transform data to long format
df_long <- df[1:5, ] %>%
  rownames_to_column("Sample") %>% 
  pivot_longer(cols = -Sample, 
               names_to = "Cluster", 
               values_to = "Percentage")

# 2. Create the plot
ggplot(df_long, aes(x = Sample, y = Percentage, fill = Cluster)) +
  geom_bar(stat = "identity") +
  theme_light() +
  theme(
    # 1. Remove the legend
    legend.position = "none",
    
    # 2. Remove X-axis title
    axis.title.x = element_blank(),
    
    # 3. Remove X-axis text (labels) and ticks
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    
    # Optional: Styling the Y-axis to match your previous plot
    axis.title.y = element_text(face = "bold"),
    panel.grid.major.x = element_blank()
  ) +
  labs(y = "Relative abundance (%)")

ggsave(file.path(path_plots, "Figure_1_A_ctcomposition.pdf"),
       width = 2.5,
       height = 2.5)
```


## Normalization

```{r}

layer <- "L1"
hist(dproc[[layer]]$cell_counts %>% unlist(), breaks=50)
hist(dproc[[layer]]$freq %>% unlist() * 100, breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_1count_all %>% unlist(), breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_0.1perc %>% unlist(), breaks=50, main="", xlab="")

layer <- "L2"
hist(dproc[[layer]]$cell_counts %>% unlist(), breaks=50)
hist(dproc[[layer]]$freq %>% unlist() * 100, breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_1count_all %>% unlist(), breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_0.1perc %>% unlist(), breaks=50, main="", xlab="")



layer <- "L3"
hist(dproc[[layer]]$cell_counts %>% unlist(), breaks=50)
hist(dproc[[layer]]$freq %>% unlist() * 100, breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_1count_all %>% unlist(), breaks=50, main="", xlab="")
hist(dproc[[layer]]$clr_0.1perc %>% unlist(), breaks=50, main="", xlab="")

df <- dproc[[layer]]$cell_counts %>% unlist() %>% as.data.frame()
p1 <- ggplot(df, aes(.))+
  geom_histogram(color="darkblue", fill="lightblue", bins = 15) +
  xlab("Counts") + ylab("") + ggtitle("Cell counts") +
  theme_classic() +
  scale_x_continuous(trans='sqrt') +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())

df <- dproc[[layer]]$freq %>% unlist() %>% as.data.frame() %>% mutate(across(everything(), ~ . * 100))
p2 <- ggplot(df, aes(.))+
  geom_histogram(color="darkblue", fill="lightblue", bins = 15) +
  xlab("%") + ylab("") + ggtitle("Frequency") +
  theme_classic() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())

df <- dproc[[layer]]$clr_1count_all %>% unlist() %>% as.data.frame()
p3 <- ggplot(df, aes(.))+
  geom_histogram(color="darkblue", fill="lightblue", bins = 15) +
  xlab("CLR") + ylab("") + ggtitle("Centered\nlogratio") +
  theme_classic() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())

patchwork::wrap_plots(p2, p3, nrow = 1)

ggsave(file.path(path_plots, "Figure_1_A_normalization_example.pdf"),
       width = 3,
       height = 2.5)
```




# ---------------------------------------------------------------

# Figure 1 B: Based on cell type composition


```{r}
feat_mat <- dproc$L3$clr_1count

# Create a mapping vector: "Old Name" = "New Name"
cell_map <- c(
  "SOX4+_naive_CD4_T_cell"          = "CD4+ Naive",
  "SOX4+_naive_CD8_T_cell"          = "CD8+ Naive",
  "Adaptive_NK_cell"                = "NK Adaptive",
  "KLRF1-_GZMB+_CD27-_EM_CD8_T_cell" = "CD8+ EM",
  "KLRF1+_effector_Vd1_gdT"         = "Vd1 Effector",
  # "KLRF1+_effector_Vd1_gdT"         = "Vd1 Effector (KLRF1+)",
  # "KLRF1-_effector_Vd1_gdT"         = "Vd1 Effector (KLRF1-)",
  "KLRF1-_GZMB+_CD27-_memory_CD4_T_cell" = "CD4+ Memory",
  "Naive_Vd1_gdT"                   = "Vd1 Naive",
  "SOX4+_Vd1_gdT"                   = "Vd1 SOX4+",
  "CD8aa"                           = "CD8aa"
)

# Apply to your feature matrix
colnames(feat_mat) <- ifelse(colnames(feat_mat) %in% names(cell_map), 
                             cell_map[colnames(feat_mat)], 
                             colnames(feat_mat))

res.pca <- prcomp(feat_mat)


feat_mat <- res.pca$x[, 1:3]


set.seed(1)
labels <- factor(kmeans(feat_mat, 4)$cluster)


levels(labels) <- c("2", "3", "4", "1")
labels <- factor(labels, levels = c("1", "2", "3", "4"))

my_colors <- c(
  "1" = "#F8766D",
  "2" = "#7CAE00",
  "3" = "#00BFC4",
  "4" = "#C77CFF",
  "CMV- Old" = "#F8766D",
  "CMV+ Old" = "#7CAE00",
  "CMV- Young" = "#00BFC4",
  "CMV+ Young" = "#C77CFF"
)

eig.val <- factoextra::get_eig(res.pca)
pc1_var <- round(eig.val[1, 2], 1)
pc2_var <- round(eig.val[2, 2], 1)


p1 <- factoextra::fviz_pca(res.pca,
                     # habillage = labels,
                     # addEllipses = TRUE,
                     # label = "var",
                     col.ind = "grey80",
                     pointsize = 1,
                     labelsize = 3,
                     invisible = c("quali"),
                     geom = "point",
                     select.var = list(contrib = 4),
                     repel = TRUE,
                     axes.linetype = "blank") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Cluster"), color=guide_legend(title="Cluster")) +
  ggtitle("PCA - ECODA sample representation") +
  # scale_color_manual(values = my_colors) +
  coord_equal() +
  theme_classic() +
  labs(x = paste0("PCA dim1 (", pc1_var, "%)"), y = paste0("PCA dim2 (", pc2_var, "%)"))  #+
  # theme(axis.text = element_blank(),
  #         axis.ticks = element_blank())
print(p1)

p2 <-factoextra::fviz_pca(res.pca,
                     habillage = dproc$meta$combined_cmv_age,
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("quali"),
                     col.var = "white",
                     arrowsize = 0,
                     labelsize = 0,
                     alpha.var = 0,
                     geom = "point",
                     select.var = list(contrib = 4),
                     axes.linetype = "blank") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Bio group"), color=guide_legend(title="Bio group")) +
  ggtitle("Known biological groups") +
  scale_color_manual(values = my_colors) + coord_equal() +
  theme_classic() +
  labs(x = paste0("PCA dim1 (", pc1_var, "%)"), y = paste0("PCA dim2 (", pc2_var, "%)"))  #+
  # theme(axis.text = element_blank(),
  #         axis.ticks = element_blank())
print(p2)
```


```{r}
# 1. Run PCA
res.pca <- prcomp(feat_mat, center = TRUE, scale. = FALSE)

# 2. Get the eigenvalues (variances)
# prcomp returns standard deviations; square them to get eigenvalues
eig <- (res.pca$sdev)^2

# 3. Calculate Variable Coordinates (Correlations)
# Formula: Loading * Standard Deviation of the PC
# This transforms the rotation matrix into the 'coordinates' used in biplots
variable_coords <- t(t(res.pca$rotation) * res.pca$sdev)

# 4. Calculate Squared Cosines (Quality of representation)
# For variables on a correlation PCA, this is just the squared coordinate
var_cos2 <- variable_coords^2

# 5. Calculate Contributions
# Formula: (cos2 / eigenvalue) * 100
# We divide each column of squared coordinates by its eigenvalue
var_contrib <- as.data.frame(t(t(var_cos2) / eig) * 100)

# Check: The sum of any column in var_contrib should be 100
colSums(var_contrib)


library(factoextra)
res.var <- get_pca_var(res.pca)

max(var_contrib - res.var$contrib)
```

# Figure 1 B: Based on pseudobulk

```{r}
pseudobulk <- readRDS(file.path(path_data, "all_bulk_sample.rds"))
meta <- readRDS(file.path(path_data, "all_bulk_meta.rds"))
meta <- meta %>%
  arrange((specimen.specimenGuid))
meta$age <- ifelse(as.numeric(as.character(meta$sample.subjectAgeAtDraw)) < 50, "young", "old")

pb_normed <- DESeq2.normalize(t(as.data.frame(pseudobulk)[, 2:ncol(pseudobulk)]), meta)
pb_normed <- t(pb_normed)
row.names(pb_normed) <- pseudobulk$specimen.specimenGuid

res.pca <- prcomp(pb_normed, scale. = FALSE)

eig.val <- factoextra::get_eig(res.pca)
pc1_var <- round(eig.val[1, 2], 1)
pc2_var <- round(eig.val[2, 2], 1)

set.seed(1)
labels <- factor(kmeans(pb_normed, 2)$cluster)

levels(labels) <- c("2", "1") # Swap these if the colors are inverted
labels <- factor(labels, levels = c("1", "2")) # Force order 1 then 2

# Ensure these are in your my_colors
my_colors <- c(
  "1" = "orange",
  "2" = "skyblue",
  "Female" = "orange",
  "Male" = "skyblue"
)

p3 <- factoextra::fviz_pca(res.pca,
                     # habillage = labels,
                     # addEllipses = TRUE,
                     col.ind = "grey80",
                     label = "var",
                     pointsize = 1,
                     invisible = c("quali"),
                     geom = "point",
                     select.var = list(contrib = 4),
                     labelsize = 3,
                     repel = TRUE,
                     axes.linetype = "blank") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Cluster"), color=guide_legend(title="Cluster")) +
  ggtitle("PCA - Pseudobulk sample representation") +
  scale_color_manual(values = my_colors) + coord_equal() +
  theme_classic() +
  labs(x = paste0("PCA dim1 (", pc1_var, "%)"), y = paste0("PCA dim2 (", pc2_var, "%)")) #+
  # theme(axis.text = element_blank(),
  #         axis.ticks = element_blank())
p3 <- p3 + 
  scale_x_continuous(breaks = seq(-60, 60, by = 20)) + 
  scale_y_continuous(breaks = seq(-40, 40, by = 20))
print(p3)

labels <- meta$subject.biologicalSex

p4 <-factoextra::fviz_pca(res.pca,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("quali"),
                     col.var = "white",
                     arrowsize = 0,
                     labelsize = 0,
                     alpha.var = 0,
                     select.var = list(contrib = 4),
                     geom = "point",
                     axes.linetype = "blank") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Bio group"), color=guide_legend(title="Bio group")) +
  ggtitle("Known biological groups") +
  scale_color_manual(values = my_colors) + coord_equal() +
  theme_classic() +
  labs(x = paste0("PCA dim1 (", pc1_var, "%)"), y = paste0("PCA dim2 (", pc2_var, "%)")) #+
  # theme(axis.text = element_blank(),
  #         axis.ticks = element_blank())
p4 <- p4 + 
  scale_x_continuous(breaks = seq(-60, 60, by = 20)) + 
  scale_y_continuous(breaks = seq(-40, 40, by = 20))
print(p4)


# p5 <- factoextra::fviz_pca(res.pca,
#                      habillage = labels,
#                      # addEllipses = TRUE,
#                      # label = "var",
#                      pointsize = 1,
#                      invisible = c("quali", "var"),
#                      geom = "point",
#                      # select.var = list(contrib = 4),
#                      labelsize = 3,
#                      repel = TRUE,
#                      axes.linetype = "blank") +
#     scale_shape_manual(values = rep(19, length(unique(labels)))) +
#   guides(shape=guide_legend(title="Cluster"), color=guide_legend(title="Cluster")) +
#   ggtitle("Unsupervised clustering") +
#   scale_color_manual(values = my_colors) + coord_equal() +
#   theme_classic() #+
#   # theme(axis.text = element_blank(),
#   #         axis.ticks = element_blank())
# p5 <- p5 + 
#   scale_x_continuous(breaks = seq(-60, 60, by = 20)) + 
#   scale_y_continuous(breaks = seq(-20, 20, by = 20))
# print(p5)
# 
# labels <- meta$subject.biologicalSex
# 
# p6 <-factoextra::fviz_pca(res.pca,
#                      habillage = labels,
#                      # addEllipses = TRUE,
#                      # label = "var",
#                      pointsize = 1,
#                      invisible = c("quali", "var"),
#                      col.var = "white",
#                      arrowsize = 0,
#                      labelsize = 0,
#                      alpha.var = 0,
#                      # select.var = list(contrib = 4),
#                      geom = "point",
#                      axes.linetype = "blank") +
#     scale_shape_manual(values = rep(19, length(unique(labels)))) +
#   guides(shape=guide_legend(title="Bio group"), color=guide_legend(title="Bio group")) +
#   ggtitle("Known biological groups") +
#   scale_color_manual(values = my_colors) + coord_equal() +
#   theme_classic() #+
#   # theme(axis.text = element_blank(),
#   #         axis.ticks = element_blank())
# p6 <- p6 + 
#   scale_x_continuous(breaks = seq(-60, 60, by = 20)) + 
#   scale_y_continuous(breaks = seq(-20, 20, by = 20))
# print(p6)

aligned_plots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
ggsave(file.path(path_plots, "Figure_1_B_GongSharma_clustering.pdf"), aligned_plots, width = 8, height = 6)

# aligned_plots <- p1 + p2 + p3 + p4 + p5 + p6 + plot_layout(ncol = 2)
# ggsave(file.path(path_plots, "TEST.pdf"), aligned_plots, width = 8, height = 8)
```



# Supp Fig 1: GongSharma PB per sex

```{r}
plot_list <- list()

for (gender in unique(meta$subject.biologicalSex)) {
  pb_normed_sub <- pb_normed[meta$subject.biologicalSex == gender, ]
  meta_sub <- meta[meta$subject.biologicalSex == gender]
  
  res.pca_sub <- prcomp(pb_normed_sub, scale. = FALSE)
  
  eig.val <- factoextra::get_eig(res.pca_sub)
  pc1_var <- round(eig.val[1, 2], 1)
  pc2_var <- round(eig.val[2, 2], 1)
  
  labels <- meta_sub$subject.cmv
  condition <- "CMV"
  title <- "Biological variable 1"
  plot_list[[paste0(gender, "_", condition)]] <- factoextra::fviz_pca(
    res.pca_sub,
    habillage = labels,
    # addEllipses = TRUE,
    label = "none",
    pointsize = 1,
    invisible = c("var", "quali"),
    # repel = TRUE,
    geom = "point",
    axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
    guides(shape=guide_legend(title=condition), color=guide_legend(title=condition)) +
    ggtitle(paste(gender, "-", condition)) +
    labs(x = paste0("PCA dim1 (", pc1_var, "%)"), y = paste0("PCA dim2 (", pc2_var, "%)")) +
    theme_classic() +
    theme(# axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      # axis.title.y=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank()) + coord_equal()
  print(plot_list[[paste0(gender, "_", condition)]])
  
  labels <- meta_sub$age
  condition <- "Age"
  plot_list[[paste0(gender, "_", condition)]] <- factoextra::fviz_pca(
    res.pca_sub,
    habillage = labels,
    # addEllipses = TRUE,
    label = "none",
    pointsize = 1,
    invisible = c("var", "quali"),
    # repel = TRUE,
    geom = "point",
    axes.linetype = NA) +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
    guides(shape=guide_legend(title=condition), color=guide_legend(title=condition)) +
    ggtitle(paste(gender, "-", condition)) +
    labs(x = paste0("PCA dim1 (", pc1_var, "%)"), y = paste0("PCA dim2 (", pc2_var, "%)")) +
    theme_classic() +
    theme(# axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      # axis.title.y=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank()) + coord_equal()
  print(plot_list[[paste0(gender, "_", condition)]])
}

patchwork::wrap_plots(plot_list, nrow = 2)
ggsave(file.path(path_plots, "Supp_fig_1_GongSharma_PB_CMV_Age.pdf"), width = 7, height = 5)
```


### Heatmp (ct comp)

```{r fig.height=12, fig.width=12}

clust_labels <- cluster::pam(feat_mat_females, k = 4)$cluster

get_label <- function(string) {
  if (grepl("Immune Variation|Stand|Day 90", string)) {
    return("Baseline")
  } else {
    match <- "Vax"
    # match <- regmatches(string, regexpr("Day \\d+", string))
    return(match)
  }
}
meta$flu_vax_day <- as.character(meta$sample.visitName)
meta$flu_vax_day <- sapply(meta$flu_vax_day, get_label)

df_heatmap <- feat_mat_females %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  Immune_type = clust_labels,
  CMV = meta_females$subject.cmv,
  Age = meta_females$age,
  BMI = ifelse(na.aggregate(as.numeric(as.character(meta_females$subject.bmi)), FUN = mean) >30, ">30", "<30")
) %>% as.data.frame()

pheatmap(df_heatmap, annotation_col = my_sample_col, scale = "none", clustering_method = "ward.D2")


df1 <- feat_mat_females
df1 <- df1[ , sqrt(colVars(feat_mat_females)) > 0.5]
df1 %>%
  prcomp() %>%
  p_pca(labels = meta_females$subject.cmv)
df1 %>%
  prcomp() %>%
  p_pca(labels = meta_females$age)

df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  Immune_type = clust_labels,
  CMV = meta_females$subject.cmv,
  Age = meta_females$age
) %>% as.data.frame()

pheatmap(df_heatmap, annotation_col = my_sample_col, scale = "none", clustering_method = "ward.D2")


feat_mat <- dproc[["L3"]]$clr_0.1perc
plot(sqrt(colVars(feat_mat)))

df1 <- feat_mat

dist_mat <- dist(df1)
hc <- stats::hclust(dist_mat, method = "ward.D2")
clust_labels <- stats::cutree(hc, k = 4)
# clust_labels <- cluster::pam(df1, k = 4)$cluster

clust_labels_reordered <- clust_labels
clust_labels_reordered[clust_labels == 4] <- 1
clust_labels_reordered[clust_labels == 1] <- 2
clust_labels_reordered[clust_labels == 2] <- 4
clust_labels_reordered <- as.factor(clust_labels_reordered)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
gg_colors <- gg_color_hue(4)

ann_colors = list(
    # Vaccination_day = c(Vax = "red", Baseline = "black")
  # BMI = c(Adipose = "red", Lean = "black")
  Immune_type = c("1" = gg_colors[1], "2" = gg_colors[2], "3" = gg_colors[3], "4" = gg_colors[4])
)

df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  # Race = meta$subject.race,
  # BMI = ifelse(na.aggregate(as.numeric(as.character(meta$subject.bmi)), FUN = mean) >30, "Adipose", "Lean"),
  # Vaccination_day = meta$flu_vax_day,
  Age = meta$age,
  CMV = meta$subject.cmv,
  Immune_type = clust_labels_reordered
) %>% as.data.frame()

p <- pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  scale = "none",
  clustering_method = "ward.D2",
  treeheight_row = 0,
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4,
  annotation_colors = ann_colors
  )
pdf(file.path(path_plots, "heatmap_gongsharma_ctcomp_full.pdf"), width=10, height=10)
grid::grid.newpage()
grid::grid.draw(p$gtable)
dev.off()



df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  # Race = meta$subject.race,
  # BMI = ifelse(na.aggregate(as.numeric(as.character(meta$subject.bmi)), FUN = mean) >30, "Adipose", "Lean"),
  # Vaccination_day = meta$flu_vax_day,
  Age = meta$age,
  CMV = meta$subject.cmv,
  Immune_type = clust_labels_reordered
) %>% as.data.frame()

p <- pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  scale = "none",
  clustering_method = "ward.D2",
  # treeheight_row = 0,
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4 ,
  annotation_colors = ann_colors
  )
pdf(file.path(path_plots, "heatmap_gongsharma_ctcomp_full_withrowclustering.pdf"), width=10, height=10)
grid::grid.newpage()
grid::grid.draw(p$gtable)
dev.off()



row_order <- p[["tree_row"]][["order"]]
col_order <- p[["tree_col"]][["order"]]
cts <- p[["tree_row"]][["labels"]][row_order]


df1 <- feat_mat[col_order, cts[1:17]]
# df1 <- df1[ , sqrt(colVars(feat_mat)) > 0.5]


df1 %>%
  prcomp() %>%
  p_pca(labels = meta$subject.cmv[col_order])
df1 %>%
  prcomp() %>%
  p_pca(labels = meta$age[col_order])

df_heatmap <- df1 %>% scale(center = TRUE, scale = FALSE)
df_heatmap <- df_heatmap %>% t() %>% as.data.frame()
my_sample_col <- list(
  Immune_type = clust_labels,
  CMV = meta$subject.cmv,
  Age = meta$age #,
  # Sex = meta$subject.biologicalSex
) %>% as.data.frame()

p <- pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 3,
  cutree_cols = 4)
pdf(file.path(path_plots, "heatmap_gongsharma_ctcomp_subset_hvc.pdf"), width=10, height=5)
grid::grid.newpage()
grid::grid.draw(p$gtable)
dev.off()

```