---
title: "Zhang et al. dataset cell type annotation"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

Zhang Yuanyuan Cancer Cell 2021; CD45+ cells; primary or metastatic tumor tissues and peripheral blood of 22 advanced TNBC patients treated with paclitaxel or its combination with atezolizumab.

paper: https://www.sciencedirect.com/science/article/pii/S1535610821004992?via%3Dihub
accession:https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE169246

This script takes output of "scRNAseq_data_processing_ZhangY_2022_34653365.Rmd"

# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::restore()
# renv::snapshot()
# 
# remotes::install_github("carmonalab/UCell", ref="f6e711c")
# remotes::install_github("carmonalab/scGate", ref="17a4ce6")
# remotes::install_github("carmonalab/STACAS", ref="0a88c5e")
# remotes::install_github("carmonalab/ProjecTILs", ref="fdc75e0")
# remotes::install_github("carmonalab/SignatuR", ref="0aded1d")
# remotes::install_github("carmonalab/HiTME", ref="73ff3ca", force = T)

# remotes::install_github("carmonalab/scooter", ref="f31eab3", force = T)

library(scooter)

# remotes::install_github("saezlab/MOFAcellulaR")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)
library(tidyverse)
library(scGate)
library(UCell)
library(STACAS)
library(scGate)
library(ProjecTILs)
library(SignatuR)
library(HiTME)
library(dplyr)
```


# Set paths
```{r}
my_seed <- 321
set.seed(my_seed)
ds_name <- "HarN_2022_36097221"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
path_output <- file.path(path_root, "output")
path_output_samples <- file.path(path_output, "samples")
```


# Load dataset
```{r}
obj <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name)))
```

# Filter out low cell count samples
```{r}
sample.size <- sort(sapply(obj, function(df) ncol(df)))
sample.size
barplot(sample.size)
abline(h = 1000)
```


# Celltype annotation with HiTME

```{r}
# Get scGate database of pre-defined gating models
scGate_models_DB <- get_scGateDB(branch = "dev", verbose = T, force_update = TRUE)
scGate_models <- scGate_models_DB$human$PBMC

# Load ProjecTILs reference maps
path_ref <- "~/Dropbox/CSI/reference_atlases/sketched_200ct"
ref.maps_sketched <- list(CD8 = load.reference.map(file.path(path_ref, "sketched_CD8T_human_ref_v1.rds")),
                          CD4 = load.reference.map(file.path(path_ref, "sketched_CD4T_human_ref_v2.rds")),
                          DC = load.reference.map(file.path(path_ref, "sketched_DC_human_ref_v2.rds")),
                          MoMac = load.reference.map(file.path(path_ref, "sketched_MoMac_human_v1.rds"))
)
ref.maps <- get.reference.maps(collection = "human")

for (i in 1:length(obj)) {
  print(i)
  obj[[i]] <- Run.HiTME(object = obj[[i]],
                        scGate.model = scGate_models,
                        ref.maps = ref.maps_sketched
  )
}

rm(i, ref.maps_sketched)
gc()

save_objs(obj, path_output_samples)
```