---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
source("functions.R")
source("Processed_dataset_metadata.R")

library(forcats)

# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
py_install("mofapy2", envname = "my_env", method="auto")
import("mofapy2")
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
# install.packages("GloScope")
library(GloScope)

# Set paths
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"

set.seed(123)
```



# Create results list to fill

```{r}
result_file <- "result_list.rds"

if (file.exists(result_file)) {
  result_list <- readRDS(result_file)
} else {
  result_list <- list()
}
```



# ---------------------------------------------------------------

# Process datasets

```{r message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
for (ds in names(datasets)) {
  seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
  
  seurat <- replace_HiTMElayer3_annot(seurat, thresh = 0.1)
  
  seurat@misc$label_col <- datasets[[ds]][["label_col"]]
  seurat@misc$low_res_ct_col <- datasets[[ds]][["low_res_ct_col"]]
  seurat@misc$hi_res_ct_col <- datasets[[ds]][["hi_res_ct_col"]]
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots)
}
```


# ---------------------------------------------------------------

# Gong and Sharma

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```



### Check for batch effect

```{r}
alpha <- 0.3

meta_cols_plot <- c(
  "Sample",
  "RNA_snn_res.0.5",
  "batch_id",
  "pool_id",
  "chip_id",
  "well_id",
  "cohort.cohortGuid",
  "AIFI_L1",
  "AIFI_L2",
  "AIFI_L3",
  "subject.cmv"
)

# Check in broad cell types (strong global batch effect)

Idents(seurat) <- "Sample"
sub <- subset(x = seurat, downsample = 100)

sub <- sub |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(dims = 1:50, verbose = FALSE) |>
  FindNeighbors() |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:50)

for (mc in meta_cols_plot) {
  Idents(sub) <- mc
  DimPlot(sub, reduction = "umap", alpha = alpha)
}

# No strong batch effect observed


# To check for more subtle batch effect, we can subset to only T cells, where the gene expression difference between cells is smaller
# Thus, by subseting to only T cells, we should be better able to see smaller batch effects

sub <- subset(x = seurat, subset = AIFI_L1 == "T cell")
sub <- subset(x = sub, downsample = 1000)

sub <- sub |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(dims = 1:50, verbose = FALSE) |>
  FindNeighbors() |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:50)

for (mc in meta_cols_plot) {
  Idents(sub) <- mc
  DimPlot(sub, reduction = "umap", alpha = alpha)
}

# No weak batch effect observed
```



## Age (females cmvneg)

```{r}
ds <- "GongSharma_age_females_cmvneg"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  seurat <- load_h5ad_to_seurat("data/Gongsharma_age_females_cmvneg.h5ad")
  seurat$Sample <- seurat$specimen.specimenGuid
  
  seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))

  seurat@misc$label_col <- "age"
  seurat@misc$low_res_ct_col <- "AIFI_L1"
  seurat@misc$hi_res_ct_col <- "AIFI_L3"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
}
```





# ---------------------------------------------------------------

# Bassez

# IMPORTANT: Needs automated annotation, as seurat$cohort c("neoadjuvant_chemo", "treatment_naive") seem to be separately and differently annotated by the authors
## Has also other non-immune cells

```{r}
ds <- "Bassez"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))

  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  ds_name <- "BassezA_2021_33958794"

  path_root <- file.path(path_nas, ds_name)
  path_output <- file.path(path_root, "output")

  seurat <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name)))

  seurat <- subset(seurat, subset = timepoint == "Pre")
  seurat <- subset(seurat, subset = expansion != "n/a")
  seurat <- subset(seurat, subset = cohort == "treatment_naive")

  seurat$Sample <- seurat$sample
  seurat$cellSubType[is.na(seurat$cellSubType)] <- seurat$cellType[is.na(seurat$cellSubType)]

  convert_seurat_to_h5ad(seurat, ds)

  seurat@misc$label_col <- "expansion"
  seurat@misc$low_res_ct_col <- "cellType"
  seurat@misc$hi_res_ct_col <- "cellSubType"

  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
}
```




# ---------------------------------------------------------------

# Test new dataset (template)

```{r}
# seurat_sub <- subset(seurat, subset = OriginalAnnotationLevel1 %in% c("B", "Mast", "Myeloid", "Plasma", "TNKILC"))
# 
# test <- run_benchmark_analysis(seurat_sub,
#                                get_labels(seurat_sub, seurat_sub@misc$label_col),
#                                seurat@misc$low_res_ct_col,
#                                seurat@misc$hi_res_ct_col,
#                                factors_test = 2,
#                                Pseudobulk = TRUE,
#                                ECODA_deconv = FALSE,
#                                ECODA_low_res = FALSE,
#                                ECODA_high_res = TRUE,
#                                Pseudobulk_PCA = FALSE,
#                                ECODA_high_res_PCA = FALSE,
#                                MOFA = FALSE,
#                                scITD = FALSE,
#                                MrVI = FALSE,
#                                show_pca_plots = TRUE,
#                                save_pca_plots = FALSE
# )
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Create list to export raw data to excel

```{r}
df_results_list <- list()
```


# ---------------------------------------------------------------
# ---------------------------------------------------------------


# Plot benchmark results

## Save and flatten the nested list into a DataFrame

```{r}
bmark_result_list <- result_list[["bmark"]]

df_results <- imap_dfr(bmark_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    if ("scores" %in% names(.x)) {
      n_samples <- nrow(.x$feat_mat)
      n_features <- ncol(.x$feat_mat)
      
      imap_dfr(.x$scores, ~ {
        tibble(
          dataset = dataset_name,
          method = method_name,
          score = .y,
          value = .x,
          n_samples = n_samples,
          n_features = n_features
        )
      })
    } else {
      tibble() # Return an empty tibble if "scores" is not found
    }
  })
})

# Drop old modularity score method
df_results <- df_results[!df_results$score == "mod_score", ]
```

## Show + sort all tested conditions

```{r}
library(gtools)
levs <- mixedsort(as.character(unique(df_results$method)))
levs

df_results$method <- factor(df_results$method, levels = levs)

df_results$dataset[df_results$dataset == "GongSharma_females_cmv"] <- "GongSharma_cmv"
df_results$dataset[df_results$dataset == "GongSharma_females_cmvneg_age"] <- "GongSharma_age"

df_results_list[["Benchmark"]] <- df_results
```


## Plot results



### All methods (overview)

```{r fig.height=8, fig.width=16}
# New facet label names for dose variable
score.labs <- c("ANOSIM score", "Adjusted rand index", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            labeller = labeller(score = score.labs, dataset = datas.labs),
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "benchmark_detailed_barplot.pdf"),
       width = 150,
       height = 7, limitsize = FALSE)


df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_bars.pdf"),
       width = 14,
       height = 7)
```


### Main methods (benchmark)

```{r fig.height=5, fig.width=4}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)


df_plot <- df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "ECODA_NULL",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"))


df_plot %>%
  filter(score == "anosim_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_authors",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_mainmethods_refECODA.pdf"),
       width = 4.5,
       height = 5.5)



df_plot %>%
  filter(score == "cluster_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_authors",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ARI score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_ari_mainmethods_refECODA.pdf"),
       width = 4.5,
       height = 5.5)



# df_plot %>%
#   filter(score == "mod_score") %>%
# 
#   ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
#   stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
#   stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
#   geom_point(position = "identity", alpha = 0.6, size = 0.8) +
#   geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
#   stat_compare_means(
#     ref.group = "ECODA",
#     method = "wilcox.test",
#     paired = TRUE,
#     angle = 25,
#     aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
#   ) +
#   ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
#   labs(x = "", y = "Modularity score") +
#   theme_classic() +
#   theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
# ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity_mainmethods_refECODA.pdf"),
#        width = 4.5,
#        height = 5.5)



df_plot %>%
  filter(score == "mod_knn3_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_authors",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "Modularity score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularityknn3_mainmethods_refECODA.pdf"),
       width = 4.5,
       height = 5.5)
```

#### Extended methods

```{r fig.height=5, fig.width=4}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)

df_plot <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "ECODA_NULL",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"))

df_plot %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_authors",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_mainmethods_with_bad_baselines.pdf"),
       width = 4.5,
       height = 5.5)
```


#### Plot all separation metrics together

```{r fig.height=6, fig.width=5}
df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "ECODA_NULL",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_allmetricscombined_mainmethods_with_bad_baselines.pdf"),
       width = 4.5,
       height = 6.5)
```

#### Plot PCoA for selected datasets and main methods

```{r}
m <- filter_methods[1]
# ds <- names(result_list[["bmark"]])[1]
ds <- "Zhang"

p <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["dist_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = FALSE,
    title = m
  )
print(p)
```


```{r fig.height=12, fig.width=12}
library(patchwork)

ds <- "Pelka"
plot_list <- list()

methods <- list(
  "ECODA_authors_HR" = "ECODA",
  "ECODA_authors_HR_NULL" = "ECODA_NULL",
  "Pseudobulk_10_PCA_dims" = "Pseudobulk",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scITD_hvg2000_factors5" = "scITD",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims5" = "scPoli",
  "PILOT_hvg2000" = "PILOT",
  "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"
)

for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 3, guides = "collect")
```



```{r fig.height=12, fig.width=12}
methods <- list(
  "ECODA_authors_HR" = "ECODA",
  "ECODA_authors_HR_NULL" = "ECODA_NULL",
  "Pseudobulk_10_PCA_dims" = "Pseudobulk",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scITD_hvg2000_factors5" = "scITD",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims5" = "scPoli",
  "PILOT_hvg2000" = "PILOT",
  "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"
)

for (ds in names(result_list[["bmark"]])) {
  plot_list <- list()
  
  for (m in names(methods)) {
    dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
    labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
    
    if (m == "ECODA_authors_HR_NULL") {
      labels <- sample(labels)
    }
    
    plot_list[[m]] <- plot_pca(
      feat_mat = dist_mat,
      labels = labels,
      coord_equal = FALSE,
      title = methods[[m]]
    )
  }
  
  wrap_plots(plot_list) +
    plot_layout(nrow = 3, guides = "collect")
  
  ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
         width = 12,
         height = 12)
}

```


### Compare PCA based on features vs dist

```{r fig.height=12, fig.width=12}
ds <- "Pelka"
plot_list <- list()

methods <- list(
  "ECODA_authors_HR" = "ECODA",
  "ECODA_authors_HR_NULL" = "ECODA_NULL",
  "Pseudobulk_10_PCA_dims" = "Pseudobulk",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scITD_hvg2000_factors5" = "scITD",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims5" = "scPoli",
  "PILOT_hvg2000" = "PILOT",
  "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"
)

for (m in names(methods)) {
  feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = feat_mat,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 3, guides = "collect")


for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 3, guides = "collect")
```



### ECODA annotation method comparison

```{r fig.height=6, fig.width=5}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_HiTME_HR_layer2",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_0.4"
)

df_plot <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
                         "ECODA_scATOMIC_HR" = "ECODA_scATOMIC",
                         "ECODA_seuratres_0.4" = "ECODA_Leiden"))
  
p <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Separation score (ANOSIM)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_ecodaonly.pdf"),
       width = 4,
       height = 5)


# For appendix
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_LR",
  "ECODA_deconv",
  "ECODA_HiTME_HR_layer2",
  "ECODA_HiTME_HR_layer3",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_0.4",
  "ECODA_seuratres_2",
  "ECODA_seuratres_5",
  "ECODA_seuratres_20"
)


df_plot <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(!grepl("Zhang|Lee", dataset)) %>% # Exclude Zhang and Lee, as they do not have high-resolution author annotation
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_seuratres_0.4" = "ECODA_leiden_res_0.4",
                         "ECODA_seuratres_2" = "ECODA_leiden_res_2",
                         "ECODA_seuratres_5" = "ECODA_leiden_res_5",
                         "ECODA_seuratres_20" = "ECODA_leiden_res_20"))
  
p <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_authors_HR",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    angle = 25,
    label.y = 1,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.15)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_mainmethods_refECODAhr.pdf"),
       width = 4,
       height = 4.5)

```


### Method parameter tests

```{r fig.height=8, fig.width=14}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_LR",
  "ECODA_deconv",
  "ECODA_HiTME_HR_layer2",
  "ECODA_HiTME_HR_layer3",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_0.4",
  "ECODA_seuratres_2",
  "ECODA_seuratres_5",
  "ECODA_seuratres_20"
)

default_methods <- c(
    "ECODA_authors_HR\\b", 
    "MOFA_hvg2000_factors15", 
    "Pseudobulk_10_PCA_dims",
    "PILOT_hvg2000",
    "scITD_hvg2000_factors5",
    "scPoli_hvg2000_dims5",
    "GloScope_hvg2000_pcadims30_sqrtmat",
    "MrVI_hvg2000"
)
default_methods_regex <- paste(default_methods, collapse = "|")

df_results %>%
  # filter(score == "anosim_score") %>%
  filter(!grepl("HiTME|scATOMIC|seuratres|ECODA_authors_HR_NULL|ECODA_authors_LR|ECODA_deconv|Freq_highres|varexp|varcts|Avg_PCA", method)) %>%
  # filter(method %in% filter_methods) %>%
  # mutate(method = recode(method,
  #                        "MOFA_2_factors" = "MOFA",
  #                        "scITD_2_factors" = "scITD",
  #                        "ECODA_highres" = "ECODA")) %>%
  
  # CREATE THE FACETING GROUP COLUMN
  mutate(
    method_plot = case_when(
      grepl("^ECODA_", method) ~ "ECODA",
      grepl("^MOFA_", method) ~ "MOFA",
      grepl("^MrVI_", method) ~ "MrVI",
      grepl("^scITD_", method) ~ "scITD",
      grepl("^GloScope_", method) ~ "GloScope",
      grepl("^PILOT_", method) ~ "PILOT",
      grepl("^scPoli_", method) ~ "scPoli",
      grepl("^Pseudobulk_", method) ~ "Pseudobulk",
      TRUE ~ method # Keep any other simple names as they are
    ),
    
    method_plot = factor(method_plot, 
                         levels = c("ECODA", "Pseudobulk", "MrVI", "GloScope",
                                    "MOFA", "scITD", "scPoli", "PILOT"))
  ) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # Highlight default methods
  scale_x_discrete(labels = function(x) highlight(
    x,
    default_methods_regex,
    "black")
  ) +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             cols = vars(method_plot), # Use the new grouping column for column faceting
             labeller = labeller(score = score.labs))


ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_mainmethods_refECODAhr.pdf"),
       width = 14,
       height = 8)
```


### Combined ECODA and PB

```{r}
result_list1 <- result_list

ecoda_weights <- c(0, 0.25, 0.5, 0.75, 1)
norm_methods <- c("max", "median", "zscore", "quantile")

for (ds in names(result_list$bmark)) {
  for (i in ecoda_weights) {
    for (norm in norm_methods) {
      result_list$bmark[[ds]][[paste0("ECODA_PB_combo_norm", norm, "_ecodaweight", i)]] <- process_ecodapb_fig(
        dist_mat_ecoda = result_list$bmark[[ds]][["ECODA_authors_HR"]][["dist_mat"]],
        dist_mat_pb = result_list$bmark[[ds]][["Pseudobulk_unsup_hvg2000"]][["dist_mat"]],
        ecoda_weight = i,
        norm_method = norm,
        labels = result_list$bmark[[ds]][["ECODA_authors_HR"]][["labels"]]
      )
    }

    result_list$bmark[[ds]][[paste0("ECODA_PB_combo_normcosine_ecodaweight", i)]] <- process_ecodapb_fig(
        feat_mat_ecoda = result_list$bmark[[ds]][["ECODA_authors_HR"]][["feat_mat"]],
        feat_mat_pb = result_list$bmark[[ds]][["Pseudobulk_unsup_hvg2000"]][["feat_mat"]],
        ecoda_weight = i,
        # norm_method = norm,
        labels = result_list$bmark[[ds]][["ECODA_authors_HR"]][["labels"]]
    )
  }
}
```


#### Plot distance distribution after different normalization methods applied

```{r}
ecoda <- result_list$bmark$Pelka$ECODA_authors_HR$dist_mat
pb <- result_list$bmark$Pelka$Pseudobulk_hvg2000$dist_mat

hist(ecoda)
hist(pb)

test1 <- ecoda / max(ecoda)
test2 <- pb / max(pb)
hist(test1, main = "ecoda max")
hist(test2, main = "pb max")

test1 <- ecoda / median(ecoda)
test2 <- pb / median(pb)
hist(test1, main = "ecoda median")
hist(test2, main = "pb median")

test1 <- global_quantile_norm_gaussian(ecoda)
test2 <- global_quantile_norm_gaussian(pb)
hist(test1, main = "ecoda quantiles")
hist(test2, main = "pb quantiles")

test1 <- zscore_transform(ecoda)
test2 <- zscore_transform(pb)
hist(test1, main = "ecoda zscore")
hist(test2, main = "pb zscore")
```


#### Plot benchmark including ECODA+PB (overview)

```{r fig.height=9, fig.width=7}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  
  paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  "ECODA_PB_combo_normcosine_ecodaweight0.5",
  
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)


df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "ECODA_NULL",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 80, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             labeller = labeller(score = score.labs))
```


#### Plot benchmark including ECODA+PB (detailed)

```{r fig.height=7, fig.width=6}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)


df_plot <- df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "ECODA_NULL",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"))


df_plot %>%
  filter(score == "anosim_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 80, vjust = 1, hjust = 1))



df_plot %>%
  filter(score == "cluster_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ARI score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))




df_plot %>%
  filter(score == "mod_knn3_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "Modularity score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
```




### Poster figures

```{r fig.height=10, fig.width=14}
# Poster figure (highlighted)

filter_methods <- c(
  "Pseudobulk",
  # "Freq_highres",
  # "ECODA_deconv",
  # "ECODA_lowres",
  "ECODA_highres",
  # "ECODA_high_res_2_PCA_dims",
  # "ECODA_highres_top_varexp0.7",
  # "ECODA_seuratres_0.4",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI",
  "scPoli",
  "PILOT",
  "GloScope"
)

p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "MOFA_2_factors" = "MOFA",
                         "scITD_2_factors" = "scITD",
                         "ECODA_highres" = "ECODA")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "ECODA", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "ECODA",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "ECODA", "black"))
print(p)
ggsave(file.path(path_plots, "POSTER_benchmark_summary_barplot_anosim_mainmethods_refECODA.pdf"),
       width = 3.5,
       height = 4)



filter_methods <- c(
  "Pseudobulk",
  "ECODA_highres",
  "ECODA_highres_top_varexp0.1"
)


p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_highres_top_varexp0" = "2 most highly variable cell types",
                         "ECODA_highres" = "ECODA")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "ECODA", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "ECODA",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "ECODA", "black"))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAvs2HVCvsBULK_refECODA.pdf"),
       width = 2,
       height = 6)




filter_methods <- c(
  "ECODA_deconv",
  "ECODA_lowres",
  "ECODA_highres",
  "ECODA_highres_HiTME",
  "ECODA_highres_HiTME_layer3",
  "ECODA_highres_scATOMIC",
  "ECODA_seuratres_0.4"
)

# my_comparisons <- list(
#   c("Low res. annot.", "High res. annot.")
#   )

p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_deconv" = "Deconvoluted",
                         "ECODA_lowres" = "Low res. author annot.",
                         "ECODA_highres" = "High res. author annot.",
                         "ECODA_highres_HiTME" = "HiTME annot.",
                         "ECODA_highres_HiTME_layer3" = "HiTME annot. w. gene. progs.",
                         "ECODA_highres_scATOMIC" = "scATOMIC annot.",
                         "ECODA_seuratres_0.4" = "Unsup. Leiden clust.")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "High res. author annot.", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.5) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "High res. author annot.",
    method = "wilcox.test",
    paired = TRUE,
    label.y = 1,
    tip.length = 0,
    aes(label = after_stat(p.signif))
  ) +
  ylim(c(-0.03, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1)) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "High res. author annot.", "black"))
print(p)
ggsave(file.path(path_plots, "POSTER_benchmark_summary_barplot_anosim_ecodamethods_refECODA_highres.pdf"),
       width = 3,
       height = 5)
```

### Other figures (not shown)

```{r}
ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores.pdf"),
       width = 7,
       height = 7)



df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

df_results_clustering <- df_results_clustering[!grepl("2|10|PCA", df_results_clustering$method), ]

for (ds in unique(df_results_clustering$dataset)) {
  number_cts <- dim(result_list$bmark[[ds]]$ECODA_highres$feat_mat)[2]
  df_results_clustering[df_results_clustering$dataset == ds, "number_celltypes"] <- number_cts
}

# df_results_clustering$dataset <- paste0(df_results_clustering$dataset, " (", df_results_clustering$number_celltypes, ")")
# legend_title <- "Dataset (number of cell types)"
legend_title <- "Dataset"

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_clustering)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)


my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "Pseudobulk"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(3, 6, 4, 5, 2, 1)]


p <- ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(group = legend_title, linetype = legend_title, shape = legend_title) +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_clustering.pdf"),
       width = 7,
       height = 5)



df_results_mod <- df_results %>%
  filter(score == "mod_score")

p <- ggplot(df_results_mod, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
  ) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_mod$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Modularity score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity.pdf"),
       width = 7,
       height = 6)
```



## Plot feature selection

Surprisingly, even at the lowest number highly variable cell types (i.e. variance explained of 10% -> >=2 cell types) the average biological separation is quite very high.
Different daatasets can do with a different number of highly variable cell types (separation plateaus out at different levels of variance explained % and number of cell types retained).

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_authors_HR", method) & !grepl("HiTME|scATOMIC|PCA|NULL|most", method)) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "100",
                         "ECODA_authors_HR_top_varexp0" = "2most_varcts")) %>%
  mutate(method = if_else(
    grepl("^top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  mutate(method = gsub("ECODA_authors_HR", "", method)) %>%
  mutate(method = gsub("top_varexp0", "2most_varcts", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

unique(df_results_sub$method)

p1 <- df_results_sub %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(0, 1)) +
  labs(x = "Variance explained (%)", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p1)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAfeatureselectionHVC.pdf"),
       width = 4,
       height = 4)

p2 <- df_results_sub %>%
  ggplot(aes(x = method, y = n_features)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  labs(x = "Variance explained (%)", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p2)

# Create p1 (the top plot) without x-axis labels
p1_clean <- p1 + theme(axis.title.x = element_blank(), 
                       axis.text.x = element_blank(),
                       axis.ticks.x = element_blank())

# Combine the plots vertically
combined_plot <- p2 / p1
print(combined_plot)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAfeatureselectionHVC_withNumCT.pdf"),
       width = 4,
       height = 5)


library(ggbreak)

filter_methods_features <- c(
  # "Pseudobulk",
  "ECODA_highres",
  "ECODA_highres_top_varexp0.7"
)

df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods_features) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = gsub("ECODA_highres", "All cell types", method)) %>%
  
  ggplot(aes(x = fct_reorder(method, n_features, .desc = TRUE, .fun = mean), y = n_features)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  labs(x = "", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_nfeatures.pdf"),
       width = 1.4,
       height = 4)
```

## Separation based on least variable cell types (sanity check)

```{r}
# Create an empty list to store the results from each dataset
temp_list <- list()

for (ds in names(result_list[["bmark"]])) {
  feat_mat <- result_list[["bmark"]][[ds]]$ECODA_highres$feat_mat
  labels <- label_list[[ds]]
  ncts <- 2
  
  # Calculate separation score for most highly variable cell types (HVC)
  top_hvcs <- get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
    get_hvcs(top_n_hvcs = ncts, variance_threshold = NULL)
  score_most_hvc <- calc_sep_score(feat_mat[, top_hvcs], labels)$anosim_score
  
  # Calculate separation score for least highly variable cell types (HVC)
  least_hvcs <- get_ct_var(feat_mat, show_plot = FALSE, descending = FALSE) %>%
    get_hvcs(top_n_hvcs = ncts, variance_threshold = NULL)
  score_least_hvc <- calc_sep_score(feat_mat[, least_hvcs], labels)$anosim_score
  
  # Store the results in a temporary data frame
  temp_df <- data.frame(
    dataset = ds,
    anosim_most_hvc = score_most_hvc,
    anosim_least_hvc = score_least_hvc
  )
  
  # Add the temporary data frame to the results list
  temp_list[[ds]] <- temp_df
}

# Combine all data frames in the list into a single data frame
df_plot <- bind_rows(temp_list)



# Reshape the data from wide to long format for ggplot
df_plot_long <- df_plot %>%
  pivot_longer(
    cols = starts_with("anosim_"),
    names_to = "group",
    values_to = "anosim_score"
  ) %>%
  # Clean up the group names for better readability
  mutate(group = gsub("anosim_|_hvc", "", group))

# Create the grouped bar plot
ggplot(df_plot_long, aes(x = dataset, y = anosim_score, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "ANOSIM Scores for Highly vs. Least Variable Cell Types",
    x = "Dataset",
    y = "ANOSIM Score",
    fill = "Cell Type Group" # Legend title
  ) +
  scale_fill_manual(values = c("most" = "darkred", "least" = "steelblue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Most highly variable cell types per dataset

```{r fig.height=8, fig.width=12}
print("Most highly variable cell types per dataset:")
for (ds in names(result_list[["bmark"]])) {
  print(ds)
  print(paste("ANOSIM score using all cell types:", round(result_list[["bmark"]][[ds]][["ECODA_highres"]]$scores$anosim_score, 3)))
  print(paste("ANOSIM score at 10% variance explained:", round(result_list[["bmark"]][[ds]][["ECODA_highres_top_varexp0.1"]]$scores$anosim_score, 3)))
  print("Most highly variable cell types (authors annotation):\n")
  print(paste(colnames(result_list[["bmark"]][[ds]][["ECODA_highres_top_varexp0"]][["feat_mat"]])))
  print("Most highly variable cell types (HiTME annotation):\n")
  print(paste(colnames(result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0"]] [["feat_mat"]])))
  
  p <- result_list[["bmark"]][[ds]][["ECODA_highres"]][["feat_mat"]] %>%
    get_ct_var() %>%
    varmeanplot(title = ds, label_points = TRUE)
  print(p)
  
  p <- plot_pca(result_list[["bmark"]][[ds]][["ECODA_highres_2least_varcts"]][["feat_mat"]], labels = label_list[[ds]], title = paste0(ds, ": Least variable cts"), invisible = "quali", labelsize = 4)
  print(p)
  p <- plot_pca(result_list[["bmark"]][[ds]][["ECODA_highres_top_varexp0"]][["feat_mat"]], labels = label_list[[ds]], title = paste0(ds, ": Most highly variable cts "), invisible = "quali", labelsize = 4)
  print(p)
}
```

```{r fig.height=8, fig.width=12}
print("Most highly variable cell types per dataset:")
for (ds in names(result_list[["bmark"]])) {
  print(ds)
  print("Most highly variable cell types (HiTME annotation):\n")
  print(paste(colnames(result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2_top_varexp0.5"]][["feat_mat"]])))
  print(paste(colnames(result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.5"]][["feat_mat"]])))
}
```




### Check IFN expression in Stephenson dataset

```{r}
ds <- "Stephenson"
seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))

plot_data <- data.frame(
  expression = seurat$HeatShock_UCell,
  status = seurat$Status
)

ggplot(plot_data, aes(x = expression, fill = status)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Expression Density by Status",
    x = "IFN_UCell Expression",
    y = "Density",
    fill = "Status"
  ) +
  xlim(c(0, 0.2)) +
  theme_minimal() +
  scale_fill_manual(values = c("Healthy" = "steelblue", "Covid" = "darkred"))
```


```{r}
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, "Status")
test <- process_coda_fig(seurat, labels, ct_col = "layer2", title = "ECODA\nHiTME layer2")
test

varexp_hvc <- 0.7
test1 <- process_coda_fig(seurat, labels, ECODA_top_varexp_hvct = varexp_hvc, ct_col = "layer2", title = paste0("ECODA layer2\nhigh res. var. exp. ", varexp_hvc*100, "%"))
test1


df_counts <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", "layer2")

feat_mat <- df_counts %>%
  impute_zeros(clr_zero_impute_method = clr_zero_impute_method, clr_zero_impute_num = clr_zero_impute_num) %>%
  calc_perc_df()

feat_mat <- feat_mat %>%
  clr()

get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
        get_hvcs(top_n_hvcs = 10, variance_threshold = NULL)
```

### Check different UCell score thresholds for HiTME layer3

```{r}
seurat <- replace_HiTMElayer3_annot(seurat)
```

```{r}
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, "Status")
test2 <- process_coda_fig(seurat, labels, ct_col = "layer3", title = "ECODA\nHiTME layer3")
test2

varexp_hvc <- 0.7
test3 <- process_coda_fig(seurat, labels, ECODA_top_varexp_hvct = varexp_hvc, ct_col = "layer3", title = paste0("ECODA layer3\nhigh res. var. exp. ", varexp_hvc*100, "%"))
test3


df_counts <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", "layer3")

feat_mat <- df_counts %>%
  impute_zeros(clr_zero_impute_method = clr_zero_impute_method, clr_zero_impute_num = clr_zero_impute_num) %>%
  calc_perc_df()

feat_mat <- feat_mat %>%
  clr()

get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
        get_hvcs(top_n_hvcs = 10, variance_threshold = NULL)
```



## Reduced separation with fewer cell types?
Why do some datasets show reduced separability when using fewer highly variable cell types while other still show good separation even with just 2 cell types?
Are there any comonalities between these datasets?

In general, datasets with HIGH ANOSIM score (~ >0.5) when using all cell types also show still good separability with very low number of cell types.
-> Maybe these biological conditions are more distinct and are easier to separate?
Datasets with LOW ANOSIM score (~ <0.5) when using all cell types also show decreasing separability when reducing the number of cell types.
-> The minimum number of cell types for maximal separability depends on the dataset

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_hi", method) & !grepl("HiTME|scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = if_else(
    grepl("^top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

ds_diffs <- df_results_sub %>%
  filter(method %in% c("10", "70")) %>%
  group_by(dataset) %>%
  summarize(
    diff = value[method == "10"] - value[method == "70"]
  ) %>%
  arrange(diff)

ggplot(ds_diffs, aes(x = reorder(dataset, diff), y = diff)) +
  geom_bar(stat = "identity") +
  labs(x = "Dataset", y = "Difference") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Which datasets have lower separation when using less cell types?\nVar. exp. 10% - Var. exp. 70%")
```


## Which datasets HiTME annotation underperforms?

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("^ECODA_highres$|^ECODA_highres_HiTME$", method) & !grepl("scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = gsub("ECODA_highres", "Author annotation", method)) %>%
  # mutate(method = if_else(
  #   grepl("^top_varexp0\\.\\d", method),
  #   paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
  #   method
  # )) %>%
  # mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

ds_diffs <- df_results_sub %>%
  group_by(dataset) %>%
  summarize(
    diff = value[method == "HiTME"] - value[method == "Author annotation"]
  ) %>%
  arrange(diff)

ggplot(ds_diffs, aes(x = reorder(dataset, diff), y = diff)) +
  geom_bar(stat = "identity") +
  labs(x = "Dataset", y = "Difference") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  ggtitle("For each dataset, how does ECODA perform based on\nHiTME annotation vs Author annotation?")
```



## Plot feature selection (HiTME layer3)

```{r fig.height=8, fig.width=12}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_hi", method) & !grepl("scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  # mutate(method = if_else(
  #   grepl("^top_varexp0\\.\\d", method),
  #   paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
  #   method
  # )) %>%
  # mutate(method = gsub("ECODA_highres", "100", method)) %>%
  # mutate(method = if_else(
  #   grepl("HiTME_layer3_top_varexp0\\.\\d", method),
  #   paste0("HiTME_layer3_", as.integer(gsub("HiTME_layer3_top_varexp0\\.", "", method)) * 10),
  #   method
  # )) %>%
  # mutate(method = gsub("^HiTME_layer3$", "HiTME_layer3_100", method)) %>%
  mutate(method = gsub("^HiTME$", "HiTME_layer2", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method)))) %>%
  # mutate(method = fct_relevel(method, "HiTME_layer2_100", after = Inf)) %>%
  mutate(method = fct_relevel(method, "HiTME_layer3", after = 5)) #%>%
  # filter(!grepl("varexp0|varcts", method))

p1 <- df_results_sub %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(0, 1)) +
  labs(x = "Variance explained (%)", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p1)
# ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAfeatureselectionHVC.pdf"),
#        width = 4,
#        height = 4)

p2 <- df_results_sub %>%
  ggplot(aes(x = method, y = n_features)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  labs(x = "Variance explained (%)", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p2)

# Create p1 (the top plot) without x-axis labels
p1_clean <- p1 + theme(axis.title.x = element_blank(), 
                       axis.text.x = element_blank(),
                       axis.ticks.x = element_blank())

# Combine the plots vertically
combined_plot <- p2 / p1
print(combined_plot)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAandHiTMEfeatureselectionHVC_withNumCT.pdf"),
       width = 5,
       height = 8)
```




## Show example PCAs for publication

```{r fig.height=7, fig.width=7}
for (ds in names(result_list[["bmark"]])) {
  p <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres"]][["plot"]][["data"]][["Groups"]],
    n_ct_show = 5,
    invisible = "quali",
    labelsize = 5
  ) + ggtitle(ds)
  print(p)
}

ds <- "Stephenson"
plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.9"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.9"]][["plot"]][["data"]][["Groups"]],
    n_ct_show = 50,
    invisible = "quali",
    labelsize = 3
) + ggtitle(ds)

```


```{r fig.height=6, fig.width=12}

levs <- c(
  "ECODA_high_res",
  "ECODA_low_res",
  "ECODA_deconv",
  "Pseudobulk",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI"
)


plot_list <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_cmv_females.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_age_females_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Lee"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Lee"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Lee.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Zhang"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Zhang"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Zhang.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Stephenson.pdf"),
       plot = p,
       width = 15,
       height = 7)

```

## Example PCAs for HiTME layer2 vs layer3


```{r}
plot_pca(result_list[["bmark"]][["Adams"]][["ECODA_highres_HiTME_layer3_top_varexp0.1"]][["feat_mat"]],
         labels = result_list[["bmark"]][["Adams"]][["ECODA_highres_HiTME_layer3_top_varexp0.1"]][["plot"]][["data"]][["Groups"]],
         invisible = "quali",
         labelsize = 4
         )
```

```{r}
test <- plist_layer2[[1]]
```


```{r}
plist_layer2 <- list()
for (ds in names(result_list$bmark)) {
  plist_layer2[[ds]] <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2_top_varexp0.5"]][["feat_mat"]],
    labels = label_list[[ds]],
    invisible = "quali",
    labelsize = 4,
    title = ds,
    n_ct_show = 5
  )
}

plist_layer3 <- list()
for (ds in names(result_list$bmark)) {
  plist_layer3[[ds]] <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.5"]][["feat_mat"]],
    labels = label_list[[ds]],
    invisible = "quali",
    labelsize = 4,
    title = ds,
    n_ct_show = 5
  )
}
```

```{r}
ggsave("pca_layer2.pdf", cowplot::plot_grid(plotlist = plist_layer2, ncol = 3), width = 20, height = 20)
ggsave("pca_layer3.pdf", cowplot::plot_grid(plotlist = plist_layer3, ncol = 3), width = 20, height = 20)
```


```{r fig.height=8, fig.width=16}
plist_vcts <- list()
for (ds in names(result_list$bmark)) {
  p1 <- result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2"]][["feat_mat"]]%>%
    get_ct_var() %>%
    varmeanplot(title = paste0(ds, " Layer2"), label_points = TRUE)
  p2 <- result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3"]][["feat_mat"]]%>%
    get_ct_var() %>%
    varmeanplot(title = paste0(ds, " Layer3"), label_points = TRUE)
  plist_vcts[[ds]] <- p1 | p2
  print(p1 | p2)
}


```

```{r}
ggsave(filename = "test.pdf", plot = cowplot::plot_grid(plotlist = plist_vcts, ncol = 1), width = 20, height = 100, limitsize = FALSE)
```


## Compare layer2 and layer3 by scores

```{r fig.height=8, fig.width=2.5}
filter_methods <- c(
  "ECODA_highres",
  "ECODA_highres_HiTME_layer2_top_varexp0.5",
  "ECODA_highres_HiTME_layer3_top_varexp0.5"
)

for (s in c("anosim_score", "mod_score", "cluster_score")) {
  p <- df_results %>%
    filter(score == s) %>%
    filter(method %in% filter_methods) %>%
    
    ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
    stat_summary(fun = mean, geom = "bar", fill = "gray70") +
    geom_point(position = "identity", alpha = 0.6) +
    geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
    labs(x = "", y = s) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
    ggtitle(s)
  print(p)
}
```


## Overlap between HVGs from single-cell and bulk
```{r}
# test <- c()
test_int <- c()
test_anosim <- c()
for (ds in names(datasets)) {
  # test <- c(test, result_list[[ds]][["Pseudobulk_unsup_hvg2000"]][["hvgs_sc_pb_intersect"]])
  test_int <- c(test_int, result_list[["bmark"]][[ds]][["Pseudobulk_unsup_hvg2000"]]$hvgs_sc_pb_intersect)
  test_anosim <- c(test_anosim, result_list[["bmark"]][[ds]][["Pseudobulk_unsup_hvg2000"]]$scores$anosim_score)
}
# plot(test)
plot(test_int, test_anosim)
```

# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot execution times

## Get execution times from R

```{r}
r_exec_times_path <- file.path(getwd(), "data", "r_execution_times.rds")

if (!file.exists(r_exec_times_path)) {
  r_exec_times <- data.frame(
    dataset = character(),
    method = character(),
    time_secs = numeric(),
    n_samples = numeric(),
    n_features = numeric()
  )
  
  bmark_result_list <- result_list[["bmark"]]
  
  ds_test <- names(bmark_result_list)
  
  # Exclude GongSharma_all_cmv and GongSharma_all_sex because they are the same 
  # dataset as GongSharma_all_age (so would just be replicated)
  ds_test <- ds_test[!ds_test %in% c("GongSharma_all_cmv", "GongSharma_all_sex")]
  
  for (ds in ds_test) {
    for (met in names(bmark_result_list[[ds]])) {
      if ("exec_time" %in% names(bmark_result_list[[ds]][[met]])) {
        temp <- data.frame(
          dataset = ds,
          method = met,
          time_secs = as.numeric(bmark_result_list[[ds]][[met]][["exec_time"]], units = "secs"),
          n_samples = nrow(bmark_result_list[[ds]][[met]][["feat_mat"]]),
          n_features = ncol(bmark_result_list[[ds]][[met]][["feat_mat"]])
        )
        
        r_exec_times <- rbind(r_exec_times, temp)
      }
    }
  }
  
  r_exec_times$dataset[r_exec_times$dataset == "GongSharma_all_age"] <- "GongSharma_all"
  
  saveRDS(object = r_exec_times, r_exec_times_path)
} else {
  r_exec_times <- readRDS(r_exec_times_path)
}
```


## Read execution times from python

```{r}
p_exec_times_path <- file.path(getwd(), "data", "p_execution_times.rds")

if (!file.exists(p_exec_times_path)) {
  p_exec_times <- read.table(
    file.path(getwd(), "data", "execution_times_formatted.txt"),
    header = TRUE
  )
  p_exec_times$n_samples <- 0
  p_exec_times$n_features <- 0
  
  for (ds in unique(p_exec_times$dataset)) {
    p_exec_times$n_samples[p_exec_times$dataset == ds] <- r_exec_times$n_samples[r_exec_times$dataset == ds][1]
    p_exec_times$n_features[p_exec_times$dataset == ds] <- r_exec_times$n_features[r_exec_times$dataset == ds][1]
  }
  
  saveRDS(object = p_exec_times, p_exec_times_path)
} else {
  p_exec_times <- readRDS(p_exec_times_path)
}
```


## Combine and get actual cell numbers used per dataset

```{r}
exec_times_path <- file.path(getwd(), "data", "exec_times.rds")

if (!file.exists(exec_times_path)) {
  exec_times <- rbind(r_exec_times, p_exec_times)
  exec_times$n_cells <- 0
  
  dss <- names(datasets)
  dss <- dss[!dss %in% c("GongSharma_all_cmv", "GongSharma_all_sex")]
  
  exec_times <- exec_times[!exec_times$dataset %in% c("Gongsharma_sex_young_cmvneg", "Gongsharma_cmv_females_young", "Gongsharma_age_females_cmvneg"), ]
  
  for (ds in dss) {
    print(ds)
    seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
    
    if (ds == "GongSharma_all_age") ds <- "GongSharma_all"
    
    exec_times$n_cells[exec_times$dataset == ds] <- nrow(seurat@meta.data)
  }
  
  saveRDS(object = exec_times, exec_times_path)
} else {
  exec_times <- readRDS(exec_times_path)
}
```

```{r}
unique(exec_times$method)
```

## Plot

```{r}
filter_methods <- c(
  "ECODA_authors_HR",
  "Pseudobulk_hvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors15",
  "GloScope_hvg2000_pcadims50",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000"
)

library(scales)
library(directlabels)

scientific_A_10n <- function(x) {
  s <- scales::scientific_format(digits = 2)(x)  # get scientific notation
  s_converted <- gsub("e\\+", " %*% 10^", s)  # replace 'e' with plotmath notation
  return(parse(text = s_converted))  # return as plotmath expression
}

exec_times %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors15" = "scITD",
                         "GloScope_hvg2000_pcadims50" = "GloScope",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "PILOT_hvg2000" = "PILOT")) %>%
  ggplot(aes(x = n_cells, y = time_secs, color = method)) +
  # geom_line() +
  geom_smooth(method = "lm", se = FALSE, size = 0.7) +
  geom_point(shape = 4, size = 1, stroke = 1.25) +
  labs(x = "Number of cells",
       y = "Time (s)") +
  theme_minimal() +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 5),  # Generate breaks
    labels = function(x) sapply(x, scientific_A_10n),  # Apply custom label function to breaks
    limits = c(min(exec_times$n_cells), max(exec_times$n_cells) * 1.15)
  ) +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                labels = trans_format("log10", math_format(10^.x))) +
  geom_dl(aes(label = method),
          method = list(dl.combine("last.points"), hjust = -0.1, cex = 0.8),
          cex = 1) +
  # xlim(min(exec_times$n_cells), max(exec_times$n_cells) * 1.15) +
  NoLegend()
ggsave(file.path(path_plots, "benchmark_exec_times.pdf"),
       width = 6,
       height = 3.5, limitsize = FALSE)
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot benchmark results

## Main plot (with funkyheatmap)

```{r fig.height=6, fig.width=16}
library(funkyheatmap)

# --- 1. SETUP & CLEANING (Same as before) ---
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_HiTME_HR_layer3",
  "ECODA_seuratres_2",
  "ECODA_PB_combo_normquantile_ecodaweight0.5",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)

min_max <- function(x) {
  if (all(is.na(x))) return(x)
  if (max(x, na.rm=T) == min(x, na.rm=T)) return(x)
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

df_results_bmark <- df_results
df_results_bmark$temp <- paste0(df_results_bmark$dataset, df_results_bmark$method)
exec_times$temp <- paste0(exec_times$dataset, exec_times$method)

df_results_bmark <- merge(df_results_bmark,
                          exec_times[, c("temp", "time_secs")], 
                          by = "temp", 
                          all.x = TRUE)

df_results_bmark$temp <- NULL
exec_times$temp <- NULL


clean_data <- df_results_bmark %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "ECODA_NULL",
                         # "ECODA_HiTME_HR_layer2" = "ECODA_automated_(HiTME_layer2)",
                         # "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "ECODA_PB_combo_normquantile_ecodaweight0.5" = "ECODA+Pseudobulk",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  mutate(score = recode(score,
                        "anosim_score" = "ANOSIM",
                        "mod_knn3_score" = "Modularity",
                        "cluster_score" = "ARI")) %>%
  mutate(dataset = recode(dataset,
                          "GongSharma_all_age" = "GongSharma_age",
                          "GongSharma_all_cmv" = "GongSharma_cmv",
                          "GongSharma_all_sex" = "GongSharma_sex"))

# --- 2. CALCULATE GEOMETRIC MEAN OF RANKS (NEW STEP) ---
df_ranks <- clean_data %>%
  # Group by specific test case
  group_by(dataset, score) %>% 
  # Rank: Higher Value = Better Rank (1). Ties get the minimum rank.
  mutate(rank = rank(-value, ties.method = "min")) %>% 
  ungroup() %>%
  # Now aggregate per method
  group_by(method) %>% 
  # Calculate Geometric Mean: exp(mean(log(x)))
  summarise(geo_mean_rank = exp(mean(log(rank), na.rm = TRUE))) %>%
  ungroup() %>%
  # Normalize so it plots nicely (0 to 1). 
  # We INVERT it so Best Rank (lowest number) = 1.0 (biggest bar)
  mutate(Rank_Score = (max(geo_mean_rank) - geo_mean_rank) / 
           (max(geo_mean_rank) - min(geo_mean_rank))) %>%
  select(method, Rank_Score)

# --- 3. SCALE DATA (0-1) FOR HEATMAP VISUALS ---
scaled_data <- clean_data %>%
  group_by(dataset, score) %>%
  mutate(value = min_max(value)) %>%
  ungroup()

# --- 4. AGGREGATE SCORES ---

# A. Overall Score (Arithmetic Mean of Scaled Values)
df_overall <- scaled_data %>%
  group_by(method) %>%
  summarise(Overall_Score = mean(value, na.rm = TRUE)) 

# B. Dataset Scores
df_datasets_wide <- scaled_data %>%
  group_by(method, dataset) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = dataset, values_from = value, names_prefix = "Dataset__")

# C. Metric Scores
df_metrics_wide <- scaled_data %>%
  group_by(method, score) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = score, values_from = value, names_prefix = "Metric__")

# D. Execution times
df_runtime <- clean_data %>%
  # distinct() is crucial! clean_data has 3 rows per dataset (one per score).
  # We only want to count the execution time once per dataset.
  distinct(method, dataset, time_secs) %>% 
  group_by(method) %>%
  summarise(value = mean(time_secs, na.rm = TRUE), .groups = "drop") %>%
  # IMPORTANT: We must scale this 0-1 because your funkyheatmap has scale_column = FALSE
  mutate(value = min_max(value)) %>% 
  rename(Runtime = value)

# --- 5. MERGE EVERYTHING ---
final_data <- df_overall %>%
  left_join(df_ranks, by = "method") %>% # <--- Join the new Rank Score
  left_join(df_datasets_wide, by = "method") %>%
  left_join(df_metrics_wide, by = "method") %>%
  left_join(df_runtime, by = "method") %>%
  arrange(desc(Rank_Score)) # Optional: sort by the new Rank Score instead

# --- 6. BUILD COLUMN INFO ---

# 1. Fixed Columns (Updated to include Rank_Score)
info_fixed <- tribble(
  ~id,             ~group,    ~name,           ~geom,  ~palette,        ~options,
  "method",        "",        "",              "text", NA,              list(hjust = 0, width = 8),
  "Rank_Score",    "Overall", "GeoMean Rank",  "bar",  "palette_avg_rank",   list(width = 4, legend = FALSE),
  "Overall_Score", "Overall", "Overall Score", "bar",  "palette_avg_mean",   list(width = 4, legend = FALSE)
)

# 2. Dataset Columns
dataset_cols <- colnames(final_data)[startsWith(colnames(final_data), "Dataset__")]
info_datasets <- tibble(id = dataset_cols) %>%
  mutate(
    group = "Datasets",
    name = str_remove(id, "Dataset__"), 
    geom = "funkyrect",
    palette = "palette_A",
    options = list(list())
  ) %>% arrange(name)

# 3. Metric Columns
metric_cols <- colnames(final_data)[startsWith(colnames(final_data), "Metric__")]
target_metric_order <- c("ANOSIM", "Modularity", "ARI")
info_metrics <- tibble(id = metric_cols) %>%
  mutate(
    group = "Metrics",
    name = str_remove(id, "Metric__"),
    geom = "funkyrect",
    palette = "palette_B",
    options = list(list()),
    name_f = factor(name, levels = target_metric_order)
  ) %>% arrange(name_f) %>% select(-name_f)

# 4. Runtime Column (Manually added since it's just one column now)
info_runtime <- tribble(
  ~id,       ~group,      ~name,         ~geom,       ~palette,     ~options,
  "Runtime", "Resources", "Runtime (s)", "funkyrect", "palette_C",  list(legend = FALSE)
)

# Combine
column_info <- bind_rows(info_fixed, info_datasets, info_metrics, info_runtime)

# --- 7. PLOT ---
funky_heatmap(
  final_data,
  column_info = column_info,
  scale_column = FALSE, 
  column_groups = tibble(
    group = c("Overall", "Datasets", "Metrics", "Resources")
  ),
  palettes = list(
    palette_avg_rank = "Greys",
    palette_avg_mean = "Greys",
    palette_A = "Blues",
    palette_B = "Reds",
    palette_C = "YlOrBr"),
  position_args = position_arguments(
    col_width = 1.5,
    col_space = 0.2
  )
)
```



# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot normalization results

## Plot results

```{r}
df_results <- Reduce(rbind, result_list[["trans"]]) %>%
  mutate(trans_method = recode(trans_method,
                         "clr" = "CLR",
                         "alr_mincvref" = "ALR_minCVref",
                         "alr_randref" = "ALR_randref",
                         "arcsine_sqrt" = "Arcsine_sqrt",
                         "counts" = "Counts",
                         "freq" = "Frequency(%)"))

df_results_list[["Normalization"]] <- df_results


df_results <- df_results %>%
  pivot_longer(
    cols = c(ANOSIM_score, Modularity_score, Adjusted_Rand_Index),
    names_to = "score",
    values_to = "value"
  ) %>%
  as.data.frame()

df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score") %>%
  group_by(trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_anosim <- df_results_anosim %>%
  arrange(desc(mean))

df_results_anosim$trans_method <- factor(df_results_anosim$trans_method, levels = df_results_anosim$trans_method)
df_results$trans_method <- factor(df_results$trans_method, levels = df_results_anosim$trans_method)



df_results %>%
  group_by(score, trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = trans_method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores_bars.pdf"),
       width = 7,
       height = 7)


df_results %>%
  filter(score == "ANOSIM_score") %>%
  
  ggplot(aes(x = trans_method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..))),
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 0.9
  ) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  ylim(c(-0.1, 1))
ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_anosim_refCLR.pdf"),
       width = 2.5,
       height = 3.5)



my_comparisons <- list(
  c("CLR", "Frequency(%)"),
  c("CLR", "Counts")
  )

df_results %>%
  filter(score == "ANOSIM_score") %>%
  filter(trans_method %in% c("Counts", "Frequency(%)", "CLR")) %>%
  
  ggplot(aes(x = fct_reorder(trans_method, value, .desc = FALSE, .fun = mean), y = value, fill = ifelse(trans_method == "CLR", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    paired = TRUE,
    size = 3,
    tip.length = 0,
    # label.y = 0.95,
    aes(label = after_stat(p.signif))
  ) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  ylim(c(-0.05, 1)) +
  scale_fill_identity() +
  scale_x_discrete(labels = c(
    "Counts",
    "Frequency(%)",
    expression(bold("CLR")) # Make label bold
  ))
ggsave(file.path(path_plots, "POSTER_transformations_real_datasets_summary_barplot_anosim_refCLR.pdf"),
       width = 2,
       height = 3.5)





# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "ANOSIM score", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))



ggplot(df_results, aes(x = trans_method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
# facet_grid(scales = "free",
#            rows = vars(score_method), cols = vars(dataset))
ggsave(file.path(path_plots, "transformations_real_datasets_detailed_barplot.pdf"),
       width = 8,
       height = 7)



# df_results %>%
#   group_by(score_method, method) %>%
#   summarize(mean = mean(values), sem = sd(values) / sqrt(length(values))) %>%
#   ungroup() %>%
#   
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Separation score") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   facet_grid(scales = "free",
#              rows = vars(score_method),
#              labeller = labeller(score_method = score.labs))


ggplot(df_results, aes(x = trans_method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores.pdf"),
       width = 5,
       height = 5)



# df_results_anosim %>%
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Adjusted Rand Index") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   NoLegend()


df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score")

df_results_anosim <- df_results_anosim[!grepl("pca", df_results_anosim$trans_method), ]

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ trans_method, data = df_results_anosim)
summary(anova_result)
stat.test <- tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)
# stat.test <- stat.test[["method"]] %>% as.data.frame()
# 
# # Extract row names where p adj < 0.05
# significant_comparisons <- rownames(stat.test)[stat.test[["p adj"]] < 0.05]
# 
# # Initialize an empty list to store results
# my_comparisons <- list()
# 
# # Split and store each comparison in the list
# for (comparison in significant_comparisons) {
#   parts <- strsplit(comparison, "-")[[1]]
#   my_comparisons <- c(my_comparisons, list(parts))
# }

my_comparisons <- purrr::map(unique(as.character(df_results_anosim$trans_method)), ~ c(.x, "Frequency(%)"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
# my_comparisons <- c(my_comparisons, list(c("clr", "arcsine_sqrt")))
my_comparisons <- my_comparisons[c(4, 3, 2, 1, 5)]

# my_comparisons <- list(
#   c("alr_mincvref", "freq"),
#   c("alr_randref", "freq"),
#   c("arcsine_sqrt", "freq"),
#   c("clr", "freq"),
#   c("clr_pca", "freq"),
#   c("counts", "freq")
#   )

ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    # ref.group = "freq",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "ANOSIM score") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #                    y.position = c(max(df_results_anosim$values)+0.1, max(df_results_anosim$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot.pdf"),
       width = 6,
       height = 4.5)


ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    # comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "ANOSIM score") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #                    y.position = c(max(df_results_anosim$values)+0.1, max(df_results_anosim$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

```

# Statistical comparison

```{r}
df_results_anosim <- df_results %>%
  filter(score_method == "ANOSIM_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(values ~ trans_method, data = df_results_anosim)
summary(anova_result)

TukeyHSD(aov(df_results_anosim$values ~ df_results_anosim$trans_method))
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot zero imputation results

```{r}
zeroimp_result_list <- result_list[["zeroimp"]]
```


## Plot results

```{r}
df_results <- imap_dfr(zeroimp_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    imap_dfr(.x, ~ {
      tibble(
        dataset = dataset_name,
        method = method_name,
        score = .y,
        value = .x
      )
    })
  })
})

df_results <- df_results[df_results$method != "impRZilr", ]

df_results_list[["Zero_imputation"]] <- df_results

df_results %>%
  filter(score == "anosim_score") %>%
  filter(method != "asinsqrt") %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_zeros_1",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.93,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.15, 1)) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_barplot_ANOSIM.pdf"),
       width = 4,
       height = 4)


# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "ANOSIM score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "zero_imputation_real_datasets_detailed_barplot.pdf"),
       width = 10,
       height = 6)




df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%

  ggplot(aes(x = method, y = mean)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_barplot_separatescores.pdf"),
       width = 3.5,
       height = 6)

ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_separatescores.pdf"),
       width = 7,
       height = 6)



df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results %>%
  filter(score == "anosim_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "counts_all_1"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(6,1,7:10, 2:5)]
# my_comparisons[[length(my_comparisons) + 1]] <- c("perc_all_0.1%", "perc_all_1%")

p <- ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_clusterscore.pdf"),
       width = 8,
       height = 6)

df_results_modularity <- df_results %>%
  filter(score == "mod_score")

p <- ggplot(df_results_modularity, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Modularity score") +
  scale_shape_manual(values=1:length(unique((df_results_modularity$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_modscore.pdf"),
       width = 8,
       height = 6)


df_results_anosim <- df_results %>%
  filter(score == "anosim_score")

p <- ggplot(df_results_anosim, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_anosimscore.pdf"),
       width = 8,
       height = 6)

```

# Test for statistical significance

To few data points for statistical significance

```{r}
library(rstatix)

df_results_cluster <- df_results %>% filter(score == "cluster_score")

outlier <- df_results_cluster %>%
  group_by(method) %>%
  rstatix::identify_outliers(value)
data.frame(outlier)

normality <- df_results_cluster %>%
  group_by(method) %>%
  rstatix::shapiro_test(value)
data.frame(normality)
cat(paste("All groups normally ditributed?", all(normality$p > 0.05),
            "\nNot normally distributed:", normality$method[normality$p <= 0.05]))

# res <- rstatix::anova_test(data = df_results_cluster, dv = value, wid = dataset, within = method) 
# rstatix::get_anova_table(res)
# 
# pair_t <- df_results_cluster %>%
#   rstatix::pairwise_t_test(value~method, paired=TRUE, p.adjust.method = "BH" )
# data.frame(pair_t)

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)




df_results_cluster <- df_results %>% filter(score == "mod_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)



df_results_cluster <- df_results %>% filter(score == "sil_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)
```

# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Export tables with raw data for paper

```{r}
library(writexl)
write_xlsx(df_results_list, path = file.path(path_data, "raw_data_for_publication.xlsx"))
```

