---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
# renv::init()
# renv::snapshot()

source("functions.R")
source("Processed_dataset_metadata.R")


# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")

# reticulate::virtualenv_create(envname = "C:/R_Envs/my_env", version = "3.11.9")
# # Define the path first
# my_path <- "C:/R_Envs/my_env"
# 
# # Install BEFORE the environment is loaded into the R session memory
# reticulate::py_install(
#     packages = "pandas==2.0.3",
#     envname = my_path,
#     method = "virtualenv",
#     pip = TRUE
# )
# reticulate::py_install(
#     packages = "h5py==3.12.1",
#     envname = my_path,
#     method = "virtualenv",
#     pip = TRUE
# )
# reticulate::py_install(
#     packages = "numpy==1.26.4",
#     envname = my_path,
#     method = "virtualenv",
#     pip = TRUE
# )
# reticulate::py_install(
#     packages = "mofapy2",
#     envname = my_path,
#     method = "virtualenv",
#     pip = TRUE
# )
library(reticulate)
use_virtualenv("C:/R_Envs/my_env", required = TRUE)
py_config()
# py_install("mofapy2")
import("mofapy2")
# library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
# BiocManager::install("GloScope")
library(GloScope)
library(funkyheatmap)
library(RColorBrewer)

library(forcats)
library(gtools)

# Set paths
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"

set.seed(123)
```

# Create results list to fill

```{r}
result_file <- "result_list.rds"

if (file.exists(result_file)) {
  result_list <- readRDS(result_file)
} else {
  result_list <- list()
}
```


# ---------------------------------------------------------------


# Process datasets

```{r message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
for (ds in names(datasets)) {
  seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
  seurat <- replace_HiTMElayer3_annot(seurat, thresh = 0.1)
  
  seurat@misc$label_col <- datasets[[ds]][["label_col"]]
  seurat@misc$low_res_ct_col <- datasets[[ds]][["low_res_ct_col"]]
  seurat@misc$hi_res_ct_col <- datasets[[ds]][["hi_res_ct_col"]]
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots)
}
```

# ---------------------------------------------------------------

# Gong and Sharma

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```

### Check for batch effect

```{r}
alpha <- 0.3

meta_cols_plot <- c(
  "Sample",
  "RNA_snn_res.0.5",
  "batch_id",
  "pool_id",
  "chip_id",
  "well_id",
  "cohort.cohortGuid",
  "AIFI_L1",
  "AIFI_L2",
  "AIFI_L3",
  "subject.cmv"
)

# Check in broad cell types (strong global batch effect)

Idents(seurat) <- "Sample"
sub <- subset(x = seurat, downsample = 100)

sub <- sub |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(dims = 1:50, verbose = FALSE) |>
  FindNeighbors() |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:50)

for (mc in meta_cols_plot) {
  Idents(sub) <- mc
  DimPlot(sub, reduction = "umap", alpha = alpha)
}

# No strong batch effect observed


# To check for more subtle batch effect, we can subset to only T cells, where the gene expression difference between cells is smaller
# Thus, by subseting to only T cells, we should be better able to see smaller batch effects

sub <- subset(x = seurat, subset = AIFI_L1 == "T cell")
sub <- subset(x = sub, downsample = 1000)

sub <- sub |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(dims = 1:50, verbose = FALSE) |>
  FindNeighbors() |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:50)

for (mc in meta_cols_plot) {
  Idents(sub) <- mc
  DimPlot(sub, reduction = "umap", alpha = alpha)
}

# No weak batch effect observed
```

## Age (females cmvneg)

```{r}
ds <- "GongSharma_age_females_cmvneg"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  seurat <- load_h5ad_to_seurat("data/Gongsharma_age_females_cmvneg.h5ad")
  seurat$Sample <- seurat$specimen.specimenGuid
  
  seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))

  seurat@misc$label_col <- "age"
  seurat@misc$low_res_ct_col <- "AIFI_L1"
  seurat@misc$hi_res_ct_col <- "AIFI_L3"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
}
```

# ---------------------------------------------------------------

# Bassez

# IMPORTANT: Needs automated annotation, as seurat\$cohort c("neoadjuvant_chemo", "treatment_naive") seem to be separately and differently annotated by the authors

## Has also other non-immune cells

```{r}
ds <- "Bassez"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))

  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  ds_name <- "BassezA_2021_33958794"

  path_root <- file.path(path_nas, ds_name)
  path_output <- file.path(path_root, "output")

  seurat <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name)))

  seurat <- subset(seurat, subset = timepoint == "Pre")
  seurat <- subset(seurat, subset = expansion != "n/a")
  seurat <- subset(seurat, subset = cohort == "treatment_naive")

  seurat$Sample <- seurat$sample
  seurat$cellSubType[is.na(seurat$cellSubType)] <- seurat$cellType[is.na(seurat$cellSubType)]

  convert_seurat_to_h5ad(seurat, ds)

  seurat@misc$label_col <- "expansion"
  seurat@misc$low_res_ct_col <- "cellType"
  seurat@misc$hi_res_ct_col <- "cellSubType"

  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
}
```


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Create list to export raw data to excel

```{r}
df_results_list <- list()
```

# ---------------------------------------------------------------

# ---------------------------------------------------------------

# Plot benchmark results

## Save and flatten the nested list into a DataFrame

```{r}
bmark_result_list <- result_list[["bmark"]]

bmark_result_list$GongSharma <- bmark_result_list$Gongsharma_cmv_young_males
bmark_result_list$Gongsharma_cmv_young_males <- NULL

df_results <- imap_dfr(bmark_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    if ("scores" %in% names(.x)) {
      n_samples <- nrow(.x$feat_mat)
      n_features <- ncol(.x$feat_mat)
      
      imap_dfr(.x$scores, ~ {
        tibble(
          dataset = dataset_name,
          method = method_name,
          score = .y,
          value = .x,
          n_samples = n_samples,
          n_features = n_features
        )
      })
    } else {
      tibble() # Return an empty tibble if "scores" is not found
    }
  })
})

# Drop old modularity score method
df_results <- df_results[!df_results$score == "mod_score", ]
```

## Show + sort all tested conditions

```{r}
levs <- mixedsort(as.character(unique(df_results$method)))
levs

df_results$method <- factor(df_results$method, levels = levs)

df_results$dataset[df_results$dataset == "GongSharma_females_cmv"] <- "GongSharma_cmv"
df_results$dataset[df_results$dataset == "GongSharma_females_cmvneg_age"] <- "GongSharma_age"

df_results_list[["Benchmark"]] <- df_results
```

## Plot results

### All methods (overview)

```{r fig.height=8, fig.width=16}
# New facet label names for dose variable
score.labs <- c("ANOSIM score", "Adjusted rand index", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            labeller = labeller(score = score.labs, dataset = datas.labs),
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "benchmark_detailed_barplot.pdf"),
       width = 150,
       height = 7, limitsize = FALSE)


df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_bars.pdf"),
       width = 14,
       height = 7)
```

### Supp fig 16: Benchmark unscaled metrics

```{r fig.width=12, fig.height=5}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  # "ECODA_HiTME_HR_layer2",
  # "ECODA_HiTME_HR_layer3",
  # "ECODA_seuratres_2",
  # "ECODA_PB_combo_normquantile_ecodaweight0.5",
  # paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  # "ECODA_authors_LR",
  # "Avg_PCA_embedding",
  # "Freq_highres",
  # "ECODA_deconv",
  # "Pseudobulk_10_PCA_dims",
  "Pseudobulk_hvg2000",
  # "Pseudobulk_hvg2000_bl",
  # "Pseudobulk_schvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims5",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat",
  "GloProp"
)


df_plot <- df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         # "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         # "ECODA_PB_combo_normquantile_ecodaweight0.5" = "ECODA+Pseudobulk",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims5" = "scPoli",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"))

pointsize <- 1.5

# Plot 1: ANOSIM
p1 <- df_plot %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

# Plot 2: Modularity
p2 <- df_plot %>%
  filter(grepl(pattern = "mod", score)) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "", y = "Modularity score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

# Plot 3: ARI
p3 <- df_plot %>%
  filter(score == "cluster_score") %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "", y = "ARI score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

# --- Combine using Patchwork ---

combined_plot <- (p1 | p2 | p3) + 
  plot_layout(guides = "collect") & # Collects all legends into one
  theme(legend.position = "right") # Places the shared legend at the bottom
combined_plot

# Save the combined plot
ggsave(file.path(path_plots, "Supp_fig_16_benchmark_unscaled_metrics_perdataset.pdf"), 
       plot = combined_plot,
       width = 12, # Increased width to fit 3 plots
       height = 5)
```

```{r fig.height=4, fig.width=4}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  "Avg_PCA_embedding",
  "Freq_highres",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims5",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30"
)


df_plot_scaled <- df_results %>%
  filter(method %in% filter_methods) %>%
  # -- Recoding names --
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_Leiden_clustering",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims5" = "scPoli",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30" = "GloScope")) %>%
  # -- CRITICAL STEP: Min-Max Scaling per Dataset and Score --
  group_by(dataset, score) %>% 
  mutate(scaled_value = min_max(value)) %>% 
  ungroup()

# 4. Generate the Plot
# We filter for ANOSIM and use 'scaled_value' instead of 'value'
df_plot_scaled %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = fct_reorder(method, scaled_value, .desc = TRUE, .fun = mean), 
             y = scaled_value)) + # Use the scaled column
  
  # Error bars and Mean Bar
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  
  # Individual Points and connecting lines (Paired data)
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  
  # Statistical Test
  stat_compare_means(
    ref.group = "ECODA_authors",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    vjust = 1,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  
  # Adjust Y-axis (Since data is 0-1, we set limit slightly higher for p-values)
  ylim(c(0, 1.15)) + 
  
  labs(x = "", y = "Separation score\n(Scaled ANOSIM score)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_ANOSIM_minmaxscaled_mainmethods_refECODA.pdf"),
       width = 4.5,
       height = 4.5)
```


#### Plot all separation metrics together

```{r fig.height=6, fig.width=5}
df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_allmetricscombined_mainmethods_with_bad_baselines.pdf"),
       width = 4.5,
       height = 6.5)
```

#### Supp fig 18: Plot PCA for selected datasets and main methods

```{r}
ds <- "Kfoury"

m <- "ECODA_authors_HR"
p1 <- plot_pca(
  result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
  labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
  coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
  title = m, n_ct_show = 5, invisible = c("quali")
)
print(p1)

plot_mds(
  result_list[["bmark"]][[ds]][[m]][["dist_mat"]],
  labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
  coord_equal = TRUE, cluster_score = FALSE, mod_score = FALSE,
  title = m
)

m <- "ECODA_HiTME_HR_layer2"
p2 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_HiTME", n_ct_show = 5, invisible = c("quali")
  )
print(p2)

m <- "ECODA_scATOMIC_HR"
p3 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_scATOMIC", n_ct_show = 5, invisible = c("quali")
  )
print(p3)

p_kfoury <- wrap_plots(list(p1, p2, p3)) +
  plot_layout(nrow = 1, guides = "collect")
ggsave(file.path(path_plots, paste0("Supp_fig_18_Kfoury_PCA_authors_vs_auto_", ds, ".pdf")),
       width = 15,
       height = 6)


ds <- "Wu"

m <- "ECODA_authors_HR"
p1 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = m, n_ct_show = 5, invisible = c("quali")
  )
print(p1)

m <- "ECODA_HiTME_HR_layer2"
p2 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_HiTME", n_ct_show = 5, invisible = c("quali")
  )
print(p2)

m <- "ECODA_scATOMIC_HR"
p3 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_scATOMIC", n_ct_show = 5, invisible = c("quali")
  )
print(p3)

p_wu <- wrap_plots(list(p1, p2, p3)) +
  plot_layout(nrow = 1, guides = "collect")
ggsave(file.path(path_plots, paste0("Supp_fig_18_Wu_PCA_authors_vs_auto_", ds, ".pdf")),
       width = 15,
       height = 6)

p_combined <- wrap_plots(list(p_kfoury + labs(tag = "Kfoury"), p_wu + labs(tag = "Wu"))) +
  plot_layout(nrow = 2) & 
  theme(plot.tag = element_text(face = "bold"))
ggsave(file.path(path_plots, paste0("Supp_fig_18_combined.pdf")),
       width = 15,
       height = 6)
```

```{r fig.height=12, fig.width=12}
ds <- "Pelka"
plot_list <- list()

methods <- list(
  "ECODA_authors_HR" = "ECODA",
  "ECODA_authors_HR_NULL" = "Shuffled_baseline",
  "Pseudobulk_10_PCA_dims" = "Pseudobulk",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scITD_hvg2000_factors5" = "scITD",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims5" = "scPoli",
  "PILOT_hvg2000" = "PILOT",
  "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"
)

for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 3, guides = "collect")
```


### Plot PCA for all datasets (PB and ECODA)

```{r fig.height=4, fig.width=8}
methods <- list(
  "Pseudobulk_hvg2000" = "Pseudobulk",
  # "ECODA_HiTME_HR_layer2" = "ECODA_automated",
  "ECODA_authors_HR" = "ECODA"
)

for (ds in names(result_list[["bmark"]])) {
  plot_list <- list()
  
  for (m in names(methods)) {
    feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
    labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
    
    if (ds=="Bassez") levels(labels) <- c("Responders", "Non-responders")
    if (ds=="Zhang") levels(labels) <- tools::toTitleCase(levels(labels))
    
    if (methods[[m]] == "ECODA") title <- "Cell type composition (ECODA)" else title <- methods[[m]]
    
    plot_list[[m]] <- plot_pca(
      feat_mat = feat_mat,
      labels = labels,
      cluster_score = FALSE, mod_score = FALSE, anosim_score = FALSE,
      coord_equal = FALSE,
      n_ct_show = 5, invisible = "quali", repel = TRUE,
      title = title
    )
    
    if (ds=="Bassez") {
      plot_list[[m]] <- plot_list[[m]] +
        scale_color_discrete(
          name = "Treatment response", 
          labels = labels
        ) +
        scale_shape_discrete(
          name = "Treatment response",
          labels = labels
        )
    }
    if (ds=="Zhang") {
      plot_list[[m]] <- plot_list[[m]] +
        scale_color_discrete(
          name = "Tissue", 
          labels = labels
        ) +
        scale_shape_discrete(
          name = "Tissue", 
          labels = labels
        )
    }
    
  }
  
  wrap_plots(plot_list) +
    plot_layout(nrow = 1, guides = "collect")
  
  ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
         width = 8,
         height = 3.5)
}

```


### Supp fig 2-12: MDS for all datasets

```{r fig.height=9, fig.width=11}
methods <- list(
  "ECODA_authors_HR" = "ECODA",
  # "ECODA_seuratres_2" = "ECODA_Leiden_res_2",
  # "ECODA_HiTME_HR_layer2" = "ECODA_automated",
  
  # "Freq_highres" = "ECODA_Freq%",
  "Pseudobulk_hvg2000" = "Pseudobulk",
  # "Avg_PCA_embedding" = "Avg_PCA_embedding",
  
  "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope",
  "GloProp" = "GloProp",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims15_highres" = "scPoli",
  
  "MOFA_hvg2000_factors15" = "MOFA",
  "PILOT_hvg2000" = "PILOT",
  "scITD_hvg2000_factors5" = "scITD"
)

idx <- 2

for (ds in names(result_list[["bmark"]])) {
  plot_list <- list()
  
  for (m in names(methods)) {
    dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
    labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
    
    if (ds == "Bassez") {
      labels <- as.character(labels)
      labels[labels == "E"] <- "Responders"
      labels[labels == "NE"] <- "Non-responders"
      labels <- factor(labels, levels = c("Responders", "Non-responders"))
    }
    
    plot_list[[m]] <- plot_mds(
      dist_mat,
      labels,
      cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
      coord_equal = TRUE,
      title = methods[[m]]
    )
  }
  
  wrap_plots(plot_list) +
    plot_layout(nrow = 3, guides = "collect")
  
  ggsave(file.path(path_plots, paste0("Supp_fig_", idx, "_benchmark_mds_", ds, ".pdf")),
         width = 11,
         height = 9)
  
  idx <- idx + 1
}

```

```{r fig.height=6, fig.width=6}
m <- "PILOT_hvg2000"
ds <- "Zhang"
dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]

plot_mds(
  dist_mat,
  labels,
  cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
  coord_equal = TRUE,
  axes = c(10,11),
  title = methods[[m]]
)
```


### Compare PCA based on features vs dist

```{r fig.height=6, fig.width=18}
ds <- "Pelka"
plot_list <- list()

methods <- list(
  "ECODA_authors_HR" = "ECODA_authors",
  "ECODA_authors_HR_NULL" = "Shuffled_baseline",
  "Freq_highres" = "ECODA_Freq%",
  "ECODA_seuratres_2" = "ECODA_unsupervised",
  "ECODA_HiTME_HR_layer2" = "ECODA_automated",
  "Pseudobulk_hvg2000" = "Pseudobulk",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scITD_hvg2000_factors5" = "scITD",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims5" = "scPoli",
  "PILOT_hvg2000" = "PILOT",
  "GloScope_hvg2000_pcadims30" = "GloScope"
)

for (m in names(methods)) {
  feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  if (ds == "Pelka") {
    labels <- as.character(labels)
    labels[labels == "N"] <- "Normal"
    labels[labels == "T"] <- "Tumor"
  }
  if (ds == "Bassez") {
    labels <- as.character(labels)
    labels[labels == "E"] <- "Expanders"
    labels[labels == "NE"] <- "Non-expanders"
  }
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  avg_sep_score <- result_list[["bmark"]][[ds]][[m]]$scores[c("mod_knn3_score", "anosim_score", "cluster_score")] %>%
    unlist() %>%
    mean() %>%
    round(2)
  
  plot_list[[m]] <- plot_pca(
    feat_mat = feat_mat,
    # cluster_score = F, mod_score = F, anosim_score = F,
    # knn_k = NULL,
    labels = labels,
    coord_equal = FALSE,
    title = paste0(methods[[m]], "\nSeparation score: ", sprintf("%.2f", avg_sep_score))
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect") +
  plot_annotation(title = ds)
# ggsave(file.path(path_plots, paste0("PCA_", ds, ".pdf")),
#        width = 18,
#        height = 6)

```

```{r}
for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  if (ds == "Pelka") {
    labels <- as.character(labels)
    labels[labels == "N"] <- "Normal"
    labels[labels == "T"] <- "Tumor"
  }
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    cluster_score = F, mod_score = F, anosim_score = F,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect") +
  plot_annotation(title = ds)


for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  if (ds == "Pelka") {
    labels <- as.character(labels)
    labels[labels == "N"] <- "Normal"
    labels[labels == "T"] <- "Tumor"
  }
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]],
    cluster_score = FALSE,
    mod_score = FALSE,
    anosim_score = FALSE
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect") +
  plot_annotation(title = ds)
```

```{r fig.height=12, fig.width=12}
for (m in names(methods)) {
  feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = feat_mat,
    labels = labels,
    n_ct_show = 5,
    invisible = "quali",
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 3, guides = "collect")

```

### Show example heatmap

```{r fig.height=7, fig.width=7}
library(pheatmap)

ds <- "Bassez"

feat_mat <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$feat_mat
labels <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(labels)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pdf(file = file.path(path_plots, paste0("Heatmap_", ds, ".pdf")),
    width = 6,
    height = 6)
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4
)
dev.off()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4
)

plot_pca(
    feat_mat = feat_mat,
    labels = labels,
    n_ct_show = 5,
    invisible = "quali",
    coord_equal = FALSE,
    title = "",
    repel = TRUE
)
ggsave(file.path(path_plots, paste0("PCA_arrows_", ds, ".pdf")),
       width = 6,
       height = 6)
```

### Figure 3 A: ECODA annotation method comparison
### and Supp fig 17

```{r fig.height=6, fig.width=5}
pointsize <- 1.5

filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_LR",
  "ECODA_deconv",
  "ECODA_HiTME_HR_layer2",
  # "ECODA_HiTME_HR_layer3",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_0.1",
  "ECODA_seuratres_0.4",
  "ECODA_seuratres_2",
  "ECODA_seuratres_5",
  "ECODA_seuratres_20"
)


df_plot <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(!grepl("Zhang|Lee", dataset)) %>% # Exclude Zhang and Lee, as they do not have high-resolution author annotation
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_seuratres_0.1" = "ECODA_Leiden_res_0.1",
                         "ECODA_seuratres_0.4" = "ECODA_Leiden_res_0.4",
                         "ECODA_seuratres_2" = "ECODA_Leiden_res_2",
                         "ECODA_seuratres_5" = "ECODA_Leiden_res_5",
                         "ECODA_seuratres_20" = "ECODA_Leiden_res_20",
                         "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
                         "ECODA_scATOMIC_HR" = "ECODA_scATOMIC"
  ))
  
p <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "ECODA_authors_HR",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 1
  ) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.15)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p)
ggsave(file.path(path_plots, "Figure_3_A_annotationmethods_barplot_anosim.pdf"),
       width = 5,
       height = 5)



df_plot <- df_results %>%
  filter(score == "mod_knn3_score") %>%
  filter(!grepl("Zhang|Lee", dataset)) %>% # Exclude Zhang and Lee, as they do not have high-resolution author annotation
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_seuratres_0.1" = "ECODA_Leiden_res_0.1",
                         "ECODA_seuratres_0.4" = "ECODA_Leiden_res_0.4",
                         "ECODA_seuratres_2" = "ECODA_Leiden_res_2",
                         "ECODA_seuratres_5" = "ECODA_Leiden_res_5",
                         "ECODA_seuratres_20" = "ECODA_Leiden_res_20",
                         "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
                         "ECODA_scATOMIC_HR" = "ECODA_scATOMIC"
  ))
  
p_mod <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "ECODA_authors_HR",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 1.1
  ) +
  labs(x = "", y = "Separation score (Modularity)") +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.15)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p_mod)
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity_annotationmethods_refECODAhr.pdf"),
       width = 2.5,
       height = 4)


df_plot <- df_results %>%
  filter(score == "cluster_score") %>%
  filter(!grepl("Zhang|Lee", dataset)) %>% # Exclude Zhang and Lee, as they do not have high-resolution author annotation
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_seuratres_0.1" = "ECODA_Leiden_res_0.1",
                         "ECODA_seuratres_0.4" = "ECODA_Leiden_res_0.4",
                         "ECODA_seuratres_2" = "ECODA_Leiden_res_2",
                         "ECODA_seuratres_5" = "ECODA_Leiden_res_5",
                         "ECODA_seuratres_20" = "ECODA_Leiden_res_20",
                         "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
                         "ECODA_scATOMIC_HR" = "ECODA_scATOMIC"
  ))
  
p_ari <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "ECODA_authors_HR",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 1.1
  ) +
  labs(x = "", y = "Separation score (ARI)") +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.15)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p_ari)
ggsave(file.path(path_plots, "benchmark_summary_barplot_ari_annotationmethods_refECODAhr.pdf"),
       width = 2.5,
       height = 4)


wrap_plots(list(p_mod, p_ari)) +
  plot_layout(nrow = 1, guides = "collect")
ggsave(file.path(path_plots, "Supp_fig_17_annotationmethods_barplot_mod_ari.pdf"),
       width = 9,
       height = 5)
```

#### Figure 3 B: Plot PCA for selected annot. methods (Adams)

```{r fig.height=8, fig.width=8}
ds <- "Adams"

methods <- list(
  "ECODA_authors_HR" = "ECODA_authors_HR",
  "ECODA_seuratres_5" = "ECODA_Leiden_res_5",
  "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
  "ECODA_authors_LR" = "ECODA_authors_LR"
  # "ECODA_scATOMIC_HR" = "ECODA_scATOMIC",
  # "ECODA_deconv" = "ECODA_deconv"
  # "ECODA_seuratres_0.1" = "ECODA_Leiden_res_0.1"
)

plot_list <- list()

for (m in names(methods)) {
  feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  plot_list[[m]] <- plot_pca(
    feat_mat = feat_mat,
    labels = labels,
    cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
    coord_equal = TRUE, pointsize = 2,
    n_ct_show = 5, invisible = "quali", repel = TRUE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect")

ggsave(file.path(path_plots, "Figure_3_B_annotationmethods_pca_adams.pdf"),
       width = 8,
       height = 7)
```



#### Supp fig 18: Plot PCA for selected annot. methods (GongSharma)

```{r fig.height=10, fig.width=12}
datas <- c("Bassez", "Lee", "Stephenson")

methods <- list(
  "ECODA_authors_HR" = "ECODA_authors_HR",
  # "ECODA_authors_LR" = "ECODA_authors_LR",
  "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
  "ECODA_scATOMIC_HR" = "ECODA_scATOMIC",
  # "ECODA_deconv" = "ECODA_deconv",
  "ECODA_seuratres_2" = "ECODA_Leiden_res_2"
  # "ECODA_seuratres_0.1" = "ECODA_Leiden_res_0.1"
)

plot_list <- list()

for (ds in datas) {
  pl <- list()
  for (m in names(methods)) {
    if (ds == "Lee" & m == "ECODA_authors_HR") next
    
    dist_mat <- result_list[["bmark"]][[ds]][[m]]$dist_mat
    labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
    
    if (ds == "Bassez") {
      labels <- as.character(labels)
      labels[labels == "E"] <- "Responders"
      labels[labels == "NE"] <- "Non-responders"
      labels <- factor(labels, levels = c("Responders", "Non-responders"))
    }
    
    pl[[m]] <- plot_mds(
      dist_mat = dist_mat,
      labels = labels,
      cluster_score = FALSE, mod_score = FALSE, anosim_score = TRUE,
      coord_equal = TRUE,
      title = methods[[m]]
    )
  }
  
  # Wrap the row patchwork in wrap_elements to preserve the annotation
  row_plot <- wrap_plots(pl, nrow = 1, guides = "collect") + 
    plot_annotation(
      title = ds, 
      theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
    )
  
  plot_list[[ds]] <- wrap_elements(row_plot)
}

# Combine the rows vertically
wrap_plots(plot_list, ncol = 1)


ggsave(file.path(path_plots, paste0("Supp_fig_18_Bassez_Lee_Stephenson.pdf")),
       width = 12,
       height = 10)
```


### Supp fig 22: Method parameter screening

```{r fig.height=8, fig.width=14}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_LR",
  "ECODA_deconv",
  "ECODA_HiTME_HR_layer2",
  "ECODA_HiTME_HR_layer3",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_0.4",
  "ECODA_seuratres_2",
  "ECODA_seuratres_5",
  "ECODA_seuratres_20"
)

default_methods <- c(
    "ECODA_authors_HR\\b", 
    "MOFA_hvg2000_factors15", 
    "Pseudobulk_hvg2000\\b",
    "PILOT_hvg2000_HR\\b",
    "scITD_hvg2000_factors5",
    "scPoli_hvg2000_dims15_HR\\b",
    "GloScope_hvg2000_pcadims30",
    "GloProp",
    "MrVI_hvg2000"
)
default_methods_regex <- paste(default_methods, collapse = "|")

score.labs <- c("ANOSIM score", "Adjusted rand index", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

df_results %>%
  # filter(score == "anosim_score") %>%
  filter(!grepl("HiTME|scATOMIC|seuratres|ECODA_authors_HR_NULL|ECODA_authors_LR|ECODA_deconv|Freq_highres|varexp|varcts|Avg_PCA|combo", method),
         !grepl("GloScope", method) | grepl("sqrtmat", method)
  ) %>%
  filter(
    if_else(grepl("scPoli", method), grepl("highres|lowres", method), TRUE)
  ) %>%
  # filter(method %in% filter_methods) %>%
  # mutate(method = recode(method,
  #                        "MOFA_2_factors" = "MOFA",
  #                        "scITD_2_factors" = "scITD",
  #                        "ECODA_highres" = "ECODA")) %>%
  
  # CREATE THE FACETING GROUP COLUMN
  mutate(
    method_plot = case_when(
      grepl("^ECODA_", method) ~ "ECODA",
      grepl("^MOFA_", method) ~ "MOFA",
      grepl("^MrVI_", method) ~ "MrVI",
      grepl("^scITD_", method) ~ "scITD",
      grepl("^GloScope_", method) ~ "GloScope",
      grepl("^GloProp", method) ~ "GloScope",
      grepl("^PILOT_", method) ~ "PILOT",
      grepl("^scPoli_", method) ~ "scPoli",
      grepl("^Pseudobulk_", method) ~ "Pseudobulk",
      TRUE ~ method # Keep any other simple names as they are
    ),
    
    method_plot = factor(method_plot, 
                         levels = c("ECODA", "Pseudobulk", "MrVI", "GloScope", "GloProp",
                                    "scPoli", "MOFA", "PILOT", "scITD"))
  ) %>%
  
  mutate(method = method %>%
           # Remove _sqrtmat and swap highres/lowres
           stringr::str_replace_all(c("_sqrtmat" = "", 
                                      "highres"  = "HR", 
                                      "lowres"   = "LR")) %>%
           # Transform Pseudobulk_x_PCA_dims -> Pseudobulk_hvg2000_pcadimsx
           stringr::str_replace("Pseudobulk_(\\d+)_PCA_dims", "Pseudobulk_hvg2000_pcadims\\1")
  ) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # Highlight default methods
  scale_x_discrete(labels = function(x) highlight(
    x,
    default_methods_regex,
    "black")
  ) +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             cols = vars(method_plot), # Use the new grouping column for column faceting
             labeller = labeller(score = score.labs))


ggsave(file.path(path_plots, "Supp_fig_22_methodparameter_screening.pdf"),
       width = 14,
       height = 8)
```

### Combined ECODA and PB

#### Plot distance distribution after different normalization methods applied

```{r}
ecoda <- result_list$bmark$Pelka$ECODA_authors_HR$dist_mat
pb <- result_list$bmark$Pelka$Pseudobulk_hvg2000$dist_mat

hist(ecoda)
hist(pb)

test1 <- ecoda / max(ecoda)
test2 <- pb / max(pb)
hist(test1, main = "ecoda max")
hist(test2, main = "pb max")

test1 <- ecoda / median(ecoda)
test2 <- pb / median(pb)
hist(test1, main = "ecoda median")
hist(test2, main = "pb median")

test1 <- global_quantile_norm_gaussian(ecoda)
test2 <- global_quantile_norm_gaussian(pb)
hist(test1, main = "ecoda quantiles")
hist(test2, main = "pb quantiles")

test1 <- zscore_transform(ecoda)
test2 <- zscore_transform(pb)
hist(test1, main = "ecoda zscore")
hist(test2, main = "pb zscore")
```

#### Plot benchmark including ECODA+PB (overview)

```{r fig.height=9, fig.width=7}
norm_methods <- c("max", "median", "zscore", "quantile")

filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  
  paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  "ECODA_PB_combo_normcosine_ecodaweight0.5",
  
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)


df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 80, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             labeller = labeller(score = score.labs))
```

#### Plot benchmark including ECODA+PB (detailed)

```{r fig.height=7, fig.width=6}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)


df_plot <- df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope"))


df_plot %>%
  filter(score == "anosim_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 80, vjust = 1, hjust = 1))



df_plot %>%
  filter(score == "cluster_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ARI score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))




df_plot %>%
  filter(score == "mod_knn3_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "Modularity score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
```

### Poster figures

```{r fig.height=10, fig.width=14}
# Poster figure (highlighted)

filter_methods <- c(
  "Pseudobulk",
  # "Freq_highres",
  # "ECODA_deconv",
  # "ECODA_lowres",
  "ECODA_highres",
  # "ECODA_high_res_2_PCA_dims",
  # "ECODA_highres_top_varexp0.7",
  # "ECODA_seuratres_0.4",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI",
  "scPoli",
  "PILOT",
  "GloScope"
)

p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "MOFA_2_factors" = "MOFA",
                         "scITD_2_factors" = "scITD",
                         "ECODA_highres" = "ECODA")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "ECODA", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "ECODA",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "ECODA", "black"))
print(p)
ggsave(file.path(path_plots, "POSTER_benchmark_summary_barplot_anosim_mainmethods_refECODA.pdf"),
       width = 3.5,
       height = 4)



filter_methods <- c(
  "Pseudobulk",
  "ECODA_highres",
  "ECODA_highres_top_varexp0.1"
)


p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_highres_top_varexp0" = "2 most highly variable cell types",
                         "ECODA_highres" = "ECODA")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "ECODA", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "ECODA",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "ECODA", "black"))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAvs2HVCvsBULK_refECODA.pdf"),
       width = 2,
       height = 6)




filter_methods <- c(
  "ECODA_deconv",
  "ECODA_lowres",
  "ECODA_highres",
  "ECODA_highres_HiTME",
  "ECODA_highres_HiTME_layer3",
  "ECODA_highres_scATOMIC",
  "ECODA_seuratres_0.4"
)

# my_comparisons <- list(
#   c("Low res. annot.", "High res. annot.")
#   )

p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_deconv" = "Deconvoluted",
                         "ECODA_lowres" = "Low res. author annot.",
                         "ECODA_highres" = "High res. author annot.",
                         "ECODA_highres_HiTME" = "HiTME annot.",
                         "ECODA_highres_HiTME_layer3" = "HiTME annot. w. gene. progs.",
                         "ECODA_highres_scATOMIC" = "scATOMIC annot.",
                         "ECODA_seuratres_0.4" = "Unsup. Leiden clust.")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "High res. author annot.", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.5) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "High res. author annot.",
    method = "wilcox.test",
    paired = TRUE,
    label.y = 1,
    tip.length = 0,
    aes(label = after_stat(p.signif))
  ) +
  ylim(c(-0.03, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1)) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "High res. author annot.", "black"))
print(p)
ggsave(file.path(path_plots, "POSTER_benchmark_summary_barplot_anosim_ecodamethods_refECODA_highres.pdf"),
       width = 3,
       height = 5)
```

### Other figures (not shown)

```{r}
ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores.pdf"),
       width = 7,
       height = 7)



df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

df_results_clustering <- df_results_clustering[!grepl("2|10|PCA", df_results_clustering$method), ]

for (ds in unique(df_results_clustering$dataset)) {
  number_cts <- dim(result_list$bmark[[ds]]$ECODA_highres$feat_mat)[2]
  df_results_clustering[df_results_clustering$dataset == ds, "number_celltypes"] <- number_cts
}

# df_results_clustering$dataset <- paste0(df_results_clustering$dataset, " (", df_results_clustering$number_celltypes, ")")
# legend_title <- "Dataset (number of cell types)"
legend_title <- "Dataset"

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_clustering)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)


my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "Pseudobulk"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(3, 6, 4, 5, 2, 1)]


p <- ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(group = legend_title, linetype = legend_title, shape = legend_title) +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_clustering.pdf"),
       width = 7,
       height = 5)



df_results_mod <- df_results %>%
  filter(score == "mod_score")

p <- ggplot(df_results_mod, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
  ) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_mod$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Modularity score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity.pdf"),
       width = 7,
       height = 6)
```

## Figure 4 AB: Plot feature selection

Surprisingly, even at the lowest number highly variable cell types (i.e. variance explained of 10% -\> \>=2 cell types) the average biological separation is quite very high. Different daatasets can do with a different number of highly variable cell types (separation plateaus out at different levels of variance explained % and number of cell types retained).

```{r}
df_results_sub <- df_results %>%
  # filter(score == "anosim_score") %>%
  filter(grepl("ECODA_authors_HR", method) & !grepl("HiTME|scATOMIC|PCA|NULL|most|least", method)) %>%
  filter(!grepl("cmv|sex", dataset)) %>%
  filter(!method == "ECODA_authors_HR_top_varexp0") %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "100",
                         "ECODA_authors_HR_top_varexp0" = "2most_varcts")) %>%
  mutate(method = gsub("ECODA_authors_HR", "", method)) %>%
  mutate(method = if_else(
    grepl("^_top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("_top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  # mutate(method = gsub("top_varexp0", "2most_varcts", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = rev(mixedsort(levels(method))))) %>%
  mutate(dataset = recode(dataset,
                          "GongSharma_all_age" = "GongSharma"))

df_results_anosim <- df_results_sub %>% filter(score == "anosim_score")

df_hvc40 <- df_results_anosim[df_results_anosim$method == 40 ,]
perc_tot <- c()
for (ds in df_hvc40$dataset) {
  tot <- df_results_anosim$n_features[df_results_anosim$dataset == ds & df_results_anosim$method == 100]
  perc_tot <- c(perc_tot, round(df_hvc40$n_features[df_hvc40$dataset == ds] / tot * 100))
}

cts <- df_hvc40$n_features
print(paste0("The top ", min(cts), "-", max(cts), " cell types ", 
             "(~", min(perc_tot), "-", max(perc_tot), "% of the total) ",
             "explain 40% of the datasets' variances."))

unique(df_results_sub$method)

pointsize <- 1.5

p1 <- df_results_sub %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = method, y = n_features)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "Variance explained (%)", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p1)

p2 <- df_results_sub %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(-0.1, 1)) +
  labs(x = "Variance explained (%)", y = "Separation score (ANOSIM)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p2)


# Combine the plots vertically
combined_plot <- (p1 / p2) + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') & 
  theme(
    plot.tag = element_text(size = 16, face = "bold"),
    legend.position = "right"
  ) &
  labs(color = "Dataset", 
       shape = "Dataset", 
       linetype = "Dataset")
print(combined_plot)
ggsave(file.path(path_plots, "Figure_4_AB_featureselectionHVC_barplot_anosim.pdf"),
       width = 5.5,
       height = 6)
```

#### Supp fig 19: HVC Mod + ARI

```{r fig.height=3.5, fig.width=8}
p3 <- df_results_sub %>%
  filter(score == "mod_knn3_score") %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(-0.1, 1)) +
  labs(x = "Variance explained (%)", y = "Separation score (Modularity)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p3)

p4 <- df_results_sub %>%
  filter(score == "cluster_score") %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(-0.1, 1)) +
  labs(x = "Variance explained (%)", y = "Separation score (ARI)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p4)


# Combine the plots vertically
combined_plot <- (p3 + p4) + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') & 
  theme(
    plot.tag = element_text(size = 16, face = "bold"),
    legend.position = "right"
  ) &
  labs(color = "Dataset", 
       shape = "Dataset", 
       linetype = "Dataset")
print(combined_plot)
ggsave(file.path(path_plots, "Supp_fig_19_featureselectionHVC_barplot_mod_ari.pdf"),
       width = 8,
       height = 3.5)
```


### Show example PCAs

#### NOTE: CLR of two parts vs simple log-ratio

The CLR of the first part is exactly half of the log-ratio of the two parts.

```{r}
test <- result_list$bmark[[ds]]$Freq_highres$feat_mat[, colnames(feat_mat_recalc)]
test_clr <- clr(test)
test_lr <- log(test$`Immature B cells` / test$TIM)
```

```{r}
test_clr[, 1]
```

```{r}
test_lr
```

```{r}
test_clr[, 1] * 2 - test_lr
```

#### Figure 4 C: Adams

```{r fig.height=3, fig.width=7}
ds <- "Adams"

feat_mat_recalc <- result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0$feat_mat
feat_mat <- result_list$bmark[[ds]]$ECODA_authors_HR$feat_mat[, colnames(feat_mat_recalc)]

labels <- result_list$bmark[[ds]]$ECODA_authors_HR$labels
df_logr <- feat_mat_recalc[, 1, drop = FALSE] * 2

df_logr$Groups <- labels
avg_sep_score <- calc_avg_sep_score(dist(feat_mat), labels = labels, digits = 3)

p1 <- plot_pca(
  feat_mat = feat_mat,
  labels = labels,
  cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
  coord_equal = TRUE,
  title = "Adams",
  n_ct_show = 4, invisible = "quali", repel = TRUE, pointsize = 2
)

p2 <- ggplot(df_logr, aes(x = Groups, y = !!sym(colnames(df_logr)[1]), fill = Groups)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Standard boxplot
  geom_jitter(width = 0.2, size = 1, alpha = 0.5) + # Optional: adds points to see distribution
  theme_classic() +
  # theme_cleveland() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate labels if group names are long
  labs(
    x = NULL,
    y = paste0(
      "Log ratio\n",
      colnames(feat_mat_recalc)[1], " / ", colnames(feat_mat_recalc)[2]
    )
  ) &
  NoLegend()


wrap_plots(list(p1, p2)) +
  plot_layout(nrow = 1, widths = c(2.5, 1))
ggsave(file.path(path_plots, "Figure_4_C_featureselectionHVC_PCA_adams.pdf"),
       width = 7,
       height = 3)
```

#### Figure 4 D: Kfoury

```{r fig.height=3, fig.width=7}
ds <- "Kfoury"

feat_mat_recalc <- result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0$feat_mat
feat_mat <- result_list$bmark[[ds]]$ECODA_authors_HR$feat_mat[, colnames(feat_mat_recalc)]

labels <- result_list$bmark[[ds]]$ECODA_authors_HR$labels
df_logr <- feat_mat_recalc[, 1, drop = FALSE] * 2
df_logr$Groups <- labels
avg_sep_score <- calc_avg_sep_score(dist(feat_mat), labels = labels, digits = 3)

p1 <- plot_pca(
  feat_mat = feat_mat,
  labels = labels,
  cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
  coord_equal = TRUE,
  title = "Kfoury",
  n_ct_show = 4, invisible = "quali", repel = TRUE, pointsize = 2
)

p2 <- ggplot(df_logr, aes(x = Groups, y = !!sym(colnames(df_logr)[1]), fill = Groups)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Standard boxplot
  geom_jitter(width = 0.2, size = 1, alpha = 0.5) + # Optional: adds points to see distribution
  theme_classic() +
  # theme_cleveland() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate labels if group names are long
  labs(
    x = NULL,
    y = paste0(
      "Log ratio\n",
      colnames(feat_mat_recalc)[1], " / ", colnames(feat_mat_recalc)[2]
    )
  ) &
  NoLegend()

wrap_plots(list(p1, p2)) +
  plot_layout(nrow = 1, widths = c(2.5, 1))
ggsave(file.path(path_plots, "Figure_4_D_featureselectionHVC_PCA_kfoury.pdf"),
       width = 7,
       height = 3)
```

## Separation based on least variable cell types (sanity check)

```{r}
# Create an empty list to store the results from each dataset
temp_list <- list()

for (ds in names(result_list[["bmark"]])) {
  feat_mat <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$feat_mat
  labels <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels
  ncts <- 2
  
  # Calculate separation score for most highly variable cell types (HVC)
  top_hvcs <- get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
    get_hvcs(top_n_hvcs = ncts, variance_threshold = NULL)
  score_most_hvc <- calc_sep_score(feat_mat[, top_hvcs], labels)$anosim_score
  
  # Calculate separation score for least highly variable cell types (HVC)
  least_hvcs <- get_ct_var(feat_mat, show_plot = FALSE, descending = FALSE) %>%
    get_hvcs(top_n_hvcs = ncts, variance_threshold = NULL)
  score_least_hvc <- calc_sep_score(feat_mat[, least_hvcs], labels)$anosim_score
  
  # Store the results in a temporary data frame
  temp_df <- data.frame(
    dataset = ds,
    anosim_most_hvc = score_most_hvc,
    anosim_least_hvc = score_least_hvc
  )
  
  # Add the temporary data frame to the results list
  temp_list[[ds]] <- temp_df
}

# Combine all data frames in the list into a single data frame
df_plot <- bind_rows(temp_list)



# Reshape the data from wide to long format for ggplot
df_plot_long <- df_plot %>%
  pivot_longer(
    cols = starts_with("anosim_"),
    names_to = "group",
    values_to = "anosim_score"
  ) %>%
  # Clean up the group names for better readability
  mutate(group = gsub("anosim_|_hvc", "", group))

# Create the grouped bar plot
ggplot(df_plot_long, aes(x = dataset, y = anosim_score, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "ANOSIM Scores for Highly vs. Least Variable Cell Types",
    x = "Dataset",
    y = "ANOSIM Score",
    fill = "Cell Type Group" # Legend title
  ) +
  scale_fill_manual(values = c("most" = "darkred", "least" = "steelblue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Supp fig 20

## Most highly variable cell types per dataset

```{r fig.height=15, fig.width=15}
# devtools::install_github("carmonalab/scECODA", force = TRUE)
library(scECODA)
plot_list <- list()

dss <- names(result_list[["bmark"]])
# dss <- dss[!grepl("cmv|sex", dss)]
# dss[dss == "GongSharma_all_age"] <- "GongSharma"


for (ds in dss) {
  counts <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$counts
  meta <- data.frame(sample_id = names(result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels),
                     group = result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels
  )
  ecoda_object <- ecoda(counts, metadata = meta, variance_explained = 0.4)
  
  if (ds == "GongSharma_all_age") {
    ec_plot <- ecoda_object
    pt <- "GongSharma"
  } else pt <- ds
  
  p <- plot_varmean(ecoda_object, plot_title = pt)
  # print(p)
  plot_list <- c(plot_list, p)
  # p <- plot_corr(ecoda_object)
  # print(p)
}

wrap_plots(plot_list) +
  plot_layout(nrow = 4, guides = "collect")

ggsave(file.path(path_plots, "Supp_fig_20_HVCs_per_dataset.pdf"),
       width = 20,
       height = 20)
```


```{r fig.height=5, fig.width=6}
plot_pca_contributions_horizontal(prcomp(ec_plot@clr), pcs = c("PC1", "PC2"), n = 10)
```



```{r fig.height=8, fig.width=12}
print("Most highly variable cell types per dataset:")
for (ds in names(result_list[["bmark"]])) {
  # print(ds)
  # print(paste("ANOSIM score using all cell types:", round(result_list[["bmark"]][[ds]]$ECODA_authors_HR$scores$anosim_score, 3)))
  # print(paste("ANOSIM score at 10% variance explained:", round(result_list[["bmark"]][[ds]]$ECODA_authors_HR_top_varexp0.1$scores$anosim_score, 3)))
  # print("Most highly variable cell types (authors annotation):\n")
  # print(paste(colnames(result_list[["bmark"]][[ds]]$ECODA_authors_HR_top_varexp0$feat_mat)))
  # print("Most highly variable cell types (HiTME annotation):\n")
  # print(paste(colnames(result_list[["bmark"]][[ds]]$ECODA_HiTME_HR_layer3_top_varexp0$feat_mat)))
  
  p <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$feat_mat %>%
    get_ct_var() %>%
    varmeanplot(title = ds, label_points = TRUE)
  print(p)
  
  # p <- plot_pca(
  #   result_list[["bmark"]][[ds]]$ECODA_authors_HR_2least_varcts$feat_mat,
  #   labels = result_list[["bmark"]][[ds]]$ECODA_authors_HR_2least_varcts$labels,
  #   title = paste0(ds, ": Least variable cts"),
  #   invisible = "quali", labelsize = 4)
  # print(p)
  # p <- plot_pca(
  #   result_list[["bmark"]][[ds]]$ECODA_authors_HR_top_varexp0$feat_mat,
  #   labels = result_list[["bmark"]][[ds]]$ECODA_authors_HR_top_varexp0$labels,
  #   title = paste0(ds, ": Most highly variable cts "),
  #   invisible = "quali", labelsize = 4)
  # print(p)
}
```

```{r fig.height=8, fig.width=12}
print("Most highly variable cell types per dataset:")
for (ds in names(result_list[["bmark"]])) {
  print(ds)
  print("Most highly variable cell types (HiTME annotation):\n")
  print(paste(colnames(result_list[["bmark"]][[ds]]$ECODA_HiTME_HR_layer2_top_varexp0.5$feat_mat)))
  print(paste(colnames(result_list[["bmark"]][[ds]]$ECODA_HiTME_HR_layer3_top_varexp0.5$feat_mat)))
}
```

### Check IFN expression in Stephenson dataset

```{r}
ds <- "Stephenson"
seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))

plot_data <- data.frame(
  expression = seurat$HeatShock_UCell,
  status = seurat$Status
)

ggplot(plot_data, aes(x = expression, fill = status)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Expression Density by Status",
    x = "IFN_UCell Expression",
    y = "Density",
    fill = "Status"
  ) +
  xlim(c(0, 0.2)) +
  theme_minimal() +
  scale_fill_manual(values = c("Healthy" = "steelblue", "Covid" = "darkred"))
```

```{r}
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, "Status")
test <- process_coda_fig(seurat, labels, ct_col = "layer2", title = "ECODA\nHiTME layer2")
test

varexp_hvc <- 0.7
test1 <- process_coda_fig(seurat, labels, ECODA_top_varexp_hvct = varexp_hvc, ct_col = "layer2", title = paste0("ECODA layer2\nhigh res. var. exp. ", varexp_hvc*100, "%"))
test1


df_counts <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", "layer2")

feat_mat <- df_counts %>%
  impute_zeros(clr_zero_impute_method = clr_zero_impute_method, clr_zero_impute_num = clr_zero_impute_num) %>%
  calc_perc_df()

feat_mat <- feat_mat %>%
  clr()

get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
        get_hvcs(top_n_hvcs = 10, variance_threshold = NULL)
```

### Check different UCell score thresholds for HiTME layer3

```{r}
seurat <- replace_HiTMElayer3_annot(seurat)
```

```{r}
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, "Status")
test2 <- process_coda_fig(seurat, labels, ct_col = "layer3", title = "ECODA\nHiTME layer3")
test2

varexp_hvc <- 0.7
test3 <- process_coda_fig(seurat, labels, ECODA_top_varexp_hvct = varexp_hvc, ct_col = "layer3", title = paste0("ECODA layer3\nhigh res. var. exp. ", varexp_hvc*100, "%"))
test3


df_counts <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", "layer3")

feat_mat <- df_counts %>%
  impute_zeros(clr_zero_impute_method = clr_zero_impute_method, clr_zero_impute_num = clr_zero_impute_num) %>%
  calc_perc_df()

feat_mat <- feat_mat %>%
  clr()

get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
        get_hvcs(top_n_hvcs = 10, variance_threshold = NULL)
```

## Reduced separation with fewer cell types?

Why do some datasets show reduced separability when using fewer highly variable cell types while other still show good separation even with just 2 cell types? Are there any comonalities between these datasets?

In general, datasets with HIGH ANOSIM score (\~ \>0.5) when using all cell types also show still good separability with very low number of cell types. -\> Maybe these biological conditions are more distinct and are easier to separate? Datasets with LOW ANOSIM score (\~ \<0.5) when using all cell types also show decreasing separability when reducing the number of cell types. -\> The minimum number of cell types for maximal separability depends on the dataset

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_hi", method) & !grepl("HiTME|scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = if_else(
    grepl("^top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

ds_diffs <- df_results_sub %>%
  filter(method %in% c("10", "70")) %>%
  group_by(dataset) %>%
  summarize(
    diff = value[method == "10"] - value[method == "70"]
  ) %>%
  arrange(diff)

ggplot(ds_diffs, aes(x = reorder(dataset, diff), y = diff)) +
  geom_bar(stat = "identity") +
  labs(x = "Dataset", y = "Difference") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Which datasets have lower separation when using less cell types?\nVar. exp. 10% - Var. exp. 70%")
```

## Which datasets HiTME annotation underperforms?

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("^ECODA_highres$|^ECODA_highres_HiTME$", method) & !grepl("scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = gsub("ECODA_highres", "Author annotation", method)) %>%
  # mutate(method = if_else(
  #   grepl("^top_varexp0\\.\\d", method),
  #   paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
  #   method
  # )) %>%
  # mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

ds_diffs <- df_results_sub %>%
  group_by(dataset) %>%
  summarize(
    diff = value[method == "HiTME"] - value[method == "Author annotation"]
  ) %>%
  arrange(diff)

ggplot(ds_diffs, aes(x = reorder(dataset, diff), y = diff)) +
  geom_bar(stat = "identity") +
  labs(x = "Dataset", y = "Difference") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  ggtitle("For each dataset, how does ECODA perform based on\nHiTME annotation vs Author annotation?")
```

## Plot feature selection (HiTME layer3)

```{r fig.height=8, fig.width=12}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_hi", method) & !grepl("scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  # mutate(method = if_else(
  #   grepl("^top_varexp0\\.\\d", method),
  #   paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
  #   method
  # )) %>%
  # mutate(method = gsub("ECODA_highres", "100", method)) %>%
  # mutate(method = if_else(
  #   grepl("HiTME_layer3_top_varexp0\\.\\d", method),
  #   paste0("HiTME_layer3_", as.integer(gsub("HiTME_layer3_top_varexp0\\.", "", method)) * 10),
  #   method
  # )) %>%
  # mutate(method = gsub("^HiTME_layer3$", "HiTME_layer3_100", method)) %>%
  mutate(method = gsub("^HiTME$", "HiTME_layer2", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method)))) %>%
  # mutate(method = fct_relevel(method, "HiTME_layer2_100", after = Inf)) %>%
  mutate(method = fct_relevel(method, "HiTME_layer3", after = 5)) #%>%
  # filter(!grepl("varexp0|varcts", method))

p1 <- df_results_sub %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(0, 1)) +
  labs(x = "Variance explained (%)", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p1)
# ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAfeatureselectionHVC.pdf"),
#        width = 4,
#        height = 4)

p2 <- df_results_sub %>%
  ggplot(aes(x = method, y = n_features)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  labs(x = "Variance explained (%)", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p2)

# Create p1 (the top plot) without x-axis labels
p1_clean <- p1 + theme(axis.title.x = element_blank(), 
                       axis.text.x = element_blank(),
                       axis.ticks.x = element_blank())

# Combine the plots vertically
combined_plot <- p2 / p1
print(combined_plot)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAandHiTMEfeatureselectionHVC_withNumCT.pdf"),
       width = 5,
       height = 8)
```

## Show example PCAs for publication

```{r fig.height=7, fig.width=7}
for (ds in names(result_list[["bmark"]])) {
  p <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres"]][["plot"]][["data"]][["Groups"]],
    n_ct_show = 5,
    invisible = "quali",
    labelsize = 5
  ) + ggtitle(ds)
  print(p)
}

ds <- "Stephenson"
plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.9"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.9"]][["plot"]][["data"]][["Groups"]],
    n_ct_show = 50,
    invisible = "quali",
    labelsize = 3
) + ggtitle(ds)

```

```{r fig.height=6, fig.width=12}

levs <- c(
  "ECODA_high_res",
  "ECODA_low_res",
  "ECODA_deconv",
  "Pseudobulk",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI"
)


plot_list <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_cmv_females.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_age_females_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Lee"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Lee"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Lee.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Zhang"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Zhang"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Zhang.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Stephenson.pdf"),
       plot = p,
       width = 15,
       height = 7)

```

## Example PCAs for HiTME layer2 vs layer3

```{r}
plot_pca(result_list[["bmark"]][["Adams"]][["ECODA_highres_HiTME_layer3_top_varexp0.1"]][["feat_mat"]],
         labels = result_list[["bmark"]][["Adams"]][["ECODA_highres_HiTME_layer3_top_varexp0.1"]][["plot"]][["data"]][["Groups"]],
         invisible = "quali",
         labelsize = 4
         )
```

```{r}
test <- plist_layer2[[1]]
```

```{r}
plist_layer2 <- list()
for (ds in names(result_list$bmark)) {
  plist_layer2[[ds]] <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2_top_varexp0.5"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2_top_varexp0.5"]]$labels,
    invisible = "quali",
    labelsize = 4,
    title = ds,
    n_ct_show = 5
  )
}

plist_layer3 <- list()
for (ds in names(result_list$bmark)) {
  plist_layer3[[ds]] <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.5"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.5"]]$labels,
    invisible = "quali",
    labelsize = 4,
    title = ds,
    n_ct_show = 5
  )
}
```

```{r}
ggsave("pca_layer2.pdf", cowplot::plot_grid(plotlist = plist_layer2, ncol = 3), width = 20, height = 20)
ggsave("pca_layer3.pdf", cowplot::plot_grid(plotlist = plist_layer3, ncol = 3), width = 20, height = 20)
```

```{r fig.height=8, fig.width=16}
plist_vcts <- list()
for (ds in names(result_list$bmark)) {
  p1 <- result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2"]][["feat_mat"]]%>%
    get_ct_var() %>%
    varmeanplot(title = paste0(ds, " Layer2"), label_points = TRUE)
  p2 <- result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3"]][["feat_mat"]]%>%
    get_ct_var() %>%
    varmeanplot(title = paste0(ds, " Layer3"), label_points = TRUE)
  plist_vcts[[ds]] <- p1 | p2
  print(p1 | p2)
}


```

```{r}
ggsave(filename = "test.pdf", plot = cowplot::plot_grid(plotlist = plist_vcts, ncol = 1), width = 20, height = 100, limitsize = FALSE)
```

## Compare layer2 and layer3 by scores

```{r fig.height=8, fig.width=2.5}
filter_methods <- c(
  "ECODA_highres",
  "ECODA_highres_HiTME_layer2_top_varexp0.5",
  "ECODA_highres_HiTME_layer3_top_varexp0.5"
)

for (s in c("anosim_score", "mod_score", "cluster_score")) {
  p <- df_results %>%
    filter(score == s) %>%
    filter(method %in% filter_methods) %>%
    
    ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
    stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
    stat_summary(fun = mean, geom = "bar", fill = "gray70") +
    geom_point(position = "identity", alpha = 0.6) +
    geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
    labs(x = "", y = s) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
    ggtitle(s)
  print(p)
}
```

## Overlap between HVGs from single-cell and bulk

```{r}
# test <- c()
test_int <- c()
test_anosim <- c()
for (ds in names(datasets)) {
  # test <- c(test, result_list[[ds]][["Pseudobulk_unsup_hvg2000"]][["hvgs_sc_pb_intersect"]])
  test_int <- c(test_int, result_list[["bmark"]][[ds]][["Pseudobulk_unsup_hvg2000"]]$hvgs_sc_pb_intersect)
  test_anosim <- c(test_anosim, result_list[["bmark"]][[ds]][["Pseudobulk_unsup_hvg2000"]]$scores$anosim_score)
}
# plot(test)
plot(test_int, test_anosim)
```

# ---------------------------------------------------------------

# ---------------------------------------------------------------

# Plot execution times

## Get execution times from R

```{r}
for (ds in names(result_list$bmark)) {
  result_list[["bmark"]][[ds]][["GloScope_hvg2000_pcadims30_sqrtmat"]]$exec_time <- result_list[["bmark"]][[ds]][["GloScope_hvg2000_pcadims30"]]$exec_time
}
```


```{r}
r_exec_times_path <- file.path(getwd(), "data", "r_execution_times.rds")

if (!file.exists(r_exec_times_path)) {
  r_exec_times <- data.frame(
    dataset = character(),
    method = character(),
    time_secs = numeric(),
    n_samples = numeric(),
    n_features = numeric()
  )
  
  bmark_result_list <- result_list[["bmark"]]
  
  ds_test <- names(bmark_result_list)
  
  for (ds in ds_test) {
    for (met in names(bmark_result_list[[ds]])) {
      if ("exec_time" %in% names(bmark_result_list[[ds]][[met]])) {
        temp <- data.frame(
          dataset = ds,
          method = met,
          time_secs = as.numeric(bmark_result_list[[ds]][[met]][["exec_time"]], units = "secs"),
          n_samples = nrow(bmark_result_list[[ds]][[met]][["feat_mat"]]),
          n_features = ncol(bmark_result_list[[ds]][[met]][["feat_mat"]])
        )
        
        r_exec_times <- rbind(r_exec_times, temp)
      }
    }
  }
  
  # r_exec_times$dataset[grepl("GongSharma_all", r_exec_times$dataset)] <- gsub("GongSharma_all", "GongSharma", r_exec_times$dataset[grepl("GongSharma_all", r_exec_times$dataset)])

  saveRDS(object = r_exec_times, r_exec_times_path)
} else {
  r_exec_times <- readRDS(r_exec_times_path)
}
```

## Read execution times from python

```{r}
p_exec_times_path <- file.path(getwd(), "data", "p_execution_times.rds")

if (!file.exists(p_exec_times_path)) {
  p_exec_times <- arrow::read_feather(
    file.path(getwd(), "data", "execution_times.feather")
  ) %>%
    as.data.frame()
  
  p_exec_times$n_samples <- 0
  p_exec_times$n_features <- 0
  
  p_exec_times$dataset[p_exec_times$dataset == "Gongsharma_all"] <- "GongSharma_all"
  p_exec_times <- p_exec_times[!p_exec_times$dataset %in% c("Gongsharma_sex_young_cmvneg", "Gongsharma_cmv_females_young", "Gongsharma_age_females_cmvneg"), ]
  
  p_exec_times$method[p_exec_times$method == "scPoli_hvg2000_dims5_lowres"] <- "scPoli_hvg2000_dims5"
  p_exec_times$method[p_exec_times$method == "PILOT_hvg2000_highres"] <- "PILOT_hvg2000"
  
  for (ds in unique(p_exec_times$dataset)) {
    p_exec_times$n_samples[p_exec_times$dataset == ds] <- r_exec_times$n_samples[r_exec_times$dataset == ds][1]
    p_exec_times$n_features[p_exec_times$dataset == ds] <- r_exec_times$n_features[r_exec_times$dataset == ds][1]
  }
  
  saveRDS(object = p_exec_times, p_exec_times_path)
} else {
  p_exec_times <- readRDS(p_exec_times_path)
}
```

## Combine and get actual cell numbers used per dataset

```{r}
exec_times_path <- file.path(getwd(), "data", "exec_times.rds")

if (!file.exists(exec_times_path)) {
  exec_times <- rbind(r_exec_times, p_exec_times)
  exec_times$n_cells <- 0
  
  dss <- names(datasets)
  dss <- dss[!dss %in% c("GongSharma_all_cmv", "GongSharma_all_sex")]
  
  exec_times <- exec_times[!exec_times$dataset %in% c("Gongsharma_sex_young_cmvneg", "Gongsharma_cmv_females_young", "Gongsharma_age_females_cmvneg"), ]
  
  for (ds in dss) {
    print(ds)
    seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
    
    if (ds == "GongSharma_all_age") ds <- "GongSharma_all"
    
    exec_times$n_cells[exec_times$dataset == ds] <- nrow(seurat@meta.data)
  }
  
  saveRDS(object = exec_times, exec_times_path)
} else {
  exec_times <- readRDS(exec_times_path)
}
```


## Supp fig 13: Plot

```{r}
filter_methods <- c(
  "ECODA_authors_HR",

  # "Pseudobulk_hvg2000",
  "Pseudobulk_hvg2000",
  
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000"
)

library(scales)
library(ggrepel)

scientific_A_10n <- function(x) {
  s <- scales::scientific_format(digits = 2)(x) 
  s_converted <- gsub("e\\+", " %*% 10^", s) 
  return(parse(text = s_converted)) 
}

# --- 1. Clean Data ---
plot_data <- exec_times %>%
  filter(method %in% filter_methods) %>%
  # filter(!dataset %in% c("GongSharma_all_cmv", "GongSharma_all_sex")) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "GloScope_hvg2000_pcadims30" = "GloScope",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims15_highres" = "scPoli",
                         "PILOT_hvg2000" = "PILOT"))

# --- 2. Calculate the End of the Regression Line ---
line_ends <- plot_data %>%
  group_by(method) %>%
  summarise(
    max_x = max(n_cells),
    model = list(lm(log10(time_secs) ~ n_cells)),
    pred_log_y = stats::predict(model[[1]], newdata = data.frame(n_cells = max_x))
  ) %>%
  mutate(
    pred_y = 10^pred_log_y 
  )

# --- NEW STEP: Create Ranked Color Palette ---

# A. Sort methods by predicted time (Fastest/Lowest -> Slowest/Highest)
ranked_methods <- line_ends %>%
  arrange(pred_y) %>%
  pull(method)

# B. Create a gradient function (Green -> Black -> Red)
# I used "forestgreen" and "firebrick" as they are usually more readable than neon green/red
grad_func <- colorRampPalette(c("dodgerblue3", "grey50", "red"))

# C. Generate the exact number of colors needed
method_colors <- grad_func(length(ranked_methods))

# D. Assign the method names to the colors so ggplot maps them correctly
names(method_colors) <- ranked_methods


# --- 3. Plot ---
ggplot(plot_data, aes(x = n_cells, y = time_secs, color = method)) +
  annotation_logticks(sides = "l", color = "grey80") +
  # --- The Actual Data ---
  geom_point(shape = 4, size = 1, stroke = 1.25, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 0.7) +
  
  # --- The Labels ---
  geom_text_repel(
    data = line_ends, 
    aes(x = max_x, y = pred_y, label = method), 
    size = 4,
    direction = "y",
    hjust = 0,
    nudge_x = max(plot_data$n_cells) * 0.05,
    segment.size = 0.2,
    segment.color = "grey50",
    show.legend = FALSE,
    bg.color = "white", # Optional: adds a white halo around text for readability
    bg.r = 0.15
  ) +
  
  # --- NEW: Apply the Custom Colors ---
  # scale_color_manual(values = method_colors) +
  
  # --- Theme & Scales ---
  labs(x = "Number of cells", y = "Time (s)") +
  theme_minimal() +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 3),
    labels = function(x) sapply(x, scientific_A_10n),
    limits = c(1e5, max(plot_data$n_cells) * 1.4)
  ) +
  scale_y_log10(
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))
  ) +
  theme(legend.position = "none")

ggsave(file.path(path_plots, "Supp_fig_13_benchmark_exec_times.pdf"),
       width = 3.5,
       height = 3.5, limitsize = FALSE)
```

# ---------------------------------------------------------------

# ---------------------------------------------------------------

# Figure 2 A: Plot benchmark results

## Main plot (with funkyheatmap)

```{r fig.height=6, fig.width=12}
# --- 1. SETUP & CLEANING (Same as before) ---
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  # "ECODA_HiTME_HR_layer2",
  # "ECODA_HiTME_HR_layer3",
  # "ECODA_seuratres_2",
  # "ECODA_PB_combo_normquantile_ecodaweight0.5",
  # paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  # "ECODA_authors_LR",
  # "Avg_PCA_embedding",
  # "Freq_highres",
  # "ECODA_deconv",
  # "Pseudobulk_10_PCA_dims",
  "Pseudobulk_hvg2000",
  # "Pseudobulk_hvg2000_bl",
  # "Pseudobulk_schvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat",
  "GloProp"
)

df_results_bmark <- df_results
df_results_bmark$temp <- paste0(df_results_bmark$dataset, df_results_bmark$method)
exec_times$temp <- paste0(exec_times$dataset, exec_times$method)

df_results_bmark <- merge(df_results_bmark,
                          exec_times[, c("temp", "time_secs")], 
                          by = "temp", 
                          all.x = TRUE)

df_results_bmark$temp <- NULL
exec_times$temp <- NULL


clean_data <- df_results_bmark %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         # "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         # "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
                         # "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "ECODA_PB_combo_normquantile_ecodaweight0.5" = "ECODA+Pseudobulk",
                         # "Pseudobulk_10_PCA_dims" = "Pseudobulk_PCA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         # "Pseudobulk_schvg2000" = "Pseudobulk_schvg",
                         # "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims15_highres" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  mutate(score = recode(score,
                        "anosim_score" = "ANOSIM",
                        "mod_knn3_score" = "Modularity",
                        "mod_score" = "Modularity",
                        "cluster_score" = "ARI")) %>%
  # mutate(dataset = recode(dataset,
  #                         "Adams" = "Adams (pulmonary fibrosis, disease type)",
  #                         "Bassez" = "Bassez (breast cancer, treatment response)",
  #                         "GongSharma_all_age" = "GongSharma (healthy, age group)",
  #                         "GongSharma_all_cmv" = "GongSharma (healthy, CMV-infection status)",
  #                         "GongSharma_all_sex" = "GongSharma (healthy, sex)",
  #                         "Kfoury" = "Kfoury (prostate metastasis, location)",
  #                         "Lee" = "Lee (primary glioblastoma, treatment status)",
  #                         "Pelka" = "Pelka (colon cancer, tumor or healthy)",
  #                         "Smillie" = "Smillie (ulcerative colitis, inflammation status)",
  #                         "Stephenson" = "Stephenson (COVID-19, infection status)",
  #                         "Wu" = "Wu (breast cancer, subtype)",
  #                         "Zhang" = "Zhang (breast cancer, tissue)"))
  mutate(dataset = recode(dataset,
                          "Adams" = "Adams (pulmonary fibrosis)",
                          "Bassez" = "Bassez (ICB treatment response)",
                          "GongSharma" = "GongSharma (healthy, CMV)",
                          "GongSharma_all_age" = "GongSharma (healthy, age)",
                          "GongSharma_all_cmv" = "GongSharma (healthy, CMV)",
                          "GongSharma_all_sex" = "GongSharma (healthy, sex)",
                          "Kfoury" = "Kfoury (prostate metastasis)",
                          "Kim" = "Kim (metastatic lung adenocarcinoma)",
                          "Lee" = "Lee (ICB treatment vs naive)",
                          "Pelka" = "Pelka (colon cancer vs control)",
                          "Smillie" = "Smillie (ulcerative colitis)",
                          "Stephenson" = "Stephenson (COVID-19)",
                          "Wu" = "Wu (breast cancer subtype)",
                          "Zhang" = "Zhang (tissue)"))

# --- 2. CALCULATE GEOMETRIC MEAN OF RANKS (NEW STEP) ---
df_ranks <- clean_data %>%
  # Group by specific test case
  group_by(dataset, score) %>%
  # Rank: Higher Value = Better Rank (1). Ties get the minimum rank.
  mutate(rank = rank(-value, ties.method = "min")) %>%
  ungroup() %>%
  # Now aggregate per method
  group_by(method) %>%
  # Calculate Geometric Mean: exp(mean(log(x)))
  summarise(geo_mean_rank = exp(mean(log(rank), na.rm = TRUE))) %>%
  ungroup() %>%
  # Normalize so it plots nicely (0 to 1).
  # We INVERT it so Best Rank (lowest number) = 1.0 (biggest bar)
  mutate(Rank_Score = (max(geo_mean_rank) - geo_mean_rank) /
           (max(geo_mean_rank) - min(geo_mean_rank))) %>%
  select(method, Rank_Score)

# --- 3. SCALE DATA (0-1) FOR HEATMAP VISUALS ---
scaled_data <- clean_data %>%
  group_by(dataset, score) %>%
  mutate(value = min_max(value)) %>%
  ungroup()

# --- 4. AGGREGATE SCORES ---

# A. Overall Score (Arithmetic Mean of Scaled Values)
df_overall <- scaled_data %>%
  group_by(method) %>%
  summarise(Overall_Score = mean(value, na.rm = TRUE)) 

# B. Dataset Scores
df_datasets_wide <- scaled_data %>%
  group_by(method, dataset) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = dataset, values_from = value, names_prefix = "Dataset__")

# C. Metric Scores
df_metrics_wide <- scaled_data %>%
  group_by(method, score) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = score, values_from = value, names_prefix = "Metric__")

# D. Execution times
df_runtime <- clean_data %>%
  distinct(method, dataset, time_secs) %>% 
  group_by(method) %>%
  summarise(raw_time = mean(time_secs, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    log_time = log10(raw_time + 1),
    # Inverting so SHORTER time (faster) = BIGGER bar/BETTER score
    # Remove '1 -' if you want long bars to mean slow
    Runtime = min_max(log_time) 
  ) %>% 
  select(method, Runtime)

# --- 5. MERGE EVERYTHING ---
final_data <- df_overall %>%
  left_join(df_ranks, by = "method") %>% # <--- Join the new Rank Score
  left_join(df_datasets_wide, by = "method") %>%
  left_join(df_metrics_wide, by = "method") %>%
  left_join(df_runtime, by = "method") %>%
  arrange(desc(Overall_Score)) # Optional: sort by the new Rank Score instead

# --- 6. BUILD COLUMN INFO ---

# 1. Fixed Columns (Now ONLY method and Rank_Score)
info_method <- tribble(
  ~id,      ~group, ~name, ~geom,  ~palette, ~options,
  "method", "",     "",    "text", NA,       list(hjust = 0, width = 6)
)

# 2. Dataset Columns
dataset_cols <- colnames(final_data)[startsWith(colnames(final_data), "Dataset__")]
info_datasets <- tibble(id = dataset_cols) %>%
  mutate(
    group = "Datasets",
    name = str_remove(id, "Dataset__"), 
    geom = "circle",
    palette = "Datasets",
    options = list(list())
  ) %>% arrange(name)

# 3. Metric Columns (INCLUDING Score Avg at the start)
metric_cols <- colnames(final_data)[startsWith(colnames(final_data), "Metric__")]
target_metric_order <- c("ANOSIM", "Modularity", "ARI")

info_metrics_main <- tibble(id = metric_cols) %>%
  mutate(
    group = "Metrics",
    name = str_remove(id, "Metric__"),
    geom = "bar",
    palette = "Metrics",
    options = list(list(width = 2)),
    name_f = factor(name, levels = target_metric_order)
  ) %>% arrange(name_f) %>% select(-name_f)

# Define Score Avg as a Metric entry
info_score_avg <- tribble(
  ~id,             ~group,    ~name,  ~geom, ~palette,           ~options,
  "Overall_Score", "Overall", "Mean", "bar", "palette_avg_mean", list(width = 4)
)

# Combine metrics: Score Avg comes FIRST
info_metrics <- bind_rows(info_score_avg, info_metrics_main)

# 4. Runtime (The Rightmost column)
info_runtime <- tribble(
  ~id,       ~group,    ~name,     ~geom, ~palette,          ~options,
  "Runtime", "Runtime", "Runtime", "bar", "palette_runtime", list(width = 3)
)

# Combine everything in the final order
column_info <- bind_rows(info_method, info_score_avg, info_datasets, info_metrics_main, info_runtime)

ncolors <- 11

# --- 7. PLOT ---
funky_heatmap(
  final_data,
  column_info = column_info,
  scale_column = FALSE, 
  column_groups = tibble(
    group = c("Overall", "Datasets", "Metrics", "Runtime")
  ),
  palettes = list(
    palette_avg_mean = brewer.pal(n = ncolors, name = "RdBu"),
    Datasets = brewer.pal(n = ncolors, name = "RdBu"),
    Metrics = brewer.pal(n = ncolors, name = "RdBu"),
    palette_runtime = rev(brewer.pal(n = ncolors, name = "RdYlGn"))
  ),
  legends = list(
    list(
      palette = "palette_avg_mean",
      enabled = TRUE,
      title = "Overall Mean",
      labels = c(". 0", rep("", round((ncolors-2)/2)+1), "0.5", rep("", round((ncolors-2)/2)), "1")
    ),
    # list(
    #   palette = "palette_avg_rank",
    #   enabled = FALSE,
    #   title = "REMOVE PLACEHOLDER"
    # ),
    list(
      palette = "Datasets",
      enabled = TRUE,
      title = "Dataset separation",
      labels = c(".     Min", rep("", ncolors-2), "Max   .")
    ),
    list(
      palette = "Metrics",
      enabled = TRUE,
      title = "Metrics",
      labels = c(". 0", rep("", round((ncolors-2)/2)+1), "0.5", rep("", round((ncolors-2)/2)), "1")
    ),
    list(
      palette = "palette_runtime",
      enabled = TRUE,
      title = "Runtime log10(s)",
      labels = c("secs", "mins", "hours") 
    )
  ),
  position_args = position_arguments(
    col_width = 1.5,
    col_space = 0.2, col_annot_offset = 5
  )
)
ggsave(file.path(path_plots, "Figure_2_A_benchmark_mainmethods.pdf"),
       width = 10,
       height = 6)
```


## Figure 2 B: MDS Bassez

```{r}
ds <- "Bassez"

methods <- list(
  "ECODA_authors_HR" = "ECODA",
  "GloProp" = "GloProp",
  "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope",
  "Pseudobulk_hvg2000" = "Pseudobulk",
  "Freq_highres" = "CellType_Frequency%",
  "MrVI_hvg2000" = "MrVI",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scPoli_hvg2000_dims5" = "scPoli"
)

plot_list <- list()

for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  labels <- as.character(labels)
  labels[labels == "E"] <- "Responders"
  labels[labels == "NE"] <- "Non-responders"
  labels <- factor(labels, levels = c("Responders", "Non-responders"))
  
  plot_list[[m]] <- plot_mds(
    dist_mat,
    labels,
    cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
    coord_equal = TRUE, pointsize = 2,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect")

ggsave(file.path(path_plots, "Figure_2_B_MDS_Bassez.pdf"),
       width = 10,
       height = 5)
```


## Main plot (rank-based)


```{r fig.height=6, fig.width=12}
# --- 1. SETUP & CLEANING (Same as before) ---
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  # "ECODA_HiTME_HR_layer2",
  # "ECODA_HiTME_HR_layer3",
  # "ECODA_seuratres_2",
  # "ECODA_PB_combo_normquantile_ecodaweight0.5",
  # paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  # "ECODA_authors_LR",
  # "Avg_PCA_embedding",
  # "Freq_highres",
  # "ECODA_deconv",
  # "Pseudobulk_10_PCA_dims",
  "Pseudobulk_hvg2000",
  # "Pseudobulk_hvg2000_bl",
  # "Pseudobulk_schvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims5",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat"
)

df_results_bmark <- df_results
df_results_bmark$temp <- paste0(df_results_bmark$dataset, df_results_bmark$method)
exec_times$temp <- paste0(exec_times$dataset, exec_times$method)

df_results_bmark <- merge(df_results_bmark,
                          exec_times[, c("temp", "time_secs")], 
                          by = "temp", 
                          all.x = TRUE)

df_results_bmark$temp <- NULL
exec_times$temp <- NULL


clean_data <- df_results_bmark %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         # "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         # "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
                         # "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "ECODA_PB_combo_normquantile_ecodaweight0.5" = "ECODA+Pseudobulk",
                         # "Pseudobulk_10_PCA_dims" = "Pseudobulk_PCA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         # "Pseudobulk_schvg2000" = "Pseudobulk_schvg",
                         # "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims5" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  mutate(score = recode(score,
                        "anosim_score" = "ANOSIM",
                        "mod_knn3_score" = "Modularity",
                        "mod_score" = "Modularity",
                        "cluster_score" = "ARI")) %>%
  mutate(dataset = recode(dataset,
                          "GongSharma_all_age" = "GongSharma_age",
                          "GongSharma_all_cmv" = "GongSharma_cmv",
                          "GongSharma_all_sex" = "GongSharma_sex"))

# --- 1. TRANSFORMATION TO RANKS ---
df_ranked <- clean_data %>%
  group_by(dataset, score) %>%
  mutate(rank = rank(-value, ties.method = "average")) %>%
  ungroup()

# --- 2. AGGREGATE RANK-BASED SCORES ---
df_overall <- df_ranked %>%
  group_by(method) %>%
  summarise(Overall_Rank = mean(rank, na.rm = TRUE))

df_datasets_wide <- df_ranked %>%
  group_by(method, dataset) %>%
  summarise(avg_rank = mean(rank, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = dataset, 
              values_from = avg_rank, names_prefix = "Dataset__")

target_metric_order <- c("ANOSIM", "Modularity", "ARI")

df_metrics_wide <- df_ranked %>%
  mutate(score = factor(score, levels = target_metric_order)) %>%
  group_by(method, score) %>%
  summarise(avg_rank = mean(rank, na.rm = TRUE), .groups = "drop") %>%
  arrange(score) %>%
  pivot_wider(id_cols = method, names_from = score, 
              values_from = avg_rank, names_prefix = "Metric__")

# --- 3. MERGE, INVERT, AND SCALE (Fixes the Error) ---
final_data <- df_overall %>%
  left_join(df_datasets_wide, by = "method") %>%
  left_join(df_metrics_wide, by = "method") %>%
  # Add mandatory 'id' column for funkyheatmap
  mutate(id = method)

# Identify all rank-based columns
rank_cols <- c("Overall_Rank", colnames(final_data)[startsWith(colnames(final_data), "Dataset__")], 
               colnames(final_data)[startsWith(colnames(final_data), "Metric__")])

# Instead of using min(x) and max(x) which vary per column,
# we use the theoretical best (1) and worst (total methods).
num_methods <- length(unique(final_data$method))

scale_ranks_global <- function(x, n_meth) {
  if(all(is.na(x))) return(x)
  # BEST rank (1) should map to 1.0
  # WORST rank (n_meth) should map to 0.0
  # Formula: (Worst - Current) / (Worst - Best)
  (n_meth - x) / (n_meth - 1)
}

final_data <- final_data %>%
  mutate(across(all_of(rank_cols), ~scale_ranks_global(.x, num_methods))) %>%
  arrange(desc(Overall_Rank))

# --- 4. COLUMN INFO CONFIGURATION ---
info_method <- tribble(
  ~id,      ~group, ~name,    ~geom, ~palette, ~options,
  "id",     "",     "",       "text", NA,       list(hjust = 0, width = 6)
)

info_overall <- tribble(
  ~id,            ~group,   ~name,      ~geom, ~palette,        ~options,
  "Overall_Rank", "Overall", "Avg Rank", "bar",  "palette_ranks", list(width = 4)
)

dataset_cols <- colnames(final_data)[startsWith(colnames(final_data), "Dataset__")]
info_datasets <- tibble(id = dataset_cols) %>%
  mutate(
    group = "Avg Rank per Dataset",
    name = str_remove(id, "Dataset__"), 
    geom = "circle",
    palette = "palette_ranks",
    options = list(list())
  )

metric_cols <- colnames(final_data)[startsWith(colnames(final_data), "Metric__")]
info_metrics <- tibble(id = metric_cols) %>%
  mutate(
    group = "Avg Rank per Metric",
    name = str_remove(id, "Metric__"),
    name_f = factor(name, levels = target_metric_order),
    geom = "bar",
    palette = "palette_ranks",
    options = list(list(width = 2))
  ) %>% 
  arrange(name_f) %>% 
  select(-name_f)

column_info <- bind_rows(info_method, info_overall, info_datasets, info_metrics)

# --- 5. PLOT ---
funky_heatmap(
  data = final_data,
  column_info = column_info,
  scale_column = FALSE, # Values are already 0-1
  palettes = list(
    palette_ranks = brewer.pal(n = ncolors, name = "RdBu")
  ),
  column_groups = tibble(
    group = c("Overall", "Avg Rank per Dataset", "Avg Rank per Metric")
  ),
  legends = list(
    list(palette = "palette_ranks", title = "Rank", labels = c(".       Worst", rep("", round((ncolors-2)/2)+1), "", rep("", round((ncolors-2)/2)), "Best    ."))
  )
)
ggsave(file.path(path_plots, "benchmark_summary_funkyheatmap_mainmethods_ranks.pdf"),
       width = 10,
       height = 5)
```



## Supp fig 14: Extended benchmark (with avg rank)

```{r fig.height=8, fig.width=14}
# --- 1. SETUP & CLEANING (Same as before) ---
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_scATOMIC_HR",
  # "ECODA_HiTME_HR_layer3",
  "ECODA_seuratres_2",
  # "ECODA_PB_combo_normquantile_ecodaweight0.5",
  # paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  # "Pseudobulk_10_PCA_dims",
  "Pseudobulk_hvg2000",
  "Pseudobulk_ct_LR_hvg2000",
  "Pseudobulk_ct_HR_hvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30_sqrtmat",
  "GloProp"
)

df_results_bmark <- df_results
df_results_bmark$temp <- paste0(df_results_bmark$dataset, df_results_bmark$method)
exec_times$temp <- paste0(exec_times$dataset, exec_times$method)

df_results_bmark <- merge(df_results_bmark,
                          exec_times[, c("temp", "time_secs")], 
                          by = "temp",
                          all.x = TRUE)

df_results_bmark$temp <- NULL
exec_times$temp <- NULL


clean_data <- df_results_bmark %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated_HiTME",
                         "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
                         "ECODA_scATOMIC_HR" = "ECODA_automated_scATOMIC",
                         "ECODA_seuratres_2" = "ECODA_Leiden_res_2",
                         # "ECODA_PB_combo_normquantile_ecodaweight0.5" = "ECODA+Pseudobulk",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk_PCA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         "Pseudobulk_ct_LR_hvg2000" = "Pseudobulk_CT_LR",
                         "Pseudobulk_ct_HR_hvg2000" = "Pseudobulk_CT_HR",
                         "Freq_highres" = "CellType_Frequency%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims15_highres" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30_sqrtmat" = "GloScope")) %>%
  mutate(score = recode(score,
                        "anosim_score" = "ANOSIM",
                        "mod_knn3_score" = "Modularity",
                        "mod_score" = "Modularity",
                        "cluster_score" = "ARI")) %>%
  mutate(dataset = recode(dataset,
                          "Adams" = "Adams (pulmonary fibrosis)",
                          "Bassez" = "Bassez (ICB treatment response)",
                          "GongSharma" = "GongSharma (healthy, cmv)",
                          "GongSharma_all_age" = "GongSharma (healthy, age)",
                          "GongSharma_all_cmv" = "GongSharma (healthy, CMV)",
                          "GongSharma_all_sex" = "GongSharma (healthy, sex)",
                          "Kfoury" = "Kfoury (prostate metastasis)",
                          "Kim" = "Kim (metastatic lung adenocarcinoma)",
                          "Lee" = "Lee (ICB treatment vs naive)",
                          "Pelka" = "Pelka (colon cancer vs control)",
                          "Smillie" = "Smillie (ulcerative colitis)",
                          "Stephenson" = "Stephenson (COVID-19)",
                          "Wu" = "Wu (breast cancer subtype)",
                          "Zhang" = "Zhang (tissue)"))

# --- 2. CALCULATE GEOMETRIC MEAN OF RANKS (NEW STEP) ---
df_ranks <- clean_data %>%
  # Group by specific test case
  group_by(dataset, score) %>% 
  # Rank: Higher Value = Better Rank (1). Ties get the minimum rank.
  mutate(rank = rank(-value, ties.method = "min")) %>% 
  ungroup() %>%
  # Now aggregate per method
  group_by(method) %>% 
  # # Calculate Geometric Mean: exp(mean(log(x)))
  # summarise(geo_mean_rank = exp(mean(log(rank), na.rm = TRUE))) %>%
  summarise(mean_rank = mean(rank, na.rm = TRUE)) %>%
  ungroup() %>%
  # Normalize so it plots nicely (0 to 1). 
  # We INVERT it so Best Rank (lowest number) = 1.0 (biggest bar)
  mutate(Rank_Score = (max(mean_rank) - mean_rank) / 
           (max(mean_rank) - min(mean_rank))) %>%
  select(method, Rank_Score)

# --- 3. SCALE DATA (0-1) FOR HEATMAP VISUALS ---
scaled_data <- clean_data %>%
  group_by(dataset, score) %>%
  mutate(value = min_max(value)) %>%
  ungroup()

# --- 4. AGGREGATE SCORES ---

# A. Overall Score (Arithmetic Mean of Scaled Values)
df_overall <- scaled_data %>%
  group_by(method) %>%
  summarise(Overall_Score = mean(value, na.rm = TRUE)) 

# B. Dataset Scores
df_datasets_wide <- scaled_data %>%
  group_by(method, dataset) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = dataset, values_from = value, names_prefix = "Dataset__")

# C. Metric Scores
df_metrics_wide <- scaled_data %>%
  group_by(method, score) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = score, values_from = value, names_prefix = "Metric__")

# D. Execution times
df_runtime <- clean_data %>%
  distinct(method, dataset, time_secs) %>% 
  group_by(method) %>%
  summarise(raw_time = mean(time_secs, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    log_time = log10(raw_time + 1),
    # Inverting so SHORTER time (faster) = BIGGER bar/BETTER score
    # Remove '1 -' if you want long bars to mean slow
    Runtime = min_max(log_time) 
  ) %>% 
  select(method, Runtime)

# --- 5. MERGE EVERYTHING ---
final_data <- df_overall %>%
  left_join(df_ranks, by = "method") %>% # <--- Join the new Rank Score
  left_join(df_datasets_wide, by = "method") %>%
  left_join(df_metrics_wide, by = "method") %>%
  left_join(df_runtime, by = "method") %>%
  arrange(desc(Overall_Score))
  # arrange(desc(Rank_Score)) # Optional: sort by the new Rank Score instead

# --- 6. BUILD COLUMN INFO ---

# 1. Method Column (The Label)
info_method <- tribble(
  ~id,      ~group, ~name, ~geom,  ~palette, ~options,
  "method", "",     "",    "text", NA,       list(hjust = 0, width = 8)
)

# 2.1 Score Avg
info_score_avg <- tribble(
  ~id,             ~group,    ~name,  ~geom, ~palette,           ~options,
  "Overall_Score", "Overall", "Metrics Mean", "bar", "palette_avg_mean", list(width = 4)
)

# 2.2 Rank Avg
info_rank_avg <- tribble(
  ~id,             ~group,    ~name,           ~geom,  ~palette,          ~options,
  "Rank_Score",    "Overall", "Rank Mean",  "bar",  "palette_avg_rank", list(width = 4, legend = FALSE)
)

# 3. Dataset Columns
dataset_cols <- colnames(final_data)[startsWith(colnames(final_data), "Dataset__")]
info_datasets <- tibble(id = dataset_cols) %>%
  mutate(
    group = "Datasets",
    name = str_remove(id, "Dataset__"), 
    geom = "circle",
    palette = "Datasets",
    options = list(list())
  ) %>% arrange(name)

# 4. Metric Columns (Individual metrics)
metric_cols <- colnames(final_data)[startsWith(colnames(final_data), "Metric__")]
target_metric_order <- c("ANOSIM", "Modularity", "ARI")

info_metrics_main <- tibble(id = metric_cols) %>%
  mutate(
    group = "Metrics",
    name = str_remove(id, "Metric__"),
    geom = "bar",
    palette = "Metrics",
    options = list(list(width = 2)),
    name_f = factor(name, levels = target_metric_order)
  ) %>% arrange(name_f) %>% select(-name_f)

# 5. Runtime (The Rightmost column)
info_runtime <- tribble(
  ~id,       ~group,    ~name,     ~geom, ~palette,          ~options,
  "Runtime", "Runtime", "Runtime", "bar", "palette_runtime", list(width = 3)
)

# --- KEY CHANGE: Re-order the bind_rows to define horizontal placement ---
# This order determines the columns from left to right.
column_info <- bind_rows(
  info_method,      # 1st: Method Name
  info_score_avg,   # 2.1: Mean Metric Score
  info_rank_avg,    # 2.2: Mean rank
  info_datasets,    # 3rd: Dataset Circles
  info_metrics_main,# 4th: Individual Metric Bars
  info_runtime      # 5th: Runtime Bar
)

# 
# # 2. Dataset Columns
# dataset_cols <- colnames(final_data)[startsWith(colnames(final_data), "Dataset__")]
# info_datasets <- tibble(id = dataset_cols) %>%
#   mutate(
#     group = "Datasets",
#     name = str_remove(id, "Dataset__"), 
#     geom = "circle",
#     palette = "Datasets",
#     options = list(list())
#   ) %>% arrange(name)
# 
# # 3. Metric Columns (INCLUDING Score Avg at the start)
# metric_cols <- colnames(final_data)[startsWith(colnames(final_data), "Metric__")]
# target_metric_order <- c("ANOSIM", "Modularity", "ARI")
# 
# # Create the specific Metric entries
# info_metrics_main <- tibble(id = metric_cols) %>%
#   mutate(
#     group = "Metrics",
#     name = str_remove(id, "Metric__"),
#     geom = "bar",
#     palette = "Metrics",
#     options = list(list(width = 2)),
#     name_f = factor(name, levels = target_metric_order)
#   ) %>% arrange(name_f) %>% select(-name_f)
# 
# # Define Score Avg as a Metric entry
# info_score_avg <- tribble(
#   ~id,             ~group,    ~name,       ~geom, ~palette,           ~options,
#   "Overall_Score", "Metrics", "Mean", "bar", "palette_avg_mean", list(width = 4, legend = FALSE)
# )
# 
# # Combine metrics: Score Avg comes FIRST
# info_metrics <- bind_rows(info_score_avg, info_metrics_main)
# 
# # 4. Runtime (The Rightmost column)
# info_runtime <- tribble(
#   ~id,       ~group,    ~name,     ~geom, ~palette,          ~options,
#   "Runtime", "Runtime", "Runtime", "bar", "palette_runtime", list(width = 3)
# )
# 
# # Combine everything in the final order
# column_info <- bind_rows(info_left, info_datasets, info_metrics, info_runtime)

ncolors <- 11

# --- 7. PLOT ---
funky_heatmap(
  final_data,
  column_info = column_info,
  scale_column = FALSE, 
  column_groups = tibble(
    group = c("Overall", "Datasets", "Metrics", "Runtime")
  ),
  palettes = list(
    palette_avg_rank = brewer.pal(n = ncolors, name = "RdBu"),
    palette_avg_mean = brewer.pal(n = ncolors, name = "RdBu"),
    Datasets = brewer.pal(n = ncolors, name = "RdBu"),
    Metrics = brewer.pal(n = ncolors, name = "RdBu"),
    palette_runtime = rev(brewer.pal(n = ncolors, name = "RdYlGn"))
  ),
  legends = list(
    list(
      palette = "palette_avg_rank",
      enabled = FALSE,
      title = "REMOVE PLACEHOLDER"
    ),
    list(
      palette = "palette_avg_mean",
      enabled = TRUE,
      title = "Overall",
      labels = c(".         Worst", rep("", ncolors-2), "Best       .")
    ),
    list(
      palette = "Datasets",
      enabled = TRUE,
      title = "Dataset separation",
      labels = c(".     Min", rep("", ncolors-2), "Max   .")
    ),
    list(
      palette = "Metrics",
      enabled = TRUE,
      labels = c(". 0", rep("", round((ncolors-2)/2)+1), "0.5", rep("", round((ncolors-2)/2)), "1")
    ),
    list(
      palette = "palette_runtime",
      enabled = TRUE,
      title = "Runtime log10(s)",
      labels = c("secs", "mins", "hours") 
    )
  ),
  position_args = position_arguments(
    col_width = 1.5,
    col_space = 0.2, col_annot_offset = 5
  )
)
ggsave(file.path(path_plots, "Supp_fig_14_benchmark_extended.pdf"),
       width = 14,
       height = 8)
```


# ---------------------------------------------------------------

# ---------------------------------------------------------------


# Supp fig 1: Correlation between metrics

```{r}
# --- 1. PREPARE WIDE DATA ---
# Pivot so each metric is a column, and calculate the mean
df_wide <- clean_data %>%
  pivot_wider(id_cols = c(dataset, method), 
              names_from = score, 
              values_from = value) %>%
  mutate(Mean_Score = (ANOSIM + Modularity + ARI) / 3)

# --- 2. DEFINE A REUSABLE PLOTTING FUNCTION ---
# This ensures all your plots have the same look and regression stats
plot_corr <- function(data, x_var, y_var, x_label, y_label, color_hex) {
  lims <- c(min(data[, 3:6])-0.05, 1.2)
  ggplot(data, aes_string(x = x_var, y = y_var)) +
    geom_abline(intercept = 0, slope = 1, 
                linetype = "dashed", 
                color = "grey50", 
                linewidth = 0.8) +
    geom_point(alpha = 0.4, color = color_hex, size = 1.5) +
    geom_smooth(method = "lm", color = "black", linewidth = 1, se = FALSE) +
    stat_cor(aes(label = ..rr.label..), label.x.npc = "left", label.y.npc = "top", label.y = max(data[, y_var]) * 1.025) +
    labs(x = x_label, y = y_label) +
    xlim(lims) + ylim(lims) +
    theme_classic()
}

# --- 3. CREATE THE INDIVIDUAL PLOTS ---

# A. Metrics vs Mean (from your previous request)
p1 <- plot_corr(df_wide, "Mean_Score", "ANOSIM", "Mean Score", "ANOSIM", "#000000")
p2 <- plot_corr(df_wide, "Mean_Score", "Modularity", "Mean Score", "Modularity", "#000000")
p3 <- plot_corr(df_wide, "Mean_Score", "ARI", "Mean Score", "ARI", "#000000")

# B. Pairwise Correlations (New request)
p4 <- plot_corr(df_wide, "Modularity", "ANOSIM", "Modularity", "ANOSIM", "#000000")
p5 <- plot_corr(df_wide, "ARI", "Modularity", "ARI", "Modularity", "#000000")
p6 <- plot_corr(df_wide, "ANOSIM", "ARI", "ANOSIM", "ARI", "#000000")

# --- 4. COMBINE AND DISPLAY ---
# Using patchwork to create a clean layout
(p4 | p5 | p6) / (p1 | p2 | p3) + 
  plot_annotation(
    # title = "Metric Consistency Analysis",
    # subtitle = "Top: Individual metrics vs. Grand Mean | Bottom: Pairwise metric correlations",
    theme = theme(plot.title = element_text(face = "bold", size = 16))
  )
ggsave(file.path(path_plots, "Supp_fig_1_metrics_correlation.pdf"),
       width = 7.5,
       height = 5)
```

# ---------------------------------------------------------------

# ---------------------------------------------------------------

# Supp fig 15 A: Plot normalization results

## Plot results

```{r}
df_results <- Reduce(rbind, result_list[["trans"]]) %>%
  mutate(trans_method = recode(trans_method,
                         "clr" = "CLR",
                         "alr_mincvref" = "ALR_minCVref",
                         "alr_randref" = "ALR_randref",
                         "arcsine_sqrt" = "Arcsine_sqrt",
                         "counts" = "Counts",
                         "freq" = "Frequency%"))

df_results <- df_results %>%
  pivot_longer(
    cols = c(ANOSIM_score, Modularity_score, Adjusted_Rand_Index),
    names_to = "score",
    values_to = "value"
  ) %>%
  as.data.frame()


df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score") %>%
  group_by(trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_anosim <- df_results_anosim %>%
  arrange(desc(mean))

df_results_anosim$trans_method <- factor(df_results_anosim$trans_method, levels = df_results_anosim$trans_method)
df_results$trans_method <- factor(df_results$trans_method, levels = df_results_anosim$trans_method)

score.labs <- c("ANOSIM score", "Adjusted rand index", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))


df_results_list[["Normalization"]] <- df_results


p1 <- df_results %>%
  filter(score == "ANOSIM_score") %>%
  
  ggplot(aes(x = trans_method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  
  stat_compare_means(
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..))),
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 0.95
  ) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  ylim(c(-0.1, 1))
print(p1)
ggsave(file.path(path_plots, "Supp_fig_15A_transformations_anosim.pdf"),
       width = 2.5,
       height = 3.5)



df_results %>%
  group_by(score, trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = trans_method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores_bars.pdf"),
       width = 7,
       height = 7)



my_comparisons <- list(
  c("CLR", "Frequency(%)"),
  c("CLR", "Counts")
  )

df_results %>%
  filter(score == "ANOSIM_score") %>%
  filter(trans_method %in% c("Counts", "Frequency(%)", "CLR")) %>%
  
  ggplot(aes(x = fct_reorder(trans_method, value, .desc = FALSE, .fun = mean), y = value, fill = ifelse(trans_method == "CLR", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    paired = TRUE,
    size = 3,
    tip.length = 0,
    # label.y = 0.95,
    aes(label = after_stat(p.signif))
  ) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  ylim(c(-0.05, 1)) +
  scale_fill_identity() +
  scale_x_discrete(labels = c(
    "Counts",
    "Frequency(%)",
    expression(bold("CLR")) # Make label bold
  ))
ggsave(file.path(path_plots, "POSTER_transformations_real_datasets_summary_barplot_anosim_refCLR.pdf"),
       width = 2,
       height = 3.5)





# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "ANOSIM score", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))



ggplot(df_results, aes(x = trans_method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
# facet_grid(scales = "free",
#            rows = vars(score_method), cols = vars(dataset))
ggsave(file.path(path_plots, "transformations_real_datasets_detailed_barplot.pdf"),
       width = 8,
       height = 7)



# df_results %>%
#   group_by(score_method, method) %>%
#   summarize(mean = mean(values), sem = sd(values) / sqrt(length(values))) %>%
#   ungroup() %>%
#   
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Separation score") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   facet_grid(scales = "free",
#              rows = vars(score_method),
#              labeller = labeller(score_method = score.labs))


ggplot(df_results, aes(x = trans_method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores.pdf"),
       width = 5,
       height = 5)



# df_results_anosim %>%
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Adjusted Rand Index") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   NoLegend()


df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score")

df_results_anosim <- df_results_anosim[!grepl("pca", df_results_anosim$trans_method), ]

my_comparisons <- purrr::map(unique(as.character(df_results_anosim$trans_method)), ~ c(.x, "Frequency(%)"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
# my_comparisons <- c(my_comparisons, list(c("clr", "arcsine_sqrt")))
my_comparisons <- my_comparisons[c(4, 3, 2, 1, 5)]


ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    # ref.group = "freq",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot.pdf"),
       width = 6,
       height = 4.5)


ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    # comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "ANOSIM score") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #                    y.position = c(max(df_results_anosim$values)+0.1, max(df_results_anosim$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

# ---------------------------------------------------------------

# ---------------------------------------------------------------

# Supp fig 15 B: Plot zero imputation results

## Plot results

```{r}
zeroimp_result_list <- result_list[["zeroimp"]]

df_results <- imap_dfr(zeroimp_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    imap_dfr(.x, ~ {
      tibble(
        dataset = dataset_name,
        method = method_name,
        score = .y,
        value = .x
      )
    })
  })
})

df_results <- df_results[!grepl("impRZilr", df_results$method), ]
df_results <- df_results[!df_results$score == "mod_score", ]

df_results_list[["Zero_imputation"]] <- df_results

df_results_anosim <- df_results %>%
  filter(score == "anosim_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_anosim <- df_results_anosim %>%
  arrange(desc(mean))

df_results_anosim$method <- factor(df_results_anosim$method, levels = df_results_anosim$method)
df_results$method <- factor(df_results$method, levels = df_results_anosim$method)



p2 <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.95,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p2)
ggsave(file.path(path_plots, "Supp_fig_15B_zero_replacement_anosim.pdf"),
       width = 2.5,
       height = 4)



df_results %>%
  filter(score == "mod_knn3_score") %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.9,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 0.9)) +
  labs(x = "", y = "Separation score (Modularity)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))



df_results %>%
  filter(score == "cluster_score") %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.9,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 0.9)) +
  labs(x = "", y = "Separation score (ARI)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))




# New facet label names for dose variable
score.labs <- c("ANOSIM score", "Adjusted rand index", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))



df_results %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 1.1,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.2, 1.2)) +
  labs(x = "", y = "Separation score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_separatescores.pdf"),
       width = 3,
       height = 8)




ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "zero_imputation_real_datasets_detailed_barplot.pdf"),
       width = 10,
       height = 6)




df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%

  ggplot(aes(x = method, y = mean)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_barplot_separatescores.pdf"),
       width = 3.5,
       height = 6)

ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_separatescores.pdf"),
       width = 7,
       height = 6)



df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results %>%
  filter(score == "anosim_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "counts_all_1"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(6,1,7:10, 2:5)]
# my_comparisons[[length(my_comparisons) + 1]] <- c("perc_all_0.1%", "perc_all_1%")

p <- ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_clusterscore.pdf"),
       width = 8,
       height = 6)

df_results_modularity <- df_results %>%
  filter(score == "mod_score")

p <- ggplot(df_results_modularity, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Modularity score") +
  scale_shape_manual(values=1:length(unique((df_results_modularity$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_modscore.pdf"),
       width = 8,
       height = 6)


df_results_anosim <- df_results %>%
  filter(score == "anosim_score")

p <- ggplot(df_results_anosim, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_anosimscore.pdf"),
       width = 8,
       height = 6)

```

# Supp fig 15AB plot

```{r}
wrap_plots(list(p1, p2)) +
  plot_layout(nrow = 1) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold"))
ggsave(file.path(path_plots, "Supp_fig_15AB_trans_zero_anosim.pdf"),
       width = 6,
       height = 4.5)
```


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Batch effect

## Joanito dataset

## Load/preprocess

```{r}
seurat <- readRDS(file.path(path_data, "JoaI_2022_35773407_Nofilt_whole_ECODAprocessed.rds"))

seurat@meta.data$cell.type_new <- seurat@meta.data$cell.type
iCMS <- !is.na(seurat@meta.data$iCMS)
seurat@meta.data$cell.type_new[iCMS] <- paste0(seurat@meta.data$cell.type[iCMS], "_", ifelse(seurat@meta.data$iCMS[iCMS] == "Normal", "Normal", "Cancer"))

meta <- seurat@meta.data %>%
  group_by(sample.ID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample.ID
meta$sample.origin <- gsub("-2", "", meta$sample.origin)
meta$seqtec <- ifelse(meta$dataset %in% c("CRC-SG1", "KUL5"), "5' seq", "3' seq")

bio_col <- "sample.origin"
batch_col <- "seqtec"

plot_list <- list()
```


## Plot PCA

### ECODA

```{r}
dist_mat <- as.data.frame.matrix(table(seurat@meta.data$sample.ID, seurat@meta.data$cell.type_new)) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 0.5) %>%
  clr() %>%
  dist()

plot_list[["p_ecoda_bio"]] <- plot_mds(dist_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = T,
         title = "ECODA - Biological groups"
)
print(plot_list[["p_ecoda_bio"]])


plot_list[["p_ecoda_batch"]] <- plot_mds(dist_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = T,
         title = "ECODA - Batch groups"
)
print(plot_list[["p_ecoda_batch"]])
```


### PB


#### Without blacklist

```{r}
pb_norm <- get_pb_deseq2(seurat, sample_col = "sample.ID", black_list = "none")

plot_list[["p_pb_bio"]] <- plot_mds(dist(pb_norm), labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = T,
         title = "Pseudobulk - Biological groups"
)
print(plot_list[["p_pb_bio"]])

plot_list[["p_pb_batch"]] <- plot_mds(dist(pb_norm), labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = T,
         title = "Pseudobulk - Batch groups"
)
print(plot_list[["p_pb_batch"]])
```

# Figure 5: Combine all plots

```{r fig.height=6.5, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = 'A') #&
  # theme(
  #   # legend.position='bottom',
  #   plot.tag = element_text(size = 16, face = "bold")
  # )
ggsave(file.path(path_plots, "Figure_5_batcheffect_Joanito_PCA.pdf"),
       width = 10,
       height = 6.5)
```

## Number of cells, total, per sample

```{r}
print(paste("Number of samples:", length(unique(seurat@meta.data$sample.ID))))
print(paste("Number of cells:", nrow(seurat@meta.data)))
print(paste("Number of cells per sample:", round(mean(table((seurat@meta.data$sample.ID)))), "+-", round(sd(table((seurat@meta.data$sample.ID))))))
print(paste("Number of cell types:", length(unique(seurat@meta.data$cell.type))))
```

# ---------------------------------------------------------------

# Stephenson dataset

## Load/preprocess

```{r}
p <- file.path(path_nas, "StephensonE_2021_39313646", "output", "StephensonE_2021_33879890_preprocessed.rds")

seurat <- readRDS(p)

seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))

seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset[seurat$Days_from_onset == "nan"] <- 999
seurat$Days_from_onset[seurat$Days_from_onset == "Not_known"] <- 999
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)


meta <- seurat@meta.data %>%
  group_by(Sample) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$Sample

pb <- get_pb(seurat, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

bio_col <- "Status"
batch_col <- "Site"


ct_remove <- c(
  "Platelets",
  "RBC",
  "C1_CD16_mono",
  "CD14_mono",
  "CD16_mono"
)
seurat_sub <- subset(seurat, subset = full_clustering %notin% ct_remove)
pb <- get_pb(seurat_sub, sample_col = "Sample")
metadata <- get_metadata(seurat, sample_col = "Sample")
pb_norm_sub <- t(DESeq2.normalize(pb, metadata = metadata, n_hvg = 2000))

plot_list <- list()
```


## Plot PCA

### ECODA

```{r}
dist_mat <- as.data.frame.matrix(table(seurat@meta.data$Sample, seurat@meta.data$full_clustering)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 0.5) %>%
  clr() %>%
  dist()

plot_mds(dist_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)


plot_mds(dist_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
```


#### Subset to remove batch effect cell types

```{r}
dist_mat <- as.data.frame.matrix(table(seurat_sub@meta.data$Sample, seurat_sub@meta.data$full_clustering)) %>%
  # select(-contains("Epithelial")) %>%
  impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 0.5) %>%
  clr() %>%
  dist()

plot_list[["p_ecoda_bio_sub"]] <- plot_mds(dist_mat, labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_ecoda_bio_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_bio.pdf"),
       width = 4,
       height = 3)


plot_list[["p_ecoda_batch_sub"]] <- plot_mds(dist_mat, labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_ecoda_batch_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_ECODA_subset_batch.pdf"),
       width = 4,
       height = 3)
```




### PB

```{r}
plot_mds(dist(pb_norm), labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)

plot_mds(dist(pb_norm), labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
```

#### Subset to remove batch effect cell types

```{r}
plot_list[["p_pb_bio_sub"]] <- plot_mds(dist(pb_norm_sub), labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_pb_bio_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_bio.pdf"),
       width = 4,
       height = 3)

plot_list[["p_pb_batch_sub"]] <- plot_mds(dist(pb_norm_sub), labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
print(plot_list[["p_pb_batch_sub"]])
ggsave(file.path(path_plots, "batcheffect_Stephenson_PCA_PB_subset_batch.pdf"),
       width = 4,
       height = 3)
```


#### With blacklist

```{r}
pb_norm_sub <- get_pb_deseq2(seurat_sub, sample_col = "Sample", black_list = "default")

plot_mds(dist(pb_norm_sub), labels = meta[[bio_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_bio.pdf"),
#        width = 4,
#        height = 3)

plot_mds(dist(pb_norm_sub), labels = meta[[batch_col]],
         cluster_score = F, mod_score = F, anosim_score = T
)
# ggsave(file.path(path_plots, "batcheffect_Joanito_PCA_PB_batch.pdf"),
#        width = 4,
#        height = 3)

```

# Supp fig 22: Combine all plots

```{r fig.height=15, fig.width=10}
wrap_plots(plot_list) +
  plot_layout(nrow = 2)+
  plot_annotation(tag_levels = 'A') &
  theme(
    # legend.position='bottom',
    plot.tag = element_text(size = 16, face = "bold")
  )
ggsave(file.path(path_plots, "Supp_fig_22_batcheffect_Stephenson_PCA_combined.pdf"),
       width = 10,
       height = 8)
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Export tables with raw data for paper

```{r}
library(writexl)
write_xlsx(df_results_list, path = file.path(path_data, "raw_data_for_publication.xlsx"))
```


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Info about datasets

```{r}
for (ds in names(result_list$bmark)) {
  print("-----------------------------------------")
  print(ds)
  feat_mat <- result_list$bmark[[ds]]$ECODA_authors_HR$feat_mat
  print(paste("Number of cell types:", ncol(feat_mat)))
  print(paste("Number of samples:", nrow(feat_mat)))
  print(paste("Number of cells:", exec_times[exec_times$dataset == ds, "n_cells"][1]))
  
  print(paste("ANOSIM using all cell types:", round(result_list$bmark[[ds]]$ECODA_authors_HR$scores$anosim_score, 3)))
  print(paste("ANOSIM at 10% HVC:", round(result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0.1$scores$anosim_score, 3)))
  print(paste("Number of HVC at 10% HVC:", ncol(result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0.1$feat_mat)))
  cat("Most highly variable cell types:", colnames(result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0$feat_mat), "\n")
  
  cat("Cell types:", colnames(feat_mat), "\n")
}
```

# Supp table 1:
## Number of cells, total, per sample

```{r}
for (ds in names(datasets)) {
  print("-----------------------------------------")
  print(ds)
  seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
  print(paste("Number of samples:", length(unique(seurat@meta.data$Sample))))
  print(paste("Number of cells:", nrow(seurat@meta.data)))
  print(paste("Number of cells per sample:", round(mean(table((seurat@meta.data$Sample)))), "+-", round(sd(table((seurat@meta.data$Sample))))))
  print(paste("Number of cell types:", length(unique(seurat@meta.data[[datasets[[ds]][["hi_res_ct_col"]]]]))))
}
```


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Test Mass Cytometry data

```{r}
df <- readxl::read_xlsx("/Users/christianhalter/Desktop/ECODA_paper/mmc5.xlsx", skip = 2)

# 2. Define Column Groups
broad_ct_cols  <- colnames(df)[5:9]
immune_ct_cols <- colnames(df)[14:21]
cd4_ct_cols    <- c("T01", "T02", "T03", "T04", "T08", "T09", "T13", "T17", "T18", "T20")
cd8_ct_cols    <- c("T05", "T06", "T07", "T10", "T11", "T12", "T14", "T15", "T16", "T19")
epi_ct_cols    <- paste0("Ep", sprintf("%02d", 1:45))
mye_ct_cols    <- paste0("M", sprintf("%02d", 1:19))

# 3. Calculate Compositions (% of Total Sample)
df_compositions <- df %>%
  mutate(
    # --- Immune Level ---
    # Convert Immune Subsets (T-cells, B-cells, etc.) from % of Immune to % of Total
    across(all_of(immune_ct_cols), 
           ~ .x * (`ImmuneCells [%]` / 100), 
           .names = "{.col}_TotalPct"),
    
    # --- T-Cell Lineage Level ---
    # Calculate CD4 and CD8 as % of Total Sample
    # Formula: (Lineage % of T) * (T % of Immune) * (Immune % of Total)
    TCells_CD4_TotalPct = (`TCells CD4+ [%]` / 100) * (`TCells [%]` / 100) * `ImmuneCells [%]`,
    TCells_CD8_TotalPct = (`TCells CD8+ [%]` / 100) * (`TCells [%]` / 100) * `ImmuneCells [%]`,
    
    # --- Cluster Level ---
    # T-Cell Clusters (relative to their respective CD4/CD8 parent)
    across(all_of(cd4_ct_cols), 
           ~ .x * TCells_CD4_TotalPct, 
           .names = "{.col}_TotalPct"),
    
    across(all_of(cd8_ct_cols), 
           ~ .x * TCells_CD8_TotalPct, 
           .names = "{.col}_TotalPct"),
    
    # Myeloid Clusters (relative to the Myeloid parent)
    # Formula: (Cluster Fraction) * (Myeloid % of Immune) * (Immune % of Total)
    across(all_of(mye_ct_cols), 
           ~ .x * (`MyeloidCells [%]` / 100) * `ImmuneCells [%]`, 
           .names = "{.col}_TotalPct"),
    
    # Epithelial Clusters (relative to the Epithelial parent)
    # Formula: (Cluster Fraction) * (Epithelial % of Total)
    across(all_of(epi_ct_cols), 
           ~ .x * `EpithelialCells [%]`, 
           .names = "{.col}_TotalPct")
  )

# 4. Verification Check
# The sum of all Epithelial cluster TotalPcts should equal the original EpithelialCells [%]
df_compositions %>% 
  select(Sample_ID = 1, `EpithelialCells [%]`, starts_with("Ep")) %>%
  mutate(Sum_Ep_Clusters = rowSums(select(., ends_with("_TotalPct")))) %>%
  head()


# 2. Load the clinical metadata
# Adjust the path to where your mmc2.xlsx is located
clinical_df <- readxl::read_xlsx("/Users/christianhalter/Desktop/ECODA_paper/mmc2.xlsx", skip = 2)

# 3. Join the dataframes
# We match 'Patient' from df (mmc5) with 'Patient ID' from clinical_df (mmc2)
df_combined <- df %>%
  left_join(clinical_df, by = c("Patient" = "Patient ID"))

# 4. (Optional) Move metadata columns
# If you want to see the Clinical Subtype or Grade immediately after the Patient ID:
df <- df_combined %>%
  relocate(any_of(c("Clinical Subtype", "Histopathology", "Grade")), .after = Patient)
```

```{r}
# broad_ct_cols  <- colnames(df)[5:9]
# immune_ct_cols <- colnames(df)[14:21]
# cd4_ct_cols    <- c("T01", "T02", "T03", "T04", "T08", "T09", "T13", "T17", "T18", "T20")
# cd8_ct_cols    <- c("T05", "T06", "T07", "T10", "T11", "T12", "T14", "T15", "T16", "T19")
# epi_ct_cols    <- paste0("Ep", sprintf("%02d", 1:45))
# mye_ct_cols    <- paste0("M", sprintf("%02d", 1:19))

feat_mat <- df[, mye_ct_cols]*100

min_val <- min(feat_mat[feat_mat > 0], na.rm = TRUE)
impute_val <- (2/3) * min_val
feat_mat[feat_mat == 0] <- impute_val

feat_mat <- clr(feat_mat)

plot_pca(feat_mat, labels = df$`Clinical Subtype`,
         cluster_score = FALSE, mod_score = FALSE, anosim_score = FALSE,invisible = "quali", n_ct_show = 5)
```

```{r}
# 1. Define Column Groups
broad_ct_cols  <- colnames(df)[5:9] 
immune_sub_cols <- c("NaturalKillerCells [%]", "Granulocytes [%]", "BCells [%]", 
                     "PlasmaCells [%]", "plasmacytoidDendriticCells [%]", "Basophils [%]")

cd4_ct_cols    <- c("T01", "T02", "T03", "T04", "T08", "T09", "T13", "T17", "T18", "T20")
cd8_ct_cols    <- c("T05", "T06", "T07", "T10", "T11", "T12", "T14", "T15", "T16", "T19")
mye_ct_cols    <- paste0("M", sprintf("%02d", 1:19))
epi_ct_cols    <- paste0("Ep", sprintf("%02d", 1:45))

# 2. Calculate Absolute Abundances (% of Total Sample)
df_abs <- df %>%
  mutate(
    # Base components
    Endo_Abs = `EndothelialCells [%]`,
    
    # Immune subsets as % of total sample
    across(all_of(immune_sub_cols), 
           ~ .x * (`ImmuneCells [%]` / 100), 
           .names = "{.col}_Abs"),
    
    # T-cell Lineages as % of total sample
    T_CD4_Abs = (`TCells CD4+ [%]` / 100) * (`TCells [%]` / 100) * `ImmuneCells [%]`,
    T_CD8_Abs = (`TCells CD8+ [%]` / 100) * (`TCells [%]` / 100) * `ImmuneCells [%]`,
    
    # Granular Clusters as % of total sample
    across(all_of(cd4_ct_cols), ~ .x * T_CD4_Abs, .names = "{.col}_Abs"),
    across(all_of(cd8_ct_cols), ~ .x * T_CD8_Abs, .names = "{.col}_Abs"),
    across(all_of(mye_ct_cols), ~ .x * (`MyeloidCells [%]` / 100) * `ImmuneCells [%]`, .names = "{.col}_Abs"),
    across(all_of(epi_ct_cols), ~ .x * `EpithelialCells [%]`, .names = "{.col}_Abs")
  )

# 3. Define the final mutually exclusive feature set 
# Note: "Other_Abs" and Fibroblasts are NOT included here
feat_cols <- c(
  "Endo_Abs",
  paste0(immune_sub_cols, "_Abs"),
  paste0(c(cd4_ct_cols, cd8_ct_cols, mye_ct_cols, epi_ct_cols), "_Abs")
)

# 4. Create the renormalized feat_mat (Sums to 100%)
feat_mat <- df_abs %>%
  # Sum only the selected granular components (excludes Fibro and Other)
  mutate(Normalization_Sum = rowSums(select(., all_of(feat_cols)))) %>%
  # Renormalize to 100
  mutate(across(all_of(feat_cols), ~ (.x / Normalization_Sum) * 100)) %>%
  select(`Sample ID`, all_of(feat_cols))

# Clean up column names
colnames(feat_mat) <- gsub("_Abs| \\[%\\]", "", colnames(feat_mat))

# 5. Impute Zeros with 2/3 of the minimum non-zero value
# We do this on the numeric columns only
numeric_cols <- colnames(feat_mat)[-1]
feat_values <- as.matrix(feat_mat[, numeric_cols])

min_nonzero <- min(feat_values[feat_values > 0], na.rm = TRUE)
feat_values[feat_values == 0] <- (2/3) * min_nonzero

# Final Matrix Construction
feat_mat[, numeric_cols] <- as.data.frame(feat_values)
# Remove Sample ID if you only want the numerical matrix
# feat_mat <- feat_mat[,-1] 

# Final Check
print(range(rowSums(feat_mat[, -1]))) # Should be exactly 100 to 100

feat_mat <- feat_mat[,-1]
```

```{r}
feat_mat <- clr(feat_mat)

plot_pca(feat_mat, labels = df$Tissue, cluster_score = FALSE, mod_score = FALSE, anosim_score = FALSE)
```

```{r}
# BiocManager::install("HDCytoData")
library(HDCytoData)
d_SE <- Krieg_Anti_PD_1_SE() # Returns SummarizedExperiment

d_SE@assays@data$exprs

# 1. Generate a contingency table directly from colData
# This counts cells per population_id for each sample_id
counts_table <- table(d_SE$sample_id, d_SE$population_id)

# 2. Convert to proportions (row-wise normalization)
# prop.table(x, 1) divides each cell by the row sum
feat_mat_pct <- prop.table(counts_table, margin = 1) * 100

# 3. Handle Imputation
min_nonzero <- min(feat_mat_pct[feat_mat_pct > 0])
feat_mat_pct[feat_mat_pct == 0] <- (2/3) * min_nonzero

# 4. Attach Sample-level Metadata
# Instead of distinct(), we can grab the unique mapping from the SE metadata
sample_info <- colData(d_SE) %>%
  as.data.frame() %>%
  select(sample_id, condition, patient_id) %>%
  unique()

# Final merge
feat_mat_final <- as.data.frame(feat_mat_pct) %>%
  rename(sample_id = Var1, cell_type = Var2, percentage = Freq) %>%
  pivot_wider(names_from = cell_type, values_from = percentage) %>%
  left_join(sample_info, by = "sample_id")

feat_mat <- clr(feat_mat)

plot_pca(feat_mat, labels = df$`Clinical Subtype`,
         cluster_score = FALSE, mod_score = FALSE, anosim_score = FALSE,invisible = "quali", n_ct_show = 5)
```

