---
title: "Zhang et al. dataset cell type annotation"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

Zhang Yuanyuan Cancer Cell 2021; CD45+ cells; primary or metastatic tumor tissues and peripheral blood of 22 advanced TNBC patients treated with paclitaxel or its combination with atezolizumab.

paper: https://www.sciencedirect.com/science/article/pii/S1535610821004992?via%3Dihub
accession:https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE169246

This script takes output of "scRNAseq_data_processing_ZhangY_2022_34653365.Rmd"

# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::restore()
# renv::snapshot()

source("functions.R")

library(scooter)

library(scGate)
library(UCell)
library(STACAS)
library(scGate)
library(ProjecTILs)
library(SignatuR)
library(HiTME)

# devtools::install_version("dlm", version = "1.1.5", repos = "http://cran.us.r-project.org")
# devtools::install_version("Rmagic", version = "2.0.3", repos = "http://cran.us.r-project.org")
library(reticulate)
library(Rmagic)
# devtools::install_github("inofechm/cutoff.scATOMIC", force = T)
library(cutoff.scATOMIC)
# install.magic("my_env")
# pymagic_is_available()
# devtools::install_github("abelson-lab/scATOMIC")
library(scATOMIC)
# Suppress scATOMIC warnings from reticulate
warnings <- import("warnings")
warnings$simplefilter("ignore")
library(R.utils)

`%notin%` <- Negate(`%in%`)
```


# Set paths
```{r}
my_seed <- 321
set.seed(my_seed)

path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
```


# Define datasets
```{r}
datasets <- list(
  "Adams" = list(
    file_path = file.path(path_nas, "AdamsT_2020_32832599", "output", "AdamsT_2020_32832599.rds"),
    subset_vars = list(
      OriginalAnnotationLevel1 = list(values = "Multiplet", op = "notin")
    ),
    tissue = "Lung",
    normal_tissue = TRUE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "Tissue", # Main biological grouping
      "Disease_Identity", # Additional biological groupings
      "OriginalAnnotationLevel1", "OriginalAnnotationLevel2" # Author annotations
    )
  ),
  
  "Bassez" = list(
    file_path = file.path(path_nas, "BassezA_2021_33958794", "output", "BassezA_2021_33958794whole.rds"),
    subset_vars = list(
      timepoint = list(values = "Pre", op = "in"),
      expansion = list(values = "n/a", op = "notin"),
      cohort = list(values = "treatment_naive", op = "in")
    ),
    tissue = "Breast tumor",
    normal_tissue_only = FALSE, # Whether the dataset contains only normal cells (= TRUE) or also cancer cells (= FALSE)
    sample_col = "sample", # Sample identifier
    meta_cols_keep = c(
      "sample",  # Sample identifier
      "timepoint", # Main biological grouping
      "expansion", "BC_type", "cohort", # Additional biological groupings
      "cellType", "cellSubType" # Author annotations
    )
  ),
  
  "GongSharma_age_females_cmvneg" = list(
    file_path = file.path(path_nas, "GongSharma_2024_PrePrintTBD", "output", "Gongsharma_age_females_cmvneg.h5ad"),
    tissue = "Blood",
    normal_tissue = TRUE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "sample.subjectAgeAtDraw", # Main biological grouping
      "AIFI_L1", "AIFI_L3" # Author annotations
    )
  ),
  
  "GongSharma_cmv_females" = list(
    file_path = file.path(path_nas, "GongSharma_2024_PrePrintTBD", "output", "Gongsharma_cmv_females.h5ad"),
    tissue = "Blood",
    normal_tissue = TRUE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "subject.cmv", # Main biological grouping
      "sample.subjectAgeAtDraw", # Additional biological groupings
      "AIFI_L1", "AIFI_L3" # Author annotations
    )
  ),
  
  "Kfoury" = list(
    file_path = file.path(path_nas, "Kfoury_2021_34719426", "output", "Kfoury_2021_34719426.rds"),
    tissue = "Bone marrow",
    normal_tissue = FALSE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "Status", # Main biological grouping
      "subject.status", # Additional biological groupings
      "cells" # Author annotations
    )
  ),
  
  "Lee" = list(
    file_path = file.path(path_nas, "LeeA_2021_34836966", "output", "LeeA_2021_34836966p_tumor_seurat.rds"),
    tissue = "Brain tumor",
    normal_tissue = FALSE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "condition", # Main biological grouping
      "MGMT", "IDH_status" # Additional biological groupings
    )
  ),
  
  "Pelka" = list(
    file_path = file.path(path_nas, "PelkaK_2022_34450029", "output", "PelkaK_2022_34450029whole.rds"),
    tissue = "Colon cancer",
    normal_tissue = FALSE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "Tissue", # Main biological grouping
      "HistologicTypeSimple", "TissueSite_detailed", "TissueSiteSimple", "Age", "TumorStage", "HistologicTypeSimple", "HistologicGrade_detailed", "HistologicGradeSimple", # Additional biological groupings
      "OriginalAnnotationLevel1",  "OriginalAnnotationLevel3" # Author annotations
    )
  ),
  
  "Smillie" = list(
    file_path = file.path(path_nas, "SmillieC_2019_31348891", "output", "SmillieC_2019_31348891.rds"),
    subset_vars = list(
      Tissue_compartment = list(values = "Lamina_propria", op = "in"),
      OriginalAnnotationLevel1 = list(values = "Immune", op = "in")
    ),
    tissue = "Colon",
    normal_tissue = TRUE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "Condition", # Main biological grouping
      "Tissue_compartment", # Additional biological groupings
      "OriginalAnnotationLevel1", "OriginalAnnotationLevel2", "OriginalAnnotationLevel3" # Author annotations
    )
  ),
  
  "Stephenson" = list(
    file_path = file.path(path_nas, "MitchelJ_2023_PrePrintTBD", "output", "MitchelJ_2023_PrePrintTBD_Stephenson_preprocessed.rds"),
    subset_vars = list(
      Site = list(values = "Ncl", op = "in"),
      Status = list(values = c("Healthy", "Covid"), op = "in"),
      Sample = list(values = c("BGCV10_CV0198", "MH8919230"), op = "notin")
    ),
    tissue = "Blood",
    normal_tissue = TRUE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "Status", # Main biological grouping
      "Status_on_day_collection_summary", "Days_from_onset", # Additional biological groupings
      "initial_clustering", "full_clustering" # Author annotations
    )
  ),
  
  "Wu" = list(
    file_path = file.path(path_nas, "WuS_2021_34493872", "output", "WuS_2021_34493872.rds"),
    tissue = "Breast tumor",
    normal_tissue = FALSE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "subtype", # Main biological grouping
      "celltype_major", "celltype_subset" # Author annotations
    )
  ),
  
  "Zhang" = list(
    file_path = file.path(path_nas, "ZhangY_2022_34653365", "output", "ZhangY_2022_34653365.rds"),
    subset_vars = list(
      Treatment = list(values = "Pre", op = "in"),
      Clinical.efficacy. = list(values = "-", op = "notin"),
      Sample = list(values = c("Pre_P012_b", "Pre_P013_t"), op = "notin")
    ),
    tissue = "Mixed",
    normal_tissue = TRUE, # Whether the dataset contains cancer cells or only normal cells
    sample_col = "Sample",
    meta_cols_keep = c(
      "Sample", # Sample identifier
      "Tissue", # Main biological grouping
      "Clinical.efficacy." # Additional biological groupings
    )
  )
)
```



# Load scGate models and ProjecTILs reference maps
```{r message=FALSE, warning=FALSE}
# Get scGate database of pre-defined gating models
scGate_models_DB <- get_scGateDB(branch = "dev", verbose = T, force_update = TRUE)
scGate_models_blood <- scGate_models_DB$human$PBMC
scGate_models_blood$MoMac <- scGate_models_blood$Monocyte
scGate_models_blood$Monocyte <- NULL
scGate_models_tumor <- scGate_models_DB$human$HiTME
# Need to have same model names in both lists for HiTME downstream processing)
scGate_models_blood <- c(scGate_models_blood, scGate_models_tumor[!names(scGate_models_tumor) %in% names(scGate_models_blood)])

# Load ProjecTILs reference maps
path_ref <- "/Users/christianhalter/Library/CloudStorage/OneDrive-SharedLibraries-UniversitédeGenève/Santiago Carmona - CSI/public/reference_atlases/sketched_200ct"
ref.maps_sketched <- list(CD8 = load.reference.map(file.path(path_ref, "sketched_CD8T_human_ref_v1.rds")),
                          CD4 = load.reference.map(file.path(path_ref, "sketched_CD4T_human_ref_v2.rds")),
                          DC = load.reference.map(file.path(path_ref, "sketched_DC_human_ref_v2.rds")),
                          MoMac = load.reference.map(file.path(path_ref, "sketched_MoMac_human_v1.rds"))
)
```


# Pre-process datasets
```{r}
# Number of highly variable features (genes) to select
n_hvg <- 2000

# Resolutions for seurat clustering
res <- c(0.4, 2, 5, 20)
### 0.4 (lowres ~ 10-15 cell types)
### 2 (medium ~ 30-50 cell types)
### 5 (~ 100 cell types)
### 20 (Ultrahires ~ 250-300 cell types)

HiTME_cols_keep <- c(
  "IFN_UCell",
  "HeatShock_UCell",
  "cellCycle.G1S_UCell",
  "cellCycle.G2M_UCell",
  "layer1", "layer2", "layer3")

scATOMIC_cols <- c(
  "layer_1",
  "layer_2",
  "layer_3",
  "layer_4",
  "layer_5",
  "layer_6",
  "scATOMIC_pred",
  "S.Score",
  "G2M.Score",
  "Phase",
  "classification_confidence" 
)


for (ds in names(datasets)) {
  input_file_path <- datasets[[ds]][["file_path"]]
  file_name_no_ending <- gsub(".rds|.h5ad", "", basename(input_file_path))
  processed_file_path <- file.path(getwd(), "data", paste0(file_name_no_ending, "_ECODAprocessed.rds"))
  
  if (!file.exists(processed_file_path)) {
    if (endsWith(input_file_path, ".rds")) {
      seurat <- readRDS(file = input_file_path)
    } else if (endsWith(input_file_path, ".h5ad")) {
      seurat <- load_h5ad_to_seurat(input_file_path)
    }
    
    
    # Remove dataset specific samples
    if (!is.null(names(datasets[[ds]][["subset_vars"]]))) {
      for (subset_var in names(datasets[[ds]][["subset_vars"]])) {
        subset_var_values <- datasets[[ds]][["subset_vars"]][[subset_var]][["values"]]
        op <- datasets[[ds]][["subset_vars"]][[subset_var]][["op"]]
        if (op == "in") {
          seurat <- subset(seurat, subset = !!sym(subset_var) %in% subset_var_values)
        } else if (op == "notin") {
          seurat <- subset(seurat, subset = !!sym(subset_var) %notin% subset_var_values)
        }
      }
    }
    
    # Drop unnecessary metadata cols
    meta_cols_keep <- datasets[[ds]][["meta_cols_keep"]]
    meta_cols_keep <- c(meta_cols_keep, HiTME_cols_keep, scATOMIC_cols)
    
    cols_keep <- intersect(meta_cols_keep, names(seurat@meta.data))
    seurat@meta.data <- seurat@meta.data[, cols_keep, drop = FALSE]
    
    # Remove low cell count samples
    seurat <- remove_low_cellcount_samples(seurat, min_cells_per_sample = 500)
    
    # Add “Sample” metadata columns
    seurat[["Sample"]] <- seurat[[datasets[[ds]][["sample_col"]]]]
    
    # Sample names starting with digits are not allowed in seurat
    sample_names_starting_with_digit <- grepl("^\\d", unique(seurat[["Sample"]]))
    if (any(sample_names_starting_with_digit)) {
      seurat[["Sample"]][grepl("^\\d", seurat[["Sample"]])] <- paste0("g", seurat[["Sample"]][grepl("^\\d", seurat[["Sample"]])])
    }
    
    
    # Ensure it's a seurat v5 object
    seurat <- create_clean_seuratv5_object(seurat)
    
    # If not file exists: Save as h5ad
    convert_seurat_to_h5ad(seurat, ds)
    
    
    # Standard Seurat preprocessing part 1
    seurat <- seurat |>
      NormalizeData() |>
      FindVariableFeatures(nfeatures = n_hvg)
    
    
    # Annotate cells
    
    obj.list <- SplitObject(seurat, split.by = "Sample")
    rm(seurat)
    gc()

    ## scAtomic
    
    for (i in 1:length(obj.list)) {
      print(paste0(ds, " Sample ", i, " / ", length(obj.list)))
      
      start_time <- Sys.time()  # Start measuring time
      
      # On a MacBook Pro with M1 chip, it takes roughly a total of 2.5 min (= 2.5 * 60 sec) to process a sample with 10'000 cells
      # So, set a timeout of twice that
      timeout <- dim(obj.list[[i]])[2] / 10000 * 2.5 * 60 * 2
      
      # Retry 3 times
      for (a in 1:3) {
        cat(paste("Attempt to run scATOMIC", a, "...\n"))
        result <- tryCatch({
          # The 'withTimeout' function will stop the execution after 5 seconds
          withTimeout({
            invisible(capture.output(sca_preds <- run_scATOMIC(obj.list[[i]]@assays$RNA$counts)))
            suppressMessages({
              suppressWarnings({
                invisible(capture.output(
                  sca_results <- create_summary_matrix(prediction_list = sca_preds,
                                                       raw_counts = obj.list[[i]]@assays$RNA$counts,
                                                       normal_tissue = datasets[[ds]][["normal_tissue"]])
                ))
              })
            })
            return("Complete")
          }, timeout = timeout)
        }, TimeoutException = function(ex) {
          # This block runs if a timeout occurs
          cat("  Timeout occurred, retrying...\n")
          return(NULL) # Return NULL to signify a failure
        })
        
        # If the result is not NULL, the function finished successfully
        if (!is.null(result)) {
          cat(paste("  Success on attempt", a, "with result:", result, "\n"))
          result <- NULL
          break # Exit the loop
        }
      }
      
      # Drop unnecessary metadata columns
      cols_keep <- intersect(meta_cols_keep, names(sca_results))
      sca_results <- sca_results[, cols_keep, drop = FALSE]
      
      obj.list[[i]] <- AddMetaData(obj.list[[i]], sca_results)
      
      end_time <- Sys.time()  # End measuring time
      execution_time <- end_time - start_time
      cat(paste("Sample", i, "completed in", "\n"))
      print(round(execution_time, 1))
    }

    
    ## HiTME
    while (length(obj.list) > 0) {
      print(paste(ds, "Samples remaining:", length(obj.list)))
      
      obj <- obj.list[[1]]
      obj.list <- obj.list[-1]
      
      # On a MacBook Pro with M1 chip, it takes roughly 60 sec to process a sample with 10'000 cells
      # The timeout is per process (so ~ 60 / 6 = ~ 10 sec)
      # So, giving 60 sec per 10'000 cells for a process/core to finish is plenty
      timeout <- min(20, dim(obj)[2] / 10000 * 60)
      start_time <- Sys.time()  # Start measuring time
      
      # Retry 3 times
      for (a in 1:3) {
        cat(paste("Attempt to Run.HiTME", a, "...\n"))
        result <- tryCatch({
          if (datasets[[ds]][["tissue"]] == "Blood") {
            obj <- Run.HiTME(object = obj,
                             scGate.model = scGate_models_blood,
                             ref.maps = ref.maps_sketched,
                             verbose = FALSE,
                             bparam = BiocParallel::MulticoreParam(workers = parallelly::availableCores() - 2,
                                                                   progressbar = TRUE,
                                                                   timeout = timeout))
          } else if (datasets[[ds]][["tissue"]] == "Mixed") {
            if (obj$Tissue[1] == "blood") {
              obj <- Run.HiTME(object = obj,
                               scGate.model = scGate_models_blood,
                               ref.maps = ref.maps_sketched,
                               verbose = FALSE,
                               bparam = BiocParallel::MulticoreParam(workers = parallelly::availableCores() - 2,
                                                                     progressbar = TRUE,
                                                                     timeout = timeout))
            }
          } else {
            obj <- Run.HiTME(object = obj,
                             scGate.model = scGate_models_tumor,
                             ref.maps = ref.maps_sketched,
                             verbose = FALSE,
                             bparam = BiocParallel::MulticoreParam(workers = parallelly::availableCores() - 2,
                                                                   progressbar = TRUE,
                                                                   timeout = timeout))
          }
          return("Complete")
        }, error = function(ex) {
          # This block runs if a timeout occurs
          cat("  Timeout occurred, retrying...\n")
          return(NULL) # Return NULL to signify a failure
        })
        
        # If the result is not NULL, the function finished successfully
        if (!is.null(result)) {
          cat(paste("  Success on attempt", a, "with result:", result, "\n"))
          result <- NULL
          break # Exit the loop
        }
      }
      
      end_time <- Sys.time()  # End measuring time
      execution_time <- end_time - start_time
      cat(paste("Sample", i, "completed in", "\n"))
      print(round(execution_time, 1))
      
      # Drop unnecessary metadata cols
      cols_keep <- intersect(meta_cols_keep, names(obj@meta.data))
      obj@meta.data <- obj@meta.data[, cols_keep, drop = FALSE]
      
      if (!exists("seurat")) {
        seurat <- obj
      } else {
        seurat <- merge(seurat, obj)
      }
    }
    
    seurat <- JoinLayers(seurat)
    
    
    # Standard Seurat preprocessing part 2
    seurat <- seurat |>
      ScaleData() |>
      RunPCA(dims = 1:50, verbose = FALSE) |>
      FindNeighbors()
    for (r in res) {
      seurat <- seurat |>
        FindClusters(resolution = r)
    }
    
    # Save processed Seurat object
    saveRDS(seurat, file = processed_file_path)
  }
}


```








# Celltype annotation with HiTME

```{r}
# Get scGate database of pre-defined gating models
scGate_models_DB <- get_scGateDB(branch = "dev", verbose = T, force_update = TRUE)
scGate_models_blood <- scGate_models_DB$human$PBMC
scGate_models_blood$MoMac <- scGate_models_blood$Monocyte
scGate_models_blood$Monocyte <- NULL
scGate_models_tumor <- scGate_models_DB$human$HiTME
# Need to have same model names in both lists for HiTME downstream processing)
scGate_models_blood <- c(scGate_models_blood, scGate_models_tumor[!names(scGate_models_tumor) %in% names(scGate_models_blood)])

# Load ProjecTILs reference maps
path_ref <- "~/Dropbox/CSI/reference_atlases/sketched_200ct"
ref.maps_sketched <- list(CD8 = load.reference.map(file.path(path_ref, "sketched_CD8T_human_ref_v1.rds")),
                          CD4 = load.reference.map(file.path(path_ref, "sketched_CD4T_human_ref_v2.rds")),
                          DC = load.reference.map(file.path(path_ref, "sketched_DC_human_ref_v2.rds")),
                          MoMac = load.reference.map(file.path(path_ref, "sketched_MoMac_human_v1.rds"))
)

obj.list <- SplitObject(obj, split.by = "Sample")
rm(obj)
gc()

for (i in 1:length(obj.list)) {
  print(i)
  if (obj.list[[i]]$Tissue[1] == "blood") {
    obj.list[[i]] <- Run.HiTME(object = obj.list[[i]],
                               scGate.model = scGate_models_blood,
                               ref.maps = ref.maps_sketched
    )
  } else {
    obj.list[[i]] <- Run.HiTME(object = obj.list[[i]],
                               scGate.model = scGate_models_tumor,
                               ref.maps = ref.maps_sketched
    )
  }
}

rm(i, ref.maps_sketched)
gc()

save_objs(obj.list, path_output_samples)
```

```{r}
annotate_samples_HiTME <- function(
    seurat,
    blood_samples,
    nonblood_samples,
    scGate_version = "latest",
    scGate_branch = "dev") {
  
  
}
```

