---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis


# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
library(rstatix)

source("functions.R")

# Set paths
root <- "/Users/christianhalter/Library/CloudStorage/OneDrive-SharedLibraries-UniversitédeGenève/Santiago Carmona - CSI/LabPapers/ECODA"
path_plots <- file.path(root, "plots")
path_data <- file.path(getwd(), "data")
```


# Create results list to fill

```{r}
factors_test <- c(1, 2, 3, 5, 10)

result_file <- "result_list.rds"

if (file.exists(result_file)) {
  result_list <- readRDS("result_list.rds")
} else {
  result_list <- list()
}

analysis_elems <- c("bmark", "trans", "zeroimp")
```


# ---------------------------------------------------------------

# Zhang
Was annotated with file Zhang_celltype_annotation.Rmd

```{r}
# Dataset
ds <- "Zhang"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  # Set paths
  ds_name <- "ZhangY_2022_34653365"
  
  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")
  path_output_samples <- file.path(path_output, "samples")
  
  
  # Load data
  obj.list <- read_objs(path_output_samples)
  
  # Exclude samples that consist mainly of one cell type as outliers
  samples_exclude <- c("Pre_P012_b", "Pre_P013_t")
  obj.list <- obj.list[!names(obj.list) %in% samples_exclude]
  
  samples_clearcut <- c()
  
  for (i in names(obj.list)) {
    if (obj.list[[i]]@meta.data$Treatment[1] %in% c("Pre") &
        !(obj.list[[i]]@meta.data$Clinical.efficacy.[1] %in% c("-"))) {
      samples_clearcut <- c(samples_clearcut, i)
    }
  }
  
  metacols_to_keep <- c("Sample", "Tissue", "layer1", "layer2")
  
  for (i in samples_clearcut) {
    print(i)
    
    obj <- obj.list[[i]]
    Idents(obj) <- "Sample"
    obj@meta.data <- obj@meta.data[colnames(obj@assays$RNA$counts), metacols_to_keep]
    
    if (i == samples_clearcut[1]) {
      seurat <- obj
    } else {
      seurat <- merge(seurat, obj)
    }
  }
  
  rm(obj.list); gc()
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)
  
  
  seurat@misc$label_col <- "Tissue"
  seurat@misc$low_res_ct_col <- "layer1"
  seurat@misc$hi_res_ct_col <- "layer2"
  
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```


# ---------------------------------------------------------------

# Stephenson

```{r}
ds <- "Stephenson"


elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  ds_name <- "MitchelJ_2023_PrePrintTBD"
  
  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")
  
  seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))
  
  seurat <- subset(seurat, subset = Site == "Ncl")
  seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))
  seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
  seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
  seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
  seurat <- subset(seurat, subset = Days_from_onset < 10)
  
  seurat$sample_id <- as.character(seurat$sample_id)
  samples_exclude <- c("BGCV10_CV0198", "MH8919230")
  samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% samples_exclude])
  seurat <- subset(seurat, subset = sample_id %in% samples_subset)
  
  seurat$Sample <- seurat$sample_id
  
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)
  

  # # Convert Seurat object to anndata for interoperability with python
  # object <- CreateSeuratObject(counts = seurat@assays$RNA$counts, meta.data = seurat@meta.data[, c("Sample", "initial_clustering")])
  # object[["RNA3"]] <- as(object = object[["RNA"]], Class = "Assay")
  # DefaultAssay(object) <- "RNA3"
  # object[["RNA"]] <- NULL
  # object <- RenameAssays(object = object, RNA3 = 'RNA')
  # convert_seurat_to_h5ad(object, ds)
  # rm(object); gc()
  

  
  seurat@misc$label_col <- "Status"
  seurat@misc$low_res_ct_col <- "initial_clustering"
  seurat@misc$hi_res_ct_col <- "full_clustering"

  
    
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```



# ---------------------------------------------------------------

# Gong and Sharma

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## CMV (females)

```{r}
ds <- "GongSharma_cmv_females"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  seurat <- load_seurat_object("data/gongsharma_cmv_females.h5ad")
  seurat$Sample <- seurat$specimen.specimenGuid
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)

  seurat@misc$label_col <- "subject.cmv"
  seurat@misc$low_res_ct_col <- "AIFI_L1"
  seurat@misc$hi_res_ct_col <- "AIFI_L3"
    
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```


## Age (females cmvneg)

```{r}
ds <- "GongSharma_age_females_cmvneg"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  seurat <- load_seurat_object("data/gongsharma_age_females_cmvneg.h5ad")
  seurat$Sample <- seurat$specimen.specimenGuid
  
    
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)
  
  seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))

  seurat@misc$label_col <- "age"
  seurat@misc$low_res_ct_col <- "AIFI_L1"
  seurat@misc$hi_res_ct_col <- "AIFI_L3"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```



# ---------------------------------------------------------------

# Lee

```{r}
ds <- "Lee"


elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  ds_name <- "LeeA_2021_34836966"
  
  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")
  
  seurat <- readRDS(file.path(path_output, sprintf("%sp_tumor_seurat.rds", ds_name)))
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)

  # # Convert Seurat object to anndata for interoperability with python
  # object <- seurat
  # object[["RNA3"]] <- as(object = object[["RNA"]], Class = "Assay")
  # DefaultAssay(object) <- "RNA3"
  # object[["RNA"]] <- NULL
  # object <- RenameAssays(object = object, RNA3 = 'RNA')
  # SaveH5Seurat(object, filename = "lee.h5Seurat")
  # Convert("lee.h5Seurat", dest = "h5ad")
  # rm(object); gc()
  
  seurat@misc$label_col <- "condition"
  seurat@misc$low_res_ct_col <- "scGate_multi"
  seurat@misc$hi_res_ct_col <- "functional.cluster"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```



# ---------------------------------------------------------------

# Pelka

## Has also other non-immune cells

```{r}
ds <- "Pelka"


elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))

  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  ds_name <- "PelkaK_2022_34450029"

  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")

  seurat <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name)))
  
  seurat$Sample <- as.character(seurat$Sample)

  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)


  seurat@misc$label_col <- "Tissue"
  seurat@misc$low_res_ct_col <- "OriginalAnnotationLevel1"
  seurat@misc$hi_res_ct_col <- "OriginalAnnotationLevel3"

  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```



# ---------------------------------------------------------------

# Bassez

<!-- ```{r} -->
<!-- ds <- "Bassez" -->

<!-- elem_null <- c() -->
<!-- for (a in analysis_elems) { -->
<!--   elem_null <- c(elem_null, is.null(result_list[[a]][[ds]])) -->

<!--   if (is.null(result_list[[a]][[ds]])) { -->
<!--     print(paste("To be analyzed:", a, ds)) -->
<!--   } -->
<!-- } -->

<!-- # If any analyses missing, process dataset for missing analyses -->
<!-- if (any(elem_null)) { -->

<!--   ds_name <- "BassezA_2021_33958794" -->

<!--   path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name) -->
<!--   path_output <- file.path(path_root, "output") -->

<!--   seurat <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name))) -->

<!--   seurat <- subset(seurat, subset = timepoint == "Pre") -->

<!--   seurat$Sample <- seurat$sample -->
<!--   seurat$cellSubType[is.na(seurat$cellSubType)] <- seurat$cellType[is.na(seurat$cellSubType)] -->

<!--   # Convert Seurat object to anndata for interoperability with python -->
<!--   mrvi_dist_file_path <- file.path("~/Dropbox/CSI/PAPERS/ECODA/data/mrvi_dists") -->
<!--   mrvi_dist_file <- file.path(mrvi_dist_file_path, paste0(ds, "_dists.nc")) -->
<!--   if (!file.exists(mrvi_dist_file)) { -->
<!--     convert_seurat_to_h5ad(seurat, ds) -->
<!--     stop("Please process MrVI to dist file") -->
<!--   } -->

<!--   seurat <- seurat |> NormalizeData() |> FindVariableFeatures(nfeatures = 2000) -->
<!--   varf <- seurat@assays[["RNA"]]@var.features -->
<!--   hvg <- varf[!is.na(varf)] -->



<!--   seurat@misc$label_col <- "cohort" -->
<!--   seurat@misc$low_res_ct_col <- "cellType" -->
<!--   seurat@misc$hi_res_ct_col <- "cellSubType" -->

<!--   result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test) -->
<!-- } -->
<!-- ``` -->




# ---------------------------------------------------------------

# Wu

```{r}
ds <- "Wu"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  ds_name <- "WuS_2021_34493872"
  
  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")
  
  seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)
  

  # # Convert Seurat object to anndata for interoperability with python
  # object <- seurat
  # object[["RNA3"]] <- as(object = object[["RNA"]], Class = "Assay")
  # DefaultAssay(object) <- "RNA3"
  # object[["RNA"]] <- NULL
  # object <- RenameAssays(object = object, RNA3 = 'RNA')
  # convert_seurat_to_h5ad(seurat, ds)
  # rm(object); gc()
  
  
  seurat@misc$label_col <- "subtype"
  seurat@misc$low_res_ct_col <- "celltype_major"
  seurat@misc$hi_res_ct_col <- "celltype_subset"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```




# ---------------------------------------------------------------

# Smillie

```{r}
ds <- "Smillie"


elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  ds_name <- "SmillieC_2019_31348891"
  
  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")
  
  
  seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))
  
  stot <- seurat
  # seurat <- subset(stot, subset = Tissue_compartment == "Epithelial")
  # seurat <- subset(stot, subset = Tissue_compartment == "Lamina_propria")
  seurat <- subset(stot, subset = stot$OriginalAnnotationLevel1 == "Immune")
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)
  
  
  seurat@misc$label_col <- "Tissue_compartment"
  seurat@misc$low_res_ct_col <- "OriginalAnnotationLevel2"
  seurat@misc$hi_res_ct_col <- "OriginalAnnotationLevel3"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```



# ---------------------------------------------------------------

# Adams
## Has also other non-immune cells

```{r}
ds <- "Adams"


elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  ds_name <- "AdamsT_2020_32832599"
  
  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")
  
  seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))
  
  seurat <- subset(seurat, subset = OriginalAnnotationLevel1 != "Multiplet")
  
  seurat$Sample <- paste0("g", seurat$Sample)
  # seurat$Sample <- gsub("-", "_" seurat$Sample)
  Idents(seurat) <- "Sample"
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)

  
  seurat@misc$label_col <- "Condition"
  seurat@misc$low_res_ct_col <- "OriginalAnnotationLevel1"
  seurat@misc$hi_res_ct_col <- "OriginalAnnotationLevel2"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```



# ---------------------------------------------------------------

# Kfoury
## Has also other non-immune cells


```{r}
ds <- "Kfoury"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  ds_name <- "Kfoury_2021_34719426"
  
  path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
  path_output <- file.path(path_root, "output")
  
  seurat <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))
  
  seurat@misc$label_col <- "Status"
  seurat@misc$low_res_ct_col <- "cells"
  seurat@misc$hi_res_ct_col <- "cells"
  
  # Convert Seurat object to anndata for interoperability with python
  convert_seurat_to_h5ad(seurat, ds)
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, factors_test)
}
```

## Test on immune cells only

```{r}
# cts <- unique(seurat$OriginalAnnotationLevel1)
# cts <- cts[!cts %in% c("Osteoblasts", "Osteoclasts", "Pericytes", "Endothelial", "Tumor", "Erythroid")]
# seurat_sub <- subset(seurat, subset = OriginalAnnotationLevel1 %in% cts)
# 
# test <- run_benchmark_analysis(seurat_sub,
#                                get_labels(seurat, seurat@misc$label_col),
#                                low_res_ct_col,
#                                hi_res_ct_col,
#                                factors_test = 2,
#                                Pseudobulk = TRUE,
#                                ECODA_deconv = FALSE,
#                                ECODA_low_res = FALSE,
#                                ECODA_high_res = TRUE,
#                                Pseudobulk_PCA = FALSE,
#                                ECODA_high_res_PCA = FALSE,
#                                MOFA = FALSE,
#                                scITD = FALSE,
#                                MrVI = FALSE,
#                                show_pca_plots = TRUE,
#                                save_pca_plots = FALSE
# )
```










# ---------------------------------------------------------------

# Test new dataset (template)

```{r}
# seurat_sub <- subset(seurat, subset = OriginalAnnotationLevel1 %in% c("B", "Mast", "Myeloid", "Plasma", "TNKILC"))
# 
# test <- run_benchmark_analysis(seurat_sub,
#                                get_labels(seurat_sub, seurat_sub@misc$label_col),
#                                seurat@misc$low_res_ct_col,
#                                seurat@misc$hi_res_ct_col,
#                                factors_test = 2,
#                                Pseudobulk = TRUE,
#                                ECODA_deconv = FALSE,
#                                ECODA_low_res = FALSE,
#                                ECODA_high_res = TRUE,
#                                Pseudobulk_PCA = FALSE,
#                                ECODA_high_res_PCA = FALSE,
#                                MOFA = FALSE,
#                                scITD = FALSE,
#                                MrVI = FALSE,
#                                show_pca_plots = TRUE,
#                                save_pca_plots = FALSE
# )
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot benchmark results

## Save and flatten the nested list into a DataFrame

```{r}
bmark_result_list <- result_list[["bmark"]]

df_results <- imap_dfr(bmark_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    if ("scores" %in% names(.x)) {
      imap_dfr(.x$scores, ~ {
        tibble(
          dataset = dataset_name,
          method = method_name,
          score = .y,
          value = .x
        )
      })
    } else {
      tibble() # Return an empty tibble if "scores" is not found
    }
  })
})
```



## Plot results

```{r fig.height=10, fig.width=14}
levs <- sort(as.character(unique(df_results$method)))
levs <- c(
  "ECODA_high_res",
  "ECODA_high_res_2_PCA_dims",
  "ECODA_high_res_5_PCA_dims",
  "ECODA_high_res_10_PCA_dims",
  "MrVI",
  "Pseudobulk",
  "Pseudobulk_2_PCA_dims",
  "Pseudobulk_5_PCA_dims",
  "Pseudobulk_10_PCA_dims",
  "MOFA_2_factors",
  "MOFA_5_factors",
  "MOFA_10_factors",
  "scITD_2_factors",
  "scITD_5_factors",
  "scITD_10_factors",
  "ECODA_low_res",
  "ECODA_deconv"
)

df_results$method <- factor(df_results$method, levels = levs)

df_results$dataset[df_results$dataset == "GongSharma_females_cmv"] <- "GongSharma_cmv"
df_results$dataset[df_results$dataset == "GongSharma_females_cmvneg_age"] <- "GongSharma_age"

df_results_clustering <- df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

# df_results_clustering <- df_results_clustering %>%
#   arrange(desc(mean))
df_results_clustering$method <- factor(df_results_clustering$method, levels = df_results_clustering$method)

df_results$method <- factor(df_results$method, levels = df_results_clustering$method)


df_results_modularity <- df_results %>%
  filter(score == "mod_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_modularity <- df_results_modularity %>%
  arrange(desc(mean))
df_results_modularity$method <- factor(df_results_modularity$method, levels = df_results_modularity$method)



# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "Silhouette score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            labeller = labeller(score = score.labs, dataset = datas.labs),
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "benchmark_detailed_barplot.pdf"),
       width = 10,
       height = 6)


# df_results %>%
#   filter(score == "mod_score" | score == "cluster_score") %>%
#   group_by(score, method) %>%
#   summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
#   ungroup() %>%
#   
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   facet_grid(scales = "free",
#              rows = vars(score),
#              labeller = labeller(score = score.labs))
# 
# ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_mod_clust.pdf"),
#        width = 2.5,
#        height = 4)



df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_bars.pdf"),
       width = 5,
       height = 7)


ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores.pdf"),
       width = 7,
       height = 7)


# 
# df_results_clustering %>%
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Adjusted Rand Index") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   NoLegend()


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

df_results_clustering <- df_results_clustering[!grepl("2|10|PCA", df_results_clustering$method), ]

for (ds in unique(df_results_clustering$dataset)) {
  number_cts <- dim(result_list$bmark[[ds]]$ECODA_high_res$feat_mat)[2]
  df_results_clustering[df_results_clustering$dataset == ds, "number_celltypes"] <- number_cts
}

# df_results_clustering$dataset <- paste0(df_results_clustering$dataset, " (", df_results_clustering$number_celltypes, ")")
# legend_title <- "Dataset (number of cell types)"
legend_title <- "Dataset"

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_clustering)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)


my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "Pseudobulk"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(3, 6, 4, 5, 2, 1)]


ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    comparisons = my_comparisons,
    # ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 15,
    p.adjust.method = "none",
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))) + # Add pairwise comparisons p-value
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # stat_compare_means(label.y = 1) +     # Add global p-value
  labs(x = "", y = "Adjusted Rand Index") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #   y.position = c(max(df_results_clustering$values)+0.1, max(df_results_clustering$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  # ylim(c(-0.1, 1.1)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(group = legend_title, linetype = legend_title, shape = legend_title) +
  theme(axis.text.x = element_text(angle = 55, vjust = 1, hjust=1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_clustering.pdf"),
       width = 7,
       height = 5)



df_results_mod <- df_results %>%
  filter(score == "mod_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_mod)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)



ggplot(df_results_mod, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    comparisons = my_comparisons,
    # ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 15,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
  ) + # Add pairwise comparisons p-value
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_mod$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # stat_compare_means(label.y = 1) +     # Add global p-value
  labs(x = "", y = "Modularity score") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #   y.position = c(max(df_results_clustering$values)+0.1, max(df_results_clustering$values)+0.2), hide.ns = TRUE) +
  theme_minimal() +
  # ylim(c(-0.12, 2)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity.pdf"),
       width = 7,
       height = 6)

# df_results_modularity %>%
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Modularity") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   NoLegend()
# ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity.pdf"),
#        width = 2.5,
#        height = 2.5)

```

## Show example PCAs for publication

```{r fig.height=6, fig.width=12}

levs <- c(
  "ECODA_high_res",
  "ECODA_low_res",
  "ECODA_deconv",
  "Pseudobulk",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI"
)


plot_list <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_cmv_females.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_age_females_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Lee"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Lee"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Lee.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Zhang"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Zhang"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Zhang.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Stephenson.pdf"),
       plot = p,
       width = 15,
       height = 7)

```


# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot transformation results

## Plot results

```{r}
df_results <- Reduce(rbind, result_list[["trans"]])

df_results <- df_results %>%
  pivot_longer(
    cols = c(Silhouette_score, Modularity_score, Adjusted_Rand_Index),
    names_to = "score_method",
    values_to = "values"
  ) %>%
  as.data.frame()

df_results_clustering <- df_results %>%
  filter(score_method == "Adjusted_Rand_Index") %>%
  group_by(method) %>%
  summarize(mean = mean(values), sem = sd(values) / sqrt(length(values)))

df_results_clustering <- df_results_clustering %>%
  arrange(desc(mean))

df_results_clustering$method <- factor(df_results_clustering$method, levels = df_results_clustering$method)
df_results$method <- factor(df_results$method, levels = df_results_clustering$method)



# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "Silhouette score")
names(score.labs) <- sort(unique(df_results$score_method))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))



ggplot(df_results, aes(x = method, y = values)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score_method ~ dataset,
             scales = "free",
             labeller = labeller(score_method = score.labs, dataset = datas.labs))
# facet_grid(scales = "free",
#            rows = vars(score_method), cols = vars(dataset))
ggsave(file.path(path_plots, "transformations_real_datasets_detailed_barplot.pdf"),
       width = 8,
       height = 7)



# df_results %>%
#   group_by(score_method, method) %>%
#   summarize(mean = mean(values), sem = sd(values) / sqrt(length(values))) %>%
#   ungroup() %>%
#   
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Separation score") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   facet_grid(scales = "free",
#              rows = vars(score_method),
#              labeller = labeller(score_method = score.labs))


ggplot(df_results, aes(x = method, y = values)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score_method),
             labeller = labeller(score_method = score.labs))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores.pdf"),
       width = 5,
       height = 5)



# df_results_clustering %>%
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Adjusted Rand Index") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   NoLegend()


df_results_clustering <- df_results %>%
  filter(score_method == "Adjusted_Rand_Index")

df_results_clustering <- df_results_clustering[!grepl("pca", df_results_clustering$method), ]

# Example assuming 'data' is your dataframe
anova_result <- aov(values ~ method, data = df_results_clustering)
summary(anova_result)
stat.test <- tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)
# stat.test <- stat.test[["method"]] %>% as.data.frame()
# 
# # Extract row names where p adj < 0.05
# significant_comparisons <- rownames(stat.test)[stat.test[["p adj"]] < 0.05]
# 
# # Initialize an empty list to store results
# my_comparisons <- list()
# 
# # Split and store each comparison in the list
# for (comparison in significant_comparisons) {
#   parts <- strsplit(comparison, "-")[[1]]
#   my_comparisons <- c(my_comparisons, list(parts))
# }

my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "freq"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
# my_comparisons <- c(my_comparisons, list(c("clr", "arcsine_sqrt")))
my_comparisons <- my_comparisons[c(4, 3, 2, 1, 5)]

# my_comparisons <- list(
#   c("alr_mincvref", "freq"),
#   c("alr_randref", "freq"),
#   c("arcsine_sqrt", "freq"),
#   c("clr", "freq"),
#   c("clr_pca", "freq"),
#   c("counts", "freq")
#   )

ggplot(df_results_clustering, aes(x = method, y = values)) +
  stat_compare_means(
    comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    # ref.group = "freq",
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Adjusted Rand Index") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #                    y.position = c(max(df_results_clustering$values)+0.1, max(df_results_clustering$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot.pdf"),
       width = 6,
       height = 4.5)

```

# Statistical comparison

```{r}
df_results_clustering <- df_results %>%
  filter(score_method == "Adjusted_Rand_Index")

# Example assuming 'data' is your dataframe
anova_result <- aov(values ~ method, data = df_results_clustering)
summary(anova_result)

TukeyHSD(aov(df_results_clustering$values ~ df_results_clustering$method))
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot zero imputation results

```{r}
zeroimp_result_list <- result_list[["zeroimp"]]
```


## Plot results

```{r}
df_results <- imap_dfr(zeroimp_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    imap_dfr(.x, ~ {
      tibble(
        dataset = dataset_name,
        method = method_name,
        score = .y,
        value = .x
      )
    })
  })
})

df_results <- df_results[df_results$method != "impRZilr", ]


# 
# df_results <- bind_rows(res_list, .id = "test_name") %>%
#   mutate(test_name = as.character(test_name)) %>%
#   dplyr::select(test_name, sil_score, mod_score, cluster_score)
# 
# strings <- df_results$test_name
# 
# # Remove the first part (split at "_") and the last part (split at "\\.")
# first_parts <- sapply(str_split(strings, "_"), "[[", 1)
# first_parts[grepl("gongsharma", first_parts)] <- paste0(first_parts[grepl("gongsharma", first_parts)], "_", sapply(str_split(strings[grepl("gongsharma", strings)], "_"), "[[", 2))
# df_results$dataset <- first_parts
# 
# # Remove the first part (split at "_") and the last part (split at "\\.")
# middle_parts <- str_remove(strings, "^[^_]*_")  # Remove everything up to and including the first "_"
# middle_parts[grepl("cmv", middle_parts)] <- str_remove(middle_parts[grepl("cmv", middle_parts)], "^[^_]*_")  # Remove everything up to and including the first "_"
# df_results$imputation_method <- middle_parts
# 
# df_results <- df_results %>%
#   pivot_longer(
#     cols = c(sil_score, mod_score, cluster_score),
#     names_to = "score_method",
#     values_to = "values"
#   ) %>%
#   as.data.frame()


# df_results$dataset[df_results$dataset == "gongsharma_allcmv"] <- "GongSharma_cmv"
# df_results$dataset[df_results$dataset == "gongsharma_agefemalescmvneg"] <- "GongSharma_age"
# 
# 
# levels <- names(res_list)[grepl("Zhang", names(res_list))]
# levels <- sub("Zhang_", "", levels)
# 
# df_results$method <- factor(df_results$method, levels = levels)


# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "Silhouette score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "zero_imputation_real_datasets_detailed_barplot.pdf"),
       width = 10,
       height = 6)




df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%

  ggplot(aes(x = method, y = mean)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_barplot_separatescores.pdf"),
       width = 7,
       height = 6)

ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_separatescores.pdf"),
       width = 7,
       height = 6)



df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()

df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "counts_all_1"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(6,1,7:10, 2:5)]
# my_comparisons[[length(my_comparisons) + 1]] <- c("perc_all_0.1%", "perc_all_1%")

ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_clusterscore.pdf"),
       width = 8,
       height = 6)

df_results_modularity <- df_results %>%
  filter(score == "mod_score")

ggplot(df_results_modularity, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_modscore.pdf"),
       width = 8,
       height = 6)

```

# Test for statistical significance

To few data points for statistical significance

```{r}
library(rstatix)

df_results_cluster <- df_results %>% filter(score == "cluster_score")

outlier <- df_results_cluster %>%
  group_by(method) %>%
  rstatix::identify_outliers(value)
data.frame(outlier)

normality <- df_results_cluster %>%
  group_by(method) %>%
  rstatix::shapiro_test(value)
data.frame(normality)
cat(paste("All groups normally ditributed?", all(normality$p > 0.05),
            "\nNot normally distributed:", normality$method[normality$p <= 0.05]))

# res <- rstatix::anova_test(data = df_results_cluster, dv = value, wid = dataset, within = method) 
# rstatix::get_anova_table(res)
# 
# pair_t <- df_results_cluster %>%
#   rstatix::pairwise_t_test(value~method, paired=TRUE, p.adjust.method = "BH" )
# data.frame(pair_t)

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)




df_results_cluster <- df_results %>% filter(score == "mod_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)



df_results_cluster <- df_results %>% filter(score == "sil_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)
```



# ---------------------------------------------------------------
# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Brain from Macnair et al.

snRNA-seq stratifies multiple sclerosis patients into distinct white matter glial responses

Data from:
https://www.sciencedirect.com/science/article/pii/S0896627324008730#sec8



## Load data

```{r}
library(readxl)

df_mac_main <- data.frame(fread(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "MacnairW_2025_Neuron", "ms_lesions_snRNAseq_col_data_2023-09-12.txt.gz")))
df_mac_val <- data.frame(fread(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "MacnairW_2025_Neuron", "ms_lesions_snRNAseq_validation_col_data_2023-09-12.txt.gz")))

feat_mat <- as.data.frame.matrix(table(df_mac_main$sample_id_anon, df_mac_main$type_fine)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_mac_main %>%
    group_by(sample_id_anon) %>%
    slice(1)
row.names(meta) <- meta$sample_id_anon

p <- plot_pca(feat_mat, labels = meta$matter)


rsub <- meta$matter == "GM"
feat_mat_gm <- feat_mat[rsub, ]
meta_gm <- meta[rsub, ]

p <- plot_pca(feat_mat_gm, labels = meta_gm$lesion_type)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$seq_pool)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$sex)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$sample_source)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$diagnosis)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$age_cat)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$pmi_cat)


rsub <- meta$matter == "WM"
feat_mat_wm <- feat_mat[rsub, ]
meta_wm <- meta[rsub, ]

p <- plot_pca(feat_mat_wm, labels = meta_wm$lesion_type)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$seq_pool)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$sex)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$sample_source)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$diagnosis)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$age_cat)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$pmi_cat)
```


# ---------------------------------------------------------------

# Tumor immune cell atlas from Nieto et al.

A single-cell tumor immune atlas for precision oncology

Data from:
https://genome.cshlp.org/content/31/10/1913.long#sec-13


## Load data

```{r fig.height=12, fig.width=12}
df_ni <- data.frame(fread(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "TICAtlas_metadata.csv")))


feat_mat <- as.data.frame.matrix(table(df_ni$patient, df_ni$lv1_annot)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_ni %>%
    group_by(patient) %>%
    slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$patient

meta$kmeans_cluster[is.na(meta$kmeans_cluster)] <- 7

p <- plot_pca(feat_mat, labels = meta$source)
# p <- plot_pca(feat_mat, labels = meta$subtype, plotly_3d = TRUE)


p <- plot_pca(feat_mat, labels = meta$subtype, invisible = "none", labelsize = 4)


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(meta$subtype),
  dataset = as.factor(meta$source)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)






feat_mat <- as.data.frame.matrix(table(df_ni$patient, df_ni$lv2_annot)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_ni %>%
    group_by(patient) %>%
    slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$patient

meta$kmeans_cluster[is.na(meta$kmeans_cluster)] <- 7

p <- plot_pca(feat_mat, labels = meta$source)
# p <- plot_pca(feat_mat, labels = meta$subtype, plotly_3d = TRUE)


p <- plot_pca(feat_mat, labels = meta$subtype, invisible = "none", labelsize = 4)


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(meta$subtype),
  dataset = as.factor(meta$source)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```




# ---------------------------------------------------------------

# NSCLC response atlas from Liu et al.

A single-cell atlas reveals immune heterogeneity in anti-PD-1-treated non-small cell lung cancer

Data from:
https://www.cell.com/cell/fulltext/S0092-8674(25)00291-0#sec-4


## Load data

```{r fig.height=12, fig.width=12}
df <- data.frame(fread(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "GSE243013_NSCLC_immune_scRNA_metadata.csv.gz")))

# Keep only samples with a reasonable amount of cells
samples_keep <- sort(table(df$sampleID)) > 1000
samples_keep_names <- names(samples_keep[samples_keep])
df <- df[df$sampleID %in% samples_keep_names, ]

feat_mat <- as.data.frame.matrix(table(df$sampleID, df$sub_cell_type)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  # impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  calc_perc_df() %>%
  clr()

meta <- df %>%
  group_by(sampleID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sampleID

p <- plot_pca(feat_mat, labels = meta$gender, plotly_3d = TRUE)

p <- plot_pca(feat_mat, labels = meta$pathological_response)
p <- plot_pca(feat_mat, labels = meta$pathological_response, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$pathological_response, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$pathological_response, invisible = "none", labelsize = 4, axes = c(1,3))

p <- plot_pca(feat_mat, labels = meta$radiological_response)
p <- plot_pca(feat_mat, labels = meta$radiological_response, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4, axes = c(1,3))


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  response = as.factor(meta$pathological_response),
  cancer_type = as.factor(meta$cancer_type),
  gender = as.factor(meta$gender),
  stage = as.factor(meta$pre_treatment_staging),
  smoking = as.factor(meta$smoking_history)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, grepl("CD4|CD8", colnames(feat_mat))]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "B")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "M")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```


```{r fig.width=18}
df_plot <- feat_mat
df_plot$sampleID <- meta$sampleID
df_plot$pathological_response <- meta$pathological_response
df_plot <- df_plot %>%
  pivot_longer(
    cols = colnames(feat_mat),
    names_to = "cts",
    values_to = "values"
  )
ggplot(df_plot, aes(x = cts, y = values, fill = pathological_response)) +
  geom_boxplot(outlier.shape = NA) +
  # stat_compare_means(
  #   # comparisons = my_comparisons,
  #   # ref.group = "Pseudobulk",
  #   method = "t.test",
  #   # paired = TRUE,
  #   angle = 15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```




# ---------------------------------------------------------------

# BCC and SCC response from Yost et al.

Clonal replacement of tumor-specific T cells following PD-1 blockade

Data from:
https://www.nature.com/articles/s41591-019-0522-3#data-availability


## Load data

```{r fig.height=12, fig.width=12}
df1 <- data.frame(fread(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "GSE123813_bcc_all_metadata.txt.gz")))
df1$sort <- NULL
df1$type <- "bcc"
df1$sample <- paste0(df1$patient, "_", df1$treatment)
df2 <- data.frame(fread(file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "data", "GSE123813_scc_metadata.txt.gz")))
df2$type <- "scc"
df2$sample <- paste0(df2$patient, "_", df2$treatment)

df <- rbind(df1, df2)

sort(table(df$sample))

table(df$sample, df$cluster)

feat_mat <- as.data.frame.matrix(table(df$sample, df$cluster)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  # impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  calc_perc_df() %>%
  clr()

meta <- df %>%
  group_by(sample) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample


p <- plot_pca(feat_mat, labels = meta$type)
p <- plot_pca(feat_mat, labels = meta$treatment)

p <- plot_pca(feat_mat, labels = meta$type, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$type, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$type, invisible = "none", labelsize = 4, axes = c(1,3))

p <- plot_pca(feat_mat, labels = meta$radiological_response)
p <- plot_pca(feat_mat, labels = meta$radiological_response, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4, axes = c(1,3))


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  response = as.factor(meta$pathological_response),
  cancer_type = as.factor(meta$cancer_type),
  gender = as.factor(meta$gender),
  stage = as.factor(meta$pre_treatment_staging),
  smoking = as.factor(meta$smoking_history)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, grepl("CD4|CD8", colnames(feat_mat))]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "B")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "M")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```