---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

This script prepares datasets for downstream analysis


# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
source("functions.R")
source("Processed_dataset_metadata.R")

# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
py_install("mofapy2", envname = "my_env", method="auto")
import("mofapy2")
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
# install.packages("GloScope")
library(GloScope)

# Set paths
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
seurat_file_ending <- "_ECODAprocessed.rds"
```


# Define functions

```{r}
# For ggplot x-axis label bolding (and optionally color)
library(ggtext)
library(glue)
highlight <- function(x, pat, color="black", family="") {
  ifelse(grepl(pat, x), glue("<b style='font-family:{family}; color:{color}'>{x}</b>"), x)
}

# To replace HiTME layer3 annotation based on other UCell score threshold
replace_HiTMElayer3_annot <- function(seurat, thresh = 0.1) {
  seurat$layer3 <- gsub(pattern = "_resting|_IFN|_cellCycle.G1S|_cellCycle.G2M|_HeatShock|_Heatshock|_Prolif", replacement = "", seurat$layer3)
  
  # Define the simple conditions to loop through
  simple_conditions <- list(
    IFN_UCell = list(threshold = thresh, suffix = "_IFN") #,
    # HeatShock_UCell = list(threshold = thresh, suffix = "_Heatshock")
  )
  
  # Loop through the simple conditions
  for (cond_col in names(simple_conditions)) {
    threshold <- simple_conditions[[cond_col]]$threshold
    suffix <- simple_conditions[[cond_col]]$suffix
    
    seurat@meta.data <- seurat@meta.data %>%
      mutate(
        layer3 = if_else(
          !is.na(layer3) & .data[[cond_col]] > threshold,
          paste0(layer3, suffix),
          layer3
        )
      )
  }
  
  # Now, handle the combined "Proliferation" condition separately
  seurat@meta.data <- seurat@meta.data %>%
    mutate(
      layer3 = if_else(
        !is.na(layer3) & (cellCycle.G1S_UCell > thresh | cellCycle.G2M_UCell > thresh),
        paste0(layer3, "_Prolif"),
        layer3
      )
    )
  
  return(seurat)
}
```


# Create results list to fill

```{r}
factors_test <- c(2, 3, 5, 10)

result_file <- "result_list.rds"

if (file.exists(result_file)) {
  result_list <- readRDS(result_file)
} else {
  result_list <- list()
}

analysis_elems <- c("bmark", "trans", "zeroimp")
```


# ---------------------------------------------------------------

# Process datasets

```{r message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
for (ds in names(datasets)) {
  elem_null <- c()
  for (a in analysis_elems) {
    elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
    
    if (is.null(result_list[[a]][[ds]])) {
      print(paste("To be analyzed:", a, ds))
    }
  }
  
  # If any analyses missing, process dataset for missing analyses
  if (any(elem_null)) {
    seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
    
    seurat <- replace_HiTMElayer3_annot(seurat, thresh = 0.1)
    
    seurat@misc$label_col <- datasets[[ds]][["label_col"]]
    seurat@misc$low_res_ct_col <- datasets[[ds]][["low_res_ct_col"]]
    seurat@misc$hi_res_ct_col <- datasets[[ds]][["hi_res_ct_col"]]
    
    result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
  }
}
```



```{r}
label_list <- list()
for (ds in names(datasets)) {
  print(ds)
  seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
  
  metadata <- get_metadata(seurat)

  label_list[[ds]] <- get_labels(seurat, datasets[[ds]][["label_col"]])
  names(label_list[[ds]]) <- metadata[["Sample"]]
}
```



# ---------------------------------------------------------------

# Gong and Sharma

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
```


## CMV (females)

```{r}
ds <- "GongSharma_cmv_females"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {
  
  seurat <- load_h5ad_to_seurat("data/Gongsharma_cmv_females.h5ad")
  seurat$Sample <- seurat$specimen.specimenGuid

  seurat@misc$label_col <- "subject.cmv"
  seurat@misc$low_res_ct_col <- "AIFI_L1"
  seurat@misc$hi_res_ct_col <- "AIFI_L3"
    
  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
}

# Error in if (!any(within)) stop("there should be replicates within groups") : 
#   missing value where TRUE/FALSE needed
# Error during wrapup: not that many frames on the stack
# Error: no more error handlers available (recursive errors?); invoking 'abort' restart
```

### Check for batch effect

```{r}
alpha <- 0.3

meta_cols_plot <- c(
  "Sample",
  "RNA_snn_res.0.5",
  "batch_id",
  "pool_id",
  "chip_id",
  "well_id",
  "cohort.cohortGuid",
  "AIFI_L1",
  "AIFI_L2",
  "AIFI_L3",
  "subject.cmv"
)

# Check in broad cell types (strong global batch effect)

Idents(seurat) <- "Sample"
sub <- subset(x = seurat, downsample = 100)

sub <- sub |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(dims = 1:50, verbose = FALSE) |>
  FindNeighbors() |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:50)

for (mc in meta_cols_plot) {
  Idents(sub) <- mc
  DimPlot(sub, reduction = "umap", alpha = alpha)
}

# No strong batch effect observed


# To check for more subtle batch effect, we can subset to only T cells, where the gene expression difference between cells is smaller
# Thus, by subseting to only T cells, we should be better able to see smaller batch effects

sub <- subset(x = seurat, subset = AIFI_L1 == "T cell")
sub <- subset(x = sub, downsample = 1000)

sub <- sub |>
  NormalizeData() |>
  FindVariableFeatures(nfeatures = 2000) |>
  ScaleData() |>
  RunPCA(dims = 1:50, verbose = FALSE) |>
  FindNeighbors() |>
  FindClusters(resolution = 0.5) |>
  RunUMAP(dims = 1:50)

for (mc in meta_cols_plot) {
  Idents(sub) <- mc
  DimPlot(sub, reduction = "umap", alpha = alpha)
}

# No weak batch effect observed
```



## Age (females cmvneg)

```{r}
ds <- "GongSharma_age_females_cmvneg"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))
  
  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  seurat <- load_h5ad_to_seurat("data/Gongsharma_age_females_cmvneg.h5ad")
  seurat$Sample <- seurat$specimen.specimenGuid
  
  seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))

  seurat@misc$label_col <- "age"
  seurat@misc$low_res_ct_col <- "AIFI_L1"
  seurat@misc$hi_res_ct_col <- "AIFI_L3"
  
  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
}
```





# ---------------------------------------------------------------

# Bassez

# IMPORTANT: Needs automated annotation, as seurat$cohort c("neoadjuvant_chemo", "treatment_naive") seem to be separately and differently annotated by the authors
## Has also other non-immune cells

```{r}
ds <- "Bassez"

elem_null <- c()
for (a in analysis_elems) {
  elem_null <- c(elem_null, is.null(result_list[[a]][[ds]]))

  if (is.null(result_list[[a]][[ds]])) {
    print(paste("To be analyzed:", a, ds))
  }
}

# If any analyses missing, process dataset for missing analyses
if (any(elem_null)) {

  ds_name <- "BassezA_2021_33958794"

  path_root <- file.path(path_nas, ds_name)
  path_output <- file.path(path_root, "output")

  seurat <- readRDS(file.path(path_output, sprintf("%swhole.rds", ds_name)))

  seurat <- subset(seurat, subset = timepoint == "Pre")
  seurat <- subset(seurat, subset = expansion != "n/a")
  seurat <- subset(seurat, subset = cohort == "treatment_naive")

  seurat$Sample <- seurat$sample
  seurat$cellSubType[is.na(seurat$cellSubType)] <- seurat$cellType[is.na(seurat$cellSubType)]

  convert_seurat_to_h5ad(seurat, ds)

  seurat@misc$label_col <- "expansion"
  seurat@misc$low_res_ct_col <- "cellType"
  seurat@misc$hi_res_ct_col <- "cellSubType"

  result_list <- run_analyses(result_list, ds, seurat, path_data, path_plots, factors_test)
}
```




# ---------------------------------------------------------------

# Test new dataset (template)

```{r}
# seurat_sub <- subset(seurat, subset = OriginalAnnotationLevel1 %in% c("B", "Mast", "Myeloid", "Plasma", "TNKILC"))
# 
# test <- run_benchmark_analysis(seurat_sub,
#                                get_labels(seurat_sub, seurat_sub@misc$label_col),
#                                seurat@misc$low_res_ct_col,
#                                seurat@misc$hi_res_ct_col,
#                                factors_test = 2,
#                                Pseudobulk = TRUE,
#                                ECODA_deconv = FALSE,
#                                ECODA_low_res = FALSE,
#                                ECODA_high_res = TRUE,
#                                Pseudobulk_PCA = FALSE,
#                                ECODA_high_res_PCA = FALSE,
#                                MOFA = FALSE,
#                                scITD = FALSE,
#                                MrVI = FALSE,
#                                show_pca_plots = TRUE,
#                                save_pca_plots = FALSE
# )
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Create list to export raw data to excel

```{r}
df_results_list <- list()
```

# ---------------------------------------------------------------
# ---------------------------------------------------------------


# Plot benchmark results

## Save and flatten the nested list into a DataFrame

```{r}
bmark_result_list <- result_list[["bmark"]]

df_results <- imap_dfr(bmark_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    if ("scores" %in% names(.x)) {
      n_samples <- nrow(.x$feat_mat)
      n_features <- ncol(.x$feat_mat)
      
      imap_dfr(.x$scores, ~ {
        tibble(
          dataset = dataset_name,
          method = method_name,
          score = .y,
          value = .x,
          n_samples = n_samples,
          n_features = n_features
        )
      })
    } else {
      tibble() # Return an empty tibble if "scores" is not found
    }
  })
})

df_results$method <- gsub("ECODA_highres", "ECODA_author_HL", df_results$method)
df_results$method <- gsub("ECODA_lowres", "ECODA_author_LR", df_results$method)
```

## Show + sort all tested conditions

```{r}
library(gtools)
levs <- mixedsort(as.character(unique(df_results$method)))
levs

df_results$method <- factor(df_results$method, levels = levs)

df_results$dataset[df_results$dataset == "GongSharma_females_cmv"] <- "GongSharma_cmv"
df_results$dataset[df_results$dataset == "GongSharma_females_cmvneg_age"] <- "GongSharma_age"

df_results_list[["Benchmark"]] <- df_results
```


## Plot results

```{r fig.height=10, fig.width=14}
# New facet label names for dose variable
score.labs <- c("ANOSIM score", "Adjusted rand index", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            labeller = labeller(score = score.labs, dataset = datas.labs),
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "benchmark_detailed_barplot.pdf"),
       width = 150,
       height = 7, limitsize = FALSE)




df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_bars.pdf"),
       width = 14,
       height = 7)



filter_methods <- c(
  "Pseudobulk",
  "Freq_highres",
  "ECODA_deconv",
  "ECODA_author_LR",
  "ECODA_author_HR",
  "ECODA_high_res_2_PCA_dims",
  "ECODA_highres_top_varexp0.7",
  # "ECODA_seuratres_0.4",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI",
  "scPoli",
  "PILOT",
  "GloScope"
)

library(forcats)

df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "Pseudobulk",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
  ) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_mainmethods_refpb.pdf"),
       width = 7,
       height = 7)





filter_methods <- c(
  "Pseudobulk",
  # "Freq_highres",
  # "ECODA_deconv",
  # "ECODA_lowres",
  "ECODA_highres",
  # "ECODA_high_res_2_PCA_dims",
  # "ECODA_highres_top_varexp0.7",
  # "ECODA_seuratres_0.4",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI",
  "scPoli",
  "PILOT",
  "GloScope"
)


p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "MOFA_2_factors" = "MOFA",
                         "scITD_2_factors" = "scITD",
                         "ECODA_highres" = "ECODA")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "ECODA", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "ECODA",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "ECODA", "black"))
print(p)
ggsave(file.path(path_plots, "POSTER_benchmark_summary_barplot_anosim_mainmethods_refECODA.pdf"),
       width = 3.5,
       height = 4)




filter_methods <- c(
  "Pseudobulk",
  "ECODA_highres",
  "ECODA_highres_top_varexp0.1"
)


p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_highres_top_varexp0" = "2 most highly variable cell types",
                         "ECODA_highres" = "ECODA")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "ECODA", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "ECODA",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "ECODA", "black"))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAvs2HVCvsBULK_refECODA.pdf"),
       width = 2,
       height = 6)





filter_methods <- c(
  "ECODA_deconv",
  "ECODA_lowres",
  "ECODA_highres",
  "ECODA_highres_HiTME",
  "ECODA_highres_HiTME_layer3",
  "ECODA_highres_scATOMIC",
  "ECODA_seuratres_0.4"
)

# my_comparisons <- list(
#   c("Low res. annot.", "High res. annot.")
#   )

p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_deconv" = "Deconvoluted",
                         "ECODA_lowres" = "Low res. author annot.",
                         "ECODA_highres" = "High res. author annot.",
                         "ECODA_highres_HiTME" = "HiTME annot.",
                         "ECODA_highres_HiTME_layer3" = "HiTME annot. w. gene. progs.",
                         "ECODA_highres_scATOMIC" = "scATOMIC annot.",
                         "ECODA_seuratres_0.4" = "Unsup. Leiden clust.")) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value, 
             fill = ifelse(method == "High res. author annot.", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.5) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "High res. author annot.",
    method = "wilcox.test",
    paired = TRUE,
    label.y = 1,
    tip.length = 0,
    aes(label = after_stat(p.signif))
  ) +
  ylim(c(-0.03, 1)) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1)) +
  scale_fill_identity() +
  scale_x_discrete(labels = function(x) highlight(x, "High res. author annot.", "black"))
print(p)
ggsave(file.path(path_plots, "POSTER_benchmark_summary_barplot_anosim_ecodamethods_refECODA_highres.pdf"),
       width = 3,
       height = 5)





p <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods) %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_highres",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    angle = 25,
    label.y = 0.88,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 0.9)) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_mainmethods_refECODAhr.pdf"),
       width = 3.5,
       height = 4)





df_results %>%
  filter(score == "cluster_score") %>%
  filter(method %in% filter_methods) %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "Pseudobulk",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
  ) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

df_results %>%
  filter(score == "cluster_score") %>%
  filter(method %in% filter_methods) %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_highres",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
  ) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))




ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores.pdf"),
       width = 7,
       height = 7)


df_results_ecoda <- df_results[df_results$method %in% c("ECODA_highres", "ECODA_highres_HiTME", "ECODA_highres_HiTME_layer3", "ECODA_highres_scATOMIC"), ]


df_results_ecoda %>%
  filter(score == "anosim_score") %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_ecoda$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Separation score (ANOSIM)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_ecodaonly.pdf"),
       width = 5,
       height = 6)


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

df_results_clustering <- df_results_clustering[!grepl("2|10|PCA", df_results_clustering$method), ]

for (ds in unique(df_results_clustering$dataset)) {
  number_cts <- dim(result_list$bmark[[ds]]$ECODA_highres$feat_mat)[2]
  df_results_clustering[df_results_clustering$dataset == ds, "number_celltypes"] <- number_cts
}

# df_results_clustering$dataset <- paste0(df_results_clustering$dataset, " (", df_results_clustering$number_celltypes, ")")
# legend_title <- "Dataset (number of cell types)"
legend_title <- "Dataset"

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_clustering)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)


my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "Pseudobulk"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(3, 6, 4, 5, 2, 1)]


p <- ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(group = legend_title, linetype = legend_title, shape = legend_title) +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_clustering.pdf"),
       width = 7,
       height = 5)



df_results_mod <- df_results %>%
  filter(score == "mod_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_mod)
summary(anova_result)
stat.test <- rstatix::tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)



p <- ggplot(df_results_mod, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(
    # comparisons = my_comparisons,
    ref.group = "Pseudobulk",
    method = "t.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
  ) +
  geom_point(aes(shape = dataset), size = 2) +
  scale_shape_manual(values=1:length(unique((df_results_mod$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "Modularity score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(p)
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity.pdf"),
       width = 7,
       height = 6)

```


## Plot feature selection

Surprisingly, even at the lowest number highly variable cell types (i.e. variance explained of 10% -> >=2 cell types) the average biological separation is quite very high.
Different daatasets can do with a different number of highly variable cell types (separation plateaus out at different levels of variance explained % and number of cell types retained).

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_hi", method) & !grepl("HiTME|scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = if_else(
    grepl("^top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

p1 <- df_results_sub %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(0, 1)) +
  labs(x = "Variance explained (%)", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p1)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAfeatureselectionHVC.pdf"),
       width = 4,
       height = 4)

p2 <- df_results_sub %>%
  ggplot(aes(x = method, y = n_features)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  labs(x = "Variance explained (%)", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p2)

# Create p1 (the top plot) without x-axis labels
p1_clean <- p1 + theme(axis.title.x = element_blank(), 
                       axis.text.x = element_blank(),
                       axis.ticks.x = element_blank())

# Combine the plots vertically
combined_plot <- p2 / p1
print(combined_plot)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAfeatureselectionHVC_withNumCT.pdf"),
       width = 4,
       height = 5)


library(ggbreak)

filter_methods_features <- c(
  # "Pseudobulk",
  "ECODA_highres",
  "ECODA_highres_top_varexp0.7"
)

df_results %>%
  filter(score == "anosim_score") %>%
  filter(method %in% filter_methods_features) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = gsub("ECODA_highres", "All cell types", method)) %>%
  
  ggplot(aes(x = fct_reorder(method, n_features, .desc = TRUE, .fun = mean), y = n_features)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  labs(x = "", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_nfeatures.pdf"),
       width = 1.4,
       height = 4)
```

## Separation based on least variable cell types (sanity check)

```{r}
# Create an empty list to store the results from each dataset
temp_list <- list()

for (ds in names(result_list[["bmark"]])) {
  feat_mat <- result_list[["bmark"]][[ds]]$ECODA_highres$feat_mat
  labels <- result_list[["bmark"]][[ds]]$ECODA_highres$plot[["data"]][["Groups"]]
  
  ncts <- 2
  
  # Calculate separation score for most highly variable cell types (HVC)
  top_hvcs <- get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
    get_hvcs(top_n_hvcs = ncts, variance_threshold = NULL)
  score_most_hvc <- calc_sep_score(feat_mat[, top_hvcs], labels)$anosim_score
  
  # Calculate separation score for least highly variable cell types (HVC)
  least_hvcs <- get_ct_var(feat_mat, show_plot = FALSE, descending = FALSE) %>%
    get_hvcs(top_n_hvcs = ncts, variance_threshold = NULL)
  score_least_hvc <- calc_sep_score(feat_mat[, least_hvcs], labels)$anosim_score
  
  # Store the results in a temporary data frame
  temp_df <- data.frame(
    dataset = ds,
    anosim_most_hvc = score_most_hvc,
    anosim_least_hvc = score_least_hvc
  )
  
  # Add the temporary data frame to the results list
  temp_list[[ds]] <- temp_df
}

# Combine all data frames in the list into a single data frame
df_plot <- bind_rows(temp_list)



# Reshape the data from wide to long format for ggplot
df_plot_long <- df_plot %>%
  pivot_longer(
    cols = starts_with("anosim_"),
    names_to = "group",
    values_to = "anosim_score"
  ) %>%
  # Clean up the group names for better readability
  mutate(group = gsub("anosim_|_hvc", "", group))

# Create the grouped bar plot
ggplot(df_plot_long, aes(x = dataset, y = anosim_score, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "ANOSIM Scores for Highly vs. Least Variable Cell Types",
    x = "Dataset",
    y = "ANOSIM Score",
    fill = "Cell Type Group" # Legend title
  ) +
  scale_fill_manual(values = c("most" = "darkred", "least" = "steelblue")) +
  theme_minimal()

```


## Most highly variable cell types per dataset

```{r}
print("Most highly variable cell types per dataset:")
for (ds in names(result_list[["bmark"]])) {
  print(ds)
  print(paste("ANOSIM score using all cell types:", round(result_list[["bmark"]][[ds]][["ECODA_highres"]]$scores$anosim_score, 3)))
  print(paste("ANOSIM score at 10% variance explained:", round(result_list[["bmark"]][[ds]][["ECODA_highres_top_varexp0.1"]]$scores$anosim_score, 3)))
  print("Most highly variable cell types (authors annotation):\n")
  print(paste(colnames(result_list[["bmark"]][[ds]][["ECODA_highres_top_varexp0"]][["feat_mat"]])))
  print("Most highly variable cell types (HiTME annotation):\n")
  print(paste(colnames(result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0"]] [["feat_mat"]])))
}
```

### Check IFN expression in Stephenson dataset

```{r}
ds <- "Stephenson"
seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))

plot_data <- data.frame(
  expression = seurat$HeatShock_UCell,
  status = seurat$Status
)

ggplot(plot_data, aes(x = expression, fill = status)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Expression Density by Status",
    x = "IFN_UCell Expression",
    y = "Density",
    fill = "Status"
  ) +
  xlim(c(0, 0.2)) +
  theme_minimal() +
  scale_fill_manual(values = c("Healthy" = "steelblue", "Covid" = "darkred"))
```


```{r}
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, "Status")
test <- process_coda_fig(seurat, labels, ct_col = "layer2", title = "ECODA\nHiTME layer2")
test

varexp_hvc <- 0.7
test1 <- process_coda_fig(seurat, labels, ECODA_top_varexp_hvct = varexp_hvc, ct_col = "layer2", title = paste0("ECODA layer2\nhigh res. var. exp. ", varexp_hvc*100, "%"))
test1


df_counts <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", "layer2")

feat_mat <- df_counts %>%
  impute_zeros(clr_zero_impute_method = clr_zero_impute_method, clr_zero_impute_num = clr_zero_impute_num) %>%
  calc_perc_df()

feat_mat <- feat_mat %>%
  clr()

get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
        get_hvcs(top_n_hvcs = 10, variance_threshold = NULL)
```

### Check different UCell score thresholds for HiTME layer3

```{r}
seurat <- replace_HiTMElayer3_annot(seurat)
```

```{r}
metadata <- get_metadata(seurat)
labels <- get_labels(seurat, "Status")
test2 <- process_coda_fig(seurat, labels, ct_col = "layer3", title = "ECODA\nHiTME layer3")
test2

varexp_hvc <- 0.7
test3 <- process_coda_fig(seurat, labels, ECODA_top_varexp_hvct = varexp_hvc, ct_col = "layer3", title = paste0("ECODA layer3\nhigh res. var. exp. ", varexp_hvc*100, "%"))
test3


df_counts <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", "layer3")

feat_mat <- df_counts %>%
  impute_zeros(clr_zero_impute_method = clr_zero_impute_method, clr_zero_impute_num = clr_zero_impute_num) %>%
  calc_perc_df()

feat_mat <- feat_mat %>%
  clr()

get_ct_var(feat_mat, show_plot = FALSE, descending = TRUE) %>%
        get_hvcs(top_n_hvcs = 10, variance_threshold = NULL)
```



## Reduced separation with fewer cell types?
Why do some datasets show reduced separability when using fewer highly variable cell types while other still show good separation even with just 2 cell types?
Are there any comonalities between these datasets?

In general, datasets with HIGH ANOSIM score (~ >0.5) when using all cell types also show still good separability with very low number of cell types.
-> Maybe these biological conditions are more distinct and are easier to separate?
Datasets with LOW ANOSIM score (~ <0.5) when using all cell types also show decreasing separability when reducing the number of cell types.
-> The minimum number of cell types for maximal separability depends on the dataset

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_hi", method) & !grepl("HiTME|scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = if_else(
    grepl("^top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

ds_diffs <- df_results_sub %>%
  filter(method %in% c("10", "70")) %>%
  group_by(dataset) %>%
  summarize(
    diff = value[method == "10"] - value[method == "70"]
  ) %>%
  arrange(diff)

ggplot(ds_diffs, aes(x = reorder(dataset, diff), y = diff)) +
  geom_bar(stat = "identity") +
  labs(x = "Dataset", y = "Difference") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Which datasets have lower separation when using less cell types?\nVar. exp. 10% - Var. exp. 70%")
```


## Which datasets HiTME annotation underperforms?

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("^ECODA_highres$|^ECODA_highres_HiTME$", method) & !grepl("scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = gsub("ECODA_highres", "Author annotation", method)) %>%
  # mutate(method = if_else(
  #   grepl("^top_varexp0\\.\\d", method),
  #   paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
  #   method
  # )) %>%
  # mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method))))

ds_diffs <- df_results_sub %>%
  group_by(dataset) %>%
  summarize(
    diff = value[method == "HiTME"] - value[method == "Author annotation"]
  ) %>%
  arrange(diff)

ggplot(ds_diffs, aes(x = reorder(dataset, diff), y = diff)) +
  geom_bar(stat = "identity") +
  labs(x = "Dataset", y = "Difference") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, hjust = 1)) +
  ggtitle("For each dataset, how does ECODA perform based on\nHiTME annotation vs Author annotation?")
```



## Plot feature selection (HiTME layer3)

```{r}
df_results_sub <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(grepl("ECODA_hi", method) & !grepl("scATOMIC|PCA", method)) %>%
  mutate(method = gsub("ECODA_highres_", "", method)) %>%
  mutate(method = if_else(
    grepl("^top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  mutate(method = gsub("ECODA_highres", "100", method)) %>%
  mutate(method = if_else(
    grepl("HiTME_layer3_top_varexp0\\.\\d", method),
    paste0("HiTME_layer3_", as.integer(gsub("HiTME_layer3_top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  mutate(method = gsub("^HiTME_layer3$", "HiTME_layer3_100", method)) %>%
  mutate(method = gsub("^HiTME$", "HiTME_layer2_100", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = mixedsort(levels(method)))) %>%
  mutate(method = fct_relevel(method, "HiTME_layer2_100", after = Inf)) %>%
  filter(!grepl("varexp0|varcts", method))

p1 <- df_results_sub %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "All cell types",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(0, 1)) +
  labs(x = "Variance explained (%)", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p1)
# ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAfeatureselectionHVC.pdf"),
#        width = 4,
#        height = 4)

p2 <- df_results_sub %>%
  ggplot(aes(x = method, y = n_features)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  labs(x = "Variance explained (%)", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p2)

# Create p1 (the top plot) without x-axis labels
p1_clean <- p1 + theme(axis.title.x = element_blank(), 
                       axis.text.x = element_blank(),
                       axis.ticks.x = element_blank())

# Combine the plots vertically
combined_plot <- p2 / p1
print(combined_plot)
ggsave(file.path(path_plots, "benchmark_summary_barplot_anosim_ECODAandHiTMEfeatureselectionHVC_withNumCT.pdf"),
       width = 5,
       height = 8)
```




## Show example PCAs for publication

```{r fig.height=7, fig.width=7}
for (ds in names(result_list[["bmark"]])) {
  p <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres"]][["plot"]][["data"]][["Groups"]],
    n_ct_show = 5,
    invisible = "quali",
    labelsize = 5
  ) + ggtitle(ds)
  print(p)
}

ds <- "Stephenson"
plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.9"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.9"]][["plot"]][["data"]][["Groups"]],
    n_ct_show = 50,
    invisible = "quali",
    labelsize = 3
) + ggtitle(ds)

```


```{r fig.height=6, fig.width=12}

levs <- c(
  "ECODA_high_res",
  "ECODA_low_res",
  "ECODA_deconv",
  "Pseudobulk",
  "MOFA_2_factors",
  "scITD_2_factors",
  "MrVI"
)


plot_list <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_cmv_females"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_cmv_females.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["GongSharma_age_females_cmvneg"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_GongSharma_age_females_cmvneg.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Lee"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Lee"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Lee.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Zhang"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Zhang"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Zhang.pdf"),
       plot = p,
       width = 15,
       height = 7)


plot_list <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["plot"]])
scores <- lapply(bmark_result_list[["Stephenson"]], function(method) method[["scores"]])
scores <- scores[levs]
scores_long <- sapply(scores, unlist) %>% t() %>%  as.data.frame()
scores_long$method <- row.names(scores_long)
scores_long <- scores_long %>%
  pivot_longer(
    cols = c(sil_score, mod_score, cluster_score),
    names_to = "score_method",
    values_to = "values"
  )
scores_long$method <- factor(scores_long$method, levels = levs)

score_p <- ggplot(scores_long, aes(x = method, y = values)) +
  # geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=score_method)) + 
  geom_line(aes(group=score_method, linetype=score_method), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  ylim(c(min(scores_long$values, 0), max(scores_long$values))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_list <- plot_list[!grepl("5_factors|10_factors|dims", names(plot_list))]
plot_list <- plot_list[c(4, 3, 2, 1, 5:7)]
plot_list <- lapply(plot_list, function(x){x + theme(aspect.ratio=1)})
for (p in names(plot_list)){
  plot_list[[p]][["labels"]][["title"]] <- sub("\\nCluster score: .*", "", plot_list[[p]][["labels"]][["title"]])
}
plot_list[["scores"]] <- score_p
p <- cowplot::plot_grid(plotlist = plot_list, ncol = 4)
p
ggsave(file.path(path_plots, "benchmark_pca_example_Stephenson.pdf"),
       plot = p,
       width = 15,
       height = 7)

```

## Example PCAs for HiTME layer2 vs layer3


```{r}
plot_pca(result_list[["bmark"]][["Adams"]][["ECODA_highres_HiTME_layer3_top_varexp0.1"]][["feat_mat"]],
         labels = result_list[["bmark"]][["Adams"]][["ECODA_highres_HiTME_layer3_top_varexp0.1"]][["plot"]][["data"]][["Groups"]],
         invisible = "quali",
         labelsize = 4
         )
```


```{r}
plist_layer2 <- list()
for (ds in names(result_list$bmark)) {
  plist_layer2[[ds]] <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2_top_varexp0.2"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer2_top_varexp0.2"]][["plot"]][["data"]][["Groups"]],
    invisible = "quali",
    labelsize = 4,
    title = ds
  )
}

plist_layer3 <- list()
for (ds in names(result_list$bmark)) {
  plist_layer3[[ds]] <- plot_pca(
    result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.2"]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][["ECODA_highres_HiTME_layer3_top_varexp0.2"]][["plot"]][["data"]][["Groups"]],
    invisible = "quali",
    labelsize = 4,
    title = ds
  )
}
```

```{r}
ggsave("pca_layer2.pdf", cowplot::plot_grid(plotlist = plist_layer2, ncol = 3), width = 20, height = 20)
ggsave("pca_layer3.pdf", cowplot::plot_grid(plotlist = plist_layer3, ncol = 3), width = 20, height = 20)
```



# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot normalization results

## Plot results

```{r}
df_results <- Reduce(rbind, result_list[["trans"]]) %>%
  mutate(trans_method = recode(trans_method,
                         "clr" = "CLR",
                         "alr_mincvref" = "ALR_minCVref",
                         "alr_randref" = "ALR_randref",
                         "arcsine_sqrt" = "Arcsine_sqrt",
                         "counts" = "Counts",
                         "freq" = "Frequency(%)"))

df_results_list[["Normalization"]] <- df_results


df_results <- df_results %>%
  pivot_longer(
    cols = c(ANOSIM_score, Modularity_score, Adjusted_Rand_Index),
    names_to = "score",
    values_to = "value"
  ) %>%
  as.data.frame()

df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score") %>%
  group_by(trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_anosim <- df_results_anosim %>%
  arrange(desc(mean))

df_results_anosim$trans_method <- factor(df_results_anosim$trans_method, levels = df_results_anosim$trans_method)
df_results$trans_method <- factor(df_results$trans_method, levels = df_results_anosim$trans_method)



df_results %>%
  group_by(score, trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = trans_method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))
ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores_bars.pdf"),
       width = 7,
       height = 7)


df_results %>%
  filter(score == "ANOSIM_score") %>%
  
  ggplot(aes(x = trans_method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..))),
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 0.9
  ) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  ylim(c(-0.1, 1))
ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_anosim_refCLR.pdf"),
       width = 2.5,
       height = 3.5)



my_comparisons <- list(
  c("CLR", "Frequency(%)"),
  c("CLR", "Counts")
  )

df_results %>%
  filter(score == "ANOSIM_score") %>%
  filter(trans_method %in% c("Counts", "Frequency(%)", "CLR")) %>%
  
  ggplot(aes(x = fct_reorder(trans_method, value, .desc = FALSE, .fun = mean), y = value, fill = ifelse(trans_method == "CLR", "lightgreen", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    paired = TRUE,
    size = 3,
    tip.length = 0,
    # label.y = 0.95,
    aes(label = after_stat(p.signif))
  ) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  ylim(c(-0.05, 1)) +
  scale_fill_identity() +
  scale_x_discrete(labels = c(
    "Counts",
    "Frequency(%)",
    expression(bold("CLR")) # Make label bold
  ))
ggsave(file.path(path_plots, "POSTER_transformations_real_datasets_summary_barplot_anosim_refCLR.pdf"),
       width = 2,
       height = 3.5)





# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "ANOSIM score", "Modularity score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))



ggplot(df_results, aes(x = trans_method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "",
       y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
# facet_grid(scales = "free",
#            rows = vars(score_method), cols = vars(dataset))
ggsave(file.path(path_plots, "transformations_real_datasets_detailed_barplot.pdf"),
       width = 8,
       height = 7)



# df_results %>%
#   group_by(score_method, method) %>%
#   summarize(mean = mean(values), sem = sd(values) / sqrt(length(values))) %>%
#   ungroup() %>%
#   
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Separation score") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   facet_grid(scales = "free",
#              rows = vars(score_method),
#              labeller = labeller(score_method = score.labs))


ggplot(df_results, aes(x = trans_method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores.pdf"),
       width = 5,
       height = 5)



# df_results_anosim %>%
#   ggplot(aes(x = method, y = mean)) + 
#   geom_bar(stat = "identity", position = position_dodge()) +
#   # stat_summary(fun.y = mean, geom = "bar") +
#   geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
#   # geom_point(position = position_jitterdodge()) +
#   labs(x = "", y = "Adjusted Rand Index") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
#   NoLegend()


df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score")

df_results_anosim <- df_results_anosim[!grepl("pca", df_results_anosim$trans_method), ]

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ trans_method, data = df_results_anosim)
summary(anova_result)
stat.test <- tukey_hsd(anova_result)
stat.test
stat.test$p.adj <- round(stat.test$p.adj, 3)
# stat.test <- stat.test[["method"]] %>% as.data.frame()
# 
# # Extract row names where p adj < 0.05
# significant_comparisons <- rownames(stat.test)[stat.test[["p adj"]] < 0.05]
# 
# # Initialize an empty list to store results
# my_comparisons <- list()
# 
# # Split and store each comparison in the list
# for (comparison in significant_comparisons) {
#   parts <- strsplit(comparison, "-")[[1]]
#   my_comparisons <- c(my_comparisons, list(parts))
# }

my_comparisons <- purrr::map(unique(as.character(df_results_anosim$trans_method)), ~ c(.x, "Frequency(%)"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
# my_comparisons <- c(my_comparisons, list(c("clr", "arcsine_sqrt")))
my_comparisons <- my_comparisons[c(4, 3, 2, 1, 5)]

# my_comparisons <- list(
#   c("alr_mincvref", "freq"),
#   c("alr_randref", "freq"),
#   c("arcsine_sqrt", "freq"),
#   c("clr", "freq"),
#   c("clr_pca", "freq"),
#   c("counts", "freq")
#   )

ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    # ref.group = "freq",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "ANOSIM score") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #                    y.position = c(max(df_results_anosim$values)+0.1, max(df_results_anosim$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot.pdf"),
       width = 6,
       height = 4.5)


ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    # comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  labs(x = "", y = "ANOSIM score") +
  # stat_pvalue_manual(stat.test, label = "p.adj", 
  #                    y.position = c(max(df_results_anosim$values)+0.1, max(df_results_anosim$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

```

# Statistical comparison

```{r}
df_results_anosim <- df_results %>%
  filter(score_method == "ANOSIM_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(values ~ trans_method, data = df_results_anosim)
summary(anova_result)

TukeyHSD(aov(df_results_anosim$values ~ df_results_anosim$trans_method))
```




# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Plot zero imputation results

```{r}
zeroimp_result_list <- result_list[["zeroimp"]]
```


## Plot results

```{r}
df_results <- imap_dfr(zeroimp_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    imap_dfr(.x, ~ {
      tibble(
        dataset = dataset_name,
        method = method_name,
        score = .y,
        value = .x
      )
    })
  })
})

df_results <- df_results[df_results$method != "impRZilr", ]

df_results_list[["Zero_imputation"]] <- df_results

df_results %>%
  filter(score == "anosim_score") %>%
  filter(method != "asinsqrt") %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "gray70") +
  geom_point(position = "identity", alpha = 0.6, size = 1) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_zeros_1",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.93,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.15, 1)) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_barplot_ANOSIM.pdf"),
       width = 4,
       height = 4)


# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "ANOSIM score")
names(score.labs) <- sort(unique(df_results$score))

# New facet label names for supp variable
datas.labs <- sort(unique(df_results$dataset))
datas.labs[datas.labs == "GongSharma_cmv_females"] <- "Gong & Sharma\nSubset: females\nSeparation by: CMV infection"
datas.labs[datas.labs == "GongSharma_age_females_cmvneg"] <- "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age"
names(datas.labs) <- sort(unique(df_results$dataset))


ggplot(df_results, aes(x = method, y = value)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            rows = vars(score), cols = vars(dataset))
ggsave(file.path(path_plots, "zero_imputation_real_datasets_detailed_barplot.pdf"),
       width = 10,
       height = 6)




df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%

  ggplot(aes(x = method, y = mean)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_barplot_separatescores.pdf"),
       width = 3.5,
       height = 6)

ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = score.labs))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_separatescores.pdf"),
       width = 7,
       height = 6)



df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results %>%
  filter(score == "anosim_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "counts_all_1"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(6,1,7:10, 2:5)]
# my_comparisons[[length(my_comparisons) + 1]] <- c("perc_all_0.1%", "perc_all_1%")

p <- ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_clusterscore.pdf"),
       width = 8,
       height = 6)

df_results_modularity <- df_results %>%
  filter(score == "mod_score")

p <- ggplot(df_results_modularity, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Modularity score") +
  scale_shape_manual(values=1:length(unique((df_results_modularity$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_modscore.pdf"),
       width = 8,
       height = 6)


df_results_anosim <- df_results %>%
  filter(score == "anosim_score")

p <- ggplot(df_results_anosim, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_anosimscore.pdf"),
       width = 8,
       height = 6)

```

# Test for statistical significance

To few data points for statistical significance

```{r}
library(rstatix)

df_results_cluster <- df_results %>% filter(score == "cluster_score")

outlier <- df_results_cluster %>%
  group_by(method) %>%
  rstatix::identify_outliers(value)
data.frame(outlier)

normality <- df_results_cluster %>%
  group_by(method) %>%
  rstatix::shapiro_test(value)
data.frame(normality)
cat(paste("All groups normally ditributed?", all(normality$p > 0.05),
            "\nNot normally distributed:", normality$method[normality$p <= 0.05]))

# res <- rstatix::anova_test(data = df_results_cluster, dv = value, wid = dataset, within = method) 
# rstatix::get_anova_table(res)
# 
# pair_t <- df_results_cluster %>%
#   rstatix::pairwise_t_test(value~method, paired=TRUE, p.adjust.method = "BH" )
# data.frame(pair_t)

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)




df_results_cluster <- df_results %>% filter(score == "mod_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)



df_results_cluster <- df_results %>% filter(score == "sil_score")

# Example assuming 'data' is your dataframe
anova_result <- aov(value ~ method, data = df_results_cluster)
summary(anova_result)
```

# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Export tables with raw data for paper

```{r}
library(writexl)
write_xlsx(df_results_list, path = file.path(path_data, "raw_data_for_publication.xlsx"))
```



# ---------------------------------------------------------------
# ---------------------------------------------------------------
# ---------------------------------------------------------------

# Brain from Macnair et al.

snRNA-seq stratifies multiple sclerosis patients into distinct white matter glial responses

Data from:
https://www.sciencedirect.com/science/article/pii/S0896627324008730#sec8



## Load data

```{r}
library(readxl)

df_mac_main <- data.frame(fread(file.path(path_data, "other", "MacnairW_2025_Neuron", "ms_lesions_snRNAseq_col_data_2023-09-12.txt.gz")))
df_mac_val <- data.frame(fread(file.path(path_data, "other", "MacnairW_2025_Neuron", "ms_lesions_snRNAseq_validation_col_data_2023-09-12.txt.gz")))

feat_mat <- as.data.frame.matrix(table(df_mac_main$sample_id_anon, df_mac_main$type_fine)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_mac_main %>%
    group_by(sample_id_anon) %>%
    slice(1)
row.names(meta) <- meta$sample_id_anon

p <- plot_pca(feat_mat, labels = meta$matter)


rsub <- meta$matter == "GM"
feat_mat_gm <- feat_mat[rsub, ]
meta_gm <- meta[rsub, ]

p <- plot_pca(feat_mat_gm, labels = meta_gm$lesion_type)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$seq_pool)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$sex)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$sample_source)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$diagnosis)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$age_cat)
# p <- plot_pca(feat_mat_gm, labels = meta_gm$pmi_cat)


rsub <- meta$matter == "WM"
feat_mat_wm <- feat_mat[rsub, ]
meta_wm <- meta[rsub, ]

p <- plot_pca(feat_mat_wm, labels = meta_wm$lesion_type)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$seq_pool)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$sex)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$sample_source)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$diagnosis)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$age_cat)
# p <- plot_pca(feat_mat_wm, labels = meta_wm$pmi_cat)
```


# ---------------------------------------------------------------

# Tumor immune cell atlas from Nieto et al.

A single-cell tumor immune atlas for precision oncology

Data from:
https://genome.cshlp.org/content/31/10/1913.long#sec-13


## Load data

```{r fig.height=12, fig.width=12}
df_ni <- data.frame(fread(file.path(path_data, "other", "TICAtlas_metadata.csv")))


feat_mat <- as.data.frame.matrix(table(df_ni$patient, df_ni$lv1_annot)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_ni %>%
    group_by(patient) %>%
    slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$patient

meta$kmeans_cluster[is.na(meta$kmeans_cluster)] <- 7

p <- plot_pca(feat_mat, labels = meta$source)
# p <- plot_pca(feat_mat, labels = meta$subtype, plotly_3d = TRUE)


p <- plot_pca(feat_mat, labels = meta$subtype, invisible = "none", labelsize = 4)


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(meta$subtype),
  dataset = as.factor(meta$source)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)






feat_mat <- as.data.frame.matrix(table(df_ni$patient, df_ni$lv2_annot)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  calc_perc_df() %>%
  clr()
  
meta <- df_ni %>%
    group_by(patient) %>%
    slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$patient

meta$kmeans_cluster[is.na(meta$kmeans_cluster)] <- 7

p <- plot_pca(feat_mat, labels = meta$source)
# p <- plot_pca(feat_mat, labels = meta$subtype, plotly_3d = TRUE)


p <- plot_pca(feat_mat, labels = meta$subtype, invisible = "none", labelsize = 4)


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(meta$subtype),
  dataset = as.factor(meta$source)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```




# ---------------------------------------------------------------

# NSCLC response atlas from Liu et al.

A single-cell atlas reveals immune heterogeneity in anti-PD-1-treated non-small cell lung cancer

Data from:
https://www.cell.com/cell/fulltext/S0092-8674(25)00291-0#sec-4


## Load data

```{r fig.height=12, fig.width=12}
df <- data.frame(fread(file.path(path_data, "other", "GSE243013_NSCLC_immune_scRNA_metadata.csv.gz")))

# Keep only samples with a reasonable amount of cells
samples_keep <- sort(table(df$sampleID)) > 1000
samples_keep_names <- names(samples_keep[samples_keep])
df <- df[df$sampleID %in% samples_keep_names, ]

feat_mat <- as.data.frame.matrix(table(df$sampleID, df$sub_cell_type)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  # impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  calc_perc_df() %>%
  clr()

meta <- df %>%
  group_by(sampleID) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sampleID

p <- plot_pca(feat_mat, labels = meta$gender, plotly_3d = TRUE)

p <- plot_pca(feat_mat, labels = meta$pathological_response)
p <- plot_pca(feat_mat, labels = meta$pathological_response, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$pathological_response, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$pathological_response, invisible = "none", labelsize = 4, axes = c(1,3))

p <- plot_pca(feat_mat, labels = meta$radiological_response)
p <- plot_pca(feat_mat, labels = meta$radiological_response, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4, axes = c(1,3))


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  response = as.factor(meta$pathological_response),
  cancer_type = as.factor(meta$cancer_type),
  gender = as.factor(meta$gender),
  stage = as.factor(meta$pre_treatment_staging),
  smoking = as.factor(meta$smoking_history)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, grepl("CD4|CD8", colnames(feat_mat))]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "B")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "M")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```


```{r fig.width=18}
df_plot <- feat_mat
df_plot$sampleID <- meta$sampleID
df_plot$pathological_response <- meta$pathological_response
df_plot <- df_plot %>%
  pivot_longer(
    cols = colnames(feat_mat),
    names_to = "cts",
    values_to = "values"
  )
ggplot(df_plot, aes(x = cts, y = values, fill = pathological_response)) +
  geom_boxplot(outlier.shape = NA) +
  # stat_compare_means(
  #   # comparisons = my_comparisons,
  #   # ref.group = "Pseudobulk",
  #   method = "t.test",
  #   # paired = TRUE,
  #   angle = 15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```




# ---------------------------------------------------------------

# BCC and SCC response from Yost et al.

Clonal replacement of tumor-specific T cells following PD-1 blockade

Data from:
https://www.nature.com/articles/s41591-019-0522-3#data-availability


## Load data

```{r fig.height=12, fig.width=12}
df1 <- data.frame(fread(file.path(path_data, "other", "GSE123813_bcc_all_metadata.txt.gz")))
df1$sort <- NULL
df1$type <- "bcc"
df1$sample <- paste0(df1$patient, "_", df1$treatment)
df2 <- data.frame(fread(file.path(path_data, "other", "GSE123813_scc_metadata.txt.gz")))
df2$type <- "scc"
df2$sample <- paste0(df2$patient, "_", df2$treatment)

df <- rbind(df1, df2)

sort(table(df$sample))

table(df$sample, df$cluster)

feat_mat <- as.data.frame.matrix(table(df$sample, df$cluster)) %>%
  impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
  # impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
  calc_perc_df() %>%
  clr()

meta <- df %>%
  group_by(sample) %>%
  slice(1) %>%
  as.data.frame()
row.names(meta) <- meta$sample


p <- plot_pca(feat_mat, labels = meta$type)
p <- plot_pca(feat_mat, labels = meta$treatment)

p <- plot_pca(feat_mat, labels = meta$type, plotly_3d = TRUE)
p <- plot_pca(feat_mat, labels = meta$type, invisible = "none", labelsize = 4)
p <- plot_pca(feat_mat, labels = meta$type, invisible = "none", labelsize = 4, axes = c(1,3))

# p <- plot_pca(feat_mat, labels = meta$radiological_response)
# p <- plot_pca(feat_mat, labels = meta$radiological_response, plotly_3d = TRUE)
# p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4)
# p <- plot_pca(feat_mat, labels = meta$radiological_response, invisible = "none", labelsize = 4, axes = c(1,3))


cor_matrix <- cor(feat_mat)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, margins = c(15, 15))

corrplot(cor_matrix, order="hclust")
corrplot(cor_matrix, order="hclust", hclust.method = c("ward.D2"))



library(pheatmap)

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  response = as.factor(meta$pathological_response),
  cancer_type = as.factor(meta$cancer_type),
  gender = as.factor(meta$gender),
  stage = as.factor(meta$pre_treatment_staging),
  smoking = as.factor(meta$smoking_history)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, grepl("CD4|CD8", colnames(feat_mat))]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "B")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)


feat_mat_sub <- feat_mat[, startsWith(colnames(feat_mat), "M")]
df_heatmap <- feat_mat_sub %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 6,
  cutree_cols = 6
)

```



# ---------------------------------------------------------------

# CART from Haradhvala et al.

Clonal replacement of tumor-specific T cells following PD-1 blockade

Data from:
https://www.nature.com/articles/s41591-019-0522-3#data-availability


## Load data

```{r}
seurat <- readRDS("/Users/christianhalter/Library/CloudStorage/Dropbox/CSI/Standardized_SingleCell_Datasets/HarN_2022_36097221/output/HarN_2022_36097221whole.rds")

library(HiTME)

# Fetch reference maps
ref.maps <- get.reference.maps(collection = "human")

# Cell type classification on a Seurat object
query <- Run.HiTME(query,
                   ref.maps = ref.maps[["human"]],
                   species = "human" # for mouse data set to "mouse"
                   )

```


# ---------------------------------------------------------------

# Explanation of negative ARI and ANOSIM scores

```{r}
# --- Load necessary packages ---
library(vegan)
library(mclust)
library(ggplot2)


# --- Main function to analyze clusters and plot results ---
analyze_clusters <- function(n_samples, cluster_specs) {
  # Generate the data for all clusters
  data_list <- list()
  true_labels <- c()
  group_labels <- c()
  for (i in 1:length(cluster_specs)) {
    spec <- cluster_specs[[i]]
    group_name <- LETTERS[i] # Assign simple group names A, B, C, ...
    
    # Generate the descriptive label for the legend
    mean_str <- paste0("Mean: (", paste(spec$mean, collapse = ", "), ")")
    sd_str <- paste0("SD: (", paste(spec$sd, collapse = ", "), ")")
    descriptive_label <- paste(group_name, mean_str, sd_str, sep = " | ")

    # Generate the data
    if (length(spec$sd) == 1) {
      cluster_data <- data.frame(
        x = rnorm(n_samples, mean = spec$mean[1], sd = spec$sd),
        y = rnorm(n_samples, mean = spec$mean[2], sd = spec$sd)
      )
    } else {
      cluster_data <- data.frame(
        x = rnorm(n_samples, mean = spec$mean[1], sd = spec$sd[1]),
        y = rnorm(n_samples, mean = spec$mean[2], sd = spec$sd[2])
      )
    }

    cluster_data$group <- group_name
    data_list[[i]] <- cluster_data
    true_labels <- c(true_labels, rep(group_name, n_samples))
    group_labels <- c(group_labels, rep(descriptive_label, n_samples))
  }
  
  # Combine data into a single data frame
  data_df <- do.call(rbind, data_list)

  # Add the descriptive labels to the dataframe
  data_df$group_label <- factor(group_labels, levels = unique(group_labels))

  # Generate shuffled labels for both statistical test and plot
  shuffled_labels <- sample(true_labels)
  shuffled_group_labels <- sample(group_labels) # This was the flawed part. Corrected below.
  
  # The fix: create a shuffled version of the descriptive labels that correctly corresponds to the shuffled simple labels
  shuffled_labels_map <- setNames(group_labels, true_labels)
  shuffled_group_labels <- as.character(shuffled_labels_map[shuffled_labels])

  # Calculate distance matrix
  dist_matrix <- dist(data_df[, c("x", "y")], method = "euclidean")
  
  # Calculate ANOSIM and ARI for correct labels
  anosim_correct <- anosim(dist_matrix, true_labels)
  ari_correct <- adjustedRandIndex(true_labels, true_labels)

  # Calculate ANOSIM and ARI for shuffled labels
  anosim_shuffled <- anosim(dist_matrix, shuffled_labels)
  ari_shuffled <- adjustedRandIndex(true_labels, shuffled_labels)

  # Create plot titles with rounded metrics
  title_correct <- paste0(
    "Correct Labels\n",
    "ANOSIM R: ", round(anosim_correct$statistic, 4),
    " | ARI: ", round(ari_correct, 4)
  )
  
  title_shuffled <- paste0(
    "Shuffled Labels\n",
    "ANOSIM R: ", round(anosim_shuffled$statistic, 4),
    " | ARI: ", round(ari_shuffled, 4)
  )
  
  # Create plots
  plot_correct <- ggplot(data_df, aes(x = x, y = y, color = group_label)) +
    geom_point(size = 2) +
    labs(title = title_correct, x = "Feature 1", y = "Feature 2", color = "Group Specs") +
    theme_minimal()
  print(plot_correct)
  
  # The fix: create the shuffled dataframe by re-assigning the labels to the original data points
  data_df_shuffled_plot <- data_df
  data_df_shuffled_plot$group_label <- factor(shuffled_group_labels, levels = unique(group_labels))
  
  plot_shuffled <- ggplot(data_df_shuffled_plot, aes(x = x, y = y, color = group_label)) +
    geom_point(size = 2) +
    labs(title = title_shuffled, x = "Feature 1", y = "Feature 2", color = "Group Specs") +
    theme_minimal()
  print(plot_shuffled)
  
  # Create a data frame to store the results
  results_df <- data.frame(
    Scenario = c("Correct", "Shuffled"),
    Anosim_R = c(anosim_correct$statistic, anosim_shuffled$statistic),
    ARI = c(ari_correct, ari_shuffled)
  )
  
  return(results_df)
}


# --- Define Scenarios ---
set.seed(4)



# Scenario 1: Well-separated clusters
n_samples <- 50
cluster_specs <- list(
  list(mean = c(0, 0), sd = 1),
  list(mean = c(5, 5), sd = 1)
)

# --- Run and print results for each scenario ---
cat("--- Scenario 1: Well-Separated Clusters (50 samples) ---\n")
results1 <- analyze_clusters(n_samples, cluster_specs)
print(results1)


# Scenario 2: Well-separated clusters few samples
n_samples <- 5
cluster_specs <- list(
  list(mean = c(0, 0), sd = 1),
  list(mean = c(5, 5), sd = 1)
)

# --- Run and print results for each scenario ---
cat("--- Scenario 2: Well-separated clusters (10 samples)---\n")
results2 <- analyze_clusters(n_samples, cluster_specs)
print(results2)


# Scenario 3: Well-separated clusters few samples
n_samples <- 2
cluster_specs <- list(
  list(mean = c(0, 0), sd = 1),
  list(mean = c(5, 5), sd = 1)
)

# --- Run and print results for each scenario ---
cat("--- Scenario 3: Well-separated clusters (4 samples) ---\n")
results3 <- analyze_clusters(n_samples, cluster_specs)
print(results3)


```


```{r}
# --- Load necessary packages ---
library(vegan)
library(mclust)
library(ggplot2)

# --- Define scenario parameters ---
# set.seed(42)
n_samples <- 2 # Number of samples per cluster

# Create two clearly separated Gaussian clusters
cluster1 <- data.frame(
  x = rnorm(n_samples, mean = 0, sd = 1),
  y = rnorm(n_samples, mean = 0, sd = 1)
)
cluster2 <- data.frame(
  x = rnorm(n_samples, mean = 5, sd = 1),
  y = rnorm(n_samples, mean = 5, sd = 1)
)

# Combine the data
data_df <- rbind(cluster1, cluster2)

# Create the correct, true labels
true_labels <- factor(c(rep("Cluster A", n_samples), rep("Cluster B", n_samples)))

# Create the "worst-case" labels by assigning them in an alternating pattern
worst_labels <- factor(rep(c("Worst A", "Worst B"), length.out = nrow(data_df)))

# --- Plot the scenarios ---
# Plot for Correct Labels
plot_correct <- ggplot(data_df, aes(x = x, y = y, color = true_labels)) +
  geom_point(size = 2) +
  labs(title = "Data with Correct Cluster Labels", x = "Feature 1", y = "Feature 2") +
  theme_minimal()
print(plot_correct)

# Plot for Worst-Case Labels
plot_worst <- ggplot(data_df, aes(x = x, y = y, color = worst_labels)) +
  geom_point(size = 2) +
  labs(title = "Data with Worst-Case (Alternating) Labels", x = "Feature 1", y = "Feature 2") +
  theme_minimal()
print(plot_worst)

# --- Calculate ANOSIM and ARI ---
# Calculate Euclidean distance matrix
dist_matrix <- dist(data_df[, c("x", "y")], method = "euclidean")

# ANOSIM: Test for separation based on correct vs. worst-case groups
cat("--- ANOSIM Results ---\n")
anosim_correct <- anosim(dist_matrix, true_labels)
cat("ANOSIM R-statistic (Correct Labels):", round(anosim_correct$statistic, 4), "\n")
# A high positive value indicates strong separation.

anosim_worst <- anosim(dist_matrix, worst_labels)
cat("ANOSIM R-statistic (Worst-Case Labels):", round(anosim_worst$statistic, 4), "\n\n")
# This value should be highly negative, indicating that the within-group
# dissimilarity is much greater than the between-group dissimilarity.

# ARI: Compare the worst-case labels to the correct labels
cat("--- Adjusted Rand Index Results ---\n")
# A perfect clustering (correct vs correct) gives a score of 1.
ari_perfect <- adjustedRandIndex(true_labels, true_labels)
cat("ARI (Perfect Match):", round(ari_perfect, 4), "\n")

# Comparing the correct labels to the worst-case labels should
# yield a highly negative score, indicating the worst possible agreement.
ari_worst <- adjustedRandIndex(true_labels, worst_labels)
cat("ARI (Correct vs. Worst-Case):", round(ari_worst, 4), "\n")
```


# Export cell type compositions

```{r}
get_sample_metadata <- function(cell_data_df, sample_col) {
  # Step 1: Group the data by `sample_col` and check for uniqueness
  distinct_counts <- cell_data_df %>%
    dplyr::group_by(!!rlang::sym(sample_col)) %>%
    dplyr::summarise(across(everything(), ~ n_distinct(.x))) %>%
    dplyr::ungroup()

  # Step 2: Identify columns that are constant across all samples
  constant_cols <- distinct_counts %>%
    dplyr::select(where(~ max(.x) == 1)) %>%
    names()

  # Step 3: Select and return the constant columns
  metadata <- cell_data_df %>%
    dplyr::select(all_of(c(sample_col, constant_cols))) %>%
    dplyr::distinct() %>%
    as.data.frame() # Convert back to data.frame

  # Set row names to the sample column for easier access
  rownames(metadata) <- metadata[[sample_col]]
  metadata[[sample_col]] <- NULL

  return(metadata)
}


ds_data_list <- list()

for (ds in names(datasets)) {
  print(ds)
  
  seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
  seurat@misc$label_col <- datasets[[ds]][["label_col"]]
  
  ds_data_list[[ds]] <- list(
    cell_counts_highresolution = get_ct_comp_df_seurat(
      seurat,
      sample_col = "Sample",
      ct_col = datasets[[ds]][["hi_res_ct_col"]]
    ),
    cell_counts_lowresolution = get_ct_comp_df_seurat(
      seurat,
      sample_col = "Sample",
      ct_col = datasets[[ds]][["low_res_ct_col"]]
    ),
    metadata = get_sample_metadata(seurat@meta.data, "Sample"),
    main_biologicalcondition_columnname = datasets[[ds]][["label_col"]]
  )
}
  
saveRDS(
  object = ds_data_list,
  file = file.path(path_data, "example_data.rds")
)
```
