---
title: Standardized Processing of scRNA-seq datasets
author: "M. Andreatta <massimo.andreatta at unil.ch>, S. Carmona <santiago.carmona at unil.ch>, J. Garnica <josep.garnicacaparros at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
editor_options: 
  markdown: 
    wrap: 72
---

## IMPORTANT!

**When the processing is finished, please make an entry to the
[Standardized SingleCell
Datasets](https://docs.google.com/spreadsheets/d/1THNE4TSbL6pa6ulCA9xsX1L7fzUCgyHpK-mgEOaUCko/edit?usp=drive_link)
by filling this [form](https://forms.gle/drtCUXZyvTQYU9ih8).**

------------------------------------------------------------------------
# Set parameters

Dataset name: Author_Year_PBMID_cohortID (eg. ZhangY_2022_34653365)

Dataset ID should have 3 letters from Author (Aaa) followed by
additional uniquifying letter (A). If multiple cohort, followed by 1
letter from cohort ID separated by dot (e.g. AndA.H; AndB.M; And.C,
etc.)

```{r dataset_name}
ds_name <- "MitchelJ_2023_PrePrintTBD"
ID <- "MitJ"
species <- "human"
my_seed <- 123
set.seed(my_seed)

```

Meta-data timestamp: 11/30/2022 17:37:18 (to create a new entry, use
<https://forms.gle/drtCUXZyvTQYU9ih8>) \*\*\*

# Set paths
```{r}
path_root <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets",ds_name)
dir.create(file.path(path_root))
path_data <- file.path("~/Dropbox/CSI/Datasets",ds_name)
dir.create(file.path(path_data))
path_output <- file.path(path_root, "output")
dir.create(file.path(path_output))
path_plots <- file.path(path_root, "plots")
dir.create(file.path(path_plots))
```

**CHECK**: Make sure the input data and metadata are available in our
repository.

If not already in the shared Dropbox folder
[Standardized_SingleCell_Datasets](https://www.dropbox.com/scl/fo/uv9er8tqd23vngyzjgqd6/h?rlkey=onviske2cxev8t61ewt1hjwmq&dl=0),
create a new folder for this particular dataset and download the
publicly available data, then **store it in
"\~/Dropbox/CSI/Standardized_SingleCell_Datasets/*your_study*/data/"**.
Here we offer tools to download data directly from GEO (package
`GEOquery`) and Zenodo (package `Zen4R`).

# Prepare packages and environtment

```{r load_packages_functions, message=FALSE}
# remotes::install_github("carmonalab/UCell",ref="5f12a9d")
# remotes::install_github("carmonalab/scGate",ref="17a4ce6")
# remotes::install_github("carmonalab/STACAS",ref="bd73fff")
# remotes::install_github("carmonalab/ProjecTILs",ref="cc73b97")
# remotes::install_github("carmonalab/SignatuR",ref="54bc3e8")

library(Seurat) # v5
library(limma)
# devtools::install_github("immunogenomics/presto")
#library(presto)
#library(SeuratData)
library(SeuratDisk)

library(ggplot2)
library(UCell)
library(scGate)
library(SingleCellExperiment)
library(viridis)
library(Matrix)
library(patchwork)
library(GEOquery)
library(dplyr)
library(data.table)
library(STACAS)
library(SignatuR)

# load custom functions
source("Rutils/functions.R")
```

Also, increase memory parameters to allocate large vectors, otherwise R
might crash

```{bash create_Renviron}
#include vector max size
max_size=200Gb

# create .Renviron file if not present
if ! [ -a ".Renviron" ]
then
  touch .Renviron
  echo R_MAX_VSIZE=$max_size > .Renviron
  echo New .Renviron file created!
else
  echo .Renviron already present!
  echo Check size of the allowed vector:
  cat .Renviron
fi
```


# Download data

If not done before, download data eg from GEO or Zenodo.

If not already in the shared Dropbox folder
[Standardized_SingleCell_Datasets](https://www.dropbox.com/scl/fo/uv9er8tqd23vngyzjgqd6/h?rlkey=onviske2cxev8t61ewt1hjwmq&dl=0),
create a new folder for this particular dataset and download the
publicly available data, then **store it in
"\~/Dropbox/CSI/Standardized_SingleCell_Datasets/*your_study*/data/"**.
Here we offer tools to download data directly from GEO (package
`GEOquery`) and Zenodo (package `Zen4R`).

# Download data
```{r, eval=FALSE}
# options(timeout = max(10000, getOption("timeout")))
# options(download.file.method.GEOquery = "curl")
# 
# # Systemic lupus erythematosus (SLE) dataset
# geo_acc <- 'GSE137029'
# gse <- getGEO(geo_acc)
# getGEOSuppFiles(geo_acc)
# dir.create(file.path(path_data, geo_acc))
# download_dir <- file.path(getwd(), geo_acc)
# for (file in list.files(download_dir)){
#   file.rename(from = file.path(download_dir, file),
#               to = file.path(path_data, geo_acc, file))
# }
# 
# # Pediatric SLE dataset
# geo_acc <- 'GSE135779'
# gse <- getGEO(geo_acc)
# getGEOSuppFiles(geo_acc)
# dir.create(file.path(path_data, geo_acc))
# download_dir <- file.path(getwd(), geo_acc)
# file.rename(from = file.path(download_dir, "GSE135779_genes.tsv.gz"),
#             to = file.path(path_data, geo_acc, "GSE135779_genes.tsv.gz"))
# system(sprintf("tar -xvf %s/%s_RAW.tar -C %s", geo_acc, geo_acc, file.path(path_data, geo_acc)))

# The Stephenson et al. COVID-19 dataset is publicly available at
# https://www.covid19cellatlas.org/index.patient.html, titled “COVID-19 PBMC Ncl-Cambridge-UCL” (select the only non-TCR non-BCR dataset)

# The van der Wijst et al. COVID-19 dataset can be found at
# https://cellxgene.cziscience.com/collections/7d7cabfd-1d1f-40af-96b7-26a0825a306d.
```

# Load Data

Depending on the data format at GEO or Zenodo, use different tools to
load the datasets into the environment.

For now, only using the Stephenson dataset, as its the largest one and has well-annotated metadata
```{r}
# datmats <- list()
# datmats[["GSE137029"]] <- list()
# datmats[["GSE137029"]][["flare"]] <- ReadMtx(mtx = file.path(path_data, 'GSE137029', "GSE137029_flare.mtx.gz"),
#                                              cells = file.path(path_data, 'GSE137029', "GSE137029_flare.barcodes.txt.gz"),
#                                              features = file.path(path_data, 'GSE137029', "GSE137029_flare.genes.txt.gz"),
#                                              cell.column = 1, feature.column = 1, mtx.transpose = T)
# datmats[["GSE137029"]][["sle"]] <- ReadMtx(mtx = file.path(path_data, 'GSE137029', "GSE137029_sle.mtx.gz"),
#                                            cells = file.path(path_data, 'GSE137029', "GSE137029_sle.barcodes.txt.gz"),
#                                            features = file.path(path_data, 'GSE137029', "GSE137029_sle.genes.txt.gz"),
#                                            cell.column = 1, feature.column = 1, mtx.transpose = T)
# 
# datmats[["GSE135779"]] <- list()
# files <- list.files(file.path(path_data, geo_acc))
# files_barcodes <- files[grepl("barcodes", files)]
# files_matrices <- files[grepl("matrix", files)]
# for (i in 1:length(files_barcodes)){
#   datmats[["GSE135779"]][[paste(strsplit(files_barcodes[1], "_")[[1]][1:2], collapse = "_")]] <-
#     ReadMtx(mtx = file.path(path_data, 'GSE135779', files_matrices[i]),
#             cells = file.path(path_data, 'GSE135779', files_barcodes[i]),
#             features = file.path(path_data, 'GSE135779', "GSE135779_genes.tsv.gz"),
#             cell.column = 1, feature.column = 1)
# }
# 
# saveRDS(datmats, file = file.path(path_data, "datmats.rds"), compress = F)
# rm(datmats)
# gc()

# Convert(file.path(path_data, "COVID19_Stephenson.h5ad"), dest = file.path(path_data, "COVID19_Stephenson.h5seurat"), overwrite = TRUE)
# obj_COVID19_Stephenson <- LoadH5Seurat(file.path(path_data, "COVID19_Stephenson.h5seurat"))
# saveRDS(obj_COVID19_Stephenson, file = file.path(path_data, "COVID19_Stephenson.rds"), compress = T)
obj <- readRDS(file.path(path_data, "COVID19_Stephenson.rds"))

# obj_COVID19_van_der_Wijst <- readRDS(file.path(path_data, "COVID19_van_der_Wijst.rds"))
# # convert v4 assay to v5 assay
# obj_COVID19_van_der_Wijst[["RNA"]] <- as(object = obj_COVID19_van_der_Wijst[["RNA"]], Class = "Assay5")
```

# Update Seurat object to newest version (v5)

```{r}
obj@version
obj <- CreateSeuratObject(obj@assays$raw$counts,
                          meta.data = obj@meta.data)
obj@version
```


**CHECK**:

1.  Assess available assays: Are raw counts available or only
    pre-normalized (eg log2TMP); if pre-normalized, specify how was the
    normalization performed!

2.  Make sure data matrix has genes as rows and cells as columns

3.  Include, adapt and standarize cell metadata

4.  Modify cell identifiers (`orig.ident`)

5.  Standardize gene symbols

6.  Add features/genes metadata

## 1. Assess available assays

Asses whether raw counts are available or only pre-normalized (eg
log2TMP), if pre-normalized, specify how was the normalization performed
at the entry form (`Expression data type`). If some datasets not include
raw counts consider discarding them, or if normalization approach is
known, de-normalize and normalize data on all datasets. *Data input
format**: e.g. pre-normalized, "by total counts, as described in
<https://doi.org/10.1016/j.cell.2015.04.044>"***

## 2. Make sure data matrix has genes as rows and cells as columns

# Metadata info
```{r}
head(colnames(obj))
paste("Number of genes:", dim(obj)[1])
paste("Number of cells:", dim(obj)[2])
paste("Number of samples:", length(unique(obj$sample_id)))
paste("Number of patients:", length(unique(obj$patient_id)))
paste("Sites:", unique(obj$Site))
paste("Number of cells per sex:")
paste(names(table(obj$Sex)), table(obj$Sex))
paste("Outcome labels:")
paste(unique(obj$Worst_Clinical_Status))
```

## 3. Include, adapt and standarize cell metadata

Generate a metadata dataframe including the following variables for each
cell in the dataset. If some are not present include NA. If needed adapt
metadata (factorize...).

+---------------------------+-------------------------------------+
| Variable                  | Example                             |
+===========================+=====================================+
| Original_cell_names       | AAACCTGAGGTTACCT.Pre_P007_b         |
+---------------------------+-------------------------------------+
| Barcode                   | AAACCTGAGGTTACCT                    |
+---------------------------+-------------------------------------+
| Study                     | Author_Year_PBMID_cohortID          |
|                           |                                     |
|                           | (e.g. ZhangY_2022_34653365)         |
+---------------------------+-------------------------------------+
| Sample                    | Pre_P007_b                          |
+---------------------------+-------------------------------------+
| Treatment                 | Pre                                 |
+---------------------------+-------------------------------------+
| Individual                | P007                                |
+---------------------------+-------------------------------------+
| Tissue                    | blood, tumor, adjacent, LN, other   |
+---------------------------+-------------------------------------+
| O                         | CD4T                                |
| riginalAnnotationLevel{n} |                                     |
+---------------------------+-------------------------------------+


```{r adapt_meta}
# Some barcodes contain the site as a prefix, e.g. site_ATGCATGC... or some other unknown identifier
# IMPORTANT: needs to be left in, otherwise duplicate names
# Maybe some samples were sequenced multiple times?

# Explore data and modifiy this code accordingly to fill the metadata slots
cell_names <- colnames(obj)
obj@meta.data$Original_cell_names <- cell_names
cell_names[grepl("_", cell_names)] <- sapply(strsplit(cell_names[grepl("_", cell_names)], "_"), function(x) x[[2]])
cell_names.split <- strsplit(cell_names, "-")
obj@meta.data$Barcode <- sapply(cell_names.split, function(x) x[[1]])

messy_barcodes <- grepl("_", obj@meta.data$Barcode)
obj@meta.data$Barcode[messy_barcodes] <- sapply(strsplit(obj@meta.data$Barcode[messy_barcodes], "_"), function(x) x[[2]])

obj@meta.data$Study <- ds_name
obj@meta.data$Sample <- obj$sample_id
obj@meta.data$Individual <- obj$patient_id
obj@meta.data$Tissue <- "blood"
rm(cell_names, cell_names.split)
gc()
```

Clarify metadata "Status" and "Swab_results"
```{r}
table(obj$Status, obj$Swab_result)

unique(obj$sample_id[obj$Status == "Covid" & obj$Swab_result == "Negative"])
# Covid are Positive. One sample (BGCV01_CV0025) is Negative but labelled as Covid (maybe other test tested positive for Covid?) -> check label later (re-assign to Non_covid?)
# Healthy are Healthy or tested Negative
# LPS are LPS
# Non_covid are presumably sick but tested Negative (e.g. flu or false negative covid test?)
```


## 4. Modify cell identifiers (`orig.ident`)

To make cell ids compatible with other datasets, append the study ID at
the beginning of the original cell name. If available, also add other
metadata elements such as Individual, tissue or treatment.

```{r modify_cellID}
#fields in prefix are separated by dots
prefix <- paste(obj$Sample, "blood", obj$Sex, obj$Status_on_day_collection_summary, obj$Site, sep=".")  #fields in prefix are separated by dots
cell_ids <- paste(prefix, obj$Original_cell_names, sep = '_') #underscore delimits end of prefix
colnames(obj) <- cell_ids
obj[1:10,1:2]
head(colnames(obj))
rm(prefix)
gc()
```

## 5. Standardize gene symbols

Map gene ids with our reference gene symbols using
`STACAS::StandardizeGeneSymbols()`. Load EnsemblGenes105_Hsa_GRCh38.p13
and EnsemblGenes105_Mmu_GRCm39 retrieved at 14.02.2022 from
[aux](https://github.com/carmonalab/scRNAseq_data_processing/master/aux),
or already in you local directory if you cloned the repository.

```{r load_geneRef}
# load gene symbols dataset
if (species == "mouse") {
   geneRef_file <- "aux/EnsemblGenes105_Mmu_GRCm39.txt.gz"
} else if (species == "human") {
   geneRef_file <- "aux/EnsemblGenes105_Hsa_GRCh38.p13.txt.gz"
}
geneRef <- fread(geneRef_file)
```

```{r Standar_gene_Symbol}
# Standarize gene symbols using STACAS
obj <- STACAS::StandardizeGeneSymbols(obj, geneRef, assay = "RNA", slots = "counts")
```


### Clean cell metadata (custom)

CHECK: cell metadata fields: Sample, Individual, Tissue
c("blood","tumor","adjacent","LN","other"), Study,
OriginalAnnotationLevel1, OriginalAnnotationLevel2,
OriginalAnnotationLevel3... Factorize or change labels if needed.

```{r clean_metadata}
obj$orig.ident <- cell_ids
obj$OriginalAnnotationLevel1 <- obj$initial_clustering
obj$OriginalAnnotationLevel2 <- obj$full_clustering
```

## 6. Normalize data

```{r norm_Data}
obj <- NormalizeData(obj) 
```

# Quality control measurement and filtering

## 1. Ribosomal and mitochondrial content, and sequencing complexity

Use `SignatuR` package, including ribosomal and mitochondrial gene sets,
to compute their relative content. Also, compute the sequencing
complexity by computing the log10 ratio of genes detected per UMI.

```{r define_qc_metrics}
#Get mitochondrial and ribosomal signatures
ribo.genes <- GetSignature(SignatuR$Hs$Compartments$Ribo)
mito.genes <- GetSignature(SignatuR$Hs$Compartments$Mito)

# Compute ribosomal and mitochondrial content and add to Seurat object metadata
obj <- AddMetaData(obj, metadata = PercentageFeatureSet(obj, features = ribo.genes[ribo.genes %in% rownames(obj)]), col.name = "percent.ribo")
obj <- AddMetaData(obj, metadata = PercentageFeatureSet(obj, features = mito.genes[mito.genes %in% rownames(obj)]), col.name = "percent.mito")

# Compute the ratio of number of genes/features and number of counts/UMIs
obj$log10GenesPerUmi <- log10(obj$nFeature_RNA) / 
                                log10(obj$nCount_RNA)
```

## 2. Evaluate QC metrics and filter dataset

QC-based metric is mostly done manually by inspecting the distribution
of the QC metrics. QC metrics can largely change between samples, cell
types included (heterogeneity), sequencing technology, and batch. Take
into account that number of reads (`nCount`) are only informative if raw
counts are available, otherwise they cannot be used of QC filtering.

Here are shown some guidelines for QC-based filtering:

\|--------------------------------------\|---------------------------------------------------------------------\|
\| QC metric \| Guidelines for filtering \|
\|--------------------------------------\|---------------------------------------------------------------------\|
\| Ribosomal content \| \> 0.5% & \< 60% \| \| Mitochondrial content \|
\< 10-20% \| \| Number of genes/features (nFeatures) \| \>400 and custom
upper limit based on your data \| \| Numbrer of counts/UMI (nCount) \|
\>800 and customand custom upper limit based on your data \| \| Ratio
Features/Counts \| \> 0.6 \| \| Special genes \| Custom: Consider remove
some of these genes for downstream analysis \|
\|--------------------------------------\|---------------------------------------------------------------------\|

### CHECK: distributions across samples

```{r qc_plots0, fig.height= 10}
Idents(obj) <- "Sample"
VlnPlot(obj,
        features = c("nFeature_RNA", "nCount_RNA",
                     "percent.ribo","percent.mito",
                     "PTPRC","MALAT1"),
        ncol = 3, pt.size=0)
```

### Custom thresholds

# Standard check
```{r}
probs <- list()
probs$percent.ribo <- quantile(obj$percent.ribo, probs=c(0.001, 0.005, 0.01, 0.02, 0.05, 0.95, 0.98, 0.99, 0.995, 0.999))
probs$percent.mito <- quantile(obj$percent.mito, probs=c(0.001, 0.005, 0.01, 0.02, 0.05, 0.95, 0.98, 0.99, 0.995, 0.999))
probs$nFeature_RNA <- quantile(obj$nFeature_RNA, probs=c(0.001, 0.005, 0.01, 0.02, 0.05, 0.95, 0.98, 0.99, 0.995, 0.999))
probs$nCount_RNA <- quantile(obj$nCount_RNA, probs=c(0.001, 0.005, 0.01, 0.02, 0.05, 0.95, 0.98, 0.99, 0.995, 0.999))
probs$log10GenesPerUmi <- quantile(obj$log10GenesPerUmi, probs=c(0.001, 0.005, 0.01, 0.02, 0.05, 0.95, 0.98, 0.99, 0.995, 0.999))
probs
```

Review after assessing the plots!

```{r set_cutoffs}
cutoffs <- list()
cutoffs[["percent.ribo"]] <- c(min=3,max=60)
cutoffs[["percent.mito"]] <- c(min=0,max=9)
cutoffs[["nFeature_RNA"]] <- c(min=500,max=6000) 
cutoffs[["nCount_RNA"]] <- c(min=600,max=40000)
cutoffs[["log10GenesPerUmi"]] <- 0.75
print(cutoffs)
```

```{r fig.height= 15}
p1 <- VlnPlot(obj, features = "nFeature_RNA", pt.size=0) & geom_hline(yintercept = cutoffs[["nFeature_RNA"]]) & NoLegend()
p2 <- VlnPlot(obj, features = "nCount_RNA", pt.size=0) & geom_hline(yintercept = cutoffs[["nCount_RNA"]]) & NoLegend()
p3 <- VlnPlot(obj, features = "percent.ribo", pt.size=0) & geom_hline(yintercept = cutoffs[["percent.ribo"]]) & NoLegend()
p4 <- VlnPlot(obj, features = "percent.mito", pt.size=0) & geom_hline(yintercept = cutoffs[["percent.mito"]]) & NoLegend()
p5 <- VlnPlot(obj, features = "log10GenesPerUmi", pt.size=0) & geom_hline(yintercept = cutoffs[["log10GenesPerUmi"]]) & NoLegend()
wrap_plots(p1, p2, p3, p4, p5, ncol = 2)
rm(p1, p2, p3, p4, p5)
```

How many cells per sample?
```{r}
sample.size.pre <- sort(table(obj$Sample))
sample.size.pre
barplot(sort(sample.size.pre), las=2)
```

### Filter out outliers & low quality cells (custom)


### Standard check

```{r probs}
for(va in names(cutoffs)){
  cat(va, "\n")
  print(quantile(obj@meta.data[[va]],
                 probs=c(0.001, 0.005, 0.01, 0.02,
                         0.98, 0.99, 0.995, 0.999))
        )
}
```

After having evaluated the QC metrics values and distribution, modify
cutoffs if needed and then proceed to subset your data based on those.
Also, evaluate how many cells are being removed based on your cutoffs.

```{r subset_seurat fig.height= 10}
# store initial number of cells
initialCellNum <- ncol(obj)
message(paste("Original number of cells:", initialCellNum))

# Perform subsetting using cutoffs list and qc metrics considered
obj <- subset(obj, subset = 
                        nFeature_RNA >= cutoffs[["nFeature_RNA"]]["min"] &
                        nFeature_RNA < cutoffs[["nFeature_RNA"]]["max"] & 
                        nCount_RNA   >= cutoffs[["nCount_RNA"]]["min"]   & 
                        nCount_RNA   < cutoffs[["nCount_RNA"]]["max"]   &  
                        percent.ribo >= cutoffs[["percent.ribo"]]["min"] & 
                        percent.ribo < cutoffs[["percent.ribo"]]["max"] &
                        percent.mito >= cutoffs[["percent.mito"]]["min"] &
                        percent.mito < cutoffs[["percent.mito"]]["max"] &
                        log10GenesPerUmi > cutoffs[["log10GenesPerUmi"]]
                      )

# Print and evaluate the cell drop out after filtering
l <- ncol(obj)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)", l, 100*l/initialCellNum))
```

How many cells per sample after QC?
```{r, fig.height = 8, fig.width = 20}
barplot(sort(sample.size.pre), las=2)

sample.size.post <- table(obj$Sample)[names(sample.size.pre)]
sample.size.post
barplot((sample.size.post), las=2)
barplot(sort(sample.size.post), las=2)

sample.size.filterRatio <- round((sample.size.pre - sample.size.post)/sample.size.pre*100,2)
sample.size.filterRatio
barplot(sample.size.filterRatio[names(sort(sample.size.post))], las=2)
```

# Downsample, process dataset and make diagnostic plots

At this point we should only have in our datasets high quality cells. We
will remove samples with very few cells and downsample large ones, in
order to have a similar contribution from all patients and samples.
Downsampling large samples also speeds up downstream computations.

We will save the datasets with all high-quality cells (whole) and
datasets without small samples and downsampling large ones (light).

```{r remove_small_samples}
min.cells.perSample <- 500
max.cells.perSample <- 10000
cells.perSample <- table(obj$Sample)
barplot(sort(cells.perSample))
abline(h=min.cells.perSample)
abline(h=max.cells.perSample)

# Remove small samples
tab <- table(obj$Sample)
keep <- names(tab)[tab > min.cells.perSample]
obj <- subset(obj, Sample %in% keep)

# Print number of samples removed
cat(length(tab[tab <= min.cells.perSample]), "samples removed out of a total of ", length(tab) )

# Print and evaluate the cell drop out after this step
l <- ncol(obj)
message(sprintf("Number of cells after removing low cell count samples: %i (%.2f %% of input)",
                l, 100*l/sum(sample.size.pre)))
```

```{r downsample}
# Downsample
Idents(obj) <- "Sample"
# Set seed for downsampling
set.seed(my_seed)
obj <- subset(obj, downsample = max.cells.perSample)

# Explore the number of cells per sample
cells.perSample <- table(obj$Sample)
barplot(sort(cells.perSample))

# Print and evaluate the cell drop out after this step
l <- ncol(obj)
message(sprintf("Number of cells after QC and removing low cell count samples and downsampling high cell count samples: %i (%.2f %% of input)",
                l, 100*l/sum(sample.size.pre)))
```

# Export standardized datasets

```{r}
saveRDS(obj,file.path(path_output, paste0(ds_name, "_Stephenson_preprocessed.rds")))
```


**IMPORTANT** CHECK: after running this workflow (on a fresh session);
commit/push final version and record in the Google form
(<https://forms.gle/drtCUXZyvTQYU9ih8>) the Github SHA/commit ref (eg
5815f1be791ecef1e0e8dffed12ef3879074a412)
