---
title: "Compare old and current benchmark results"
subtitle: "Bassez diagnostics and targeted Zhang, Pelka, and Stephenson audit"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false
---

# Purpose

This report compares the Bassez entries in the March 26 snapshot
`result_list_old.rds` with the regenerated result bundles used by
`notebooks/benchmark_analysis.rmd`.

The comparison covers the two high-resolution methods used in the benchmark:

- **GloProp**: `feat_mat` is a sample-by-sample distance matrix, not a sample-by-feature matrix.
- **ECODA_authors_HR**: `feat_mat` is the sample-by-cell-type CLR feature matrix.

```{r setup}
repo_root <- normalizePath(getwd(), mustWork = TRUE)
if (!file.exists(file.path(repo_root, "result_list_old.rds"))) {
  repo_root <- normalizePath(file.path(repo_root, ".."), mustWork = TRUE)
}
source(file.path(repo_root, "src/utils/load_all_functions.R"))

old_file <- file.path(repo_root, "result_list_old.rds")
new_composition_file <- file.path(
  repo_root, "data", "benchmark", "results", "Bassez_composition.rds"
)

if (!file.exists(old_file)) stop("Missing old result file: ", old_file)
if (!file.exists(new_composition_file)) {
  stop("Missing regenerated composition bundle: ", new_composition_file)
}

old_result_list <- suppressWarnings(readRDS(old_file))
new_composition <- readRDS(new_composition_file)

old_bassez <- old_result_list[["bmark"]][["Bassez"]]
old_gloprop <- old_bassez[["GloProp"]]
old_ecoda <- old_bassez[["ECODA_authors_HR"]]
new_gloprop <- new_composition[["GloProp"]]
new_ecoda <- new_composition[["ECODA_authors_HR"]]

if (is.null(old_gloprop) || is.null(old_ecoda)) {
  stop("The old result list does not contain both requested Bassez methods.")
}
if (is.null(new_gloprop) || is.null(new_ecoda)) {
  stop("The regenerated composition bundle does not contain both requested methods.")
}

library(ggplot2)
library(patchwork)

fmt <- function(x, digits = 3) {
  formatC(x, format = "f", digits = digits)
}

matrix_dimensions <- function(x) {
  paste(dim(x), collapse = " x ")
}

labels_by_sample <- function(result) {
  labels <- as.character(result[["labels"]])
  sample_ids <- names(result[["labels"]])
  if (is.null(sample_ids)) sample_ids <- rownames(result[["feat_mat"]])
  stats::setNames(labels, sample_ids)
}

show_matrix <- function(x, caption) {
  display <- as.data.frame(as.matrix(x), check.names = FALSE)
  display <- cbind(Sample = rownames(display), display)
  rownames(display) <- NULL
  knitr::kable(
    display,
    format = "html",
    digits = 4,
    caption = caption,
    table.attr = 'class="table table-sm table-striped"'
  )
}

compare_glo <- function(old_result, new_result) {
  old_ids <- rownames(old_result[["feat_mat"]])
  new_ids <- rownames(new_result[["feat_mat"]])
  common <- intersect(old_ids, new_ids)
  old_mat <- as.matrix(old_result[["feat_mat"]][common, common, drop = FALSE])
  new_mat <- as.matrix(new_result[["feat_mat"]][common, common, drop = FALSE])
  delta <- new_mat - old_mat
  off_diagonal <- upper.tri(delta)
  old_values <- old_mat[off_diagonal]
  new_values <- new_mat[off_diagonal]
  label_old <- labels_by_sample(old_result)
  label_new <- labels_by_sample(new_result)

  list(
    old_dim = matrix_dimensions(old_result[["feat_mat"]]),
    new_dim = matrix_dimensions(new_result[["feat_mat"]]),
    common_samples = common,
    only_old = setdiff(old_ids, new_ids),
    only_new = setdiff(new_ids, old_ids),
    labels_equal = all(label_old[common] == label_new[common]),
    max_abs = max(abs(delta[off_diagonal])),
    mae = mean(abs(delta[off_diagonal])),
    rmse = sqrt(mean(delta[off_diagonal]^2)),
    correlation = stats::cor(old_values, new_values)
  )
}

compare_ecoda <- function(old_result, new_result) {
  old_ids <- rownames(old_result[["feat_mat"]])
  new_ids <- rownames(new_result[["feat_mat"]])
  common_samples <- intersect(old_ids, new_ids)
  old_features <- colnames(old_result[["feat_mat"]])
  new_features <- colnames(new_result[["feat_mat"]])
  common_features <- intersect(old_features, new_features)
  old_mat <- as.matrix(old_result[["feat_mat"]][common_samples, common_features, drop = FALSE])
  new_mat <- as.matrix(new_result[["feat_mat"]][common_samples, common_features, drop = FALSE])
  delta <- new_mat - old_mat
  label_old <- labels_by_sample(old_result)
  label_new <- labels_by_sample(new_result)

  list(
    old_dim = matrix_dimensions(old_result[["feat_mat"]]),
    new_dim = matrix_dimensions(new_result[["feat_mat"]]),
    common_samples = common_samples,
    only_old_samples = setdiff(old_ids, new_ids),
    only_new_samples = setdiff(new_ids, old_ids),
    common_features = common_features,
    only_old_features = setdiff(old_features, new_features),
    only_new_features = setdiff(new_features, old_features),
    labels_equal = all(label_old[common_samples] == label_new[common_samples]),
    max_abs = max(abs(delta)),
    mae = mean(abs(delta)),
    rmse = sqrt(mean(delta^2)),
    correlation = stats::cor(as.vector(old_mat), as.vector(new_mat))
  )
}

glo_comparison <- compare_glo(old_gloprop, new_gloprop)
ecoda_comparison <- compare_ecoda(old_ecoda, new_ecoda)

# These are the exact relationships used by the benchmark code. GloProp stores
# sqrt(gloscopeProp(...)) as feat_mat and then wraps it with as.dist(); ECODA
# computes dist(feat_mat).
stopifnot(
  max(abs(
    as.numeric(old_gloprop[["dist_mat"]]) -
      as.numeric(as.dist(old_gloprop[["feat_mat"]]))
  )) < 1e-10,
  max(abs(
    as.numeric(new_gloprop[["dist_mat"]]) -
      as.numeric(as.dist(new_gloprop[["feat_mat"]]))
  )) < 1e-10,
  max(abs(
    as.numeric(old_ecoda[["dist_mat"]]) -
      as.numeric(stats::dist(old_ecoda[["feat_mat"]]))
  )) < 1e-10,
  max(abs(
    as.numeric(new_ecoda[["dist_mat"]]) -
      as.numeric(stats::dist(new_ecoda[["feat_mat"]]))
  )) < 1e-10
)
```

# Result-bundle layout

The regenerated bundle is `data/benchmark/results/Bassez_composition.rds`,
which is the composition bundle loaded by the current benchmark pipeline. The
old object is accessed as `result_list_old.rds$bmark$Bassez`.

```{r bundle-summary}
bundle_summary <- data.frame(
  Method = c("GloProp (sample distances)", "ECODA_authors_HR (CLR features)"),
  Old_feat_mat = c(
    glo_comparison$old_dim,
    ecoda_comparison$old_dim
  ),
  New_feat_mat = c(
    glo_comparison$new_dim,
    ecoda_comparison$new_dim
  ),
  Old_dist_mat = c(
    matrix_dimensions(as.matrix(old_gloprop[["dist_mat"]])),
    matrix_dimensions(as.matrix(old_ecoda[["dist_mat"]]))
  ),
  New_dist_mat = c(
    matrix_dimensions(as.matrix(new_gloprop[["dist_mat"]])),
    matrix_dimensions(as.matrix(new_ecoda[["dist_mat"]]))
  ),
  row.names = NULL,
  check.names = FALSE
)
knitr::kable(bundle_summary, format = "html", caption = "Compared result-bundle dimensions.")
```

# GloProp

## What `feat_mat` means for GloProp

The current implementation calls `gloscopeProp(...)`, takes its square root,
and stores that sample-by-sample matrix as `feat_mat`. It stores
`as.dist(feat_mat)` as `dist_mat`. Therefore, the GloProp `feat_mat` below is
already a distance matrix; it should not be passed to `dist()` a second time.
The MDS panels use `as.dist(feat_mat)`, exactly as the benchmark does.

## Old GloProp `feat_mat`

```{r old-gloprop-matrix, results='asis'}
show_matrix(
  old_gloprop[["feat_mat"]],
  "Old Bassez GloProp feat_mat (March 26 snapshot; sqrt pairwise distances)."
)
```

## Regenerated GloProp `feat_mat`

```{r new-gloprop-matrix, results='asis'}
show_matrix(
  new_gloprop[["feat_mat"]],
  "Regenerated Bassez GloProp feat_mat (current composition bundle; sqrt pairwise distances)."
)
```

## GloProp comparison diagnostics

```{r gloprop-diagnostics}
glo_diag <- data.frame(
  Quantity = c(
    "Old / new matrix dimensions",
    "Common samples",
    "Samples only in old matrix",
    "Samples only in regenerated matrix",
    "Labels equal on common samples",
    "Shared off-diagonal max absolute change",
    "Shared off-diagonal mean absolute change",
    "Shared off-diagonal RMSE",
    "Shared off-diagonal correlation"
  ),
  Value = c(
    paste(glo_comparison$old_dim, "/", glo_comparison$new_dim),
    length(glo_comparison$common_samples),
    if (length(glo_comparison$only_old) == 0) "none" else paste(glo_comparison$only_old, collapse = ", "),
    if (length(glo_comparison$only_new) == 0) "none" else paste(glo_comparison$only_new, collapse = ", "),
    glo_comparison$labels_equal,
    fmt(glo_comparison$max_abs),
    fmt(glo_comparison$mae),
    fmt(glo_comparison$rmse),
    fmt(glo_comparison$correlation)
  ),
  row.names = NULL,
  check.names = FALSE
)
knitr::kable(glo_diag, format = "html", caption = "GloProp old-versus-regenerated diagnostics.")
```

# ECODA high-resolution composition

## Old ECODA `feat_mat`

```{r old-ecoda-matrix, results='asis'}
show_matrix(
  old_ecoda[["feat_mat"]],
  "Old Bassez ECODA_authors_HR feat_mat (sample x cell-type CLR matrix)."
)
```

## Regenerated ECODA `feat_mat`

```{r new-ecoda-matrix, results='asis'}
show_matrix(
  new_ecoda[["feat_mat"]],
  "Regenerated Bassez ECODA_authors_HR feat_mat (sample x cell-type CLR matrix)."
)
```

## ECODA comparison diagnostics

```{r ecoda-diagnostics}
ecoda_diag <- data.frame(
  Quantity = c(
    "Old / new matrix dimensions",
    "Common samples",
    "Samples only in old matrix",
    "Samples only in regenerated matrix",
    "Cell types only in old matrix",
    "Cell types only in regenerated matrix",
    "Labels equal on common samples",
    "Shared-value max absolute change",
    "Shared-value mean absolute change",
    "Shared-value RMSE",
    "Shared-value correlation"
  ),
  Value = c(
    paste(ecoda_comparison$old_dim, "/", ecoda_comparison$new_dim),
    length(ecoda_comparison$common_samples),
    if (length(ecoda_comparison$only_old_samples) == 0) "none" else paste(ecoda_comparison$only_old_samples, collapse = ", "),
    if (length(ecoda_comparison$only_new_samples) == 0) "none" else paste(ecoda_comparison$only_new_samples, collapse = ", "),
    if (length(ecoda_comparison$only_old_features) == 0) "none" else paste(ecoda_comparison$only_old_features, collapse = ", "),
    paste(ecoda_comparison$only_new_features, collapse = ", "),
    ecoda_comparison$labels_equal,
    fmt(ecoda_comparison$max_abs),
    fmt(ecoda_comparison$mae),
    fmt(ecoda_comparison$rmse),
    fmt(ecoda_comparison$correlation)
  ),
  row.names = NULL,
  check.names = FALSE
)
knitr::kable(ecoda_diag, format = "html", caption = "ECODA old-versus-regenerated diagnostics.")
```

The regenerated bundle's stored `counts` table makes the feature-set change
explicit. The seven regenerated-only cell types are broad `cellType` labels
that were not present as columns in the old high-resolution composition.

```{r regenerated-only-cell-types}
new_only_totals <- data.frame(
  cell_type = ecoda_comparison$only_new_features,
  regenerated_counts_total = as.numeric(
    colSums(new_ecoda[["counts"]][, ecoda_comparison$only_new_features, drop = FALSE])
  ),
  row.names = NULL,
  check.names = FALSE
)
knitr::kable(
  new_only_totals,
  format = "html",
  digits = 1,
  caption = "Nonzero regenerated-only cell-type composition entries (stored counts table)."
)
```

# MDS comparison

The four panels below use the same MDS calculation as `plot_mds()`:

- GloProp: `cmdscale(as.dist(feat_mat))`;
- ECODA: `cmdscale(dist(feat_mat))`, equivalently its stored `dist_mat`.

MDS axes can rotate or flip sign between independent fits. The meaningful
comparison is the pairwise geometry and the responder/non-responder pattern,
not the sign or absolute orientation of an axis.

```{r mds-comparison, fig.width=12, fig.height=9}
plot_mds_comparison <- function(result, title, feature_is_distance = FALSE) {
  dist_mat <- if (feature_is_distance) {
    stats::as.dist(result[["feat_mat"]])
  } else {
    result[["dist_mat"]]
  }
  mds <- stats::cmdscale(dist_mat, k = 2, eig = TRUE)
  positive_eigenvalues <- mds$eig[mds$eig > 0]
  variance_percent <- 100 * mds$eig[1:2] / sum(positive_eigenvalues)

  labels <- labels_by_sample(result)
  labels <- labels[rownames(mds$points)]
  labels[labels == "E"] <- "Responders"
  labels[labels == "NE"] <- "Non-responders"
  labels <- factor(labels, levels = c("Responders", "Non-responders"))

  plot_data <- data.frame(
    Sample = rownames(mds$points),
    MDS1 = mds$points[, 1],
    MDS2 = mds$points[, 2],
    Group = labels,
    row.names = NULL
  )

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(x = MDS1, y = MDS2, color = Group, shape = Group)
  ) +
    ggplot2::geom_point(size = 2.8) +
    ggplot2::scale_color_manual(
      values = c("Responders" = "#D55E00", "Non-responders" = "#0072B2"),
      drop = FALSE
    ) +
    ggplot2::scale_shape_manual(
      values = c("Responders" = 16, "Non-responders" = 17),
      drop = FALSE
    ) +
    ggplot2::labs(
      title = paste0(title, " (n = ", nrow(plot_data), ")"),
      x = paste0("MDS dimension 1 (", fmt(variance_percent[1], 1), "%)"),
      y = paste0("MDS dimension 2 (", fmt(variance_percent[2], 1), "%)"),
      color = "Treatment response",
      shape = "Treatment response"
    ) +
    ggplot2::theme_classic() +
    ggplot2::coord_equal()
}

patchwork::wrap_plots(
  plot_mds_comparison(old_gloprop, "Old GloProp", feature_is_distance = TRUE),
  plot_mds_comparison(new_gloprop, "Regenerated GloProp", feature_is_distance = TRUE),
  plot_mds_comparison(old_ecoda, "Old ECODA_authors_HR"),
  plot_mds_comparison(new_ecoda, "Regenerated ECODA_authors_HR"),
  ncol = 2
)
```


# Targeted old/current method audit

The targeted audit below keeps the comparison deliberately narrow: Zhang
scPoli/PILOT, Pelka MrVI, and the restored Stephenson ordinary benchmark scope.
It recomputes current Python scores from the released Feather inputs, compares
sample IDs with the old result snapshot, reports current annotation coverage,
and reads the current benchmark registry.

```{r targeted-audit}
current_metadata <- function(dataset) {
  readRDS(file.path(
    repo_root,
    "data",
    "benchmark",
    "results",
    paste0(dataset, "_metadata.rds")
  ))
}

score_value <- function(result, name, fallback = NULL) {
  value <- result[["scores"]][[name]]
  if (is.null(value) && !is.null(fallback)) {
    value <- result[["scores"]][[fallback]]
  }
  if (is.null(value)) NA_real_ else as.numeric(value)
}

sample_id_summary <- function(old_result, current_result) {
  old_ids <- rownames(old_result[["feat_mat"]])
  current_ids <- rownames(current_result[["feat_mat"]])
  common <- intersect(old_ids, current_ids)
  list(
    old_n = length(old_ids),
    current_n = length(current_ids),
    shared_n = length(common),
    only_old = setdiff(old_ids, current_ids),
    only_current = setdiff(current_ids, old_ids)
  )
}

targeted_cases <- list(
  list(
    dataset = "Zhang",
    method = "scPoli_hvg2000_dims15_highres",
    old_key = "scPoli_hvg2000_dims15_highres",
    current_file = file.path(
      repo_root,
      "data",
      "benchmark",
      "embeddings",
      "Zhang_hvg2000_highres_scpoli_dims15_embs.feather"
    ),
    current_reader = process_scpoli_fig
  ),
  list(
    dataset = "Zhang",
    method = "PILOT_hvg2000_highres",
    old_key = "PILOT_hvg2000",
    current_file = file.path(
      repo_root,
      "data",
      "benchmark",
      "embeddings",
      "Zhang_hvg2000_highres_pilot_dists.feather"
    ),
    current_reader = process_pilot_fig
  ),
  list(
    dataset = "Pelka",
    method = "MrVI_hvg2000",
    old_key = "MrVI_hvg2000",
    current_file = file.path(
      repo_root,
      "data",
      "benchmark",
      "embeddings",
      "Pelka_hvg2000_mrvi_dists.feather"
    ),
    current_reader = process_mrvi_fig
  ),
  list(
    dataset = "Stephenson",
    method = "ECODA_authors_HR",
    old_key = "ECODA_authors_HR",
    current_file = file.path(
      repo_root,
      "data",
      "benchmark",
      "results",
      "Stephenson_ECODA_authors_HR.rds"
    ),
    current_reader = function(path, labels) readRDS(path)
  )
)

targeted_results <- lapply(targeted_cases, function(case) {
  old_result <- old_result_list[["bmark"]][[case$dataset]][[case$old_key]]
  if (is.null(old_result)) {
    stop("Missing old targeted result: ", case$dataset, " / ", case$old_key)
  }
  if (!file.exists(case$current_file)) {
    stop("Missing current targeted artifact: ", case$current_file)
  }
  labels <- current_metadata(case$dataset)[["labels"]]
  current_result <- case$current_reader(case$current_file, labels)
  ids <- sample_id_summary(old_result, current_result)
  data.frame(
    Dataset = case$dataset,
    Method = case$method,
    Old_ANOSIM = score_value(old_result, "anosim_score"),
    Current_ANOSIM = score_value(current_result, "anosim_score"),
    Old_Cluster = score_value(old_result, "cluster_score"),
    Current_Cluster = score_value(current_result, "cluster_score"),
    Old_samples = ids$old_n,
    Current_samples = ids$current_n,
    Shared_samples = ids$shared_n,
    Only_old = if (length(ids$only_old) == 0) "none" else paste(ids$only_old, collapse = ", "),
    Only_current = if (length(ids$only_current) == 0) "none" else paste(ids$only_current, collapse = ", "),
    row.names = NULL,
    check.names = FALSE
  )
})
targeted_score_table <- do.call(rbind, targeted_results)
knitr::kable(
  targeted_score_table,
  format = "html",
  digits = 4,
  caption = "Targeted old/current score and sample-universe comparison."
)

metadata_coverage <- do.call(rbind, lapply(
  c("Zhang", "Pelka", "Stephenson"),
  function(dataset) {
    metadata <- current_metadata(dataset)
    data.frame(
      Dataset = dataset,
      Current_cells = as.integer(metadata[["n_cells"]]),
      Current_samples = as.integer(metadata[["n_samples"]]),
      High_resolution_levels = as.integer(metadata[["n_cell_types_high_res"]]),
      row.names = NULL
    )
  }
))

read_h5ad_obs <- function(path) {
  if (!file.exists(path)) stop("Missing current processed h5ad: ", path)
  anndata <- reticulate::import("anndata", convert = FALSE)
  handle <- anndata$read_h5ad(path, backed = "r")
  on.exit(handle$file$close(), add = TRUE)
  reticulate::py_to_r(handle$obs)
}

zhang_obs <- read_h5ad_obs(file.path(
  repo_root,
  "data",
  "Zhang",
  "output",
  "ZhangY_2022_34653365_benchmark_analysis_ECODAprocessed.h5ad"
))
zhang_layer2 <- zhang_obs[["layer2"]]
annotation_coverage <- data.frame(
  Dataset = "Zhang",
  Annotation_column = "layer2",
  Total_cells = length(zhang_layer2),
  Valid_nonmissing = sum(!is.na(zhang_layer2)),
  Missing = sum(is.na(zhang_layer2)),
  row.names = NULL
)
knitr::kable(
  metadata_coverage,
  format = "html",
  caption = "Current sample and high-resolution level coverage."
)
knitr::kable(
  annotation_coverage,
  format = "html",
  caption = "Current Zhang layer2 cell-level annotation coverage."
)

registry_dataset_scope <- data.frame(
  Dataset = names(read_datasets_json(view = "benchmark_analysis")),
  row.names = NULL
)
knitr::kable(
  registry_dataset_scope,
  format = "html",
  caption = "Datasets currently declared for ordinary benchmark analysis."
)
```

# Why are the results different?

1. **The regenerated run contains two additional samples.** GloProp and
   ECODA both have 26 rows in the old snapshot and 28 rows in the regenerated
   bundle. The recovered samples are
   `r paste(glo_comparison$only_new, collapse = ", ")`. The labels for the 26
   shared samples are identical, so label recoding is not the explanation.

2. **The Bassez high-resolution annotation contract changed.** The current
   preprocessing repair fills missing/unknown `cellSubType` values with their
   corresponding broad `cellType` values before benchmark analysis
   (`src/2_dataset_specific_preprocessing/1.6.1_fill_bassez_cellsubtype.R`).
   The regenerated ECODA matrix consequently has 40 cell-type columns instead
   of the old 33. Its seven additional columns are
   `r paste(ecoda_comparison$only_new_features, collapse = ", ")`.
   Those columns have nonzero entries in the regenerated stored composition
   counts, so they are not cosmetic column additions.

3. **CLR values must change when the composition feature set changes.** ECODA
   recomputes the CLR from the updated sample-by-cell-type composition. Adding
   broad cell-type components changes the geometric mean used by every row.
   That is why the shared 33 CLR columns remain highly correlated
   (`r fmt(ecoda_comparison$correlation)`) but are not numerically identical.
   The updated rows also include the two recovered samples.

4. **GloProp is sensitive to the same cell-level annotation input.** Its
   stored matrix is a transformed pairwise distance matrix, so filtering
   invalid annotations, filling missing subtypes, and recovering samples alter
   many pairwise entries. Among the 26 shared samples, the GloProp distances
   have correlation `r fmt(glo_comparison$correlation)`, mean absolute change
   `r fmt(glo_comparison$mae)`, and maximum absolute change
   `r fmt(glo_comparison$max_abs)`. This demonstrates that the difference is
   not only the two extra rows/columns.

5. **MDS is not itself the source of the discrepancy.** Both old and new
   panels are recomputed from their stored pairwise distances. MDS can rotate
   or reflect an embedding, but it cannot explain changed sample distances;
   the changed input matrices do.

The short version: the March 26 object is a result from the older Bassez
annotation/sample universe. The current pipeline uses the repaired high-resolution
annotation, filters invalid annotation sentinels, and reruns the composition
methods. The resulting GloProp distance matrix and ECODA CLR feature matrix are
therefore expected to differ.
