---
title: "Compositional data analysis of the Gong & Sharma et al. dataset"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---


# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

library(pROC)
source("functions.R")
source("Processed_dataset_metadata.R") # Loads "datasets" variable

my_seed <- 123
set.seed(my_seed)
```

# Set paths

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets")
path_root <- file.path(path_nas, ds_name)
# path_data <- file.path(path_root, "data")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")
```

# Define functions

```{r}
library(scales)

# Function to add the log2 scale to a plot
add_log_scale <- function(p, scale = "log2") {
  p + scale_x_continuous(trans = scale)
}

plot_results <- function(results, n_ct_to_select, log_perturbation = FALSE) {
  plot_list <- list()
  for (nct in n_ct_to_select) {
    res_all <- results[results$n_celltypes == nct, ]
    res_avg <- res_all %>%
      group_by(trans_method, Amount_of_perturbation) %>%
      mutate(Silhouette_score = mean(Silhouette_score),
             ANOSIM_score = mean(ANOSIM_score),
             Modularity_score = mean(Modularity_score),
             Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
      ungroup() %>% 
      distinct(interaction(trans_method, Amount_of_perturbation), .keep_all=TRUE)
    
    p_list <- list()
    p_list[["plot_sil"]] <- ggplot() +
      geom_line(data = res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method), size = 1.2) +
      ylim(c(min(res_avg$Silhouette_score), 1)) +
      theme_classic() +
      # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method)) +
      ggtitle(paste0("Differential cell types: ", nct, "\nSilhouette score"))
    p_list[["plot_mod"]] <- ggplot() +
      geom_line(data = res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method), size = 1.2) +
      ylim(c(min(res_avg$Modularity_score), 1)) +
      theme_classic() +
      # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method)) +
      ggtitle(paste0("Differential cell types: ", nct, "\nModularity score"))
    p_list[["plot_clust"]] <- ggplot() +
      geom_line(data = res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = trans_method), size = 1.2) +
      ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
      theme_classic() +
      # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method)) +
      ggtitle(paste0("Differential cell types: ", nct, "\nClustering score"))
    p_list[["plot_anosim"]] <- ggplot() +
      geom_line(data = res_avg, aes(x = Amount_of_perturbation, y = ANOSIM_score, color = trans_method), size = 1.2) +
      ylim(c(min(res_avg$ANOSIM_score), 1)) +
      theme_classic() +
      # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method)) +
      ggtitle(paste0("Differential cell types: ", nct, "\nANOSIM score"))
    
    if (log_perturbation) {
      p_list <- lapply(p_list, add_log_scale)
    }
    
    p <- ggarrange(plotlist = p_list, nrow=1, common.legend = TRUE)
    plot_list[[as.character(nct)]] <- p
  }
  
  return(plot_list)
}
```




# Load dataset
Manually downloaded from:
"<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>"

```{r}
files <- list(L1 = file.path(path_data, "sound_life_AIFI_L1_frequencies.csv"),
              L2 = file.path(path_data, "sound_life_AIFI_L2_frequencies.csv"),
              L3 = file.path(path_data, "sound_life_AIFI_L3_frequencies.csv"))

data <- list()

for (f in names(files)) {
  data[[f]] <- read.csv(files[[f]])
}
```



# Process data into wide dataframe

```{r}
dproc <- list()

for (layer in names(data)) {

  dproc$meta <- data[[layer]] %>%
    dplyr::select(c(1:14)) %>%
    unique() %>% 
    mutate_at(1:14, as.factor)
  
  ctable <- data[[layer]] %>%
    dplyr::select(c(14,
             !!sym(paste0("AIFI_", layer)),
             !!sym(paste0("AIFI_", layer, "_count")),
             !!sym(paste0("AIFI_", layer, "_frac_total")),
             !!sym(paste0("AIFI_", layer, "_clr")))) %>%
    pivot_wider(names_from = !!sym(paste0("AIFI_", layer)),
                values_from = c(!!sym(paste0("AIFI_", layer, "_count")),
                                !!sym(paste0("AIFI_", layer, "_frac_total")),
                                !!sym(paste0("AIFI_", layer, "_clr")))) %>% 
    column_to_rownames(var = "specimen.specimenGuid")
  
  i <- 1
  n_cts <- length(unique(data[[layer]][[paste0("AIFI_", layer)]]))
  
  dproc[[layer]]$cell_counts <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_count_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)

  i <- i + n_cts
  dproc[[layer]]$freq <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_frac_total_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)
  i <- i + n_cts
  dproc[[layer]]$clr_authors <- ctable[, i:(i + n_cts - 1)] %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_clr_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    dplyr::select(-Platelet, -Erythrocyte)
  
  df_counts <- dproc[[layer]]$cell_counts

  df_counts_1count_zero <- df_counts
  df_counts_1count_zero[df_counts_1count_zero == 0] <- 1
  dproc[[layer]]$clr_1count_zero <- df_counts_1count_zero %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_1count_all <- (df_counts + 1) %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_0.1perc <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()
}

# saveRDS(dproc, file.path(path_data, "cell_type_compositions.rds"))

n_cts_tot <- dim(dproc[[layer]]$clr_0.1perc)[2]
```



# ---------------------------------------------------------------

# Sensitivity: freq vs clr vs alr vs ilr

CMV negative BR1 female samples are used, as the biggest sources of variation has been removed (subset to specific groups according to major axis of variance).
Subsequently, one or multiple cell types will be artificially increased/decreased to simulate differential abundance.


## L1 anotations


```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L1"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.1, 1.2, 1.5, 2, 4, 8, 16, 32, 64, 128, 256)
n_ct_to_select <- c(1, 2, 3)

results_l1 <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select)
```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l1, n_ct_to_select, log_perturbation = TRUE)
print(p)
```


### Show example PCA

```{r fig.height=4, fig.width=12}
nct <- 2
da <- 4
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts + 1
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]


plot_list <- list()

plot_list[["Counts"]] <- plot_pca(df_counts_temp, labels, title = "Counts", sil_score = TRUE) + theme(aspect.ratio=1)
plot_list[["Frequency"]] <- plot_pca(df_freq, labels, title = "Frequency (%)", sil_score = TRUE) + theme(aspect.ratio=1)
data <- asin(sqrt(df_freq/100))
plot_list[["asinsqrt"]] <- plot_pca(data, labels, title = "asinsqrt", sil_score = TRUE) + theme(aspect.ratio=1)
data <- clr(df_freq)
plot_list[["CLR"]] <- plot_pca(data, labels, title = "CLR", sil_score = TRUE) + theme(aspect.ratio=1)

p <- ggarrange(plotlist = plot_list, nrow=1, common.legend = TRUE)
p
# ggsave(file.path(path_plots, paste0("semisynthetic_L1_PCA_example_nct", nct, "_da", da, ".pdf")),
#        width = 10,
#        height = 10)
```




## L2 anotations

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L2"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.1, 1.25, 1.5, 2, 4, 8, 16, 32, 64, 128)
n_ct_to_select <- c(1, 3, 5, 10)

results_l2 <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select)
```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l2, n_ct_to_select, log_perturbation = TRUE)
print(p)
```


### Show example PCA
#### Why Silhouette can be misleading

```{r fig.height=4, fig.width=12}
set.seed(1)

nct <- 5
da <- 8
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts + 1
ct_da <- c("Memory_B_cell", "Proliferating_NK_cell", "pDC", "ASDC", "MAIT")
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]


plot_list <- list()

plot_list[["Counts"]] <- plot_pca(df_counts_temp, labels, title = "Counts", sil_score = TRUE) + theme(aspect.ratio=1)
plot_list[["Frequency"]] <- plot_pca(df_freq, labels, title = "Frequency (%)", sil_score = TRUE) + theme(aspect.ratio=1)
data <- asin(sqrt(df_freq/100))
plot_list[["asinsqrt"]] <- plot_pca(data, labels, title = "asinsqrt", sil_score = TRUE) + theme(aspect.ratio=1)
data <- clr(df_freq)
plot_list[["CLR"]] <- plot_pca(data, labels, title = "CLR", sil_score = TRUE) + theme(aspect.ratio=1)

p <- ggarrange(plotlist = plot_list, nrow=1, common.legend = TRUE)
p
ggsave(file.path(path_plots, paste0("semisynthetic_L2_PCA_example_WhySilhouetteCanBeMisleading.pdf")),
       height = 4,
       width = 12)
```


#### Random

```{r fig.height=4, fig.width=12}
nct <- 5
da <- 8
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts + 1
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]


plot_list <- list()

plot_list[["Counts"]] <- plot_pca(df_counts_temp, labels, title = "Counts", sil_score = TRUE) + theme(aspect.ratio=1)
plot_list[["Frequency"]] <- plot_pca(df_freq, labels, title = "Frequency (%)", sil_score = TRUE) + theme(aspect.ratio=1)
data <- asin(sqrt(df_freq/100))
plot_list[["asinsqrt"]] <- plot_pca(data, labels, title = "asinsqrt", sil_score = TRUE) + theme(aspect.ratio=1)
data <- clr(df_freq)
plot_list[["CLR"]] <- plot_pca(data, labels, title = "CLR", sil_score = TRUE) + theme(aspect.ratio=1)

p <- ggarrange(plotlist = plot_list, nrow=1, common.legend = TRUE)
p
```



### Mean-variance relationship

```{r}
df_freq_unperturbed <- df_counts %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()
plot(colMeans(df_freq_unperturbed), main = "colMeans(df_freq_unperturbed)")
plot(Rfast::colVars(df_freq_unperturbed), main = "colVars(df_freq_unperturbed)")
plot(colMeans(df_freq_unperturbed), Rfast::colVars(df_freq_unperturbed))

means <- colMeans(clr(df_freq))
vars <- Rfast::colVars(data.frame(clr(df_freq)))

df1 <- data.frame(mean = means, variance = vars)

ggplot(df1, aes(x = mean, y = variance)) +
  geom_point() +
  geom_text(data = df1, aes(label = row.names(df1)), vjust = -0.5)

```



```{r}
means <- colMeans((df_clr))
vars <- Rfast::colVars(data.frame((df_clr)))

df1 <- data.frame(mean = means, variance = vars)

ggplot(df1, aes(x = mean, y = variance)) +
  geom_point() +
  geom_text(data = df1, aes(label = row.names(df1)), vjust = -0.5)


test_var <- get_ct_var(df_clr)

test1 <- get_hvcs(test_var, variance_threshold = 0.8)
test2 <- get_hvcs(test_var, top_n_hvcs = 0.25)
```





## L2 annotations (CLR vs CLR_PCA)

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L2"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")


# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.1, 1.25, 1.5, 2, 3, 4, 6, 8, 16, 32, 64, 128)
n_ct_to_select <- c(1, 3, 5, 10)

results_l2_clr <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select,
                          trans_method = c(
                            "arcsine_sqrt",
                            "arcsine_sqrt_pca",
                            "clr",
                            "clr_pca"
                          )
                          )
```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l2_clr, n_ct_to_select, log_perturbation = TRUE)
print(p)
```




## L2 annotations only low abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L2"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128)
n_ct_to_select <- c(1, 3, 5, 10)

ct_counts <- colMeans(df_counts)
# Find the threshold for the lowest fourth
threshold <- quantile(ct_counts, 1/4)

# Get the column names with means below the threshold
cts <- colnames(df_counts)[ct_counts < threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results_l2_OnlyLowAbundantCts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select, cts = cts)

```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l2_OnlyLowAbundantCts, n_ct_to_select, log_perturbation = TRUE)
print(p)
```



### Show example PCA

```{r fig.height=4, fig.width=16}
nct <- 5
da <- 8
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(cts, size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]

plot_list <- list()

data <- prcomp(df_counts_temp)$x[, 1:2]
p <- plot_pca(df_counts_temp, labels, title = "Counts PCA",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))

data <- prcomp(df_freq)$x[, 1:2]
p <- plot_pca(df_freq, labels, title = "Freq PCA",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))

data <- asin(sqrt(df_freq/100))
data <- prcomp(data)$x[, 1:2]
p <- plot_pca(data, labels, title = "asinsqrt PCA",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))

df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()
data <- clr(df_freq)
data <- prcomp(data)$x[, 1:2]
p <- plot_pca(data, labels, title = "CLR PCA",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE)
plot_list <- append(plot_list, list(p))


p_wrapped <- patchwork::wrap_plots(plot_list, nrow = 1)
p_wrapped
```


## L2 annotations only highly abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L2"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")


# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128)
n_ct_to_select <- c(1, 3, 5, 10)

ct_counts <- colMeans(df_counts)
# Find the threshold for the highest fourth
threshold <- quantile(ct_counts, 3/4)

# Get the column names with means above the threshold
cts <- colnames(df_counts)[ct_counts >= threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results_l2_OnlyHighAbundantCts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select, cts = cts)
```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l2_OnlyHighAbundantCts, n_ct_to_select, log_perturbation = TRUE)
print(p)
```





## L3 annotations (CLR vs CLR_PCA)

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

# df_counts <- df_counts[1:30, 30:ncol(df_counts)]
df_counts <- df_counts[1:30,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# df_counts <- df_counts[, !colnames(df_counts) %in% c("KLRF1minus_GZMBplus_CD27minus_EM_CD8_T_cell", "GZMBplus_Vd2_gdT", "KLRF1minus_GZMBplus_CD27minus_memory_CD4_T_cell")]
# df_counts <- df_counts[, !colnames(df_counts) %in% max_cv_cts]

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128, 256)
n_ct_to_select <- c(1, 3, 5, 10, 20, round(ncol(df_counts)/2))

results_l3_clr <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select,
                          trans_method = c(
                            "arcsine_sqrt",
                            "arcsine_sqrt_pca",
                            "clr",
                            "clr_pca"
                          )
                          )
```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l3_clr, n_ct_to_select, log_perturbation = TRUE)
print(p)
```


# Test different groups

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

df_counts <- df_counts[1:30, ]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128, 320)
n_ct_to_select <- c(1, 3, 5, 10, 20)

results_l3_pca <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select)
```

### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l3_pca, n_ct_to_select, log_perturbation = TRUE)
print(p)
```


### Mean-variance relationship

```{r}
df_freq_unperturbed <- df_counts %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()
plot(sort(rowSums(df_counts)), xlab = "Number of cells", ylab = "Samples", main = "Number of total cells per sample")
plot(sort(Rfast::colVars(df_freq_unperturbed)))
plot(sort(log(Rfast::colVars(df_freq_unperturbed))))

plot_data <- data.frame(
  mean = colMeans(df_freq_unperturbed),
  variance = Rfast::colVars(df_freq_unperturbed)
)
# Plot using ggplot2
ggplot(plot_data, aes(x = mean, y = variance)) +
  geom_point() +
  labs(
    x = "Mean (in %)",
    y = "Variance (in %)"
  ) +
  theme_classic2() # Optional: a clean theme
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_percent.pdf"), width = 3, height = 2.5)
get_ct_var(df_freq_unperturbed, smooth_method = "gam")
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_percent_fit.pdf"), width = 3, height = 2.5)


plot_data <- data.frame(
  mean = log(colMeans(df_freq_unperturbed)),
  variance = log(Rfast::colVars(df_freq_unperturbed))
)
# Plot using ggplot2
ggplot(plot_data, aes(x = mean, y = variance)) +
  geom_point() +
  # geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "red") +
  labs(
    x = "log( Mean (in %) )",
    y = "log( Variance (in %) )"
  ) +
  theme_classic2() # Optional: a clean theme
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_percentLogLog.pdf"), width = 3, height = 2.5)


means <- colMeans(clr(df_freq_unperturbed))
vars <- Rfast::colVars(data.frame(clr(df_freq_unperturbed)))


df1 <- data.frame(mean = means, variance = vars)

max_cv_cts <- df1 %>%
  arrange(desc(variance))
max_cv_cts <- row.names(max_cv_cts)

ggplot(df1, aes(x = mean, y = variance)) +
  geom_point() +
  xlab("Mean (in CLR)") +
  ylab("Variance (in CLR)") +
  theme_classic2()
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_CLR.pdf"), width = 3, height = 2.5)
get_ct_var(df_freq_unperturbed %>% clr())
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_CLR_fit.pdf"), width = 3, height = 2.5)


ggplot(df1, aes(x = mean, y = variance)) +
  geom_point() +
  geom_text(data = df1, aes(label = row.names(df1)), vjust = -0.5)


means <- colMeans(asin(sqrt(((df_counts %>% calc_perc_df()) / 100 ))))
vars <- Rfast::colVars(data.frame(asin(sqrt(((df_counts %>% calc_perc_df()) / 100 )))))


df1 <- data.frame(mean = means, variance = vars)

max_cv_cts <- df1 %>%
  arrange(desc(variance))
max_cv_cts <- row.names(max_cv_cts)

ggplot(df1, aes(x = mean, y = variance)) +
  geom_point() +
  xlab("Mean (asin_sqrt)") +
  ylab("Variance (asin_sqrt)") +
  theme_classic2()
# ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_CLR.pdf"), width = 3, height = 2.5)
get_ct_var(asin(sqrt(((df_counts %>% calc_perc_df()) / 100 ))))
# ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_CLR_fit.pdf"), width = 3, height = 2.5)


plot_data <- data.frame(
  mean = log(means),
  variance = log(vars)
)
# Plot using ggplot2
ggplot(plot_data, aes(x = mean, y = variance)) +
  geom_point() +
  # geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "red") +
  labs(
    x = "log( Mean (asin_sqrt) )",
    y = "log( Variance (asin_sqrt) )"
  ) +
  theme_classic2() # Optional: a clean theme
```



## L3 annotations (cmv_negative, BR1, females)

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

df_counts <- df_counts[1:30, ]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128, 256)
n_ct_to_select <- c(1, 3, 5, 10, 20, round(ncol(df_counts)/2))

results_l3_pca <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select)
```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l3_pca, n_ct_to_select, log_perturbation = TRUE)
print(p)
```



```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results_l3_pca[results_l3_pca$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(trans_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(trans_method, Amount_of_perturbation), .keep_all=TRUE)
  
  p_sil <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_mod <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_clust <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))

  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```




## Example of homogenous samples

```{r}
feat_mat <- df_counts %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df() %>% clr()
res.pca <- prcomp(feat_mat, scale. = FALSE)

factoextra::fviz_pca(res.pca,
                          label = "var",
                          invisible = c("var", "quali"),
                          geom = "point") +
  scale_shape_manual(values = rep(19, length(unique(labels)))) + ggtitle("Homogenous population")

ggsave(filename = file.path(path_plots, paste0("semisynthetic_example_1group.pdf")), width = 4, height = 4)
```

```{r}
nct <- 5
da <- 3
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]


data <- clr(df_freq)
p <- plot_pca(data, labels, title = paste0("2 groups of samples\nDifferential abundance:\n", nct, " cell types\nx ", da),
              pointsize = 2,
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = TRUE) + theme(legend.position = "none")
print(p)
ggsave(filename = file.path(path_plots, paste0("semisynthetic_example_2groups_weakseparation.pdf")), plot = p, width = 4, height = 4)


nct <- 5
da <- 6
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
# ct_da <- sample(colnames(df_counts_temp), size = nct)
# half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]


data <- clr(df_freq)
p <- plot_pca(data, labels, title = paste0("2 groups of samples\nDifferential abundance:\n", nct, " cell types\nx ", da),
              pointsize = 2,
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = TRUE) + theme(legend.position = "none")
print(p)
ggsave(filename = file.path(path_plots, paste0("semisynthetic_example_2groups_mediumseparation.pdf")), plot = p, width = 4, height = 4)



nct <- 20
da <- 6
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]


data <- clr(df_freq)
p <- plot_pca(data, labels, title = paste0("2 groups of samples\nDifferential abundance:\n", nct, " cell types\nx ", da),
              pointsize = 2,
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = TRUE) + theme(legend.position = "none")
print(p)
ggsave(filename = file.path(path_plots, paste0("semisynthetic_example_2groups_strongseparation.pdf")), plot = p, width = 4, height = 4)
```

```{r}

my_comparisons <- list( c("freq", "clr"), c("freq", "arcsine_sqrt"), c("clr", "arcsine_sqrt"))
method <- "wilcox.test"
y_lim_atleast <- 1.5


nct <- 5
da <- 8
df_long <- results_l3_pca[results_l3_pca$n_celltypes == nct &
                     results_l3_pca$Amount_of_perturbation == da &
                     results_l3_pca$trans_method %in% c("freq_pca", "arcsine_sqrt_pca", "clr_pca"), ]
df_long$trans_method <- gsub("_pca", "", df_long$trans_method)

df_long$trans_method <- factor(df_long$trans_method, levels = c("freq", "clr", "arcsine_sqrt"))
ggplot(data = df_long, aes(x = trans_method, y = ANOSIM_score)) +
  geom_boxplot() +
  stat_compare_means(method = method, comparisons = my_comparisons, paired = TRUE, aes(label = after_stat(p.signif))) +
  geom_jitter(width = 0.2, height = 0.05) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12, angle = 15, hjust = 1, vjust = 1),
    axis.title.x = element_blank()  # Remove x-axis title
  ) +
  ylim(c(min(df_long$Adjusted_Rand_Index), y_lim_atleast)) +
  ggtitle(paste0("Differential cell types: ", nct, "\nDA multiplier: ", da))
ggsave(filename = file.path(path_plots, paste0("semisynthetic_example_diffct", nct, "_diffabu", da, ".pdf")), width = 3, height = 3)


nct <- 10
da <- 4
df_long <- results_l3_pca[results_l3_pca$n_celltypes == nct &
                     results_l3_pca$Amount_of_perturbation == da &
                     results_l3_pca$trans_method %in% c("freq_pca", "arcsine_sqrt_pca", "clr_pca"), ]
df_long$trans_method <- gsub("_pca", "", df_long$trans_method)
df_long$trans_method <- factor(df_long$trans_method, levels = c("freq", "clr", "arcsine_sqrt"))

ggplot(data = df_long, aes(x = trans_method, y = ANOSIM_score)) +
  geom_boxplot() +
  stat_compare_means(method = method, comparisons = my_comparisons, paired = TRUE, aes(label = after_stat(p.signif))) +
  geom_jitter(width = 0.2, height = 0.05) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12, angle = 15, hjust = 1, vjust = 1),
    axis.title.x = element_blank()  # Remove x-axis title
  ) +
  ylim(c(min(df_long$Adjusted_Rand_Index), y_lim_atleast)) +
  ggtitle(paste0("Differential cell types: ", nct, "\nDA multiplier: ", da))
ggsave(filename = file.path(path_plots, paste0("semisynthetic_example_diffct", nct, "_diffabu", da, ".pdf")), width = 3, height = 3)



nct <- 20
da <- 4
df_long <- results_l3_pca[results_l3_pca$n_celltypes == nct &
                     results_l3_pca$Amount_of_perturbation == da &
                     results_l3_pca$trans_method %in% c("freq_pca", "arcsine_sqrt_pca", "clr_pca"), ]
df_long$trans_method <- gsub("_pca", "", df_long$trans_method)

df_long$trans_method <- factor(df_long$trans_method, levels = c("freq", "clr", "arcsine_sqrt"))
ggplot(data = df_long, aes(x = trans_method, y = ANOSIM_score)) +
  geom_boxplot() +
  stat_compare_means(method = method, comparisons = my_comparisons, paired = TRUE, aes(label = after_stat(p.signif))) +
  geom_jitter(width = 0.2, height = 0.05) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12, angle = 15, hjust = 1, vjust = 1),
    axis.title.x = element_blank()  # Remove x-axis title
  ) +
  ylim(c(min(df_long$Adjusted_Rand_Index), y_lim_atleast)) +
  ggtitle(paste0("Differential cell types: ", nct, "\nDA multiplier: ", da))
ggsave(filename = file.path(path_plots, paste0("semisynthetic_example_diffct", nct, "_diffabu", da, ".pdf")), width = 3, height = 3)
```



```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(trans_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(trans_method, Amount_of_perturbation), .keep_all=TRUE)
  
  # res_avg$linetype <- ifelse(grepl("pca", res_avg$trans_method), "PCA_2dims", "Default")
  # res_avg$color <- gsub("_pca", "", res_avg$trans_method)
  
  p_sil <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types:\n", nct, " / ", n_cts_tot))
  p_mod <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types:\n", nct, " / ", n_cts_tot))
  p_clust <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types:\n", nct, " / ", n_cts_tot))
  
  ggsave(filename = file.path(path_plots, paste0("transformations_semisynthetic_default", nct, "celltypesda.pdf")), plot = p_clust, width = 4, height = 3)
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "transformations_semisynthetic_default.pdf"), plot = p_clust + ggtitle("10 random cell types"), width = 4, height = 3)
  }
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```


### Show example PCA

```{r}
nct <- 10
da <- 4
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]

plot_list <- list()

p <- plot_pca(df_counts_temp, labels, title = "Counts",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(df_counts_temp, labels)

p <- plot_pca(df_freq, labels, title = "Freq",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
p <- plot_pca(data, labels, title = "asinsqrt",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
p <- plot_pca(data, labels, title = "CLR",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

# data <- prcomp(data)$x[, 1:3]
# p <- plot_pca(data, labels, title = "CLR PCA",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

# data <- alr(df_freq, ct_ref_mincv)
# p <- plot_pca(data, labels, title = "ALR mincvref",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))
# plot_roc_pcad1(data, labels)

p_wrapped <- patchwork::wrap_plots(plot_list, nrow = 2)
ggsave(filename = file.path(path_plots, paste0("transformations_semisynthetic_default_", nct, "celltypesda_", da, "diffabu.pdf")), plot = p_wrapped, width = 4*2, height = 4*2)
```




## L3 annotations (cmv_positive BR1, males)

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Positive") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Male")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

# df_counts <- df_counts[1:30, ]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128, 256)
n_ct_to_select <- c(1, 3, 5, 10, 20, round(ncol(df_counts)/2))

results_l3_pca <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select)
```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l3_pca, n_ct_to_select, log_perturbation = TRUE)
print(p)
```





## L3 annotations unbalanced counts

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

df_counts <- df_counts[1:30, ]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# Generate random numbers between -1 and 1 for each row
random_numbers <- runif(nrow(df_counts), min = -1, max = 1)

# Exponentiate 10 with each random number
multiplier <- 10^random_numbers

# Multiply each row by the corresponding multiplier
df_counts <- sweep(df_counts, 1, multiplier, `*`)

df_counts <- round(df_counts)

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128)
n_ct_to_select <- c(1, 3, 5, 10, 20, round(ncol(df_counts)/2))

results_l3_UnbalancedCounts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select)

```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l3_UnbalancedCounts, n_ct_to_select, log_perturbation = TRUE)
print(p)
```


```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(trans_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(trans_method, Amount_of_perturbation), .keep_all=TRUE)
  
  res_avg$linetype <- ifelse(grepl("pca", res_avg$trans_method), "PCA_2dims", "Default")
  res_avg$color <- gsub("_pca", "", res_avg$trans_method)
  
  p_sil <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  p_mod <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  p_clust <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "transformations_semisynthetic_unbalanced_counts.pdf"),
           plot = p_clust + ggtitle("10 random cell types\nUnbalanced counts:\nEach sample * rand[-10, 10]"),
           width = 4, height = 3)
  }
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 10
da <- 4
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
# ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_list <- list()

p <- plot_pca(df_counts_temp, labels, title = "Counts",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(df_counts_temp, labels)

p <- plot_pca(df_freq, labels, title = "Freq",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
p <- plot_pca(data, labels, title = "asinsqrt",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
p <- plot_pca(data, labels, title = "CLR",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

# data <- prcomp(data)$x[, 1:3]
# p <- plot_pca(data, labels, title = "CLR PCA",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

# data <- alr(df_freq, ct_ref_mincv)
# p <- plot_pca(data, labels, title = "ALR mincvref",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))
# plot_roc_pcad1(data, labels)

p_wrapped <- patchwork::wrap_plots(plot_list, nrow = 2)
ggsave(filename = file.path(path_plots, paste0("transformations_semisynthetic_unbalanced_counts_", nct, "celltypesda_", da, "diffabu.pdf")), plot = p_wrapped, width = 4*2, height = 4*2)
```



## L3 annotations only low abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

df_counts <- df_counts[1:30, ]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64, 128, 256, 512)
n_ct_to_select <- c(1, 3, 5, 10, 20, round(ncol(df_counts)/2))

ct_counts <- colMeans(df_counts)
# Find the threshold for the lowest cell types
threshold <- quantile(ct_counts, 1/2)

# Get the column names with means below the threshold
cts <- colnames(df_counts)[ct_counts < threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results_l3_OnlyLowAbundantCts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select, cts = cts)

```


### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l3_OnlyLowAbundantCts, n_ct_to_select, log_perturbation = TRUE)
print(p)
```


```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(trans_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(trans_method, Amount_of_perturbation), .keep_all=TRUE)
  
  res_avg$linetype <- ifelse(grepl("pca", res_avg$trans_method), "PCA_2dims", "Default")
  res_avg$color <- gsub("_pca", "", res_avg$trans_method)
  
  p_sil <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  p_mod <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  p_clust <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "transformations_semisynthetic_lowly_abundant.pdf"), plot = p_clust + ggtitle("10 low abundant cell types"), width = 4, height = 3)
  }
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 10
da <- 10
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(cts, size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_list <- list()
# 
# p <- plot_pca(df_counts_temp, labels, title = "Counts",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE) + theme(legend.position = "none")
# plot_list <- append(plot_list, list(p))
# plot_roc_pcad1(df_counts_temp, labels)
# 
# p <- plot_pca(df_freq, labels, title = "Freq",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE) + theme(legend.position = "none")
# plot_list <- append(plot_list, list(p))
# plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
p <- plot_pca(data, labels, title = "asinsqrt", pointsize = 2,
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))

df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()
data <- clr(df_freq)
p <- plot_pca(data, labels, title = "CLR", pointsize = 2,
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))

# data <- prcomp(data)$x[, 1:3]
# p <- plot_pca(data, labels, title = "CLR PCA",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))

# data <- alr(df_freq, ct_ref_mincv)
# p <- plot_pca(data, labels, title = "ALR mincvref",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))
# plot_roc_pcad1(data, labels)

p_wrapped <- patchwork::wrap_plots(plot_list, nrow = 1)
# ggsave(filename = file.path(path_plots, paste0("transformations_semisynthetic_lowly_abundant_", nct, "celltypesda_", da, "diffabu.pdf")), plot = p_wrapped, width = 3*2, height = 3)
p_wrapped
```

```{r}
ct_da
data <- clr(df_freq)
library(corrplot)
test <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

test <- cor(t(data))
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)
```


```{r}
data <- clr(df_freq)
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)



data <- df_counts_temp
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)

```



## L3 annotations only highly abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

df_counts <- df_counts[1:30, ]

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 2, 3, 4, 8, 16, 32, 64)
n_ct_to_select <- c(1, 3, 5, 10, 20, round(ncol(df_counts)/2))

ct_counts <- colMeans(df_counts)
# Find the threshold for the highest cell types
threshold <- quantile(ct_counts, 1/2)

# Get the column names with means above the threshold
cts <- colnames(df_counts)[ct_counts >= threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results_l3_OnlyHighAbundantCts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select, cts = cts)
```


```{r}
test_both <- df_counts[, cts]
get_ct_var(test_both)
df_freq <- test_both %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()
get_ct_var(df_freq)
plot((colMeans(df_freq)), (Rfast::colVars(df_freq)))
plot(log(colMeans(df_freq)), log(Rfast::colVars(df_freq)))
get_ct_var(df_freq %>% clr())
```




### Summarized results

```{r fig.height=4, fig.width=12}
p <- plot_results(results_l3_OnlyHighAbundantCts, n_ct_to_select, log_perturbation = TRUE)
print(p)
```




```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(trans_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(trans_method, Amount_of_perturbation), .keep_all=TRUE)
  
  res_avg$linetype <- ifelse(grepl("pca", res_avg$trans_method), "PCA_2dims", "Default")
  res_avg$color <- gsub("_pca", "", res_avg$trans_method)
  
  p_sil <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  p_mod <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  p_clust <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = trans_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct))
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "transformations_semisynthetic_highly_abundant.pdf"), plot = p_clust + ggtitle("10 high abundant cell types"), width = 4, height = 3)
  }
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 10
da <- 3
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(cts, size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * da)
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_list <- list()

p <- plot_pca(df_counts_temp, labels, title = "Counts",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(df_counts_temp, labels)

p <- plot_pca(df_freq, labels, title = "Freq",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
p <- plot_pca(data, labels, title = "asinsqrt",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
p <- plot_pca(data, labels, title = "CLR",
              sil_score = FALSE,
              mod_score = FALSE,
              cluster_score = FALSE) + theme(legend.position = "none")
plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

# data <- prcomp(data)$x[, 1:3]
# p <- plot_pca(data, labels, title = "CLR PCA",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))
plot_roc_pcad1(data, labels)

# data <- alr(df_freq, ct_ref_mincv)
# p <- plot_pca(data, labels, title = "ALR mincvref",
#               sil_score = FALSE,
#               mod_score = FALSE,
#               cluster_score = FALSE)
# plot_list <- append(plot_list, list(p))
# plot_roc_pcad1(data, labels)

p_wrapped <- patchwork::wrap_plots(plot_list, nrow = 2)
ggsave(filename = file.path(path_plots, paste0("transformations_semisynthetic_highly_abundant_", nct, "celltypesda_", da, "diffabu.pdf")), plot = p_wrapped, width = 4*2, height = 4*2)
```

```{r}
cohens_d <- function(x, group) {
  mean_a <- mean(x[group == levels(group)[1]])
  mean_b <- mean(x[group == levels(group)[2]])
  sd_a <- sd(x[group == levels(group)[1]])
  sd_b <- sd(x[group == levels(group)[2]])
  pooled_sd <- sqrt((sd_a^2 + sd_b^2) / 2)
  (mean_b - mean_a) / pooled_sd
}


data <- df_freq[, ct_da[1]]
hist(data)
data <- data.frame(values = data,
                   groups = factor(labels))
p <- ggplot(data, aes(x = values, fill = groups)) +
  geom_density(alpha=0.4) + 
  geom_histogram(aes(y=..density..), alpha=0.3, colour="black", 
                 position="identity") +
  labs(title = "Density Plot of PC1 Scores",
       x = "PC1 Scores",
       y = "Density") +
  theme_classic()
print(p)

d_value <- cohens_d(data$values, data$groups)
print(paste("Cohen's d:", d_value))

wilcox_result <- wilcox.test(values ~ groups, data = data)
print(paste("Wilcoxon p-value:", wilcox_result$p.value))


data <- clr(df_freq)[, ct_da[1]]
hist(data)
data <- data.frame(values = data,
                   groups = factor(labels))
p <- ggplot(data, aes(x = values, fill = groups)) +
  geom_density(alpha=0.4) + 
  geom_histogram(aes(y=..density..), alpha=0.3, colour="black", 
                 position="identity") +
  labs(title = "Density Plot of PC1 Scores",
       x = "PC1 Scores",
       y = "Density") +
  theme_classic()
print(p)

d_value <- cohens_d(data$values, data$groups)
print(paste("Cohen's d:", d_value))

wilcox_result <- wilcox.test(values ~ groups, data = data)
print(paste("Wilcoxon p-value:", wilcox_result$p.value))
```



```{r}
data <- clr(df_freq)
plot_pca(data, labels)
plot_pca(data, labels, scale. = TRUE)

plot_roc_pcad1(data, labels)
```


```{r}
ct_da

data <- clr(df_freq)
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, col = col)
hist(data[, ct_da[1]])
diptest::dip.test(data[, ct_da[1]])

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)

plot_pca(data[,ct_da], labels)



data <- df_counts_temp
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, col = col)
hist(data[, ct_da[1]])
diptest::dip.test(data[, ct_da[1]])


# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)


plot_pca(data[,ct_da], labels)

```



# ---------------------------------------------------------------

# Sensitivity: zero imputation methods

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(2, 4, 6, 8, 16, 32)
n_ct_to_select <- c(5, 10)

results_zeroimp <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select,
                   trans_method = "clr_pca",
                   zero_imp_method = c(
                      "counts_zeros__1",
                      "counts_all__1",
                      "percentage_all__0.001",
                      "percentage_all__0.01",
                      "percentage_all__0.1",
                      "percentage_all__1",
                      "multLN",
                      "multRepl__0.01",
                      "multRepl__0.1",
                      "multRepl__1"
                   )
)

```



### Summarized results


```{r fig.height=4, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results_zeroimp[results_zeroimp$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(zero_imp_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(zero_imp_method, Amount_of_perturbation), .keep_all=TRUE)
  
  zim <- res_avg$zero_imp_method
  zim[grepl("multRepl|percentage", zim)] <- paste0(zim[grepl("multRepl|percentage", zim)], "%")
  res_avg$zero_imp_method <- zim
  
  p_list <- list()
  p_list[["p_sil"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)")) + scale_x_continuous(trans = scale)
  p_list[["p_mod"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)")) + scale_x_continuous(trans = scale)
  p_list[["p_clust"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)")) + scale_x_continuous(trans = scale)
  p_list[["p_anosim"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = ANOSIM_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=ANOSIM_score-ari_sem, ymax=ANOSIM_score+ari_sem)) +
    ylim(c(min(res_avg$ANOSIM_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)")) + scale_x_continuous(trans = scale)
  
  ggsave(filename = file.path(path_plots, paste0("zero_imputation_semisynthetic_default", nct, "celltypesda.pdf")), plot = p_list[["p_clust"]], width = 4, height = 3)
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "zero_imputation_semisynthetic_default.pdf"), plot = p_list[["p_clust"]] + ggtitle("10 random cell types"), width = 5, height = 4)
  }
  
  p <- ggarrange(plotlist = p_list, nrow=1, common.legend = TRUE)
  print(p)
}

```



```{r}
nct <- 10
da <- 3
df_long <- results[results$n_celltypes == nct &
                     results$Amount_of_perturbation == da &
                     results$zero_imp_method %in% c("counts_all__1", "percentage_all__0.1", "percentage_all__1"), ]
df_long$zero_imp_method <- factor(df_long$zero_imp_method, levels = c("counts_all__1", "percentage_all__0.1", "percentage_all__1"))
my_comparisons <- list( c("counts_all__1", "percentage_all__0.1"), c("counts_all__1", "percentage_all__1"), c("percentage_all__0.1", "percentage_all__1"))
ggplot(data = df_long, aes(x = zero_imp_method, y = Adjusted_Rand_Index)) +
  geom_boxplot() +
  stat_compare_means(method = "t.test", comparisons = my_comparisons, paired = TRUE, aes(label = after_stat(p.signif))) +
  geom_jitter(width = 0.2, height = 0.05) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12, angle = 15, hjust = 1, vjust = 1),
    axis.title.x = element_blank()  # Remove x-axis title
  ) +
  ylim(c(-0.1, 1.3)) +
  ggtitle(paste0("Differential cell types: ", nct, "\nDA multiplier: ", da))
ggsave(filename = file.path(path_plots, paste0("zero_imputation_semisynthetic_example_diffct", nct, "_diffabu", da, ".pdf")), width = 4, height = 4)
```



## L3 annotations only low abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 1.75, 2, 3, 4, 6)
n_ct_to_select <- c(5, 10)

ct_counts <- colMeans(df_counts)
# Find the threshold for the lowest fourth
threshold <- quantile(ct_counts, 1/4)

# Get the column names with means below the threshold
cts <- colnames(df_counts)[ct_counts < threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results_zeroimp_OnlyLowAbundantCts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select,
                   trans_method = "clr_pca",
                   zero_imp_method = c(
                      "counts_zeros__1",
                      "counts_all__1",
                      "percentage_all__0.001",
                      "percentage_all__0.01",
                      "percentage_all__0.1",
                      "percentage_all__1",
                      "multLN",
                      "multRepl__0.01",
                      "multRepl__0.1",
                      "multRepl__1"
                   )
)

```

```{r fig.height=4, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results_zeroimp_OnlyLowAbundantCts[results_zeroimp_OnlyLowAbundantCts$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(zero_imp_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(zero_imp_method, Amount_of_perturbation), .keep_all=TRUE)
  
  zim <- res_avg$zero_imp_method
  zim[grepl("multRepl|percentage", zim)] <- paste0(zim[grepl("multRepl|percentage", zim)], "%")
  res_avg$zero_imp_method <- zim
  
  p_list <- list()
  p_list[["p_sil"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_mod"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_clust"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_anosim"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = ANOSIM_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=ANOSIM_score-ari_sem, ymax=ANOSIM_score+ari_sem)) +
    ylim(c(min(res_avg$ANOSIM_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  
  ggsave(filename = file.path(path_plots, paste0("zero_imputation_semisynthetic_lowly_abu", nct, "celltypesda.pdf")), plot = p_list[["p_clust"]], width = 4, height = 4)
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "zero_imputation_semisynthetic_lowly_abu.pdf"), plot = p_list[["p_clust"]] + ggtitle("10 low abundant cell types"), width = 5, height = 4)
  }
  
  p <- ggarrange(plotlist = p_list, nrow=1, common.legend = TRUE)
  print(p)
}

```



## L3 annotations only highly abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")


# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 1.75, 2, 3, 4, 6)
n_ct_to_select <- c(5, 10)

ct_counts <- colMeans(df_counts)
# Find the threshold for the highest fourth
threshold <- quantile(ct_counts, 3/4)

# Get the column names with means above the threshold
cts <- colnames(df_counts)[ct_counts >= threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)


results_zeroimp_OnlyHighAbundantCts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select,
                   trans_method = "clr_pca",
                   zero_imp_method = c(
                      "counts_zeros__1",
                      "counts_all__1",
                      "percentage_all__0.001",
                      "percentage_all__0.01",
                      "percentage_all__0.1",
                      "percentage_all__1",
                      "multLN",
                      "multRepl__0.01",
                      "multRepl__0.1",
                      "multRepl__1"
                   )
)

```



```{r fig.height=4, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results_zeroimp_OnlyHighAbundantCts[results_zeroimp_OnlyHighAbundantCts$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(zero_imp_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(zero_imp_method, Amount_of_perturbation), .keep_all=TRUE)
  
  zim <- res_avg$zero_imp_method
  zim[grepl("multRepl|percentage", zim)] <- paste0(zim[grepl("multRepl|percentage", zim)], "%")
  res_avg$zero_imp_method <- zim
  
  p_list <- list()
  p_list[["p_sil"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_mod"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_clust"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_anosim"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = ANOSIM_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=ANOSIM_score-ari_sem, ymax=ANOSIM_score+ari_sem)) +
    ylim(c(min(res_avg$ANOSIM_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  
  ggsave(filename = file.path(path_plots, paste0("zero_imputation_semisynthetic_highly_abu", nct, "celltypesda.pdf")), plot = p_list[["p_clust"]], width = 4, height = 4)
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "zero_imputation_semisynthetic_highly_abu.pdf"), plot = p_list[["p_clust"]] + ggtitle("10 high abundant cell types"), width = 5, height = 4)
  }
  
  p <- ggarrange(plotlist = p_list, nrow=1, common.legend = TRUE)
  print(p)
}

```



## L3 annotations lower number of cells

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 1.75, 2, 3, 4, 6)
n_ct_to_select <- c(5, 10)

df_counts <- round(df_counts / 10)

results_zeroimp_LowCellCounts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select,
                   trans_method = "clr_pca",
                   zero_imp_method = c(
                      "counts_zeros__1",
                      "counts_all__1",
                      "percentage_all__0.001",
                      "percentage_all__0.01",
                      "percentage_all__0.1",
                      "percentage_all__1",
                      "multLN",
                      "multRepl__0.01",
                      "multRepl__0.1",
                      "multRepl__1"
                   )
)

```

```{r fig.height=4, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results_zeroimp_LowCellCounts[results_zeroimp_LowCellCounts$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(zero_imp_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(zero_imp_method, Amount_of_perturbation), .keep_all=TRUE)
  
  zim <- res_avg$zero_imp_method
  zim[grepl("multRepl|percentage", zim)] <- paste0(zim[grepl("multRepl|percentage", zim)], "%")
  res_avg$zero_imp_method <- zim
  
  p_list <- list()
  p_list[["p_sil"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_mod"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_clust"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_anosim"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = ANOSIM_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=ANOSIM_score-ari_sem, ymax=ANOSIM_score+ari_sem)) +
    ylim(c(min(res_avg$ANOSIM_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  
  ggsave(filename = file.path(path_plots, paste0("zero_imputation_semisynthetic_lowcells", nct, "celltypesda.pdf")), plot = p_list[["p_clust"]], width = 4, height = 4)
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "zero_imputation_semisynthetic_lowcells.pdf"), plot = p_list[["p_clust"]] + ggtitle("10 random cell types\nAll counts / 10"), width = 5, height = 4)
  }
  
  p <- ggarrange(plotlist = p_list, nrow=1, common.legend = TRUE)
  print(p)
}

```



## L3 annotations unbalanced counts

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Amount_of_perturbation <- c(1.5, 1.75, 2, 3, 4, 6)
n_ct_to_select <- c(5, 10)

# Generate random numbers between -1 and 1 for each row
random_numbers <- runif(nrow(df_counts), min = -1, max = 1)

# Exponentiate 10 with each random number
multiplier <- 10^random_numbers

# Multiply each row by the corresponding multiplier
df_counts <- sweep(df_counts, 1, multiplier, `*`)

df_counts <- round(df_counts)

results_zeroimp_UnbalancedCounts <- datrans(count_mat = df_counts, Amount_of_perturbation = Amount_of_perturbation, n_ct_to_select = n_ct_to_select,
                   trans_method = "clr_pca",
                   zero_imp_method = c(
                      "counts_zeros__1",
                      "counts_all__1",
                      "percentage_all__0.001",
                      "percentage_all__0.01",
                      "percentage_all__0.1",
                      "percentage_all__1",
                      "multLN",
                      "multRepl__0.01",
                      "multRepl__0.1",
                      "multRepl__1"
                   )
)

```

```{r fig.height=4, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results_zeroimp_UnbalancedCounts[results_zeroimp_UnbalancedCounts$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(zero_imp_method, Amount_of_perturbation) %>%
    mutate(sil_sem = sd(Silhouette_score) / sqrt(length(Silhouette_score)), Silhouette_score = mean(Silhouette_score),
           mod_sem = sd(Modularity_score) / sqrt(length(Modularity_score)), Modularity_score = mean(Modularity_score),
           ari_sem = sd(Adjusted_Rand_Index) / sqrt(length(Adjusted_Rand_Index)), Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(zero_imp_method, Amount_of_perturbation), .keep_all=TRUE)
  
  zim <- res_avg$zero_imp_method
  zim[grepl("multRepl|percentage", zim)] <- paste0(zim[grepl("multRepl|percentage", zim)], "%")
  res_avg$zero_imp_method <- zim
  
  p_list <- list()
  p_list[["p_sil"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Silhouette_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Silhouette_score-sil_sem, ymax=Silhouette_score+sil_sem)) +
    ylim(c(min(res_avg$Silhouette_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Silhouette_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_mod"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Modularity_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Modularity_score-mod_sem, ymax=Modularity_score+mod_sem)) +
    ylim(c(min(res_avg$Modularity_score), 0.5)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_clust"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = Adjusted_Rand_Index, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=Adjusted_Rand_Index-ari_sem, ymax=Adjusted_Rand_Index+ari_sem)) +
    ylim(c(min(res_avg$Adjusted_Rand_Index), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  p_list[["p_anosim"]] <- ggplot(res_avg, aes(x = Amount_of_perturbation, y = ANOSIM_score, color = zero_imp_method)) +
    geom_line() +
    geom_errorbar(aes(ymin=ANOSIM_score-ari_sem, ymax=ANOSIM_score+ari_sem)) +
    ylim(c(min(res_avg$ANOSIM_score), 1)) +
    theme_classic() +
    # geom_point(data = res_all, aes(x = Amount_of_perturbation, y = Modularity_score)) +
    ggtitle(paste0("Differential cell types: ", nct, "\n(CLR-transformation)"))
  
  ggsave(filename = file.path(path_plots, paste0("zero_imputation_semisynthetic_unbalancedcounts", nct, "celltypesda.pdf")), plot = p_list[["p_clust"]], width = 4, height = 4)
  
  if (nct == 10) {
    ggsave(filename = file.path(path_plots, "zero_imputation_semisynthetic_unbalancedcounts.pdf"),
           plot = p_list[["p_clust"]] + ggtitle("10 random cell types\nUnbalanced counts:\nEach sample * rand[-10, 10]"),
           width = 5, height = 4)
  }
  
  p <- ggarrange(plotlist = p_list, nrow=1, common.legend = TRUE)
  print(p)
}

```



# ---------------------------------------------------------------

# Mean-variance across all datasets

```{r}
ctcomp_file <- "ctcomp_list.rds"

if (file.exists(ctcomp_file)) {
  ctcomp_list <- readRDS(ctcomp_file)
} else {
  ctcomp_list <- list()
}
```

```{r}
for (ds in names(datasets)) {
  if (is.null(ctcomp_list[[ds]])) {
    print(ds)
    seurat <- readRDS(file.path(path_data, paste0(datasets[[ds]][["ds_name"]], seurat_file_ending)))
    
    seurat@misc$label_col <- datasets[[ds]][["label_col"]]
    seurat@misc$low_res_ct_col <- datasets[[ds]][["low_res_ct_col"]]
    seurat@misc$hi_res_ct_col <- datasets[[ds]][["hi_res_ct_col"]]
    
    labels <- get_labels(seurat, seurat@misc$label_col)
    
    seurat <- subset(x = seurat, subset = !!sym(seurat@misc$label_col) == labels[1])
    
    ctcomp <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = seurat@misc$hi_res_ct_col)
    
    ctcomp_list[[ds]][["ctcomp"]] <- ctcomp
    ctcomp_list[[ds]][["group"]] <- labels[1]
  }
}
saveRDS(ctcomp_list, file = ctcomp_file)
```

```{r}
for (ds in names(datasets)) {
  print(ds)
  df_freq_unperturbed <- ctcomp_list[[ds]][["ctcomp"]] %>% impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>% calc_perc_df()

  # plot(sort(Rfast::colVars(df_freq_unperturbed)))
  # plot(sort(log(Rfast::colVars(df_freq_unperturbed))))
  
  plot_data <- data.frame(
    mean = colMeans(df_freq_unperturbed),
    variance = Rfast::colVars(df_freq_unperturbed)
  )
  # Plot using ggplot2
  p <- ggplot(plot_data, aes(x = mean, y = variance)) +
    geom_point() +
    labs(
      x = "Mean (in %)",
      y = "Variance (in %)"
    ) +
    theme_classic2() # Optional: a clean theme
  print(p)
  # ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_percent.pdf"), width = 3, height = 2.5)
  # get_ct_var(df_freq_unperturbed, smooth_method = "gam")
  # ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_percent_fit.pdf"), width = 3, height = 2.5)
  
  
  plot_data <- data.frame(
    mean = log(colMeans(df_freq_unperturbed)),
    variance = log(Rfast::colVars(df_freq_unperturbed))
  )
  # Plot using ggplot2
  p <- ggplot(plot_data, aes(x = mean, y = variance)) +
    geom_point() +
    # geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "red") +
    labs(
      x = "log( Mean (in %) )",
      y = "log( Variance (in %) )"
    ) +
    theme_classic2() # Optional: a clean theme
  print(p)
  # ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_percentLogLog.pdf"), width = 3, height = 2.5)
  
  
  means <- colMeans(clr(df_freq_unperturbed))
  vars <- Rfast::colVars(data.frame(clr(df_freq_unperturbed)))
  
  df1 <- data.frame(mean = means, variance = vars)
  
  max_cv_cts <- df1 %>%
    arrange(desc(variance))
  max_cv_cts <- row.names(max_cv_cts)
  
  p <- ggplot(df1, aes(x = mean, y = variance)) +
    geom_point() +
    xlab("Mean (in CLR)") +
    ylab("Variance (in CLR)") +
    theme_classic2()
  print(p)
  # ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_CLR.pdf"), width = 3, height = 2.5)
  # get_ct_var(df_freq_unperturbed %>% clr())
  # ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_CLR_fit.pdf"), width = 3, height = 2.5)
}
```

```{r}
# Create empty lists to store data from all datasets for each plot type
all_data_plot1 <- list()
all_data_plot2 <- list()
all_data_plot3 <- list()

for (ds in names(datasets)) {
  # Your existing data processing logic
  df_freq_unperturbed <- ctcomp_list[[ds]][["ctcomp"]] %>%
    impute_zeros(clr_zero_impute_method = "counts_all", clr_zero_impute_num = 1) %>%
    calc_perc_df()

  # --- Data for Plot 1 ---
  plot1_data <- data.frame(
    mean = colMeans(df_freq_unperturbed),
    variance = Rfast::colVars(df_freq_unperturbed),
    dataset = ds # Add a 'dataset' column
  )
  all_data_plot1[[ds]] <- plot1_data

  # --- Data for Plot 2 (Log-Log) ---
  plot2_data <- data.frame(
    mean = log(colMeans(df_freq_unperturbed)),
    variance = log(Rfast::colVars(df_freq_unperturbed)),
    dataset = ds
  )
  all_data_plot2[[ds]] <- plot2_data

  # --- Data for Plot 3 (CLR) ---
  means <- colMeans(clr(df_freq_unperturbed))
  vars <- Rfast::colVars(data.frame(clr(df_freq_unperturbed)))
  plot3_data <- data.frame(
    mean = means,
    variance = vars,
    dataset = ds
  )
  all_data_plot3[[ds]] <- plot3_data

  # (Continue with your individual plots within the loop if needed)
}
```

```{r}
# Combine all data frames for each plot type
final_data_plot1 <- bind_rows(all_data_plot1)
final_data_plot2 <- bind_rows(all_data_plot2)
final_data_plot3 <- bind_rows(all_data_plot3)

# Create the three overlay plots outside of the loop

## Overlay Plot 1: Mean-Variance (Percent)
p1_overlay <- ggplot(final_data_plot1, aes(x = mean, y = variance, color = dataset)) +
  geom_point() +
  labs(
    title = "Mean-Variance: All Datasets (Percent)",
    x = "Mean (in %)",
    y = "Variance (in %)"
  ) +
  theme_classic() +
  guides(color = guide_legend(title = "Dataset"))
print(p1_overlay)
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_allDatasets_percent.pdf"), width = 7, height = 3.5)


## Overlay Plot 2: Mean-Variance (Log-Log)
p2_overlay <- ggplot(final_data_plot2, aes(x = mean, y = variance, color = dataset)) +
  geom_point() +
  labs(
    title = "Mean-Variance: All Datasets (Log-Log)",
    x = "log(Mean (in %))",
    y = "log(Variance (in %))"
  ) +
  theme_classic() +
  guides(color = guide_legend(title = "Dataset"))
print(p2_overlay)
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_allDatasets_percentLogLog.pdf"), width = 7, height = 3.5)


## Overlay Plot 3: Mean-Variance (CLR)
p3_overlay <- ggplot(final_data_plot3, aes(x = mean, y = variance, color = dataset)) +
  geom_point() +
  labs(
    title = "Mean-Variance: All Datasets (CLR)",
    x = "Mean (in CLR)",
    y = "Variance (in CLR)"
  ) +
  theme_classic() +
  guides(color = guide_legend(title = "Dataset"))
print(p3_overlay)
ggsave(filename = file.path(path_plots, "semisynthetic_MeanVariance_allDatasets_CLR.pdf"), width = 7, height = 3.5)

```

