---
title: "Compositional data analysis of the Gong & Sharma et al. dataset"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---


# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

# install.packages("/Users/christianhalter/Desktop/scooter_1.0.0.tar.gz", repos = NULL, type = "source")

library(scooter)
library(tidyverse)
library(plotly)
library(ggplot2)
library(ggpubr)
library(factoextra)
library(compositions)
library(pROC)
library(mclust)
source("functions.R")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)

# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
library(ncdf4)
library(arrow)

library(dplyr)

my_seed <- 123
set.seed(my_seed)
```

# Set paths

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_data <- file.path(path_root, "data")
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# Load dataset
Manually downloaded from:
"<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>"

```{r}
files <- list(L1 = file.path(path_data, "sound_life_AIFI_L1_frequencies.csv"),
              L2 = file.path(path_data, "sound_life_AIFI_L2_frequencies.csv"),
              L3 = file.path(path_data, "sound_life_AIFI_L3_frequencies.csv"))

data <- list()

for (f in names(files)) {
  data[[f]] <- read.csv(files[[f]])
}
```



# Process data into wide dataframe

```{r}
dproc <- list()

for (layer in names(data)) {

  dproc$meta <- data[[layer]] %>%
    select(c(1:14)) %>%
    unique() %>% 
    mutate_at(1:14, as.factor)
  
  ctable <- data[[layer]] %>%
    select(c(14,
             !!sym(paste0("AIFI_", layer)),
             !!sym(paste0("AIFI_", layer, "_count")),
             !!sym(paste0("AIFI_", layer, "_frac_total")),
             !!sym(paste0("AIFI_", layer, "_clr")))) %>%
    pivot_wider(names_from = !!sym(paste0("AIFI_", layer)),
                values_from = c(!!sym(paste0("AIFI_", layer, "_count")),
                                !!sym(paste0("AIFI_", layer, "_frac_total")),
                                !!sym(paste0("AIFI_", layer, "_clr")))) %>% 
    column_to_rownames(var = "specimen.specimenGuid")
  
  i <- 1
  n_cts <- length(unique(data[[layer]][[paste0("AIFI_", layer)]]))
  
  dproc[[layer]]$cell_counts <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_count_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    select(-Platelet, -Erythrocyte)

  i <- i + n_cts
  dproc[[layer]]$freq <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_frac_total_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    select(-Platelet, -Erythrocyte)
  i <- i + n_cts
  dproc[[layer]]$clr_authors <- ctable[, i:(i + n_cts - 1)] %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_clr_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    select(-Platelet, -Erythrocyte)
  
  df_counts <- dproc[[layer]]$cell_counts

  df_counts_1count_zero <- df_counts
  df_counts_1count_zero[df_counts_1count_zero == 0] <- 1
  dproc[[layer]]$clr_1count_zero <- df_counts_1count_zero %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_1count_all <- (df_counts + 1) %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_0.1perc <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()
}

saveRDS(dproc, file.path(path_data, "cell_type_compositions.rds"))
```



# ---------------------------------------------------------------

# Sensitivity: freq vs clr vs alr vs ilr

CMV negative BR1 female samples are used, as the biggest sources of variation has been removed (subset to specific groups according to major axis of variance).
Subsequently, one or multiple cell types will be artificially increased/decreased to simulate differential abundance.


## L1 anotations


```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L1"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Percent_difference <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 2, 3)

results <- datrans(count_mat = df_counts, labels = NULL, Percent_difference = Percent_difference, n_ct_to_select = n_ct_to_select)
```

### Summarized results

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, Percent_difference) %>%
    mutate(Silhouette_score = mean(Silhouette_score),
           Modularity_score = mean(Modularity_score),
           Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(method, Percent_difference), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Silhouette_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Silhouette_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Modularity_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Adjusted_Rand_Index, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```



### Show example PCA

```{r}
nct <- 3
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```




## L2 anotations

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L2"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Percent_difference <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10)
reps <- 1:20

results <- datrans(count_mat = df_counts, labels = NULL, Percent_difference = Percent_difference, n_ct_to_select = n_ct_to_select)
```


### Summarized results

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, Percent_difference) %>%
    mutate(Silhouette_score = mean(Silhouette_score),
           Modularity_score = mean(Modularity_score),
           Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(method, Percent_difference), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Silhouette_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Silhouette_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Modularity_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Adjusted_Rand_Index, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```


### Show example PCA

```{r}
nct <- 10
pd <- 100
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

res.pca <- prcomp(data)
p <- factoextra::fviz_pca(res.pca,
                          habillage = labels,
                          label = "var",
                          addEllipses = TRUE,
                          pointsize = 3,
                          invisible = c("var", "quali"),
                          geom = "point") +
  scale_shape_manual(values = rep(19, length(unique(labels)))) +
  coord_equal() + NoLegend() + ggtitle("")
ggsave(file.path(path_plots, "workflow_schema_Adjusted_Rand_Index_example.pdf"), width = 3, height = 3)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```



### Mean-variance relationship

```{r}
df_freq_unperturbed <- df_counts %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()
plot(colMeans(df_freq_unperturbed), main = "colMeans(df_freq_unperturbed)")
plot(Rfast::colVars(df_freq_unperturbed), main = "colVars(df_freq_unperturbed)")
plot(colMeans(df_freq_unperturbed), Rfast::colVars(df_freq_unperturbed))

means <- colMeans(clr(df_freq))
vars <- Rfast::colVars(data.frame(clr(df_freq)))

df1 <- data.frame(mean = means, variance = vars)

ggplot(df1, aes(x = mean, y = variance)) +
  geom_point() +
  geom_text(data = df1, aes(label = row.names(df1)), vjust = -0.5)

```



## L3 annotations

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Percent_difference <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10, 20)

results <- datrans(count_mat = df_counts, labels = NULL, Percent_difference = Percent_difference, n_ct_to_select = n_ct_to_select)

```


```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, Percent_difference) %>%
    mutate(Silhouette_score = mean(Silhouette_score),
           Modularity_score = mean(Modularity_score),
           Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(method, Percent_difference), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Silhouette_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Silhouette_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Modularity_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Adjusted_Rand_Index, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```


### Show example PCA

```{r}
nct <- 10
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```



## L3 annotations unbalanced counts

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# Generate random numbers between -1 and 1 for each row
random_numbers <- runif(nrow(df_counts), min = -1, max = 1)

# Exponentiate 10 with each random number
multiplier <- 10^random_numbers

# Multiply each row by the corresponding multiplier
df_counts <- sweep(df_counts, 1, multiplier, `*`)

df_counts <- round(df_counts)

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Percent_difference <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10, 20)

results <- datrans(count_mat = df_counts, labels = NULL, Percent_difference = Percent_difference, n_ct_to_select = n_ct_to_select)

```

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, Percent_difference) %>%
    mutate(Silhouette_score = mean(Silhouette_score),
           Modularity_score = mean(Modularity_score),
           Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(method, Percent_difference), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Silhouette_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Silhouette_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Modularity_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Adjusted_Rand_Index, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 20
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```



## L3 annotations only low abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Percent_difference <- c(100, 200, 300, 500, 1000)
n_ct_to_select <- c(1, 3, 5, 10, 20)

ct_counts <- colMeans(df_counts)
# Find the threshold for the lowest fourth
threshold <- quantile(ct_counts, 1/4)

# Get the column names with means below the threshold
cts <- colnames(df_counts)[ct_counts < threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results <- datrans(count_mat = df_counts, labels = NULL, Percent_difference = Percent_difference, n_ct_to_select = n_ct_to_select, cts = cts)

```

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, Percent_difference) %>%
    mutate(Silhouette_score = mean(Silhouette_score),
           Modularity_score = mean(Modularity_score),
           Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(method, Percent_difference), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Silhouette_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Silhouette_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Modularity_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Adjusted_Rand_Index, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 5
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```

```{r}
ct_da
data <- clr(df_freq)
library(corrplot)
test <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

test <- cor(t(data))
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)
```


```{r}
data <- clr(df_freq)
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)



data <- df_counts_temp
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)

```




## L3 annotations only highly abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")


# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
Percent_difference <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10, 20)

ct_counts <- colMeans(df_counts)
# Find the threshold for the lowest fourth
threshold <- quantile(ct_counts, 3/4)

# Get the column names with means above the threshold
cts <- colnames(df_counts)[ct_counts >= threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results <- datrans(count_mat = df_counts, labels = NULL, Percent_difference = Percent_difference, n_ct_to_select = n_ct_to_select, cts = cts)
```



```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, Percent_difference) %>%
    mutate(Silhouette_score = mean(Silhouette_score),
           Modularity_score = mean(Modularity_score),
           Adjusted_Rand_Index = mean(Adjusted_Rand_Index)) %>%
    ungroup() %>% 
    distinct(interaction(method, Percent_difference), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Silhouette_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Silhouette_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Modularity_score, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = Percent_difference, y = Adjusted_Rand_Index, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = Percent_difference, y = Modularity_score, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 5
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```

```{r}
cohens_d <- function(x, group) {
  mean_a <- mean(x[group == levels(group)[1]])
  mean_b <- mean(x[group == levels(group)[2]])
  sd_a <- sd(x[group == levels(group)[1]])
  sd_b <- sd(x[group == levels(group)[2]])
  pooled_sd <- sqrt((sd_a^2 + sd_b^2) / 2)
  (mean_b - mean_a) / pooled_sd
}


data <- df_freq[, ct_da[1]]
hist(data)
data <- data.frame(values = data,
                   groups = factor(labels))
p <- ggplot(data, aes(x = values, fill = groups)) +
  geom_density(alpha=0.4) + 
  geom_histogram(aes(y=..density..), alpha=0.3, colour="black", 
                 position="identity") +
  labs(title = "Density Plot of PC1 Scores",
       x = "PC1 Scores",
       y = "Density") +
  theme_classic()
print(p)

d_value <- cohens_d(data$values, data$groups)
print(paste("Cohen's d:", d_value))

wilcox_result <- wilcox.test(values ~ groups, data = data)
print(paste("Wilcoxon p-value:", wilcox_result$p.value))


data <- clr(df_freq)[, ct_da[1]]
hist(data)
data <- data.frame(values = data,
                   groups = factor(labels))
p <- ggplot(data, aes(x = values, fill = groups)) +
  geom_density(alpha=0.4) + 
  geom_histogram(aes(y=..density..), alpha=0.3, colour="black", 
                 position="identity") +
  labs(title = "Density Plot of PC1 Scores",
       x = "PC1 Scores",
       y = "Density") +
  theme_classic()
print(p)

d_value <- cohens_d(data$values, data$groups)
print(paste("Cohen's d:", d_value))

wilcox_result <- wilcox.test(values ~ groups, data = data)
print(paste("Wilcoxon p-value:", wilcox_result$p.value))
```



```{r}
data <- clr(df_freq)
plot_pca(data, labels)
plot_pca(data, labels, scale. = TRUE)

plot_roc_pcad1(data, labels)
```


```{r}
ct_da

data <- clr(df_freq)
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, col = col)
hist(data[, ct_da[1]])
diptest::dip.test(data[, ct_da[1]])

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)

plot_pca(data[,ct_da], labels)



data <- df_counts_temp
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, col = col)
hist(data[, ct_da[1]])
diptest::dip.test(data[, ct_da[1]])


# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)


plot_pca(data[,ct_da], labels)

```
