---
title: "Compositional data analysis of the Gong & Sharma et al. dataset"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---


# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::snapshot()

# install.packages("/Users/christianhalter/Desktop/scooter_1.0.0.tar.gz", repos = NULL, type = "source")

library(scooter)
library(tidyverse)
library(plotly)
library(ggplot2)
library(ggpubr)
library(factoextra)
library(compositions)
library(pROC)
library(mclust)
source("functions.R")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)

# install.packages("devtools")
# devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)
# BiocManager::install("MOFA2")
library(reticulate)
library(basilisk)
library("MOFA2")
# BiocManager::install("fgsea")
# BiocManager::install("sva")
# devtools::install_github("kharchenkolab/scITD")
library(scITD)
# install.packages("ncdf4")
library(ncdf4)
library(arrow)

library(dplyr)

my_seed <- 123
set.seed(my_seed)
```

# Set paths

```{r}
ds_name <- "GongSharma_2024_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_data <- file.path(path_root, "data")
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# Load dataset
Manually downloaded from:
"<https://apps.allenimmunology.org/aifi/insights/dynamics-imm-health-age/downloads/>"

```{r}
files <- list(L1 = file.path(path_data, "sound_life_AIFI_L1_frequencies.csv"),
              L2 = file.path(path_data, "sound_life_AIFI_L2_frequencies.csv"),
              L3 = file.path(path_data, "sound_life_AIFI_L3_frequencies.csv"))

data <- list()

for (f in names(files)) {
  data[[f]] <- read.csv(files[[f]])
}
```



# Process data into wide dataframe

```{r}
dproc <- list()

for (layer in names(data)) {

  dproc$meta <- data[[layer]] %>%
    select(c(1:14)) %>%
    unique() %>% 
    mutate_at(1:14, as.factor)
  
  ctable <- data[[layer]] %>%
    select(c(14,
             !!sym(paste0("AIFI_", layer)),
             !!sym(paste0("AIFI_", layer, "_count")),
             !!sym(paste0("AIFI_", layer, "_frac_total")),
             !!sym(paste0("AIFI_", layer, "_clr")))) %>%
    pivot_wider(names_from = !!sym(paste0("AIFI_", layer)),
                values_from = c(!!sym(paste0("AIFI_", layer, "_count")),
                                !!sym(paste0("AIFI_", layer, "_frac_total")),
                                !!sym(paste0("AIFI_", layer, "_clr")))) %>% 
    column_to_rownames(var = "specimen.specimenGuid")
  
  i <- 1
  n_cts <- length(unique(data[[layer]][[paste0("AIFI_", layer)]]))
  
  dproc[[layer]]$cell_counts <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_count_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    select(-Platelet, -Erythrocyte)

  i <- i + n_cts
  dproc[[layer]]$freq <- ctable[, i:(i + n_cts - 1)] %>%
    mutate_all(~replace_na(., 0)) %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_frac_total_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    select(-Platelet, -Erythrocyte)
  i <- i + n_cts
  dproc[[layer]]$clr_authors <- ctable[, i:(i + n_cts - 1)] %>%
    rename_all(~ str_replace_all(., paste0("AIFI_", layer, "_clr_"), "") %>%
                 str_replace_all(., " ", "_")) %>%
    select(-Platelet, -Erythrocyte)
  
  df_counts <- dproc[[layer]]$cell_counts

  df_counts_1count_zero <- df_counts
  df_counts_1count_zero[df_counts_1count_zero == 0] <- 1
  dproc[[layer]]$clr_1count_zero <- df_counts_1count_zero %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_1count_all <- (df_counts + 1) %>% calc_perc_df() %>% clr()
  
  dproc[[layer]]$clr_0.1perc <- df_counts %>%
    impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>%
    calc_perc_df() %>%
    clr()
}

saveRDS(dproc, file.path(path_data, "cell_type_compositions.rds"))
```


## Create scoot objects based on author annotation

```{r}
for (layer in names(data)) {
  df_counts <- dproc[[layer]]$cell_counts
  df_freq <- dproc[[layer]]$freq * 100
  df_clr <- dproc[[layer]]$clr_0.1perc
  
  cols <- colnames(df_counts)
  
  df_counts$scoot_sample <- rownames(df_counts)
  df_freq$scoot_sample <- rownames(df_freq)
  df_clr$scoot_sample <- rownames(df_clr)

  df_counts_long <- df_counts %>%
    pivot_longer(cols = cols,
                 names_to = "celltype",
                 values_to = "cell_counts")
  df_freq_long <- df_freq %>%
    pivot_longer(cols = cols,
                 names_to = "celltype",
                 values_to = "freq")
  df_clr_long <- df_clr %>%
    pivot_longer(cols = cols,
                 names_to = "celltype",
                 values_to = "clr")
  df_long <- dplyr::left_join(df_counts_long,
                              df_freq_long,
                              by = c("scoot_sample", "celltype"))
  df_long <- dplyr::left_join(df_long,
                              df_clr_long,
                              by = c("scoot_sample", "celltype"))
  
  ctable <- list(layer_1 = df_long)
  
  meta <- dproc$meta
  colnames(meta)[colnames(meta) == "specimen.specimenGuid"] <- "scoot_sample"
  
  assign(paste0("scoot_summary_", layer, "_authors"),
         methods::new("scoot",
                      data = list(composition = ctable),
                      metadata = meta,
                      version = packageVersion("scooter"))
  )
}
```

# ---------------------------------------------------------------

# Introduction schematic
GongSharma highest resolution -> best clustering example so far

```{r}
feat_mat <- dproc$L3$clr_0.1perc
res.pca <- prcomp(feat_mat)

eigenvalues <- res.pca$sdev^2
round(eigenvalues/sum(eigenvalues),3)
plot(cumsum(eigenvalues/sum(eigenvalues)))

# Create a scree plot
plot(eigenvalues, type = "b",
     xlab = "Principal Component",
     ylab = "Eigenvalue")
plot(eigenvalues/sum(eigenvalues)*100, type = "b",
     xlab = "Principal Component",
     ylab = "Percentage of Variance Explained")


feat_mat <- res.pca$x[, 1:3]

dist_mat <- dist(feat_mat)
# Perform hierarchical clustering
hc <- stats::hclust(dist_mat, method = "ward.D2")
labels <- stats::cutree(hc, k = 4)

# labels <- kmeans(feat_mat, 2)$cluster
# labels <- cluster::pam(feat_mat, k = 2)$cluster

# adj_graph <- igraph::as.undirected(cccd::nng(feat_mat, k = 2))
# r <- quantile(igraph::strength(adj_graph))[2] / (igraph::gorder(adj_graph) - 1) / 4
# r <- 0.0000000001
# labels <- igraph::cluster_leiden(adj_graph, resolution = r)$membership
# 
# dbscan_clustres <- dbscan::hdbscan(feat_mat, minPts = 20)
# plot(dbscan_clustres, show_flat = TRUE)
# labels <- dbscan_clustres$cluster

factoextra::fviz_pca(res.pca,
                     habillage = labels,
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     geom = "point") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Cluster"), color=guide_legend(title="Cluster")) +
  ggtitle("Unsupervised clustering")
ggsave(file.path(path_plots, "workflow_schema_clustering_unsupervised.pdf"),
       width = 4,
       height = 3)

factoextra::fviz_pca(res.pca,
                     habillage = dproc$meta$subject.cmv,
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     geom = "point") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="CMV"), color=guide_legend(title="CMV")) +
  ggtitle("Biological variable 1\nCMV infection status")
ggsave(file.path(path_plots, "workflow_schema_clustering_CMV.pdf"),
       width = 4,
       height = 3)

factoextra::fviz_pca(res.pca,
                     habillage = factor(ifelse(dproc$meta$cohort.cohortGuid == "BR1", "Young", "Advanced"), levels = c("Young", "Advanced")),
                     # addEllipses = TRUE,
                     label = "var",
                     pointsize = 1,
                     invisible = c("var", "quali"),
                     geom = "point") +
    scale_shape_manual(values = rep(19, length(unique(labels)))) +
  guides(shape=guide_legend(title="Age"), color=guide_legend(title="Age")) +
  ggtitle("Biological variable 2\nAge")
ggsave(file.path(path_plots, "workflow_schema_clustering_age.pdf"),
       width = 4,
       height = 3)
```

```{r fig.height=6, fig.width=16}
composition_boxplot(scoot_object = scoot_summary_L3_authors)
composition_boxplot(scoot_object = scoot_summary_L3_authors,
                    group_by = "subject.cmv")
composition_boxplot(scoot_object = scoot_summary_L3_authors,
                    group_by = "cohort.cohortGuid")
composition_boxplot(scoot_object = scoot_summary_L3_authors,
                    group_by = "cohort.cohortGuid",
                    facet_by = "subject.cmv")
```



# ---------------------------------------------------------------

# Plot number of samples and cell types lost per granularity

```{r}
# Apply function to each dataframe
df1 <- dproc[["L1"]][["cell_counts"]]
df2 <- dproc[["L2"]][["cell_counts"]]
df3 <- dproc[["L3"]][["cell_counts"]]

# Assume df1, df2, df3 are your three dataframes
dfs <- list(Low = df1, Medium = df2, High = df3)


# Function to calculate percentage of rows/columns with at least one zero
compute_zero_percentage <- function(df, granularity) {
  total_samples <- nrow(df)
  total_celltypes <- ncol(df)
  
  samples_with_zeros <- sum(rowSums(df == 0) > 0) / total_samples * 100
  celltypes_with_zeros <- sum(colSums(df == 0) > 0) / total_celltypes * 100
  
  granularity_labels <- paste0(granularity, "\n(", total_celltypes, " cell types)")
  
  data.frame(
    Granularity = factor(granularity_labels, levels = granularity_labels),
    Category = c("Samples", "Cell Types"),
    Percentage = c(samples_with_zeros, celltypes_with_zeros)
  )
}

# Compute percentages for all dataframes
zero_percentages <- bind_rows(
  lapply(names(dfs), function(n) compute_zero_percentage(dfs[[n]], n))
)

# Plot
ggplot(zero_percentages, aes(x = Granularity, y = Percentage, group = Category, color = Category, label = round(Percentage, 1))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(vjust=-1, show.legend  = F) +
  theme_minimal() +
  labs(title = "Percentage of samples or cell types\nwith at least one zero",
       x = "Granularity",
       y = "Percentage (%)",
       color = "Category") +
  theme(text = element_text(size = 14)) +
  ylim(0,100)
ggsave(file.path(path_plots, "coda_numberzeros.pdf"), width = 6, height = 4)
```

# ---------------------------------------------------------------


## Test on real datasets

```{r}
results_list <- list()
```


### Zhang et al. dataset

```{r}
ds_name <- "ZhangY_2022_34653365"
path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")
path_output_samples <- file.path(path_output, "samples")

obj.list <- read_objs(path_output_samples)

# Exclude samples that consist mainly of one cell type as outliers
samples_exclude <- c("Pre_P012_b", "Pre_P013_t")
obj.list <- obj.list[!names(obj.list) %in% samples_exclude]

samples_clearcut <- c()

for (i in names(obj.list)) {
  if (obj.list[[i]]@meta.data$Treatment[1] %in% c("Pre") &
      !(obj.list[[i]]@meta.data$Clinical.efficacy.[1] %in% c("-"))) {
    samples_clearcut <- c(samples_clearcut, i)
  }
}

metacols_to_keep <- c("Sample", "Tissue", "layer1", "layer2")

for (i in samples_clearcut) {
  print(i)
  
  obj <- obj.list[[i]]
  Idents(obj) <- "Sample"
  obj@meta.data <- obj@meta.data[colnames(obj@assays$RNA$counts), metacols_to_keep]

  if (i == samples_clearcut[1]) {
    seurat <- obj
  } else {
    seurat <- merge(seurat, obj)
  }
}

rm(obj.list); gc()

label_condition <- "Tissue"
low_res_ct_col <- "layer1"
hi_res_ct_col <- "layer2"


count_mat <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = hi_res_ct_col)
labels <- as.factor(get_labels(seurat, label_condition))

ds <- "zhang"
results_list[[ds]] <- datrans(count_mat = count_mat, labels = labels, perc_diffs = 0, n_ct_to_select = 0, reps = 20, n_cores = 1,
                                   methods = c(
                                     "counts",
                                     "freq",
                                     "asin_sqrt",
                                     "alr_randref",
                                     "alr_mincvref",
                                     "clr"
                                   ))
results_list[[ds]] <- results_list[[ds]] %>%
  group_by(method) %>%
  summarize(silhouette = mean(silhouette),
            modularity = mean(modularity),
            clustering = mean(clustering)) %>%
  ungroup()
results_list[[ds]]$dataset <- ds
```



### Stephenson et al. dataset


```{r}
# Set paths
ds_name <- "MitchelJ_2023_PrePrintTBD"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")

seurat <- readRDS(file.path(path_output, sprintf("%s_Stephenson_preprocessed.rds", ds_name)))

seurat <- subset(seurat, subset = Site == "Ncl")
seurat <- subset(seurat, subset = Status %in% c("Healthy", "Covid"))
seurat$Days_from_onset <- as.character(seurat$Days_from_onset)
seurat$Days_from_onset[seurat$Days_from_onset == "Healthy"] <- 0
seurat$Days_from_onset <- as.numeric(seurat$Days_from_onset)
seurat <- subset(seurat, subset = Days_from_onset < 10)

seurat$sample_id <- as.character(seurat$sample_id)
samples_exclude <- c("BGCV10_CV0198", "MH8919230")
samples_subset <- unique(seurat$sample_id[!seurat$sample_id %in% samples_exclude])
seurat <- subset(seurat, subset = sample_id %in% samples_subset)


label_condition <- "Status"
low_res_ct_col <- "initial_clustering"
hi_res_ct_col <- "full_clustering"


count_mat <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = hi_res_ct_col)
labels <- as.factor(get_labels(seurat, label_condition))

ds <- "stephenson"
results_list[[ds]] <- datrans(count_mat = count_mat, labels = labels, perc_diffs = 0, n_ct_to_select = 0, reps = 20, n_cores = 1,
                                   methods = c(
                                     "counts",
                                     "freq",
                                     "asin_sqrt",
                                     "alr_randref",
                                     "alr_mincvref",
                                     "clr"
                                   ))
results_list[[ds]] <- results_list[[ds]] %>%
  group_by(method) %>%
  summarize(silhouette = mean(silhouette),
            modularity = mean(modularity),
            clustering = mean(clustering)) %>%
  ungroup()
results_list[[ds]]$dataset <- ds
```


### Gong and Sharma et al. dataset (all CMV)


```{r}
# Set paths
ds_name <- "GongSharma_2024_PrePrintTBD"
path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_output <- file.path(path_root, "output")

seurat <- load_seurat_object("data/gongsharma_all.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid


label_condition <- "subject.cmv"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

count_mat <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = hi_res_ct_col)
labels <- as.factor(get_labels(seurat, label_condition))

ds <- "gongsharma_allcmv"
results_list[[ds]] <- datrans(count_mat = count_mat, labels = labels, perc_diffs = 0, n_ct_to_select = 0, reps = 20, n_cores = 1,
                                   methods = c(
                                     "counts",
                                     "freq",
                                     "asin_sqrt",
                                     "alr_randref",
                                     "alr_mincvref",
                                     "clr"
                                   ))
results_list[[ds]] <- results_list[[ds]] %>%
  group_by(method) %>%
  summarize(silhouette = mean(silhouette),
            modularity = mean(modularity),
            clustering = mean(clustering)) %>%
  ungroup()
results_list[[ds]]$dataset <- ds
```


#### Age (females cmvneg)

```{r}
seurat <- load_seurat_object("data/gongsharma_age_females_cmvneg.h5ad")
seurat$Sample <- seurat$specimen.specimenGuid
seurat$age <- as.factor(ifelse(seurat$sample.subjectAgeAtDraw < 50, "young", "old"))

label_condition <- "age"
low_res_ct_col <- "AIFI_L1"
hi_res_ct_col <- "AIFI_L3"

count_mat <- get_ct_comp_df_seurat(seurat, sample_col = "Sample", ct_col = hi_res_ct_col)
labels <- as.factor(get_labels(seurat, label_condition))

ds <- "gongsharma_agefemalescmvneg"
results_list[[ds]] <- datrans(count_mat = count_mat, labels = labels, perc_diffs = 0, n_ct_to_select = 0, reps = 20, n_cores = 1,
                                   methods = c(
                                     "counts",
                                     "freq",
                                     "asin_sqrt",
                                     "alr_randref",
                                     "alr_mincvref",
                                     "clr"
                                   ))
results_list[[ds]] <- results_list[[ds]] %>%
  group_by(method) %>%
  summarize(silhouette = mean(silhouette),
            modularity = mean(modularity),
            clustering = mean(clustering)) %>%
  ungroup()
results_list[[ds]]$dataset <- ds
```

## Plot results

```{r}
df_results <- Reduce(rbind, results_list)

df_results <- df_results %>%
  pivot_longer(
    cols = c(silhouette, modularity, clustering),
    names_to = "score_method",
    values_to = "values"
  ) %>%
  as.data.frame()

df_results_clustering <- df_results %>%
  filter(score_method == "clustering") %>%
  group_by(method) %>%
  summarize(mean = mean(values), sem = sd(values) / sqrt(length(values)))

df_results_clustering <- df_results_clustering %>%
  arrange(desc(mean))

df_results_clustering$method <- factor(df_results_clustering$method, levels = df_results_clustering$method)
df_results$method <- factor(df_results$method, levels = df_results_clustering$method)



# New facet label names for dose variable
score.labs <- c("Adjusted rand index", "Modularity score", "Silhouette score")
names(score.labs) <- sort(unique(df_results$score_method))

# New facet label names for supp variable
datas.labs <- c(
  "Gong & Sharma\nSubset: all\nSeparation by: CMV infection",
  "Gong & Sharma\nSubset: CMV negative females\nSeparation by: age",
  "Stephenson",
  "Zhang")
names(datas.labs) <- sort(unique(df_results$dataset))



ggplot(df_results, aes(x = method, y = values)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none") +
  facet_grid(score_method ~ dataset,
             scales = "free",
             labeller = labeller(score = score.labs, dataset = datas.labs))
  # facet_grid(scales = "free",
  #            rows = vars(score_method), cols = vars(dataset))
ggsave(file.path(path_plots, "transformations_real_datasets_detailed_barplot.pdf"),
       width = 8,
       height = 7)



df_results %>%
  group_by(score_method, method) %>%
  summarize(mean = mean(values), sem = sd(values) / sqrt(length(values))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score_method),
             labeller = labeller(score_method = score.labs))

ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores.pdf"),
       width = 5,
       height = 7)



df_results_clustering %>%
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()
ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot.pdf"),
       width = 2.5,
       height = 2.5)

```


## Plot all

```{r}
# df_plot <- data.frame(values = unlist(res_list) / length(unique(dataset)) / length(unique(scores)),
#                       dataset = dataset,
#                       imputation_method = middle_parts,
#                       score = sapply(str_split(names(unlist(res_list)), "\\."), function(x) x[length(x)]))
# summarized_df <- df_plot %>%
#   group_by(imputation_method) %>%
#   summarize(values = sum(values))
# summarized_df$score <- factor(sapply(str_split(summarized_df$imputation_method, "\\."), function(x) x[length(x)]))
# summarized_df$imputation_method <- gsub("\\.sil_score|\\.mod_score", "", summarized_df$imputation_method)

summarized_df <- Reduce(rbind, results_list)
summarized_df <- summarized_df %>%
  pivot_longer(
    cols = c(silhouette, modularity, clustering),
    names_to = "score_method",
    values_to = "values"
  )

ggplot(summarized_df, aes(x = method, y = values, fill = score_method)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score_method), cols = vars(dataset))
ggsave(file.path(path_plots, "transformations_real_datasets_detailed_barplot_separatescores.pdf"),
       width = 6,
       height = 4)



summarized_df <- summarized_df %>%
  group_by(score_method, method) %>%
  summarize(values = mean(values)) %>%
  ungroup()

ggplot(summarized_df, aes(x = method, y = values, fill = score_method)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggplot(summarized_df, aes(x = method, y = values, fill = score_method)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Imputation method",
       y = "Separation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score_method))
ggsave(file.path(path_plots, "transformations_real_datasets_summary_barplot_separatescores.pdf"),
       width = 4,
       height = 4)
```



# ---------------------------------------------------------------

# Sensitivity: freq vs clr vs alr vs ilr

CMV negative BR1 female samples are used, as the biggest sources of variation has been removed (subset to specific groups according to major axis of variance).
Subsequently, one or multiple cell types will be artificially increased/decreased to simulate differential abundance.


## L1 anotations


```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L1"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
perc_diffs <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 2, 3)

results <- datrans(count_mat = df_counts, labels = NULL, perc_diffs = perc_diffs, n_ct_to_select = n_ct_to_select)
```

### Summarized results

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, perc_diff) %>%
    mutate(silhouette = mean(silhouette),
           modularity = mean(modularity),
           clustering = mean(clustering)) %>%
    ungroup() %>% 
    distinct(interaction(method, perc_diff), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = silhouette, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = silhouette, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = modularity, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = clustering, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```



### Show example PCA

```{r}
nct <- 3
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```




## L2 anotations

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L2"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
perc_diffs <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10)
reps <- 1:20

results <- datrans(count_mat = df_counts, labels = NULL, perc_diffs = perc_diffs, n_ct_to_select = n_ct_to_select)
```


### Summarized results

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, perc_diff) %>%
    mutate(silhouette = mean(silhouette),
           modularity = mean(modularity),
           clustering = mean(clustering)) %>%
    ungroup() %>% 
    distinct(interaction(method, perc_diff), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = silhouette, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = silhouette, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = modularity, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = clustering, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```


### Show example PCA

```{r}
nct <- 10
pd <- 100
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

res.pca <- prcomp(data)
p <- factoextra::fviz_pca(res.pca,
                          habillage = labels,
                          label = "var",
                          addEllipses = TRUE,
                          pointsize = 3,
                          invisible = c("var", "quali"),
                          geom = "point") +
  scale_shape_manual(values = rep(19, length(unique(labels)))) +
  coord_equal() + NoLegend() + ggtitle("")
ggsave(file.path(path_plots, "workflow_schema_clustering_example.pdf"), width = 3, height = 3)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```



### Mean-variance relationship

```{r}
df_freq_unperturbed <- df_counts %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()
plot(colMeans(df_freq_unperturbed), main = "colMeans(df_freq_unperturbed)")
plot(Rfast::colVars(df_freq_unperturbed), main = "colVars(df_freq_unperturbed)")
plot(colMeans(df_freq_unperturbed), Rfast::colVars(df_freq_unperturbed))

means <- colMeans(clr(df_freq))
vars <- Rfast::colVars(data.frame(clr(df_freq)))

df1 <- data.frame(mean = means, variance = vars)

ggplot(df1, aes(x = mean, y = variance)) +
  geom_point() +
  geom_text(data = df1, aes(label = row.names(df1)), vjust = -0.5)

```



## L3 annotations

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
perc_diffs <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10, 20)

results <- datrans(count_mat = df_counts, labels = NULL, perc_diffs = perc_diffs, n_ct_to_select = n_ct_to_select)

```


```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, perc_diff) %>%
    mutate(silhouette = mean(silhouette),
           modularity = mean(modularity),
           clustering = mean(clustering)) %>%
    ungroup() %>% 
    distinct(interaction(method, perc_diff), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = silhouette, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = silhouette, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = modularity, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = clustering, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```


### Show example PCA

```{r}
nct <- 10
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)


# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```



## L3 annotations unbalanced counts

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# Generate random numbers between -1 and 1 for each row
random_numbers <- runif(nrow(df_counts), min = -1, max = 1)

# Exponentiate 10 with each random number
multiplier <- 10^random_numbers

# Multiply each row by the corresponding multiplier
df_counts <- sweep(df_counts, 1, multiplier, `*`)

df_counts <- round(df_counts)

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
perc_diffs <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10, 20)

results <- datrans(count_mat = df_counts, labels = NULL, perc_diffs = perc_diffs, n_ct_to_select = n_ct_to_select)

```

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, perc_diff) %>%
    mutate(silhouette = mean(silhouette),
           modularity = mean(modularity),
           clustering = mean(clustering)) %>%
    ungroup() %>% 
    distinct(interaction(method, perc_diff), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = silhouette, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = silhouette, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = modularity, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = clustering, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 20
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```



## L3 annotations only low abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")

# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
perc_diffs <- c(100, 200, 300, 500, 1000)
n_ct_to_select <- c(1, 3, 5, 10, 20)

ct_counts <- colMeans(df_counts)
# Find the threshold for the lowest fourth
threshold <- quantile(ct_counts, 1/4)

# Get the column names with means below the threshold
cts <- colnames(df_counts)[ct_counts < threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results <- datrans(count_mat = df_counts, labels = NULL, perc_diffs = perc_diffs, n_ct_to_select = n_ct_to_select, cts = cts)

```

```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, perc_diff) %>%
    mutate(silhouette = mean(silhouette),
           modularity = mean(modularity),
           clustering = mean(clustering)) %>%
    ungroup() %>% 
    distinct(interaction(method, perc_diff), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = silhouette, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = silhouette, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = modularity, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = clustering, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 5
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```

```{r}
ct_da
data <- clr(df_freq)
library(corrplot)
test <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

test <- cor(t(data))
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)
```


```{r}
data <- clr(df_freq)
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)



data <- df_counts_temp
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = test, col = col, symm = TRUE)

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)

```




## L3 annotations only highly abundant cell types DA

```{r}
# Subset to CMV negative BR1 female samples
cmv_neg_BR1_female_samples <- (dproc$meta$subject.cmv == "Negative") & (dproc$meta$cohort.cohortGuid == "BR1") & (dproc$meta$subject.biologicalSex == "Female")

df_counts <- dproc[["L3"]]$cell_counts[cmv_neg_BR1_female_samples,]

colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\-", "minus")
colnames(df_counts) <- stringr::str_replace_all(colnames(df_counts), "\\+", "plus")


# To one or multiple cell types, add +1%, +5%, +10%, +20%, +50%
perc_diffs <- c(10, 25, 50, 75, 100, 200, 300, 500)
n_ct_to_select <- c(1, 3, 5, 10, 20)

ct_counts <- colMeans(df_counts)
# Find the threshold for the lowest fourth
threshold <- quantile(ct_counts, 3/4)

# Get the column names with means above the threshold
cts <- colnames(df_counts)[ct_counts >= threshold]

n_ct_to_select[n_ct_to_select >= length(cts)] <- length(cts)

results <- datrans(count_mat = df_counts, labels = NULL, perc_diffs = perc_diffs, n_ct_to_select = n_ct_to_select, cts = cts)
```



```{r fig.height=12, fig.width=12}
for (nct in n_ct_to_select) {
  res_all <- results[results$n_celltypes == nct, ]
  res_avg <- res_all %>%
    group_by(method, perc_diff) %>%
    mutate(silhouette = mean(silhouette),
           modularity = mean(modularity),
           clustering = mean(clustering)) %>%
    ungroup() %>% 
    distinct(interaction(method, perc_diff), .keep_all=TRUE)
  
  p_sil <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = silhouette, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = silhouette, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nSilhouette score"))
  p_mod <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = modularity, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nModularity score"))
  p_clust <- ggplot() +
    geom_line(data = res_avg, aes(x = perc_diff, y = clustering, color = method), size = 1.2) +
    # geom_point(data = res_all, aes(x = perc_diff, y = modularity, color = method)) +
    ggtitle(paste0("Differentially abundant cell types: ", nct, "\nClustering score"))
  
  p <- ggarrange(p_sil, p_mod, p_clust)
  print(p)
}

```

### Show example PCA

```{r}
nct <- 5
pd <- 200
n_half_samples <- round(dim(df_counts)[1] / 2)

# Prepare data
df_counts_temp <- df_counts
ct_da <- sample(colnames(df_counts_temp), size = nct)
half_samples_da <- sample(row.names(df_counts_temp), size = n_half_samples)
labels <- as.numeric(row.names(df_counts) %in% half_samples_da)

# Simulate differential abundance
rsums_before <- rowSums(df_counts_temp)
df_counts_temp[half_samples_da, ct_da] <- round(df_counts_temp[half_samples_da, ct_da] * (1 + pd/100))
rsums_after <- rowSums(df_counts_temp)
df_counts_temp <- round(df_counts_temp / (rsums_after / rsums_before))
df_freq <- df_counts_temp %>% impute_zeros(clr_zero_impute_method = "percentage_all", clr_zero_impute_num = 0.1) %>% calc_perc_df()

cvs <- apply(df_freq, 2, cv)
min_cv <- min(cvs[!colnames(df_freq) %in% ct_da])
ct_ref_mincv <- colnames(df_freq)[which(cvs == min_cv)]



plot_pca(df_counts_temp, labels, title = "Counts")
plot_roc_pcad1(df_counts_temp, labels)

plot_pca(df_freq, labels, title = "Freq")
plot_roc_pcad1(df_freq, labels)

data <- asin(sqrt(df_freq/100))
plot_pca(data, labels, title = "asinsqrt")
plot_roc_pcad1(data, labels)

data <- clr(df_freq)
plot_pca(data, labels, title = "CLR")
plot_roc_pcad1(data, labels)

data <- prcomp(data)$x[, 1:3]
plot_pca(data, labels, title = "CLR PCA")
plot_roc_pcad1(data, labels)

data <- alr(df_freq, ct_ref_mincv, title = "ALR mincvref")
plot_pca(data, labels)
plot_roc_pcad1(data, labels)
```

```{r}
cohens_d <- function(x, group) {
  mean_a <- mean(x[group == levels(group)[1]])
  mean_b <- mean(x[group == levels(group)[2]])
  sd_a <- sd(x[group == levels(group)[1]])
  sd_b <- sd(x[group == levels(group)[2]])
  pooled_sd <- sqrt((sd_a^2 + sd_b^2) / 2)
  (mean_b - mean_a) / pooled_sd
}


data <- df_freq[, ct_da[1]]
hist(data)
data <- data.frame(values = data,
                   groups = factor(labels))
p <- ggplot(data, aes(x = values, fill = groups)) +
  geom_density(alpha=0.4) + 
  geom_histogram(aes(y=..density..), alpha=0.3, colour="black", 
                 position="identity") +
  labs(title = "Density Plot of PC1 Scores",
       x = "PC1 Scores",
       y = "Density") +
  theme_classic()
print(p)

d_value <- cohens_d(data$values, data$groups)
print(paste("Cohen's d:", d_value))

wilcox_result <- wilcox.test(values ~ groups, data = data)
print(paste("Wilcoxon p-value:", wilcox_result$p.value))


data <- clr(df_freq)[, ct_da[1]]
hist(data)
data <- data.frame(values = data,
                   groups = factor(labels))
p <- ggplot(data, aes(x = values, fill = groups)) +
  geom_density(alpha=0.4) + 
  geom_histogram(aes(y=..density..), alpha=0.3, colour="black", 
                 position="identity") +
  labs(title = "Density Plot of PC1 Scores",
       x = "PC1 Scores",
       y = "Density") +
  theme_classic()
print(p)

d_value <- cohens_d(data$values, data$groups)
print(paste("Cohen's d:", d_value))

wilcox_result <- wilcox.test(values ~ groups, data = data)
print(paste("Wilcoxon p-value:", wilcox_result$p.value))
```



```{r}
data <- clr(df_freq)
plot_pca(data, labels)
plot_pca(data, labels, scale. = TRUE)

plot_roc_pcad1(data, labels)
```


```{r}
ct_da

data <- clr(df_freq)
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, col = col)
hist(data[, ct_da[1]])
diptest::dip.test(data[, ct_da[1]])

# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)

plot_pca(data[,ct_da], labels)



data <- df_counts_temp
library(corrplot)
cor_matrix <- cor(data)
col<- colorRampPalette(c("blue", "white", "red"))(20)
heatmap(x = cor_matrix, col = col)
hist(data[, ct_da[1]])
diptest::dip.test(data[, ct_da[1]])


# 2. Variance Filtering
variances <- apply(data, 2, var)
plot(variances)
hist(variances, breaks = 20)
threshold <- quantile(variances, 0.8) # Example threshold
filtered_df <- data[, variances > threshold]

plot_pca(data, labels)
plot_pca(filtered_df, labels)


plot_pca(data[,ct_da], labels)

```
