---
title: "clr-transform - How to handle zeros"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

# Load libraries

```{r}
library(ggplot2)

# Set paths
path_plots <- path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
```


# Toy data examples

## +1 count, +1%, KNN, model

```{r fig.height=8, fig.width=25}
library(gridExtra)

calculate_geometric_mean <- function(row) {
  exp(mean(log(row)))
}

a <- t(data.frame(low = c( 0, 0,  0,   10),
                  medium = c( 0, 0,  2,  100),
                  high = c( 0, 1, 20, 1000) #, one_low = c( 0, 2, 20, 1000)
                  ))
colnames(a) <- c("ct1", "ct2", "ct3", "ct4")
a_plot <- as.data.frame(a)
a_plot$geom <- round(apply(a_plot, 1, calculate_geometric_mean),2)

b <- a
b[b==0] <- 1
b_plot <- as.data.frame(b)
b_plot$geom <- round(apply(b_plot, 1, calculate_geometric_mean),2)
for(row in 1:dim(a)[1]){
  b[row,] <- Hotelling::clr(t(b[row,] / sum(b[row,]) * 100))
}
b_clr_plot <- round(b, 2)
b <- cbind(rownames(a), rep("+1", dim(a)[1]), b)
df <- b

b1 <- a
b1 <- b1+1
b1_plot <- as.data.frame(b1)
b1_plot$geom <- round(apply(b1_plot, 1, calculate_geometric_mean),2)
for(row in 1:dim(a)[1]){
  b1[row,] <- Hotelling::clr(t(b1[row,] / sum(b1[row,]) * 100))
}
b1_clr_plot <- round(b1, 2)
b1 <- cbind(rownames(a), rep("+1 to all", dim(a)[1]), b1)
df <- rbind(df, b1)

c <- a
for(row in 1:dim(a)[1]){
  c[row,][c[row,] == 0] <- sum(c[row,]/100)
}
c_plot <- as.data.frame(c)
c_plot$geom <- round(apply(c_plot, 1, calculate_geometric_mean),2)
for(row in 1:dim(a)[1]){
  c[row,] <- Hotelling::clr(t(c[row,] / sum(c[row,]) * 100))
}
c_clr_plot <- round(c, 2)
c <- cbind(rownames(a), rep("+1%", dim(a)[1]), c)
df <- rbind(df, c)

d <- a
for(row in 1:dim(a)[1]){
  d[row,] <- d[row,] + sum(d[row,]/100)
}
d_plot <- as.data.frame(d)
d_plot$geom <- round(apply(d_plot, 1, calculate_geometric_mean),2)
for(row in 1:dim(a)[1]){
  d[row,] <- Hotelling::clr(t(d[row,] / sum(d[row,]) * 100))
}
d_clr_plot <- round(d, 2)
d <- cbind(rownames(a), rep("+1% to all", dim(a)[1]), d)
df <- rbind(df, d)

d1 <- a
for(row in 1:dim(a)[1]){
  d1[row,] <- d1[row,] + sum(d1[row,]/1000)
}
d1_plot <- round(as.data.frame(d1), 2)
d1_plot$geom <- round(apply(d1_plot, 1, calculate_geometric_mean),2)
for(row in 1:dim(a)[1]){
  d1[row,] <- Hotelling::clr(t(d1[row,] / sum(d1[row,]) * 100))
}
d1_clr_plot <- round(d1, 2)
d1 <- cbind(rownames(a), rep("+0.1% to all", dim(a)[1]), d1)
df <- rbind(df, d1)


df <- data.frame(df)
df$V1 <- factor(df$V1, levels = c("low", "medium", "high", "one_low"))
df$V2 <- factor(df$V2, levels = c("+1", "+1 to all", "+1%", "+1% to all", "+0.1% to all"))


df_long <- reshape2::melt(df, id = c("V1", "V2"))
df_long$value <- as.numeric(df_long$value)

p1 <- ggplot(df_long, aes(x = V1, y = value, color = V1)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position=position_jitterdodge(dodge.width = 1), size = 1) +
  theme_classic() + xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  facet_grid(~V2, scale = "free") + NoLegend()
print(p1)


p2 <- ggplot(df_long, aes(x = V1, y = value, fill = variable)) +
  geom_col(position=position_dodge()) +
  theme_classic(base_size = 16) +
  xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  facet_grid(~V2, scale = "free") +
  guides(fill=guide_legend(title="Cell types"))
print(p2)

# Extract legend from p2
legend <- ggpubr::get_legend(p2)

# Remove legend from p2
p2 <- p2 + theme(legend.position = "none")

#Plot your table with table Grob in the library(gridExtra)
a_grob <- tableGrob(a_plot)
b_grob <- tableGrob(b_plot)
b1_grob <- tableGrob(b1_plot)
c_grob <- tableGrob(c_plot)
d_grob <- tableGrob(d_plot)
d1_grob <- tableGrob(d1_plot)

b_clr_grob <- tableGrob(b_clr_plot)
b1_clr_grob <- tableGrob(b1_clr_plot)
c_clr_grob <- tableGrob(c_clr_plot)
d_clr_grob <- tableGrob(d_clr_plot)
d1_clr_grob <- tableGrob(d1_clr_plot)

#Arrange them as you want with grid.arrange
lay <- rbind(c(1,2,2,2,2,2),
             c(1,2,2,2,2,2),
             c(1,2,2,2,2,2),
             c(1,2,2,2,2,2),
             c(3:8),
             c(NA, 9:13))

grid.arrange(
  legend,
  p2,
  a_grob,
  b_grob,
  b1_grob,
  c_grob,
  d_grob,
  d1_grob,
  b_clr_grob,
  b1_clr_grob,
  c_clr_grob,
  d_clr_grob,
  d1_clr_grob,
  layout_matrix = lay
)

pdf(file = file.path(path_plots, "zero_imputation.pdf"),   # The directory you want to save the file in
    width = 22, # The width of the plot in inches
    height = 9) # The height of the plot in inches
grid.arrange(
  legend,
  p2,
  a_grob,
  b_grob,
  b1_grob,
  c_grob,
  d_grob,
  d1_grob,
  b_clr_grob,
  b1_clr_grob,
  c_clr_grob,
  d_clr_grob,
  d1_clr_grob,
  layout_matrix = lay
)
dev.off()
```

## Number of zeros

```{r fig.height=6, fig.width=3}
a <- t(data.frame(c( 0, 0, 0,  3, 5, 100),
                  c( 0, 1, 1,  3, 5, 100)))


b <- a
for(col in 1:dim(b)[1]){
  b[col,] <- b[col,] + sum(b[col,]/100)
  b[col,] <- b[col,] / sum(b[col,]) * 100
}


df <- data.frame(t(data.frame(
  # Adding +1 pseudocount to all zeros
  c("More_zeros_sample", "# zeros", Hotelling::clr(t(b[1, ]))),
  c("Less_zeros_sample", "# zeros", Hotelling::clr(t(b[2, ])))
)))

df_long <- reshape2::melt(df, id = c("X1", "X2"))
df_long$value <- as.numeric(df_long$value)

ggplot(df_long, aes(x = X1, y = value, color = X1)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position=position_jitterdodge(dodge.width = 1), size = 1) +
  theme_classic() + xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  facet_grid(~factor(X2, levels = c("# zeros")), scale = "free") + NoLegend()
```

# Real data example
## Load data
```{r fig.height=4, fig.width=8}
obj.list <- read_objs(path_data)
celltype.compositions <- calc_CTcomp(obj.list, annot.cols = c("scGate_multi", "CD8_subtypes", "CD4_subtypes", "DC_subtypes", "MoMac_subtypes"))
```

## Plot data per patient
```{r fig.height=4, fig.width=10}
a <- celltype.compositions$MoMac_subtypes$cell_counts
samples <- rownames(a)

b <- a
b[b==0] <- 1
for(col in 1:dim(b)[1]){
  b[col,] <- b[col,] / sum(b[col,]) * 100
}

c <- a
for(col in 1:dim(c)[1]){
  c[col,][c[col,] == 0] <- sum(c[col,]/100)
  c[col,] <- c[col,] / sum(c[col,]) * 100
}

d <- a
for(col in 1:dim(d)[1]){
  d[col,][d[col,] == 0] <- sum(d[col,]/1000)
  d[col,] <- d[col,] / sum(d[col,]) * 100
}

e <- a[1,] * 10
e1 <- e
e2 <- e
e3 <- e
e1[e1==0] <- 1
e1 <- e1 / sum(e1) * 100
e2[e2==0] <- sum(e/100)
e2 <- e2 / sum(e2) * 100
e3[e3==0] <- sum(e/1000)
e3 <- e3 / sum(e3) * 100

df <- data.frame(t(data.frame(
  unlist(c(samples[1], "Count+1", Hotelling::clr(b[1,]))),
  unlist(c(samples[2], "Count+1", Hotelling::clr(b[2,]))),
  unlist(c(samples[3], "Count+1", Hotelling::clr(b[3,]))),
  unlist(c(samples[4], "Count+1", Hotelling::clr(b[4,]))),
  unlist(c(samples[5], "Count+1", Hotelling::clr(b[5,]))),
  unlist(c(samples[6], "Count+1", Hotelling::clr(b[6,]))),

  unlist(c(samples[1], "+1%", Hotelling::clr(c[1,]))),
  unlist(c(samples[2], "+1%", Hotelling::clr(c[2,]))),
  unlist(c(samples[3], "+1%", Hotelling::clr(c[3,]))),
  unlist(c(samples[4], "+1%", Hotelling::clr(c[4,]))),
  unlist(c(samples[5], "+1%", Hotelling::clr(c[5,]))),
  unlist(c(samples[6], "+1%", Hotelling::clr(c[6,]))),
  
  unlist(c(samples[1], "+0.1%", Hotelling::clr(d[1,]))),
  unlist(c(samples[2], "+0.1%", Hotelling::clr(d[2,]))),
  unlist(c(samples[3], "+0.1%", Hotelling::clr(d[3,]))),
  unlist(c(samples[4], "+0.1%", Hotelling::clr(d[4,]))),
  unlist(c(samples[5], "+0.1%", Hotelling::clr(d[5,]))),
  unlist(c(samples[6], "+0.1%", Hotelling::clr(d[6,]))),
  
  unlist(c(samples[1], ">1000_cells count+1", Hotelling::clr(e1))),
  unlist(c(samples[1], ">1000_cells_+1%", Hotelling::clr(e2))),
  unlist(c(samples[1], ">1000_cells_+0.1%", Hotelling::clr(e3)))
)))

df_long <- reshape2::melt(df, id = c("X1", "X2"))
df_long$value <- as.numeric(df_long$value)

ggplot(df_long, aes(x = X1, y = value, color = X1)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position=position_jitterdodge(), size = 1) +
  theme_classic() + xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  facet_grid(~factor(X2, levels = c("Count+1", "+1%", "+0.1%", ">1000_cells count+1", ">1000_cells_+1%", ">1000_cells_+0.1%")), scale = "free") + NoLegend()


f <- a
f <- f + 1
f <- f / sum(f) * 100

g <- a
for(col in 1:dim(g)[1]){
  g[col,] <- g[col,] + sum(g[col,]/100)
  g[col,] <- g[col,] / sum(g[col,]) * 100
}

h <- a
for(col in 1:dim(h)[1]){
  h[col,] <- h[col,] + sum(h[col,]/1000)
  h[col,] <- h[col,] / sum(h[col,]) * 100
}

i <- a[1,]*10
i1 <- i + 1
i1 <- i1 / sum(i1) * 100
i2 <- i + sum(i/100)
i2 <- i2 / sum(i2) * 100
i3 <- i + ifelse(sum(i/100) > 1, 1, sum(i/100))
i3 <- i3 / sum(i3) * 100

df <- data.frame(t(data.frame(
  unlist(c(samples[1], "Count+1", Hotelling::clr(f[1,]))),
  unlist(c(samples[2], "Count+1", Hotelling::clr(f[2,]))),
  unlist(c(samples[3], "Count+1", Hotelling::clr(f[3,]))),
  unlist(c(samples[4], "Count+1", Hotelling::clr(f[4,]))),
  unlist(c(samples[5], "Count+1", Hotelling::clr(f[5,]))),
  unlist(c(samples[6], "Count+1", Hotelling::clr(f[6,]))),

  unlist(c(samples[1], "+1%", Hotelling::clr(g[1,]))),
  unlist(c(samples[2], "+1%", Hotelling::clr(g[2,]))),
  unlist(c(samples[3], "+1%", Hotelling::clr(g[3,]))),
  unlist(c(samples[4], "+1%", Hotelling::clr(g[4,]))),
  unlist(c(samples[5], "+1%", Hotelling::clr(g[5,]))),
  unlist(c(samples[6], "+1%", Hotelling::clr(g[6,]))),
  
  unlist(c(samples[1], "+0.1%", Hotelling::clr(h[1,]))),
  unlist(c(samples[2], "+0.1%", Hotelling::clr(h[2,]))),
  unlist(c(samples[3], "+0.1%", Hotelling::clr(h[3,]))),
  unlist(c(samples[4], "+0.1%", Hotelling::clr(h[4,]))),
  unlist(c(samples[5], "+0.1%", Hotelling::clr(h[5,]))),
  unlist(c(samples[6], "+0.1%", Hotelling::clr(h[6,]))),
  
  unlist(c(samples[1], "x10 count+1", Hotelling::clr(i1))),
  unlist(c(samples[1], "x10 +1%", Hotelling::clr(i2))),
  unlist(c(samples[1], "x10 +1% max 1", Hotelling::clr(i3)))
)))

df_long <- reshape2::melt(df, id = c("X1", "X2"))
df_long$value <- as.numeric(df_long$value)

p <- ggplot(df_long, aes(x = X1, y = value, color = X1)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position=position_jitterdodge(), size = 1) +
  theme_classic() + xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  facet_grid(~factor(X2, levels = c("Count+1", "+1%", "+0.1%",
                                    "x10 count+1", "x10 +1%", "x10 +1% max 1")), scale = "free") + NoLegend()
print(p)
ggsave(plot = p, file.path(path_plots,"MoMac_perPatient_clr_zero_removed.png"),width = 10, height = 4, limitsize = F)
```

## Plot data per cell type
```{r fig.height=6, fig.width=10}
a <- celltype.compositions$MoMac_subtypes$cell_counts
samples <- rownames(a)

f <- a
for(col in 1:dim(f)[1]){
  f[col,] <- f[col,] + 1
  f[col,] <- Hotelling::clr(f[col,] / sum(f) * 100)
}
f <- t(f)
f <- cbind(rownames(f), rep("Count+1", dim(f)[1]), f)
df <- f

g <- a
for(col in 1:dim(g)[1]){
  g[col,] <- g[col,] + sum(g[col,]/100)
  g[col,] <- Hotelling::clr(g[col,] / sum(g[col,]) * 100)
}
g <- t(g)
g <- cbind(rownames(g), rep("+1%", dim(g)[1]), g)
df <- rbind(df, g)

h <- a
for(col in 1:dim(h)[1]){
  h[col,] <- h[col,] + sum(h[col,]/1000)
  h[col,] <- Hotelling::clr(h[col,] / sum(h[col,]) * 100)
}
h <- t(h)
h <- cbind(rownames(h), rep("+0.1%", dim(h)[1]), h)
df <- rbind(df, h)

i <- a
for(col in 1:dim(i)[1]){
  i[col,] <- i[col,] + sum(i[col,]/10)
  i[col,] <- Hotelling::clr(i[col,] / sum(i[col,]) * 100)
}
i <- t(i)
i <- cbind(rownames(i), rep("+10%", dim(i)[1]), i)
df <- rbind(df, i)

df_long <- reshape2::melt(data.frame(df), id = c("V1", "V2"))
df_long$value <- as.numeric(df_long$value)

p <- ggplot(df_long, aes(x = V1, y = value, color = V1)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position=position_jitterdodge(), size = 1) +
  theme_classic() + xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  facet_grid(~factor(V2, levels = c("Count+1", "+1%", "+0.1%", "+10%")), scale = "free") + NoLegend()
print(p)
ggsave(plot = p, file.path(path_plots,"MoMac_perCelltype_clr_zero_removed.png"),width = 8, height = 4, limitsize = F)
```


## Plot data per cell type CD8 for comparison
```{r fig.height=6, fig.width=5}
a <- celltype.compositions$CD8_subtypes$cell_counts
samples <- rownames(a)

f <- a
for(col in 1:dim(f)[1]){
  if(any(f[col,] == 0)) {
    f[col,] <- f[col,] + 1
  }
  f[col,] <- Hotelling::clr(f[col,] / sum(f) * 100)
}
f <- t(f)
f <- cbind(rownames(f), rep("Count+1", dim(f)[1]), f)
df <- f

g <- a
for(col in 1:dim(g)[1]){
  if(any(g[col,] == 0)) {
    g[col,] <- g[col,] + sum(g[col,]/100)
  }
  g[col,] <- Hotelling::clr(g[col,] / sum(g[col,]) * 100)
}
g <- t(g)
g <- cbind(rownames(g), rep("+1%", dim(g)[1]), g)
df <- rbind(df, g)

df_long <- reshape2::melt(data.frame(df), id = c("V1", "V2"))
df_long$value <- as.numeric(df_long$value)

p <- ggplot(df_long, aes(x = V1, y = value, color = V1)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position=position_jitterdodge(), size = 1) +
  theme_classic() + xlab("") + ylab("") +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  facet_grid(~factor(V2, levels = c("Count+1", "+1%")), scale = "free") + NoLegend()
print(p)
ggsave(plot = p, file.path(path_plots,"CD8_perCelltype_clr_zero_removed.png"),width = 6, height = 5, limitsize = F)
```
# Synthetic data

## functions

```{r}
library(dplyr)
calc_sil <- function(feat_mat,
                     labels,
                     return_mean = TRUE) {
  
  dist_mat <- dist(feat_mat)

  sils <- cluster::silhouette(x = as.numeric(factor(labels)),
                              dist = as.data.frame(dist_mat)) %>%
    as.data.frame() %>%
    dplyr::rename(group = cluster) %>%
    dplyr::mutate(group = labels) %>%
    dplyr::group_by(group) %>%
    dplyr::arrange(dplyr::desc(sil_width), .by_group = TRUE) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(rowid = dplyr::row_number())

  if (return_mean) {
    avg_sil <- mean(sils[["sil_width"]])
    return(avg_sil)
  } else {
    return(sils)
  }
}
```


## Counts vs freq vs clr

Why not use counts?

```{r fig.height=3, fig.width=5}
library(simstudy)
library(factoextra)
library(easyCODA)

# Define group
def <- defData(varname = "group", formula = "1;1;1", dist = "trtAssign")
def <- defRepeat(def, 2, "ct_const_", formula = "10", variance = 0.1,
    dist = "normal")
def <- defRepeat(def, 1, "ct_var1_", formula = "10 * group", variance = 0.1,
    dist = "normal")

df <- genData(9, def)

# Sort by group
df <- df[order(df$group), ]
# df$sample_id <- NULL
df$id <- 1:dim(df)[1]

df[c(2,5,8),3:5] <- df[c(2,5,8),3:5] * 10
df[c(3,6,9),3:5] <- df[c(3,6,9),3:5] * 100

df[,3:5] <- round(df[,3:5])

habillage = factor(df$group)

df_counts <- df
df_long <- reshape2::melt(df_counts, id = c("id", "group"))
ggplot(df_long, aes(x = id, y = value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap(. ~ group, scales = "free_x") +
  xlab("Sample") +
  ylab("Cell counts")
# calc_sil(feat_mat = df_counts[, 3:5],
#          labels = df_counts[, 2])
fviz_pca(prcomp(df_counts[,3:5]),
         habillage = habillage,
         invisible = c("var", "quali"),
         # label = "none",
         mean.point = FALSE)

df_perc <- df
df_perc[,3:5] <- df_perc[,3:5] / rowSums(df_perc[,3:5]) * 100
df_perc <- as.data.frame(df_perc)
df_long <- reshape2::melt(df_perc, id = c("id", "group"))
ggplot(df_long, aes(x = id, y = value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap(. ~ group, scales = "free_x") +
  xlab("Sample") +
  ylab("Relative abundance (%)")
fviz_pca(prcomp(df_perc[,3:5]),
         habillage = habillage,
         # invisible = c("var", "quali"),
         # label = "none",
         mean.point = FALSE)

df_clr <- df_perc
df_clr[,3:5] <- Hotelling::clr(df_clr[,3:5])
df_long <- reshape2::melt(df_clr, id = c("id", "group"))
ggplot(df_long, aes(x = id, y = value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap(. ~ group, scales = "free_x") +
  xlab("Sample") +
  ylab("Relative abundance (clr-transformed)")
fviz_pca(prcomp(df_clr[,3:5]),
         habillage = habillage,
         # invisible = c("var", "quali"),
         # label = "none",
         mean.point = FALSE)
```

## freq vs clr

Why not use freq?

```{r}
library(simstudy)
library(factoextra)
library(easyCODA)

# Define group
def <- defData(varname = "group", formula = "1;1;1", dist = "trtAssign")
def <- defRepeat(def, 1, "ct_const_low_", formula = "3", variance = 0,
    dist = "normal")
def <- defRepeat(def, 3, "ct_const_medium_", formula = "10", variance = 5,
    dist = "normal")
def <- defRepeat(def, 3, "ct_const_high_", formula = "100", variance = 5,
    dist = "normal")
def <- defRepeat(def, 1, "ct_var1_", formula = "1 * group", variance = 0,
    dist = "normal")

df <- genData(9, def)

# Sort by group
df <- df[order(df$group), ]
# df$id <- NULL
df$id <- 1:dim(df)[1]

# df[c(2,5,8),3:6] <- df[c(2,5,8),3:6] * 10
# df[c(3,6,9),3:6] <- df[c(3,6,9),3:6] * 100

df_cols <- dim(df)[2]

df[,3:df_cols] <- round(df[,3:df_cols])

habillage = factor(df$group)

df_perc <- df
df_perc[,3:df_cols] <- df_perc[,3:df_cols] / rowSums(df_perc[,3:df_cols]) * 100
df_perc <- as.data.frame(df_perc)
df_long <- reshape2::melt(df_perc, id = c("id", "group"))
ggplot(df_long, aes(x = id, y = value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") +
  xlab("Sample") +
  ylab("Relative abundance (%)")
fviz_pca(prcomp(df_perc[,3:df_cols]),
         habillage = habillage,
         # invisible = c("var", "quali"),
         # label = "none",
         mean.point = FALSE)

df_clr <- df_perc
df_clr[,3:df_cols] <- Hotelling::clr(df_clr[,3:df_cols])
df_long <- reshape2::melt(df_clr, id = c("id", "group"))
ggplot(df_long, aes(x = id, y = value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") +
  xlab("Sample") +
  ylab("Relative abundance (clr-transformed)")
fviz_pca(prcomp(df_clr[,3:df_cols]),
         habillage = habillage,
         # invisible = c("var", "quali"),
         # label = "none",
         mean.point = FALSE)
```

## +1 count, +1%, KNN, model

Why not use freq?

```{r}
library(simstudy)
library(factoextra)
library(easyCODA)

# Define group
def <- defData(varname = "group", formula = "1;1;1", dist = "trtAssign")
def <- defRepeat(def, 1, "ct_zero_imputed", formula = "3", variance = 0,
    dist = "normal")
def <- defRepeat(def, 1, "ct_10_", formula = "10", variance = 0,
    dist = "normal")
def <- defRepeat(def, 1, "ct_const_high_", formula = "100", variance = 0,
    dist = "normal")
def <- defRepeat(def, 1, "ct_var1_", formula = "1 * group", variance = 0,
    dist = "normal")

df <- genData(9, def)

# Sort by group
df <- df[order(df$group), ]
# df$id <- NULL
df$id <- 1:dim(df)[1]

# df[c(2,5,8),3:6] <- df[c(2,5,8),3:6] * 10
# df[c(3,6,9),3:6] <- df[c(3,6,9),3:6] * 100

df_cols <- dim(df)[2]

df[,3:df_cols] <- round(df[,3:df_cols])

habillage = factor(df$group)

df_perc <- df
df_perc[,3:df_cols] <- df_perc[,3:df_cols] / rowSums(df_perc[,3:df_cols]) * 100
df_perc <- as.data.frame(df_perc)
df_long <- reshape2::melt(df_perc, id = c("id", "group"))
ggplot(df_long, aes(x = id, y = value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") +
  xlab("Sample") +
  ylab("Relative abundance (%)")
fviz_pca(prcomp(df_perc[,3:df_cols]),
         habillage = habillage,
         # invisible = c("var", "quali"),
         # label = "none",
         mean.point = FALSE)

df_clr <- df_perc
df_clr[,3:df_cols] <- Hotelling::clr(df_clr[,3:df_cols])
df_long <- reshape2::melt(df_clr, id = c("id", "group"))
ggplot(df_long, aes(x = id, y = value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") +
  xlab("Sample") +
  ylab("Relative abundance (clr-transformed)")
fviz_pca(prcomp(df_clr[,3:df_cols]),
         habillage = habillage,
         # invisible = c("var", "quali"),
         # label = "none",
         mean.point = FALSE)
```



show example why ratios matter

rel abu vs clr (real-world data)(clustering e.g. tissue)
rel abu vs clr (real-world data)(clustering e.g. CD8.Tex)