---
title: "Zhang et al. dataset cell type annotation"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

Zhang Yuanyuan Cancer Cell 2021; CD45+ cells; primary or metastatic tumor tissues and peripheral blood of 22 advanced TNBC patients treated with paclitaxel or its combination with atezolizumab.

paper: https://www.sciencedirect.com/science/article/pii/S1535610821004992?via%3Dihub
accession:https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE169246

This script takes output of "scRNAseq_data_processing_ZhangY_2022_34653365.Rmd"

# Load libraries

```{r, include=FALSE, fig.width=16, fig.height=12}
# renv::init()
# renv::restore()
# renv::snapshot()
# 
# remotes::install_github("carmonalab/UCell", ref="f6e711c")
# remotes::install_github("carmonalab/scGate", ref="17a4ce6")
# remotes::install_github("carmonalab/STACAS", ref="0a88c5e")
# remotes::install_github("carmonalab/ProjecTILs", ref="fdc75e0")
# remotes::install_github("carmonalab/SignatuR", ref="0aded1d")
# remotes::install_github("carmonalab/HiTME", ref="73ff3ca", force = T)

# remotes::install_github("carmonalab/scooter", ref="f31eab3", force = T)

library(scooter)

# remotes::install_github("saezlab/MOFAcellulaR")

library(Seurat)
# devtools::install_github('satijalab/seurat-data')
library(SeuratData)
# install.packages("hdf5r")
# remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)
library(tidyverse)
library(scGate)
library(UCell)
library(STACAS)
library(scGate)
library(ProjecTILs)
library(SignatuR)
library(HiTME)
library(dplyr)
```


# Set paths
```{r}
my_seed <- 321
set.seed(my_seed)
ds_name <- "ZhangY_2022_34653365"

path_root <- file.path("~", "Dropbox", "CSI", "Standardized_SingleCell_Datasets", ds_name)
path_plots <- file.path("~", "Dropbox", "CSI", "PAPERS", "ECODA", "plots")
path_output <- file.path(path_root, "output")
path_output_samples <- file.path(path_output, "samples")
```


# Load dataset
```{r}
obj <- readRDS(file.path(path_output, sprintf("%s.rds", ds_name)))
```

# Filter out low cell count samples
```{r}
sample.size <- sort(table(obj$Sample))
sample.size
barplot(sort(table(obj$Sample)))
rm(sample.size)
```

# ------------------

# FOR PUBLICATION

## Number of cells per sample

```{r, fig.width=10, fig.height=5}
cells.perSample <- table(obj$Sample)

# Minimal theme + blue fill color
df <- data.frame(sort(cells.perSample))
p <- ggplot(data=df, aes(x = Var1, y = Freq)) +
  geom_bar(stat="identity", fill="steelblue") +
  xlab("") +
  ylab("Cells") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10))
ggsave(file.path(path_plots, "number_of_cells_per_sample.pdf"),
       plot = p,
       width = 8,
       height = 4)
```

# ------------------


```{r, fig.width=10, fig.height=5}
min.cells.perSample <- 1000
max.cells.perSample <- 12000
cells.perSample <- table(obj$Sample)

# Minimal theme + blue fill color
df <- data.frame(sort(cells.perSample))
p <- ggplot(data=df, aes(x = Var1, y = Freq)) +
  geom_bar(stat="identity", fill="steelblue") +
  xlab("") +
  ylab("Cells") +
  geom_abline(intercept = min.cells.perSample) +
  geom_abline(intercept = max.cells.perSample) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p
rm(p, df)
```


## Remove samples with low count
```{r}
Idents(obj) <- "Sample"
set.seed(my_seed)
obj <- subset(obj, idents = names(cells.perSample[cells.perSample > min.cells.perSample]))
# obj <- subset(obj, downsample = max.cells.perSample)
obj$Sample <- factor(obj$Sample)
rm(min.cells.perSample, max.cells.perSample)
```

```{r}
cells.perSample <- table(obj$Sample)
barplot(sort(cells.perSample))
rm(cells.perSample)
```



# Celltype annotation with HiTME

```{r}
# Get scGate database of pre-defined gating models
scGate_models_DB <- get_scGateDB(branch = "dev", verbose = T, force_update = TRUE)
scGate_models_blood <- scGate_models_DB$human$PBMC
scGate_models_blood$MoMac <- scGate_models_blood$Monocyte
scGate_models_blood$Monocyte <- NULL
scGate_models_tumor <- scGate_models_DB$human$HiTME
# Need to have same model names in both lists for HiTME downstream processing)
scGate_models_blood <- c(scGate_models_blood, scGate_models_tumor[!names(scGate_models_tumor) %in% names(scGate_models_blood)])

# Load ProjecTILs reference maps
path_ref <- "~/Dropbox/CSI/reference_atlases/sketched_200ct"
ref.maps_sketched <- list(CD8 = load.reference.map(file.path(path_ref, "sketched_CD8T_human_ref_v1.rds")),
                          CD4 = load.reference.map(file.path(path_ref, "sketched_CD4T_human_ref_v2.rds")),
                          DC = load.reference.map(file.path(path_ref, "sketched_DC_human_ref_v2.rds")),
                          MoMac = load.reference.map(file.path(path_ref, "sketched_MoMac_human_v1.rds"))
)

obj.list <- SplitObject(obj, split.by = "Sample")
rm(obj)
gc()

for (i in 1:length(obj.list)) {
  print(i)
  if (obj.list[[i]]$Tissue[1] == "blood") {
    obj.list[[i]] <- Run.HiTME(object = obj.list[[i]],
                               scGate.model = scGate_models_blood,
                               ref.maps = ref.maps_sketched
    )
  } else {
    obj.list[[i]] <- Run.HiTME(object = obj.list[[i]],
                               scGate.model = scGate_models_tumor,
                               ref.maps = ref.maps_sketched
    )
  }
}

rm(i, ref.maps_sketched)
gc()

save_objs(obj.list, path_output_samples)
```