---
title: Standardized Processing of scRNA-seq datasets, WuS_2021_34493872
author: "M. Andreatta <massimo.andreatta at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

***
Dataset name: WuS_2021_34493872 (Author_Year_PBMID_cohortID)
***

scRNA-Seq (Chromium, 10X Genomics) on 26 primary tumors from three major clinical subtypes of breast cancer, including 11 ER+, 5 HER2+ and 10 TNBC.

```{r, message=FALSE}
renv::activate()
renv::restore()

#remotes::install_github("carmonalab/scGate",ref="v1.2.0")
#remotes::install_github("carmonalab/SignatuR",ref="716d829")
#renv::snapshot()

library(ggplot2)
library(Seurat)
library(UCell)
library(scGate)
library(SingleCellExperiment)
library(viridis)
library(Matrix)
library(patchwork)
library(R.utils)
library(data.table)
source("functions.R")
my_seed <- 321
```

# Set paths
```{r}
set.seed(my_seed)
ds_name <- "WuS_2021_34493872"
ID <- "WuS"
species <- "human"

input_data_path_root <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets",ds_name,"data")

output_data_path <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets",ds_name,"output")
dir.create(file.path(output_data_path), recursive = T)

output_plots_path <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets",ds_name,"plots")
dir.create(file.path(output_plots_path), recursive = T)
```


#Load data

GEO entry: GSE176078

```{r}
data.files <- list()
fpath <- file.path(input_data_path_root,"BrCa_Atlas_Count_out")

aggr <- ReadMtx(mtx=sprintf("%s/matrix.mtx.gz", fpath),
               cells=sprintf("%s/barcodes.tsv.gz", fpath),
               features=sprintf("%s/features.tsv.gz", fpath)
               )
```

**CHECK**:
* Make sure data matrix has genes as rows and cells as columns

* Map gene ids with our reference gene symbols. EnsemblGenes105_Hsa_GRCh38.p13 and EnsemblGenes105_Mmu_GRCm39 retreived 14.02.2022 and stored in EnsemblGenes105_Mmu_GRCm39.rds (only unique Gene stable ID)

* To make cell ids compatible with other datasets, append the study ID at the beginning of cell name

* Are raw counts available or only pre-normalized (eg log2TMP); if pre-normalize specify how was the normalization performed

***
Data input format:   pre-normalized, "by total counts, as described in https://doi.org/10.1016/j.cell.2015.04.044"
***



# Standardize gene symbols in expression matrix  (custom)
```{r}
if (species == "mouse") {
   geneRef_file <- "aux/EnsemblGenes105_Mmu_GRCm39.txt.gz"
} else {
   geneRef_file <- "aux/EnsemblGenes105_Hsa_GRCh38.p13.txt.gz"
}
geneRef <- data.table::fread(geneRef_file)

aggr <- standardizeGeneSymbols(aggr, EnsemblGeneTable=geneRef)
```

# Create Seurat objects (custom)
```{r}
data_seurat<- CreateSeuratObject(counts=aggr, project=ds_name, 
                                 assay="RNA", min.cells = 0, min.features = 0)

```

# Add metadata (if available)
```{r}
mpath <- file.path(input_data_path_root,"Whole_miniatlas_meta.csv")

meta <- read.csv(mpath)
#first row is an extra header
meta <- meta[-1,]

rownames(meta) <- meta$NAME

#check barcodes are consistent
identical(rownames(meta), rownames(data_seurat@meta.data))

data_seurat <- AddMetaData(data_seurat, metadata=meta)
```

# append the study ID at the beginning of cell name  (custom)
```{r}
#We do it on the Seurat object, after adding metadata

data_seurat <- RenameCells(data_seurat, add.cell.id = 'WuS')
```

CHECK slots

TODO If data pre-normalized apply inverse of NormalizeData() into "counts" slot

```{r}
head(data_seurat@meta.data$nCount_RNA)
```

# Clean cell metadata  (custom)

CHECK: cell metadata fields: 
Sample, Individual, Tissue c("blood","tumor","adjacent","LN","other"), Study, OriginalAnnotationLevel1, OriginalAnnotationLevel2, OriginalAnnotationLevel3
```{r}
colnames(data_seurat@meta.data)

table(data_seurat@meta.data$celltype_major)
table(data_seurat@meta.data$celltype_minor)

data_seurat$Study <- ID
data_seurat$Tissue <- "tumor"
data_seurat$Sample <- data_seurat$Patient

data_seurat$OriginalAnnotationLevel1 <- data_seurat@meta.data$celltype_major
data_seurat$OriginalAnnotationLevel2 <- data_seurat@meta.data$celltype_minor
data_seurat$OriginalAnnotationLevel3 <- NA

data_seurat$nCount_RNA <- as.numeric(data_seurat$nCount_RNA)
data_seurat$nFeature_RNA <- as.numeric(data_seurat$nFeature_RNA)

head(data_seurat@meta.data)
```


CHECK are data pre-normalized? decide whether to Normalize
```{r}
data_seurat <- NormalizeData(data_seurat) 
```

# Calculate variable genes

```{r}
#Get blocklists from SignatuR
library(SignatuR)

#genesBlocklist <- readRDS("aux/genes.blocklist.rds")
if (species == "mouse") {
   sig <- SignatuR$Mm
   
   genesBlocklist <- GetSignature(sig$Programs)
   genesBlocklist <- c(genesBlocklist, GetSignature(sig$Compartments))
} else {
   sig <- SignatuR$Hs
   
   genesBlocklist <- GetSignature(sig$Programs)
   genesBlocklist <- c(genesBlocklist, GetSignature(sig$Compartments))
   genesBlocklist <- c(genesBlocklist, GetSignature(sig$Blocklists))
}

set.seed(my_seed)
data_seurat <- select.variable.genes(data_seurat,
                                 nfeat=1000,
                                 blacklist=genesBlocklist,
                                 plot=TRUE)

paste("# variable genes",length(data_seurat@assays$RNA@var.features))

```


# Calculate Ribosomal and mitochondrial content
```{r}
if (species == "mouse") {
  patterns <- c("^Rp","^mt-")
} else {
  patterns <- c("^RP[LS]","^MT-")
}

pc.ribo <- PercentageFeatureSet(data_seurat, pattern = patterns[1])
pc.mito <- PercentageFeatureSet(data_seurat, pattern = patterns[2])

data_seurat <- AddMetaData(data_seurat, metadata = pc.ribo, col.name = "percent.ribo")
data_seurat <- AddMetaData(data_seurat, metadata = pc.mito, col.name = "percent.mito")
```


# Filter out outliers & low quality cells (custom)
```{r fig.height= 5}
Idents(data_seurat) <- "Sample"
VlnPlot(data_seurat, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0.01)
```
# Standard check
```{r}
probs <- list()

probs$percent.ribo <- quantile(data_seurat$percent.ribo, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))
probs$percent.mito <- quantile(data_seurat$percent.mito, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))
probs$nFeature_RNA <- quantile(data_seurat$nFeature_RNA, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))
probs$nCount_RNA <- quantile(data_seurat$nCount_RNA, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))

as.data.frame(probs)
```

# Custom thresholds
```{r}
cutoffs <- list()
cutoffs[["percent.ribo"]] <- c(min=0,max=50)
cutoffs[["percent.mito"]] <- c(min=0,max=20)
cutoffs[["nFeature_RNA"]] <- c(min=200,max=7500) 
cutoffs[["nCount_RNA"]] <- c(min=300,max=50000)
print(cutoffs)
```

# (custom)

CHECK: if pre-processed nCount_RNA makes no sense

```{r fig.height= 10}

initialCellNum <- ncol(data_seurat)
message(paste("Original number of cells:",ncol(data_seurat)))
data_seurat <- subset(data_seurat, subset = 
                        nFeature_RNA > cutoffs[["nFeature_RNA"]]["min"] & nFeature_RNA < cutoffs[["nFeature_RNA"]]["max"] & 
                        nCount_RNA   > cutoffs[["nCount_RNA"]]["min"]   & nCount_RNA   < cutoffs[["nCount_RNA"]]["max"]   &
                        percent.ribo > cutoffs[["percent.ribo"]]["min"] & percent.ribo < cutoffs[["percent.ribo"]]["max"] &
                        percent.mito > cutoffs[["percent.mito"]]["min"] & percent.mito < cutoffs[["percent.mito"]]["max"] 
                      )

l <- ncol(data_seurat)

message(sprintf("Number of cells after QC: %i (%.2f %% of input)", l, 100*l/initialCellNum))
```


```{r fig.height= 10}
Idents(data_seurat) <- "Sample"
VlnPlot(data_seurat, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito","PTPRC","CD2"), ncol = 3, pt.size=0.01)
ggsave(file.path(output_plots_path,"QC_Vln.png"),width = 10, height = 8)
```

# Scale data; run default PCA & UMAP
```{r}
set.seed(my_seed)
ndim <- 30

data_seurat <- ScaleData(data_seurat)
data_seurat <- RunPCA(data_seurat, npcs = ndim, ndims.print = 1:5, nfeatures.print = 5)
data_seurat <- RunUMAP(data_seurat, reduction = "pca", dims = 1:ndim, seed.use=my_seed)
```

# UMAP plots
```{r fig.height=5}
pll <- list()
pll[[1]] <- DimPlot(data_seurat, reduction = "umap", group.by = "Patient") + theme(aspect.ratio=1)
pll[[2]] <- DimPlot(data_seurat, reduction = "umap", group.by = "OriginalAnnotationLevel1", label=T) + 
  theme(aspect.ratio=1) + NoLegend()
pll[[3]] <- DimPlot(data_seurat, reduction = "umap", group.by = "OriginalAnnotationLevel2", label=T) + 
  theme(aspect.ratio=1) + NoLegend()

wrap_plots(pll)
ggsave(file.path(output_plots_path,"UMAP_meta.png"),width = 14, height = 4)
```



```{r fig.height=6}
myFeatures <- c("PTPRC","EPCAM","CSF1R","CD68","CD2","CD3D","MS4A1","BANK1","JCHAIN","PECAM1","PDGFRB","KLRD1","FAP","PDGFRA","COL1A1")

FeaturePlot(data_seurat, reduction = "umap", features = myFeatures, coord.fixed=T, ncol=5) 

ggsave(file.path(output_plots_path,"UMAP_features.png"),width = 20, height = 10)
```


# (optional) run scGate
```{r, eval=T}
set.seed(my_seed)
#Load default models from DB
models.DB <- scGate::get_scGateDB(version="v0.7")  ### Specifying a version ensure reproducibility of signatures

if (species=="mouse") {
   models <- models.DB$mouse$generic
} else {
   models <- models.DB$human$generic  ## Pre-compiled list of models for human cell types
}
names(models)

models <- models[c("Bcell","Tcell","Erythrocyte","Megakaryocyte","MoMacDC","NK","Plasma_cell")]


models$CAF <- scGate::gating_model(name="CAF",signature = c("FAP","PDGFRB","PDGFRA","COL1A1","PTPRC-"))
models$Endothelial <- scGate::gating_model(name="Endothelial",signature = c("CD34","PECAM1","PTPRC-"))
models$Epithelial <- scGate::gating_model(name="Epith",signature = c("EPCAM","PTPRC-"))

data_seurat <- scGate(data_seurat, model = models, ncores = 8, reduction = "pca", verbose = T)
```

See "multi" classification
```{r}
DimPlot(data_seurat, reduction="umap", group.by = "scGate_multi", label = T)

ggsave(file.path(output_plots_path,"UMAP_scGate_multi.png"),width = 9, height = 5)
```

```{r , eval=T, fig.height=4}
pll2 <- list()
for(gate in names(models)){
    cname <- paste0("is.pure_", gate)
    pll2[[gate]] <- DimPlot(data_seurat, reduction = "umap", group.by = cname,
                            cols = c(list(Impure = "gray", Pure = "green"))) + 
      theme(aspect.ratio=1)
}
wrap_plots(pll2)
ggsave(file.path(output_plots_path,"UMAP_scGate_single.png"),width = 12, height = 8)
```



# Export standardized dataset
```{r}
saveRDS(data_seurat,file.path(output_data_path,paste0(ds_name,".rds")))
```

