---
title: Standardized Processing of ZhangY_2022_34653365
author: "S. Carmona <santiago.carmona at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

***
Dataset name:   ZhangY_2022_34653365 (Author_Year_PBMID_cohortID)
***

**CHECK**: Make sure the input data and metadata are available in our repository

```{r, message=FALSE}
# renv::activate()
# renv::restore()

library(Seurat)
# library(limma)
# library(presto)
# library(BPCells)

library(ggplot2)
library(UCell)
library(scGate)
library(SingleCellExperiment)
library(viridis)
library(Matrix)
library(patchwork)
library(GEOquery)
library(dplyr)
library(data.table)
library(STACAS)
library(SignatuR)
my_seed <- 321
```

# Set paths
```{r}
set.seed(my_seed)
ds_name <- "ZhangY_2022_34653365"
ID <- "ZhyA"
species <- "human"

path_root <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets",ds_name)
dir.create(file.path(path_root))
path_data <- file.path(path_root, "data")
dir.create(file.path(path_data))
path_output <- file.path(path_root, "output")
dir.create(file.path(path_output))
path_plots <- file.path(path_root, "plots")
dir.create(file.path(path_plots))
```


#If not done before, download data eg fro GEO
```{r, eval=FALSE}
# geo_acc <- 'GSE169246'
# gse <- getGEO(geo_acc)
# getGEOSuppFiles(geo_acc)
# system(sprintf("tar -xvf %s/GSE169246_RAW.tar -C %s", geo_acc, path_root))
```


#Load Data
```{r}
data_matrix <- ReadMtx(mtx = file.path(path_data, "GSE169246_TNBC_RNA.counts.mtx.gz"),
                       cells = file.path(path_data, "GSE169246_TNBC_RNA.barcode.tsv"),
                       features = file.path(path_data, "GSE169246_TNBC_RNA.feature.tsv"), cell.column = 1, feature.column = 1)
```


**CHECK**:
* Make sure data matrix has genes as rows and cells as columns

* Map gene ids with our reference gene symbols. EnsemblGenes105_Hsa_GRCh38.p13 and EnsemblGenes105_Mmu_GRCm39 retreived 14.02.2022 and stored in EnsemblGenes105_Mmu_GRCm39.rds (only unique Gene stable ID)

* To make cell ids compatible with other datasets, append the study ID at the beginning of cell name

* Are raw counts available or only pre-normalized (eg log2TMP); if pre-normalized, specify how was the normalization performed

***
Data input format:   pre-normalized, "by total counts, as described in https://doi.org/10.1016/j.cell.2015.04.044"
***

```{r}
dim(data_matrix)
data_matrix[1:10,1:10]
head(colnames(data_matrix))
```

# extract meta data from colnames
```{r}
cell_names <- colnames(data_matrix)
cell_names.split <- strsplit(cell_names,"\\.")
meta <- list()
meta[["Original_cell_names"]] <- cell_names
meta[["Barcode"]] <- sapply(cell_names.split,function(x) x[[1]])
cell_names.split.2 <- sapply(cell_names.split,function(x) x[[2]])
meta[["Sample"]] <- cell_names.split.2
cell_names.split.2.split <- strsplit(cell_names.split.2,"\\_")
meta[["Treatment"]] <- sapply(cell_names.split.2.split,function(x) x[[1]])
meta[["Individual"]] <- sapply(cell_names.split.2.split,function(x) x[[2]])
meta[["Tissue"]] <- sapply(cell_names.split.2.split,function(x) x[[3]])

sapply(meta,head)
meta <- as.data.frame(meta)
```


# append the study ID at the beginning of cell name  (custom)
```{r}
prefix <- paste(ID, meta$Individual, meta$Tissue,meta$Treatment,sep=".")  #fields in prefix are separated by dots
cell_ids <- paste(prefix, meta$Barcode,sep = '_') #underscore delimits end of prefix
colnames(data_matrix)<- cell_ids
data_matrix[1:10,1:2]
```

# Adapt metadata  (custom)
```{r}
rownames(meta) <- cell_ids
head(meta)
```


# Collect additional metadata from Supplemental information:
https://ars.els-cdn.com/content/image/1-s2.0-S1535610821004992-mmc2.xlsx
```{r}
excel_file_path <- file.path(path_data, "metadata.xlsx")
download.file(url = "https://ars.els-cdn.com/content/image/1-s2.0-S1535610821004992-mmc2.xlsx",
              destfile = excel_file_path)
```

```{r}
library(tidyverse)

# Read the xlsx file
excel_data <- readxl::read_xlsx(excel_file_path)

# Combine Row2 and Row3 into a single row and set as header
new_header <- paste(excel_data[1,], excel_data[2,], sep = "_")
new_header <- gsub("_NA", "", new_header)
new_header <- gsub("NA_", "", new_header)
names(excel_data) <- new_header

excel_data <- excel_data[3:24, c(1:9, 16:19)]

library(tidyr)
excel_data <- excel_data %>% fill(Treatment)

names(excel_data) <- make.names(names(excel_data))

names(excel_data)[names(excel_data) == "Treatment"] <- "Treatment_type"
names(excel_data)[names(excel_data) == "Patient.ID"] <- "Individual"

rnames <- rownames(meta)
meta <- meta %>%
  left_join(excel_data, by = "Individual")
rownames(meta) <- rnames
```


# Create Seurat object (custom)
```{r}
data_seurat <- CreateSeuratObject(counts=data_matrix, project=ds_name, meta.data = meta, assay="RNA", min.cells = 0, min.features = 0)
rm(data_matrix,meta)
```

# Standardize gene symbols 
```{r}
if (species == "mouse") {
   geneRef_file <- "aux/EnsemblGenes105_Mmu_GRCm39.txt.gz"
} else {
   geneRef_file <- "aux/EnsemblGenes105_Hsa_GRCh38.p13.txt.gz"
}
geneRef <- fread(geneRef_file)
```


```{r}
data_seurat <- STACAS::StandardizeGeneSymbols(data_seurat, EnsemblGeneFile = geneRef_file, slots = "counts")
```

CHECK slots

TODO If data pre-normalized, apply inverse of NormalizeData() into "counts" slot. Then we can run NormalizeData() rountinely

```{r}
head(data_seurat@meta.data)
```

# Clean cell metadata  (custom)

CHECK: cell metadata fields: 
Sample, Individual, Tissue c("blood","tumor","adjacent","LN","other"), Study, OriginalAnnotationLevel1, OriginalAnnotationLevel2, OriginalAnnotationLevel3

```{r}
data_seurat$orig.ident <- cell_ids
data_seurat$Sample <- factor(data_seurat$Sample)
data_seurat$Individual <- factor(data_seurat$Individual)
data_seurat$Tissue <- factor(data_seurat$Tissue)
data_seurat$Tissue <- recode(data_seurat$Tissue, b = "blood", t = "tumor")
#data_seurat$OriginalAnnotationLevel1 <- 
#data_seurat$OriginalAnnotationLevel2 <- 
#data_seurat$OriginalAnnotationLevel3 <- 
head(data_seurat)
```

# Add features/genes metadata
```{r}
features_meta_data <- geneRef[!duplicated(geneRef$`Gene name`),]
setDF(features_meta_data)
rownames(features_meta_data) <- features_meta_data$`Gene name`
rownames(features_meta_data) <- gsub("_","-",rownames(features_meta_data)) # in Seurat, underscores ('_'), are replaced with dashes ('-') in gene names; 5S_rRNA; but these just a few
features_meta_data <- features_meta_data[rownames(data_seurat),]

if ( all.equal(rownames(features_meta_data),as.character(rownames(data_seurat))) ) {
  data_seurat[["RNA"]] <- AddMetaData(data_seurat[["RNA"]],metadata = features_meta_data)
  rm(geneRef,features_meta_data)
}

```


CHECK are data pre-normalized? decide whether to Normalize
```{r}
data_seurat <- NormalizeData(data_seurat) 
```


# Calculate Ribosomal and mitochondrial content

Package SignatuR contains gene sets including mito/ribo genes
```{r}
SignatuR
```


```{r}
ribo.genes <- GetSignature(SignatuR$Hs$Compartments$Ribo)$Ribo
mito.genes <- GetSignature(SignatuR$Hs$Compartments$Mito)$Mito

data_seurat <- AddMetaData(data_seurat, metadata = PercentageFeatureSet(data_seurat, features = ribo.genes[ribo.genes %in% rownames(data_seurat)]), col.name = "percent.ribo")
data_seurat <- AddMetaData(data_seurat, metadata = PercentageFeatureSet(data_seurat, features = mito.genes[mito.genes %in% rownames(data_seurat)]), col.name = "percent.mito")
```

CHECK: distributions across samples
```{r fig.height= 10}
Idents(data_seurat) <- "Sample"
VlnPlot(data_seurat, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito","PTPRC","MALAT1"), ncol = 3, pt.size=0)
```


# Filter out outliers & low quality cells (custom)
```{r fig.height= 10}
Idents(data_seurat) <- "Sample"
VlnPlot(data_seurat, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0.01)
```

# Standard check
```{r}
probs <- list()

probs$percent.ribo <- quantile(data_seurat$percent.ribo, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))
probs$percent.mito <- quantile(data_seurat$percent.mito, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))
probs$nFeature_RNA <- quantile(data_seurat$nFeature_RNA, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))
probs$nCount_RNA <- quantile(data_seurat$nCount_RNA, probs=c(0.001, 0.005, 0.01, 0.02, 0.98, 0.99, 0.995, 0.999))

probs
```

# Custom thresholds
```{r}
cutoffs <- list()
cutoffs[["percent.ribo"]] <- c(min=0,max=60)
cutoffs[["percent.mito"]] <- c(min=0,max=20)
cutoffs[["nFeature_RNA"]] <- c(min=400,max=6000) 
cutoffs[["nCount_RNA"]] <- c(min=600,max=40000)
print(cutoffs)
```

How many cells per sample?
```{r}
sample.size.pre <- sort(table(data_seurat$Sample))
sample.size.pre
```


CHECK: if pre-processed nCount_RNA is informative (eg yes if UMI counts, no if TPM)

```{r fig.height= 10}

initialCellNum <- ncol(data_seurat)
message(paste("Original number of cells:", initialCellNum))
data_seurat <- subset(data_seurat, subset = 
                        nFeature_RNA >= cutoffs[["nFeature_RNA"]]["min"] & nFeature_RNA < cutoffs[["nFeature_RNA"]]["max"] & 
                        nCount_RNA   >= cutoffs[["nCount_RNA"]]["min"]   & nCount_RNA   < cutoffs[["nCount_RNA"]]["max"]   &  
                        percent.ribo >= cutoffs[["percent.ribo"]]["min"] & percent.ribo < cutoffs[["percent.ribo"]]["max"] &
                        percent.mito >= cutoffs[["percent.mito"]]["min"] & percent.mito < cutoffs[["percent.mito"]]["max"] 
                      )

l <- ncol(data_seurat)

message(sprintf("Number of cells after QC: %i (%.2f %% of input)", l, 100*l/initialCellNum))
```

How many cells per sample after QC?
```{r}
sample.size <- table(data_seurat$Sample)[names(sample.size.pre)]
sample.size
barplot(sort(sample.size))
sample.size.filterRatio <- round((sample.size.pre - sample.size)/sample.size.pre*100,2)
sample.size.filterRatio
barplot(sample.size.filterRatio[names(sort(sample.size))])
```


For QC, downsample and make some diagnostic analyses/plots

Set params:
```{r}
ds <- 1000 # number of cells per sample
nfeatures <- 2000 # number of variable genes selected for dim reduction 
ndim <- 20 # number of principal components for dim. reduction
```


Samples cell downsampling
```{r}
Idents(data_seurat) <- "Sample"
hist(table(data_seurat$Sample))
set.seed(123)
data_seurat.ds <- subset(data_seurat, downsample=1000)
```

Find variable genes (we usually exclude some gene sets)
```{r}

my.genes.blocklist <- GetSignature(SignatuR$Hs)[c("Pseudogenes", "HSP", "Non-coding", "cellCycle.G1S", "cellCycle.G2M", 
"Mito", "Ribo", "TCR", "Immunoglobulins")]

data_seurat.ds <- STACAS::FindVariableFeatures.STACAS(data_seurat.ds, nfeat=nfeatures, genesBlockList = my.genes.blocklist)
```


# Scale data; run default PCA & UMAP
```{r}
set.seed(my_seed)
data_seurat.ds <- ScaleData(data_seurat.ds)
data_seurat.ds <- RunPCA(data_seurat.ds, ndims.print = 1:5, nfeatures.print = 5, npcs=ndim)
data_seurat.ds <- RunUMAP(data_seurat.ds, reduction = "pca", dims = 1:ndim, seed.use=my_seed)
```

# UMAP plots
```{r fig.height=10}
pll <- list()
pll[[1]] <- DimPlot(data_seurat.ds, reduction = "umap", group.by = "Sample") + theme(aspect.ratio=1)
pll[[2]] <- DimPlot(data_seurat.ds, reduction = "umap", group.by = "Individual") + theme(aspect.ratio=1)
pll[[3]] <- DimPlot(data_seurat.ds, reduction = "umap", group.by = "Treatment") + theme(aspect.ratio=1)
pll[[4]] <- DimPlot(data_seurat.ds, reduction = "umap", group.by = "Tissue") + theme(aspect.ratio=1)
wrap_plots(pll)
ggsave(file.path(path_plots,"UMAP_metadata.png"),width = 40, height = 40)
```

```{r fig.height=10}
myFeatures <- c("PTPRC","CD2","FOXP3","CD8A","SPI1","CD79A","MKI67","COL1A1") # just to explore
FeaturePlot(data_seurat.ds, reduction = "umap", features = myFeatures, coord.fixed=T, ncol=5, order = T) 
ggsave(file.path(path_plots,"UMAP_features.png"),width = 40, height = 30)
```

# Export standardized dataset
```{r}
saveRDS(data_seurat, file.path(path_output, paste0(ds_name,".rds")))
```
