---
title: Standardized Processing of scRNA-seq datasets
author: "M. Andreatta <massimo.andreatta at unil.ch>, S. Carmona <santiago.carmona at unil.ch>, C. Halter <christian.halter at unil.ch>, J. Garnica <josep.garnicacaparros at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

***
Dataset name:   Author_Year_PBMID_cohortID (eg. ZhangY_2022_34653365)
```{r dataset_name}
ds_name <- "Kfoury_2021_34719426"
```
***

Set root directory for processing
```{r set_dir}
# Set path to the directory were
# Set path to the directory were to get input and output
home.path <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets", ds_name)

knitr::opts_chunk$set(echo = T, warning = F,include=T, message=F)
```

**CHECK**: Make sure the input data and metadata are available in our repository. 

# Prepare packages and environtment
```{r load_packages_functions, message=FALSE}
# restore packages versions with renv.lock file from github repository
renv::activate()
renv::restore()

# load packages
pckgs <- c("tidyverse", "Seurat", "UCell", "scGate",
           "SingleCellExperiment","Matrix", "patchwork",
           "data.table", "STACAS", "SignatuR")
for(package in pckgs){
  library(package, character.only = T)
}

# load custom functions
source("Rutils/functions.R")

options(timeout=4000) # increase time out for downloading
```

Also, increase memory parameters to allocate large vectors, otherwise R might crash
```{bash create_Renviron}
#include vector max size
max_size=200Gb

# create .Renviron file if not present
if ! [ -a ".Renviron" ]
then
  touch .Renviron
  echo R_MAX_VSIZE=$max_size > .Renviron
  echo New .Renviron file created!
else
  echo .Renviron already present!
  echo Check size of the allowed vector:
  cat .Renviron
fi
```


# Set parameters
Dataset ID should have 3 letters from Author (Aaa) followed by additional uniquifying letter (A). If multiple cohort, followed by 1 letter from cohort ID separated by dot (e.g. AndA.H; AndB.M; And.C, etc.)
```{r define_parameters}
# Assign a short id for this data source
ID <- "KfoY"

# Assign species of the data
species <- "human"

#set seed
my_seed <- 321
set.seed(my_seed)

# max number of cells per sample (if higher, they will downsampled)
# make sure to set seed for this step
ds <- 1000

# Optional: min number of cells per sample, samples with lower number will be discarded
min.cells <- 500 
```


# Set paths
Declare the paths in your directory or Dropbox
```{r}
# Create directory to store input, output and plots
dir.create(home.path)

# Create directories for input, output and plots
input_data_path_root <- file.path(home.path, "data")
dir.create(file.path(input_data_path_root))
output_data_path <- file.path(home.path, "output")
dir.create(file.path(output_data_path))
output_plots_path <- file.path(home.path,"plots")
dir.create(file.path(output_plots_path))
```

# Download data

```{r download_GEO}
download.file(url = "https://ghoffman-cdn.s3.us-east-2.amazonaws.com/dreamlet_analysis/data/Kfoury_CancerCell_2021.h5ad")
```



# Load Data
Depending on the data format at GEO or Zenodo, use different tools to load the datasets into the environment.

RDS files
```{r load_rds, eval=F}
install.packages("anndata")
library(anndata)
data <- read_h5ad(file.path(input_data_path_root, "Kfoury_CancerCell_2021.h5ad"))
data_seurat <- CreateSeuratObject(counts = t(as.matrix(data$X)), meta.data = data$obs, min.features = 1, min.cells = 1)
```

**CHECK**:

1. Assess available assays: Are raw counts available or only pre-normalized (eg log2TMP); if pre-normalized, specify how was the normalization performed!

2. Make sure data matrix has genes as rows and cells as columns

3. Include, adapt and standarize cell metadata

4. Modify cell identifiers (`orig.ident`)

5. Standardize gene symbols

6. Add features/genes metadata


## 3. Include, adapt and standarize cell metadata

Generate a metadata dataframe including the following variables for each cell in the dataset. If some are not present include NA. If needed adapt metadata (factorize...).

+----------------------------+--------------------------------------+
| Variable                   | Example                              |
+============================+======================================+
| Original_cell_names        | AAACCTGAGGTTACCT.Pre_P007_b          |
+----------------------------+--------------------------------------+
| Barcode                    | AAACCTGAGGTTACCT                     |
+----------------------------+--------------------------------------+
| Study                      | Author_Year_PBMID_cohortID           |
|                            |                                      |
|                            | (e.g. ZhangY_2022_34653365)          |
+----------------------------+--------------------------------------+
| Sample                     | Pre_P007_b                           |
+----------------------------+--------------------------------------+
| Treatment                  | Pre                                  |
+----------------------------+--------------------------------------+
| Individual                 | P007                                 |
+----------------------------+--------------------------------------+
| Tissue                     | blood, tumor, adjacent, LN, other    |
+----------------------------+--------------------------------------+
| OriginalAnnotationLevel{n} | CD4T                                 |
+----------------------------+--------------------------------------+

```{r adapt_meta}
# Explore data and modifiy this code accordinly to fill the metadata slots

data_seurat[["Original_cell_names"]] <- row.names(data_seurat@meta.data)
names(data_seurat@meta.data)[names(data_seurat@meta.data) == 'tissue'] <- 'Tissue'
names(data_seurat@meta.data)[names(data_seurat@meta.data) == 'subject.id'] <- 'Individual'
names(data_seurat@meta.data)[names(data_seurat@meta.data) == 'geoid'] <- 'Sample'
names(data_seurat@meta.data)[names(data_seurat@meta.data) == 'barcode'] <- 'Barcode'
data_seurat$Barcode <- sapply(strsplit(data_seurat$Barcode, "_"),function(x) x[[2]])

head(data_seurat@meta.data)
```

## 4. Modify cell identifiers (`orig.ident`)
To make cell ids compatible with other datasets, append the study ID at the beginning of the original cell name. If available, also add other metadata elements such as Individual, tissue or treatment.

```{r modify_cellID}
#fields in prefix are separated by dots
prefix <- paste(ID, data_seurat$Individual, data_seurat$Tissue, data_seurat$Status,sep=".")

# underscore delimits end of prefix
cell_ids <- paste(prefix, data_seurat$Barcode, sep = '_') 
colnames(data_seurat) <- cell_ids
data_seurat[1:10,1:2]
```

## 5. Standardize gene symbols 
Map gene ids with our reference gene symbols using `STACAS::StandardizeGeneSymbols()`. Load EnsemblGenes105_Hsa_GRCh38.p13 and EnsemblGenes105_Mmu_GRCm39 retrieved at 14.02.2022 from [aux](https://github.com/carmonalab/scRNAseq_data_processing/master/aux), or already in you local directory if you cloned the repository.

```{r load_geneRef}
# load gene symbols dataset
if (species == "mouse") {
   geneRef_file <- "aux/EnsemblGenes105_Mmu_GRCm39.txt.gz"
} else if (species == "human") {
   geneRef_file <- "aux/EnsemblGenes105_Hsa_GRCh38.p13.txt.gz"
}
geneRef <- fread(geneRef_file)
```


```{r Standar_gene_Symbol}
# Standarize gene symbols using STACAS
data_seurat <- STACAS::StandardizeGeneSymbols(data_seurat, geneRef, assay = "RNA", slots = "counts")

# adapt metafeatures
# there incompatibilites with newer version of STACAS and Seurat v5
data_seurat[["RNA"]]@meta.features <- data.frame(row.names = rownames(data_seurat))
```

CHECK slots
TODO If data pre-normalized, apply inverse of NormalizeData() into "counts" slot. Then we can run NormalizeData() routinely

```{r}
head(data_seurat@meta.data)
```

### Clean cell metadata  (custom)

CHECK: cell metadata fields: 
Sample, Individual, Tissue c("blood","tumor","adjacent","LN","other"), Study, OriginalAnnotationLevel1, OriginalAnnotationLevel2, OriginalAnnotationLevel3...
Factorize or change labels if needed.

```{r clean_metadata}
data_seurat$orig.ident <- cell_ids

data_seurat$OriginalAnnotationLevel1 <- data_seurat$cells
#data_seurat$OriginalAnnotationLevel2 <- 
#data_seurat$OriginalAnnotationLevel3 <- 
head(data_seurat)
```

## 6. Add features/genes metadata
```{r add_gene_metadata}
# Remove duplicates from Gene symbols reference
features_meta_data <- geneRef[!duplicated(geneRef$`Gene name`),]
# Convert to dataframe
setDF(features_meta_data)

rownames(features_meta_data) <- features_meta_data$`Gene name`
# in Seurat, underscores ('_'), are replaced with dashes ('-')
# in gene names; 5S_rRNA; but these just a few
rownames(features_meta_data) <- gsub("_","-",rownames(features_meta_data)) 
features_meta_data <- features_meta_data[rownames(data_seurat),]

if ( all.equal(rownames(features_meta_data),
               as.character(rownames(data_seurat))) ) {
  data_seurat[["RNA"]] <- AddMetaData(data_seurat[["RNA"]],
                                      metadata = features_meta_data)
  rm(geneRef,features_meta_data)
}
```



# Export standardized datasets
```{r}
saveRDS(data_seurat,file.path(output_data_path,paste0(ds_name,".rds")))
# saveRDS(data_seurat,file.path(output_data_path,paste0(ds_name,"light.rds")))
```
