---
title: All cells processing of Bassez Breast cancer dataset
author: "J. Garnica <josep.garnicacaparros at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

***
Dataset name:   Author_Year_PBMID_cohortID (eg. ZhangY_2022_34653365)
```{r dataset_name}
ds_name <- "BassezA_2021_33958794"
```
***

Set root directory for processing
```{r set_dir}
# Set path to the directory were
# Set path to the directory were to get input and output
home.path <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets", ds_name)

knitr::opts_chunk$set(echo = T, warning = F, include=T, message=F)
```

# Prepare packages and environtment
```{r load_packages_functions, message=FALSE}
# restore packages versions with renv.lock file from github repository
renv::restore()

# load packages
pckgs <- c("tidyverse", "Seurat", "UCell", "scGate",
           "SingleCellExperiment","Matrix", "patchwork",
           "data.table", "STACAS", "SignatuR")
for(package in pckgs){
  library(package, character.only = T)
}

# load custom functions
source("Rutils/functions.R")

options(timeout=4000) # increase time out for downloading
```

Also, increase memory parameters to allocate large vectors, otherwise R might crash
```{bash create_Renviron}
#include vector max size
max_size=200Gb

# create .Renviron file if not present
if ! [ -a ".Renviron" ]
then
  touch .Renviron
  echo R_MAX_VSIZE=$max_size > .Renviron
  echo New .Renviron file created!
else
  echo .Renviron already present!
  echo Check size of the allowed vector:
  cat .Renviron
fi
```


# Set parameters
Dataset ID should have 3 letters from Author (Aaa) followed by additional uniquifying letter (A). If multiple cohort, followed by 1 letter from cohort ID separated by dot (e.g. AndA.H; AndB.M; And.C, etc.)
```{r define_parameters}
# Assign a short id for this data source
ID <- "Bassez"

# Assign species of the data
species <- "human"

#set seed
my_seed <- 321
set.seed(my_seed)

# max number of cells per sample (if higher, they will downsampled)
# make sure to set seed for this step
ds <- 1000

# Optional: min number of cells per sample, samples with lower number will be discarded
min.cells <- 200 
```


# Set paths
Declare the paths in your directory or Dropbox
```{r}
# Create directory to store input, output and plots
dir.create(home.path)

# Create directories for input, output and plots
input_data_path_root <- file.path(home.path, "data")
dir.create(file.path(input_data_path_root))
output_data_path <- file.path(home.path, "output")
dir.create(file.path(output_data_path))
output_plots_path <- file.path(home.path,"plots")
dir.create(file.path(output_plots_path))
```

# Fetch data
Raw data and metadata had been previously downloaded by Massimo/Santiago on May 2021 and saved in dropbox ("~/Dropbox/CSI/Datasets/Bassez/"). Here it is included metadata and counts.

# Load Data

## Load counts
```{r load_rds, eval=T}
# directory
dir <- "~/Dropbox/CSI/Datasets/Bassez/data"

# load counts
c1 <- readRDS(file.path(dir, "1863-counts_cells_cohort1.rds"))
c2 <- readRDS(file.path(dir, "1867-counts_cells_cohort2.rds"))

# are names repeated between cohorts?
table(colnames(c1) %in% colnames(c2))

```

## Load metadata
There are different metadata files, i.e. 1872-BIOKEY_metaData_cohort1_web.csv file contain general information of cohort 1 but only broad classification. Then, metadata for each cell type is inclued (with respective cell barcode) with finer classification by authors. This is not the case for cohort2, where only broad classification is provided.
Let's try to put togehter all the metadata.
### Cohort1
```{r}
list.meta.files <- list.files(dir, "[.]csv$") %>% 
                    # exclude clonotypes combined
                  grep("clonotypes_combined", ., invert = T, value = T)

# read coarase metadata cohort1
meta1 <- read.csv(file.path(dir, "1872-BIOKEY_metaData_cohort1_web.csv"))

# now add the finer cell type prediction of other files
ct <- list.meta.files %>% grep("cohort1", ., value = T)
fine.meta <- list()
for(a in ct){
  tmp <- read.csv(file.path(dir, a))
  # for clonotype info the first column is not named Cell, but barcode...
  if(grepl("vdj", a)){
  names(tmp)[1] <- "Cell"
  # add also suffix not included in vdj metadata
  tmp$Cell <- paste0(tmp$Cell, "-1")
  }
  # only get different name columns
  sel <- c("Cell", names(tmp)[!names(tmp) %in% names(meta1)])
  cat(a, " :  ", sel, "##############\n")

  
  if(length(sel)>1){
    fine.meta[[a]] <- tmp[,sel]
  }
}

# join metadata
# we can rbind the subtypes, but need to left_join the clonotype info
fine <- data.table::rbindlist(fine.meta[1:3], fill = T)
fine.meta <- left_join(fine, fine.meta$`1879-BIOKEY_barcodes_vdj_combined_cohort1.csv`,
                       by = "Cell")
#join with rest of metadata
meta1 <- left_join(meta1, fine.meta, by="Cell")

# how many cells are in count names?
table(meta1$Cell %in% colnames(c1))

# convert Cell to rownames
meta1 <- meta1 %>% column_to_rownames("Cell")

rm(fine.meta, fine, tmp)
gc()
```

### Cohort2
```{r}
list.meta.files %>% grep("cohort2", ., value = T)
# read coarase metadata cohort1
meta2 <- read.csv(file.path(dir, "1871-BIOKEY_metaData_cohort2_web.csv"))

# add VDJ data, we don't have fine prediction by authors for cohort2
vdj <- read.csv(file.path(dir, "1880-BIOKEY_barcodes_vdj_combined_cohort2.csv"))
names(vdj)[1] <- "Cell"
# add also suffix not included in vdj metadata
vdj$Cell <- paste0(vdj$Cell, "-1")

meta2 <- left_join(meta2, vdj, by="Cell")

table(duplicated(meta2$Cell))

# how many cells are in count names?
table(meta2$Cell %in% colnames(c2))

# convert Cell to rownames
meta2 <- meta2 %>% column_to_rownames("Cell")

rm(vdj)
gc()
```




**CHECK**:

1. Assess available assays: Are raw counts available or only pre-normalized (eg log2TMP); if pre-normalized, specify how was the normalization performed!

2. Make sure data matrix has genes as rows and cells as columns

3. Include, adapt and standarize cell metadata

4. Modify cell identifiers (`orig.ident`)

5. Standardize gene symbols

6. Add features/genes metadata


## 1. Assess available assays
Asses whether raw counts are available or only pre-normalized (eg log2TMP), if pre-normalized, specify how was the normalization performed at the entry form (`Expression data type`). If some datasets not include raw counts consider discarding them, or if normalization approach is known, de-normalize and normalize data on all datasets.
***
**Data input format**: e.g.  pre-normalized, "by total counts, as described in https://doi.org/10.1016/j.cell.2015.04.044"
***

## 2.  Make sure data matrix has genes as rows and cells as columns
```{r check_mtx}
dim(c1)
c1[1:10,1:10]
head(colnames(c1))

dim(c2)
c2[1:10,1:10]
head(colnames(c2))
```


## 3. Modify cell identifiers (`orig.ident`)
To make cell ids compatible with other datasets, append the study ID at the beginning of the original cell name. If available, also add other metadata elements such as Individual, tissue or treatment.

```{r modify_cellID}
identical(rownames(meta1), colnames(c1))

# underscore delimits end of prefix
cell_ids <- paste(ds_name, rownames(meta1), sep = '_') 
colnames(c1) <- cell_ids
rownames(meta1) <- cell_ids
c1[1:10,1:2]


identical(rownames(meta2), colnames(c2))

# underscore delimits end of prefix
cell_ids <- paste(ds_name, rownames(meta2), sep = '_') 
colnames(c2)<- cell_ids
rownames(meta2) <- cell_ids
c2[1:10,1:2]
```


## Create Seurat object (custom)
```{r Create_Seurat}
seu.list <- list()

seu.list[["cohort1"]] <- CreateSeuratObject(counts=c1,
                                            project=ds_name,
                                            meta.data = meta1,
                                            assay="RNA",
                                            min.cells = 0,
                                            min.features = 0)

seu.list[["cohort2"]] <- CreateSeuratObject(counts=c2,
                                            project=ds_name,
                                            meta.data = meta2,
                                            assay="RNA",
                                            min.cells = 0,
                                            min.features = 0)
rm(cell_ids, c1, c2, meta1, meta2)
# free up rstudio memory
gc() 
```

Merge files into Seurat object
```{r}
seu <- merge(seu.list$cohort1, seu.list$cohort2)
rm(seu.list)
gc()

# agreement of metafeatures??
table(rownames(seu[["RNA"]]@data) == rownames(seu[["RNA"]]@meta.features))
table(rownames(seu[["RNA"]]@counts) == rownames(seu[["RNA"]]@meta.features))
```

## Define samples
There are samples from different patietns (`patient_id`) but each patient have 2 samples, `pre` and `on`. Hence, there are 2 samples per patients to be separated.

```{r}
seu$sample <- paste(seu$patient_id, seu$timepoint, sep = "_")

```



## 4. Standardize gene symbols 
Map gene ids with our reference gene symbols using `STACAS::StandardizeGeneSymbols()`. Load EnsemblGenes105_Hsa_GRCh38.p13 and EnsemblGenes105_Mmu_GRCm39 retrieved at 14.02.2022 from [aux](https://github.com/carmonalab/scRNAseq_data_processing/master/aux), or already in you local directory if you cloned the repository.

```{r load_geneRef}
# load gene symbols dataset
if (species == "mouse") {
   geneRef_file <- "aux/EnsemblGenes105_Mmu_GRCm39.txt.gz"
} else if (species == "human") {
   geneRef_file <- "aux/EnsemblGenes105_Hsa_GRCh38.p13.txt.gz"
}
geneRef <- fread(geneRef_file)
```


```{r Standar_gene_Symbol}
# Standarize gene symbols using STACAS
seu <- STACAS::StandardizeGeneSymbols(seu, EnsemblGeneTable =  geneRef)


# adapt metafeatures
seu[["RNA"]]@meta.features <- data.frame(row.names = rownames(seu))

# agreement of metafeatures??
table(rownames(seu[["RNA"]]@data) == rownames(seu[["RNA"]]@meta.features))
table(rownames(seu[["RNA"]]@counts) == rownames(seu[["RNA"]]@meta.features))
```


CHECK are data pre-normalized? decide whether to Normalize

```{r norm_Data}
seu <- NormalizeData(seu) 
```



# 5. Quality control measurement and filtering

## 5.1. Ribosomal and mitochondrial content, and sequencing complexity
Use `SignatuR` package, including ribosomal and mitochondrial gene sets, to compute their relative content. Also, compute the sequencing complexity by computing the log10 ratio of genes detected per UMI.


```{r define_qc_metrics}
#Get mitochondrial and ribosomal signatures
ribo.genes <- GetSignature(SignatuR$Hs$Compartments$Ribo)
mito.genes <- GetSignature(SignatuR$Hs$Compartments$Mito)

# Compute ribosomal and mitochondrial content and add to Seurat object metadata
seu <- AddMetaData(seu, metadata = PercentageFeatureSet(seu, features = ribo.genes[ribo.genes %in% rownames(seu)]), col.name = "percent.ribo")
seu <- AddMetaData(seu, metadata = PercentageFeatureSet(seu, features = mito.genes[mito.genes %in% rownames(seu)]), col.name = "percent.mito")

# Compute the ratio of number of genes/features and number of counts/UMIs
seu$log10GenesPerUmi <- log10(seu$nFeature_RNA) / 
                                log10(seu$nCount_RNA)
```

## 2. Evaluate QC metrics and filter dataset
QC-based metric is mostly done manually by inspecting the distribution of the QC metrics. QC metrics can largely change between samples, cell types included (heterogeneity), sequencing technology, and batch. Take into account that number of reads (`nCount`) are only informative if raw counts are available, otherwise they cannot be used of QC filtering.

Here are shown some guidelines for QC-based filtering:

|--------------------------------------|---------------------------------------------------------------------|
| QC metric                            | Guidelines for filtering                                            |
|--------------------------------------|---------------------------------------------------------------------|
| Ribosomal content                    | \> 0.5% & \< 60%                                                    |
| Mitochondrial content                | \< 10-20%                                                           |
| Number of genes/features (nFeatures) | \>400 and custom upper limit based on your data                     |
| Numbrer of counts/UMI (nCount)       | \>800 and customand custom upper limit based on your data           |
| Ratio Features/Counts                | \> 0.6                                                              |
| Special genes                        | Custom: Consider remove some of these genes for downstream analysis |
|--------------------------------------|---------------------------------------------------------------------|

### CHECK: distributions across samples
#### Custom thresholds
Review after assessing the plots!

```{r set_cutoffs}
cutoffs <- list()
cutoffs[["percent.ribo"]] <- c(min=0.5,max=60)
cutoffs[["percent.mito"]] <- c(min=0,max=17)
cutoffs[["nFeature_RNA"]] <- c(min=600,max=6000) 
cutoffs[["nCount_RNA"]] <- c(min=1300,max=45000) # raw counts present??
cutoffs[["log10GenesPerUmi"]] <- 0.7
print(cutoffs)
```

#### Filter out outliers & low quality cells (custom)
```{r qc_plots}
# When using qc_plot cutoffs should be a list with the same names for 
# QC metrics as in the metadata
qclist <- lapply(names(cutoffs), # names of QC metrics to evaluate
                 qc_plot, # qc_plot function on functions.R script
                 metadata = seu@meta.data,
                 cutoffs = cutoffs,
                 fill = "sample",
                 split = F) # Seurat metadata

pdf(file.path(output_plots_path, "QC_metrics_sample.pdf"), width = 12, height = 10)
lapply(qclist, print)
dev.off()
```

```{r}
qclist <- lapply(names(cutoffs), # names of QC metrics to evaluate
                 qc_plot, # qc_plot function on functions.R script
                 metadata = seu@meta.data %>%
                   filter(grepl("myeloid", cellType, ignore.case = T)),
                 cutoffs = cutoffs,
                 fill = "sample",
                 split = T)
```


### Standard check

```{r probs}
for(va in names(cutoffs)){
  cat(va, "\n")
  print(quantile(seu@meta.data[[va]],
                 probs=c(0.001, 0.005, 0.01, 0.02,
                         0.98, 0.99, 0.995, 0.999))
        )
}
```

After having evaluated the QC metrics values and distribution, modify cutoffs if needed and then proceed to subset your data based on those. Also, evaluate how many cells are being removed based on your cutoffs.

```{r subset_seurat, eval = F}
# store initial number of cells
initialCellNum <- ncol(seu)
message(paste("Original number of cells:", initialCellNum))

# Perform subsetting using cutoffs list and qc metrics considered
seuqc <- subset(seu, subset = 
                        nFeature_RNA >= cutoffs[["nFeature_RNA"]]["min"] &
                        nFeature_RNA < cutoffs[["nFeature_RNA"]]["max"] & 
                        # raw counts present??
                        nCount_RNA   >= cutoffs[["nCount_RNA"]]["min"]   & 
                        nCount_RNA   < cutoffs[["nCount_RNA"]]["max"]   &  
                        percent.ribo >= cutoffs[["percent.ribo"]]["min"] & 
                        percent.ribo < cutoffs[["percent.ribo"]]["max"] &
                        percent.mito >= cutoffs[["percent.mito"]]["min"] &
                        percent.mito < cutoffs[["percent.mito"]]["max"] &
                        log10GenesPerUmi > cutoffs[["log10GenesPerUmi"]]
                      )

# Print and evaluate the cell drop out after filtering
l <- ncol(seuqc)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)", l, 100*l/initialCellNum))
```


# Downsample, process dataset and make diagnostic plots

At this point we should only have in our datasets high quality cells. We will remove samples with very few cells and downsample large ones, in order to have a similar contribution from all patients and samples. Downsampling large samples also speeds up downstream computations.

We will save the datasets with all high-quality cells (whole) and datasets without small samples and downsampling large ones (light).

```{r remove_small_samples}
initialCellNum <- ncol(seu)
# Check distribution of samples cell number
hist(table(seu$sample))

# Remove small samples
tab <- table(seu$sample)
keep <- names(tab)[tab > min.cells]
seu <- subset(seu, sample %in% keep)

# Print number of samples removed
cat(length(tab[tab <= min.cells]), "samples removed out of a total of ", length(tab) )

# Print and evaluate the cell drop out after this step
l <- ncol(seu)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)",
                l, 100*l/initialCellNum))
```


```{r downsample}
# Downsample

Idents(seu) <- "sample"
set.seed(my_seed)
seu.ds <- subset(seu, downsample = ds)

# Explore the number of cells per sample
table(seu.ds$sample)

# Print and evaluate the cell drop out after this step
l <- ncol(seu.ds)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)",
                l, 100*l/initialCellNum))
```




# Export standardized datasets
```{r}
saveRDS(seu.ds, file.path(output_data_path,paste0(ds_name,"_downsampled.rds")))
saveRDS(seu, file.path(output_data_path,paste0(ds_name,"whole.rds")))
```

# Consider saving as H5 (https://github.com/mojaveazure/seurat-disk) for interoperability with Bioconductor and Python; eg SaveH5Seurat() 



# Export 3 patients subset for HiTME demo
```{r}
# load whole samples (not downsampled)
whole <- readRDS(file.path(output_data_path, "BassezA_2021_33958794whole.rds"))

# selection based on low cancer cells
cc <- whole@meta.data %>% 
  # only keep sample from cohort 1
  filter(cohort == "treatment_naive") %>% 
  # only keep large samples
  group_by(sample) %>% 
  filter(n() > 1000) %>% 
  {table(.$cellType == "Cancer_cell", .$sample)} %>%
  prop.table(margin = 2) %>% 
  as.data.frame() %>% 
  filter(Var1 == "TRUE" &
           grepl("Pre", Var2)) %>% 
  arrange(Freq)
(sel <- cc[1:3, "Var2"])

# subset seurat object
sv <- whole[,whole$sample %in% sel]

# save rds
saveRDS(sv, file.path(output_data_path,paste0(ds_name,"_3patients.rds")))
```