---
title: Standardized Processing of scRNA-seq datasets
author: "M. Andreatta <massimo.andreatta at unil.ch>, S. Carmona <santiago.carmona at unil.ch>, J. Garnica <josep.garnicacaparros at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------
# Set parameters

Dataset name: Author_Year_PBMID_cohortID (eg. ZhangY_2022_34653365)

Dataset ID should have 3 letters from Author (Aaa) followed by
additional uniquifying letter (A). If multiple cohort, followed by 1
letter from cohort ID separated by dot (e.g. AndA.H; AndB.M; And.C,
etc.)

```{r dataset_name}
ds_name <- "StephensonE_2021_33879890"
ID <- "SteE"
species <- "human"
my_seed <- 123
set.seed(my_seed)

```

# Set paths
```{r}
path_nas <- file.path("", "Volumes", "Shared", "DataCollections", "Standardized_SingleCell_Datasets", ds_name)
dir.create(file.path(path_nas))
path_data <- file.path(path_nas, "data")
dir.create(file.path(path_data))
path_output <- file.path(path_nas, "output")
dir.create(file.path(path_output))
path_plots <- file.path(path_nas, "plots")
dir.create(file.path(path_plots))
```

**CHECK**: Make sure the input data and metadata are available in our
repository.


# Prepare packages and environtment

```{r load_packages_functions, message=FALSE}
# remotes::install_github("carmonalab/UCell",ref="5f12a9d")
# remotes::install_github("carmonalab/scGate",ref="17a4ce6")
# remotes::install_github("carmonalab/STACAS",ref="bd73fff")
# remotes::install_github("carmonalab/ProjecTILs",ref="cc73b97")
# remotes::install_github("carmonalab/SignatuR",ref="54bc3e8")

library(Seurat) # v5
library(limma)
# devtools::install_github("immunogenomics/presto")
#library(presto)
#library(SeuratData)
# library(SeuratDisk)

library(ggplot2)
library(UCell)
library(scGate)
library(SingleCellExperiment)
library(viridis)
library(Matrix)
library(patchwork)
library(GEOquery)
library(dplyr)
library(data.table)
library(STACAS)
library(SignatuR)

# load custom functions
source("Rutils/functions.R")
```

Also, increase memory parameters to allocate large vectors, otherwise R
might crash

```{bash create_Renviron}
#include vector max size
max_size=200Gb

# create .Renviron file if not present
if ! [ -a ".Renviron" ]
then
  touch .Renviron
  echo R_MAX_VSIZE=$max_size > .Renviron
  echo New .Renviron file created!
else
  echo .Renviron already present!
  echo Check size of the allowed vector:
  cat .Renviron
fi
```


# Download data

# Load Data

Depending on the data format at GEO or Zenodo, use different tools to
load the datasets into the environment.

For now, only using the Stephenson dataset, as its the largest one and has well-annotated metadata
```{r}

# File "haniffa21.processed.h5ad" downloaded from:
# https://www.covid19cellatlas.org/index.patient.html
# -> COVID-19 PBMC Ncl-Cambridge-UCL

Convert(file.path(path_data, "haniffa21.processed.h5ad"), dest = file.path(path_data, "COVID19_Stephenson.h5seurat"))
obj_COVID19_Stephenson <- LoadH5Seurat(file.path(path_data, "COVID19_Stephenson.h5seurat"))
saveRDS(obj_COVID19_Stephenson, file = file.path(path_data, "COVID19_Stephenson.rds"), compress = T)
obj <- readRDS(file.path(path_data, "COVID19_Stephenson.rds"))

# obj_COVID19_van_der_Wijst <- readRDS(file.path(path_data, "COVID19_van_der_Wijst.rds"))
# # convert v4 assay to v5 assay
# obj_COVID19_van_der_Wijst[["RNA"]] <- as(object = obj_COVID19_van_der_Wijst[["RNA"]], Class = "Assay5")
```

# Update Seurat object to newest version (v5)

```{r}
obj@version
obj <- CreateSeuratObject(obj@assays$raw$counts,
                          meta.data = obj@meta.data)
obj@version
```


**CHECK**:

1.  Assess available assays: Are raw counts available or only
    pre-normalized (eg log2TMP); if pre-normalized, specify how was the
    normalization performed!

2.  Make sure data matrix has genes as rows and cells as columns

3.  Include, adapt and standarize cell metadata

4.  Modify cell identifiers (`orig.ident`)

5.  Standardize gene symbols

6.  Add features/genes metadata

## 1. Assess available assays

Asses whether raw counts are available or only pre-normalized (eg
log2TMP), if pre-normalized, specify how was the normalization performed
at the entry form (`Expression data type`). If some datasets not include
raw counts consider discarding them, or if normalization approach is
known, de-normalize and normalize data on all datasets. *Data input
format**: e.g. pre-normalized, "by total counts, as described in
<https://doi.org/10.1016/j.cell.2015.04.044>"***

## 2. Make sure data matrix has genes as rows and cells as columns

# Metadata info
```{r}
head(colnames(obj))
paste("Number of genes:", dim(obj)[1])
paste("Number of cells:", dim(obj)[2])
paste("Number of samples:", length(unique(obj$sample_id)))
paste("Number of patients:", length(unique(obj$patient_id)))
paste("Sites:", unique(obj$Site))
paste("Number of cells per sex:")
paste(names(table(obj$Sex)), table(obj$Sex))
paste("Outcome labels:")
paste(unique(obj$Worst_Clinical_Status))
```

## 3. Include, adapt and standarize cell metadata

Generate a metadata dataframe including the following variables for each
cell in the dataset. If some are not present include NA. If needed adapt
metadata (factorize...).

+---------------------------+-------------------------------------+
| Variable                  | Example                             |
+===========================+=====================================+
| Original_cell_names       | AAACCTGAGGTTACCT.Pre_P007_b         |
+---------------------------+-------------------------------------+
| Barcode                   | AAACCTGAGGTTACCT                    |
+---------------------------+-------------------------------------+
| Study                     | Author_Year_PBMID_cohortID          |
|                           |                                     |
|                           | (e.g. ZhangY_2022_34653365)         |
+---------------------------+-------------------------------------+
| Sample                    | Pre_P007_b                          |
+---------------------------+-------------------------------------+
| Treatment                 | Pre                                 |
+---------------------------+-------------------------------------+
| Individual                | P007                                |
+---------------------------+-------------------------------------+
| Tissue                    | blood, tumor, adjacent, LN, other   |
+---------------------------+-------------------------------------+
| O                         | CD4T                                |
| riginalAnnotationLevel{n} |                                     |
+---------------------------+-------------------------------------+


```{r adapt_meta}
# Some barcodes contain the site as a prefix, e.g. site_ATGCATGC... or some other unknown identifier
# IMPORTANT: needs to be left in, otherwise duplicate names
# Maybe some samples were sequenced multiple times?

# Explore data and modifiy this code accordingly to fill the metadata slots
cell_names <- colnames(obj)
obj@meta.data$Original_cell_names <- cell_names
cell_names[grepl("_", cell_names)] <- sapply(strsplit(cell_names[grepl("_", cell_names)], "_"), function(x) x[[2]])
cell_names.split <- strsplit(cell_names, "-")
obj@meta.data$Barcode <- sapply(cell_names.split, function(x) x[[1]])

messy_barcodes <- grepl("_", obj@meta.data$Barcode)
obj@meta.data$Barcode[messy_barcodes] <- sapply(strsplit(obj@meta.data$Barcode[messy_barcodes], "_"), function(x) x[[2]])

obj@meta.data$Study <- ds_name
obj@meta.data$Sample <- obj$sample_id
obj@meta.data$Sample <- gsub("-", ".", obj@meta.data$Sample)
obj@meta.data$Individual <- obj$patient_id
obj@meta.data$Tissue <- "blood"
rm(cell_names, cell_names.split)
gc()
```

Clarify metadata "Status" and "Swab_results"
```{r}
table(obj$Status, obj$Swab_result)

unique(obj$sample_id[obj$Status == "Covid" & obj$Swab_result == "Negative"])
```


## 4. Modify cell identifiers (`orig.ident`)

To make cell ids compatible with other datasets, append the study ID at
the beginning of the original cell name. If available, also add other
metadata elements such as Individual, tissue or treatment.

```{r modify_cellID}
#fields in prefix are separated by dots
prefix <- paste(obj$Sample, "blood", obj$Sex, obj$Status_on_day_collection_summary, obj$Site, sep=".")  #fields in prefix are separated by dots
cell_ids <- paste(prefix, gsub("-", ".", obj$Original_cell_names), sep = '_') #underscore delimits end of prefix
colnames(obj) <- cell_ids
obj[1:10,1:2]
head(colnames(obj))
rm(prefix)
gc()
```

## 5. Standardize gene symbols

Map gene ids with our reference gene symbols using
`STACAS::StandardizeGeneSymbols()`. Load EnsemblGenes105_Hsa_GRCh38.p13
and EnsemblGenes105_Mmu_GRCm39 retrieved at 14.02.2022 from
[aux](https://github.com/carmonalab/scRNAseq_data_processing/master/aux),
or already in you local directory if you cloned the repository.

```{r load_geneRef}
# load gene symbols dataset
if (species == "mouse") {
   geneRef_file <- "aux/EnsemblGenes105_Mmu_GRCm39.txt.gz"
} else if (species == "human") {
   geneRef_file <- "aux/EnsemblGenes105_Hsa_GRCh38.p13.txt.gz"
}
geneRef <- fread(geneRef_file)
```

```{r Standar_gene_Symbol}
# Standarize gene symbols using STACAS
obj <- STACAS::StandardizeGeneSymbols(obj, geneRef, assay = "RNA", slots = "counts")
```


### Clean cell metadata (custom)

CHECK: cell metadata fields: Sample, Individual, Tissue
c("blood","tumor","adjacent","LN","other"), Study,
OriginalAnnotationLevel1, OriginalAnnotationLevel2,
OriginalAnnotationLevel3... Factorize or change labels if needed.

```{r clean_metadata}
obj$orig.ident <- cell_ids
obj$OriginalAnnotationLevel1 <- obj$initial_clustering
obj$OriginalAnnotationLevel2 <- obj$full_clustering
```

How many cells per sample?
```{r}
sample.size.pre <- sort(table(obj$Sample))
sample.size.pre
barplot(sort(sample.size.pre), las=2)
```

# Downsample, process dataset and make diagnostic plots

At this point we should only have in our datasets high quality cells. We
will remove samples with very few cells and downsample large ones, in
order to have a similar contribution from all patients and samples.
Downsampling large samples also speeds up downstream computations.

We will save the datasets with all high-quality cells (whole) and
datasets without small samples and downsampling large ones (light).

```{r remove_small_samples}
min.cells.perSample <- 500
# max.cells.perSample <- 10000
cells.perSample <- table(obj$Sample)
barplot(sort(cells.perSample))
abline(h=min.cells.perSample)
# abline(h=max.cells.perSample)

# Remove small samples
tab <- table(obj$Sample)
keep <- names(tab)[tab > min.cells.perSample]
obj <- subset(obj, Sample %in% keep)

# Print number of samples removed
cat(length(tab[tab <= min.cells.perSample]), "samples removed out of a total of ", length(tab) )

# # Print and evaluate the cell drop out after this step
# l <- ncol(obj)
# message(sprintf("Number of cells after removing low cell count samples: %i (%.2f %% of input)",
#                 l, 100*l/sum(sample.size.pre)))
```


# Export standardized datasets

```{r}
saveRDS(obj,file.path(path_output, paste0(ds_name, "_Stephenson_preprocessed.rds")))
```
