---
title: Joanito et al., 2022 processing (Nature genetics)
author: "M. Andreatta <massimo.andreatta at unil.ch>, S. Carmona <santiago.carmona at unil.ch>, J. Garnica <josep.garnicacaparros at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

***
Dataset name:   Author_Year_PBMID_cohortID (eg. ZhangY_2022_34653365)
```{r dataset_name}
ds_name <- "JoaI_2022_35773407"
```
***

Set root directory for processing
```{r set_dir}
# Set path to the directory were
# Set path to the directory were to get input and output
home.path <- file.path("~/Dropbox/CSI/Standardized_SingleCell_Datasets", ds_name)

```


# Prepare packages and environtment
```{r load_packages_functions, message=FALSE}
# restore packages versions with renv.lock file from github repository
renv::restore()

# load packages
pckgs <- c("tidyverse", "Seurat", "rhdf5",
           "Matrix", "patchwork",
           "data.table")
for(package in pckgs){
  library(package, character.only = T)
}

# load custom functions
source("Rutils/functions.R")

options(timeout=4000) # increase time out for downloading
```

Also, increase memory parameters to allocate large vectors, otherwise R might crash
```{bash create_Renviron}
#include vector max size
max_size=200Gb

# create .Renviron file if not present
if ! [ -a ".Renviron" ]
then
  touch .Renviron
  echo R_MAX_VSIZE=$max_size > .Renviron
  echo New .Renviron file created!
else
  echo .Renviron already present!
  echo Check size of the allowed vector:
  cat .Renviron
fi
```


# Set parameters
Dataset ID should have 3 letters from Author (Aaa) followed by additional uniquifying letter (A). If multiple cohort, followed by 1 letter from cohort ID separated by dot (e.g. AndA.H; AndB.M; And.C, etc.)
```{r define_parameters}
# Assign a short id for this data source
ID <- "JoaI"

# Assign species of the data
species <- "human"

#set seed
my_seed <- 321
set.seed(my_seed)

# max number of cells per sample (if higher, they will downsampled)
# make sure to set seed for this step
ds <- 1000

# Optional: min number of cells per sample, samples with lower number will be discarded
min.cells <- 200 
```


# Set paths
Declare the paths in your directory or Dropbox
```{r set_directories}
# Create directory to store input, output and plots
dir.create(home.path)

# Create directories for input, output and plots
input_data_path_root <- file.path(home.path, "data")
dir.create(file.path(input_data_path_root))
output_data_path <- file.path(home.path, "output")
dir.create(file.path(output_data_path))
output_plots_path <- file.path(home.path,"plots")
dir.create(file.path(output_plots_path))
```

# Download data
If not done before, download data 

The processed data can be found at https://www.synapse.org/#!Synapse:syn26844071/wiki/615389

## Download from synapse 



# Load Data

HDF5 files
```{r load_rds}
list.h5 <- list.files(input_data_path_root,
                       "matrix[.]h5$",
                       recursive = T,
                       full.names = T)
names.h5 <- list.files(input_data_path_root,
                       "matrix[.]h5$",
                       recursive = T,
                       full.names = F) %>% 
              gsub("_Count_matrix[.]h5", "", .) %>% 
              strsplit(., "/", fixed = T) %>% 
              sapply("[", 2 )
list.meta <- list.files(input_data_path_root,
                       "metadata[.]csv$",
                       recursive = T,
                       full.names = T)


clin.meta <- read.csv(file.path(input_data_path_root, "patient_clinical_information.csv"))
```

Load data in a list of 2 for epithelial and non-epithelial cells
```{r, eval = F}
list.seurat <- list()
for(r in seq_along(list.h5)){
  mat <- Read10X_h5(list.h5[[r]])
  dim(mat)
  meta <- read.csv(list.meta[r]) %>% 
    # join with clinical metadata
        left_join(., clin.meta[,-1],
                  by = "patient.ID") %>% 
    column_to_rownames("cell.ID") 
  dim(meta)
  
  if(identical(rownames(meta), colnames(mat))){
    list.seurat[[names.h5[r]]] <- CreateSeuratObject(counts = mat,
                                      project = names.h5[r],
                                      assay = "RNA",
                                      meta.data = meta,
                                      min.cells = 0,
                                      min.features = 0)
  } else {
    message("Discrepancies between count matrix and metadata")
  }
}
```

## Load matrix with all genes

```{r}
nofilt.h5 <- list.files(input_data_path_root,
                       "Filtered[.]h5$",
                       recursive = T,
                       full.names = T)

mat.full <- Read10X_h5(nofilt.h5)
```

## Load metadata 
Load both metadata files and rbind
```{r}
meta.list <- list()
for(r in seq_along(list.meta)){
  meta.list[[names.h5[r]]] <- read.csv(list.meta[r]) %>% 
                          # join with clinical metadata
                              left_join(., clin.meta[,-1],
                                        by = "patient.ID")
}

# 2 columns in epithelial metadata is not present in nonepithelial
setdiff(names(meta.list$Epithelial),
        names(meta.list$NonEpithelial))
# let's add them as NA

meta <- data.table::rbindlist(meta.list,
                              use.names = T, fill = T) %>% 
          column_to_rownames("cell.ID")
rm(meta.list)
```

## Creat seurat object
```{r}
if(identical(sort(rownames(meta)), sort(colnames(mat.full)))){
  seurat <- CreateSeuratObject(counts = mat.full,
                                    project = ds_name,
                                    assay = "RNA",
                                    meta.data = meta,
                                    min.cells = 0,
                                    min.features = 0)
  rm(mat.full)
  rm(meta)
} else {
  message("Discrepancies between count matrix and metadata")
}

dim(seurat)
```
Check mitochondrial and ribosomal genes
```{r}
grep("^MT-", rownames(seurat), value = T)
grep("^RP[LS]", rownames(seurat), value = T)
```



**CHECK**:

1. Assess available assays: Are raw counts available or only pre-normalized (eg log2TMP); if pre-normalized, specify how was the normalization performed!

2. Make sure data matrix has genes as rows and cells as columns

3. Include, adapt and standarize cell metadata

4. Modify cell identifiers (`orig.ident`)

5. Standardize gene symbols

6. Add features/genes metadata


## 1. Assess available assays
Asses whether raw counts are available or only pre-normalized (eg log2TMP), if pre-normalized, specify how was the normalization performed at the entry form (`Expression data type`). If some datasets not include raw counts consider discarding them, or if normalization approach is known, de-normalize and normalize data on all datasets.
***
**Data input format**: e.g.  pre-normalized, "by total counts, as described in https://doi.org/10.1016/j.cell.2015.04.044"
***


Both CD4 and CD8 datasets contain count matrix!



## 2.  Make sure data matrix has genes as rows and cells as columns
```{r check_mtx, eval = F}
dim(seurat@assays$RNA@counts)
seurat@assays$RNA@counts[1:10,1:10]
head(colnames(seurat@assays$RNA@counts))
```

## 3. Include, adapt and standarize cell metadata

Generate a metadata dataframe including the following variables for each cell in the dataset. If some are not present include NA. If needed adapt metadata (factorize...).

+----------------------------+--------------------------------------+
| Variable                   | Example                              |
+============================+======================================+
| Original_cell_names        | AAACCTGAGGTTACCT.Pre_P007_b          |
+----------------------------+--------------------------------------+
| Barcode                    | AAACCTGAGGTTACCT                     |
+----------------------------+--------------------------------------+
| Study                      | Author_Year_PBMID_cohortID           |
|                            |                                      |
|                            | (e.g. ZhangY_2022_34653365)          |
+----------------------------+--------------------------------------+
| Sample                     | Pre_P007_b                           |
+----------------------------+--------------------------------------+
| Treatment                  | Pre                                  |
+----------------------------+--------------------------------------+
| Individual                 | P007                                 |
+----------------------------+--------------------------------------+
| Tissue                     | blood, tumor, adjacent, LN, other    |
+----------------------------+--------------------------------------+
| OriginalAnnotationLevel{n} | CD4T                                 |
+----------------------------+--------------------------------------+

```{r inspect_metadata}
seurat@meta.data

```
Adapt metada and modify cell identifiers
As all cells are labeled with cell[Number] to define unequivocally each cell, we will just add as a prefix the name of the dataset, and the rest of metadata.

```{r adapt_meta}
# Explore data and modifiy this code accordinly to fill the metadata slots
meta <- seurat@meta.data %>% 
            mutate(new.names = paste(ID,
                                rownames(.),
                                patient.ID,
                                sample.origin,
                                dataset,
                                sep = ".")
                   )
          
  # rename cells accordingly
seurat <- RenameCells(seurat, new.names = meta$new.names)

rm(meta)
gc()

```




## 4. MERGE DATA
At this point we won't merge the data.




## 5. Standardize gene symbols 
Map gene ids with our reference gene symbols using `STACAS::StandardizeGeneSymbols()`. Load EnsemblGenes105_Hsa_GRCh38.p13 and EnsemblGenes105_Mmu_GRCm39 retrieved at 14.02.2022 from [aux](https://github.com/carmonalab/scRNAseq_data_processing/master/aux), or already in you local directory if you cloned the repository.

```{r load_geneRef}
# load gene symbols dataset
if (species == "mouse") {
   geneRef_file <- "aux/EnsemblGenes105_Mmu_GRCm39.txt.gz"
} else if (species == "human") {
   geneRef_file <- "aux/EnsemblGenes105_Hsa_GRCh38.p13.txt.gz"
}
geneRef <- fread(geneRef_file)

#change names with _ to -, as only accepted by seurat object features
geneRef <- geneRef %>% mutate(`Gene name` = gsub("_", "-", `Gene name`))
```


```{r Standar_gene_Symbol}
# Standarize gene symbols using STACAS
# store in a list genes exluded in CD4 and CD8
diff <- list()

# split by dataset, to avoid get rid of synonims 
li <- SplitObject(seurat, split.by = "dataset")
sli <- list()
diff <- list()
for(a in names(li)){
  sli[[a]] <- STACAS::StandardizeGeneSymbols(li[[a]],
                                            geneRef)
  # different genes
  diff[[a]] <- setdiff(rownames(li[[a]]), rownames(sli[[a]]))
}

seurat <- scCustomize::Merge_Seurat_List(sli)
  
rm(li)
rm(sli)
gc()
```




CHECK slots
TODO If data pre-normalized, apply inverse of NormalizeData() into "counts" slot. Then we can run NormalizeData() routinely

```{r, eval = F}
head(seurat)
```



## 6. Add features/genes metadata
```{r add_gene_metadata}
# Remove duplicates from Gene symbols reference
features_meta_data <- geneRef[!duplicated(geneRef$`Gene name`)]
# Convert to dataframe
setDF(features_meta_data)

rownames(features_meta_data) <- features_meta_data$`Gene name`

# in Seurat, underscores ('_'), are replaced with dashes ('-')
# in gene names; 5S_rRNA; but these just a few

rownames(features_meta_data) <- gsub("_","-",rownames(features_meta_data)) 

# add genes metadata
features_meta_data <- features_meta_data[rownames(seurat),]

rownames(features_meta_data)[!rownames(features_meta_data) %in% rownames(seurat)]
rownames(seurat)[!rownames(seurat) %in% rownames(features_meta_data)]

if ( all.equal(rownames(features_meta_data),rownames(seurat)) ) {
  seurat[["RNA"]] <- AddMetaData(seurat[["RNA"]],
                                      metadata = features_meta_data)
}

rm(geneRef,features_meta_data)
```



## 7. Load and store patient metadata
This data is already in the metadata.



## NORMALIZE DATA 
CHECK are data pre-normalized? decide whether to Normalize

In this case we have both counts (unnormalized) and normalized data (data slot). 

We will normalize the count slot using `NormalizeData`.

```{r norm_Data}
seurat <- NormalizeData(seurat) 

```


# Quality control measurement and filtering

## 1. Ribosomal and mitochondrial content, and sequencing complexity
Use `SignatuR` package, including ribosomal and mitochondrial gene sets, to compute their relative content. Also, compute the sequencing complexity by computing the log10 ratio of genes detected per UMI.


```{r define_qc_metrics}
require(SignatuR)
#Get mitochondrial and ribosomal signatures
ribo.genes <- GetSignature(SignatuR$Hs$Compartments$Ribo)
mito.genes <- GetSignature(SignatuR$Hs$Compartments$Mito)



# Compute ribosomal and mitochondrial content and add to Seurat object metadata
seurat <- AddMetaData(seurat, metadata = PercentageFeatureSet(seurat,
                                  features = ribo.genes[ribo.genes %in% rownames(seurat)]),
                      col.name = "percent.ribo")
seurat <- AddMetaData(seurat,
                      metadata = PercentageFeatureSet(seurat,
                                  features = mito.genes[mito.genes %in% rownames(seurat)]),
                      col.name = "percent.mito")

# Compute the ratio of number of genes/features and number of counts/UMIs
seurat$log10GenesPerUmi <- log10(seurat$nFeature_RNA) / 
                                log10(seurat$nCount_RNA)

```

## 2. Evaluate QC metrics and filter dataset
QC-based metric is mostly done manually by inspecting the distribution of the QC metrics. QC metrics can largely change between samples, cell types included (heterogeneity), sequencing technology, and batch. Take into account that number of reads (`nCount`) are only informative if raw counts are available, otherwise they cannot be used of QC filtering.

Here are shown some guidelines for QC-based filtering:

|--------------------------------------|---------------------------------------------------------------------|
| QC metric                            | Guidelines for filtering                                            |
|--------------------------------------|---------------------------------------------------------------------|
| Ribosomal content                    | \> 0.5% & \< 60%                                                    |
| Mitochondrial content                | \< 10-20%                                                           |
| Number of genes/features (nFeatures) | \>400 and custom upper limit based on your data                     |
| Numbrer of counts/UMI (nCount)       | \>800 and customand custom upper limit based on your data           |
| Ratio Features/Counts                | \> 0.6                                                              |
| Special genes                        | Custom: Consider remove some of these genes for downstream analysis |
|--------------------------------------|---------------------------------------------------------------------|

### CHECK: distributions across samples
```{r qc_plots0, fig.height= 10}

Idents(seurat) <- "dataset"
print(VlnPlot(seurat,
              features = c("nFeature_RNA", "nCount_RNA",
                           "percent.ribo","percent.mito",
                           "PTPRC","MALAT1"),
                ncol = 3, pt.size=0)
    )

```

### Custom thresholds
Review after assessing the plots!

```{r set_cutoffs}
cutoffs <- list()
cutoffs[["percent.ribo"]] <- c(min=0.5,max=60)
cutoffs[["percent.mito"]] <- c(min=0,max=20)
cutoffs[["nFeature_RNA"]] <- c(min=400,max=6000) 
cutoffs[["nCount_RNA"]] <- c(min=800,max=40000) # raw counts present??
cutoffs[["log10GenesPerUmi"]] <- 0.6
print(cutoffs)
```

### Filter out outliers & low quality cells (custom)
```{r qc_plots}

# When using qc_plot cutoffs should be a list with the same names for 
# QC metrics as in the metadata
qclist <- lapply(names(cutoffs), # names of QC metrics to evaluate
                 qc_plot, # qc_plot function on functions.R script
                 metadata = seurat@meta.data,
                 cutoffs = cutoffs,
                 fill = "cell.type", 
                 show.legend =T) # seurat metadata

# Join all plots into a single one
require(ggpubr)
qclistgg <- ggarrange(plotlist = qclist,
                      common.legend = T,
                      legend = "bottom")


qclistgg
```

```{r}
var <- "percent.mito"
fill <- "cell.type"
nc <- seurat@meta.data %>%
    ggplot(aes(.data[[var]])) +
    {if(is.null(fill)){
      geom_density(fill = "grey30",
                   color = "grey10")
    } else {
      geom_density(aes(fill = .data[[fill]]),
                   color = "grey10",
                   alpha = 0.5,
                   show.legend = F)
    }}+
    {if(max(seurat@meta.data[[var]])>100){
          scale_x_log10()}} +
    {if(!all(is.na(cutoffs[[var]]))){
      geom_vline(xintercept = cutoffs[[var]],
                 color = "red",
                 linetype="dashed")}} +
    facet_wrap(~cell.type, ncol = 4) +
    theme_bw() +
    theme(aspect.ratio = 1)
```


### Standard check

```{r probs, eval = F}
for(va in names(cutoffs)){
  cat(va, "\n")
  print(quantile(seurat@meta.data[[va]],
                 probs=c(0.001, 0.005, 0.01, 0.02,
                         0.98, 0.99, 0.995, 0.999))
        )
}
```

After having evaluated the QC metrics values and distribution, modify cutoffs if needed and then proceed to subset your data based on those. Also, evaluate how many cells are being removed based on your cutoffs.

IMPORTANT
I decided to not filter out cells based on QC metrics. 
1) QC metrics seem acceptable
2) in order to better compare with original paper annotation

*in this case we won't filter out cells, we will keep them all, trusting authors QC metrics*


```{r subset_seurat, fig.height= 10, eval = F}
# store initial number of cells
initialCellNum <- ncol(seurat)
message(paste("Original number of cells:", initialCellNum))

# Perform subsetting using cutoffs list and qc metrics considered
seurat <- subset(seurat, subset = 
                        nFeature_RNA >= cutoffs[["nFeature_RNA"]]["min"] &
                        nFeature_RNA < cutoffs[["nFeature_RNA"]]["max"] & 
                        # raw counts present??
                        nCount_RNA   >= cutoffs[["nCount_RNA"]]["min"]   & 
                        nCount_RNA   < cutoffs[["nCount_RNA"]]["max"]   &  
                        percent.ribo >= cutoffs[["percent.ribo"]]["min"] & 
                        percent.ribo < cutoffs[["percent.ribo"]]["max"] &
                        percent.mito >= cutoffs[["percent.mito"]]["min"] &
                        percent.mito < cutoffs[["percent.mito"]]["max"] &
                        log10GenesPerUmi > cutoffs[["log10GenesPerUmi"]]
                      )

# Print and evaluate the cell drop out after filtering
l <- ncol(seurat)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)", l, 100*l/initialCellNum))
```


# Downsample, process dataset and make diagnostic plots

We will save the datasets with all cells (whole) and datasets without small samples and downsampling large ones (light).

## Small samples
```{r}
col.meta <- c("sample.ID")
pl.list <- list()

for(a in col.meta){
  pl <- seurat@meta.data %>% 
    ggplot(aes( y = .data[[a]], fill = .data[[a]])) +
    geom_bar(show.legend = F) +
    geom_vline(xintercept = 200,
               color = "black",
               linetype="dashed") +
    scale_x_log10() +
    labs(x = "Number of cells pre QC filtering") +
    ggprism::theme_prism()
  pl.list[[a]] <- pl
}

wrap_plots(pl.list, ncol = 1)

```
We will remove sample below `r min.cells` (200)

```{r remove_small_samples, eval = F}
# Check distribution of samples cell number
hist(table(seurat$sample.ID))

# Remove small samples
tab <- table(seurat$sample.ID)
keep <- names(tab)[tab > min.cells]
seurat.ds <- subset(seurat, sample.ID %in% keep)

# Print number of samples removed
cat(length(tab[tab <= min.cells]), "samples removed out of a total of ", length(tab) )

# Print and evaluate the cell drop out after this step
l <- ncol(seurat.ds)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)",
                l, 100*l/ncol(seurat)))
```

We keep most of the cells: Number of cells after QC: 372200 (99.77 % of input)
Removed a total of 8 samples out of 189.

## Downsample data
We will downsample based on sample.id

```{r downsample, eval = T}
(initial <- ncol(seurat.ds))
# Downsample
Idents(seurat.ds) <- "sample.ID"
# Set seed for downsampling
set.seed(my_seed)
seurat.ds <- subset(seurat.ds,
                       downsample = ds)

l <- ncol(seurat.ds)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)",
                l, 100*l/initial))
```

We kept: Number of cells after QC: 166586 (44.76 % of input)



# Export standardized datasets
At this point we save the file without QC filtering, removing small samples nor downsampling. 
```{r}
saveRDS(seurat,
        file.path(output_data_path,paste0(ds_name,"_Nofilt_whole.rds")))

saveRDS(seurat.ds,
        file.path(output_data_path,paste0(ds_name,"_Nofilt_Downsampled_light.rds")))
```

# Consider saving as H5 (https://github.com/mojaveazure/seurat-disk) for interoperability with Bioconductor and Python; eg SaveH5Seurat() 

## Save H5 file
SaveH5Seurat save the whole seurat object, including metadata. For tools such as CellBender we require only the matrix count to be saved in a .h5 file, not a .h5Seurat.

```{r, eval = F}
for(e in names(list.seurat)){
filename <- paste0(ds_name,"_Nofilt_Downsampled")
write_dgCMatrix_h5(GetAssayData(object = list.ds[[e]], slot = "counts"),
                    cols_are = "sample_names",
                    h5_target = paste0(filename, ".h5"),
                    ref_name = ds_name)

filename <- paste0(ds_name,"_Nofilt_whole")
write_dgCMatrix_h5(GetAssayData(object = list.seurat[[e]], slot = "counts"),
                    cols_are = "sample_names",
                    h5_target = paste0(filename, ".h5"),
                    ref_name = ds_name)

}
```
