---
title: "Uncorrected batch-effect analysis"
author: "Christian Halter"
date: "`r format(Sys.Date(), '%e de %B, %Y')`"
output:
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
---

# Scope

This exploratory pass compares the twelve canonical batch-effect cohorts using
only the persisted `batch_effect_uncorrected` distance artifacts. It does not
rerun preprocessing, recompute embeddings, recalculate CLR distances, read the
legacy batch notebook inputs, or inspect corrected artifacts. Biological labels
are evaluation metadata only.

```{r setup, message=FALSE, warning=FALSE}
if (basename(getwd()) == "notebooks") {
  setwd("..")
  if (file.exists(".Rprofile")) source(".Rprofile")
}

suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(rmarkdown)
})
source("src/utils/load_all_functions.R")
source("src/utils/batch_effect_analysis.R")

set.seed(42)
n_permutations <- 999L
permanova_permutations <- 99L

# vegan::anosim()/adonis2() use forked permutation workers on macOS. Keep the
# default bounded to eight physical cores; lower this value for interactive
# work or constrained machines.
detected_cores <- parallel::detectCores(logical = FALSE)
if (is.na(detected_cores)) detected_cores <- 1L
parallel_cores <- min(8L, max(1L, detected_cores))
options(mc.cores = parallel_cores)
input_root <- file.path(getwd(), "data", "batch_effect", "uncorrected")
metadata_root <- file.path(input_root, "metadata")
plot_root <- file.path(getwd(), "plots", "batch_effect_uncorrected")
analysis_output_root <- file.path(input_root, "analysis")
dir.create(plot_root, recursive = TRUE, showWarnings = FALSE)
dir.create(analysis_output_root, recursive = TRUE, showWarnings = FALSE)

batch_datasets <- c(
  "Alzheimer",
  "Breast_cancer",
  "Covid19_PBMC",
  "Kidney_KPMP",
  "Myocardial_infarction",
  "Diabetes",
  "Lupus_PBMC",
  "Lung",
  "Parkinson",
  "Joanito",
  "Stephenson",
  "CombinedPBMC"
)

datasets <- read_datasets_json(view = "batch_effect_uncorrected")
if (!identical(names(datasets)[match(batch_datasets, names(datasets))], batch_datasets)) {
  stop("datasets.json does not contain the exact twelve canonical batch-effect datasets")
}
if (any(vapply(datasets[batch_datasets], function(x) is.null(x$label_col), logical(1)))) {
  stop("one or more canonical batch-effect datasets lack a primary biological label")
}
registry <- batch_candidate_registry(datasets, batch_datasets)
method_specs <- batch_uncorrected_method_specs()
selected_methods <- names(method_specs)
method_labels <- unname(method_label_map_main[selected_methods])
if (anyNA(method_labels) || length(unique(method_labels)) != length(selected_methods)) {
  stop("selected methods are not covered by method_label_map_main")
}
method_colors <- stats::setNames(
  grDevices::hcl(seq(15, 375, length.out = length(selected_methods) + 1L)[-1L], l = 65, c = 100),
  method_labels
)


# The preflight is intentionally before any figure work. Every sidecar, summary
# bundle, distance matrix, sample ID, and label must validate or the pass stops.
dataset_results <- lapply(batch_datasets, function(dataset) {
  load_batch_uncorrected_dataset(input_root, metadata_root, dataset, registry)
})
names(dataset_results) <- batch_datasets
stopifnot(identical(names(dataset_results), batch_datasets))
```

# Per-dataset sample and covariate contract

```{r metadata-contract, message=FALSE, warning=FALSE}
metadata_summary <- do.call(rbind, lapply(dataset_results, function(dataset_result) {
  dataset <- dataset_result$dataset
  ds_registry <- dataset_result$registry
  data.frame(
    dataset = dataset,
    sample_count = length(dataset_result$sample_ids),
    primary_label = dataset_result$label_col,
    candidate_count = nrow(ds_registry),
    biological_secondary_count = sum(ds_registry$is_secondary_biology),
    technical_count = sum(ds_registry$is_technical),
    stringsAsFactors = FALSE
  )
}))
metadata_summary
```

# Plot and metric helpers

```{r plotting-helpers, message=FALSE, warning=FALSE}
`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0L || all(is.na(x))) y else x
}

.metric_warning <- function(...) {
  pieces <- as.character(c(...))
  pieces <- pieces[!is.na(pieces) & nzchar(pieces)]
  paste(unique(pieces), collapse = "; ")
}

.metric_plot <- function(data, value_column, status_column, title, y_label, exclude_primary = FALSE) {
  if (exclude_primary) data <- data[!data$is_primary, , drop = FALSE]
  data <- data[data[[status_column]] == "ESTIMABLE" & is.finite(data[[value_column]]), , drop = FALSE]
  if (nrow(data) == 0L) {
    return(ggplot() + theme_void() + labs(title = paste0(title, "\nNo estimable candidate")))
  }
  data$method_label <- factor(data$method_label, levels = method_labels)
  data$candidate <- factor(data$candidate, levels = unique(data$candidate))
  data$candidate_class <- factor(
    data$candidate_class,
    levels = c(
      "biological_primary",
      "biological_secondary",
      "technical",
      "shared_biological_technical"
    )
  )
  ggplot(data, aes(x = candidate, y = .data[[value_column]], fill = method_label)) +
    geom_col(position = position_dodge(width = 0.82), width = 0.76) +
    facet_wrap(~candidate_class, scales = "free_x", nrow = 1, drop = TRUE) +
    scale_fill_manual(values = method_colors, drop = FALSE) +
    labs(title = title, x = NULL, y = y_label, fill = "Method") +
    theme_classic(base_size = 9) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      legend.position = "bottom",
      strip.background = element_blank(),
      strip.text = element_text(face = "bold")
    )
}

.blank_plot <- function(title, subtitle = NULL) {
  ggplot() + theme_void() + labs(title = title, subtitle = subtitle)
}

.save_plot <- function(plot, path, width = 12, height = 7) {
  ggplot2::ggsave(path, plot = plot, width = width, height = height, units = "in", device = "pdf")
  if (!file.exists(path) || file.info(path)$size <= 0) stop("empty plot output: ", path)
  invisible(path)
}


build_batch_funky_heatmap <- function(metric_data, output_path) {
  metric_order <- c("ANOSIM", "Modularity", "ARI")
  clean_data <- metric_data %>%
    filter(score %in% metric_order, is.finite(value)) %>%
    mutate(score = factor(score, levels = metric_order)) %>%
    group_by(dataset, score) %>%
    mutate(value = min_max(value)) %>%
    ungroup()
  if (nrow(clean_data) == 0L) stop("batch funky heatmap has no estimable metrics")

  df_overall <- clean_data %>%
    group_by(method) %>%
    summarise(Overall_Score = mean(value, na.rm = TRUE), .groups = "drop")
  df_datasets_wide <- clean_data %>%
    group_by(method, dataset) %>%
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(id_cols = method, names_from = dataset, values_from = value, names_prefix = "Dataset__")
  df_metrics_wide <- clean_data %>%
    group_by(method, score) %>%
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(id_cols = method, names_from = score, values_from = value, names_prefix = "Metric__")
  final_data <- df_overall %>%
    left_join(df_datasets_wide, by = "method") %>%
    left_join(df_metrics_wide, by = "method") %>%
    mutate(id = method) %>%
    arrange(desc(Overall_Score))

  info_method <- tibble::tribble(
    ~id, ~group, ~name, ~geom, ~palette, ~options,
    "id", "", "", "text", NA, list(hjust = 0, width = 6)
  )
  info_overall <- tibble::tribble(
    ~id, ~group, ~name, ~geom, ~palette, ~options,
    "Overall_Score", "Overall", "Mean", "bar", "palette_overall", list(width = 4)
  )
  dataset_cols <- colnames(final_data)[startsWith(colnames(final_data), "Dataset__")]
  info_datasets <- tibble(id = dataset_cols) %>%
    mutate(
      group = "Datasets",
      name = stringr::str_remove(id, "Dataset__"),
      geom = "circle",
      palette = "Datasets",
      options = list(list())
    ) %>%
    arrange(name)
  metric_cols <- colnames(final_data)[startsWith(colnames(final_data), "Metric__")]
  metric_order <- c("ANOSIM", "Modularity", "ARI")
  info_metrics <- tibble(id = metric_cols) %>%
    mutate(
      group = "Metrics",
      name = stringr::str_remove(id, "Metric__"),
      name_f = factor(name, levels = metric_order),
      geom = "bar",
      palette = "Metrics",
      options = list(list(width = 2))
    ) %>%
    arrange(name_f) %>%
    select(-name_f)
  column_info <- bind_rows(info_method, info_overall, info_datasets, info_metrics)
  ncolors <- 11
  p <- funky_heatmap(
    data = final_data,
    column_info = column_info,
    scale_column = FALSE,
    palettes = list(
      palette_overall = RColorBrewer::brewer.pal(ncolors, "RdBu"),
      Datasets = RColorBrewer::brewer.pal(ncolors, "RdBu"),
      Metrics = RColorBrewer::brewer.pal(ncolors, "RdBu")
    ),
    column_groups = tibble(group = c("Overall", "Datasets", "Metrics")),
    legends = list(
      list(palette = "palette_overall", enabled = TRUE, title = "Mean normalized score", labels = c(". Min", rep("", ncolors - 2), "Max .")),
      list(palette = "Datasets", enabled = TRUE, title = "Dataset score", labels = c(". Min", rep("", ncolors - 2), "Max .")),
      list(palette = "Metrics", enabled = TRUE, title = "Metric score", labels = c(". Min", rep("", ncolors - 2), "Max ."))
    ),
    position_args = position_arguments(col_width = 1.5, col_space = 0.2, col_annot_offset = 5)
  )
  ggsave(output_path, p, width = 12, height = 7)
  if (!file.exists(output_path) || file.info(output_path)$size <= 0) {
    stop("empty batch funky heatmap output: ", output_path)
  }
  invisible(p)
}
```

# MDS, metrics, joint decomposition, and NMI

```{r batch-analysis, message=FALSE, warning=FALSE, results='hide'}
metric_tables <- list()
joint_tables <- list()
decomposition_tables <- list()
nmi_tables <- list()
funky_metric_tables <- list()
metric_index <- 0L
joint_index <- 0L
decomposition_index <- 0L
nmi_index <- 0L
funky_metric_index <- 0L

for (dataset in batch_datasets) {
  dataset_result <- dataset_results[[dataset]]
  ds_registry <- dataset_result$registry
  primary <- dataset_result$label_col
  labels <- factor(as.character(dataset_result$metadata[[primary]]))
  dataset_heatmap_metrics <- list()
  for (method in selected_methods) {
    method_result <- dataset_result$methods[[method]]
    stored_scores <- method_result$scores
    score_names <- c("anosim_score", "mod_knn3_score", "cluster_score")
    if (!is.null(stored_scores)) {
      if (!is.list(stored_scores) || !all(score_names %in% names(stored_scores))) {
        stop(dataset, "/", method, ": persisted RDS scores lack the benchmark heatmap metrics")
      }
      heatmap_values <- c(
        stored_scores[["anosim_score"]],
        stored_scores[["mod_knn3_score"]],
        stored_scores[["cluster_score"]]
      )
      heatmap_source <- "persisted_scores"
    } else {
      heatmap_anosim <- compute_batch_anosim(
        method_result$dist_mat,
        labels,
        permutations = n_permutations,
        parallel = parallel_cores
      )
      heatmap_values <- c(
        heatmap_anosim$statistic,
        calc_modularity(method_result$dist_mat, labels, 3),
        clust_eval(method_result$dist_mat, labels)
      )
      heatmap_source <- "persisted_distance"
    }
    dataset_heatmap_metrics[[method]] <- data.frame(
      dataset = dataset,
      method = unname(method_label_map_main[[method]]),
      score = c("ANOSIM", "Modularity", "ARI"),
      value = as.numeric(heatmap_values),
      source = heatmap_source,
      stringsAsFactors = FALSE
    )
  }

  null_bundle_path <- file.path(
    input_root,
    "results",
    paste0(dataset, "_batch_effect_uncorrected_ECODA_authors_HR_NULL.rds")
  )
  validate_batch_artifact(null_bundle_path, "ECODA shuffled baseline RDS")
  null_bundle <- readRDS(null_bundle_path)
  null_scores <- null_bundle$scores
  if (is.null(null_scores) || !all(score_names %in% names(null_scores))) {
    stop(dataset, "/ECODA_authors_HR_NULL: persisted RDS scores lack heatmap metrics")
  }
  dataset_heatmap_metrics[["ECODA_authors_HR_NULL"]] <- data.frame(
    dataset = dataset,
    method = unname(method_label_map_main[["ECODA_authors_HR_NULL"]]),
    score = c("ANOSIM", "Modularity", "ARI"),
    value = as.numeric(c(
      null_scores[["anosim_score"]],
      null_scores[["mod_knn3_score"]],
      null_scores[["cluster_score"]]
    )),
    source = "persisted_scores",
    stringsAsFactors = FALSE
  )

  mds_plots <- lapply(selected_methods, function(method) {
    method_result <- dataset_result$methods[[method]]
    anosim_value <- dataset_heatmap_metrics[[method]]$value[
      dataset_heatmap_metrics[[method]]$score == "ANOSIM"
    ]
    plot_mds(
      method_result$dist_mat,
      labels,
      title = paste0(dataset, " — ", unname(method_label_map_main[[method]])),
      cluster_score = FALSE,
      mod_score = FALSE,
      sil_score = FALSE,
      anosim_score = TRUE,
      anosim_value = anosim_value,
      digits = 3
    )
  })
  names(mds_plots) <- selected_methods
  mds_panel <- patchwork::wrap_plots(mds_plots, ncol = 2, guides = "collect") &
    theme(legend.position = "bottom")
  .save_plot(mds_panel, file.path(plot_root, paste0(dataset, "_mds.pdf")), width = 13, height = 10)


  dataset_metrics <- list()
  dataset_joint <- list()
  dataset_decomposition <- list()
  for (method in selected_methods) {
    method_result <- dataset_result$methods[[method]]
    funky_metric_index <- funky_metric_index + 1L
    funky_metric_tables[[funky_metric_index]] <- dataset_heatmap_metrics[[method]]
    metric <- make_batch_metric_table(
      dataset = dataset,
      method = method,
      metadata = dataset_result$metadata,
      candidate_registry = registry,
      dist_mat = method_result$dist_mat,
      artifact_path = method_result$path,
      permutations = n_permutations,
      parallel = parallel_cores,
      permanova_permutations = permanova_permutations,
      precomputed_anosim = dataset_heatmap_metrics[[method]]$value[
        dataset_heatmap_metrics[[method]]$score == "ANOSIM"
      ]
    )
    metric_index <- metric_index + 1L
    metric_tables[[metric_index]] <- metric
    dataset_metrics[[method]] <- metric

    joint <- compute_batch_joint_permanova(
      method_result$dist_mat,
      dataset_result$metadata,
      ds_registry,
      permutations = permanova_permutations,
      parallel = parallel_cores
    )
    joint_table <- make_batch_joint_table(dataset, method, joint, method_result$path)
    joint_index <- joint_index + 1L
    joint_tables[[joint_index]] <- joint_table
    dataset_joint[[method]] <- joint_table

    if (is.null(joint$decomposition)) {
      decomposition <- data.frame(
        component = c("Unique Biological", "Unique Technical", "Shared / Confounded", "Residual / Unexplained"),
        r2 = NA_real_,
        status = joint$status,
        warning = joint$warning,
        n_samples = as.integer(joint$n_samples),
        full_model_r2 = NA_real_,
        stringsAsFactors = FALSE
      )
    } else {
      decomposition <- joint$decomposition
    }
    decomposition$dataset <- dataset
    decomposition$method <- method
    decomposition$method_label <- unname(method_label_map_main[[method]])
    decomposition$artifact_path <- method_result$path
    decomposition_index <- decomposition_index + 1L
    decomposition_tables[[decomposition_index]] <- decomposition
    dataset_decomposition[[method]] <- decomposition
  }

  funky_metric_index <- funky_metric_index + 1L
  funky_metric_tables[[funky_metric_index]] <- dataset_heatmap_metrics[["ECODA_authors_HR_NULL"]]

  metric_data <- do.call(rbind, dataset_metrics)
  metric_data$method_label <- factor(metric_data$method_label, levels = method_labels)
  .save_plot(
    .metric_plot(
      metric_data,
      "anosim_statistic",
      "anosim_status",
      paste0(dataset, " — ANOSIM by metadata covariate"),
      "ANOSIM statistic"
    ),
    file.path(plot_root, paste0(dataset, "_anosim.pdf")),
    width = 14,
    height = 7
  )
  .save_plot(
    .metric_plot(
      metric_data,
      "univariate_r2",
      "univariate_status",
      paste0(dataset, " — univariate PERMANOVA"),
      "Marginal R²"
    ),
    file.path(plot_root, paste0(dataset, "_permanova_univariate.pdf")),
    width = 14,
    height = 7
  )
  .save_plot(
    .metric_plot(
      metric_data,
      "biology_informed_r2",
      "biology_informed_status",
      paste0(dataset, " — biology-informed PERMANOVA"),
      "Candidate marginal R² after primary biology",
      exclude_primary = TRUE
    ),
    file.path(plot_root, paste0(dataset, "_permanova_bio_informed.pdf")),
    width = 14,
    height = 7
  )

  joint_data <- do.call(rbind, dataset_joint)
  joint_data$method_label <- factor(joint_data$method_label, levels = method_labels)
  joint_data$plot_candidate <- ifelse(
    joint_data$status == "ESTIMABLE" & !is.na(joint_data$joint_model_term),
    joint_data$joint_model_term,
    joint_data$candidate
  )
  joint_data$candidate <- joint_data$plot_candidate
  joint_data$candidate_class <- ifelse(
    joint_data$status == "ESTIMABLE" & !is.na(joint_data$joint_model_class),
    joint_data$joint_model_class,
    joint_data$candidate_class
  )
  design_status <- unique(joint_data$joint_design_status)
  design_title <- if (length(design_status) == 1L && design_status == "ESTIMABLE_REDUCED") {
    "joint PERMANOVA candidate terms (reduced estimable design)"
  } else {
    "joint PERMANOVA candidate terms"
  }
  upper <- .metric_plot(
    joint_data,
    "r2",
    "status",
    paste0(dataset, " — ", design_title),
    "Reduced-model marginal R²"
  )
  decomposition_data <- do.call(rbind, dataset_decomposition)
  decomposition_data$method_label <- factor(decomposition_data$method_label, levels = method_labels)
  decomposition_data <- decomposition_data[decomposition_data$status %in% c("ESTIMABLE", "ESTIMABLE_REDUCED") & is.finite(decomposition_data$r2), , drop = FALSE]
  if (nrow(decomposition_data) == 0L) {
    lower <- .blank_plot(
      paste0(dataset, " — global decomposition"),
      "No estimable reduced joint design"
    )
  } else {
    lower <- ggplot(decomposition_data, aes(x = method_label, y = r2, fill = component)) +
      geom_col(width = 0.75) +
      scale_fill_brewer(palette = "Set2", drop = FALSE) +
      labs(title = paste0(dataset, " — global decomposition"), x = NULL, y = "R² component", fill = NULL) +
      theme_classic(base_size = 9) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
  }
  joint_panel <- upper / lower + patchwork::plot_layout(heights = c(1.2, 1))
  .save_plot(
    joint_panel,
    file.path(plot_root, paste0(dataset, "_permanova_joint.pdf")),
    width = 14,
    height = 11
  )

  nmi <- compute_batch_nmi(dataset_result$metadata, ds_registry)
  nmi_table <- make_batch_nmi_table(dataset, nmi, ds_registry)
  nmi_index <- nmi_index + 1L
  nmi_tables[[nmi_index]] <- nmi_table
  .save_plot(
    plot_batch_nmi_heatmap(nmi, dataset),
    file.path(plot_root, paste0(dataset, "_nmi.pdf")),
    width = 11,
    height = 9
  )
}

scores <- do.call(rbind, metric_tables)
scores$.row_order <- seq_len(nrow(scores))
joint_scores <- do.call(rbind, joint_tables)
join_columns <- c(
  "dataset", "method", "candidate", "r2", "pseudo_f", "p_value",
  "p_adjusted_holm", "status", "warning", "reduction_status",
  "reduction_warning", "joint_model_term", "joint_model_class",
  "joint_reduction_group", "joint_design_status", "joint_design_warning",
  "n_samples", "full_model_r2"
)
joint_scores <- joint_scores[, join_columns, drop = FALSE]
colnames(joint_scores)[match(
  c("r2", "pseudo_f", "p_value", "p_adjusted_holm", "status", "warning", "n_samples"),
  colnames(joint_scores)
)] <- c(
  "joint_r2", "joint_pseudo_f", "joint_p_value", "joint_p_adjusted_holm",
  "joint_status", "joint_warning", "joint_n_samples"
)
scores <- merge(scores, joint_scores, by = c("dataset", "method", "candidate"), all.x = TRUE, sort = FALSE)
scores <- scores[order(scores$.row_order), , drop = FALSE]
scores$.row_order <- NULL
row.names(scores) <- NULL

funky_metric_data <- do.call(rbind, funky_metric_tables)
build_batch_funky_heatmap(
  funky_metric_data,
  file.path(plot_root, "batch_effect_uncorrected_funkyheatmap.pdf")
)

decomposition <- do.call(rbind, decomposition_tables)
nmi <- do.call(rbind, nmi_tables)

write_batch_analysis_table(
  scores,
  file.path(analysis_output_root, "batch_effect_uncorrected_scores.csv")
)
write_batch_analysis_table(
  decomposition,
  file.path(analysis_output_root, "batch_effect_uncorrected_decomposition.csv")
)
write_batch_analysis_table(
  nmi,
  file.path(analysis_output_root, "batch_effect_uncorrected_nmi.csv")
)

stopifnot(
  identical(sort(unique(scores$dataset)), sort(batch_datasets)),
  identical(sort(unique(scores$method)), sort(selected_methods)),
  all(vapply(list(
    file.path(analysis_output_root, "batch_effect_uncorrected_scores.csv"),
    file.path(analysis_output_root, "batch_effect_uncorrected_decomposition.csv"),
    file.path(analysis_output_root, "batch_effect_uncorrected_nmi.csv")
  ), function(path) validate_batch_artifact(path, "analysis table")[["path"]] == normalizePath(path), logical(1)))
)
```

# Outputs

The pass writes six base exploratory PDFs per canonical dataset:

- `<dataset>_mds.pdf`
- `<dataset>_anosim.pdf`
- `<dataset>_permanova_univariate.pdf`
- `<dataset>_permanova_bio_informed.pdf`
- `<dataset>_permanova_joint.pdf`
- `<dataset>_nmi.pdf`


The batch-level funky heatmap is written as
`plots/batch_effect_uncorrected/batch_effect_uncorrected_funkyheatmap.pdf`.
It follows the Figure 2A layout with overall, per-dataset, and per-metric
normalized score columns; runtime is intentionally excluded.

Its metric columns are ANOSIM, Modularity, and ARI; the displayed methods also
include the persisted `Shuffled_baseline` scores from
`ECODA_authors_HR_NULL`.

The normalized score, global decomposition, and metadata-only NMI tables are
written under `data/batch_effect/uncorrected/analysis/`, each with an atomic
`MD5/SIZE/PATH` sidecar. Invalid or non-estimable candidates remain in the
score tables with their status and warning; they are never plotted as zero.
