---
title: "Benchmark of exploratory data analysis methods for scRNA-seq data"
author: "Christian Halter"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

# Description

Core script to run the benchmark and generate paper figures.

## Inputs

-   HPC benchmark result bundles (`${NAS_TARGET_DIR}/benchmark/results/`),
    including the composition metadata bundles (`<ds>_metadata.rds`: labels,
    cell/sample counts) and the python-method sample distance/embedding
    `.feather` files + the merged `execution_times.feather`
    (`${NAS_TARGET_DIR}/benchmark/embeddings/`), all rsynced by the benchmark
    submit scripts from the HPC pipelines in `src/5_run_benchmark_methods/`.
    **No h5ad files are read anywhere in the notebook anymore** (all
    benchmark methods — heavy and composition-based — run on HPC).

## Outputs

-   Figures for the manuscript
-   Summarized raw data for appendix (e.g. cell type compositions)

# ---------------------------------------------------------------

# Load libraries and set parameters

```{r message=FALSE, warning=FALSE}
# Setup: See README.md for initial environment setup
if (basename(getwd()) == "notebooks") {
  setwd("..")
  if (file.exists(".Rprofile")) source(".Rprofile")
}
source("src/utils/load_all_functions.R")
datasets <- read_datasets_json(view = "benchmark_analysis")
# ============================================================
# ANALYSIS SCOPE — select which methods + datasets the analysis includes.
#   analysis_scope:    "preprint" = methods and datasets evaluated in the preprint
#                      (calculated with the new pipeline, but excluding post-preprint
#                      methods/datasets);
#                      "all" / "extended" = everything currently in the pipeline.
#   analysis_methods:  NULL = all methods found in the loaded results;
#                      or exact result keys, e.g.
#                      c("ECODA_authors_HR", "MrVI_hvg2000", "QOT_hvg2000")
#   analysis_datasets: NULL = all non-debug datasets from datasets.json;
#                      or dataset keys, e.g. c("Adams", "Pelka") (allows "_debug").
# Defaults reproduce the preprint analysis.
# ============================================================
analysis_scope <- "preprint"   # "preprint" | "all" (or "extended")
analysis_methods <- NULL
analysis_datasets <- NULL

# Result-key regex of methods added after the preprint analysis (excluded
# when analysis_scope == "preprint"). Convention for future methods: (a) cover
# every recode map they appear in, (b) give main methods a default_methods
# regex + screening branch, and (c) add them to extended_method_regex or the
# preprint/all distinction silently stops covering them.
extended_method_regex <- "^QOT_|^PILOT-GM-VAE_"
# Dataset keys added after the preprint analysis (excluded when "preprint").
extended_datasets <- character(0)

stopifnot(analysis_scope %in% c("preprint", "extended", "all"))
if (!is.null(analysis_datasets)) {
  datasets <- datasets[names(datasets) %in% analysis_datasets]
} else {
  # Skip the _debug dataset (Joanito 5-sample subset): the `_*`-prefix
  # convention marks debug-only entries in datasets.json.
  datasets <- datasets[!grepl("^_", names(datasets))]
}
if (analysis_scope == "preprint") {
  datasets <- datasets[!names(datasets) %in% extended_datasets]
}
stopifnot(length(datasets) > 0)

# Scope helpers (closed over analysis_methods, resolved after df_results is
# built; only called from figure chunks, which run later).
methods_in_scope <- function(x) intersect(x, analysis_methods)
methods_in_scope_named <- function(m) m[names(m) %in% analysis_methods]
dataset_display_map <- sapply(datasets, `[[`, "display_name")
# The internal key stays Gongsharma_cmv_young_males (HPC bundle/feather file
# names use it); figures display it as GongSharma.
dataset_display_map[["GongSharma"]] <- dataset_display_map[["Gongsharma_cmv_young_males"]]

# Set paths. Default: LOCAL mirror of the NAS layout pulled into ./data
# (data/benchmark/{results,embeddings,pseudobulks} + slim per-dataset h5ads
# data/<ds>/output/...). To run against the NAS mount instead (slower),
# comment the line below and uncomment the /Volumes/... line.
path_nas_project <- file.path(getwd(), "data")
# path_nas_project <- file.path("", "Volumes", "Shared", "Projects", "ECODA_paper")
# HPC-computed R benchmark results (Pipeline A methods incl. composition,
# <ds>_metadata.rds bundles, trans/zeroimp, rsynced by the submit scripts to
# ${NAS_TARGET_DIR}/benchmark/).
path_nas_benchmark <- file.path(path_nas_project, "benchmark", "results")
# Python-method sample distance/embedding feathers (MrVI/PILOT/scPoli/
# QOT/PILOT-GM-VAE) and the merged execution_times.feather.
path_nas_embeddings <- file.path(path_nas_project, "benchmark", "embeddings")
path_data <- file.path(getwd(), "data")
path_plots <- file.path(getwd(), "plots")

# Separation metrics used in the benchmark figures. Any combination of
# score_label_map names works, e.g. add "sil_score", "mod_knn9_score".
benchmark_metrics <- c("anosim_score", "mod_knn3_score", "cluster_score")
stopifnot(all(benchmark_metrics %in% names(score_label_map)))

# Methods treated as "default" (highlighted in the parameter-screening
# figures): main methods at their default parameter setting. The regex is
# matched against the *recoded* method names in the screening figure.
default_methods <- c(
  "ECODA_authors_HR\\b",
  "MOFA_hvg2000_factors15",
  "Pseudobulk_hvg2000\\b",
  "Pseudobulk_CT_HR_hvg2000",
  "PILOT_hvg2000_HR\\b",
  "scITD_hvg2000_factors5",
  "scPoli_hvg2000_embdims15_HR\\b",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "MrVI_hvg2000",
  "QOT_hvg2000_HR\\b",
  "PILOT-GM-VAE_hvg2000_HR\\b"
)
default_methods_regex <- paste(default_methods, collapse = "|")

set.seed(123)
```

# Create results list to fill

```{r}
# No result_list.rds persistence (Option A): every knit starts fresh and
# reloads all HPC bundles (fast readRDS) — a stale file on disk must not be
# read back in, and nothing writes it anymore (the old stale-in-memory
# clobber hazard is gone; knits are fully deterministic).
result_list <- list()
```

# ---------------------------------------------------------------

# Process datasets

NOTE: Pre-processed (e.g. including cell type annotation, clustering, etc.) dataset .rds files (with shortened names) are deposited online at: <https://doi.org/10.6084/m9.figshare.31456522> . They can be used as input for the following code (after adapting the names according to the names defined in `"datasets.json"`):

```{r message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
# Fail fast if the HPC benchmark results are unavailable: the chunk below
# suppresses warnings, and silently missing bundles would render the whole
# notebook without GloScope/MOFA/Pseudobulk/scITD/composition/trans/zeroimp.
if (!dir.exists(path_nas_benchmark)) {
  stop(
    "HPC benchmark results dir not found: ", path_nas_benchmark,
    "\nRun the HPC benchmark pipelines first (src/5_run_benchmark_methods/",
    "run_r_sample_embedding_methods/ and run_transformation_zeroimp_analysis/)",
    " and make sure the NAS is mounted."
  )
}
if (length(list.files(path_nas_benchmark, pattern = "\\.rds$")) == 0) {
  stop(
    "No RDS result bundles found in ", path_nas_benchmark,
    "\nRun the HPC benchmark pipelines first (see TODO.md Phase 3.2)."
  )
}
if (!dir.exists(path_nas_embeddings)) {
  stop(
    "Python benchmark embeddings dir not found: ", path_nas_embeddings,
    "\nRun the HPC python benchmark pipeline first (src/5_run_benchmark_methods/",
    "run_python_sample_embedding_methods/) and make sure the NAS is mounted."
  )
}

sample_col <- "Sample"
# HVG levels / factor (PCA-dim) levels mirror run_benchmark_analysis'
# defaults (now computed on HPC by run_composition_methods_hpc).
HVGs <- c(1000, 2000, 3000)
factors_test <- c(2, 3, 5, 10, 15)
# Per-dataset cell-count/statistics collected during the loop (display keys;
# used by the exec-time figures and Supp table 1 instead of full reloads).
n_cells_by_ds <- list()
ds_stats_list <- list()

for (ds in names(datasets)) {
  # Per-dataset inputs: HPC result bundles + the composition metadata bundle
  # (<ds>_metadata.rds, written by the composition worker: labels, n_cells,
  # n_samples, cells_per_sample, n_cell_types_high_res). No h5ad/pseudobulk
  # reads anywhere in the notebook anymore.
  metadata_file <- file.path(path_nas_benchmark, paste0(ds, "_metadata.rds"))
  if (!file.exists(metadata_file)) {
    stop(
      "Composition metadata bundle not found: ", metadata_file,
      "\nRun the HPC composition benchmark for this dataset first ",
      "(src/5_run_benchmark_methods/run_r_sample_embedding_methods/ ",
      "--methods composition; auto-prepends prepare_pseudobulk)."
    )
  }
  ds_meta <- readRDS(metadata_file)
  labels <- ds_meta$labels

  # All R benchmark methods — heavy (GloScope, MOFA, Pseudobulk, scITD) AND
  # composition-based (ECODA_*, GloProp, Freq_highres, Avg_PCA_embedding,
  # ECODA_deconv) — run on HPC (src/5_run_benchmark_methods/
  # run_r_sample_embedding_methods/); the transformation and zero-imputation
  # analyses too. Bundles are loaded FRESH from the NAS every knit (no
  # result_list.rds persistence, no rerun semantics). NOTE:
  # ECODA_authors_HR_NULL (shuffle_labels null control) is computed on HPC
  # with set.seed(123) per dataset and is NOT bit-identical to the old
  # notebook-local run (different RNG stream) — irrelevant for a null
  # control.
  result_list <- load_hpc_benchmark_results(
    result_list, ds, path_nas_benchmark,
    methods = c("gloscope", "mofa", "pseudobulk", "scitd", "composition")
  )

  # Python-feather methods (MrVI/PILOT/scPoli/QOT/PILOT-GM-VAE): fast
  # sample-distance/embedding feather reads + scoring, kept in the notebook
  # (seconds, no h5ad I/O). Labels come from the metadata bundle.
  for (i in HVGs) {
    if (i == 2000) {
      scpoli_dims <- factors_test
    } else {
      scpoli_dims <- 15
    }

    file_mrvi <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg", i, "_mrvi_dists.feather")
    )
    file_pilot <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg", i, "_highres_pilot_dists.feather")
    )
    files_scpoli <- file.path(
      path_nas_embeddings,
      paste0(
        ds, "_hvg", i, "_highres_scpoli_dims", scpoli_dims, "_embs.feather"
      )
    )
    file_highres_pilot <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg2000_highres_pilot_dists.feather")
    )

    files_to_check <- c(file_mrvi, file_pilot, files_scpoli, file_highres_pilot)
    missing_files <- files_to_check[!file.exists(files_to_check)]

    if (length(missing_files) > 0) {
      stop(
        "The following file(s) are missing:\n",
        paste(missing_files, collapse = "\n")
      )
    }
  }
  for (i in HVGs) {
    # --- MrVI (Runs once per HVG) ---
    mrvi_dist_file <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg", i, "_mrvi_dists.feather")
    )
    result_list[["bmark"]][[ds]][[paste0("MrVI_hvg", i)]] <- process_mrvi_fig(
      mrvi_dist_file = mrvi_dist_file,
      labels
    )

    # --- PILOT (Runs once per HVG) ---
    pilot_dist_file <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg", i, "_highres_pilot_dists.feather")
    )
    result_list[["bmark"]][[ds]][[paste0("PILOT_hvg", i)]] <- process_pilot_fig(
      pilot_dist_file = pilot_dist_file,
      labels
    )

    # --- QOT (Runs once per HVG) ---
    # Pending method (TODO.md Phase 3): feathers may be absent -> skip with a
    # message instead of failing the whole dataset.
    qot_dist_file <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg", i, "_highres_qot_dists.feather")
    )
    if (file.exists(qot_dist_file)) {
      result_list[["bmark"]][[ds]][[paste0("QOT_hvg", i)]] <- process_qot_fig(
        qot_dist_file = qot_dist_file,
        labels
      )
    } else {
      message("QOT_hvg", i, " skipped for ", ds,
              ": feather not found (pending method, TODO.md Phase 3)")
    }

    # --- PILOT-GM-VAE (Runs once per HVG) ---
    # Pending method (TODO.md Phase 3): feathers may be absent -> skip with a
    # message instead of failing the whole dataset.
    pilotgm_dist_file <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg", i, "_highres_pilotgm_dists.feather")
    )
    if (file.exists(pilotgm_dist_file)) {
      result_list[["bmark"]][[ds]][[paste0("PILOT-GM-VAE_hvg", i)]] <-
        process_pilotgm_fig(
          pilotgm_dist_file = pilotgm_dist_file,
          labels
        )
    } else {
      message("PILOT-GM-VAE_hvg", i, " skipped for ", ds,
              ": feather not found (pending method, TODO.md Phase 3)")
    }

    scpoli_emb_file <- file.path(
      path_nas_embeddings,
      paste0(ds, "_hvg", i, "_highres_scpoli_dims15_embs.feather")
    )
    result_list[["bmark"]][[ds]][[paste0(
      "scPoli_hvg",
      i,
      "_dims15_highres"
    )]] <- process_scpoli_fig(scpoli_emb_file = scpoli_emb_file, labels)

    # --- scPoli (Runs once OR multiple times depending on HVG) ---
    if (i == 2000) {
      target_dims <- factors_test
      for (f in target_dims) {
        scpoli_emb_file <- file.path(
          path_nas_embeddings,
          paste0(
            ds,
            "_hvg",
            i,
            "_highres_scpoli_dims",
            f,
            "_embs.feather"
          )
        )
        result_list[["bmark"]][[ds]][[paste0(
          "scPoli_hvg",
          i,
          "_dims",
          f,
          "_highres"
        )]] <- process_scpoli_fig(scpoli_emb_file = scpoli_emb_file, labels)
      }

      pilot_dist_file <- file.path(
        path_nas_embeddings,
        paste0(ds, "_hvg", i, "_lowres_pilot_dists.feather")
      )
      result_list[["bmark"]][[ds]][[paste0("PILOT_hvg", i, "_lowres")]] <-
        process_pilot_fig(
          pilot_dist_file = pilot_dist_file,
          labels
        )

      qot_dist_file <- file.path(
        path_nas_embeddings,
        paste0(ds, "_hvg", i, "_lowres_qot_dists.feather")
      )
      if (file.exists(qot_dist_file)) {
        result_list[["bmark"]][[ds]][[paste0("QOT_hvg", i, "_lowres")]] <-
          process_qot_fig(
            qot_dist_file = qot_dist_file,
            labels
          )
      } else {
        message("QOT_hvg", i, "_lowres skipped for ", ds,
                ": feather not found (pending method, TODO.md Phase 3)")
      }

      pilotgm_dist_file <- file.path(
        path_nas_embeddings,
        paste0(ds, "_hvg", i, "_lowres_pilotgm_dists.feather")
      )
      if (file.exists(pilotgm_dist_file)) {
        result_list[["bmark"]][[ds]][[paste0("PILOT-GM-VAE_hvg", i, "_lowres")]] <-
          process_pilotgm_fig(
            pilotgm_dist_file = pilotgm_dist_file,
            labels
          )
      } else {
        message("PILOT-GM-VAE_hvg", i, "_lowres skipped for ", ds,
                ": feather not found (pending method, TODO.md Phase 3)")
      }

      scpoli_emb_file <- file.path(
        path_nas_embeddings,
        paste0(ds, "_hvg", i, "_lowres_scpoli_dims15_embs.feather")
      )
      result_list[["bmark"]][[ds]][[paste0(
        "scPoli_hvg",
        i,
        "_dims15_lowres"
      )]] <- process_scpoli_fig(scpoli_emb_file = scpoli_emb_file, labels)
    }
  }

  # Per-dataset stats for the exec-time figures + Supp table 1 (from the
  # metadata bundle; no obs reads).
  ds_key_out <- if (ds == "Gongsharma_cmv_young_males") "GongSharma" else ds
  n_cells_by_ds[[ds_key_out]] <- ds_meta$n_cells
  ct_per_sample <- ds_meta$cells_per_sample
  ds_stats_list[[ds_key_out]] <- data.frame(
    dataset = ds_key_out,
    n_samples = ds_meta$n_samples,
    n_cells = ds_meta$n_cells,
    cells_per_sample_mean = mean(ct_per_sample),
    cells_per_sample_sd = sd(ct_per_sample),
    n_cell_types = ds_meta$n_cell_types_high_res
  )

  rm(ds_meta, labels, ct_per_sample)
  gc()
}

# Prune any loaded entries of excluded datasets so every later
# names(result_list[["bmark"]]) loop is scoped (nothing persists to disk;
# pruning is in-memory only).
result_list[["bmark"]] <- result_list[["bmark"]][names(datasets)]
result_list[["trans"]] <- result_list[["trans"]][names(datasets)]
result_list[["zeroimp"]] <- result_list[["zeroimp"]][names(datasets)]

# Combined per-dataset stats (used by Supp table 1).
ds_stats_df <- bind_rows(ds_stats_list)
```

# ---------------------------------------------------------------

# Create list to export raw data to excel

```{r}
df_results_list <- list()
```

# ---------------------------------------------------------------

# Plot benchmark results

## Save and flatten the nested list into a DataFrame

```{r}
bmark_result_list <- result_list[["bmark"]]

bmark_result_list$GongSharma <- bmark_result_list$Gongsharma_cmv_young_males
bmark_result_list$Gongsharma_cmv_young_males <- NULL

df_results <- imap_dfr(bmark_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, ~ {
    method_name <- .y
    if ("scores" %in% names(.x)) {
      n_samples <- nrow(.x$feat_mat)
      n_features <- ncol(.x$feat_mat)
      
      imap_dfr(.x$scores, ~ {
        tibble(
          dataset = dataset_name,
          method = method_name,
          score = .y,
          value = .x,
          n_samples = n_samples,
          n_features = n_features
        )
      })
    } else {
      tibble() # Return an empty tibble if "scores" is not found
    }
  })
})

# Drop old modularity score method
df_results <- df_results[!df_results$score == "mod_score", ]

# Resolve the analysis scope: analysis_methods NULL = all methods found in
# the loaded results; "preprint" drops the extended (post-preprint) methods.
# Runs BEFORE the factor() below so the method levels are exactly the scoped
# set.
if (is.null(analysis_methods)) {
  analysis_methods <- unique(as.character(df_results$method))
}
if (analysis_scope == "preprint") {
  analysis_methods <- analysis_methods[
    !grepl(extended_method_regex, analysis_methods)
  ]
}
df_results <- df_results[df_results$method %in% analysis_methods, ]

# datasets.json `not_suitable_for_auto_annotation` flag (see AGENTS.md):
# datasets flagged for the annotation method families are dropped from the
# HiTME/scATOMIC annotation-method figures (Figure 3 A + Supp fig 19) with
# this documented message. The dataset stays in Figure 2a, where the author
# CT columns drive the methods. Absent/empty flag = suitable for all.
flag_status_for <- function(ds_key) {
  e <- datasets[[ds_key]]
  if (is.null(e) || is.null(e$not_suitable_for_auto_annotation)) {
    character(0)
  } else {
    e$not_suitable_for_auto_annotation
  }
}
hitme_flagged_ds <- names(datasets)[vapply(names(datasets), function(k) "hitme" %in% flag_status_for(k), logical(1))]
scatomic_flagged_ds <- names(datasets)[vapply(names(datasets), function(k) "scatomic" %in% flag_status_for(k), logical(1))]
if (length(hitme_flagged_ds) > 0 || length(scatomic_flagged_ds) > 0) {
  message("not_suitable_for_auto_annotation: HiTME excluded for [",
          paste(hitme_flagged_ds, collapse = ", "), "], scATOMIC excluded for [",
          paste(scatomic_flagged_ds, collapse = ", "),
          "] in the annotation-comparison figures")
}
flag_drop_annotation_methods <- function(df) {
  df %>%
    filter(!(dataset %in% hitme_flagged_ds &
               method %in% c("ECODA_HiTME_HR_layer2", "ECODA_HiTME_HR_layer3"))) %>%
    filter(!(dataset %in% scatomic_flagged_ds & method == "ECODA_scATOMIC_HR"))
}
```

## Show + sort all tested conditions

```{r}
levs <- mixedsort(as.character(unique(df_results$method)))
levs

df_results$method <- factor(df_results$method, levels = levs)

df_results_list[["Benchmark"]] <- df_results
```

## Plot results

### All methods (overview)

```{r fig.height=8, fig.width=16}
df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = as_labeller(score_facet_labels)))
ggsave(file.path(path_plots, "benchmark_summary_barplot_separatescores_bars.pdf"),
       width = 14,
       height = 7)
```

### Supp fig 17: Benchmark unscaled metrics

```{r fig.width=12, fig.height=4}
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  # "ECODA_HiTME_HR_layer2",
  # "ECODA_HiTME_HR_layer3",
  # "ECODA_seuratres_2",
  # "ECODA_PB_combo_normquantile_ecodaweight0.5",
  # paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  # "ECODA_authors_LR",
  # "Avg_PCA_embedding",
  # "Freq_highres",
  # "ECODA_deconv",
  # "Pseudobulk_10_PCA_dims",
  "Pseudobulk_hvg2000",
  # "Pseudobulk_hvg2000_bl",
  # "Pseudobulk_schvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))


df_plot <- df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         # "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         # "ECODA_PB_combo_normquantile_ecodaweight0.5" = "ECODA+Pseudobulk",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims15_highres" = "scPoli",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30" = "GloScope",
                         "QOT_hvg2000" = "QOT",
                         "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"))

pointsize <- 1.5

# Plot 1: ANOSIM
p1 <- df_plot %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

# Plot 2: Modularity
p2 <- df_plot %>%
  filter(grepl(pattern = "mod_knn3_score", score)) %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "", y = "Modularity score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

# Plot 3: ARI
p3 <- df_plot %>%
  filter(score == "cluster_score") %>%
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "", y = "ARI score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))

# --- Combine using Patchwork ---

combined_plot <- (p1 | p2 | p3) + 
  plot_layout(guides = "collect") & # Collects all legends into one
  theme(legend.position = "right") # Places the shared legend at the bottom
combined_plot

# Save the combined plot
ggsave(file.path(path_plots, "Supp_fig_17_benchmark_unscaled_metrics_perdataset.pdf"), 
       plot = combined_plot,
       width = 12, # Increased width to fit 3 plots
       height = 5)
```

```{r fig.height=4, fig.width=4}
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  "Avg_PCA_embedding",
  "Freq_highres",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims5_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))


df_plot_scaled <- df_results %>%
  filter(method %in% filter_methods) %>%
  # -- Recoding names --
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_Leiden_clustering",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims5_highres" = "scPoli",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30" = "GloScope",
                         "QOT_hvg2000" = "QOT",
                         "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE")) %>%
  # -- CRITICAL STEP: Min-Max Scaling per Dataset and Score --
  group_by(dataset, score) %>% 
  mutate(scaled_value = min_max(value)) %>% 
  ungroup()

# 4. Generate the Plot
# We filter for ANOSIM and use 'scaled_value' instead of 'value'
df_plot_scaled %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = fct_reorder(method, scaled_value, .desc = TRUE, .fun = mean), 
             y = scaled_value)) + # Use the scaled column
  
  # Error bars and Mean Bar
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  
  # Individual Points and connecting lines (Paired data)
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  
  # Statistical Test
  stat_compare_means(
    ref.group = "ECODA_authors",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    vjust = 1,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  
  # Adjust Y-axis (Since data is 0-1, we set limit slightly higher for p-values)
  ylim(c(0, 1.15)) + 
  
  labs(x = "", y = "Separation score\n(Scaled ANOSIM score)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
ggsave(file.path(path_plots, "benchmark_summary_barplot_ANOSIM_minmaxscaled_mainmethods_refECODA.pdf"),
       width = 4.5,
       height = 4.5)
```

#### Plot all separation metrics together

```{r fig.height=6, fig.width=5}
df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30" = "GloScope",
                         "QOT_hvg2000" = "QOT",
                         "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE")) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 65, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             labeller = as_labeller(score_facet_labels))
ggsave(file.path(path_plots, "benchmark_summary_barplot_allmetricscombined_mainmethods_with_bad_baselines.pdf"),
       width = 4.5,
       height = 6.5)
```

#### Plot PCA for selected datasets and main methods

```{r}
ds <- "Kfoury"

m <- "ECODA_authors_HR"
p1 <- plot_pca(
  result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
  labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
  coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
  title = m, n_ct_show = 5, invisible = c("quali")
)
print(p1)

plot_mds(
  result_list[["bmark"]][[ds]][[m]][["dist_mat"]],
  labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
  coord_equal = TRUE, cluster_score = FALSE, mod_score = FALSE,
  title = m
)

m <- "ECODA_HiTME_HR_layer2"
p2 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_HiTME", n_ct_show = 5, invisible = c("quali")
  )
print(p2)

m <- "ECODA_scATOMIC_HR"
p3 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_scATOMIC", n_ct_show = 5, invisible = c("quali")
  )
print(p3)

p_kfoury <- wrap_plots(list(p1, p2, p3)) +
  plot_layout(nrow = 1, guides = "collect")
# NOTE (review): the dataset name was previously duplicated in the filename
# (Supp_fig_18_Kfoury_PCA_authors_vs_auto_Kfoury.pdf); fixed to drop the
# trailing duplicate.
ggsave(file.path(path_plots, "Supp_fig_18_Kfoury_PCA_authors_vs_auto.pdf"),
       width = 15,
       height = 6)


ds <- "Wu"

m <- "ECODA_authors_HR"
p1 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = m, n_ct_show = 5, invisible = c("quali")
  )
print(p1)

m <- "ECODA_HiTME_HR_layer2"
p2 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_HiTME", n_ct_show = 5, invisible = c("quali")
  )
print(p2)

m <- "ECODA_scATOMIC_HR"
p3 <- plot_pca(
    result_list[["bmark"]][[ds]][[m]][["feat_mat"]],
    labels = result_list[["bmark"]][[ds]][[m]][["labels"]],
    coord_equal = TRUE, repel = TRUE, cluster_score = FALSE, mod_score = FALSE,
    title = "ECODA_scATOMIC", n_ct_show = 5, invisible = c("quali")
  )
print(p3)

p_wu <- wrap_plots(list(p1, p2, p3)) +
  plot_layout(nrow = 1, guides = "collect")
ggsave(file.path(path_plots, "Supp_fig_18_Wu_PCA_authors_vs_auto.pdf"),
       width = 15,
       height = 6)

p_combined <- wrap_plots(list(p_kfoury + labs(tag = "Kfoury"), p_wu + labs(tag = "Wu"))) +
  plot_layout(nrow = 2) & 
  theme(plot.tag = element_text(face = "bold"))
ggsave(file.path(path_plots, paste0("Supp_fig_18_combined.pdf")),
       width = 15,
       height = 6)
```

```{r fig.height=12, fig.width=12}
ds <- "Pelka"
plot_list <- list()

methods <- list(
  "ECODA_authors_HR" = "ECODA",
  "ECODA_authors_HR_NULL" = "Shuffled_baseline",
  "Pseudobulk_10_PCA_dims" = "Pseudobulk",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scITD_hvg2000_factors5" = "scITD",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims5_highres" = "scPoli",
  "PILOT_hvg2000" = "PILOT",
  "GloScope_hvg2000_pcadims30" = "GloScope",
  "QOT_hvg2000" = "QOT",
  "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"
)
methods <- methods_in_scope_named(methods)

for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 3, guides = "collect")
```

### Plot PCA for all datasets (PB and ECODA)

```{r fig.height=4, fig.width=8}
methods <- list(
  "Pseudobulk_hvg2000" = "Pseudobulk",
  # "ECODA_HiTME_HR_layer2" = "ECODA_automated",
  "ECODA_authors_HR" = "ECODA"
)
methods <- methods_in_scope_named(methods)

for (ds in names(result_list[["bmark"]])) {
  plot_list <- list()
  
  for (m in names(methods)) {
    feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
    labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
    
    if (ds=="Bassez") levels(labels) <- c("Responders", "Non-responders")
    if (ds=="Zhang") levels(labels) <- tools::toTitleCase(levels(labels))
    
    if (methods[[m]] == "ECODA") title <- "Cell type composition (ECODA)" else title <- methods[[m]]
    
    plot_list[[m]] <- plot_pca(
      feat_mat = feat_mat,
      labels = labels,
      cluster_score = FALSE, mod_score = FALSE, anosim_score = FALSE,
      coord_equal = FALSE,
      n_ct_show = 5, invisible = "quali", repel = TRUE,
      title = title
    )
    
    if (ds=="Bassez") {
      plot_list[[m]] <- plot_list[[m]] +
        scale_color_discrete(
          name = "Treatment response", 
          labels = labels
        ) +
        scale_shape_discrete(
          name = "Treatment response",
          labels = labels
        )
    }
    if (ds=="Zhang") {
      plot_list[[m]] <- plot_list[[m]] +
        scale_color_discrete(
          name = "Tissue", 
          labels = labels
        ) +
        scale_shape_discrete(
          name = "Tissue", 
          labels = labels
        )
    }
    
  }
  
  wrap_plots(plot_list) +
    plot_layout(nrow = 1, guides = "collect")
  
  ggsave(file.path(path_plots, paste0("benchmark_pca_", ds, ".pdf")),
         width = 8,
         height = 3.5)
}

```

### Supp fig 3-13: MDS for all datasets

```{r fig.height=9, fig.width=11}
methods <- list(
  "ECODA_authors_HR" = "ECODA",
  # "ECODA_seuratres_2" = "ECODA_Leiden_res_2",
  # "ECODA_HiTME_HR_layer2" = "ECODA_automated",
  
  # "Freq_highres" = "ECODA_Freq%",
  "Pseudobulk_hvg2000" = "Pseudobulk",
  # "Avg_PCA_embedding" = "Avg_PCA_embedding",
  
  "GloScope_hvg2000_pcadims30" = "GloScope",
  "GloProp" = "GloProp",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims15_highres" = "scPoli",
  
  "MOFA_hvg2000_factors15" = "MOFA",
  "PILOT_hvg2000" = "PILOT",
  "scITD_hvg2000_factors5" = "scITD",
  "QOT_hvg2000" = "QOT",
  "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"
)
methods <- methods_in_scope_named(methods)

idx <- 3

for (ds in names(result_list[["bmark"]])) {
  plot_list <- list()
  
  all_possible_labels <- unique(unlist(lapply(names(methods), function(m) {
    as.character(result_list[["bmark"]][[ds]][[m]][["labels"]])
  })))
  
  if (ds == "Bassez") {
    all_possible_labels <- c("Responders", "Non-responders")
  }
  
  for (m in names(methods)) {
    dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
    labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
    
    if (ds == "Bassez") {
      labels <- as.character(labels)
      labels[labels == "E"] <- "Responders"
      labels[labels == "NE"] <- "Non-responders"
      labels <- factor(labels, levels = c("Responders", "Non-responders"))
    }
    
    labels <- factor(labels, levels = sort(all_possible_labels))
    
    plot_list[[m]] <- plot_mds(
      dist_mat,
      labels,
      cluster_score = FALSE, mod_score = FALSE, anosim_score = FALSE,
      coord_equal = TRUE,
      title = methods[[m]]
    )
  }
  
  wrap_plots(plot_list) +
    plot_layout(nrow = 3, guides = "collect")
  
  ggsave(file.path(path_plots, paste0("Supp_fig_", idx, "_benchmark_mds_", ds, ".pdf")),
         width = 11,
         height = 9)
  
  idx <- idx + 1
}

```

# Plot MDS of selected datasets and methods

### Compare PCA based on features vs dist

```{r fig.height=6, fig.width=18}
ds <- "Pelka"
plot_list <- list()

methods <- list(
  "ECODA_authors_HR" = "ECODA_authors",
  "ECODA_authors_HR_NULL" = "Shuffled_baseline",
  "Freq_highres" = "ECODA_Freq%",
  "ECODA_seuratres_2" = "ECODA_unsupervised",
  "ECODA_HiTME_HR_layer2" = "ECODA_automated",
  "Pseudobulk_hvg2000" = "Pseudobulk",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scITD_hvg2000_factors5" = "scITD",
  "MrVI_hvg2000" = "MrVI",
  "scPoli_hvg2000_dims5_highres" = "scPoli",
  "PILOT_hvg2000" = "PILOT",
  "GloScope_hvg2000_pcadims30" = "GloScope",
  "QOT_hvg2000" = "QOT",
  "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"
)
methods <- methods_in_scope_named(methods)

for (m in names(methods)) {
  feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  if (ds == "Pelka") {
    labels <- as.character(labels)
    labels[labels == "N"] <- "Normal"
    labels[labels == "T"] <- "Tumor"
  }
  if (ds == "Bassez") {
    labels <- as.character(labels)
    labels[labels == "E"] <- "Expanders"
    labels[labels == "NE"] <- "Non-expanders"
  }
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  avg_sep_score <- result_list[["bmark"]][[ds]][[m]]$scores[c("mod_knn3_score", "anosim_score", "cluster_score")] %>%
    unlist() %>%
    mean() %>%
    round(2)
  
  plot_list[[m]] <- plot_pca(
    feat_mat = feat_mat,
    # cluster_score = F, mod_score = F, anosim_score = F,
    # knn_k = NULL,
    labels = labels,
    coord_equal = FALSE,
    title = paste0(methods[[m]], "\nSeparation score: ", sprintf("%.2f", avg_sep_score))
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect") +
  plot_annotation(title = ds)

```

```{r}
for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  if (ds == "Pelka") {
    labels <- as.character(labels)
    labels[labels == "N"] <- "Normal"
    labels[labels == "T"] <- "Tumor"
  }
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    cluster_score = F, mod_score = F, anosim_score = F,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect") +
  plot_annotation(title = ds)


for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  if (ds == "Pelka") {
    labels <- as.character(labels)
    labels[labels == "N"] <- "Normal"
    labels[labels == "T"] <- "Tumor"
  }
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = dist_mat,
    labels = labels,
    coord_equal = FALSE,
    title = methods[[m]],
    cluster_score = FALSE,
    mod_score = FALSE,
    anosim_score = FALSE
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect") +
  plot_annotation(title = ds)
```

```{r fig.height=12, fig.width=12}
for (m in names(methods)) {
  feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  if (m == "ECODA_authors_HR_NULL") {
    labels <- sample(labels)
  }
  
  plot_list[[m]] <- plot_pca(
    feat_mat = feat_mat,
    labels = labels,
    n_ct_show = 5,
    invisible = "quali",
    coord_equal = FALSE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 3, guides = "collect")

```

### Show example heatmap

```{r fig.height=7, fig.width=7}
ds <- "Bassez"

feat_mat <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$feat_mat
labels <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels

df_heatmap <- feat_mat %>% scale(center = TRUE, scale = FALSE) %>% t() %>% as.data.frame()
my_sample_col <- list(
  subtype = as.factor(labels)
) %>% as.data.frame()
row.names(my_sample_col) <- colnames(df_heatmap)

pdf(file = file.path(path_plots, paste0("Heatmap_", ds, ".pdf")),
    width = 6,
    height = 6)
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4
)
dev.off()
pheatmap(
  df_heatmap,
  annotation_col = my_sample_col,
  # cluster_rows = FALSE,
  # cluster_cols = FALSE,
  scale = "none",
  clustering_method = "ward.D2",
  show_colnames = FALSE,
  cutree_rows = 4,
  cutree_cols = 4
)

plot_pca(
    feat_mat = feat_mat,
    labels = labels,
    n_ct_show = 5,
    invisible = "quali",
    coord_equal = FALSE,
    title = "",
    repel = TRUE
)
ggsave(file.path(path_plots, paste0("PCA_arrows_", ds, ".pdf")),
       width = 6,
       height = 6)
```

### Figure 3 A: ECODA annotation method comparison

### and Supp fig 19

```{r fig.height=6, fig.width=5}
pointsize <- 1.5

filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_LR",
  "ECODA_deconv",
  "ECODA_HiTME_HR_layer2",
  # "ECODA_HiTME_HR_layer3",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_0.1",
  "ECODA_seuratres_0.4",
  "ECODA_seuratres_2",
  "ECODA_seuratres_5",
  "ECODA_seuratres_20"
)


df_plot <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(!grepl("Zhang|Lee", dataset)) %>% # Exclude Zhang and Lee, as they do not have high-resolution author annotation
  filter(method %in% filter_methods) %>%
  flag_drop_annotation_methods() %>%
  apply_method_labels(method_label_map_annotation)
  
p <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "ECODA_authors_HR",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 1
  ) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.15)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p)
ggsave(file.path(path_plots, "Figure_3_A_annotationmethods_barplot_anosim.pdf"),
       width = 5,
       height = 5)



df_plot <- df_results %>%
  filter(score == "mod_knn3_score") %>%
  filter(!grepl("Zhang|Lee", dataset)) %>% # Exclude Zhang and Lee, as they do not have high-resolution author annotation
  filter(method %in% filter_methods) %>%
  flag_drop_annotation_methods() %>%
  apply_method_labels(method_label_map_annotation)
  
p_mod <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "ECODA_authors_HR",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 1.1
  ) +
  labs(x = "", y = "Separation score (Modularity)") +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.15)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p_mod)
ggsave(file.path(path_plots, "benchmark_summary_barplot_modularity_annotationmethods_refECODAhr.pdf"),
       width = 2.5,
       height = 4)


df_plot <- df_results %>%
  filter(score == "cluster_score") %>%
  filter(!grepl("Zhang|Lee", dataset)) %>% # Exclude Zhang and Lee, as they do not have high-resolution author annotation
  filter(method %in% filter_methods) %>%
  flag_drop_annotation_methods() %>%
  apply_method_labels(method_label_map_annotation)
  
p_ari <- ggplot(df_plot, aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_plot$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "ECODA_authors_HR",
    method = "wilcox.test",
    paired = TRUE, # Set to FALSE as the groups are not paired
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 1.1
  ) +
  labs(x = "", y = "Separation score (ARI)") +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.15)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p_ari)
ggsave(file.path(path_plots, "benchmark_summary_barplot_ari_annotationmethods_refECODAhr.pdf"),
       width = 2.5,
       height = 4)


wrap_plots(list(p_mod, p_ari)) +
  plot_layout(nrow = 1, guides = "collect")
ggsave(file.path(path_plots, "Supp_fig_18_annotationmethods_barplot_mod_ari.pdf"),
       width = 9,
       height = 5)
```

#### Figure 3 B: Plot PCA for selected annot. methods (Adams)

```{r fig.height=8, fig.width=8}
ds <- "Adams"

methods <- list(
  "ECODA_authors_HR" = "ECODA_authors_HR",
  "ECODA_seuratres_5" = "ECODA_Leiden_res_5",
  "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
  "ECODA_authors_LR" = "ECODA_authors_LR"
  # "ECODA_scATOMIC_HR" = "ECODA_scATOMIC",
  # "ECODA_deconv" = "ECODA_deconv"
  # "ECODA_seuratres_0.1" = "ECODA_Leiden_res_0.1"
)
methods <- methods_in_scope_named(methods)

plot_list <- list()

for (m in names(methods)) {
  feat_mat <- result_list[["bmark"]][[ds]][[m]][["feat_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  
  plot_list[[m]] <- plot_pca(
    feat_mat = feat_mat,
    labels = labels,
    cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
    coord_equal = TRUE, pointsize = 2,
    n_ct_show = 5, invisible = "quali", repel = TRUE,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect")

ggsave(file.path(path_plots, "Figure_3_B_annotationmethods_pca_adams.pdf"),
       width = 8,
       height = 7)
```

#### Supp fig 19: Plot PCA for selected annot. methods (GongSharma)

```{r fig.height=10, fig.width=12}
datas <- c("Bassez", "Lee", "Stephenson")

methods <- list(
  "ECODA_authors_HR" = "ECODA_authors_HR",
  # "ECODA_authors_LR" = "ECODA_authors_LR",
  "ECODA_HiTME_HR_layer2" = "ECODA_HiTME",
  "ECODA_scATOMIC_HR" = "ECODA_scATOMIC",
  # "ECODA_deconv" = "ECODA_deconv",
  "ECODA_seuratres_2" = "ECODA_Leiden_res_2"
  # "ECODA_seuratres_0.1" = "ECODA_Leiden_res_0.1"
)
methods <- methods_in_scope_named(methods)

plot_list <- list()

for (ds in datas) {
  pl <- list()
  for (m in names(methods)) {
    if (ds == "Lee" & m == "ECODA_authors_HR") next
    
    dist_mat <- result_list[["bmark"]][[ds]][[m]]$dist_mat
    labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
    
    if (ds == "Bassez") {
      labels <- as.character(labels)
      labels[labels == "E"] <- "Responders"
      labels[labels == "NE"] <- "Non-responders"
      labels <- factor(labels, levels = c("Responders", "Non-responders"))
    }
    
    pl[[m]] <- plot_mds(
      dist_mat = dist_mat,
      labels = labels,
      cluster_score = FALSE, mod_score = FALSE, anosim_score = TRUE,
      coord_equal = TRUE,
      title = methods[[m]]
    )
  }
  
  # Wrap the row patchwork in wrap_elements to preserve the annotation
  row_plot <- wrap_plots(pl, nrow = 1, guides = "collect") + 
    plot_annotation(
      title = ds, 
      theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5))
    )
  
  plot_list[[ds]] <- wrap_elements(row_plot)
}

# Combine the rows vertically
wrap_plots(plot_list, ncol = 1)


ggsave(file.path(path_plots, paste0("Supp_fig_19_Bassez_Lee_Stephenson.pdf")),
       width = 12,
       height = 10)
```

### Supp fig 2: Method parameter screening

```{r fig.height=8, fig.width=14}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_LR",
  "ECODA_deconv",
  "ECODA_HiTME_HR_layer2",
  "ECODA_HiTME_HR_layer3",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_0.4",
  "ECODA_seuratres_2",
  "ECODA_seuratres_5",
  "ECODA_seuratres_20"
)

# default_methods / default_methods_regex are defined once in the setup chunk
# (highlighting of default methods in this screening figure).

df_results %>%
  # filter(score == "anosim_score") %>%
  filter(!grepl("HiTME|scATOMIC|seuratres|ECODA_authors_HR_NULL|ECODA_authors_LR|ECODA_deconv|Freq_highres|varexp|varcts|Avg_PCA|combo|CT_LR|schvg", method)
  ) %>%
  filter(
    if_else(grepl("scPoli", method), grepl("highres|lowres", method), TRUE)
  ) %>%
  # filter(method %in% filter_methods) %>%
  # mutate(method = recode(method,
  #                        "MOFA_2_factors" = "MOFA",
  #                        "scITD_2_factors" = "scITD",
  #                        "ECODA_highres" = "ECODA")) %>%
  
  # CREATE THE FACETING GROUP COLUMN
  mutate(
    method_plot = case_when(
      grepl("^ECODA_", method) ~ "ECODA",
      grepl("^MOFA_", method) ~ "MOFA",
      grepl("^MrVI_", method) ~ "MrVI",
      grepl("^scITD_", method) ~ "scITD",
      grepl("^GloScope_", method) ~ "GloScope",
      grepl("^GloProp", method) ~ "GloScope",
      grepl("^PILOT_", method) ~ "PILOT",
      grepl("^PILOT-GM-VAE_", method) ~ "PILOT-GM-VAE",
      grepl("^QOT_", method) ~ "QOT",
      grepl("^scPoli_", method) ~ "scPoli",
      grepl("^Pseudobulk_", method) ~ "Pseudobulk",
      TRUE ~ method # Keep any other simple names as they are
    ),
    
    method_plot = factor(method_plot, 
                         levels = c("ECODA", "Pseudobulk", "MrVI", "GloScope", "GloProp",
                                    "scPoli", "MOFA", "PILOT", "scITD", "QOT",
                                    "PILOT-GM-VAE"))
  ) %>%
  
  mutate(method = method %>%
           # Swap highres/lowres
           stringr::str_replace_all(c("highres"  = "HR", 
                                      "lowres"   = "LR",
                                      "0_dims"    = "0_embdims")) %>%
           stringr::str_replace("Pseudobulk_(\\d+)_PCA_dims", "Pseudobulk_hvg2000_pcadims\\1") %>%
           stringr::str_replace("(\\d+)_PCA_dims", "pcadims\\1") %>%
           stringr::str_replace("PILOT_hvg(\\d+)", "PILOT_hvg\\1_HR") %>%
           stringr::str_replace("QOT_hvg(\\d+)", "QOT_hvg\\1_HR") %>%
           stringr::str_replace("PILOT-GM-VAE_hvg(\\d+)", "PILOT-GM-VAE_hvg\\1_HR")
  ) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # Highlight default methods
  scale_x_discrete(labels = function(x) highlight(
    x,
    default_methods_regex,
    "black")
  ) +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             cols = vars(method_plot), # Use the new grouping column for column faceting
             labeller = labeller(score = as_labeller(score_facet_labels)))


ggsave(file.path(path_plots, "Supp_fig_2_methodparameter_screening.pdf"),
       width = 14,
       height = 8)
```

### Combined ECODA and PB

LEGACY (internal testing only, NOT shown in the publication): the
`ECODA_PB_combo_*` methods are not implemented in the benchmark pipeline
(`run_benchmark_analysis` is now deprecated — the notebook loads all R
methods from the HPC bundles), so the three chunks below would produce
nothing. Kept as documentation — see TODO.md "Ideas for later".

#### Plot distance distribution after different normalization methods applied

```{r eval=FALSE}
# LEGACY: ECODA+Pseudobulk distance-distribution demo (ECODA_PB_combo_*
# methods not in the pipeline).
ecoda <- result_list$bmark$Pelka$ECODA_authors_HR$dist_mat
pb <- result_list$bmark$Pelka$Pseudobulk_hvg2000$dist_mat

hist(ecoda)
hist(pb)

test1 <- ecoda / max(ecoda)
test2 <- pb / max(pb)
hist(test1, main = "ecoda max")
hist(test2, main = "pb max")

test1 <- ecoda / median(ecoda)
test2 <- pb / median(pb)
hist(test1, main = "ecoda median")
hist(test2, main = "pb median")

test1 <- global_quantile_norm_gaussian(ecoda)
test2 <- global_quantile_norm_gaussian(pb)
hist(test1, main = "ecoda quantiles")
hist(test2, main = "pb quantiles")

test1 <- zscore_transform(ecoda)
test2 <- zscore_transform(pb)
hist(test1, main = "ecoda zscore")
hist(test2, main = "pb zscore")
```

#### Plot benchmark including ECODA+PB (overview)

```{r fig.height=9, fig.width=7, eval=FALSE}
norm_methods <- c("max", "median", "zscore", "quantile")

filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  
  paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  "ECODA_PB_combo_normcosine_ecodaweight0.5",
  
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30"
)


df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30" = "GloScope")) %>%
  
  group_by(method, score) %>%
  mutate(mean_value = mean(value[score == "anosim_score"])) %>%
  ungroup() %>%
  
  ggplot(aes(x = fct_reorder(method, mean_value, .desc = TRUE),
             y = value, 
             fill = ifelse(grepl(default_methods_regex, method), "cornflowerblue", "gray70"))) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  labs(x = "", y = "") +
  theme_classic2(base_size = 12) +
  theme(axis.text.x = element_markdown(angle = 80, vjust = 1, hjust = 1),
  panel.grid.major.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey"), 
  panel.grid.minor.y = element_line(size = 0.25, linetype = 'solid',
                                colour = "lightgrey")) +
  scale_fill_identity() +
  # ylim(c(-0.4, 1)) +
  facet_grid(scales = "free_x",
             space='free',
             rows = vars(score),
             labeller = labeller(score = as_labeller(score_facet_labels)))
```

#### Plot benchmark including ECODA+PB (detailed)

```{r fig.height=7, fig.width=6, eval=FALSE}
filter_methods <- c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  # "ECODA_authors_LR",
  "Avg_PCA_embedding",
  "Freq_highres",
  # "ECODA_deconv",
  "Pseudobulk_10_PCA_dims",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30"
)


df_plot <- df_results %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA_authors",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "Pseudobulk_10_PCA_dims" = "Pseudobulk",
                         "Freq_highres" = "Cell_type_composition_%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30" = "GloScope"))


df_plot %>%
  filter(score == "anosim_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 80, vjust = 1, hjust = 1))



df_plot %>%
  filter(score == "cluster_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "ARI score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))




df_plot %>%
  filter(score == "mod_knn3_score") %>%
  
  ggplot(aes(x = fct_reorder(method, value, .desc = TRUE, .fun = mean), y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "ECODA_PB_combo_ecodaweight0.5",
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(min(df_plot$value), max(df_plot$value) + 0.05)) +
  labs(x = "", y = "Modularity score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
```

## Figure 4 AB: Plot feature selection

Surprisingly, even at the lowest number highly variable cell types (i.e. variance explained of 10% -\> \>=2 cell types) the average biological separation is quite very high. Different daatasets can do with a different number of highly variable cell types (separation plateaus out at different levels of variance explained % and number of cell types retained).

```{r}
df_results_sub <- df_results %>%
  # filter(score == "anosim_score") %>%
  filter(grepl("ECODA_authors_HR", method) & !grepl("HiTME|scATOMIC|PCA|NULL|most|least", method)) %>%
  filter(!grepl("cmv|sex", dataset)) %>%
  filter(!method == "ECODA_authors_HR_top_varexp0") %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "100",
                         "ECODA_authors_HR_top_varexp0" = "2most_varcts")) %>%
  mutate(method = gsub("ECODA_authors_HR", "", method)) %>%
  mutate(method = if_else(
    grepl("^_top_varexp0\\.\\d", method),
    paste0(as.integer(gsub("_top_varexp0\\.", "", method)) * 10),
    method
  )) %>%
  # mutate(method = gsub("top_varexp0", "2most_varcts", method)) %>%
  mutate(method = fct_relevel(method)) %>%
  mutate(method = factor(method, levels = rev(mixedsort(levels(method))))) %>%
  mutate(dataset = recode(dataset,
                          "GongSharma_all_age" = "GongSharma"))

df_results_anosim <- df_results_sub %>% filter(score == "anosim_score")

df_hvc40 <- df_results_anosim[df_results_anosim$method == 40 ,]
perc_tot <- c()
for (ds in df_hvc40$dataset) {
  tot <- df_results_anosim$n_features[df_results_anosim$dataset == ds & df_results_anosim$method == 100]
  perc_tot <- c(perc_tot, round(df_hvc40$n_features[df_hvc40$dataset == ds] / tot * 100))
}

cts <- df_hvc40$n_features
print(paste0("The top ", min(cts), "-", max(cts), " cell types ", 
             "(~", min(perc_tot), "-", max(perc_tot), "% of the total) ",
             "explain 40% of the datasets' variances."))

unique(df_results_sub$method)

pointsize <- 1.5

p1 <- df_results_sub %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = method, y = n_features)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  labs(x = "Variance explained (%)", y = "Number of cell types") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p1)

p2 <- df_results_sub %>%
  filter(score == "anosim_score") %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "100",  # recoded ECODA_authors_HR level (was "All cell types")
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(-0.1, 1)) +
  labs(x = "Variance explained (%)", y = "Separation score (ANOSIM)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p2)


# Combine the plots vertically
combined_plot <- (p1 / p2) + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') & 
  theme(
    plot.tag = element_text(size = 16, face = "bold"),
    legend.position = "right"
  ) &
  labs(color = "Dataset", 
       shape = "Dataset", 
       linetype = "Dataset")
print(combined_plot)
ggsave(file.path(path_plots, "Figure_4_AB_featureselectionHVC_barplot_anosim.pdf"),
       width = 5.5,
       height = 6)
```

#### Supp fig 20: HVC Mod + ARI

```{r fig.height=3.5, fig.width=8}
p3 <- df_results_sub %>%
  filter(score == "mod_knn3_score") %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "100",  # recoded ECODA_authors_HR level (was "All cell types")
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(-0.1, 1)) +
  labs(x = "Variance explained (%)", y = "Separation score (Modularity)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p3)

p4 <- df_results_sub %>%
  filter(score == "cluster_score") %>%
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(aes(shape = dataset, color = dataset), size = pointsize) + # Color added for legend consistency
  scale_shape_manual(values=1:length(unique((df_results_sub$dataset)))) +
  geom_line(aes(group=dataset, linetype=dataset, color = dataset), alpha = 1) +
  stat_compare_means(
    ref.group = "100",  # recoded ECODA_authors_HR level (was "All cell types")
    method = "wilcox.test",
    paired = TRUE,
    angle = 25,
    label.y = 0.93,
    aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  # ylim(c(-0.1, 1)) +
  labs(x = "Variance explained (%)", y = "Separation score (ARI)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1),
        legend.position = "none")
print(p4)


# Combine the plots vertically
combined_plot <- (p3 + p4) + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') & 
  theme(
    plot.tag = element_text(size = 16, face = "bold"),
    legend.position = "right"
  ) &
  labs(color = "Dataset", 
       shape = "Dataset", 
       linetype = "Dataset")
print(combined_plot)
ggsave(file.path(path_plots, "Supp_fig_20_featureselectionHVC_barplot_mod_ari.pdf"),
       width = 8,
       height = 3.5)
```

### Show example PCAs

#### Figure 4 CE: Adams

```{r fig.height=3, fig.width=7}
ds <- "Adams"

feat_mat_recalc <- result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0$feat_mat
feat_mat <- result_list$bmark[[ds]]$ECODA_authors_HR$feat_mat[, colnames(feat_mat_recalc)]

labels <- result_list$bmark[[ds]]$ECODA_authors_HR$labels
df_logr <- feat_mat_recalc[, 1, drop = FALSE] * 2

df_logr$Groups <- labels

p1 <- plot_pca(
  feat_mat = feat_mat,
  labels = labels,
  cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
  coord_equal = TRUE,
  title = "Adams",
  n_ct_show = 4, invisible = "quali", repel = TRUE, pointsize = 2
)

p2 <- ggplot(df_logr, aes(x = Groups, y = !!sym(colnames(df_logr)[1]), fill = Groups)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Standard boxplot
  geom_jitter(width = 0.2, size = 1, alpha = 0.5) + # Optional: adds points to see distribution
  theme_classic() +
  # theme_cleveland() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate labels if group names are long
  labs(
    x = NULL,
    y = paste0(
      "Log ratio\n",
      colnames(feat_mat_recalc)[1], " / ", colnames(feat_mat_recalc)[2]
    )
  ) &
  NoLegend()


wrap_plots(list(p1, p2)) +
  plot_layout(nrow = 1, widths = c(2.5, 1))
ggsave(file.path(path_plots, "Figure_4_CE_featureselectionHVC_PCA_adams.pdf"),
       width = 7,
       height = 3)
```

#### Figure 4 DF: Kfoury

```{r fig.height=3, fig.width=7}
ds <- "Kfoury"

feat_mat_recalc <- result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0$feat_mat
feat_mat <- result_list$bmark[[ds]]$ECODA_authors_HR$feat_mat[, colnames(feat_mat_recalc)]

labels <- result_list$bmark[[ds]]$ECODA_authors_HR$labels
df_logr <- feat_mat_recalc[, 1, drop = FALSE] * 2
df_logr$Groups <- labels

p1 <- plot_pca(
  feat_mat = feat_mat,
  labels = labels,
  cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
  coord_equal = TRUE,
  title = "Kfoury",
  n_ct_show = 4, invisible = "quali", repel = TRUE, pointsize = 2
)

p2 <- ggplot(df_logr, aes(x = Groups, y = !!sym(colnames(df_logr)[1]), fill = Groups)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Standard boxplot
  geom_jitter(width = 0.2, size = 1, alpha = 0.5) + # Optional: adds points to see distribution
  theme_classic() +
  # theme_cleveland() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate labels if group names are long
  labs(
    x = NULL,
    y = paste0(
      "Log ratio\n",
      colnames(feat_mat_recalc)[1], " / ", colnames(feat_mat_recalc)[2]
    )
  ) &
  NoLegend()

wrap_plots(list(p1, p2)) +
  plot_layout(nrow = 1, widths = c(2.5, 1))
ggsave(file.path(path_plots, "Figure_4_DF_featureselectionHVC_PCA_kfoury.pdf"),
       width = 7,
       height = 3)
```

## Supp fig 21

## Most highly variable cell types per dataset

```{r fig.height=15, fig.width=15}
# devtools::install_github("carmonalab/scECODA", force = TRUE)
plot_list <- list()

dss <- names(result_list[["bmark"]])
# dss <- dss[!grepl("cmv|sex", dss)]
# dss[dss == "GongSharma_all_age"] <- "GongSharma"


for (ds in dss) {
  counts <- result_list[["bmark"]][[ds]]$ECODA_authors_HR$counts
  meta <- data.frame(
    sample_id = names(result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels),
    group = result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels,
    row.names = names(result_list[["bmark"]][[ds]]$ECODA_authors_HR$labels)
  )
  ecoda_object <- ecoda(t(counts), metadata = meta, variance_explained = 0.4)
  
  if (ds == "GongSharma_all_age") {
    ec_plot <- ecoda_object
    pt <- "GongSharma"
  } else pt <- ds
  
  p <- plot_varmean(ecoda_object, plot_title = pt)
  # print(p)
  plot_list[[ds]] <- p
  # p <- plot_corr(ecoda_object)
  # print(p)
}

wrap_plots(plot_list) +
  plot_layout(nrow = 4, guides = "collect")

ggsave(file.path(path_plots, "Supp_fig_21_HVCs_per_dataset.pdf"),
       width = 20,
       height = 20)
```

# Most highly variable cell types per dataset

```{r fig.height=8, fig.width=12}
print("Most highly variable cell types per dataset:")
for (ds in names(result_list[["bmark"]])) {
  print(ds)
  print("Most highly variable cell types (HiTME annotation):\n")
  print(paste(colnames(result_list[["bmark"]][[ds]]$ECODA_HiTME_HR_layer2_top_varexp0.5$feat_mat)))
  print(paste(colnames(result_list[["bmark"]][[ds]]$ECODA_HiTME_HR_layer3_top_varexp0.5$feat_mat)))
}
```

# ---------------------------------------------------------------

# ---------------------------------------------------------------

# Plot execution times

## Unified execution times (HPC feather + bundle rows)

```{r}
# Single source of execution times, regenerated fresh on EVERY knit (no
# data/exec_times.rds cache: a stale cache would pin old rows, e.g. NA
# mem_GB for R methods computed before the workers logged peak RAM):
# 1. The NAS merged feather (${NAS_TARGET_DIR}/benchmark/embeddings/
#    execution_times.feather): HPC rows (R heavy methods, composition
#    methods, python methods, trans/zeroimp, prepare_pseudobulk_*), schema
#    dataset/method/time_secs/mem_GB (mem_GB = worker peak RSS: python
#    ru_maxrss, R /proc/self/status VmHWM).
# 2. Bundle-derived rows for methods the feather does not cover — union by
#    (dataset, method); mem_GB taken from the bundle when present.
exec_times_feather <- file.path(path_nas_embeddings, "execution_times.feather")
if (!file.exists(exec_times_feather)) {
  stop(
    "Merged execution-times feather not found: ", exec_times_feather,
    "\nRun the benchmark pipelines (they merge + sync execution_times.feather",
    " to the NAS) and make sure the NAS is mounted."
  )
}
feather_exec_times <- arrow::read_feather(exec_times_feather) %>%
  as.data.frame()

bundle_exec_times <- imap_dfr(result_list[["bmark"]], ~ {
  dataset_name <- .y
  imap_dfr(.x, function(bundle, method_name) {
    if ("exec_time" %in% names(bundle)) {
      data.frame(
        dataset = dataset_name,
        method = method_name,
        time_secs = as.numeric(bundle[["exec_time"]], units = "secs"),
        mem_GB = if ("mem_GB" %in% names(bundle)) {
          as.numeric(bundle[["mem_GB"]])
        } else {
          NA_real_
        },
        stringsAsFactors = FALSE
      )
    } else {
      NULL # no timing (e.g. methods that failed to compute)
    }
  })
})

# Normalize keys so the merge with df_results (paste0(dataset, method)
# join, helpers.R::merge_exec_times) hits:
# - GongSharma: internal key Gongsharma_cmv_young_males is displayed as
#   GongSharma (same rename as bmark_result_list above);
# - PILOT/QOT/PILOT-GM-VAE: the feather rows are <method>_hvg<n>_highres,
#   the bundles <method>_hvg<n>.
normalize_exec_keys <- function(df) {
  df %>%
    mutate(
      dataset = recode(dataset, "Gongsharma_cmv_young_males" = "GongSharma"),
      method = gsub("^(PILOT_hvg[0-9]+)_highres$", "\\1", method),
      method = gsub("^(QOT_hvg[0-9]+)_highres$", "\\1", method),
      method = gsub("^(PILOT-GM-VAE_hvg[0-9]+)_highres$", "\\1", method)
    )
}
feather_exec_times <- normalize_exec_keys(feather_exec_times)
bundle_exec_times <- normalize_exec_keys(bundle_exec_times)

# Feather rows win for the HPC methods; bundle rows fill the methods
# missing from the feather.
exec_times <- bind_rows(
  feather_exec_times,
  bundle_exec_times %>%
    anti_join(feather_exec_times, by = c("dataset", "method"))
)

# n_cells per dataset (from the metadata bundles via the dataset loop;
# display keys).
n_cells_lookup <- unlist(n_cells_by_ds)
exec_times$n_cells <- n_cells_lookup[exec_times$dataset]

# NAS feather covers all benchmark datasets; keep only this run's datasets
# (n_cells_by_ds keys are display keys, matching exec_times$dataset).
exec_times <- exec_times[exec_times$dataset %in% names(n_cells_by_ds), ]
```

## Supp fig 14: Plot

```{r}
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",

  # "Pseudobulk_hvg2000",
  "Pseudobulk_hvg2000",
  
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))

scientific_A_10n <- function(x) {
  s <- scales::scientific_format(digits = 2)(x) 
  s_converted <- gsub("e\\+", " %*% 10^", s) 
  return(parse(text = s_converted)) 
}

# --- 1. Clean Data ---
plot_data <- exec_times %>%
  filter(method %in% filter_methods) %>%
  # filter(!dataset %in% c("GongSharma_all_cmv", "GongSharma_all_sex")) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "GloScope_hvg2000_pcadims30" = "GloScope",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims15_highres" = "scPoli",
                         "PILOT_hvg2000" = "PILOT",
                         "QOT_hvg2000" = "QOT",
                         "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"))

# --- 2. Calculate the End of the Regression Line ---
line_ends <- plot_data %>%
  group_by(method) %>%
  dplyr::summarise(
    max_x = max(n_cells),
    model = list(lm(log10(time_secs) ~ n_cells)),
    pred_log_y = stats::predict(model[[1]], newdata = data.frame(n_cells = max_x))
  ) %>%
  mutate(
    pred_y = 10^pred_log_y 
  )

# --- NEW STEP: Create Ranked Color Palette ---

# A. Sort methods by predicted time (Fastest/Lowest -> Slowest/Highest)
ranked_methods <- line_ends %>%
  arrange(pred_y) %>%
  pull(method)

# B. Create a gradient function (Green -> Black -> Red)
# I used "forestgreen" and "firebrick" as they are usually more readable than neon green/red
grad_func <- colorRampPalette(c("dodgerblue3", "grey50", "red"))

# C. Generate the exact number of colors needed
method_colors <- grad_func(length(ranked_methods))

# D. Assign the method names to the colors so ggplot maps them correctly
names(method_colors) <- ranked_methods


# --- 3. Plot ---
ggplot(plot_data, aes(x = n_cells, y = time_secs, color = method)) +
  annotation_logticks(sides = "l", color = "grey80") +
  # --- The Actual Data ---
  geom_point(shape = 4, size = 1, stroke = 1.25, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 0.7) +
  
  # --- The Labels ---
  geom_text_repel(
    data = line_ends, 
    aes(x = max_x, y = pred_y, label = method), 
    size = 4,
    direction = "y",
    hjust = 0,
    nudge_x = max(plot_data$n_cells) * 0.05,
    segment.size = 0.2,
    segment.color = "grey50",
    show.legend = FALSE,
    bg.color = "white", # Optional: adds a white halo around text for readability
    bg.r = 0.15
  ) +
  
  # --- NEW: Apply the Custom Colors ---
  # scale_color_manual(values = method_colors) +
  
  # --- Theme & Scales ---
  labs(x = "Number of cells", y = "Time (s)") +
  theme_minimal() +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 3),
    labels = function(x) sapply(x, scientific_A_10n),
    limits = c(min(plot_data$n_cells), max(plot_data$n_cells) * 1.4)
  ) +
  scale_y_log10(
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))
  ) +
  theme(legend.position = "none")

ggsave(file.path(path_plots, "Supp_fig_14_benchmark_exec_times.pdf"),
       width = 3.5,
       height = 3.5, limitsize = FALSE)
```

## Supp fig 14B: Peak RAM usage (all HPC methods)

```{r fig.width=3.5, fig.height=3.5}
# Info-only figure: every HPC worker logs its process peak RSS in mem_GB
# (python: getrusage ru_maxrss; R: /proc/self/status VmHWM via
# peak_rss_gb(), incl. prepare_pseudobulk + composition rows). Rows stay NA
# only for bundles computed before 2026-08-16 or off-Linux. Same method
# recoding as Supp fig 14; log-log axes, points only.
ram_data <- exec_times %>%
  filter(!is.na(mem_GB)) %>%
  filter(method %in% analysis_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "GloScope_hvg2000_pcadims30" = "GloScope",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims15_highres" = "scPoli",
                         "PILOT_hvg2000" = "PILOT",
                         "QOT_hvg2000" = "QOT",
                         "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"))

ggplot(ram_data, aes(x = n_cells, y = mem_GB, color = method)) +
  annotation_logticks(sides = "bl", color = "grey80") +
  geom_point(shape = 4, size = 1.25, stroke = 1.25, alpha = 0.5) +
  scale_x_log10(
    breaks = scales::pretty_breaks(n = 3),
    labels = function(x) sapply(x, scientific_A_10n)
  ) +
  scale_y_log10(
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))
  ) +
  labs(x = "Number of cells", y = "Peak RAM usage (GB)") +
  theme_minimal() +
  theme(legend.position = "right")

ggsave(file.path(path_plots, "Supp_fig_14B_benchmark_mem_GB.pdf"),
       width = 3.5,
       height = 3.5, limitsize = FALSE)
```

# ---------------------------------------------------------------

# Figure 2 A: Plot benchmark results

## Main plot (with funkyheatmap)

```{r fig.height=6, fig.width=12}
# Parameterized funky-heatmap builder (src/utils/plotting.R): merges exec
# times, filters/recodes, ranks + min-max scales, aggregates and plots.
# Metrics are restricted to the notebook-level benchmark_metrics parameter
# (default: ANOSIM / Modularity / ARI).
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "Pseudobulk_hvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))

build_funky_heatmap(
  df_results = df_results,
  exec_times = exec_times,
  filter_methods = filter_methods,
  method_recode = c(
    "ECODA_authors_HR" = "ECODA",
    "ECODA_authors_HR_NULL" = "Shuffled_baseline",
    "Pseudobulk_hvg2000" = "Pseudobulk",
    "MOFA_hvg2000_factors15" = "MOFA",
    "scITD_hvg2000_factors5" = "scITD",
    "MrVI_hvg2000" = "MrVI",
    "scPoli_hvg2000_dims15_highres" = "scPoli",
    "PILOT_hvg2000" = "PILOT",
    "GloScope_hvg2000_pcadims30" = "GloScope",
    "QOT_hvg2000" = "QOT",
    "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"
  ),
  dataset_recode = dataset_display_map,
  score_set = benchmark_metrics,
  output_file = "Figure_2_A_benchmark_mainmethods.pdf",
  output_dir = path_plots,
  plot_width = 10,
  plot_height = 6
)
```

## Figure 2 B: MDS Bassez

```{r}
ds <- "Bassez"

methods <- list(
  "ECODA_authors_HR" = "ECODA",
  "GloProp" = "GloProp",
  "GloScope_hvg2000_pcadims30" = "GloScope",
  "Pseudobulk_hvg2000" = "Pseudobulk",
  "Freq_highres" = "CellType_Frequency%",
  "MrVI_hvg2000" = "MrVI",
  "MOFA_hvg2000_factors15" = "MOFA",
  "scPoli_hvg2000_dims5_highres" = "scPoli"
)
methods <- methods_in_scope_named(methods)

plot_list <- list()

for (m in names(methods)) {
  dist_mat <- result_list[["bmark"]][[ds]][[m]][["dist_mat"]]
  labels <- result_list[["bmark"]][[ds]][[m]][["labels"]]
  labels <- as.character(labels)
  labels[labels == "E"] <- "Responders"
  labels[labels == "NE"] <- "Non-responders"
  labels <- factor(labels, levels = c("Responders", "Non-responders"))
  
  plot_list[[m]] <- plot_mds(
    dist_mat,
    labels,
    cluster_score = FALSE,mod_score = FALSE, anosim_score = FALSE,
    coord_equal = TRUE, pointsize = 2,
    title = methods[[m]]
  )
}

wrap_plots(plot_list) +
  plot_layout(nrow = 2, guides = "collect")

ggsave(file.path(path_plots, "Figure_2_B_MDS_Bassez.pdf"),
       width = 10,
       height = 5)
```

## Main plot (rank-based)

```{r fig.height=6, fig.width=12}
# --- 1. SETUP & CLEANING (Same as before) ---
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  # "ECODA_HiTME_HR_layer2",
  # "ECODA_HiTME_HR_layer3",
  # "ECODA_seuratres_2",
  # "ECODA_PB_combo_normquantile_ecodaweight0.5",
  # paste0("ECODA_PB_combo_norm", norm_methods, "_ecodaweight0.5"),
  # "ECODA_authors_LR",
  # "Avg_PCA_embedding",
  # "Freq_highres",
  # "ECODA_deconv",
  # "Pseudobulk_10_PCA_dims",
  "Pseudobulk_hvg2000",
  # "Pseudobulk_hvg2000_bl",
  # "Pseudobulk_schvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims5_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))

df_results_bmark <- merge_exec_times(df_results, exec_times)

clean_data <- df_results_bmark %>%
  filter(method %in% filter_methods) %>%
  mutate(method = recode(method,
                         "ECODA_authors_HR" = "ECODA",
                         "ECODA_authors_HR_NULL" = "Shuffled_baseline",
                         # "ECODA_HiTME_HR_layer2" = "ECODA_automated",
                         # "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
                         # "ECODA_seuratres_2" = "ECODA_unsupervised",
                         "ECODA_PB_combo_normquantile_ecodaweight0.5" = "ECODA+Pseudobulk",
                         # "Pseudobulk_10_PCA_dims" = "Pseudobulk_PCA",
                         "Pseudobulk_hvg2000" = "Pseudobulk",
                         # "Pseudobulk_schvg2000" = "Pseudobulk_schvg",
                         # "Freq_highres" = "ECODA_Freq%",
                         "MOFA_hvg2000_factors15" = "MOFA",
                         "scITD_hvg2000_factors5" = "scITD",
                         "MrVI_hvg2000" = "MrVI",
                         "scPoli_hvg2000_dims5_highres" = "scPoli",
                         "scITD_2_factors" = "scITD",
                         "PILOT_hvg2000" = "PILOT",
                         "GloScope_hvg2000_pcadims30" = "GloScope",
                         "QOT_hvg2000" = "QOT",
                         "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE")) %>%
  mutate(score = recode(score, !!!score_label_map)) %>%
  mutate(dataset = recode(dataset,
                          "GongSharma_all_age" = "GongSharma_age",
                          "GongSharma_all_cmv" = "GongSharma_cmv",
                          "GongSharma_all_sex" = "GongSharma_sex"))

# --- 1. TRANSFORMATION TO RANKS ---
df_ranked <- clean_data %>%
  group_by(dataset, score) %>%
  mutate(rank = rank(-value, ties.method = "average")) %>%
  ungroup()

# --- 2. AGGREGATE RANK-BASED SCORES ---
df_overall <- df_ranked %>%
  group_by(method) %>%
  dplyr::summarise(Overall_Rank = mean(rank, na.rm = TRUE))

df_datasets_wide <- df_ranked %>%
  group_by(method, dataset) %>%
  dplyr::summarise(avg_rank = mean(rank, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(id_cols = method, names_from = dataset, 
              values_from = avg_rank, names_prefix = "Dataset__")

target_metric_order <- c("ANOSIM", "Modularity", "ARI")

df_metrics_wide <- df_ranked %>%
  mutate(score = factor(score, levels = target_metric_order)) %>%
  group_by(method, score) %>%
  dplyr::summarise(avg_rank = mean(rank, na.rm = TRUE), .groups = "drop") %>%
  arrange(score) %>%
  pivot_wider(id_cols = method, names_from = score, 
              values_from = avg_rank, names_prefix = "Metric__")

# --- 3. MERGE, INVERT, AND SCALE (Fixes the Error) ---
final_data <- df_overall %>%
  left_join(df_datasets_wide, by = "method") %>%
  left_join(df_metrics_wide, by = "method") %>%
  # Add mandatory 'id' column for funkyheatmap
  mutate(id = method)

# Identify all rank-based columns
rank_cols <- c("Overall_Rank", colnames(final_data)[startsWith(colnames(final_data), "Dataset__")], 
               colnames(final_data)[startsWith(colnames(final_data), "Metric__")])

# Instead of using min(x) and max(x) which vary per column,
# we use the theoretical best (1) and worst (total methods).
num_methods <- length(unique(final_data$method))

scale_ranks_global <- function(x, n_meth) {
  if(all(is.na(x))) return(x)
  # BEST rank (1) should map to 1.0
  # WORST rank (n_meth) should map to 0.0
  # Formula: (Worst - Current) / (Worst - Best)
  (n_meth - x) / (n_meth - 1)
}

final_data <- final_data %>%
  mutate(across(all_of(rank_cols), ~scale_ranks_global(.x, num_methods))) %>%
  arrange(desc(Overall_Rank))

# --- 4. COLUMN INFO CONFIGURATION ---
info_method <- tribble(
  ~id,      ~group, ~name,    ~geom, ~palette, ~options,
  "id",     "",     "",       "text", NA,       list(hjust = 0, width = 6)
)

info_overall <- tribble(
  ~id,            ~group,   ~name,      ~geom, ~palette,        ~options,
  "Overall_Rank", "Overall", "Avg Rank", "bar",  "palette_ranks", list(width = 4)
)

dataset_cols <- colnames(final_data)[startsWith(colnames(final_data), "Dataset__")]
info_datasets <- tibble(id = dataset_cols) %>%
  mutate(
    group = "Avg Rank per Dataset",
    name = str_remove(id, "Dataset__"), 
    geom = "circle",
    palette = "palette_ranks",
    options = list(list())
  )

metric_cols <- colnames(final_data)[startsWith(colnames(final_data), "Metric__")]
info_metrics <- tibble(id = metric_cols) %>%
  mutate(
    group = "Avg Rank per Metric",
    name = str_remove(id, "Metric__"),
    name_f = factor(name, levels = target_metric_order),
    geom = "bar",
    palette = "palette_ranks",
    options = list(list(width = 2))
  ) %>% 
  arrange(name_f) %>% 
  dplyr::select(-name_f)

column_info <- bind_rows(info_method, info_overall, info_datasets, info_metrics)

# --- 5. PLOT ---
ncolors <- 11
funky_heatmap(
  data = final_data,
  column_info = column_info,
  scale_column = FALSE, # Values are already 0-1
  palettes = list(
    palette_ranks = brewer.pal(n = ncolors, name = "RdBu")
  ),
  column_groups = tibble(
    group = c("Overall", "Avg Rank per Dataset", "Avg Rank per Metric")
  ),
  legends = list(
    list(palette = "palette_ranks", title = "Rank", labels = c(".       Worst", rep("", round((ncolors-2)/2)+1), "", rep("", round((ncolors-2)/2)), "Best    ."))
  )
)
ggsave(file.path(path_plots, "benchmark_summary_funkyheatmap_mainmethods_ranks.pdf"),
       width = 10,
       height = 5)
```

## Supp fig 15: Extended benchmark (with avg rank)

```{r fig.height=8, fig.width=14}
# Parameterized funky-heatmap builder (see Figure 2A chunk): extended method
# set + Rank_Score column (include_rank = TRUE).
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_scATOMIC_HR",
  "ECODA_seuratres_2",
  "Freq_highres",
  "Pseudobulk_hvg2000",
  "Pseudobulk_CT_HR_hvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))

build_funky_heatmap(
  df_results = df_results,
  exec_times = exec_times,
  filter_methods = filter_methods,
  method_recode = c(
    "ECODA_authors_HR" = "ECODA_authors",
    "ECODA_authors_HR_NULL" = "Shuffled_baseline",
    "ECODA_HiTME_HR_layer2" = "ECODA_automated_HiTME",
    "ECODA_HiTME_HR_layer3" = "ECODA_automated_(HiTME_layer3)",
    "ECODA_scATOMIC_HR" = "ECODA_automated_scATOMIC",
    "ECODA_seuratres_2" = "ECODA_Leiden_res_2",
    "Pseudobulk_10_PCA_dims" = "Pseudobulk_PCA",
    "Pseudobulk_hvg2000" = "Pseudobulk",
    "Pseudobulk_CT_HR_hvg2000" = "Pseudobulk_CT_HR",
    "Freq_highres" = "CellType_Frequency%",
    "MOFA_hvg2000_factors15" = "MOFA",
    "scITD_hvg2000_factors5" = "scITD",
    "MrVI_hvg2000" = "MrVI",
    "scPoli_hvg2000_dims15_highres" = "scPoli",
    "PILOT_hvg2000" = "PILOT",
    "GloScope_hvg2000_pcadims30" = "GloScope",
    "QOT_hvg2000" = "QOT",
    "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"
  ),
  dataset_recode = dataset_display_map,
  score_set = benchmark_metrics,
  output_file = "Supp_fig_15_benchmark_extended.pdf",
  output_dir = path_plots,
  plot_width = 14,
  plot_height = 8,
  method_width = 8,
  overall_col_name = "Metrics Mean",
  include_rank = TRUE,
  overall_legend_title = "Overall",
  overall_legend_labels = c(".         Worst", rep("", 9), "Best       .")
)
```
```

## Difference between Pseudobulk_hvg2000 and Pseudobulk_CT_HR

```{r}
pb2000_scores <- list()
pb_ct_hr_2000_scores <- list()
pb_ct_hr_500_scores <- list()

for (ds in names(result_list[["bmark"]])) {
  pb2000_scores[[ds]] <- result_list[["bmark"]][[ds]][["Pseudobulk_hvg2000"]]$scores
  pb_ct_hr_2000_scores[[ds]] <- result_list[["bmark"]][[ds]][["Pseudobulk_CT_HR_hvg2000"]]$scores
  pb_ct_hr_500_scores[[ds]] <- result_list[["bmark"]][[ds]][["Pseudobulk_CT_HR_hvg500"]]$scores
}


# Define a function to get the global average excluding 'mod_score'
get_global_avg <- function(score_list) {
  # 1. Calculate the mean for each dataset
  ds_means <- sapply(score_list, function(ds) {
    # Filter out 'mod_score' by name
    relevant_vals <- unlist(ds[names(ds) != "mod_score"])
    # Return the mean of remaining scores for this dataset
    mean(relevant_vals, na.rm = TRUE)
  })
  
  # 2. Return the average of all dataset means
  return(mean(ds_means, na.rm = TRUE))
}

# Apply to your three result objects
avg_pb2000   <- get_global_avg(pb2000_scores)
avg_ct_2000 <- get_global_avg(pb_ct_hr_2000_scores)
avg_ct_500  <- get_global_avg(pb_ct_hr_500_scores)

cat("Pseudobulk_CT_HR_hvg2000 vs Pseudobulk_hvg2000: ", round(avg_ct_2000 / avg_pb2000, 2), "\n")
cat("Pseudobulk_CT_HR_hvg500 vs Pseudobulk_hvg2000: ", round(avg_ct_500 / avg_pb2000, 2))
```

## For presentation: Extended benchmark

```{r fig.height=8, fig.width=14}
# Parameterized funky-heatmap builder (see Figure 2A chunk): presentation
# variant of Supp fig 15 (no scATOMIC / Freq_highres columns).
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "ECODA_HiTME_HR_layer2",
  "ECODA_seuratres_2",
  "Pseudobulk_hvg2000",
  "Pseudobulk_CT_HR_hvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))

build_funky_heatmap(
  df_results = df_results,
  exec_times = exec_times,
  filter_methods = filter_methods,
  method_recode = c(
    "ECODA_authors_HR" = "ECODA_authors",
    "ECODA_authors_HR_NULL" = "Shuffled_baseline",
    "ECODA_HiTME_HR_layer2" = "ECODA_automated",
    "ECODA_scATOMIC_HR" = "ECODA_automated_scATOMIC",
    "ECODA_seuratres_2" = "ECODA_unsupervised",
    "Pseudobulk_10_PCA_dims" = "Pseudobulk_PCA",
    "Pseudobulk_hvg2000" = "Pseudobulk",
    "Pseudobulk_CT_HR_hvg2000" = "Pseudobulk_CT_HR",
    "Freq_highres" = "CellType_Frequency%",
    "MOFA_hvg2000_factors15" = "MOFA",
    "scITD_hvg2000_factors5" = "scITD",
    "MrVI_hvg2000" = "MrVI",
    "scPoli_hvg2000_dims15_highres" = "scPoli",
    "PILOT_hvg2000" = "PILOT",
    "GloScope_hvg2000_pcadims30" = "GloScope",
    "QOT_hvg2000" = "QOT",
    "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"
  ),
  dataset_recode = dataset_display_map,
  score_set = benchmark_metrics,
  output_file = "Presentation_benchmark_extended.pdf",
  output_dir = path_plots,
  plot_width = 14,
  plot_height = 8,
  method_width = 8,
  overall_col_name = "Metrics Mean",
  include_rank = TRUE,
  overall_legend_title = "Overall",
  overall_legend_labels = c(".         Worst", rep("", 9), "Best       .")
)
```

# ---------------------------------------------------------------

# Supp fig 1: Correlation between metrics

```{r}
# --- 0. CLEAN DATA (self-contained: same filter/recode as Figure 2A, built
# locally so this chunk has no cross-chunk dependency on clean_data). All
# score columns are kept (no score_set restriction) — the correlation matrix
# below includes Silhouette and LISI.
filter_methods <- methods_in_scope(c(
  "ECODA_authors_HR",
  "ECODA_authors_HR_NULL",
  "Pseudobulk_hvg2000",
  "MOFA_hvg2000_factors15",
  "scITD_hvg2000_factors5",
  "MrVI_hvg2000",
  "scPoli_hvg2000_dims15_highres",
  "PILOT_hvg2000",
  "GloScope_hvg2000_pcadims30",
  "GloProp",
  "QOT_hvg2000",
  "PILOT-GM-VAE_hvg2000"
))

clean_data <- prepare_benchmark_data(
  df_results = df_results,
  exec_times = exec_times,
  filter_methods = filter_methods,
  method_recode = c(
    "ECODA_authors_HR" = "ECODA",
    "ECODA_authors_HR_NULL" = "Shuffled_baseline",
    "Pseudobulk_hvg2000" = "Pseudobulk",
    "MOFA_hvg2000_factors15" = "MOFA",
    "scITD_hvg2000_factors5" = "scITD",
    "MrVI_hvg2000" = "MrVI",
    "scPoli_hvg2000_dims15_highres" = "scPoli",
    "PILOT_hvg2000" = "PILOT",
    "GloScope_hvg2000_pcadims30" = "GloScope",
    "QOT_hvg2000" = "QOT",
    "PILOT-GM-VAE_hvg2000" = "PILOT-GM-VAE"
  ),
  dataset_recode = dataset_display_map
)

# --- 1. PREPARE WIDE DATA ---
# Dynamically calculate the mean based on whatever metrics are present
metrics <- c("ANOSIM", "Modularity", "ARI", "Silhouette", "LISI")

df_wide <- clean_data %>%
  pivot_wider(id_cols = c(dataset, method), 
              names_from = score, 
              values_from = value) %>%
  mutate(Mean_Score = rowMeans(dplyr::select(., all_of(metrics)), na.rm = TRUE))

# --- 2. REUSABLE PLOTTING FUNCTION ---
# Updated to use tidy evaluation (.data[[]]) instead of the deprecated aes_string()
plot_mean_corr <- function(data, metric_var) {
  # Calculate limits dynamically
  lims <- c(min(data %>% dplyr::select(all_of(metrics)), na.rm = TRUE) - 0.05, 1.2)
  
  ggplot(data, aes(x = Mean_Score, y = .data[[metric_var]])) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50", linewidth = 0.8) +
    geom_point(alpha = 0.4, color = "#000000", size = 1.5) +
    geom_smooth(method = "lm", color = "black", linewidth = 1, se = FALSE) +
    stat_cor(aes(label = ..rr.label..), label.x.npc = "left", label.y.npc = "top") +
    labs(x = "Mean Score", y = metric_var) +
    coord_cartesian(xlim = lims, ylim = lims) + # Better than xlim()/ylim() to prevent dropping data in smooths
    theme_classic()
}

# --- 3. DYNAMICALLY GENERATE "METRIC VS MEAN" PLOTS ---
# purrr::map iterates over your metrics and creates a list of plots
mean_plots <- map(metrics, ~plot_mean_corr(df_wide, .x))

# Combine them using patchwork's wrap_plots
mean_row <- wrap_plots(mean_plots, nrow = 1)

# --- 4. PAIRWISE CORRELATION MATRIX ---
# ggpairs handles the "all vs all" automatically
pairwise_matrix <- ggpairs(
  df_wide, 
  columns = metrics,
  lower = list(continuous = wrap("smooth", alpha = 0.4, color = "black")),
  diag = list(continuous = wrap("densityDiag", fill = "grey80"))
) + theme_classic()

# --- 5. COMBINE AND DISPLAY ---
# Note: ggpairs returns a ggmatrix, which behaves slightly differently with patchwork, 
# so saving them as two separate panels or using cowplot/gridExtra might be necessary 
# depending on your exact layout needs. 

# Save the pairwise matrix
ggsave(file.path(path_plots, "Supp_fig_1A_pairwise.pdf"), pairwise_matrix, width = 7, height = 7)

# Save the mean row
ggsave(file.path(path_plots, "Supp_fig_1B_means.pdf"), mean_row, width = 10, height = 3.5)
```

# ---------------------------------------------------------------

# Supp fig 16 A: Plot normalization results

## Plot results

```{r}
if (length(result_list[["trans"]]) == 0) {
  stop("No transformation-analysis results found (result_list$trans is empty).",
       "\nRun the HPC trans pipeline first (run_transformation_zeroimp_analysis/).")
}

# Per-dataset trans bundles are summarized data.frames without a dataset
# column; add it here so the paired lines/points below can group by dataset.
df_results <- imap_dfr(result_list[["trans"]], ~ mutate(.x, dataset = .y)) %>%
  mutate(
    trans_method = recode(
      trans_method,
      "clr" = "CLR",
      "alr_mincvref" = "ALR_minCVref",
      "alr_randref" = "ALR_randref",
      "arcsine_sqrt" = "Arcsine_sqrt",
      "counts" = "Counts",
      "freq" = "Frequency%"
    )
  )

df_results <- df_results %>%
  pivot_longer(
    cols = c(ANOSIM_score, Modularity_score, Adjusted_Rand_Index),
    names_to = "score",
    values_to = "value"
  ) %>%
  as.data.frame()


df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score") %>%
  group_by(trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_anosim <- df_results_anosim %>%
  arrange(desc(mean))

df_results_anosim$trans_method <- factor(
  df_results_anosim$trans_method,
  levels = df_results_anosim$trans_method
)
df_results$trans_method <- factor(
  df_results$trans_method,
  levels = df_results_anosim$trans_method
)


df_results_list[["Normalization"]] <- df_results


p1 <- df_results %>%
  filter(score == "ANOSIM_score") %>%

  ggplot(aes(x = trans_method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +

  stat_compare_means(
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..))),
    aes(label = after_stat(p.signif)),
    size = 3,
    label.y = 0.95
  ) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  ylim(c(-0.1, 1))
print(p1)
ggsave(
  file.path(path_plots, "Supp_fig_16A_transformations_anosim.pdf"),
  width = 2.5,
  height = 3.5
)


df_results %>%
  group_by(score, trans_method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%

  ggplot(aes(x = trans_method, y = mean)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(
    aes(ymin = mean - sem, ymax = mean + sem),
    width = 0.2,
    colour = "black",
    alpha = 0.9
  ) +
  labs(x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust = 1)) +
  facet_grid(
    scales = "free",
    rows = vars(score),
    labeller = labeller(score = as_labeller(score_facet_labels))
  )
ggsave(
  file.path(
    path_plots,
    "transformations_real_datasets_summary_barplot_separatescores_bars.pdf"
  ),
  width = 7,
  height = 7
)


my_comparisons <- list(
  c("CLR", "Frequency(%)"),
  c("CLR", "Counts")
)

df_results %>%
  filter(score == "ANOSIM_score") %>%
  filter(trans_method %in% c("Counts", "Frequency(%)", "CLR")) %>%

  ggplot(aes(
    x = fct_reorder(trans_method, value, .desc = FALSE, .fun = mean),
    y = value,
    fill = ifelse(trans_method == "CLR", "lightgreen", "gray70")
  )) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar") +
  geom_point(position = "identity", alpha = 0.6, size = 0.7) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.1) +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    paired = TRUE,
    size = 3,
    tip.length = 0,
    # label.y = 0.95,
    aes(label = after_stat(p.signif))
  ) +
  labs(x = "", y = "Separation of biological groups\n(ANOSIM score)") +
  theme_classic2(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ylim(c(-0.05, 1)) +
  scale_fill_identity() +
  scale_x_discrete(
    labels = c(
      "Counts",
      "Frequency(%)",
      expression(bold("CLR")) # Make label bold
    )
  )
ggsave(
  file.path(
    path_plots,
    "POSTER_transformations_real_datasets_summary_barplot_anosim_refCLR.pdf"
  ),
  width = 2,
  height = 3.5
)


ggplot(df_results, aes(x = trans_method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = dataset)) +
  geom_line(
    aes(group = dataset, linetype = dataset),
    alpha = 1,
    color = "steelblue"
  ) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  facet_grid(
    scales = "free",
    rows = vars(score),
    labeller = labeller(score = as_labeller(score_facet_labels))
  )

ggsave(
  file.path(
    path_plots,
    "transformations_real_datasets_summary_barplot_separatescores.pdf"
  ),
  width = 5,
  height = 5
)

df_results_anosim <- df_results %>%
  filter(score == "ANOSIM_score")

df_results_anosim <- df_results_anosim[
  !grepl("pca", df_results_anosim$trans_method),
]

my_comparisons <- purrr::map(
  unique(as.character(df_results_anosim$trans_method)),
  ~ c(.x, "Frequency(%)")
)
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
# my_comparisons <- c(my_comparisons, list(c("clr", "arcsine_sqrt")))
my_comparisons <- my_comparisons[c(4, 3, 2, 1, 5)]


ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    # ref.group = "freq",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = dataset)) +
  scale_shape_manual(values = 1:length(unique((df_results_anosim$dataset)))) +
  geom_line(
    aes(group = dataset, linetype = dataset),
    alpha = 1,
    color = "steelblue"
  ) +
  labs(x = "", y = "ANOSIM score") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggsave(
  file.path(path_plots, "transformations_real_datasets_summary_barplot.pdf"),
  width = 6,
  height = 4.5
)


ggplot(df_results_anosim, aes(x = trans_method, y = value)) +
  stat_compare_means(
    # comparisons = my_comparisons,
    # p.adjust.method = "bonferroni",
    ref.group = "CLR",
    method = "wilcox.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape = dataset)) +
  scale_shape_manual(values = 1:length(unique((df_results_anosim$dataset)))) +
  geom_line(
    aes(group = dataset, linetype = dataset),
    alpha = 1,
    color = "steelblue"
  ) +
  labs(x = "", y = "ANOSIM score") +
  # stat_pvalue_manual(stat.test, label = "p.adj",
  #                    y.position = c(max(df_results_anosim$values)+0.1, max(df_results_anosim$values)+0.2), hide.ns = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# ---------------------------------------------------------------

# Supp fig 16 B: Plot zero imputation results

## Plot results

```{r}
zeroimp_result_list <- result_list[["zeroimp"]]
if (length(zeroimp_result_list) == 0) {
  stop("No zero-imputation results found (result_list$zeroimp is empty).",
       "\nRun the HPC zeroimp pipeline first (run_transformation_zeroimp_analysis/).")
}

df_results <- imap_dfr(zeroimp_result_list, ~ {
  dataset_name <- .y
  imap_dfr(.x, function(method_value, method_name) {
    if (is.list(method_value) && !is.data.frame(method_value)) {
      if (length(method_value) == 1 && is.list(method_value[[1]])) {
        method_value <- method_value[[1]]
      }
      if (!all(sapply(method_value, is.numeric))) {
        message(
          "Dropping non-numeric zeroimp rows for ", dataset_name, "/",
          method_name
        )
        return(tibble())
      }
    }
    imap_dfr(method_value, ~ tibble(
      dataset = dataset_name,
      method = method_name,
      score = .y,
      value = as.numeric(.x)
    ))
  })
})

df_results <- df_results[!grepl("impRZilr", df_results$method), ]
df_results <- df_results[!df_results$score == "mod_score", ]

df_results_list[["Zero_imputation"]] <- df_results

df_results_anosim <- df_results %>%
  filter(score == "anosim_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value)))

df_results_anosim <- df_results_anosim %>%
  arrange(desc(mean))

df_results_anosim$method <- factor(df_results_anosim$method, levels = df_results_anosim$method)
df_results$method <- factor(df_results$method, levels = df_results_anosim$method)



p2 <- df_results %>%
  filter(score == "anosim_score") %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.95,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 1)) +
  labs(x = "", y = "Separation score (ANOSIM)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))
print(p2)
ggsave(file.path(path_plots, "Supp_fig_16B_zero_replacement_anosim.pdf"),
       width = 2.5,
       height = 4)



df_results %>%
  filter(score == "mod_knn3_score") %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.9,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 0.9)) +
  labs(x = "", y = "Separation score (Modularity)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))



df_results %>%
  filter(score == "cluster_score") %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 0.9,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.1, 0.9)) +
  labs(x = "", y = "Separation score (ARI)") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1))


df_results %>%
  filter(method != "asinsqrt") %>%
  filter(!grepl("perc", method)) %>%
  
  ggplot(aes(x = method, y = value)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "bar", fill = "white", color = "black") +
  geom_point(position = "identity", alpha = 0.6, size = 0.8) +
  geom_line(aes(group = dataset), color = "grey20", alpha = 0.2) +
  stat_compare_means(
    ref.group = "counts_all_0.5",
    method = "wilcox.test",
    paired = TRUE,
    # angle = 25,
    label.y = 1.1,
    size = 3,
    aes(label = after_stat(p.signif))
    # aes(label = sprintf("%5.2f", as.numeric(..p.format..)))
  ) +
  ylim(c(-0.2, 1.2)) +
  labs(x = "", y = "Separation score") +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust = 1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = as_labeller(score_facet_labels))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_separatescores.pdf"),
       width = 3,
       height = 8)



df_results %>%
  group_by(score, method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  ungroup() %>%

  ggplot(aes(x = method, y = mean)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 65, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = as_labeller(score_facet_labels)))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_barplot_separatescores.pdf"),
       width = 3.5,
       height = 6)

ggplot(df_results, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Separation score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_grid(scales = "free",
             rows = vars(score),
             labeller = labeller(score = as_labeller(score_facet_labels)))

ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_separatescores.pdf"),
       width = 7,
       height = 6)



df_results %>%
  filter(score == "cluster_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results %>%
  filter(score == "anosim_score") %>%
  group_by(method) %>%
  summarize(mean = mean(value), sem = sd(value) / sqrt(length(value))) %>%
  
  ggplot(aes(x = method, y = mean)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  # stat_summary(fun.y = mean, geom = "bar") +
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=0.2, colour="black", alpha=0.9) +
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  NoLegend()


df_results_clustering <- df_results %>%
  filter(score == "cluster_score")

my_comparisons <- purrr::map(unique(as.character(df_results_clustering$method)), ~ c(.x, "counts_all_1"))
my_comparisons <- Filter(function(x) length(unique(x)) == 2, my_comparisons)
my_comparisons <- my_comparisons[c(6,1,7:10, 2:5)]
# my_comparisons[[length(my_comparisons) + 1]] <- c("perc_all_0.1%", "perc_all_1%")

p <- ggplot(df_results_clustering, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "Adjusted Rand Index") +
  scale_shape_manual(values=1:length(unique((df_results_clustering$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_clusterscore.pdf"),
       width = 8,
       height = 6)

# (removed: mod_score boxplot — mod_score rows are dropped above, so the
# plot was always empty)

df_results_anosim <- df_results %>%
  filter(score == "anosim_score")

p <- ggplot(df_results_anosim, aes(x = method, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_point(aes(shape=dataset)) + 
  geom_line(aes(group=dataset, linetype=dataset), alpha = 1, color = "steelblue") +
  stat_compare_means(
    comparisons = my_comparisons,
    method = "t.test",
    paired = TRUE,
    # aes(label = sprintf("p = %5.2f", as.numeric(..p.format..)))
    aes(label = after_stat(p.signif))
  ) + # Add pairwise comparisons p-value
  # geom_point(position = position_jitterdodge()) +
  labs(x = "", y = "ANOSIM score") +
  scale_shape_manual(values=1:length(unique((df_results_anosim$dataset)))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(p)
ggsave(file.path(path_plots, "zero_imputation_real_datasets_summary_boxplot_anosimscore.pdf"),
       width = 8,
       height = 6)

```

# Supp fig 16AB plot

```{r}
wrap_plots(list(p1, p2)) +
  plot_layout(nrow = 1) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold"))
ggsave(file.path(path_plots, "Supp_fig_16AB_trans_zero_anosim.pdf"),
       width = 6,
       height = 4.5)
```

# ---------------------------------------------------------------

# Export tables with raw data for paper

```{r}
write_xlsx(df_results_list, path = file.path(path_data, "raw_data_for_publication.xlsx"))
```

# ---------------------------------------------------------------

# Info about datasets

```{r}
for (ds in names(result_list$bmark)) {
  print("-----------------------------------------")
  print(ds)
  feat_mat <- result_list$bmark[[ds]]$ECODA_authors_HR$feat_mat
  print(paste("Number of cell types:", ncol(feat_mat)))
  print(paste("Number of samples:", nrow(feat_mat)))
  # n_cells was collected during the dataset loop (display keys, e.g.
  # GongSharma for Gongsharma_cmv_young_males).
  ds_display <- if (ds == "Gongsharma_cmv_young_males") "GongSharma" else ds
  print(paste("Number of cells:", n_cells_by_ds[[ds_display]]))
  
  print(paste("ANOSIM using all cell types:", round(result_list$bmark[[ds]]$ECODA_authors_HR$scores$anosim_score, 3)))
  print(paste("ANOSIM at 10% HVC:", round(result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0.1$scores$anosim_score, 3)))
  print(paste("Number of HVC at 10% HVC:", ncol(result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0.1$feat_mat)))
  cat("Most highly variable cell types:", colnames(result_list$bmark[[ds]]$ECODA_authors_HR_top_varexp0$feat_mat), "\n")
  
  cat("Cell types:", colnames(feat_mat), "\n")
}
```

# Supp table 1:

## Number of cells, total, per sample

```{r}
# Computed from the per-dataset stats collected during the dataset loop
# (backed obs reads; no full reloads).
for (i in seq_len(nrow(ds_stats_df))) {
  print("-----------------------------------------")
  print(ds_stats_df$dataset[i])
  print(paste("Number of samples:", ds_stats_df$n_samples[i]))
  print(paste("Number of cells:", ds_stats_df$n_cells[i]))
  print(paste("Number of cells per sample:", round(ds_stats_df$cells_per_sample_mean[i]), "+-", round(ds_stats_df$cells_per_sample_sd[i])))
  print(paste("Number of cell types:", ds_stats_df$n_cell_types[i]))
}
```


# ---------------------------------------------------------------

